### 2015-06-02 Appendices/AppendixCPETRawData.tex
Latex files added for first time

% Appendix CPET Raw Data

\chapter{Breath-by-breath CPET sample data} % Main appendix title

\label{AppendixCPETRawData} % For referencing this appendix elsewhere, use \ref{AppendixA}

\lhead{Appendix C. \emph{Raw CPET Data}} % This is for the header on each page - perhaps a shortened title
\begin{figure}
	\centering
	\caption{Breath-by-breath sample data with values averaged every 10 seconds - Part 1.}
	\includegraphics[width=\textwidth,page=1]{Figures/cpet_9p_rawdata.pdf}
\end{figure}
\begin{figure}
	\centering
	\caption{Breath-by-breath sample data with values averaged every 10 seconds - Part 2.}
	\includegraphics[width=\textwidth,page=2]{Figures/cpet_9p_rawdata.pdf}
\end{figure}

### 2015-06-02 Appendices/AppendixExcelMacros.tex
Latex files added for first time

% Appendix A

\chapter{Visual Basic for Applications For CPET Analysis} % Main appendix title

\label{AppendixExcelMacros} % For referencing this appendix elsewhere, use \ref{AppendixA}

\lhead{Appendix B. \emph{Excel VBA For CPET Analysis}} % This is for the header on each page - perhaps a shortened title

\setstretch{1.0}
The following VBA code was written by me to facilitate detailed analysis of large volumes of raw CPET data that would not have been otherwise available from the final report generated for clinical use. Where necessary, I have used information from the MSDN knowledgebase and internet fora to supplement my knowledge of coding.

\begin{lstlisting}
Option Explicit
Sub RenameSortSheets() 
   Dim ws As Worksheet
   Dim strName() As String

   For Each ws In Worksheets
     If ws.Range("$B$1").Value = "" Then
       Exit Sub
     End If
     If ws.Name <> "aaa_main" Then
       strName = Split(CStr(ws.Range("$B$1").Value), "(")
       ws.Name = LCase(strName(0))
       ws.Activate
       ws.Range("B4").Select
       ActiveWindow.FreezePanes = True
     End If
   Next ws

   Dim N As Integer
   Dim M As Integer
   Dim FirstWSToSort As Integer
   Dim LastWSToSort As Integer
   Dim SortDescending As Boolean

   SortDescending = False

   If ActiveWindow.SelectedSheets.Count = 1 Then
     'Change the 1 to the worksheet you want sorted first
     FirstWSToSort = 1
     LastWSToSort = Worksheets.Count
    Else
     With ActiveWindow.SelectedSheets
       For N = 2 To .Count
         If .Item(N - 1).Index <> .Item(N).Index - 1 Then
           MsgBox "You cannot sort non-adjacent sheets"
           Exit Sub
         End If
       Next N

       FirstWSToSort = .Item(1).Index
       LastWSToSort = .Item(.Count).Index
     End With
   End If

   For M = FirstWSToSort To LastWSToSort
     For N = M To LastWSToSort
       If SortDescending = True Then
         If UCase(Worksheets(N).Name) > UCase(Worksheets(M).Name) Then
           Worksheets(N).Move Before:=Worksheets(M)
         End If
       Else
         If UCase(Worksheets(N).Name) < UCase(Worksheets(M).Name) Then
           Worksheets(N).Move Before:=Worksheets(M)
         End If
       End If
     Next N
   Next M

   Dim x As Integer
   x = 2
   Sheets("aaa_main").Range("A:A").Clear
   Sheets("aaa_main").Cells(1, 1) = "SheetName"
   For Each ws In Worksheets
     Sheets("aaa_main").Cells(x, 1) = ws.Name
     x = x + 1
   Next ws
End Sub

Sub GotoSheet()
  If ActiveWindow.ActiveSheet.Name = "aaa_main" Then
    Dim selName As String
    selName = ActiveWindow.ActiveCell.Value
    Worksheets(selName).Activate
  Else
    Worksheets("aaa_main").Activate
  End If
End Sub

Sub VCO2VO2()
  Dim ws As Worksheet
  For Each ws In Worksheets
    Dim x As Integer
    x = 4
    While ws.Cells(x, 2) > 0
      If ws.Cells(x, 9) <> "-" And ws.Cells(x, 6) <> "-" Then
        ws.Cells(x, 22) = CDbl(ws.Cells(x, 9)) / CDbl(ws.Cells(x, 6))
      End If
      x = x + 1
    Wend
  Next ws  
End Sub

Sub ClearColumn()
  Dim ws As Worksheet
  For Each ws In Worksheets
    ws.Range("AD9") = ws.Range("A1")
    ws.Range("A1").Clear
  Next ws
End Sub

Sub DeleteAllCharts()
  Dim ws As Worksheet
  Dim chart As ChartObject
  For Each ws In Worksheets
    For Each chart In ws.ChartObjects
      chart.Delete
    Next chart
  Next ws
End Sub

Sub AddRCValues()
  Dim x As Integer
  For x = 3 To 88
    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD8") = _
      Sheets("aaa_main").Cells(x, 27).Value
    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD9") = _
      Sheets("aaa_main").Cells(x, 28).Value
  Next x
End Sub

Sub CopyLineForSPSS()
  Dim rng As Range
  Dim selrow As Integer
  Dim selrng As String  
  selrow = ActiveWindow.ActiveCell.Row
  selrng = "AF" & CStr(selrow) & ":BX" & CStr(selrow)
  Range(selrng).Select
  Selection.Copy
End Sub

%----------------------------------
Sub ValuesAT()
  Dim r As Integer
  r = 3 - Selection.Cells(1, 1).Row
  ActiveSheet.Range("X29") = Selection.Cells(r, 1)
  ActiveSheet.Range("Y29") = Selection.Cells(r, 3)
  ActiveSheet.Range("Z29") = Selection.Cells(r, 4) 'VE
  ActiveSheet.Range("AA29") = Selection.Cells(r, 5) 'VT
  ActiveSheet.Range("AB29") = Selection.Cells(r, 6) 'VO2
  ActiveSheet.Range("AC29") = Selection.Cells(r, 7) 'VO2/KG
  ActiveSheet.Range("AD29") = Selection.Cells(r, 8) 'VE/VO2
  ActiveSheet.Range("AE29") = Selection.Cells(r, 9) 'VCO2
  ActiveSheet.Range("AF29") = Selection.Cells(r, 10) 'VE/VCO2
  ActiveSheet.Range("AG29") = Selection.Cells(r, 11) 'RER
  ActiveSheet.Range("AH29") = Selection.Cells(r, 12) 'PETO2
  ActiveSheet.Range("AI29") = Selection.Cells(r, 13) 'PETCO2
  ActiveSheet.Range("AJ29") = Selection.Cells(r, 14) 'O2PULSE
  ActiveSheet.Range("AK29") = Selection.Cells(r, 15) 'HR
  ActiveSheet.Range("AL29") = Selection.Cells(r, 16) 'Bf

  ActiveSheet.Range("W30") = "AT"
  ActiveSheet.Range("X30") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
  ActiveSheet.Range("Y30") = WorksheetFunction.Average(Selection.Columns(3)) 'LOAD
  ActiveSheet.Range("Z30") = WorksheetFunction.Average(Selection.Columns(4)) 'VE
  ActiveSheet.Range("AA30") = WorksheetFunction.Average(Selection.Columns(5)) 'VT
  ActiveSheet.Range("AB30") = WorksheetFunction.Average(Selection.Columns(6)) 'VO2
  ActiveSheet.Range("AC30") = WorksheetFunction.Average(Selection.Columns(7)) 'VO2/KG
  ActiveSheet.Range("AD30") = WorksheetFunction.Average(Selection.Columns(8)) 'VE/VO2
  ActiveSheet.Range("AE30") = WorksheetFunction.Average(Selection.Columns(9)) 'VCO2
  ActiveSheet.Range("AF30") = WorksheetFunction.Average(Selection.Columns(10)) 'VE/VCO2
  ActiveSheet.Range("AG30") = WorksheetFunction.Average(Selection.Columns(11)) 'RER
  ActiveSheet.Range("AH30") = WorksheetFunction.Average(Selection.Columns(12)) 'PETO2
  ActiveSheet.Range("AI30") = WorksheetFunction.Average(Selection.Columns(13)) 'PETCO2
  ActiveSheet.Range("AJ30") = WorksheetFunction.Average(Selection.Columns(14)) 'O2PULSE
  ActiveSheet.Range("AK30") = WorksheetFunction.Average(Selection.Columns(15)) 'HR
  ActiveSheet.Range("AL30") = WorksheetFunction.Average(Selection.Columns(16)) 'Bf
End Sub

Sub ValuesPeak()
  ActiveSheet.Range("W31") = "Peak"
  ActiveSheet.Range("X31") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
  ActiveSheet.Range("Y31") = WorksheetFunction.Max(Selection.Columns(3)) 'LOAD
  ActiveSheet.Range("Z31") = WorksheetFunction.Max(Selection.Columns(4)) 'VE
  ActiveSheet.Range("AA31") = WorksheetFunction.Max(Selection.Columns(5)) 'VT
  ActiveSheet.Range("AB31") = WorksheetFunction.Max(Selection.Columns(6)) 'VO2
  ActiveSheet.Range("AC31") = WorksheetFunction.Max(Selection.Columns(7)) 'VO2/KG
  ActiveSheet.Range("AD31") = WorksheetFunction.Max(Selection.Columns(8)) 'VE/VO2
  ActiveSheet.Range("AE31") = WorksheetFunction.Max(Selection.Columns(9)) 'VCO2
  ActiveSheet.Range("AF31") = WorksheetFunction.Max(Selection.Columns(10)) 'VE/VCO2
  ActiveSheet.Range("AG31") = WorksheetFunction.Max(Selection.Columns(11)) 'RER
  ActiveSheet.Range("AH31") = WorksheetFunction.Max(Selection.Columns(12)) 'PETO2
  ActiveSheet.Range("AI31") = WorksheetFunction.Max(Selection.Columns(13)) 'PETCO2
  ActiveSheet.Range("AJ31") = WorksheetFunction.Max(Selection.Columns(14)) 'O2PULSE
  ActiveSheet.Range("AK31") = WorksheetFunction.Max(Selection.Columns(15)) 'HR
  ActiveSheet.Range("AL31") = WorksheetFunction.Max(Selection.Columns(16)) 'Bf
End Sub

Sub ValuesOther()
  ActiveSheet.Range("W32") = "Other"
  ActiveSheet.Range("X32") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
  ActiveSheet.Range("Y32") = WorksheetFunction.Average(Selection.Columns(3)) 'LOAD
  ActiveSheet.Range("Z32") = WorksheetFunction.Average(Selection.Columns(4)) 'VE
  ActiveSheet.Range("AA32") = WorksheetFunction.Average(Selection.Columns(5)) 'VT
  ActiveSheet.Range("AB32") = WorksheetFunction.Average(Selection.Columns(6)) 'VO2
  ActiveSheet.Range("AC32") = WorksheetFunction.Average(Selection.Columns(7)) 'VO2/KG
  ActiveSheet.Range("AD32") = WorksheetFunction.Average(Selection.Columns(8)) 'VE/VO2
  ActiveSheet.Range("AE32") = WorksheetFunction.Average(Selection.Columns(9)) 'VCO2
  ActiveSheet.Range("AF32") = WorksheetFunction.Average(Selection.Columns(10)) 'VE/VCO2
  ActiveSheet.Range("AG32") = WorksheetFunction.Average(Selection.Columns(11)) 'RER
  ActiveSheet.Range("AH32") = WorksheetFunction.Average(Selection.Columns(12)) 'PETO2
  ActiveSheet.Range("AI32") = WorksheetFunction.Average(Selection.Columns(13)) 'PETCO2
  ActiveSheet.Range("AJ32") = WorksheetFunction.Average(Selection.Columns(14)) 'O2PULSE
  ActiveSheet.Range("AK32") = WorksheetFunction.Average(Selection.Columns(15)) 'HR
  ActiveSheet.Range("AL32") = WorksheetFunction.Average(Selection.Columns(16)) 'Bf
End Sub

Sub AutomatePeakOtherValues()
  Dim ws As Worksheet
  Dim topRow As Integer
  Dim bottomRow As Integer
  For Each ws In Worksheets
    If ws.Name <> "aaa_main" Then
      ws.Activate
      'Fill other values
      Dim rng As Range
      Dim fdrng As Range
      Set rng = ws.Range("C:C")
      Set fdrng = rng.Find(30, LookIn:=xlValues)
      If Not fdrng Is Nothing Then
        topRow = fdrng(0, 1).Row
        bottomRow = fdrng(2, 0).Row
        Set rng = ws.Range(topRow & ":" & bottomRow)
        rng.Select
        Call ValuesOther
      End If

      'Fill peak values
      Dim maxVO2
      maxVO2 = WorksheetFunction.Max(ws.Range("G:G"))
      Set rng = ws.Range("G:G")
      Set fdrng = rng.Find(maxVO2)
        If Not fdrng Is Nothing Then
          topRow = fdrng(-1, 1).Row
          bottomRow = fdrng(3, 0).Row
        If ws.Range("W31").Value = "" Then
          Set rng = ws.Range(topRow & ":" & bottomRow)
          rng.Select
          Call ValuesPeak
        End If
      End If

    End If
  Next ws
End Sub

Sub ChartResize()
  Range("Y35").Select
  ActiveSheet.ChartObjects("Chart 1").Activate
  ActiveChart.Axes(xlCategory).Select
  ActiveChart.ChartArea.Select
  ActiveSheet.Shapes("Chart 1").ScaleWidth 0.5, msoFalse, _
  msoScaleFromBottomRight
  ActiveSheet.Shapes("Chart 1").IncrementLeft 102#
  ActiveSheet.Shapes("Chart 1").IncrementTop 253.5
End Sub

Sub VECharts()
  Dim ws As Worksheet
  For Each ws In Worksheets
    With ws.ChartObjects.Add _
      (Left:=800, Top:=725, Width:=400, Height:=225)
      .chart.ChartType = xlLine
      .chart.SetSourceData Source:=ws.Range("A:A,H:H,J:J") _
      , PlotBy:=xlColumns

      .chart.HasAxis(xlCategory, xlPrimary) = True
      .chart.HasAxis(xlValue, xlPrimary) = True

      .chart.Axes(xlCategory, xlPrimary).CategoryType = xlAutomatic
    End With
  Next ws
End Sub

Sub CopyCPETParametersFromSheetsToMain()
  Dim x As Integer
  For x = 3 To 99
    Dim sheetname As String
    Sheets("aaa_main").Activate
    sheetname = ActiveSheet.Cells(x, 1)
    ActiveSheet.Cells(x, 32).Select
    Sheets(sheetname).Select
    Range("X30:AL30").Select
    Selection.Copy
    Sheets("aaa_main").Select
    ActiveSheet.Paste

    ActiveSheet.Cells(x, 47).Select
    Sheets(sheetname).Select
    Range("X31:AL31").Select
    Selection.Copy
    Sheets("aaa_main").Select
    ActiveSheet.Paste

    ActiveSheet.Cells(x, 62).Select
    Sheets(sheetname).Select
    Range("X32:AL32").Select
    Selection.Copy
    Sheets("aaa_main").Select
    ActiveSheet.Paste
  Next x

End Sub

Sub CopyVEVO2VEVCO2FromSheetsToMain()
  Dim x As Integer
  For x = 3 To 99
    Dim sheetname As String
    Sheets("aaa_main").Activate
    sheetname = ActiveSheet.Cells(x, 1)
    ActiveSheet.Cells(x, 79).Select
    Sheets(sheetname).Select
    Range("Y22").Select
    Selection.Copy
    Sheets("aaa_main").Select
    ActiveSheet.Paste

    ActiveSheet.Cells(x, 80).Select
    Sheets(sheetname).Select
    Range("Y23").Select
    Selection.Copy
    Sheets("aaa_main").Select
    ActiveSheet.Paste
  Next x

End Sub

Sub FindStartEndTimeofExercise()
  Dim x As Integer
  Dim y As Integer
  Dim starttime As String
  Dim endtime As String
  Dim sheetname As String
  For x = 3 To 99
    Sheets("aaa_main").Activate
    sheetname = ActiveSheet.Cells(x, 1)
    Sheets(sheetname).Select
    y = 15
    While ActiveSheet.Cells(y, 3) = "-"
      y = y + 1
    Wend
    starttime = ActiveSheet.Cells(y, 1)
    y = y + 5
    While ActiveSheet.Cells(y, 3) > "0"
      y = y + 1
    Wend
    endtime = ActiveSheet.Cells(y - 1, 1)
    Sheets("aaa_main").Activate
    ActiveSheet.Cells(x, 77) = starttime
    ActiveSheet.Cells(x, 78) = endtime
  Next x
End Sub

\end{lstlisting}

\setstretch{2.0}

### 2015-06-02 Chapters/BodyComp.tex
Latex files added for first time

% Chapter 04

\chapter{An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing major pancreatic surgery.}
\label{ch_bodycomp}

\lhead{Chapter \ref{ch_bodycomp}. \emph{CPET and Body Composition}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------

\section{Introduction}
Major abdominal surgery especially for pancreatic disease is associated with significant morbidity and mortality. Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes. \parencite{balthazar_acute_2002}

\subsection{Role of preoperative CPET}
The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established. A number of studies have shown that poor aerobic fitness demonstrated by a low anaerobic threshold or low peak VO$_2$ or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.

\subsection{The pathophysiological basis of CPET}
Aerobic fitness, as defined by the ability to perform physical exercise, is dependant on and often limited by the ability of the cardiorespiratory and circulatory systems (henceforth simply the cardiorespiratory system) to supply O2 to skeletal muscles at times of increased demand as well as remove the main end product of aerobic metabolism, namely CO2. Several factors play an important role in this increased response of the cardiorespiratory system. The most important factor is an increase in cardiac output which in healthy adults can increase by upto six-fold during exercise. Aside from increased stroke volume and heart rate, the redistribution of blood volume from the splanchnic circulation increases venous return to the heart. A consequent increase in pulmonary blood flow and skeletal blood flow occurs which in turn is assisted by vasodilation in these circulatory beds.

Oxygenation of the increased pulmonary blood flow and removal of the excess CO2 generated by aerobic exercise is effected by increased minute ventilation as a result of increase in its constituent factors namely respiratory rate and tidal volume. Oxygenation of skeletal muscle is further dependant on numerous other factors including the oxygen carrying capacity of blood (primary determinant being haemoglobin), adequate peripheral circulation and the ability of the mitochondria within the skeletal muscle to utilise the oxygen that is being delivered to them.

It is clear that limitations in the patient's physiology resulting in inadequate or inappropriate response in any of the above mentioned factors will result in overall limitation of their aerobic fitness. Cardiopulmonary exercise testing allows the accurate measurement of most of these factors either directly or indirectly during dynamic exercise thus allowing identifying not only limitations in aerobic fitness but also the cause for such limitation.

\subsection{Factors influencing aerobic fitness}
A low anaerobic threshold and/or low peak VO$_2$ have universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise. This is often thought to be due to cardiorespiratory disease either overt or subclinical. Occasionally other factors such as anaemia, peripheral vascular disease and rarely mitochondrial diseases have been recognised as factors contributing to low anaerobic threshold/peak VO$_2$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but this is uncommon in patients undergoing major abdominal surgery.

The most common parameters used to quantify perioperative risk in surgical patients are oxygen consumption at the anaerobic threshold (VO$_2$AT) and at peak exercise capacity (VO$_2$Peak). Conventionally these have been reported as per weight ratios in mls/kg/min. However, numerous studies on cardiorespiratory exercise physiology have reported that normalising VO$_2$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}

\subsection{Aims}
In chapter \ref{ch_cpet_outcomes}, we reported that low anaerobic threshold in patients undergoing pancreaticoduodenectomy was associated with an increased incidence of postoperative pancreatic fistula and prolonged hospital stay. We also reported that patients with a VO$_2$AT less than 10mls/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and likely due to lack of physiological reserve post-surgery to be fit enough to undergo chemotherapy.

However, we noted that high BMI was associated with a low VO$_2$AT independent of all other clinicopathological characteristics. Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of VO$_2$AT.
The aim of the present study was to explore the association between body composition, total body weight and the physiological parameters measured at cardiopulmonary exercise testing.

\clearpage
\section{Methods}

\subsection{Patients}
Patients who underwent major abdominal surgery for malignant or benign disease involving the head of the pancreas and periampullary region at a single institution between August 2008 and October 2010 were included in this study. All data were recorded in a prospectively maintained database. Data was collected on demographics, preoperative clinicopathological characteristics including blood tests, body mass index, weight, height and the underlying surgical pathology. Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. A detailed description of methodology of cardiopulmonary exercise testing and a description of the measured parameters is provided in CHAPTERX.

\subsection{Body composition calculation}
Preoperative computed tomography that had been performed as part of the routine assessment of these patients was used to calculate body composition. Previously published and well established methods were used were used to calculate body composition information from single CT slices.\parencite{bredella_comparison_2010,shen_total_2004}

The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae.The cross-sectional images at these levels where then exported as bitmap images with C40 W350 settings [speak to a radiologist about what these numbers mean]. The scale in millimeters was included with every image. A representative image is shown in Fig. 1.The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org). The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us. \parencite{anblagan_measurement_2013}

The first step involved converted the bitmap images into JPEG images using lossy compression set at 85\% to minimise sharp transitions between grey areas of very similar colour values. This aided easier automatic selection of contiguous areas of similar grey shades.

The next step involved standardising the scale of all images by dividing the length of the scale on every image by the number of pixels along the scale thus providing a length in millimetres for each pixel in each image. As pixels on a CT image are square, the area of each pixel was calculated as a square of its length.

The Fuzzy Select (Magic Wand) tool was used to select contiguous areas of similar colour while simultaneously using visual confirmation that the correct anatomical structures had been selected without overspill into unwanted areas. The number of pixels within the selection was obtained using the 'Histogram' dialog window and entered into an excel spreadsheet against the selected area of interest. The area in mm2 was calculated by multiplying the number of pixels by the area of each pixel.

\textbf{Body compartment selection methodology}:

The sequence of steps is depicted in Fig. \ref{fig:bc_ct_gimp} on p	\pageref{fig:bc_ct_gimp}. The total cross-sectional area of the abdomen at the level of L3/L4 was calculated by first selecting all the empty space outside the image followed by inverting this selection. This is depicted in Fig. \ref{fig:bc_ct_csa}. Subcutaneous fat in the image was selected using the Fuzzy Select tool (if necessary by choosing multiple times and removing any unnecessary areas) as depicted in Fig. \ref{fig:bc_ct_sat}. The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig. \ref{fig:bc_ct_vat} and Fig. \ref{fig:bc_ct_sm} respectively. Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the area under selection as shown in the insets in each of the images.
%CT slices images x 4
\begin{figure}[htbp]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_csa}
		\caption{Total Cross-sectional Area}
		\label{fig:bc_ct_csa}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sat}
		\caption{Subcutaneous Adipose Tissue$^*$}
		\label{fig:bc_ct_sat}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_vat}
		\caption{Visceral Adipose Tissue$^*$}
		\label{fig:bc_ct_vat}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_ct_sm}
		\caption{Skeletal Muscle$^*$}
		\label{fig:bc_ct_sm}
	\end{subfigure}

	\caption{Selection of components of body composition from CT images using GIMP.}($^*$ The selected area has been removed for representation purposes. The inset confirms the area selected.)
	\label{fig:bc_ct_gimp}

\end{figure}

\subsection{Cardiopulmonary exercise testing}
All patients performed cardiopulmonary exercise testing on a cycle ergometer as described in chapterx. Raw data of all breath-by-breath parameters averaged every 10 seconds was collected for analysis. The first three minutes of the recorded data were during the rest period when the patients were on the exercise bike but did not do exercise. The average of each parameter measured between the first and second minute was treated as the rest value. Anaerobic threshold was identified using previously established methods. \parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.

\subsection{Statistics}

All analyses were performed using the SPSS statistical package for Microsoft Windows (version 22 ). Comparisons between body composition and cardiopulmonary exercise testing parameters were done using the partial correlations controlling for the effect of gender (and/or age). All p-values reported are two-sided. The relationship between body composition and various preoperative clinico-pathological characteristics (in the form of categorical variables) was analysed using the Mann-Whitney U test for variables with two categories and the Kruskal-Wallis Test for variables with more than two categories. Previously established cut-offs were used for categorising continuous variables where applicable. The level of significance was set at $p<0.05$.

\clearpage
\section{Results}

\subsection{Body composition and Clinico-pathological characteristics}
Eighty-two patients (35 male) were included in the study. The clinico-pathological characteristics of the study patients and their relationship to body composition is shown in Table \ref{table:bc_clinical} on page \pageref{table:bc_clinical}. There were several significant associations between clinico-pathological variables and body composition as depicted in this table.

	\input{Tables/c4_t1}

\subsection{Body Composition in Normal BMI vs Overweight/Obese Patients}
The body composition differences between patients with a normal BMI and patients who are overweight or obese is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females. Men had generally larger cross-sectional area, less SAT but greater VAT and SM areas. However, the proportion of skeletal muscle in both males and females decreased significantly with increasing BMI.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
	\caption{Differences in body composition according to gender and body mass index.}
	\label{fig:bc_gender_bmi}
\end{figure}

The proportion of skeletal muscle area at L3/L4 decreases from 38\% in male patients with normal BMI to 22\% in males who are obese. There was a greater decrease in the proportion of skeletal muscle area in females with normal BMI (32\%) and obese females (14\%). The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. Moreover, the distribution of the adipose tissue differed between males and females with visceral adipose tissue contributing more to weight in obese males (43\% VAT vs. 35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (68\% SAT vs. 18\% VAT)

\subsection{Correlation with Pulmonary Function Tests}

Partial correlation analysis was performed to study the relationship between pulmonary function tests and body composition. It has been well-established in previous studies that pulmonary function tests are correlated with age and gender and the analysis was therefore adjusted for these two variables. Forced Vital Capacity (FVC, litres), Forced Expiratory Volume in 1 second (FEV1, litres) and the ratio FEV1/FVC (Tiffeneau-Pinelli index,\%) were compared against the various components of body composition. Both FVC and FEV1 were positively correlated with skeletal muscle area but not with adipose tissue area or total cross-sectional area. FEV1/FVC was not correlated with any of the body composition components. This would indicate that pulmonary function was dependent on skeletal muscle area while FEV1/FVC, a calculated index to quantify restrictive or obstructive lung disease, was not associated with skeletal muscle area. These results are shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.

\subsection{Correlation with Exercise Load}

\input{Tables/c4_t2}

Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area and subcutaneous adipose tissue area measured at L3/L4 to create scatter-plots(Fig. \ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}. Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig. \ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig. \ref{fig:bc_pk_load_vs_sm}). However, no correlation was identified between exercise loads achieved and subcutaneous adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$).

\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_atLoad_skeletal}
		\caption{Anaerobic Threshold}
		\label{fig:bc_at_load_vs_sm}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_pkLoad_skeletal}
		\caption{Peak Exercise}
		\label{fig:bc_pk_load_vs_sm}
	\end{subfigure}
	\caption{Correlation between exercise load and skeletal muscle area.}	
	\label{fig:bc_load_vs_sm}

\end{figure}

%This would suggest as one would expect that exercise capacity is mostly related to skeletal muscle volume rather than subcutaneous adipose tissue. - Move this to discussion

\subsection{Correlation with Oxygen consumption}

The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender. Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to VO$_2$AT or VO$_2$Peak and therefore no adjustments were made for age. The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).

% Absolute oxygen consumption at the anaerobic threshold in litres per minute was plotted against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area. This was repeated by plotting the oxygen consumption at anaerobic threshold corrected for total body weight (in millilitres/kg/min) against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area. These scatter-plots are shown in FigureX1. The same was repeated for oxygen consumption at peak exercise and at rest.

Tidal volume (litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise. There was a statistically significant but weak positive correlation between Minute Ventilation (Tidal Volume x Respiratory Rate) and skeletal muscle at anaerobic threshold and peak exercise but not at rest. There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.

%----------------------------------------------------------------------------------------------
% Correlation between body composition and VO$_2$AT before and after correction for total body weight.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal}
		\caption{VO$_2$AT vs. Skeletal Muscle}
		\label{fig:bc_scatter_VO2_skeletal}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
		\caption{VO$_2$AT vs. Total Adipose Tissue}
		\label{fig:bc_scatter_VO2_TAT}
	\end{subfigure}
	\caption{Correlation between body composition and VO$_2$AT before and after correction for total body weight.}	
	\label{fig:bc_scatter_VO2_bodycomp_reversal}
\end{figure}
%----------------------------------------------------------------------------------------------

Absolute oxygen consumption (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$). However, this correlation was lost after correction of oxygen consumption for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.

Absolute oxygen consumption (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold. However, when it was corrected for total body weight, there was a strong correlation between corrected oxygen consumption (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).

The loss of the physiological relationship between VO$_2$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig. \ref{fig:bc_scatter_VO2_TAT}.

%VO$_2$ corrected to height seems to correlate best with BC and everything else
\clearpage

\section{Discussion}

The results of this study show that the most important cardiopulmonary exercise test parameters as used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition.

\subsection{Oxygen consumption and body composition}

The positive correlation between absolute oxygen consumption and skeletal muscle area is easily explained by the physiology of aerobic exercise. During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.

Current convention is to report oxygen consumption measured at cardiopulmonary exercise testing according the following formula: 
\[Corrected\ VO_2 (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ VO_2\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]

In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between oxygen consumption at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.

The results of the present study suggest that the negative correlation between corrected $VO_{2}$ (mls/kg/min) and BMI is consequent to the reporting convention rather than due to any pathophysiological effect of obesity.

The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for body weight further supports the argument that the corrected value under-reports aerobic capacity in obese patients. Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between O2Pulse and adipose tissue area appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.

\subsection{Comparison with previous studies}
Our findings are similar to those reported by several authors previously. The relationship between body size, body composition and aerobic capacity both at rest and during exercise has been studied extensively for over a hundred years.

Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{"Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation"}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass. In a detailed anaysis comparing oxygen consumption and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt. must be expected to decrease purely as a result of the method used for representing the data." }

Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas.\parencite{batterham_modeling_1999} The authors report that as body mass increased, the proportion composed of fat-free mass decreased. They also found that fat-free mass had a linear relationship with oxygen consumption while total body mass did not. They suggest that ideally estimates of fat-free mass should be used in the representation of oxygen consumption to allow more reliable comparison between subjects.

Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of VO$_2$peak was evaluated annually in 126 children (mean age 10.3 years) for five years. Body composition changes were also tracked over this period. They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. They reported on the significant difficulties in normalising VO$_2$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising VO$_2$. They found that VO$_2$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. However, this underestimation was greater in girls than in boys.

Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on VO$_2$max in obese women before and after weight loss. VO$_2$max corrected for total body weight was significantly lower in the obese state while VO$_2$max corrected for fat-free mass did not change significantly after weight loss. They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities. This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.

These findings have been replicated by several other authors in different subject groups.\parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014} Several of the above studies also recommend using allometric scaling to avoid the confounding effects of total body weight. However, this has not gained widespread clinical use.

In a study aimed at determining the optimal method of expressing VO$_2$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with VO$_2$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat. However, when VO$_2$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. They, similar to Goran et al\parencite{goran_total_2000}, recommend that VO$_2$ be normalised to lean body mass rather than total body weight.

The conclusion from the above studies would be that oxygen consumption normalised for total body weight unfairly penalises obese patients in the absence of true impairment of cardio-respiratory function. This has significant clinical implications as outlined below.

\subsection{Clinical implications of spurious correlation}
Older et al in their pioneering study in 1993 reported that VO$_2$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a VO$_2$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between VO$_2$AT < 11mls/kg/min and mortality due to cardiovascular causes. \parencite{older_cardiopulmonary_1999} The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}

Swart and Carlisle reported that VO$_2$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low VO$_2$AT group was significantly greater than that in the normal VO$_2$AT groups (24\% vs 51\%). The average VO$_2$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min. In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low VO$_2$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.

This is similar to the findings in our cohort of patients. This may have been due to the increased incidence of obesity especially in the subcutaneous plane as we have found in our cohort of patients as shown in Fig. \ref{fig:bc_figure1}.

It is clear from the review presented in Chapter 1, that cardiopulmonary exercise testing is useful in predicting risk after major surgery. Cardiopulmonary exercise testing has become ubiquitous in the preoperative workup of complex surgical patients. However, the results of the present study suggest that the results especially in the obese, female patient must be interpreted with caution, especially when used to select patients who may be declined surgery based on their cardiopulmonary exercise test results.

\subsection{Measuring impact of Prehabilitation}

Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery and mitigating the effects of neoadjuvant oncological therapy. Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients.\parencite{west_effect_2015}

The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs. Instead, improvement in the absolute values of VO$_2$AT and VO$_2$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.

### 2015-06-02 Chapters/CPET_Jaundice.tex
Latex files added for first time

%CPET and Jaundice

\chapter{An investigation into the relationship between obstructive jaundice and preoperative pathophysiology in patients undergoing major pancreatic surgery.}

\label{ch_cpet_jaundice}

\lhead{Chapter \ref{ch_cpet_jaundice}. \emph{CPET and Obstructive Jaundice}} % This is for the header on each page - perhaps a shortened title

\clearpage

%----------------------------------------------------------------------------------------

\section{Introduction}
Patients with tumours involving the pancreatic head or the periampullary region often present with inoperable disease. In the minority of patients with operable disease, resectional surgery in the form of a pancreaticoduodenectomy remains the main modality of treatment and only chance of a potential cure. However, major pancreatic surgery is associated with significant morbidity and mortality and is only undertaken in specialist centres. Patient selection, preoperative optimisation, good surgical technique and improvements in postoperative care have all contributed to a reduction in mortality\parencite{winter_1423_2006} but morbidity remains high. While several technical strategies have been described in recent years to minimise morbidity, these strategies are not necessarily based on a better understanding of the physiological basis of postoperative complications in these patients.

The anatomical relationship between the distal bile duct, distal pancreatic duct, head of the pancreas and the duodenum is responsible for obstructive jaundice being the most common presenting symptom in patients with tumours affecting this region. Distal bile duct strictures also occur in a small proportion of patients with severe chronic pancreatitis involving the pancreatic head. The perioperative management of the patient with obstructive jaundice is complex and management algorithms are still evolving.

Obstructive jaundice has been reported to be associated with abnormal cardiovascular physiology in several animal and human studies. Surgery in the jaundiced patient has been reported to be associated with adverse postoperative haemodynamic events and renal dysfunction.\parencite{pain_perioperative_1985,green_systemic_1995} The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death.\parencite{king_effect_1909} Green and co-workers (1986) described the effects of ‘cholemia' in dogs that were subjected to choledochocaval anastomosis. The resultant myocardial depression was described by them as the ‘jaundiced heart'\parencite{green_jaundiced_1986} and has been reported to be associated with poor myocardial response to inotropic stimulation in dogs\parencite{binah_obstructive_1985, bomzon_systemic_1986} as well as humans.\parencite{lumlertgul_jaundiced_1991}

Preoperative biliary drainage used to be advocated before subjecting a patient to pancreaticoduodenectomy with the intention of reducing postoperative morbidity. However, several recent studies have reported that routine PBD is associated with increased complication rates as a consequence of the drainage procedure itself as well as increased incidence of postoperative complications. The DROP trial reported that PBD was associated with drainage related complication as well as postoperative infectious complications. However, this trial excluded patients with a bilirubin levels greater than 250 mg/dl from the study.

We have recently reported that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy resulting in an increased incidence of POPF and prolonged hospital stay. However, the effects of 'severe jaundice' where bilirubin levels exceed 250 on preoperative patient physiology have not been studied adequately.

The aim of the present study was to evaluate the relationship between obstructive jaundice and preoperative pathophysiology including cardiopulmonary exercise physiology in patients undergoing pancreaticoduodenectomy.

\clearpage

\section{Patients and Methods}
Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative workup at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study. Established criteria for resectability in patients with malignant disease were used as outlined in previous published work. Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.

\subsection{Preoperative Data}
Patient demographics, preoperative clinico-pathological characteristics including cardiorespiratory comorbidity, results of preoperative blood tests, chest x-ray, ECG and cardiopulmonary exercise tests were collected from prospectively held databases. The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease, respiratory disease, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) and was used as an objective score of comorbidity. Cardiovascular comorbidity was defined as a score of 2 or more for either the cardiac disease or ECG component of the POSSUM score. Respiratory comorbidity was defined as a score or 2 or more for the respiratory disease component of the POSSUM score.

\subsection{Obstructive Jaundice}
Serum bilirubin levels were measured in all patients on the day before surgery. Obstructive jaundice (OJ) was defined as bilirubin levels greater than 35 micromol/litre and severe obstructive jaundice (sOJ) was defined as bilirubin levels greater than 250 micromol/litre. This threshold was selected because the DROP trial did not investigate patients with bilirubin levels greater than 250 micromol/litre and this study aimed to evaluate preoperative pathophysiology in this particular group.

Data on PBD (PBD) was also recorded. Serum bilirubin levels before and after biliary stenting were recorded [I will expand this section when I get the updated stent data]

\subsection{Cardiopulmonary Exercise Test}
Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA) (9). All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. A cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons. Peak oxygen consumption achieved at this stage was defined as VO$_2$Peak. The VO$_2$AT was calculated using the V-slope\parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents\parencite{society_ats/accp_2003} methods. VO$_2$AT less than 10 ml/kg/min was considered to be low based on previous work by us\parencite{chandrabalan_pre-operative_2013} as well as Ausania and co-workers\parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients. Oxygen consumption at peak exercise (VO$_2$Peak) was dichotomised using a cut-off of 16 ml/kg/min. Detailed description of cardiopulmonary exercise testing as well as the physiological parameters described in this study are published elsewhere.\parencite{balady_clinicians_2010}

\subsection{Statistics}
Grouping of the variables was carried out using standard or previously published thresholds. In the absence of such thresholds, the variables were treated as continuous variables. Non-parametric tests were used to analyse the association between categorical and continuous variables while Chi-square tests were used to analyse the association between categorical variables. Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and VO$_2$AT / VO$_2$Peak. SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.

\clearpage

\section{Results}
One-hundred and thirty eight patients had undergone pancreaticoduodenectomy (n=138), with preoperative cardiopulmonary exercise testing during the study period. Over half the patients were male (n=93, 67\%). Approximately half the number of patients were over the age of 65 (n=68, 49\%) and overweight or obese (n=69, 50\%). Cardiovascular comorbidity was present in 58 patients (42\%) and respiratory comorbidity was present in 12 patients (9\%). Fifty patients (36\%) had a history of cigarette smoking. The POSSUM Physiology Score was greater than 14 in 61 patients (44\%). Obstructive jaundice (serum bilirubin 35 – 250) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin > 250) was present in 19 (14\%) patients. 
The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_table1}. A larger proportion of jaundiced patients were females compared to the non-jaundiced cohort (p<0.05) and smokers (p<0.05). Patients with jaundice were more likely to have an elevated POSSUM Physiology Score (p<0.005). Patients with cancer were more likely to be jaundiced (p<0.001). However, there was no statistically significant difference in age, BMI, cardiovascular comorbidity, or respiratory comorbidity between the non-jaundiced and jaundiced patients.

\input{Tables/c3_t1}

The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_table2}. While obstructive jaundice was statistically associated with multiple haematological and biochemical abnormalities, most of these did not appear to be of clinically significance. As expected, obstructive jaundice was associated with markedly elevated liver enzymes with severity of derangement associated with severity of jaundice. Obstructive jaundice and sOJ were associated with increasing CRP levels (p<0.001) and decreasing serum albumin levels (p<0.001). Obstructive jaundice was not associated with deranged renal function with both urea and creatinine remaining similar across all cohorts (p=0.09 and p=0.22 respectively).

\input{Tables/c3_t2}

There was no association between obstructive jaundice and preoperative pulmonary function tests (Table \ref{table:cpet_oj_table3}).

\input{Tables/c3_t3}

\subsubsection{Univariate analysis of obstructive jaundice versus CPET}
The relationship between obstructive jaundice and multiple physiological parameters measured at cardiopulmonary exercise testing is shown in Table \ref{table:cpet_oj_table4}. There was an inverse relationship between oxygen consumption at the anaerobic threshold (VO$_2$AT) and increasing severity of jaundice (p<0.05). However, no such linear relationship was noted between any of the other parameters measured both at anaerobic threshold and at peak exercise in spite of apparent statistically significant associations.

\input{Tables/c3_t4}

\subsubsection{Association between preoperative clinico-pathological factors and VO$_2$AT}
On multivariate analysis female sex (HR 3.75 CI 1.57-8.95 p<0.005), high BMI (HR 3.65 CI 1.61-8.26 p<0.005), presence of cancer (HR 4.02 CI 1.33-12.16 p<0.05) and raised CRP (HR 2.98 CI 1.29-6.86 p<0.05) were independently associated with low VO$_2$AT (<10mls/kg/min). However, jaundice was not associated with low VO$_2$AT. These results are shown in Table \ref{table:cpet_oj_table5}

\input{Tables/c3_t5}

\subsubsection{Scatter-plot analysis}
Scatter-plot analysis comparing serum bilirubin and VO$_2$AT as continuous variables is depicted in Figure 1. This shows that the relationship between serum bilirubin and AT is weak with an r2 value of only 0.04 (I will have to confirm this but it is not more than 0.1).

\clearpage

\section{Discussion}
The optimal preoperative management of obstructive jaundice, especially with extremely high serum bilirubin levels, in the patient with periampullary cancer requiring pancreaticoduodenectomy is still unclear. The results of the present study also show for the first time that while obstructive jaundice is associated with a range of biochemical and haematological abnormalities, it does not affect cardiopulmonary physiology as measured by cardiopulmonary exercise testing.

The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers.\parencite{older_preoperative_1993} Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general\parencite{snowden_submaximal_2010}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, oesophagogastric\parencite{nagamatsu_preoperative_2001} as well as vascular\parencite{carlisle_mid-term_2007} surgery. Cardiopulmonary exercise testing has been reported to be superior to conventional measures of comorbidity chiefly due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress. However, the factors responsible for poor aerobic capacity in preoperative patients have not been adequately studied.

The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death.\parencite{king_effect_1909}

Jaundice has been reported to be associated with myocardial depression\parencite{green_jaundiced_1986}, poor myocardial response to inotropic stimulation\parencite{lumlertgul_jaundiced_1991}, impaired sympathetic baroreflex sensitivity\parencite{song_baroreflex_2009}, deranged atrial natriuretic peptide levels\parencite{pereira_increased_1994,gallardo_increased_1998} as well as multiple other bile-acid receptor mediated effects on the cardiovascular system.\parencite{khurana_bile_2011} Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and coworkers.\parencite{padillo_improved_2001}

Historically, obstructive jaundice has also been reported to be associated with adverse haemodynamic events in patients undergoing major surgery. Intraopertive blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were all more common in patients with obstructive jaundice. This increased incidence of complications as a consequence of obstructive jaundice resulted in routine PBD being recommended in these patients in order to alleviate their jaundice before undertaking major surgery. In fact, Whipple described his earliest pancreaticoduodenectomy as a two-stage operation, with the first stage aimed at performing a biliary bypass to reduce jaundice levels before undertaking the resection at a later second operation.

However, more recently, there has been increasing evidence that such routine PBD may itself be associated with increased complications both associated with the drainage procedure itself as well as the effects of PBD on surgical outcomes.

Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications.\parencite{pitt_does_1985} But, this study looked at a heterogenous group of patients of which only 7 underwent pancreaticoduodenectomy.

A recent meta-analysis\parencite{sewnath_meta-analysis_2002} analysed data from 5 randomised controlled trials comparing surgery with PBD versus surgery without PBD and concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself. All five RCTs included in this meta-analysis included a heterogenous group of operations with only a few undergoing pancreaticoduodenectomy while more than 50\% of patients underwent palliative bypass or exploratory laparotomy making comparison of outcomes difficult. A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings.\parencite{wang_preoperative_2008}

The DROP trial sought to clarify the role of PBD in patients undergoing pancreaticoduodenectomy.\parencite{van_der_gaag_preoperative_2010} It randomised patients with bilirubin levels between 40 and 250 either to undergo surgery without PBD or to undergo PBD followed by surgery after 4 - 6 weeks. The authors reported that PBD resulted in an increase in incidence of complications of which the majority were related to the drainage procedure itself. However, this trial excluded patients with bilirubin levels over 250.

While the aforementioned studies have undermined the role of PBD in jaundiced patients undergoing pancreaticoduodenectomy, the results of the present study show for the first time that the premise for performing PBD, namely the adverse effect of jaundice on cardiopulmonary physiology may itself be flawed in patients undergoing pancreaticoduodenectomy. In our study, obstructive jaundice including severe obstructive jaundice did not affect cardiopulmonary exercise capacity as measured by VO$_2$AT or the peak oxygen consumption. These findings taken together with previously published findings of adverse effects of PBD further support the fact that major surgery may be safe in jaundiced patients without subjecting them to preoperative biliary drainage. 
The basis of the relationship between low VO$_2$AT and raised BMI is not clear.

However, such an association has been previously reported.\parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate VO$_2$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction. Other authors have suggested that different thresholds for CPET parameters may have to be considered in obese patients to improve risk-prediction.\parencite{donnelly_criteria_1990,hulens_exercise_2001} Cardiopulmonary exercise testing measures oxygen delivery to skeletal muscle. Adipose tissue, however, does not contribute to the metabolic activity that is measured during CPET. However, AT as normally reported, is calculated by dividing the oxygen consumption per minute at the ‘anaerobic threshold' into the weight of the patient. However, this does not account for the disproportionately higher amount of adipose tissue in overweight/obese patients resulting in a spuriously low AT (in mls/kg/min). The present study found no association between cardiorespiratory comorbidity and VO$_2$AT. Low VO$_2$AT in female patients and overweight/obese patients should be interpreted with caution as this may not be due to true poor aerobic capacity.

\section{Conclusions}
Obstructive jaundice, including severe obstructive jaundice (serum bilirubin>250 mg/dl) does not affect preoperative cardiopulmonary exercise physiology. Reduction of cardiovascular adverse events can no longer be the rationale for preoperative biliary drainage even in patients with severe obstructive jaundice. Future studies must evaluate the safety of elective surgery in patients with severe jaundice and show comparable outcomes to non-jaundiced patients before PBD can be completely abandoned except in special circumstances.

### 2015-06-02 Chapters/CPET_Outcomes.tex
Latex files added for first time

%CPET and Outcomes

\chapter{An investigation into the role of preoperative cardiopulmonary exercise testing in predicting adverse postoperative events after major pancreatic surgery.}

\label{ch_cpet_outcomes}

\lhead{Chapter \ref{ch_cpet_outcomes}. \emph{CPET and Postoperative Outcomes}} % This is for the header on each page - perhaps a shortened title
\clearpage
%----------------------------------------------------------------------------------------

\section{Introduction}
Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years. \parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999} In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy\parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.

The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors.\parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010} Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery. \parencite{mann_review_2010, mayo_management_2012} However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy.\parencite{teh_patient_2009}

There have been a number of attempts to objectively define patient fitness and its relationship with postoperative outcome. Copeland and co-workers (1991) reported that the Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) criteria, in particular the POSSUM physiology score (PPS) could be used to quantify the risk of postoperative morbidity and mortality.\parencite{copeland_possum:_1991} However, the role of POSSUM in predicting postoperative outcome after surgery for pancreatic cancer is not entirely clear.\parencite{de_castro_evaluation_2009, khan_evaluation_2003, kocher_risk-adjustment_2005, pratt_possum_2008, tamijmarane_application_2008} The physiological component of POSSUM as well as other similar risk scoring systems such as E-PASS (Estimation of Physiologic Ability and Surgical Stress)\parencite{haga_estimation_1999} are calculated based on known comorbidities, clinically evident abnormalities in patient physiology or blood tests.

More recently, there has been some evidence that the presence of an ongoing systemic inflammatory response before surgery is associated with the development of postoperative complications in patients undergoing surgery for colorectal cancer\parencite{moyes_preoperative_2009}, oesophageal cancer\parencite{vashist_glasgow_2010} as well as pancreatic cancer.\parencite{knight_evaluation_2010}

Older and co-workers (1993) reported that cardiopulmonary exercise testing (CPET) was an objective evaluation of the response of the cardiovascular and respiratory systems to an increase in oxygen demand during exercise and was useful in predicting perioperative morbidity and mortality in patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}

The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.

\clearpage

\section{Methods}
Patients who underwent pancreaticoduodenectomy or total pancreatectomy for pancreatic head lesions between August 2008, when cardiopulmonary exercise testing was first used for fitness assessment at our hospital, and January 2012 were considered for this retrospective study. Patients who had not undergone cardiopulmonary exercise testing as part of their preoperative assessment and patients who underwent cardiopulmonary exercise testing but did not undergo surgery were excluded.

Data on patient demographics, comorbidity including cardiovascular and respiratory disease, preoperative blood tests, chest x-ray and cardiopulmonary exercise tests were collected from prospectively maintained databases (march 2009 - January 2012) and case note review (August 2008 - March 2009). Data was also collected for patients who did not undergo cardiopulmonary exercise testing to allow comparison with the study group. The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described previously.

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). An electrically-braked cycle ergometer was used to perform a symptom-limited, incremental work-load test preceded by a 3-minute rest period. The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. The VO$_2$AT was calculated using the V-slope\parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents\parencite{sue_metabolic_1988} methods. Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers\parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.

The decision to operate was based on overall preoperative evaluation of the patient’s comorbid conditions and performance status and not exclusively on the result of cardiopulmonary exercise testing. Whilst the results of cardiopulmonary exercise tests were available to the clinicians before surgery, no specific changes were made to perioperative management based exclusively on these results. These results were used in conjunction with other established forms of preoperative evaluation for risk assessment and perioperative care. All patients were routinely admitted to the surgical high dependency unit unless intra-operative events or postoperative complications required admission to the intensive care unit. Patients were discharged after resolution of organ dysfunction and/or sepsis and when nutrition, analgesia and mobilisation were adequately established to the clinician's and patient's satisfaction.

Postoperative adverse events were recorded using internationally recognised definitions. The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae\parencite{bassi_postoperative_2005} and post-operative haemorrhage\parencite{wente_postpancreatectomy_2007}. The Clavien-Dindo (CD) classification\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major. Multiple admissions to critical care as well as re-operations were recorded. Operative mortality was defined as postoperative death in-hospital regardless of duration of stay or occurring within 30 days of the surgery. All complications were discussed at a weekly multidisciplinary meeting attended by three pancreatic surgeons and a radiologist with a specialist interest in pancreatic diseases and recorded in a prospective database.

Primary outcome measures were length of stay in hospital, major postoperative adverse events including operative mortality and fitness to undergo adjuvant therapy when indicated. Secondary outcome measures included cumulative length of stay in critical care and number of critical care admissions.

\subsection{Statistics}
Grouping of the variables was carried out using standard or previously published thresholds. In the absence of such thresholds, the variables were treated as continuous variables and analysed using non-parametric statistical methods. Cox proportional hazards regression analysis was used to study the relationship between preoperative risk factors and length of hospital stay. Chi-square test was used to examine the relationship between complications and VO$_2$AT as a categorical variable. Univariate binary logistic regression analysis with calculation of hazard ratios (HR) and 95\% confidence intervals was used to explore the association between perioperative clinico-pathological factors and receipt of adjuvant therapy. Multivariate binary logistic regression analysis was performed on all variables showing a significant association on univariate analysis. Backward stepwise regression was used starting with a saturated model and variables with P-value$>$  0.1 were excluded at each step until no more variables could be excluded. SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.

\clearpage

\section{Results}
One hundred and twenty-nine patients had undergone pancreaticoduodenectomy (n=127), sub-total pancreatectomy (n=1) or total pancreatectomy (n=1) during the study period. Sub-total and total pancreatectomy were performed in patients scheduled for a pancreaticoduodenectomy but were found to have pancreatic remnants either too friable or too atrophic during the operation to perform an anastomosis. Of these, 100 patients (pancreaticoduodenectomy - 98, sub-total/total pancreatectomy - 2) had undergone cardiopulmonary exercise testing as part of their preoperative assessment and were included in the study. Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intraductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).

Twenty-nine patients did not undergo cardiopulmonary exercise testing due to reasons including subjective assessment of fitness, resource constraints and logistics. Table \ref{table:cpet_outcomes_table1} shows the clinico-pathological characteristics of patients included in the study compared to the excluded patients. The median age in the study cohort was higher than in the excluded cohort (66 vs. 54 years, p=0.001). However, there was no difference in gender, body mass index, preoperative biliary drainage, jaundice at the time of surgery, modified Glasgow Prognostic Score, POSSUM physiology score, preoperative blood tests including haemoglobin and liver function tests and length of critical care/hospital stay. The overall postoperative mortality during the study period was 5.4\% (7/129) with all deaths occurring in the study cohort (p=0.144).

\input{Tables/c2_t1}

The median VO$_2$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). The VO$_2$AT was less than 10ml/kg/min in 49 patients. The distribution of VO$_2$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.

%Figure 1
\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_dist_of_AT}
	\caption{Distribution of VO$_2$AT across the study population.}
	\label{fig:cpet_outcomes_dist_of_AT}
\end{figure}

The relationship between VO$_2$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. Patients with VO$_2$AT less than 10ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042). While there was an association between low VO$_2$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). There was no association between low VO$_2$AT and cardiopulmonary complications or postoperative mortality. Major cardiopulmonary complications occurred more often in patients with major intra-abdominal adverse events including major intra-abdominal abscesses or Grade B and C pancreatic fistulae or haemorrhage than in patients who did not have these complications (5/31,16.1\% vs. 2/69,2.9\%, p=0.017). Postoperative mortality was not associated with VO$_2$AT (HR 0.77, 95\% CI 0.16-3.61, p 0.737) or the POSSUM Physiology Score (HR 0.39, 95\% CI 0.07-2.12, p 0.277). Postoperative mortality was associated with postoperative pancreatic fistula (n=5), post-pancreatectomy haemorrhage (n=3), major intra-abdominal sepsis (n=6) and major cardiorespiratory complications (n=4) with 6 patients requiring radiological or operative intervention.

\input{Tables/c2_t2}

The median length of postoperative stay was 17 days (IQR 13 - 26). The median cumulative length of stay in critical care was 7 days (IQR 6 - 12). Twenty-six patients were admitted to critical care more than once. The relationship between preoperative clinico-pathological characteristics and length of postoperative stay in patients who were discharged from hospital (n=93) is shown in Table \ref{table:cpet_outcomes_table3}. On univariate analysis, age over 65 years (p=0.072) and low VO$_2$AT (p=0.010) were associated with prolonged postoperative stay. On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay. A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low and normal VO$_2$ATs is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. Patients with a low VO$_2$AT stayed a median 6 days longer in hospital (14 versus 20 days, Mann-Whitney Test p=0.001). There was no significant association between any of the preoperative factors including VO$_2$AT and length of critical care stay or number of critical care admissions.

\input{Tables/c2_t3}

% Figure 2
\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_km_at_los}
\includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10ml/kg/min versus $<$ 10ml/kg/min.}
\label{fig:cpet_outcomes_km_at_los}
\end{figure}

The relationship between clinico-pathological patient factors and receipt of adjuvant therapy is shown in Table \ref{table:cpet_outcomes_table4}. Fifty-five patients were included in the analysis. Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6). On binary logistic regression analysis, VO$_2$AT less than 10ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).

\input{Tables/c2_t4}

\clearpage

\section{Discussion}
The results of the present study show that a low VO$_2$AT is associated with prolonged postoperative stay in hospital, postoperative pancreatic fistula and intra-abdominal abscesses in patients undergoing major resections for pancreatic head lesions. The results of this study also show that patients with low VO$_2$AT are less likely to receive adjuvant therapy.

Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.

Patients with a low VO$_2$AT stayed longer in hospital after their operation. While length of stay in hospital is influenced by multiple factors including postoperative complications, it would appear that patients with a low VO$_2$AT take longer to recover from the physiological stress placed by major pancreatic surgery and its sequelae.

The incidence of pancreatic fistula was greater in patients with a low VO$_2$AT. This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss.\parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006} It is possible that local or operative factors may be compounded by poor oxygen delivery and organ perfusion as measured by cardiopulmonary exercise testing. There was a non-significant trend towards clinically relevant pancreatic fistulae (ISGPS Grades B and C) as well as a significant association with major intra-abdominal abscesses (Clavien-Dindo Grades 3-5 i.e., requiring intervention, associated with organ dysfunction requiring intensive care or resulting in mortality). This would suggest that complications in patients with low VO$_2$AT are more likely to be severe than in patients with normal VO$_2$AT. However, there was no difference in mortality between patients with normal or low VO$_2$AT, indicating that multiple factors including preoperative patient fitness, local and operative factors, systemic inflammatory response, number of complications as well as perioperative critical care all play a role.

The results of this study also show that patients with a low VO$_2$AT were less likely to receive adjuvant therapy. Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly.\parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade\parencite{winter_1423_2006} even in elderly patients\parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high.\parencite{mann_review_2010} The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.

In the present study, VO$_2$AT was less than 10ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit. While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this.\parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010} Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}

However, Snowden and co-workers\parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers\parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy. The lack of association between low VO$_2$AT and cardiopulmonary complications in this study may have been due to two reasons. Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy.\parencite{braga_prognostic_2011} Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.

The results of this study are consistent with the results of the study by Ausania and co-workers\parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min. However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.

The physiological demands placed on a patient undergoing major pancreatic surgery are significant, both during and after the operation. It is not entirely surprising therefore, that conventional parameters of patient fitness like the POSSUM Physiology Score or the modified Glasgow Prognostic Score are limited in their ability to distinguish patients based on their performance under physiological stress. Cardiopulmonary exercise testing overcomes this disadvantage by replicating some of the physiological burden major pancreatic surgery places on the functional capacity of the patient's cardiovascular and respiratory systems.

This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’.\parencite{topp_effect_2002} It has been suggested that prehabilitation not only improves aerobic capacity\parencite{jones_effects_2007} but may also improve postoperative recovery.\parencite{mayo_impact_2011, pehlivan_effects_2011} The results of this study show that impaired aerobic capacity is associated with postoperative adverse events. Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.

Further work needs to be carried out to study the value of cardiopulmonary exercise testing in predicting postoperative complications in conjunction with previously established factors such as pancreatic morphology and operative factors before it can be used on its own to select or exclude patients for pancreaticoduodenectomy. Cardiopulmonary exercise testing would play an important role not only in identifying patients who will benefit from prehabilitation, but also in the objective measurement of the effects of such interventions on aerobic capacity as well as in identifying high risk patients who may not be able to complete oncological treatment. Prehabilitation and optimised perioperative care may allow a greater proportion of high risk patients to progress to oncological treatment after surgery.

### 2015-06-02 Chapters/CPET_Survival.tex
Latex files added for first time

% Chapter 05 - CPET, complications and survival

\chapter{An investigation into the relationship between cardiopulmonary exercise testing, comorbidity, systemic inflammation and survival after pancreaticoduodenectomy for cancer.}
\label{ch_survival}

\lhead{Chapter \ref{ch_survival}. \emph{CPET, Complications and Survival}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------

\section{Introduction}
Median survival after pancreaticoduodenectomy for pancreatic ductal adenocarcinoma varies from approximately 18 months to 24 months.\parencite{winter_1423_2006,neoptolemos_adjuvant_2010}

Selecting patients who will benefit from the survival advantage that a pancreaticoduodnectomy offers is important to maximise the usefulness of this morbid procedure.

Comorbidity is not only associated with postoperative morbidity and mortality (Section \ref{sec:comorbidity_risk_stratification}) but has also been reported to be associated with poor long-term survival in patients undergoing surgery several different cancers including colorectal cancer \parencite{} and breast cancer.\parencite{} With 10-year survival rates approaching 60\% and 80\% in these patients, it is not surprising that some patients with significant comorbidity may die from their comorbid disease rather than from cancer recurrence. However, cancer-specific survival is also shorter in patients with significant comorbidities.

Systemic inflammation has been proposed as one of the intermediary mechanism in these patients that increases rates of recurrence and decreases disease free survival. The modified Glasgow Prognostic Score, a measure of preoperative systemic inflammation in cancer patients, is associated with poor survival regardless of the site or stage of cancer. This is discussed in detail in Section \ref{sec:intro_systemic_inflammation_outcome}.

However, an objective method to measure comorbidity itself remains elusive and various scores have been used for this purpose. The Charlson Comorbidity Index is one such score and has been reported to predict long-term survival in cancer patients. Cardiopulmonary exercise testing is an objective measure of aerobic fitness and of cardiorespiratory comorbidity and has been shown to be useful in predicting complications after major abdominal surgery including pancreatic surgery. (\parencite{ausania_effects_2012} and Chapter \ref{ch_cpet_outcomes})

Moreover, cardiopulmonary exercise testing has been used to predict medium term survival after aortic aneurysm surgery \parencite{carlisle_mid-term_2007} as well as overall survival in patients with medical diseases such as chronic heart failure or chronic obstructive airways disease. The relationship between cardiopulmonary exercise testing and long-term survival in patients undergoing pancreaticoduodenectomy for cancer has not been reported before.

\section{Aim}
The aim of the present study was to investigate the relationship between cardiopulmonary exercise testing, comorbidity, systemic inflammation and survival in patients undergoing pancreaticoduodenectomy for pancreatic ductal adenocarcinoma.

\section{Patients and Methods}
All patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma between August 2008 and July 2012 were included in the study. Data was collected prospectively in a structured database and included demographics, preoperative clinico-pathological characteristics, cardiopulmonary exercise testing, postoperative complications, tumour characteristics and long-term survival. Survival data was collected using the Greater Glasgow and Clyde NHS Clinical Portal and the Scottish National Statutory Register of Deaths. The modified Glasgow Prognostic Score was calculated as described in Table \ref{table:mGPS} on p\pageref{table:mGPS}. The POSSUM Physiology Score was calculated as described in \ref{}.

The Scottish Index of Multiple Deprivation (SMID) combines 38 indicators across 7 domains including  income, employment, health, education, skills and training, housing, geographic access and crime.  All of Scotland's population is placed into 6505 geographical groups ranked in descending order of deprivation. SMID quintiles place these into 5 categories with 1 representing the most deprived areas and 5 representing the least deprived. The SMID quintile for each patient was derived from the postcode of their primary residence.

\subsection{Statistics}
Standard thresholds were used to categorise continuous variables where applicable. Kaplan-Meir survival analysis and Cox-regression analysis were used to study the relationship between preoperative clinico-pathological characteristics and long-term survival. SPSS version 22 statistical software package was used for all analysis.

\section{Results}

\section{Discussion}

 This is considerably less than other common cancers such as colorectal cancer or breast cancer where 5-year survivals across all stages are 60\% and 90\% respectively. However, only 10-20\% of pancreatic cancers are suitable for potentially curative surgery and of patients who undergo curative surgery 5-year survival remains low at approximately 20\%.\parencite{cancerresearchuk_cancer_2014}

### 2015-06-02 Chapters/Discussion.tex
Latex files added for first time

% Chapter 06 - Conclusion

\chapter{Conclusion}
\label{ch_conclusion}

\lhead{Chapter \ref{ch_conclusion}. \emph{Conclusion}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------

This is the easy bit

### 2015-06-02 Chapters/Introduction.tex
Latex files added for first time

% Chapter 01 - Introduction

\chapter{Introduction}
\label{ch_intro}

\lhead{Chapter \ref{ch_intro}. \emph{Introduction}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------

\section{Pancreatic Neoplasia}

\subsection{Epidemiology of pancreatic cancer}
\parencite{crozier_preoperative_2007}
Tumours involving the head of the pancreas and the periampullary region account for a small proportion of gastrointestinal tumours. They may be broadly classified as benign and malignant. Most pancreatic neoplasia are malignant and arise from the exocrine component of the gland, the ductal epithelium.

Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. These include the distal common bile duct that can give rise to cholangiocarcinoma, the duodenum that can give rise to duodenal adenocarcinoma and the ampulla that can give rise to ampullary adenocarcinoma. The endocrine portion of the pancreas can give rise to a variety of tumours that are collectively called neuroendocrine tumours (NET). The milieu of tumours is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours. Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.

Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 21\% surviving beyond the first year and 3\% surviving beyond 5 years.\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}

In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} remains the primary modality of cure. However, major pancreatic surgery places significant physiological stresses on multiple organ systems. The ability of the cardiac and respiratory systems, in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.

\subsection{Clinical presentation}

The anatomical location of the pancreas, deep within the retroperitoneum surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours. Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. Obstructive jaundice is the most common presenting symptom and painless, obstructive jaundice in an elderly patient should always raise the suspicion of a neoplastic process in the head of the pancreas or the periampullary region. Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back.

\subsection{Diagnosis and staging}

Aside from a thorough history, clinical examination, blood tests including liver function tests, diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase). CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced MRI Liver in specific cases. CECT-pancreas is also useful for assessing local resectability with regards to vascular involvement. Endoscopic ultrasound (EUS) is also useful in assessing vascular involvement and for obtaining tissue samples for histological examination. In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile ducts, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile ducts for cytological examination. The role of preoperative biliary drainage is discussed in more detail in section %\ref{sec:preoperative_biliary_drainage].

\subsection{Treatment of pancreatic cancer}
Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. Assessing the resectability is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.

\section{Surgical treatment of pancreatic cancer}
Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description. The procedure was performed as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}. The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}

\subsection{Patient selection}
\subsubsection{Resectability criteria}
Resectable pancreatic cancer is defined as a tumour that
- does not involve the coeliac axis or the superior mesenteric artery
- and is not associated with distant metastatic disease

Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection. Research is ongoing to assess the role of neoadjuvant therapy and newer treatment modalities such as electroporation in these patients to improve resectability.

\subsubsection{Patient factors}
\subsection{Operative technique}
Pancreaticoduodenectomy is considered one of the most technically challenging operations on the gastrointestinal tract. While the procedure is carried out in a broadly similar fashion in all major centres, there remain some variations in perioperative care as well as some operative steps. The following is a description of the procedure as performed at the West of Scotland Pancreatic Unit.

After a comprehensive preoperative work-up including both assessments of the tumour as well as patient fitness, informed consent was obtained. Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital. General anaesthesia with complete muscle relaxation was used in all patients. Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine. Antibiotic prophylaxis is administered at induction. While the use of Octreotide, a somatostatin analogue, to reduce the risk of postoperative pancreatic fistula formation is still debated, it was routinely used in all patients at this centre. Octreotide was administered intra-operatively (200 mcg s.c.) and was continued for 5 days postoperatively (200 mcg s.c., t.d.s.).

A roof-top incision was used for access. After assessing the peritoneal cavity for absence of metastatic disease, an early assessment was made for local resectability. This involved complete Kocherisation of the duodenum to assess the retroperitoneum. Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). The rest of the procedure was performed as described extensively elsewhere. The gastrocolic omentum was divided to enter the lesser sac. The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein. If less than half the circumference of the SMV or PV was involved, an en-bloc resection was performed with vein repair at the same time. The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. The gastro-duodenal artery was divided. Resection was then completed by dividing the stomach (classical Whipple procedure) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy, PPPD) and transecting the pancreatic neck.

Reconstruction was performed as follows: Either a pancreatico-jejunostomy was performed using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was performed using 3/0 Biosyn sutures placed in a similar manner. Hepaticojejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastrojeunonostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 PDS sutures in a 2-layers. One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.

\subsection{Postoperative care}
All patients were routinely admitted to the Surgical High Dependency Unit unless intra-operative events necessitated admission to the Intensive Care Unit. A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion. All patients received proton pump inhibitors and octreotide. Patients were discharged to the general surgical ward as early as possible.

\subsection{Complications}
The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006,emick_hospital_2006,yeo_six_1997, winter_1423_2006,teh_patient_2009,gouma_rates_2000}

\subsubsection{Postoperative pancreatic fistula}
Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability. The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.\parencite{yeo_six_1997,deoliveira_assessment_2006,bassi_postoperative_2005,winter_biochemical_2007,pratt_risk_2008} 
The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula. It can be a result of breakdown or poor healing at the pancreaticojejunostomy/pancreaticogastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis. It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing. Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. Grade B and C fistulae are considered to be clinically significant in that they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage, delayed gastric emptying as well as need for intervention (either radiological or operative) and/or prolonged critical care support.

\subsubsection{Post-pancreatectomy haemorrhage}
Post-pancreatectomy haemorrhage is reported to occur in 1 to 8\% of patients undergoing pancreaticoduodenectomy. However, it accounts for 11\% to 38\% of mortality after pancreaticoduodenectomy. Post-pancreatectomy haemorrhage may either be intra-luminal into the gastrointestinal tract or intra-abdominal into the peritoneal/retro-peritoneal space. Post-pancreatectomy haemorrhage may be from any of a number of potential sources although bleeding from the stump of the gastroduodenal artery is the most common cause. Other potential sources include suture lines at the anastomoses, gastric/duodenal ulcers or diffuse gastritis, pseudoaneurysms of the gastro-duodenal, splenic or rarely the hepatic artery or rarely, haemobilia.

Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodnenal artery. Post-pancreatectomy haemorrhage is often managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.

\input{Tables/c1_isgps}

\subsubsection{Clavien-Dindo classification of complications}
A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrythmias, pneumonia, pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal sepsis including intra-abdominal sepsis, leakage from the hepaticojejunostomy or the gastrojejunostomy, renal dysfunction, etc. The Clavien-Dindo method grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis.

\input{Tables/c1_clavien_dindo}

\section{Adjuvant and Neoadjuvant treatment}

\clearpage

\section{Comorbidity and Risk Stratification}
\label{sec:comorbidity_risk_stratification}
\subsection{Comorbidity}
Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or chemoradiotherapy.\parencite{sandroussi_sociodemographics_2010}

Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009}

Major pancreatic surgery requires the patient to have adequate physiological reserve to cope with the increased demand during and immediately after surgery. However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery. \parencite{shah_limitations_2012}

\subsection{Risk Stratification}
Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery. Major surgery places a significant physiological stress on multiple organ systems, especially the cardiorespiratory system. The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and the postoperative recovery plays a major role in determining short-term outcomes.

Accurate measurement of physiological fitness

\subsection{Static Versus Dynamic Testing}

Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase. Identification of such high-risk patients allows not only for improved patient selection, but also for risk-stratified, anaesthetic and postoperative critical care. Preoperative risk stratification will also allow for prehabilitation of these patients in an attempt to improve outcomes.

Several tests have been used for preoperative assessment of cardiac function. These include 
- electrocardiography 
- echocardiography 
- exercise tolerance testing 
- myocardial perfusion scans

Tests of respiratory function that are commonly performed in selected patients undergoing major surgery include 
- pulmonary function tests including forced expiratory volume and forced vital capacity 
- spirometry

However, neither of the above cardiac or respiratory function tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand.

\section{Cardiopulmonary Exercise Testing}
%History of CPET
\subsection{History of CPET in Surgery}

\subsection{Cardiopulmonary Exercise Test Methodology}
CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.

Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). The equipment was calibrated regularly to the standards set by the manufacturer and currently published guidelines[]. All tests were performed by specialist respiratory physiologists. Suitable equipment for cardiopulmonary resuscitation were available in the department in case of unexpected problems. The department was situated within the main hospital premises and therefore was easily accessible to the hospital cardiac arrest team. All patients were fully informed of the steps involved in the procedure, the reasons for performing the test as well as the risks involved.

Spirometry was performed in all patients prior to CPET. Capillary blood gases were measured in all patients after CPET. An electronically braked cycle ergometer was used to increase resistance to pedalling in preset increments. A tight-fitting face mask was placed on the patient covering the nose and the mouth. This allowed breath-by-breath gas analysis thus allowing measurement of several respiratory parameters as listed in table []. 12-lead electocardiogram was recorded at the same time.

The test started with an initial 3-minute rest period to allow measurement of baseline parameters. This was followed by an incremental work-load test that involved the patient pedalling approximately at 60 revoultions per minute while the resistance to pedalling was gradually increased in preset increments. The test was terminated when patients reached volitional fatigue (maximal exercise tolerance), significant ischaemic changes on ECG or for other safety reasons.

The parameters measured at spirometry are shown in Table \ref{table:spirometry} and those measured during cardiopulmonary exercise testing are shown in Table \ref{table:cpet_parameters} on p\pageref{table:cpet_parameters}.

\subsection{Measuring the Anaerobic Threshold}
The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply. Therefore, muscle tissues use anaerobic respiration to supplement aerobic respiration to continue generation of ATP. The resulting metabolic lactic acidosis is almost immediately compensated by the bicarbonate buffer as below: 
\begin{equation} \label{eq:bicarb_buffer}
	H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2
\end{equation}
The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing. This transition from aerobic to anaerobic respiration may be determined using the V-slope method\parencite{sue_metabolic_1988} or the ventilatory equivalents method.\parencite{beaver_new_1986} Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing.\parencite{society_ats/accp_2003}

The software presents a standard 9-panel view of trending plots of various parameters measured during incremental exercise. All of these trends are taken into consideration rather than any one particular parameter value in determining the overall outcome of the test. A sample 9-panel view derived from parameters belonging to one of the patients studied is shown in Figure \ref{fig:cpet_9panel}. The data used to generate these plots is included in Appendix \ref{AppendixCPETRawData}.

\begin{figure}[p]
	\centering
	\includegraphics[width=0.9\textwidth]{Figures/cpet_9panel.pdf}
	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
	\label{fig:cpet_9panel}
\end{figure}

\subsubsection{V-slope method}

During aerobic exercise, VO2 and VCO2 share a linear relationship as shown in segment A of the graph in figure 1. However, as anaerobic respiration starts to supplement aerobic respiration, VCO2 increases disproportionate to VO2 as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}. This results in a distinct difference in the slope of the initial part of the graph (seg A) and the later part (seg B). The point at which the two slopes intersect is the anaerobic threshold and the V02 at this point in exercise is commonly referred to as the anaerobic threshold, VO2at or simply AT.

\subsubsection{Ventilatory equivalents method}
[...]

\section{Description of CPET Parameters}
\input{Tables/c1_cpet_parameters}

\subsection{Exercise Load}
The most common form of cardiopulmonary exercise testing for clinical purposes involves a cycle ergometer with steadily increasing resistance delivered through electric braking allowing accurate measurement of work load in Watts. The relationship between $\dot{V}_{O_2}$ and work rate is usually linear and the slope of this relationship is independent of sex, age or height. An abnormality in this relationship is usually due to cardiopulmonary or circulatory causes.

\subsection{Minute Ventilation, $\dot{V}_E$}
Minute ventilation or respiratory minute volume is the volume of air that is inhaled/expired in a minute.
\begin{equation} \label{eq:VE=VTxBf}
	\dot{V}_E = \dot{V}_T \times Bf
\end{equation}
where $\dot{V}_T$ = Tidal Volume and $Bf$ = Breathing Frequency.

Increasing $\dot{V}_E$ is one of the main mechanisms involved in increasing oxygen delivery during exercise. It is also an important factor in clearing $CO_2$ from the blood.

\subsection{Oxygen Uptake, $\dot{V}_{O_2}$}

$\dot{V}_{O_2}$ or oxygen uptake is measured breath-by-breath using digital analysis of the inspired and expired gases. This is then averaged, usually over time, to smooth-out any signficant breath-by-breath variation. $\dot{V}_{O_2}$ increases with increasing work load and is influenced by several factors that have a role in the transport and utilisation of oxygen. These may be broadly classified as cardiac, pulmonary, circulatory and tissue factors. Some of the factors are encompassed in the following formula for $\dot{V}_{O_2}$.
\begin{equation} \label{eq:VO2=CaO2xCO}
	\dot{V}_{O_2} = CaO_2 \times Cardiac\ Output
\end{equation}
where $CaO_2$ is $O_2$ content per ml of blood and is defined by,
\begin{equation} \label{eq:CaO2=Hbx1.34xSaO2}
	CaO_2 = Haemoglobin \times 1.34 \times SaO_2
\end{equation}
and
cardiac output, the primary cardiac factor that influences $\dot{V}_{O_2}$, is:
\begin{equation} \label{eq:CO=SVxHR}
	Cardiac\ Output = Stroke\ Volume \times Heart\ Rate
\end{equation}
Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.

Pulmonary gas exchange plays an important role in the oxygenation of blood and removal of $CO_2$ and is influenced by numerous factors, the detailed discussion of which is beyond the scope of this chapter. However, ventilation, pulmonary blood flow, gas-exchange across the alveolar membrane and ventilation-perfusion mismatches (V/Q mismatch) all play an important role in determining the response of the lungs to exercise.

The quality of the peripheral circulation, both anatomical and its physiologic response to exercise which involves redistribution of blood flow to exercising muscle, has an important role in increasing availability of oxygen. The oxygen carrying capacity of blood determined by haemoglobin concentration, its saturation and the $O_2$ dissociation curve as well as the ability of tissues to extract and utilise oxygen are equally important factors that influence $\dot{V}_{O_2}$.

\subsection{Oxygen Pulse, $O_2Pulse$}
Oxygen pulse is defined as the oxygen uptake per heart beat.
\begin{equation} \label{eq:O2Pulse=VO2/HR}
	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
\end{equation}
While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. The clinical application of oxygen pulse in surgical patients remains unclear.

\subsection{Respiratory Exchange Ratio, RER}
The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio. An RER greater that 1.0 may be caused either by lactic acidosis or due to hyperventilation. The RER is also a marker of the fuel being used for metabolism with RER less than 1.0 indicating mixed fuel source in the form of carbohydrate and fat while an RER of 1.0 or greater indicates a primarily carbohydrate source.

\subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
The change in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise. Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. However, as the anaerobic threshold is passed, $\dot{V}_E/\dot{V}_{O_2}$ starts increasing before $\dot{V}_E/\dot{V}_{CO_2}$. This change in direction is yet another method to confirm the anaerobic threshold.  $\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as well as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.

\subsection{End-tidal $O_2$ and $CO_2$, $P_{ET_{O_2}}$, $P{ET_{CO_2}}$}
$PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to PaO2 and PaCO2 respectively. $P{ET_{CO_2}}$ is dependent on pulmonary gas-exchange which is in turn influenced by the right ventricular output, pulmonary blood flow and alveolar gas exchange. The changes in $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ during exercise help identify ventilation-perfusion mismatch as well as hyperventilation.

\subsection{Heart Rate, HR}
\label{sec:heart_rate}
The heart rate response during exercise in healthy individuals is a linear function of $\dot{V}_{O_2}$ increasing linearly with increasing work load and increasing $\dot{V}_{O_2}$. The difference between the predicted peak heart rate and the observed peak heart rate is called the Heart Rate Reserve or HRR. Failure to achieve the predicted peak heart rate or a wide HRR may be due to cardiac disease or due to medication used to treat cardiovascular disorders such as beta-blockers or calcium-channel blockers. This information in conjunction with 12-lead ECG evidence of ischaemia provides undeniable evidence of primary cardiac dysfunction.

\subsection{Breathing frequency, $B_f$}

\section{Role of CPET in preoperative assessment}
\subsection{General Surgery}
\subsection{Oesophago-gastric Surgery}
\subsection{Colorectal Surgery}
\subsection{Vascular Surgery}
\subsection{Hepato-pancreato-biliary surgery and Transplantation}
\subsection{Thoracic Surgery}

\clearpage

\section{Systemic inflammation and outcome}
\label{sec:intro_systemic_inflammation_outcome}

%Preop inflammation - short-term and long-term outcomes
%Postop inflammation -> complications -> long-term ouctomes
%Possible mechanisms behind poor outcomes

The host inflammatory response to cancer, comorbidity and surgical trauma has been known to influence both short-term and long-term outcomes after major cancer surgery. Moreover, postoperative complications have been reported to be associated with poorer oncologic outcomes and cancer-specific survival in patients undergoing potentially curative surgery for cancer. The complex interactions between pro-inflammatory cytokines and anti-inflammatory cytokines at different phases during the perioperative period further impact upon the incidence of complications as well as survival.

\subsection{Measuring systemic inflammation}
Numerous tests are available to not only measure systemic inflammation in general but also to quantify the various components of the inflammatory response. The most commonly employed measures in the clinical setting are the serum levels of C-reactive protein (CRP)and the differential leucocyte count.

One of the earliest reports on the use of CRP to predict cancer-specific survival was by McMillan and co-workers in 1995 when they reported that an elevated CRP 4 months after curative resection for colorectal cancer was associated with earlier recurrence.\parencite{mcmillan_prospective_1995} The modified Glasgow Prognostic Score (mGPS)\parencite{elahi_score_2004} is based on a combination of C-reactive protein and serum albumin and is outlined in Table \ref{table:mGPS}. Since its introduction, mGPS has been validated in over a hundred studies looking at several thousand patients with a wide-range of cancers and an increasing score is associated with poorer long-term survival in patients with operable as well as inoperable cancers.

\begin{table}[h]
	\centering
	\caption{The modified Glasgow Prognostic Score}
	\label{table:mGPS}
	\begin{tabular}{c c c}
		mGPS & CRP (mg/dL) & Albumin  (mg/dL) \\ \hline
		 0   & $\leq 10$   & $\geq 35$        \\
		 1   & $> 10$      & $\geq 35$        \\
		 2   & $> 10$      & $< 35$
	\end{tabular}
\end{table}

\subsection{Systemic inflammation and long-term survival}
Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable pancreatic cancer.\parencite{glen_evaluation_2006} Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated at over 10 mg/dl approximately a month after their operation.\parencite{jamieson_systemic_2005} Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR). A selection of these studies are presented in Table %Create new table here for this.

\subsection{Systemic inflammation and postoperative complications}

Abnormalities of systemic inflammatory processes present as a continuum that starts in the preoperative phase possibly as a consequence of underlying comorbid illnesses, presence of cancer, or an abnormality of the immune system or a due to a combination of all of these factors. Surgical trauma in such 'primed' patients results in a cascade of events that trigger several inflammatory pathways that have now shown to have a direct impact not only on the incidence of postoperative complications but also on cancer recurrence and long-term survival.

\subsubsection{Preoperative systemic inflammation}
Elevated levels of interleukin-6, alpha-1 antitrypsin and CRP and decreased levels of albumin and prealbumin before surgery have been reported to be associated with a more exaggerated postoperative systemic inflammatory response and infectious complications after major abdominal surgery.\parencite{haupt_association_1997}

Preoperative systemic inflammation has been reported to be associated with infectious complications in patients undergoing potentially curative surgery for colorectal cancer.\parencite{moyes_preoperative_2009} In a study of 455 patients, Moyes and coworkers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery. They postulated that several mechanisms may have a role including disregulation of cell-mediated immunity, impaired T-lymphocyte response, disorders in the complement pathway and possibly due to loss of lean tissue and protein as a consequence of systemic inflammation. Preoperative mGPS has also been shown to predict postoperative morbidity in patients undergoing oesophageal resection for cancer.\parencite{vashist_glasgow_2010}

\subsubsection{Postoperative systemic inflammation}
An exaggerated and persistent systemic inflammatory response in the early postoperative period is associated with an increased incidence of complications. One of the earliest studies comparing several 'acute-phase proteins' and their role in predicting postoperative complications reported that in patients who developed surgical inflammatory complications, CRP remained elevated after the third postoperative day while other acute-phase proteins such as ceruloplasmin and alpha-1 antitrypsin where not useful in monitoring the postoperative course.\parencite{fischer_quantitation_1976}

Further studies have established the value of monitoring trends in serum CRP levels in predicting complications after both elective and emergency surgery.\parencite{mustard_c-reactive_1987}

In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak.\parencite{welsch_c-reactive_2007} They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreaticojejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.

Similar findings have been reported after elective colorectal surgery\parencite{ortega-deballon_c-reactive_2010, woeste_increased_2010}, oesophago-gastric surgery\parencite{dutta_persistent_2011}, spinal surgery\parencite{meyer_c-reactive_1995,mok_use_2008}, neurosurgery\parencite{al-jabi_value_2010}, simultaneous pancreas-kidney transplantation\parencite{wullstein_high_2004}, stem-cell transplantation\parencite{mcneer_early_2010} and paediatric surgery\parencite{laporta_baez_c-reactive_2011}.

While CRP level between the third and fifth postoperative day has been reported to be most predictive of complications,  the complications themselves do not become clinically apparent until a later in the postperative course, ofter after the fifth postoperative period. This has led some authors to postulate that the elevated CRP levels may in fact be due to an abnormally modulated postoperative inflammatory response resulting in an initial exaggerated systemic inflammatory response syndrome (SIRS) followed by a compensatory anti-inflammatory response syndrome (CARS).

\subsubsection{Compensatory Anti-inflammatory Response Syndrome (CARS)}
The compensatory anti-inflammatory response syndrome is characterised by several features including reduction in lymphocyte numbers by apoptosis, decreased responsiveness of monocytes to cytokines, reduced number of human leukocyte antigen presenting receptors on monocytes, expression of cytokines that suppress Tumour Necrosis Factor (TNF) and clonal anergy.

In their seminal work on the role of SIRS and CARS in the pathogenesis of sepsis and organ dysfunction, Bone and co-workers described a state of 'immunologic dissonance' where a 'pre-primed' immune system may result in an inappropriate, out-of-balance massive pro-inflammatory response which is followed by a proportionately large compensatory anti-inflammatory response that leaves the patient immunosuppressed and prone to further organ dysfunction, infections and death. \parencite{bone_sepsis:_1997, bone_immunologic_1996} It is very likely that similar mechanisms are involved in surgical patients except that the initial stressor in this case is surgical trauma rather than a bacterial infection as in sepsis.

This form of 'immunoparalysis' was first described in patients after major trauma with tissue damage \parencite{abraham_effects_1985,bandyopadhyay_negative_2007} or after haemorrhage on its own without associated tissue trauma. \parencite{stephan_hemorrhage_1987} In a detailed review of the mechanisms underlying the compensatory anti-inflammatory response syndrome, Ward and coworkers describe SIRS and CARS to be mirror images suggesting that a disproportionately high SIRS is followed by a period of immunosuppression that leaves the patient prone to further complications.\parencite{ward_compensatory_2008}

Patients who developed infectious complications after major cancer surgery had higher levels of interleukin-10 (IL-10), an anti-inflammatory cytokine and marker of the compensatory anti-inflammatory process.\parencite{mokart_early_2002} Major surgery and the associated surgical trauma is associated with elevated levels of IL-10 which in turn is associated with increase in lymphocyte apoptosis \parencite{delogu_interleukin-10_2001}, reduced monocyte expression of HLA-DR antigens \parencite{klava_interleukin-10._1997} and a blunted response to endotoxins \parencite{ogata_role_2000, kawasaki_surgical_2001}, all considered to be key features of a compensatory anti-inflammatory response syndrome.

Yamaguchi and co-workers compared the levels of pro- and anti-inflammatory cytokines in patients undergoing cholecystectomy versus patients undergoing trans-thoracic oesophagectomy. They reported that the initial inflammatory phase was followed by an immunosuppressive phase that started around the seventh postoperative day in patients undergoing oesophagectomy. However, patient who underwent underwent an open cholecystectomy did not experience this immunosuppressive phase, leading them to postulate that the degree of immunosuppression was directly proportional to the intial pro-inflammatory process. This in turn was related to the greater degree of surgical stress and tissue trauma that occurs with a trans-thoracic oesophagectomy. They also reported that in a randomised cohort that received an infusion of lymphokine-activated natural killer cells immediately after oesophagectomy, there was a trend towards fewer infectious complications.\parencite{yamaguchi_postoperative_2006}

\subsection{Postoperative complications and long-term survival}
There has been increasing evidence that postoperative complications not only have an impact on the short-term outcomes but also on long-term survival after major cancer surgery. A recent meta-analysis of 21 studies including 21,902 patients found that anastomotic leakage was associated with earlier local recurrence after rectal cancer surgery, a trend towards early local recurrence in other colonic cancer surgery and a significant reduction in overall survival.\parencite{mirnezami_increased_2011} The reviewers suggested that several mechanisms may be involved in early recurrence including local spillage of cancer cells from within the bowel lumen. However, the role of the local inflammatory processes that occur as a consequence of anastomotic leakage may play a more important role. This inflammatory process with the attendant milieu of pro-inflammatory cytokines and angiogenic factors may provide a fertile ground for tumour seeding and proliferation.

McArdle and co-workers reported in their study of 2235 patients undergoing colorectal cancer surgery that anastomotic leakage was associated with early local recurrence and reduced survival. They suggested that the 'double-hit' of surgery followed by anastomotic leak may result in an inflammatory response that is greater and more protracted and that this may explain the poorer cancer outcomes in these patients.\parencite{mcardle_impact_2005} In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP 2 weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications.\parencite{katoh_anastomotic_2011} Wound infections and intra-abdominal infections have also been associated with poorer survival in colorectal cancer patients.\parencite{nespoli_impact_2006} Similar findings have been reported after curative surgery for advanced gastric cancer with patients who develop an anastomotic leak surviving for 30.5 months while patients who did not have a leak survived for a median of 96.2 months (p$<$0.001). \parencite{yoo_negative_2011}

Patients who develop severe postoperative complications after pancreaticoduodenectomy for cancer had significantly shortened survival in a study involving 428 patients (16.5 vs. 12.4 months, p=0.002) and this was independent of other recognised risk factors such as tumour grade and lymph node status. \parencite{kamphues_postoperative_2011} Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007} and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head. \parencite{kang_detrimental_2009}

These reports in conjunction with the studies on preoperative inflammation, sepsis, SIRS and CARS emphasise the important role of peri;operative systemic inflammation as a causative factor in postoperative complications and the impact of the 'second-hit' of postoperative complications on long-term survival after curative cancer surgery.

\subsection{Relationship between systemic inflammation and comorbidity}

\section{The Jaundiced Patient}
\subsection{Impact of jaundice on cardiovascular physiology}
\subsection{Impact of jaundice on renal physiology}
\subsection{Impact of jaundice on the immune system}
\subsection{Jaundice and postoperative outcomes}
\subsection{Role of preoperative biliary drainage}

### 2015-06-02 Tables/c1_clavien_dindo.tex
Latex files added for first time

\begin{sidewaystable}[htbp]
\caption{The Clavien-Dindo Classification of Surgical Complications}
\label{table:clavien-dindo}
\centering
\begin{tabular}{|l m{15cm}|}
 \hline
 Grade  &   Description  \\ \hline

 Grade I     &   Any deviation from the normal postoperative course without the need for pharmacological treatment or surgical, endoscopic and radiological interventions. \\

Grade II      &   Requiring pharmacological treatment with drugs other than such allowed for grade I complications. Blood transfusions and total parenteral nutrition are also included.\\ 
Grade III     &   Requiring surgical, endoscopic or radiological intervention\\
                &  Grade III-a: - intervention not under general anesthesia\\
                &  Grade III-b: - intervention under general anesthesia\\ 
Grade IV       &  Life-threatening complication (including CNS complications)‡ requiring IC/ICU-management\\
                &   Grade IV-a: - single organ dysfunction (including dialysis)\\
                &   Grade IV-b: - multi organ dysfunction\\ 
Grade V        &   Death of a patient\\ \hline
Suffix 'd':     &   If the patients suffers from a complication at the time of discharge,  the suffix  “d”  (for ‘disability’) is added to the respective grade of complication. This label indicates the need for a follow-up to fully evaluate the complication.\\ \hline

\end{tabular}

\end{sidewaystable}

### 2015-06-02 Tables/c1_cpet_parameters.tex
Latex files added for first time

\begin{table}[p]
\centering
\caption{Parameters measured at spirometry.}
\label{table:spirometry}
\begin{tabular}{| m{2cm}  m{2cm}  m{7cm} |}
	\hline
	Parameter & Units  & Description                          \\ \hline
	FVC       & litres & Forced Vital Capacity                \\
	FEV1      & litres & Forced Expiratory Volume in 1 second \\
	FEV1/FVC  & \%     & Tiffeneau-Pinelli[1] index           \\ \hline
\end{tabular}

\vspace{2cm}

\caption{Common parameters measured at cardiopulmonary exercise testing.}
\label{table:cpet_parameters}
\begin{tabular}{| m{2cm}  m{2cm}  m{7cm} |}
	\hline
	Parameter & Units       & Description                                  \\ \hline
	\%peakVO2 & \%          & VO2 as a \% of predicted VO2Peak             \\
	Load      & Watts       & Exercise Workload                            \\
	VE        & litres/min  & Ventilatory Equivalent                       \\
	Vt        & litres      & Tidal volume                                 \\
	VO2       & litres/min  & Absolute Oxygen uptake/consumption           \\
	VO2/kg    & ml/(kg*min) & Corrected Oxygen uptake/consumption          \\
	VE/VO2    &             & Ventilatory Equivalent for O$_2$             \\
	VCO2      & litres/min  & Carbon-dioxide output                        \\
	VE/VCO2   &             & Ventilatory Equivalent for CO$_2$            \\
	RER       &             & Respiratory Exchange Ratio                   \\
	PETO2     & mmHg        & End Tidal O2                                 \\
	PETCO2    & mmHg        & End Tidal CO2                                \\
	O2Pulse   & ml/beat     & Oxygen pulse                                 \\
	HR        & beats/min   & Heart Rate                                   \\
	Bf        & /min        & Breathing Frequency                          \\
	P(A-a)O2  & mmHg        & Alveolar-arterial PO2 difference             \\
	Vd/Vt     &             & Physiologic dead space-to-tidal volume ratio \\
	SBP       & mmHg        & Systolic blood pressure                      \\
	DBP       & mmHg        & Diastolic blood pressure                     \\
	O2sat     & \%          & Oxygen saturation                            \\ \hline
\end{tabular}

\end{table}

### 2015-06-02 Tables/c1_isgps.tex
Latex files added for first time

\begin{sidewaystable}[htbp]
\centering
\caption{Postoperative pancreatic fistula: ISGPF definition.}
\label{table:isgps_popf}
\begin{tabular}{|m{4cm}|m{4cm}|m{5cm}|m{5cm}|}
	\hline
	Grade                                & A        & B                 & C                 \\ \hline
	Clinical conditions                  & Well     & Often well        & Ill appearing/bad \\
	Specific treatment                   & No       & Yes/no            & Yes               \\
	US/CT (if obtained)                  & Negative & Negative/positive & Positive          \\
	Persistent drainage (after 3 weeks)† & No       & Usually yes       & Yes               \\
	Reoperation                          & No       & No                & Yes               \\
	Death related to POPF                & No       & No                & Possibly yes      \\
	Signs of infections                  & No       & Yes               & Yes               \\
	Sepsis                               & No       & No                & Yes               \\
	Readmission                          & No       & Yes/no            & Yes/no            \\ \hline
\end{tabular}

\caption{Postpancreatectomy haemorrhage: ISGPS definition.}
\label{table:isgps_pph}
\begin{tabular}{|m{4cm}|m{4cm}|m{5cm}|m{5cm}|}
	\hline
	Grade                                                             & A                                                                                & B                                                                                                                             & C                                                                                         \\ \hline
	Time of onset, location, severity and clinical impact of bleeding & Early, intra- or extraluminal, mild                                              & Early, intra- or extraluminal, severe	or Late, intra- or extraluminal, mild                                                   & Late, intra- or extraluminal, severe                                                      \\
	Clinical condition                                                & Well                                                                             & Often well/ intermediate, very rarely life-threatening                                                                        & Severely impaired, life-threatening                                                       \\
	Diagnostic consequence                                            & Observation, blood count, ultrasonography and, if necessary, computed tomography & Observation, blood count, ultrasonography, computed tomography, angiography, endoscopy                                        & Angiography, computed tomography, endoscopy                                               \\
	Therapeutic consequence                                           & No                                                                               & Transfusion of fluid/blood, intermediate care unit (or ICU), therapeutic endoscopy,† embolization, relaparotomy for early PPH & Localization of bleeding, angiography and embolization, (endoscopy†) or relaparotomy, ICU \\ \hline
\end{tabular}
\end{sidewaystable}

### 2015-06-02 main.tex
Latex files added for first time

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Masters/Doctoral Thesis 
% LaTeX Template
% Version 1.43 (17/5/14)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original authors:
% Steven Gunn 
% http://users.ecs.soton.ac.uk/srg/softwaretools/document/templates/
% and
% Sunil Patel
% http://www.sunilpatel.co.uk/thesis-template/
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
% Note:
% Make sure to edit document variables in the Thesis.cls file
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[12pt, oneside]{Thesis} % The default font size, sides to print and document class

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%--------------------PACKAGES------------------------------------------------------------

\usepackage[backend=biber,
			style=authoryear,
			url=false]{biblatex}
%For Square Brackets around citations
%http://tex.stackexchange.com/questions/16765/biblatex-author-year-square-brackets
\makeatletter

\newrobustcmd*{\parentexttrack}[1]{%
	\begingroup
	\blx@blxinit
	\blx@setsfcodes
	\blx@bibopenparen#1\blx@bibcloseparen
	\endgroup}

\AtEveryCite{%
	\let\parentext=\parentexttrack%
	\let\bibopenparen=\bibopenbracket%
	\let\bibcloseparen=\bibclosebracket}

\makeatother
%------------------------------------------------------

\usepackage{graphicx}
\usepackage{caption} 
\usepackage{subcaption} % For Subfigures - not sure if I have used this anywhere
\usepackage{parskip} %Adds extra space between paragraphs
\usepackage{listings} %This is for the VBA code in the appendices
\usepackage{amsmath} %For aligning all equations on the equal to sign
\usepackage{pdfpages} %for including pdf files or individual pages of pdf files
\usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
\linenumbers

%--------------Table customisations------------------------------------------------------
% The following allows creation of columns of defined width within tables
%http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\usepackage{array}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

% The next few lines are for customising tables
\usepackage[justification=justified,singlelinecheck=false]{caption}
\usepackage{rotating} % for landscape tables use begin{sidewaystable}
%----------------Figures and hyperlinks----------------------------------------------------
%\graphicspath{Figures/} % Specifies the directory where pictures are stored
%\hypersetup{urlcolor=black, colorlinks=true} % Colors hyperlinks in blue - change to black if annoying

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%Bibligraphy file from Zotero
\addbibresource{zot.bib}
%----------------------------------------------------------------------------------------

\title{\ttitle} % Defines the thesis title - don't touch this

%----------------------------------------------------------------------------------------
% BEGIN DOCUMENT -- BEGIN DOCUMENT -- BEGIN DOCUMENT -- BEGIN DOCUMENT -- BEGIN DOCUMENT
%----------------------------------------------------------------------------------------
\begin{document}

\frontmatter % Use roman page numbering style (i, ii, iii, iv...) for the pre-content pages

\setstretch{2.0} % Line spacing of 2.0 although I have set \doublespace set somewhere else

% Define the page headers using the FancyHdr package and set up for one-sided printing
\fancyhead{} % Clears all page headers and footers
\rhead{\thepage} % Sets the right side header to show the page number
\lhead{} % Clears the left side page header

\pagestyle{fancy} % Finally, use the "fancy" page style to implement the FancyHdr headers

\newcommand{\HRule}{\rule{\linewidth}{0.5mm}} % New command to make the lines in the title page

% PDF meta-data
\hypersetup{pdftitle={\ttitle}}
\hypersetup{pdfsubject=\subjectname}
\hypersetup{pdfauthor=\authornames}
\hypersetup{pdfkeywords=\keywordnames}

%----------------------------------------------------------------------------------------
%	TITLE PAGE - Reformatted to Univ of Glasgow guidelines
%----------------------------------------------------------------------------------------

\begin{titlepage}
	\begin{center}

	\HRule \\[0.4cm] % Horizontal line
	{\huge \bfseries \ttitle}\\[0.4cm] % Thesis title
	\HRule \\[1.5cm] % Horizontal line

	\begin{minipage}{1\textwidth}
		\begin{center} \large
			\setstretch{1.3}
			\textbf{by} \\
			\vspace{1.1cm}
			\textbf{\authornames}\\
			\textbf{M.B.B.S., M.R.C.S., Dip.N.B.}\\
			\vspace{1.1cm}
			\normalsize
			\textbf{SUBMITTED IN FULFILMENT OF THE REQUIREMENTS}\\
			\textbf{FOR THE DEGREE OF DOCTOR OF MEDICINE}\\
			\vspace{1.1cm}
			\textbf{to}\\
			\vspace{1.1cm}
			\textbf{THE UNIVERSITY OF GLASGOW}\\
			\vspace{1.1cm}
			\textbf{BASED ON RESEARCH CONDUCTED IN THE} \\
			\textbf{UNIVERSITY DEPARTMENT OF SURGERY,}\\
			\textbf{GLASGOW ROYAL INFIRMARY}\\
			\textbf{AUGUST 2015}\\
			\vspace{1.1cm}
			\textcopyright \textbf{Vishnu V Chandrabalan 2015}
		\end{center}

	\end{minipage}
	\vfill
	\end{center}

\end{titlepage}

%----------------------------------------------------------------------------------------
%	QUOTATION PAGE
%----------------------------------------------------------------------------------------

\pagestyle{empty} % No headers or footers for the following pages

\dedicatory{Pebbles on the beach.}%Never again.
\clearpage % Start a new page

%----------------------------------------------------------------------------------------
%	DEDICATION
%----------------------------------------------------------------------------------------

\pagestyle{empty} % Page style needs to be empty for this page

\dedicatory{Dedicated to A, I, A} % Dedication text

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

%----------------------------------------------------------------------------------------
%	ABSTRACT PAGE
%----------------------------------------------------------------------------------------

\addtotoc{Abstract} % Add the "Abstract" page entry to the Contents

\abstract{\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics

To be finalised...
}
\clearpage

%----------------------------------------------------------------------------------------
%	LIST OF CONTENTS/FIGURES/TABLES PAGES
%----------------------------------------------------------------------------------------

\pagestyle{fancy} % The page style headers have been "empty" all this time, now use the "fancy" headers as defined before to bring them back

\lhead{\emph{Contents}} % Set the left side page header to "Contents"
\tableofcontents % Write out the Table of Contents

\lhead{\emph{List of Figures}} % Set the left side page header to "List of Figures"
\listoffigures % Write out the List of Figures

\lhead{\emph{List of Tables}} % Set the left side page header to "List of Tables"
\listoftables % Write out the List of Tables

\clearpage

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\acknowledgements{\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics

I would like to thank the following people, for their help, advice and encouragement:

Professor Donald C McMillan,\\
University Department of Surgery, Glasgow Royal Infirmary

Professor Paul Horgan,\\
University Department of Surgery, Glasgow Royal Infirmary
}

\clearpage % Start a new page

%----------------------------------------------------------------------------------------
%	DECLARATION PAGE
%	Your institution may give you a different text to place here
%----------------------------------------------------------------------------------------

\Declaration{

	\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics

	I declare that the work presented in this thesis was carried out solely by me, as a clinical research fellow in the University Dept of Surgery, Royal Infirmary, Glasgow, except where indicated below:

	Measurement of biochemical and haematological data was performed by the hospital laboratory service.

	Statistical analysis was performed with the assistance of Prof Donald C McMillan, University Dept of Surgery, Royal Infirmary, Glasgow.

	In addition, no work referred to in this thesis has been submitted in support of an application for another degree or qualification in this or any other university.
}

%----------------------------------------------------------------------------------------
%	ABBREVIATIONS
%----------------------------------------------------------------------------------------

\input{Chapters/Abbreviations}

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - CHAPTERS
%----------------------------------------------------------------------------------------

\mainmatter % Begin numeric (1,2,3...) page numbering

\pagestyle{fancy} % Return the page headers back to the "fancy" style

% Include the chapters of the thesis as separate files from the Chapters folder
\input{Chapters/Introduction}
\input{Chapters/CPET_Outcomes} 
\input{Chapters/CPET_Jaundice}
\input{Chapters/BodyComp}
\input{Chapters/CPET_Survival}
\input{Chapters/Discussion}

%----------------------------------------------------------------------------------------
%	THESIS CONTENT - APPENDICES
%----------------------------------------------------------------------------------------

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

\appendix % Cue to tell LaTeX that the following 'chapters' are Appendices

%\input{Appendices/AppendixExcelMacros}
\input{Appendices/AppendixCPETRawData}
%\input{Appendices/AppendixC}

\addtocontents{toc}{\vspace{2em}} % Add a gap in the Contents, for aesthetics

\backmatter

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\label{Bibliography}

\lhead{\emph{Bibliography}} % Change the page header to say "Bibliography"

%\bibliographystyle{unsrtnat} % Use the "unsrtnat" BibTeX style for formatting the Bibliography
\printbibliography % The references (bibliography) information are stored in the file named "Bibliography.bib"

\end{document}

### 2015-06-02 Chapters/Introduction.tex
Jaundice introduction - first sentence only
Committing to Jaundice branch

@@
 \section{The Jaundiced Patient}
+Obstructive jaundice is the most common presenting symptom in patients with pancreatic cancer involving the head of the pancreas and the periampullary region due to the anatomical location of the distal bile duct.
+Obstructive jaundice is known to have a wide range of effects on multiple organ systems including the cardiovascular system, immune system, coagulation cascade, as well as hepatic function.
+
 \subsection{Impact of jaundice on cardiovascular physiology}

### 2015-06-02 Chapters/Introduction.tex
complete first paragraph of jaundice introduction

@@
 
+Until recently, major surgery in the jaundiced patient has been considered to be more prone to adverse postoperative events.
+While this concept has been recently challenged, surgeons remain wary of the severely jaundiced patient.
+
+Pancreaticoduodenectomy was initially described as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out.
+
 \subsection{Impact of jaundice on cardiovascular physiology}

### 2015-06-02 Chapters/Introduction.tex
Testing bash

@@
 
-Pancreaticoduodenectomy was initially described as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out.
+Pancreaticoduodenectomy -Whipple was initially described as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out.
 

### 2015-06-02 Chapters/Introduction.tex
Added papers to read re jaundice physiology - remove these in final version.

--- 2015-06-02 Chapters/Introduction.tex
@@
 
+Until recently, major surgery in jaundiced patients has been considered to be more prone to adverse postoperative events.
+While this concept has been recently challenged, surgeons remain wary of operating on the severely jaundiced patient.
+In fact, pancreaticoduodenectomy was initially described as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out.
+
 \subsection{Impact of jaundice on cardiovascular physiology}
+%Read the following papers:
+%yang_clinical_2010
+%padillo_improved_2001
+%khurana_bile_2011
+%green_jaundiced_1986
+
 \subsection{Impact of jaundice on renal physiology}
+
 \subsection{Impact of jaundice on the immune system}
+%nehez_compromise_2002 - get other citations from this one
+
 \subsection{Jaundice and postoperative outcomes}
 \subsection{Role of preoperative biliary drainage}
+
+\section{Body Composition}
+Make this a brief description of body composition.

### 2015-06-03 Chapters/Introduction.tex
Impact of jaundice on renal physiology

@@
 
+%clarke_current_2006 - This is probably the best paper to base this entire section on.
+
 \subsection{Impact of jaundice on cardiovascular physiology}
@@
 
+Several studies have reported that obstructive jaundice is associated with significant abnormalities in fluid homeostasis.
+Obstructive jaundice is associated with reduction in the interstitial volume as well as the circulating plasma volume.\parencite{sitges-serra_body_1992, padillo_preoperative_1999} In a study of 63 patients with obstructive jaundice, Padillo and co-workers reported that severity of jaundice, age of patient and reduced urinary sodium excretion were independently related to postoperative renal dysfunction.
+They also reported that these variables were related to abnormalities in the levels of hormones responsible for sodium and water homeostasis including atrial natriuretic peptide.\parencite{padillo_multivariate_2005} Endotoxemia consequent to lack of bile salts in the gut has also been postulated as a possible mechanism for renal dysfunction in patients with severe jaundice.\parencite{bailey_endotoxin_1976}
+
+While expansion of the intravascular volume and avoiding dehydration by the judicious use of intravenous fluids has been recommended in avoiding renal dysfunction in jaundiced patients \parencite{parks_prospective_1994}, others have reported that relief of obstructive jaundice by restoring bile flow improves renal function independent of fluid therapy.\parencite{padillo_randomized_2005} Nonetheless, adequate perioperative fluid management, avoidance of hypotension, sepsis control and relief of obstructive jaundice play important roles in the prevention of renal failure in jaundiced patients.
+
 \subsection{Impact of jaundice on the immune system}
@@
 \subsection{Role of preoperative biliary drainage}
+%fang_pre-operative_2012 - Cochrane review
+%wang_preoperative_2008 - Cochrane review
 

### 2015-06-03 Chapters/Introduction.tex
Jaundice and cardiovascular physiology Please enter the commit message for your changes. Lines starting

@@
 
-\subsection{Impact of jaundice on renal physiology}
+The detrimental effects of jaundice on the heart and the circuilatory system has been recognised for over 150 years now.
+King and coworkers reported that the intravenous injection of bile caused bradycardia, hypotension and eventually death in dogs.\parencite{king_effect_1909} The concept of a \textit{'jaundiced heart'} was fist put forth by Green and coworkers in 1986.\parencite{green_jaundiced_1986} They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure.
+They reported that 'cholemia' was associated with impaired left ventricular function and blunted reponse to sympathomimmetic agents.
+Similar findings have been reported by other authors in animal studies.
+\parencite{binah_obstructive_1985,bomzon_systemic_1986}
+
+The role of atrial natriuretic peptide (ANP) has also been investigated.
+Obstructive jaundice is associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits.\parencite{pereira_increased_1994} Similar findings have been reported in humans as well.
+\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998)
+
+Moreover, relief of obstructive jaundice is associated with improvement in endocrine markers of fluid homeostasis as well as cardiac function.
+\parencite{padillo_improved_2001, gallardo_increased_1998) Padillo and co-workers reported that there was a negative correlation between serum bilirubin levels and left ventricular systolic work and this was associated with elevated ANP and BNP (brain natriuretic peptide) levels.
+However, both ANP and BNP levels decreased after biliary drainage and there was a significant improvement in cardiac output, cardiac index, systolic volume and left ventricular systolic work.\parencite{padillo_improved_2001}
+
+More recently, TNF-$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF-$\alpha$ agent restored myocardial contractility.\parencite{yang_mechanisms_2010} Obstructive jaundice is also associated with systemic hypotension and there has been increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree.\parencite{green_systemic_1995, lefebvre_role_2009} Bile acids can thus cause vasodilation by decreasing arterial tone and this may partly explain some of the haemodynamic adverse events that occur after surgery in jaundiced patients.
+
+\subsection{Impact of jaundice on renal function}
 
@@
 Obstructive jaundice is associated with reduction in the interstitial volume as well as the circulating plasma volume.\parencite{sitges-serra_body_1992, padillo_preoperative_1999} In a study of 63 patients with obstructive jaundice, Padillo and co-workers reported that severity of jaundice, age of patient and reduced urinary sodium excretion were independently related to postoperative renal dysfunction.
-They also reported that these variables were related to abnormalities in the levels of hormones responsible for sodium and water homeostasis including atrial natriuretic peptide.\parencite{padillo_multivariate_2005} Endotoxemia consequent to lack of bile salts in the gut has also been postulated as a possible mechanism for renal dysfunction in patients with severe jaundice.\parencite{bailey_endotoxin_1976}
+They also reported that these variables were related to abnormalities in the levels of hormones responsible for sodium and water homeostasis including atrial natriuretic peptide.\parencite{padillo_multivariate_2005} Endotoxemia consequent to lack of bile salts in the gut has also been postulated as a possible mechanism for renal dysfunction in patients with severe jaundice.\parencite{bailey_endotoxin_1976} Jaundice is one of the leading causes of acute renal failure in tertiary-care hospitals.\parencite{liano_epidemiology_1996}
 

### 2015-06-03 Chapters/Introduction.tex
Bibilography file updated

@@
 Obstructive jaundice is associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits.\parencite{pereira_increased_1994} Similar findings have been reported in humans as well.
-\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998)
+\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}
 
 Moreover, relief of obstructive jaundice is associated with improvement in endocrine markers of fluid homeostasis as well as cardiac function.
-\parencite{padillo_improved_2001, gallardo_increased_1998) Padillo and co-workers reported that there was a negative correlation between serum bilirubin levels and left ventricular systolic work and this was associated with elevated ANP and BNP (brain natriuretic peptide) levels.
+\parencite{padillo_improved_2001, gallardo_increased_1998} Padillo and co-workers reported that there was a negative correlation between serum bilirubin levels and left ventricular systolic work and this was associated with elevated ANP and BNP (brain natriuretic peptide) levels.
 However, both ANP and BNP levels decreased after biliary drainage and there was a significant improvement in cardiac output, cardiac index, systolic volume and left ventricular systolic work.\parencite{padillo_improved_2001}

### 2015-06-04 Chapters/Introduction.tex
More updates to introduction

@@
 
-\subsection{Jaundice and postoperative outcomes}
+\subsection{Role of preoperative biliary drainage and outcome after pancreaticoduodenectomy}
+The increased incidence of adverse events after surgery in patients with obstructive jaundice resulted in routine preoperative biliary drainage becoming the standard practice in patients undergoing major pancreatic surgery.
+However, several studies over the past 20 years have challenged this paradigm.
+There appears to be increasing evidence that preoperative biliary drainage may be associated with increased morbidity both as a consequence of the drainage procedure itself as well as from infectious complications after surgery.
+
+The paucity of good quality evidence on the role of preoperative biliary drainage in patients with obstructive jaundice undergoing major surgery and the lack of clear guidelines on the perioperative management of the jaundiced patients has resulted in wide variation in practice across different centres.
+
+The effect of obstructive jaundice on cardiopulmonary exercise testing in patients with pancreatic disease has not been reported previously.
+Chapter \ref{ch_cpet_jaundice} reports on the investigation into the relationship between obstructive jaundice and preoperative cardiopulmonary exercise testing.
+
 \subsection{Role of preoperative biliary drainage}

### 2015-06-04 Chapters/Introduction.tex

@@
 \section{The Jaundiced Patient}
+
 Obstructive jaundice is the most common presenting symptom in patients with pancreatic cancer involving the head of the pancreas and the periampullary region due to the anatomical location of the distal bile duct.

### 2015-06-04 Chapters/Introduction.tex
CPET General Abdominal Surgery - Complete I think

@@
 \subsection{General Surgery}
+Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery.
+
+Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery.
+All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring.
+The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg.
+There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes.
+They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$).
+Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993}
+In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise.
+High risk patients ($\dot{V}_{O_2}AT<11$ml/kg/min, n=153) were admitted to intensive care after surgery and had a mortality due to cardiovascular causes of 4.6\%.
+Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%.
+Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality.\parencite{older_cardiopulmonary_1999} Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient.
+\parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}
+
+These findings have since been replicated in several other studies although the threshold value of $\dot{V}_{O_2}$AT as well as the inclusion of other CPET parameters to attribute risk differ in these studies.
+
+Hightower and co-workers studied 32 patients over the age of 18 undergoing a variety of major elective abdominal surgery.
+In this heterogenous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic thresold were all independently associated with postoperative morbidity.
+While it is difficult to extrapolate results from this small heterogenous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients.\parencite{hightower_pilot_2010}
+
+Snowden and coworkers reported in a study of 171 patients of who 123 underwent surgery, that $\dot{V}_{O_2}AT<10.1$ ml/kg/min was associated with not only cardiovascular complications but also with other complications including pulmonary, renal, gastrointestinal, infective and haematological complications.
+One of the strengths of this study was the fact that the clinicians treating these patients were blinded to the results of preoperative cardiopulmonary exercise testing results thus avoiding management bias.
+They also included other cardiopulmonary exercise test parameters such as $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ at the anaerobic threshold in addition to $\dot{V}_{O_2}$AT in their analysis and showed that $\dot{V}_{O_2}$AT was more predictive of postoperative morbidity than POSSUM derived morbidity, cardiac risk scoring index or a validated activity questionnaire.
+However, this study also included a heterogenous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas.\parencite{snowden_submaximal_2010}
+
+In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9).
+This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
+Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) and in patients without ischaemic heart disease.
+It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.\parencite{wilson_impaired_2010}
+
 \subsection{Oesophago-gastric Surgery}
@@
 \subsection{Hepato-pancreato-biliary surgery and Transplantation}
+Preoperative cardiopulmonary exercise testing is associated with increased 100-day mortality in patients undergoing liver transplantation.\parencite{epstein_aerobic_2004} Patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery.
+\parencite{iscar_functional_2009} However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
 \subsection{Thoracic Surgery}

### 2015-06-04 Chapters/Introduction.tex
IntroCPET - HPB changed to liver transplantation and section completed, No studies in HPB as of 2010.

@@
 \subsection{Oesophago-gastric Surgery}
+
 \subsection{Colorectal Surgery}
 \subsection{Vascular Surgery}
-\subsection{Hepato-pancreato-biliary surgery and Transplantation}
-Preoperative cardiopulmonary exercise testing is associated with increased 100-day mortality in patients undergoing liver transplantation.\parencite{epstein_aerobic_2004} Patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery.
+\subsection{Liver Transplantation}
+Peak $\dot{V}_{O_2}<60\%$ predicted and $\dot{V}_{O_2}AT<50\%$ predicted have been reported to be associated with increased 100-day mortality in patients undergoing liver transplantation.
+In patients with cirrhosis awaiting hepatic transplantation, reduced aerobic capacity may not only be due to primary cardiorespiratory insufficiency but may also be due to the secondary effects of hepatic dysfunction itself.
+These include cirrhotic cardiomyopathy, hepatopulmonary syndrome and decreased peripheral oxygen utilisation due to cirrhotic myopathy.\parencite{epstein_aerobic_2004} This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery.
 \parencite{iscar_functional_2009} However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.

### 2015-06-04 Chapters/Introduction.tex
OG and bariatric section complete

@@
 
-In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9).
-This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
+In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9).\parencite{wilson_impaired_2010} This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
 Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) and in patients without ischaemic heart disease.
-It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.\parencite{wilson_impaired_2010}
+It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.
 
-\subsection{Oesophago-gastric Surgery}
+\subsection{Oesophago-gastric and bariatric surgery}
+In a retrospective study involving 91 patients who underwent curative oesophagectomy with 3-field lymphadenectomy via a
+right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2}<800$ml/min/m$^2$ predicted cardiopulmonary complications.\parencite{nagamatsu_preoperative_2001} They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications.
+They recommended that in patients with $\dot{V}_{O_2}max/m^2}<800$ml/min/m$^2$, surgical treatment be modified either into a 2-stage procedure or a trans-hiatal procedure avoiding a thoracotomy where possible.
+
+Cardiopulmonary exercise testing has also been useful in predicting short-term complications after bariatric surgery.
+McCullough and co-workers defined a composite primary outcome measure that included myocardial infarction, unstable angina, deep vein thrombosis, pulmonary embolism, renal failure, stroke and death and applied this in a cohort of 109 consecutive patients who underwent Roux-en-Y gastric bypass surgery.
+This composite adverse outcome was more likely in patients with high BMI ($>45$) and low $\dot{V}_{O_2}$Peak ($<15.9 ml/kg/min$).
+In fact, patients with neither of these two risk factors had no complications after surgery.
+They also observed a significant negative relationship between BMI and $\dot{V}_{O_2}$ ($p<0.0001$) and age($p<0.0001$).
+Poor aerobic capacity was also associated with the presence of diabetes and hypertension suggesting further evidence of the metabolic cost of obesity in these patients.\parencite{mccullough_cardiorespiratory_2006}
+
+However, Forshaw and co-workers reported that cardiopulmonary exercise testing was of limited use in predicting complications after oesophagectomy.\parencite{forshaw_is_2008} They studied 78 patients undergoing either trans-hiatal (n=39) or trans-thoracic (n=39) oesophagectomy.
+They evaluated several thresholds for both $\dot{V}_{O_2}$Peak and $\dot{V}_{O_2}$AT and found no useful correlation with cardiopulmonary complications, non-cardiopulmonary complications, length of stay in critical care or in hospital.
+While $\dot{V}_{O_2}$Peak was significantly lower in patients who developed postoperative cardiopulmonary complications, receiver-operator characteristics (ROC) analysis did not identify a clinically useful threshold that would stratify patients into different risk groups.
+They postulated that the fact that post-oesophagectomy complications such as anastomotic leak or sepsis often happen within the thorax and are not necessarily due to cardiopulmonary dysfunction, preoperative cardiopulmonary exercise testing may not have identified these patients.
+Older and Hall, the pioneers of cardiopulmonary exercise test in surgical patients, also raise this issue in their letter to the authors of the above study.
+\parencite{hall_cardiopulmonary_2009}
 

### 2015-06-04 Chapters/Introduction.tex
Corrected a few missing curly brackets - that's all. And, pdf has been added to gitignore

@@
 In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9).\parencite{wilson_impaired_2010} This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
-Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) and in patients without ischaemic heart disease.
+Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) {a}nd in patients without ischaemic heart disease.
 It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.
@@
 In a retrospective study involving 91 patients who underwent curative oesophagectomy with 3-field lymphadenectomy via a
-right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2}<800$ml/min/m$^2$ predicted cardiopulmonary complications.\parencite{nagamatsu_preoperative_2001} They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications.
-They recommended that in patients with $\dot{V}_{O_2}max/m^2}<800$ml/min/m$^2$, surgical treatment be modified either into a 2-stage procedure or a trans-hiatal procedure avoiding a thoracotomy where possible.
+right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$ predicted cardiopulmonary complications.\parencite{nagamatsu_preoperative_2001} They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications.
+They recommended that in patients with $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$, surgical treatment be modified either into a 2-stage procedure or a trans-hiatal procedure avoiding a thoracotomy where possible.
 

### 2015-06-05 Chapters/Introduction.tex
Modified main.tex to include several package options for bibiblatex, including maxcitenames, mincitenames, uniquelist and uniquenames options - may need modification.
Introduction - CPET Vascular Surgery
- Add few studies pre 2010.
- Several studies post 2010 in zotero but not discussed.
- Check with Donny re adding studies post 2010

@@
 These findings have since been replicated in several other studies although the threshold value of $\dot{V}_{O_2}$AT as well as the inclusion of other CPET parameters to attribute risk differ in these studies.
-
 Hightower and co-workers studied 32 patients over the age of 18 undergoing a variety of major elective abdominal surgery.
@@
 
-\subsection{Colorectal Surgery}
 \subsection{Vascular Surgery}
+The earliest report of the application of cardiopulmonary exercise test in elective abdominal aortic aneurysm surgery was by Nugent and co-workers in 1998.
+They reported on cardiopulmonary exercise testing using a treadmill in 36 patients undergoing elective abdominal aortic aneurysm surgery.
+They found no difference in the $\dot{V}_{O_2}$Peak in patients who developed complications versus those who did not (18.6 vs 2.8 ml/kg/min, p$>$0.05).
+However, they reported that in the 4 patient who were denied surgery as they were deemed medically unfit, $\dot{V}_{O_2}$Peak was less than 20 ml/kg/min.
+However, this threshold did not discriminate patients who developed complications and the authors concluded that cardiopulmonary exercise testing should not be used on its own to guide clinical management of these patients.\parencite{nugent_cardiopulmonary_1998}
+
+However, cardiopulmonary exercise testing has been reported to predict 2-year survival in patients undergoing elective aortic aneurysm surgery.
+Carlisle and Swart reported that a Revised Cardiac Risk Index (RCRI) score $>1$ in combination with a $\dot{V}_E/\dot{V}_{CO_2}>42$ was associated with a 2-year survival rate of 55\% while the 2-year survival rate was 97\% if neither risk factor was present.
+$\dot{V}_{O_2}$AT was not a predictor of survival in multi-variate analysis in this study.\parencite{carlisle_mid-term_2007}
+
+While studies have reported that repeated cardiopulmonary exercise testing provides reproducible results in patients with abdominal aortic aneurysm and preoperative exercise regimens are safe and effective at increasing aerobic capacity in patients awaiting surgery, there is no published literature on the association between cardiopulmonary exercise testing and postoperative complications in patients undergoing vascular surgery.
+%True as of 2010.
+There are several papers post 2010.
+
 \subsection{Liver Transplantation}
@@
 \parencite{iscar_functional_2009} However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
+
 \subsection{Thoracic Surgery}

### 2015-06-05 Chapters/Introduction.tex
CPET in Thoracic Surgery - complete
- 2 reviews included
- First - a meta-analysis by Benzo et al
- Second, ACCP recommendations by Colice et al

@@
 \subsection{Thoracic Surgery}
+The role of cardiopulmonary exercise testing in patient undergoing non-cardiac thoracic surgery has been well established for over a decade.
+In a recent meta-analysis of 14 studies, Benzo and co-workers reported on 955 patients who underwent cardiopulmonary exercise tests as part of their preoperative work-up before lung resection for cancer.
+$\dot{V}_{O_2}$Max (ml/kg/min), $\dot{V}_{O_2}$Max (\% predicted), maximum exercise load achieved (Watts) and FEV1 (\% predicted) were all significantly higher in patients who recovered with no postoperative complications.
+They also found that the mean $\dot{V}_{O_2}$Max in patients who did not develop complications was approximately 20 ml/kg/min and the mean $\dot{V}_{O_2}$Max in patients who developed complications was 15 ml/kg/min across all studies.
+This would suggest that cardiopulmonary exercise testing can offer clinically useful thresholds across different populations that can be used to risk stratify patients.\parencite{benzo_complications_2007}
+
+The 2$^nd$ edition of the American College of Chest Physicians' Evidence-Based Clinical Practice Guidelines for the 'Physiologic evaluation of the patient with lung cancer being considered for resectional surgery' recommends that patients with a low $\dot{V}_{O_2}$Max ($<10$ ml/kg/min) or with low $\dot{V}_{O_2}$Max ($<15$ ml/kg/min) associated with other risk factors on spirometry and cardiopulmonary exercise testing should be counselled about non-operative treatment options and non-standard surgery as the risk of perioperative death and cardiopulmonary complications was very high.
+The 1C grading of this recommendation not only represents the observational nature of most of the studies it is based on but also that such observational evidence is sufficiently consistent\parencite{benzo_complications_2007} to make a strong recommendation.\parencite{colice_physiologic_2007}
 

### 2015-06-06 Chapters/Introduction.tex
Introduction - Preoperative biliary drainage
- Complete section
- Merge this section with jaundice and outcomes section

@@
 
-\subsection{Role of preoperative biliary drainage and outcome after pancreaticoduodenectomy}
-The increased incidence of adverse events after surgery in patients with obstructive jaundice resulted in routine preoperative biliary drainage becoming the standard practice in patients undergoing major pancreatic surgery.
-However, several studies over the past 20 years have challenged this paradigm.
-There appears to be increasing evidence that preoperative biliary drainage may be associated with increased morbidity both as a consequence of the drainage procedure itself as well as from infectious complications after surgery.
-
-The paucity of good quality evidence on the role of preoperative biliary drainage in patients with obstructive jaundice undergoing major surgery and the lack of clear guidelines on the perioperative management of the jaundiced patients has resulted in wide variation in practice across different centres.
-
-The effect of obstructive jaundice on cardiopulmonary exercise testing in patients with pancreatic disease has not been reported previously.
-Chapter \ref{ch_cpet_jaundice} reports on the investigation into the relationship between obstructive jaundice and preoperative cardiopulmonary exercise testing.
-
 \subsection{Role of preoperative biliary drainage}
@@
 %wang_preoperative_2008 - Cochrane review
+The increased incidence of adverse events after surgery in patients with obstructive jaundice resulted in routine preoperative biliary drainage becoming the standard practice in patients undergoing major pancreatic surgery.
+Moreover, while significant improvements in perioperative care has resulted in reduction in postoperative mortality, morbidity rates have remained high.
+Obstructive jaundice has provided a potential target for preoperative optimisation in an attempt to improve postoperative outcomes.
+
+Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated ith increased risk and in fact preoperative biliary drainage increased hospitalisation costs.
+\parencite{snellen_influence_1985, bakkevold_morbidity_1993, pitt_does_1985} Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications.
+\parencite{hatfield_preoperative_1982}
+
+Similar findings have been reported by other authors as well.
+\parencite{lai_preoperative_1994, lai_preoperative_1994, jagannath_effect_2005} However, some authors have recommended that expandable metal stents were associated with fewer stent related complications than plastic stents and may be considered in jaundiced patients where surgery is delayed for other reasons.
+\parencite{wasan_use_2005, mullen_pancreaticoduodenectomy_2005}
+
+A recent randomised controlled trial of 202 patients reported that there was no significant difference in post-operative morbidity (47\% vs, 37\%, p=0.14), mortality or length of stay between the group that underwent preoperative biliary drainage and the group that underwent early surgery.
+There was a high incidence of procedure related complications in the drainage group (46\%).
+The high incidence of procedure related complications, the use of plastic stents, the 25\% failure to drain at first attempt and exclusion of patients with serum bilirubin $>$ 250 $\mu$mol/L were significant limitations of the study.
+However, the study does show that early surgery in the jaundiced patient with serum bilirubin $<$ 250 $\mu$mol/L is safe and not associated with increased morbidity or mortality as once feared.
+Mezhir and co-workers also reported a high (23\%) stent-related complication rate in their retrospective study of 340 patients where 96\% of 201 patients in the biliary drainage group had plastic stents placed.
+However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage.
+
+The paucity of good quality evidence on the role of preoperative biliary drainage before pancreatic resection \parencite{wang_preoperative_2008} and the limitations in our understanding of the complex patho-physiological effects of jaundice on multiple organ systems has resulted in persistent wide variation in the perioperative management of jaundiced patients across different centres.
+
+The historical indication was biliary drainage was to minimise the haemodynamic and renal adverse effects of obstructive jaundice.
+The above mentioned studies show that surgery in the jaundiced patient is safe and feasible.
+However, the effect of obstructive jaundice on cardiopulmonary exercise testing in patients with pancreatic disease has not been reported previously.
+Chapter \ref{ch_cpet_jaundice} is an investigation into the relationship between obstructive jaundice and preoperative cardiopulmonary exercise testing.
 

### 2015-06-06 Chapters/BodyComp.tex
Chapter Body Composition - numerous small changes - fixed all VO2 to correct format - fixed some problems with style - fixed a few spelling/grammar issues - needs a final check - see comments at end of discussion - consider separate tables for each phase of exercise

@@
 %----------------------------------------------------------------------------------------
-
+$\dot{V}_{O_2}$
 \section{Introduction}
@@
 The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established.
-A number of studies have shown that poor aerobic fitness demonstrated by a low anaerobic threshold or low peak VO$_2$ or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
+A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
 Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.
@@
 \subsection{Factors influencing aerobic fitness}
-A low anaerobic threshold and/or low peak VO$_2$ have universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise.
-This is often thought to be due to cardiorespiratory disease either overt or subclinical.
-Occasionally other factors such as anaemia, peripheral vascular disease and rarely mitochondrial diseases have been recognised as factors contributing to low anaerobic threshold/peak VO$_2$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but this is uncommon in patients undergoing major abdominal surgery.
+A low $\dot{V}_{O_2}$ has universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise.
+This is often thought to be due to cardiorespiratory disease, overt or sub-clinical.
+Occasionally, other factors such as anaemia, peripheral vascular disease and rarely, mitochondrial diseases have been recognised as factors contributing to low $\dot{V}_{O_2}$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but these causes are less common in patients undergoing major abdominal surgery.
 
-The most common parameters used to quantify perioperative risk in surgical patients are oxygen consumption at the anaerobic threshold (VO$_2$AT) and at peak exercise capacity (VO$_2$Peak).
-Conventionally these have been reported as per weight ratios in mls/kg/min.
-However, numerous studies on cardiorespiratory exercise physiology have reported that normalising VO$_2$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}
+The most common parameters used to quantify perioperative risk in surgical patients are oxygen consumption at the anaerobic threshold ($\dot{V}_{O_2}$AT) and at peak exercise capacity ($\dot{V}_{O_2}$Peak).
+Conventionally these have been reported as per weight ratios (ml/kg/min) to allow comparison between patients.
+However, numerous studies on cardiorespiratory exercise physiology have reported that normalising $\dot{V}_{O_2}$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}
 
 \subsection{Aims}
-In chapter \ref{ch_cpet_outcomes}, we reported that low anaerobic threshold in patients undergoing pancreaticoduodenectomy was associated with an increased incidence of postoperative pancreatic fistula and prolonged hospital stay.
-We also reported that patients with a VO$_2$AT less than 10mls/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and likely due to lack of physiological reserve post-surgery to be fit enough to undergo chemotherapy.
+In chapter \ref{ch_cpet_outcomes}, we reported that low $\dot{V}_{O_2}$AT in patients undergoing pancreaticoduodenectomy was associated with increased incidence of postoperative pancreatic fistula and prolonged hospital stay.
+We also reported that patients with a $\dot{V}_{O_2}$AT less than 10 ml/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and possibly due to lack of physiological reserve after surgery.
+However, we also observed that high BMI was associated with a low $\dot{V}_{O_2}$AT independent of all other clinicopathological characteristics.
+Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of $\dot{V}_{O_2}$AT.
 
-However, we noted that high BMI was associated with a low VO$_2$AT independent of all other clinicopathological characteristics.
-Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of VO$_2$AT.
 The aim of the present study was to explore the association between body composition, total body weight and the physiological parameters measured at cardiopulmonary exercise testing.
@@
 \subsection{Patients}
-Patients who underwent major abdominal surgery for malignant or benign disease involving the head of the pancreas and periampullary region at a single institution between August 2008 and October 2010 were included in this study.
+Patients scheduled to undergo pancreaticoduodenectomy for malignant or benign disease involving the head of the pancreas and periampullary region between August 2008 and October 2010 were included in this study.
 All data were recorded in a prospectively maintained database.
@@
 Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database.
-A detailed description of methodology of cardiopulmonary exercise testing and a description of the measured parameters is provided in CHAPTERX.
+Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (p\pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (p\pageref{sec:cpx_parameters}).
 
-\subsection{Body composition calculation}
-Preoperative computed tomography that had been performed as part of the routine assessment of these patients was used to calculate body composition.
-Previously published and well established methods were used were used to calculate body composition information from single CT slices.\parencite{bredella_comparison_2010,shen_total_2004}
+\subsection{Calculation of body composition}
+Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods.\parencite{bredella_comparison_2010,shen_total_2004}
 
-The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae.The cross-sectional images at these levels where then exported as bitmap images with C40 W350 settings [speak to a radiologist about what these numbers mean].
+%http://regionstraumapro.com/post/16349545265 - Excellent description of CT window width(W) and level/center(C)
+The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae.
+The CT window width (W) was set at 400 and the level/center (C) was set at 40.
+This allowed tissue between -160 HU and +240 HU to be represented in grayscale with adequate contrast between tissues of interest.
+The cross-sectional images at these levels where then exported as gray-scale bitmap images.
 The scale in millimeters was included with every image.
@@
 
-The first step involved converted the bitmap images into JPEG images using lossy compression set at 85\% to minimise sharp transitions between grey areas of very similar colour values.
-This aided easier automatic selection of contiguous areas of similar grey shades.
+The first step involved converting the bitmap images into JPEG images using lossy compression set at 85\% to minimise sharp transitions between grey areas of very similar colour values.
+This allowed more consistent selection of contiguous areas of similar grey shades.
 
@@
 
-\textbf{Body compartment selection methodology}:
+\textbf{Selecting tissue compartments}:
 
-The sequence of steps is depicted in Fig.
+The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig.
 \ref{fig:bc_ct_gimp} on p	\pageref{fig:bc_ct_gimp}.
@@
 \subsection{Cardiopulmonary exercise testing}
-All patients performed cardiopulmonary exercise testing on a cycle ergometer as described in chapterx.
+All patients performed cardiopulmonary exercise testing on a cycle ergometer as described in Section \ref{sec:cpx_method}.
 Raw data of all breath-by-breath parameters averaged every 10 seconds was collected for analysis.
-The first three minutes of the recorded data were during the rest period when the patients were on the exercise bike but did not do exercise.
-The average of each parameter measured between the first and second minute was treated as the rest value.
-Anaerobic threshold was identified using previously established methods.
-\parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.
+The first three minutes of the recorded data were during the rest period when the patients were on the exercise bike but did not exercise.
+The average of each parameter measured between the first and second minute (6 readings) was treated as the rest value.
+Anaerobic threshold was identified using previously established methods and all corresponding parameters at this point were recorded.\parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.
 
@@
 All analyses were performed using the SPSS statistical package for Microsoft Windows (version 22 ).
-Comparisons between body composition and cardiopulmonary exercise testing parameters were done using the partial correlations controlling for the effect of gender (and/or age).
+Comparison between body composition and cardiopulmonary exercise testing parameters was done using Partial correlations controlling for the effect of gender (and/or age).
 All p-values reported are two-sided.
-The relationship between body composition and various preoperative clinico-pathological characteristics (in the form of categorical variables) was analysed using the Mann-Whitney U test for variables with two categories and the Kruskal-Wallis Test for variables with more than two categories.
+The relationship between body composition and preoperative clinico-pathological characteristics (as categorical variables) was analysed using the Mann-Whitney U test for variables with two categories or the Kruskal-Wallis test for variables with more than two categories.
 Previously established cut-offs were used for categorising continuous variables where applicable.
@@
 The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender.
-Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to VO$_2$AT or VO$_2$Peak and therefore no adjustments were made for age.
+Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age.
 The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).
@@
 
-Tidal volume (litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise.
-There was a statistically significant but weak positive correlation between Minute Ventilation (Tidal Volume x Respiratory Rate) and skeletal muscle at anaerobic threshold and peak exercise but not at rest.
+Tidal volume ($\dot{V}_T$, litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise.
+There was a statistically significant but weak positive correlation between minute ventilation ($\dot{V}_E$, litres) and skeletal muscle at anaerobic threshold and peak exercise but not at rest.
 There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.
@@
 %----------------------------------------------------------------------------------------------
-% Correlation between body composition and VO$_2$AT before and after correction for total body weight.
+% Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
 \begin{figure}[htb]
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal}
-		\caption{VO$_2$AT vs. Skeletal Muscle}
+		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
 		\label{fig:bc_scatter_VO2_skeletal}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
-		\caption{VO$_2$AT vs. Total Adipose Tissue}
+		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
 		\label{fig:bc_scatter_VO2_TAT}
 	\end{subfigure}
-	\caption{Correlation between body composition and VO$_2$AT before and after correction for total body weight.}	
+	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.}	
 	\label{fig:bc_scatter_VO2_bodycomp_reversal}
@@
 
-Absolute oxygen consumption (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$).
-However, this correlation was lost after correction of oxygen consumption for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.
+Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$).
+However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.
 
-Absolute oxygen consumption (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold.
-However, when it was corrected for total body weight, there was a strong correlation between corrected oxygen consumption (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
+Absolute $\dot{V}_{O_2}$ (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold.
+However, when it was corrected for total body weight, there was a strong correlation between corrected $\dot{V}_{O_2}$ (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
 
-The loss of the physiological relationship between VO$_2$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig.
+The loss of the physiological relationship between $\dot{V}_{O_2}$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig.
 \ref{fig:bc_scatter_VO2_TAT}.
 
-%VO$_2$ corrected to height seems to correlate best with BC and everything else
+%$\dot{V}_{O_2}$ corrected to height seems to correlate best with BC and everything else
 \clearpage
@@
 
-The positive correlation between absolute oxygen consumption and skeletal muscle area is easily explained by the physiology of aerobic exercise.
+The positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area is easily explained by the physiology of aerobic exercise.
 During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.
 
-Current convention is to report oxygen consumption measured at cardiopulmonary exercise testing according the following formula: 
+Current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according the following formula: 
 \[Corrected\ VO_2 (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ VO_2\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
 
-In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between oxygen consumption at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.
+In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between $\dot{V}_{O_2}$ at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.
 
@@
 Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{"Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation"}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass.
-In a detailed anaysis comparing oxygen consumption and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt.
+In a detailed anaysis comparing $\dot{V}_{O_2}$ and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt.
 must be expected to decrease purely as a result of the method used for representing the data." }
@@
 
-Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of VO$_2$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
+Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
 Body composition changes were also tracked over this period.
 They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls.
-They reported on the significant difficulties in normalising VO$_2$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising VO$_2$.
-They found that VO$_2$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls.
+They reported on the significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$.
+They found that $\dot{V}_{O_2}$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls.
 However, this underestimation was greater in girls than in boys.
 
-Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on VO$_2$max in obese women before and after weight loss.
-VO$_2$max corrected for total body weight was significantly lower in the obese state while VO$_2$max corrected for fat-free mass did not change significantly after weight loss.
+Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss.
+$\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss.
 They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities.
@@
 
-In a study aimed at determining the optimal method of expressing VO$_2$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with VO$_2$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
-However, when VO$_2$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups.
-They, similar to Goran et al\parencite{goran_total_2000}, recommend that VO$_2$ be normalised to lean body mass rather than total body weight.
+In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
+However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups.
+They, similar to Goran et al\parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.
 
@@
 \subsection{Clinical implications of spurious correlation}
-Older et al in their pioneering study in 1993 reported that VO$_2$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
-However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a VO$_2$AT < 11mls/kg/min represented cardiac failure.
-This association is repeated in their later work on 548 patients which also showed a clear association between VO$_2$AT < 11mls/kg/min and mortality due to cardiovascular causes.
+Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
+However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure.
+This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes.
 \parencite{older_cardiopulmonary_1999} The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}
 
-Swart and Carlisle reported that VO$_2$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low VO$_2$AT group was significantly greater than that in the normal VO$_2$AT groups (24\% vs 51\%).
-The average VO$_2$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
-In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low VO$_2$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.
+Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%).
+The average $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
+In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.
 
@@
 The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs.
-Instead, improvement in the absolute values of VO$_2$AT and VO$_2$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.
+Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.
+
+%This needs to be expanded significantly.
+
+%Mention studies that have used VE/VCO2. 
+%I think there is a study in bariatrics where bmi and at were independent predictors. 
+%Mention studies that used predicted VO2 or absolute $\dot{V}_{O_2}$ where calculated AT was not useful. 
+%This should give me at least another 2 paragraphs

### 2015-06-06 Chapters/Introduction.tex
Chapter Body Composition - numerous small changes - fixed all VO2 to correct format - fixed some problems with style - fixed a few spelling/grammar issues - needs a final check - see comments at end of discussion - consider separate tables for each phase of exercise

@@
 \subsection{Cardiopulmonary Exercise Test Methodology}
+\label{sec:cpx_method}
 CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.
@@
 \section{Description of CPET Parameters}
+\label{sec:cpx_parameters}
 \input{Tables/c1_cpet_parameters}

### 2015-06-06 main.tex
Modified main.tex to include several package options for bibiblatex, including maxcitenames, mincitenames, uniquelist and uniquenames options - may need modification.
Added package /todo/ to main.tex
Changed todo to easy-todo package. Works

@@
 \usepackage[backend=biber,
-			style=authoryear,
+			citestyle=authoryear,
+			bibstyle=authoryear,
+			uniquelist=false,
+			uniquename=false,
+			maxcitenames=1,
+			mincitenames=1,
 			url=false]{biblatex}
@@
 %http://tex.stackexchange.com/questions/16765/biblatex-author-year-square-brackets
+%http://ctan.mirrorcatalogs.com/macros/latex/contrib/biblatex/doc/biblatex.pdf - for info on all the above options. File is also stored in the Glasgow Research folder
 \makeatletter
@@
 \usepackage{pdfpages} %for including pdf files or individual pages of pdf files
+\usepackage{easy-todo} %Adds todos to the document. Use [diable] as package option in final verion. Also see \listoftodos below.
 \usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
@@
 
+\listoftodos
+
 \end{document}

### 2015-06-06 Chapters/Introduction.tex
Changed todo to easy-todo package. Works
Added Todos in introduction
Added a whole pile of Todos
- Time to take a break
- There is a pile of stuff to do in introduction still
- AAaaaaaaaaaaaaaaaaaarggggggggggggggggghhhhhhhhhhhhhh

@@
 \label{sec:comorbidity_risk_stratification}
+\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
 \subsection{Comorbidity}
@@
 \subsection{Risk Stratification}
+\todo{Intro - risk stratification}
+%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
 Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery.
@@
 \subsection{Static Versus Dynamic Testing}
-
+\todo{Static Versus Dynamic Testing}
 Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase.
@@
 \section{Cardiopulmonary Exercise Testing}
-%History of CPET
-\subsection{History of CPET in Surgery}
+\todo{A brief history of CPET in general}
+\todo{The physiological basis of CPET}
 
@@
 \subsubsection{Ventilatory equivalents method}
-[...]
+\todo{Ventilatory equivalents method}
 
@@
 \subsection{Breathing frequency, $B_f$}
+In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake.
+However, a variety of respiratory disorders including obstructive disorders such as chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response.
+Morbidly obese patients may also have additional restrictions on respiratory response due to their body habitus.
+On the other hand, hyperventilation in an otherwise healthy patient may make interpretation of the results difficult although the analysis of trends in other parameters such as $P_{ET_{O_2}}$ and $P{ET_{CO_2}}$ aids in the diagnosis of hyperventilation versus true respiratory problems.
 
 \section{Role of CPET in preoperative assessment}
+\label{sec:cpx_roleInAssessment}
 \subsection{General Surgery}
+
 Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery.
-
 Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery.
@@
 Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993}
+
 In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise.
@@
 \subsection{Relationship between systemic inflammation and comorbidity}
+\todo{Relationship between systemic inflammation and comorbidity}
 
@@
 %nehez_compromise_2002 - get other citations from this one
+\todo{Impact of jaundice on the immune system}
 
@@
 \section{Body Composition}
-Make this a brief description of body composition.
+\todo{Introduction - body composition}

### 2015-06-13 Chapters/CPET_Survival.tex
Some minor changes before starting work today

@@
 
-\chapter{An investigation into the relationship between cardiopulmonary exercise testing, comorbidity, systemic inflammation and survival after pancreaticoduodenectomy for cancer.}
+\chapter{An investigation into the relationship between cardiopulmonary exercise testing, comorbidity, systemic inflammation and survival after pancreaticoduodenectomy for pancreatic ductal adenocarcinoma.}
 \label{ch_survival}
@@
 
-Comorbidity is not only associated with postoperative morbidity and mortality (Section \ref{sec:comorbidity_risk_stratification}) but has also been reported to be associated with poor long-term survival in patients undergoing surgery several different cancers including colorectal cancer \parencite{} and breast cancer.\parencite{} With 10-year survival rates approaching 60\% and 80\% in these patients, it is not surprising that some patients with significant comorbidity may die from their comorbid disease rather than from cancer recurrence.
-However, cancer-specific survival is also shorter in patients with significant comorbidities.
+Comorbidity is not only associated with postoperative morbidity and mortality (Section \ref{sec:comorbidity_risk_stratification}) but has also been reported to be associated with poor long-term survival in patients undergoing surgery for several different cancers including colorectal cancer \parencite{} and breast cancer.\parencite{} With 10-year survival rates approaching 60\% and 80\% respectively in patients with these cancers, it is not surprising that some patients with significant comorbidity die of their comorbid disease rather than from cancer recurrence.
+However, reports suggest that cancer-specific survival is also shorter in patients with significant comorbidities.
 

### 2015-06-15 Chapters/BodyComp.tex
Some minor changes before starting work today
Additions to results section of CPET Survival
- Add description of results of Cox regression to CPET Survival
- Shorten cpet_survival_cox_regression table by condensing rows into one per parameter
- Save long form of above table in separate file - not currently in use
- New table - cpet_survival_6months - needs fleshing out
- Add todo to BodyComp chapter - ?relationship with systemic inflammation
- Changes to main are simply variations in included files

@@
 % Chapter 04
-
 \chapter{An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing major pancreatic surgery.}
@@
 %----------------------------------------------------------------------------------------
-$\dot{V}_{O_2}$
+\todo{Consider including relationship with systemic inflammation}
+
+zotero - richards-relationships-2012
+
 \section{Introduction}
@@
 Current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according the following formula: 
-\[Corrected\ VO_2 (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ VO_2\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
+\[Corrected\ \dot{V}_{O_2} (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
 

### 2015-06-15 Chapters/CPET_Survival.tex
CPET Survival Cox Univariate regression
- Complete CPET vs Survival Univariate table
- Create new SPSS dataset for this chapter -ds_at_pdac
- Remove redundant variables in this dataset
- Changes to main.tex are simply commenting out other chapters from output
Add cpet_survival_cox_regression table to Tables
- Include table in chapter CPET Survival
Additions to results section of CPET Survival
- Add description of results of Cox regression to CPET Survival
- Shorten cpet_survival_cox_regression table by condensing rows into one per parameter
- Save long form of above table in separate file - no

@@
-% Chapter 05 - CPET, complications and survival
+% Chapter 05 - CPET, comorbidity and survival
 
@@
 
+\input{Tables/cpet_survival_cox_regression}
+
+Fifty patients (31 male) with pancreatic ductal adenocarcinoma underwent pancreaticoduodenectomy (n=48) or sub-total pancreatectomy (n=2) and had cardiopulmonary exercise test as part of their preoperative assessment.
+The median age of the patients was xx (range ).
+Three patients died within 30 days of the operation as a result of postoperative complications and were excluded from the survival analysis.
+
+%UNIVARIATE Cox-regression analysis
+The clinico-pathological characteristics of the patients studied (n=47) are shown in \ref{table:cpet_survival_cox} on p\pageref{table:cpet_survival_cox}.
+POSSUM Physiology score $>$14 (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014) and preoperative serum C-reactive protein $>$ 10 mg/dl (HR 2.14, 95\% CI 1.04 - 4.41, p = 0.04) were associated with survival on univariate Cox-regression analysis.
+Cardiopulmonary exercise test parameters including $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ were not associated with survival on univariate analysis.
+%MULTIVARIATE Cox-regression analysis
+On multi-variate backward conditional regression analysis using POSSUM Physiology Score and serum C-reactive protein as independent variables, only POSSUM Physiology Score $>$14  was found to be an independent predictor of survival (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014).
+
+%Create a table looking at means of cardiopulmonary exercise test parameters in survivors vs. non-survivors at 6 months, 1 year and 2 years. This should give three tables. as below.
+
+\input{Tables/cpet_survival_6months}
+
 \section{Discussion}

### 2015-06-23 Chapters/CPET_Survival.tex
Add CPET vs 6, 12, 24 month survivor tables
- Compare CPET in survivors vs non-survivors at 6, 12, 24 months
- Include these tables in chapter
- TODO - describe these tables in the results section
- TODO - Kaplan-Meir plots

@@
 Fifty patients (31 male) with pancreatic ductal adenocarcinoma underwent pancreaticoduodenectomy (n=48) or sub-total pancreatectomy (n=2) and had cardiopulmonary exercise test as part of their preoperative assessment.
-The median age of the patients was xx (range ).
+The median age of the patients was 66 years (IQR 57 - 70 years ).
 Three patients died within 30 days of the operation as a result of postoperative complications and were excluded from the survival analysis.
@@
 
-%Create a table looking at means of cardiopulmonary exercise test parameters in survivors vs. non-survivors at 6 months, 1 year and 2 years. This should give three tables. as below.
+%Create a table looking at means of cardiopulmonary exercise test parameters in survivors vs. non-survivors at 6 months, 1 year and 2 years. This should give three tables as below.
+
+Cardiopulmonary exercise testing parameters in survivors and non-survivors at 6 months, 12 months and 24 months are shown in tables \ref{table:cpet_survival_6months}, \ref{table:cpet_survival_12months}, \ref{table:cpet_survival_24months}.
 
 \input{Tables/cpet_survival_6months}
+\input{Tables/cpet_survival_12months}
+\input{Tables/cpet_survival_24months}
 

### 2015-06-23 Chapters/CPET_Survival.tex
Kaplan-Meier, Cox Regression - tumour factors
- Complete and include KM plot
- Minor edits of all other tables
- Include new table on Cox regression incl tumour factors
- TODO - ?Split tables into uni and multi
- TODO - Fix tables spilling over margins

@@
 \input{Tables/cpet_survival_cox_regression}
+\input{Tables/cpet_survival_cox_tumour_factors}
 
@@
 %MULTIVARIATE Cox-regression analysis
-On multi-variate backward conditional regression analysis using POSSUM Physiology Score and serum C-reactive protein as independent variables, only POSSUM Physiology Score $>$14  was found to be an independent predictor of survival (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014).
+On multi-variate backward conditional regression analysis using POSSUM Physiology Score and serum C-reactive protein as independent variables, only POSSUM Physiology Score $>14$  was found to be an independent predictor of survival (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014).
+
+Kaplan-Meier survival analysis comparing survival in patients with a low ($leq14$) versus high ($>14$) POSSUM Phsyiology Score is shown in Fig.
+\ref{fig:cpet_survival_km_possum}.
+Estimated median survival was  32 months (95\% CI 17.02 - 46.99) in patients with a low POSSUM Physiology Score but reduced to 14 months (95\% CI 9.51 - 18.49) in a high POSSUM Physiology Score (p = 0.012).
+
+%Figure 1
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width=0.8\linewidth]{Figures/cpet_survival_km_possum}
+	\caption{Kaplan-Meier Survival Analysis of POSSUM Physiology Score ($\leq14$ vs. $>14$)}
+	\label{fig:cpet_survival_km_possum}
+\end{figure}
 
@@
 Cardiopulmonary exercise testing parameters in survivors and non-survivors at 6 months, 12 months and 24 months are shown in tables \ref{table:cpet_survival_6months}, \ref{table:cpet_survival_12months}, \ref{table:cpet_survival_24months}.
+Load achieved at peak exercise was significantly higher in patients who survived at least 6 months (118.5 vs.
+76.5 watts, p = 0.026) However, there was no association between any of the other cardiopulmonary exercise testing parameters and survival at 6 months, 12 months or 24 months.
 
@@
 
+%Compare significant preoperative factors with tumour factors
+
 \section{Discussion}

### 2015-06-26 Chapters/CPET_Survival.tex
CPET Survival - Tidy up tables
- Tidy up all tables in this chapter
- Change table titles
- Format numbers in tables
- Fix some references to sections in introduction

@@
 
-\lhead{Chapter \ref{ch_survival}. \emph{CPET, Complications and Survival}} % This is for the header on each page - perhaps a shortened title
+\lhead{Chapter \ref{ch_survival}. \emph{CPET, Comorbidity and Survival}} % This is for the header on each page - perhaps a shortened title
 
@@
 \section{Introduction}
-Median survival after pancreaticoduodenectomy for pancreatic ductal adenocarcinoma varies from approximately 18 months to 24 months.\parencite{winter_1423_2006,neoptolemos_adjuvant_2010}
+Median survival after pancreaticoduodenectomy for pancreatic ductal adenocarcinoma varies from approximately 18 months to 24 months.\parencite{winter_1423_2006,neoptolemos_adjuvant_2010}  Selecting patients who will benefit from the survival advantage that a pancreaticoduodenectomy offers is important in improving long-term outcomes after this morbid procedure.
 
-Selecting patients who will benefit from the survival advantage that a pancreaticoduodnectomy offers is important to maximise the usefulness of this morbid procedure.
-
-Comorbidity is not only associated with postoperative morbidity and mortality (Section \ref{sec:comorbidity_risk_stratification}) but has also been reported to be associated with poor long-term survival in patients undergoing surgery for several different cancers including colorectal cancer \parencite{} and breast cancer.\parencite{} With 10-year survival rates approaching 60\% and 80\% respectively in patients with these cancers, it is not surprising that some patients with significant comorbidity die of their comorbid disease rather than from cancer recurrence.
+Comorbidity is not only associated with postoperative morbidity and mortality (Section \ref{sec:comorbidity_risk_stratification} but has also been reported to be associated with poor long-term survival in patients undergoing surgery for several different cancers including colorectal cancer \parencite{} and breast cancer.\parencite{} With 10-year survival rates approaching 60\% and 80\% respectively in patients with these cancers, it is not surprising that some patients with significant comorbidity die of their comorbid disease rather than from cancer recurrence.
 However, reports suggest that cancer-specific survival is also shorter in patients with significant comorbidities.
@@
 The modified Glasgow Prognostic Score, a measure of preoperative systemic inflammation in cancer patients, is associated with poor survival regardless of the site or stage of cancer.
-This is discussed in detail in Section \ref{sec:intro_systemic_inflammation_outcome}.
+This is discussed in detail in Section \ref{sec:intro_systemic_inflammation_outcome}
+.
 
@@
 The aim of the present study was to investigate the relationship between cardiopulmonary exercise testing, comorbidity, systemic inflammation and survival in patients undergoing pancreaticoduodenectomy for pancreatic ductal adenocarcinoma.
+\clearpage
 
@@
 
+\clearpage
+
 \section{Results}
@@
 Fifty patients (31 male) with pancreatic ductal adenocarcinoma underwent pancreaticoduodenectomy (n=48) or sub-total pancreatectomy (n=2) and had cardiopulmonary exercise test as part of their preoperative assessment.
-The median age of the patients was 66 years (IQR 57 - 70 years ).
 Three patients died within 30 days of the operation as a result of postoperative complications and were excluded from the survival analysis.
@@
 %UNIVARIATE Cox-regression analysis
-The clinico-pathological characteristics of the patients studied (n=47) are shown in \ref{table:cpet_survival_cox} on p\pageref{table:cpet_survival_cox}.
+The clinico-pathological characteristics of the patients studied (n=47) are shown in Table \ref{table:cpet_survival_cox} on p\pageref{table:cpet_survival_cox}.
+The median age of the patients was 66 years (IQR 57 - 70 years).
+
 POSSUM Physiology score $>$14 (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014) and preoperative serum C-reactive protein $>$ 10 mg/dl (HR 2.14, 95\% CI 1.04 - 4.41, p = 0.04) were associated with survival on univariate Cox-regression analysis.
@@
 %MULTIVARIATE Cox-regression analysis
-On multi-variate backward conditional regression analysis using POSSUM Physiology Score and serum C-reactive protein as independent variables, only POSSUM Physiology Score $>14$  was found to be an independent predictor of survival (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014).
+On multi-variate backward conditional Cox-regression analysis using POSSUM Physiology Score and serum C-reactive protein as independent variables, only POSSUM Physiology Score $>14$  was found to be an independent predictor of survival (HR 2.38, 95\% CI 1.19 - 4.75, p = 0.014).
+However, in a multi-variate model including POSSUM Physiology Score and tumour related factors, lymph node ratio $\geq$0.15 (HR 2.3, 95\% CI 1.14 - 4.66, p = 0.021) and positive resection margin (HR 3.33, 95\% CI 1.36 - 8.16, p = 0.008) were significantly associated with poor survival while POSSUM Physiology Score was not (p = 0.391).
 
@@
 
-%Create a table looking at means of cardiopulmonary exercise test parameters in survivors vs. non-survivors at 6 months, 1 year and 2 years. This should give three tables as below.
-
 Cardiopulmonary exercise testing parameters in survivors and non-survivors at 6 months, 12 months and 24 months are shown in tables \ref{table:cpet_survival_6months}, \ref{table:cpet_survival_12months}, \ref{table:cpet_survival_24months}.
@@
 
+\clearpage
+
 %Compare significant preoperative factors with tumour factors
@@
 
- This is considerably less than other common cancers such as colorectal cancer or breast cancer where 5-year survivals across all stages are 60\% and 90\% respectively.
-However, only 10-20\% of pancreatic cancers are suitable for potentially curative surgery and of patients who undergo curative surgery 5-year survival remains low at approximately 20\%.\parencite{cancerresearchuk_cancer_2014}
+Long-term survival after potentialy curative surgery for pancreatic ductal adenocarcinoma remains poor with 5-year survival rates after surgery of only 20\%.
+This is considerably lower than surgery after other cancers such as breast or colorectal cancer.\parencite{cancerresearchuk_cancer_2014}
+
+The results of the present

### 2015-07-04 Chapters/CPET_Survival.tex
CPET_Survival Todos and comments
- The next few commits will be for a new chapter
- Sat out in my camping chair listening to aerosmith

@@
 
+\todo{Should complications-survival also be included in this chapter?}
+
 \section{Introduction}
@@
 The modified Glasgow Prognostic Score, a measure of preoperative systemic inflammation in cancer patients, is associated with poor survival regardless of the site or stage of cancer.
-This is discussed in detail in Section \ref{sec:intro_systemic_inflammation_outcome}
-.
+This is discussed in detail in Section \ref{sec:intro_systemic_inflammation_outcome}.
 
@@
 The clinico-pathological characteristics of the patients studied (n=47) are shown in Table \ref{table:cpet_survival_cox} on p\pageref{table:cpet_survival_cox}.
-The median age of the patients was 66 years (IQR 57 - 70 years).
+The median age of the patients was 66 years (Inter-Quartile Range, IQR 57 - 70 years).
 
@@
 \ref{fig:cpet_survival_km_possum}.
-Estimated median survival was  32 months (95\% CI 17.02 - 46.99) in patients with a low POSSUM Physiology Score but reduced to 14 months (95\% CI 9.51 - 18.49) in a high POSSUM Physiology Score (p = 0.012).
+Estimated median survival was  32 months (95\% CI 17.02 - 46.99) in patients with a low POSSUM Physiology Score but reduced to 14 months (95\% CI 9.51 - 18.49) in patients with a high POSSUM Physiology Score (p = 0.012).
 
@@
 
-The results of the present
+The results of the present study suggest that comorbidity as measured by the POSSUM Physiology Score may be associated with shorter long-term survival in patients with pancreatic ductal adenocarcinoma.
+
+%Lack of assocation with CPET
+
+%Tumour factors trump comorbidity
+
+Overall median survival in this study was [\todo{pending result}?
+months] which is considerably shorter than that reported in other studies on the impact of comorbidty on long-term survival.
+The short median survival is a consequence of the aggressive nature of pancreatic ductal adenocarcinoma and is similar to previously published literature.
+[\todo{pending reference}] This is also demonstrated in Table \ref{table:cpet_survival_cox_tumour_factors}.
+This
+
+%Discuss limitations
+% - Small numbers
+% - Selection bias - patients with significant comorbidity would have not been offered surgery
+% - No information on adjuvant therapy
+
+There are several limitations to this study.
+The number of patients included in this study is small (47) which increases the risk of a Type 2 error.
+Patients with resectable disease who underwent surgery were by definition a fitter cohort with fewer comorbdities.
+This introduces a selection bias and may have affected the results.
+The impact of adjuvant therapy

### 2015-07-04 Chapters/CRP_Complications.tex
New Chapter - CRP_Complications
- Complete Introduction
- Create sections

% Chapter 06 - Postop CRP and complications

\chapter{An investigation into the relationship between postoperative systemic inflammation and complications after pancreaticoduodenectomy.}
\label{ch_survival}

\lhead{Chapter \ref{ch_survival}. \emph{Postoperative CRP and complications}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------
\section{Introduction}
Pancreaticoduodenectomy is associated with significant morbidity in spite of advances in patient selection, perioperative care and surgical technique. Early identification of complications can help improve outcomes by better allocation of critical care resources as well as goal-directed therapy. Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can lead to a cascade of other complications including delayed haemorrhage, infected intra-abdominal collections, delayed gastric emptying, prolonged hospitalisation and in some cases, death. The International Study Group for Pancreatic Fistula (ISGPF) have not only defined what constitutes a post-operative pancreatic fistula but have also graded the severity of this complication based on its impact on the management of the patient. However, these definitions are applied after the event and there is no clear way of predicting the severity of complications.

Postoperative CRP levels have been shown to predict infectious complications after colorectal surgery as well as oesophago-gastric surgery. It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing. The role of postoperative CRP in predicting the severity of complications after major pancreatic surgery has not been reported before. Stratifying patients based on the predicted severity of complications will allow identifying low risk patients who will be suitable to continue on enhanced recovery pathways and high risk patients who may require prolonged critical care, prolonged hospitalisation or further interventions.

The aim of this study was to investigate the association between postoperative systemic inflammation and severity of complications after pancreaticoduodenectomy.

\clearpage
\section{Methods}
Patients who underwent pancreaticoduodenectomy between August 2008 and July 2012 were included in the study. Data were collected from a prospectively maintained database. All complications were discussed at a weekly meeting and prospectively recorded. Postoperative pancreatic fistula and post-pancreatectomy haemorrhage were graded according to the International Study Group classifications. All other complications were graded using the Clavien-Dindo system. Postoperative mortality was defined as death within 30-days of the operation or while still in hospital after the operation. Re-intervention in the form of radiological, endoscopic or surgical procedures was recorded prospectively. Preoperative blood results as well as postoperative blood results for the first 7 days after surgery were collected from the hospital laboratory databases.

\subsection{Statistics}
SPSS version 22 was used for all analysis. ROC analysis was used to identify the optimum thresholds of CRP for predicting complications.
\todo{Complete this section - most of this can be modified from the stats sections in other chapters}

\clearpage
\section{Results}

\clearpage
\section{Discussion}

### 2015-07-07 Tables/crp_comp_vs_POPF_ISGPS_p_values_only.tex
New table POPF_ISGPS_p_values only
- Table comparing crp in leak vs no leak as well as between ISGPS grades
- This table has only the p-values

%06/07/15 - Started this table
\begin{table}[p]
	\centering
	\caption{The relationship between CRP during the first postoperative week and post-operative pancreatic fistula (POPF) in patients undergoing pancreaticoduodenectomy (Mann-Whitney U test) }
	\label{table:crp_comp_vs_POPF_ISGPS_p_values_only}
	\renewcommand{\arraystretch}{1.2} %Increases space between rows
	%\setlength{\tabcolsep}{9pt} %sets the space between columns
	\begin{tabular}{l | c c c c}
		           & multicolumn{4}{c}{POPF} &  \\
		Postop Day & Yes/No                  & No vs. A & A vs. B & B vs. C \\
		0          & 0.722                   & 0.582    & 0.849   & 0.730   \\
		1          & 0.242                   & 0.331    & 0.740   & 0.295   \\
		2          & 0.002                   & 0.161    & 0.222   & 0.563   \\
		3          & $<$0.001                & 0.022    & 0.602   & 0.530   \\
		4          & $<$0.001                & 0.007    & 0.906   & 0.834   \\
		5          & $<$0.001                & 0.035    & 0.611   & 0.959   \\
		6          & $<$0.001                & 0.012    & 0.638   & 0.781   \\
		7          & $<$0.001                & 0.006    & 0.235   & 0.356
	\end{tabular}
\end{table}

### 2015-07-07 Chapters/CRP_Complications.tex
CRP_Complications Chapter - CRP Trends Graphs and Tables
- Create 4 new graphs for crp vs popf vs infectious complications
- Create one new table for crp vs infectious complications in patients with/without popf
- Modify crp_complications chapter text to include above

@@
 \section{Methods}
-Patients who underwent pancreaticoduodenectomy between August 2008 and July 2012 were included in the study.
+\todo{Check dates for this study}
+Patients who underwent pancreaticoduodenectomy between January 2008 and July 2012 were included in the study.
 Data were collected from a prospectively maintained database.
@@
 \section{Results}
+One hundred and eighty patient (131 male) underwent pancreaticoduodenectomy during the study period.
+
+%CRP trends and pancreatic fistula
+Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula.
+However, CRP levels in the first postoperative week were not associated with the clinical severity of the pancreatic fistula.
+These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig.
+\ref{fig:crp_comp_crp_popf}.
+Fig.
+\ref{fig:crp_comp_crp_popf_isgps} shows that CRP levels were not significantly different between the different grades of pancreatic fistula while CRP levels in patients who did not develop a pancreatic fistula had significantly lower CRP levels.
+
+%SPSS results and graphs are stored in crp_complications_graphs_etc.spv file
+
+%==============================================================================
+%CRP trends in a 2-panel figure and table showing p-values below it
+\clearpage
+\begin{figure}[htbp]
+	\centering
+	\begin{subfigure}{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_crp_popf_yes_no}
+		\caption{POPF Absent vs. Any Grade}
+		\label{fig:crp_comp_crp_popf_yes_no}
+	\end{subfigure}
+	\begin{subfigure}{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_crp_popf_isgps}
+		\caption{POPF ISGPS Grades}
+		\label{fig:crp_comp_crp_popf_isgps}
+	\end{subfigure}
+
+	\caption{Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula (POPF).}
+	\label{fig:crp_comp_crp_popf}
+
+\end{figure}
+\input{Tables/crp_comp_vs_POPF_ISGPS_p_values_only}
+%==============================================================================
+
+%Ability of CRP to predict infectious complications in leak vs. no leak complications
+%Rationale - POPF is diagnosed with a separate definition - ISGPS on Day 3.
+Therefore, there is no role for CRP in diagnosing POPF.
+Above results show that it does not discriminate between the grades either.
+Therefore, the following analysis is to find out if the first week CRP has any role in predicting other outcome measures.
+
+%==============================================================================
+%CRP trends in a 2-panel figure and table showing p-values below it
+\clearpage
+\begin{figure}[htbp]
+	\centering
+	\begin{subfigure}{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_infectious_leak0}
+		\caption{Patients with \textbf{\underline{no}} POPF}
+		\label{fig:crp_comp_infectious_leak0}
+	\end{subfigure}
+	\begin{subfigure}{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_infectious_leak1}
+		\caption{Patients with POPF}
+		\label{fig:crp_comp_infectious_leak1}
+	\end{subfigure}
+
+	\caption{Relationship between postoperative CRP and clinically significant infectious complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_infectious_leak}
+
+\end{figure}
+\input{Tables/crp_comp_vs_infections_popf_y1n0}
+%==============================================================================
 

### 2015-07-08 Chapters/CRP_Complications.tex
CRP Complications - ROC Analysis - Infective complications
- Create d3-7 ROC figures
- Create 2 new tables - crp vs. infections, ROC
- Modify chapter to include above

@@
 
-%==============================================================================
+%========================CRP vs POPF============================================
 %CRP trends in a 2-panel figure and table showing p-values below it
@@
 
-%==============================================================================
-%CRP trends in a 2-panel figure and table showing p-values below it
+%========================CRP vs Infections====================================
+%CRP trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
 \clearpage
-\begin{figure}[htbp]
+\begin{figure}[t]
 	\centering
@@
 
+%========================ROC Curves==========================================
+\clearpage
+\begin{figure}[t]
+	\centering
+	\begin{subfigure}{0.3\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_ROC_infection_D3}
+		\caption{Day 3}
+		\label{fig:crp_comp_ROC_infection_D3}
+	\end{subfigure}
+	\begin{subfigure}{0.3\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_ROC_infection_D4}
+		\caption{Day 4}
+		\label{fig:crp_comp_ROC_infection_D4}
+	\end{subfigure}
+	\begin{subfigure}{0.3\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_ROC_infection_D5}
+		\caption{Day 5}
+		\label{fig:crp_comp_ROC_infection_D5}
+	\end{subfigure}
+	\begin{subfigure}{0.3\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_ROC_infection_D6}
+		\caption{Day 6}		
+		\label{fig:crp_comp_ROC_infection_D6}
+	\end{subfigure}
+	\begin{subfigure}{0.3\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_ROC_infection_D7}
+		\caption{Day 7}
+		\label{fig:crp_comp_ROC_infection_D7}
+	\end{subfigure}
+
+	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications.}
+	\label{fig:crp_comp_ROC_infection}
+
+\end{figure}
+\input{Tables/crp_comp_ROC_infections_noPOPF}
+
 \clearpage

### 2015-07-08 Tables/crp_comp_ROC_infections_noPOPF.tex
CRP Complications - ROC Analysis - Infective complications
- Create d3-7 ROC figures
- Create 2 new tables - crp vs. infections, ROC
- Modify chapter to include above

%07/07/15 - Started this table
\begin{table}[h]
	\centering
	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications in patients who did not develop a POPF}
	\label{table:crp_comp_ROC_infections_noPOPF}
	\renewcommand{\arraystretch}{1.2} %Increases space between rows
	%\setlength{\tabcolsep}{9pt} %sets the space between columns
	\begin{tabular}{| c | c c c | c c c c c |}
		\hline
		Day & AUC  & p        & 95\% CI    & CRP   & Spec. & Sens. & NPV  & PPV              \\ \hline
		3   & 0.64 & 0.011    & 0.54 - .74 & 177.5 & 0.66  & 0.64  & 0.79 & 0.48             \\
		4   & 0.71 & $<$0.001 & 0.61 - .80 & 125   & 0.64  & 0.74  & 0.83 & 0.5              \\
		5   & 0.70 & $<$0.001 & 0.61 - .80 & 106.5 & 0.64  & 0.63  & 0.78 & 0.45             \\
		6   & 0.73 & $<$0.001 & 0.65 - .82 & 91.5  & 0.68  & 0.73  & 0.84 & 0.53             \\
		7   & 0.80 & $<$0.001 & 0.72 - .88 & 87    & 0.65  & 0.81  & 0.87 & 0.52             \\ \hline
		\multicolumn{9}{l}{AUC - Area under curve; Spec. - Specificity, Sens. - Sensitivity} \\
		\multicolumn{9}{l}{NPV - Negative Predictive Value; PPV - Positive Predictive Value}
	\end{tabular}
\end{table}

### 2015-07-10 Tables/crp_comp_vs_POPF_ISGPS_p_values_only.tex
CRP_Complications Chapter - CRP Trends Graphs and Tables
- Create 4 new graphs for crp vs popf vs infectious complications
- Create one new table for crp vs infectious complications in patients with/without popf
- Modify crp_complications chapter text to include above
CRP_Complications - Minor Changes
- Multiple minor cosmetic changes

@@
 %06/07/15 - Started this table
-\begin{table}[p]
+\begin{table}[h]
 	\centering
-	\caption{The relationship between CRP during the first postoperative week and post-operative pancreatic fistula (POPF) in patients undergoing pancreaticoduodenectomy (Mann-Whitney U test) }
+	\caption{The relationship between CRP during the first postoperative week and post-operative pancreatic fistula (POPF) in patients undergoing pancreaticoduodenectomy.}
 	\label{table:crp_comp_vs_POPF_ISGPS_p_values_only}
@@
 	%\setlength{\tabcolsep}{9pt} %sets the space between columns
-	\begin{tabular}{l | c c c c}
-		           & multicolumn{4}{c}{POPF} &  \\
-		Postop Day & Yes/No                  & No vs. A & A vs. B & B vs. C \\
+	\begin{tabular}{| c | c c c c |}
+		\hline
+		           &       \multicolumn{4}{c}{POPF ISGPS Grades - \textit{p}}        \\
+		Postop Day & No vs. Any Grade & No vs. A & A vs. B & B vs. C                 \\ \hline
 		0          & 0.722            & 0.582    & 0.849   & 0.730                   \\
@@
 		6          & $<$0.001         & 0.012    & 0.638   & 0.781                   \\
-		7          & $<$0.001                & 0.006    & 0.235   & 0.356
+		7          & $<$0.001         & 0.006    & 0.235   & 0.356                   \\ \hline
+		\multicolumn{5}{l}{\textit{p} - Mann-Whitney U test}                         \\
+		\multicolumn{5}{l}{ISGPS - International Study Group for Pancreatic Surgery}	\\
+				\multicolumn{5}{l}{POPF - Postoperative Pancreatic Fistula}
 	\end{tabular}
+
 \end{table}

### 2015-07-12 Chapters/CRP_Complications.tex
CRP_Complications - Minor Changes
- Multiple minor cosmetic changes
CRP_Comp - More Tables
- Include table for preop factors vs. infectious complications
- Include table for infectious complications vs. other outcomes
- Describe these within the text
- Flesh out methods, stats and results sections properly

@@
 
-\clearpage
 \section{Methods}
 \todo{Check dates for this study}
-Patients who underwent pancreaticoduodenectomy between January 2008 and July 2012 were included in the study.
-Data were collected from a prospectively maintained database.
-All complications were discussed at a weekly meeting and prospectively recorded.
-Postoperative pancreatic fistula and post-pancreatectomy haemorrhage were graded according to the International Study Group classifications.
-All other complications were graded using the Clavien-Dindo system.
+Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 were included in the study.
+Patients who underwent only a trial dissection or trial dissection and surgical bypass during this period were excluded.
+All patients were given antibiotic prophylaxis at induction and this was continued for 24 hours after surgery.
+All patients had general anaesthesia supplemented either with patient-controlled epidural analgesia or spinal diamorphine and patient-controlled opiate analgesia for postoperative pain relief.
+Patients were routinely admitted to the surgical high dependency unit where a standardised regimen was followed for early mobilisation, chest physiotherapy and early enteral nutrition.
+All patients had one or two surgical drain(s) placed at the time of surgery which were removed at the clinician's discretion during the postoperative period based on the presence or absence of postoperative pancreatic fistula or other intra-abdominal collections.
+
+Patients had routine measurements of inflammatory markers including C-reactive protein (CRP) and differential white cell counts from the day before surgery and every day for at least a week after surgery.
+These results including the preoperative blood tests and the postoperative blood tests for the first 7 days after surgery were collected from the hospital laboratory databases.
+
+All complications were discussed at a weekly morbidity meeting and prospectively recorded in an electronic database.
+The diagnosis and grading of pancreas-specific complications including postoperative pancreatic fistula (POPF, Section \ref{sec:ch_intro_POPF}) and post-pancreatectomy haemorrhage (PPH, Section \ref{sec:ch_intro_PPH}) were made according to the International Study Group classifications as shown in Table \ref{table:isgps_popf} and Table \ref{table:isgps_pph} on p\pageref{table:isgps_popf} respectively.
+
+All other complications were graded using the Clavien-Dindo system as shown in Table \ref{table:clavien-dindo}.
+This allows the grading of the severity of the complications based on the magnitude of the intervention(s) required to treat them.
 Postoperative mortality was defined as death within 30-days of the operation or while still in hospital after the operation.
 Re-intervention in the form of radiological, endoscopic or surgical procedures was recorded prospectively.
-Preoperative blood results as well as postoperative blood results for the first 7 days after surgery were collected from the hospital laboratory databases.
 
 \subsection{Statistics}
-SPSS version 22 was used for all analysis. ROC analysis was used to identify the optimum thresholds of CRP for predicting complications.
+Mann-Whitney U test was used to compare the distribution of postoperative CRP (as a continuous variable) in patients who had a complication and those who did not.
+Data are presented as median (inter-quartile range, IQR) unless otherwise specified.
+Line plots with error bars depicting 95\% confidence intervals are used to depict the trends in CRP over time across different patient groups.
+
+Receiver operating characteristic (ROC) analysis was used to identify the optimum thresholds for postoperative CRP for predicting infectious complications with a preference for thresholds that had a greater negative predictive value.
+Patients were categorised using these thresholds to analyse the relationship between CRP (as a categorical variable) and re-operation, hospital stay, critical care stay, number of critical care admissions as well as postoperative mortality using the Chi-square test of proportions.
+
+SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis.
+Effects were considered significant at $\alpha \leq0.05$.
+
 \todo{Complete this section - most of this can be modified from the stats sections in other chapters}
 
-\clearpage
 \section{Results}
-One hundred and eighty patient (131 male) underwent pancreaticoduodenectomy during the study period.
+Pancreaticoduodenectomy was performed in 188 patients (126 male, 67\%) during the study period. The median age was 63.5 years (IQR 54 - 70 years). 79 (42\%) patients were over the age of 65 years.
+
+Post-operative pancreatic fistula (POPF) as defined by the International Study Group for Pancreatic Fistula (ISGPF) occurred in 61 (33\%) patients.
+Of these patients, 18 (10.1\%) had a Grade A POPF, 25 (13.3\%) had a Grade B POPF and 18 (9.6\%) had a Grade C POPF.
+Significant infectious complications of Clavien-Dindo grade 3 or higher occurred in 84 patients.
+Infectious complications were more common in patients with age $>$65 years (35.6\% vs 50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 mls/kg/min (28.8\% vs.
+52.4\%, p=0.006) were significantly associated with infectious complications.
+Gender, body mass index, smoking status, modified Glasgow Prognostic Score, preoperative serum bilirubin and POSSUM physiology score were not associated with infectious complications.
+
+Infectious complications were also significantly associated with other complications including post-operative pancreatic fistula, post-pancreatectomy haemorrhage, length of stay in hospital, critical care stay, number of critical care admissions, reoperation rates and in-hospital mortality.
+(Table \ref{table:crp_comp_infect_vs_other_complications})
+
+\input{Tables/crp_comp_preop_factors}
+\input{Tables/crp_comp_infect_vs_other_complications}
 
 %CRP trends and pancreatic fistula
-Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula.
+Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p<0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
 However, CRP levels in the first postoperative week were not associated with the clinical severity of the pancreatic fistula.
 These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig.
-\ref{fig:crp_comp_crp_popf}.
+\ref{fig:crp_comp_crp_popf} on p\pageref{fig:crp_comp_crp_popf}.
 Fig.
-\ref{fig:crp_comp_crp_popf_isgps} shows that CRP levels were not significantly different between the different grades of pancreatic fistula while CRP levels in patients who did not develop a pancreatic fistula had significantly lower CRP levels.
+\ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not signficantly different between postoperative pancreatic fistulae of ISGPS Grade A, B and C.
+
+However, there is no useful clinical application for the association between CRP in the early postoperative period and  pancreatic fistula as the diagnosis of POPF is based on the the amylase content in surgical drains on the third postoperative day.
+Moreover, early postoperative CRP did not predict the severity of the pancreatic fistula either.
+
+The role of early postoperative CRP in predicting other infectious complications in the 137 patients who did not develop a POPF was analysed.
+The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infectious complications in the absence of a POPF (Fig.
+\ref{fig:crp_comp_infectious_leak0}).
+Moreover, the median difference in CRP also increases from 56 mg/dl on the third day to 79 mg/dl on the seventh day.
+However, there was no such association in patients who did develop a POPF (Fig.
+\ref{fig:crp_comp_infectious_leak1}).
+For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infectious complication was 97 mg/dl (IQR 71 - 174) but was 173 mg/dl if they had an infectious complication.
+(p$<$0.001).
+However, in patients who had POPF, the CRP on day 4 was 214 mg/dl (IQR 151 - 250) in the absence of infectious complications and 190 mg/dl (IQR 137 - 252) in the presence of infectious complications.
+(p = 0.559)
+
+Therefore, CRP on or after the third postoperative day appears to be useful in predicting infectious complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_infectious_leak} on p\pageref{fig:crp_comp_infectious_leak}.
+
+Receiver operating characteristic (ROC) analysis was undertaken in patients without POPF to find the optimum threshold of CRP that was associated with infectious complications.
+The area under curve (AUC) was significantly higher than 0.5 for CRP on days 3 though 7.
+The optimum thresholds alongwith the corresponding sensitivity, specificity, negative predictive value, postive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
+\ref{fig:crp_comp_ROC_infection} on p\pageref{fig:crp_comp_ROC_infection}.
+
+The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infectious complication.
+A CRP level of 125 mg/dl had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infectious complications.
+The AUC improved to 0.80 on day 7, with corresponding sensitivity, specificity and negative predictive values of 65\%, 81\% and 87\% respectively.
 
 %SPSS results and graphs are stored in crp_complications_graphs_etc.spv file
-
 %========================CRP vs POPF============================================
@@
 \clearpage
-\begin{figure}[htbp]
+\begin{figure}[t]
+	\caption{Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula (POPF).}
+	\label{fig:crp_comp_crp_popf}
 	\centering
-	\begin{subfigure}{0.45\textwidth}
+	\begin{subfigure}{0.48\textwidth}
 		\centering
@@
 	\end{subfigure}
-	\begin{subfigure}{0.45\textwidth}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
 		\centering
@@
 	\end{subfigure}
-
-	\caption{Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula (POPF).}
-	\label{fig:crp_comp_crp_popf}
-
 \end{figure}
+\hfill
 \input{Tables/crp_comp_vs_POPF_ISGPS_p_values_only}
@@
 \begin{figure}[t]
+	\caption{Relationship between postoperative CRP and clinically significant infectious complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_infectious_leak}
 	\centering
-	\begin{subfigure}{0.45\textwidth}
+	\begin{subfigure}{0.48\textwidth}
 		\centering
@@
 	\end{subfigure}
-
-	\caption{Relationship between postoperative CRP and clinically significant infectious complications in patients without (A) and with (B) POPF.}
-	\label{fig:crp_comp_infectious_leak}
-
 \end{figure}
+\vfill
 \input{Tables/crp_comp_vs_infections_popf_y1n0}
@@
 \begin{figure}[t]
+	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications.}
+	\label{fig:crp_comp_ROC_infection}
+
 	\centering
@@
 	\end{subfigure}
+	\hfill
 	\begin{subfigure}{0.3\textwidth}
@@
 	\end{subfigure}
-
-	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications.}
-	\label{fig:crp_comp_ROC_infection}
-
 \end{figure}
+\vfill
 \input{Tables/crp_comp_ROC_infections_noPOPF}
 
+%=============================================================================
+%Quick analysis of CPET vs postop CRP - maybe one table
+%ROC analysis of CRP vs LOS
+%Multivariate analysis of CRP as continuous or categorical vs complications, 
 \clearpage

### 2015-07-12 Chapters/Introduction.tex
Minor changes to Introduction
- Include section labels for POPF, PPH
- vfill/hfill in ISGPS tables

@@
 \subsubsection{Postoperative pancreatic fistula}
+\label{sec:ch_intro_POPF}
 Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability.
@@
 \subsubsection{Post-pancreatectomy haemorrhage}
+\label{sec:ch_intro_PPH}
 Post-pancreatectomy haemorrhage is reported to occur in 1 to 8\% of patients undergoing pancreaticoduodenectomy.

### 2015-07-15 Chapters/CRP_Complications.tex
CRP_Comp - Massive amount of work
- See individual files for changes
- Epic all-nighter while on-call

@@
 
-Postoperative CRP levels have been shown to predict infectious complications after colorectal surgery as well as oesophago-gastric surgery.
+Postoperative CRP levels have been shown to predict infective complications after colorectal surgery as well as oesophago-gastric surgery.
 It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
-The role of postoperative CRP in predicting the severity of complications after major pancreatic surgery has not been reported before.
-Stratifying patients based on the predicted severity of complications will allow identifying low risk patients who will be suitable to continue on enhanced recovery pathways and high risk patients who may require prolonged critical care, prolonged hospitalisation or further interventions.
 
-The aim of this study was to investigate the association between postoperative systemic inflammation and severity of complications after pancreaticoduodenectomy.
+But, the mechanisms underlying the pathogenesis of post-operative pancreatic fistula are largely attributed to anatomical factors including soft texture of the pancreas and small pancreatic duct diameter as well as surgical technique with increased intra-operative blood loss and pancreatico-jejunostomy rather than pancreatico-gastrostomy being associated with increased incidence of anastomotic leakage.
+The association between the early postoperative systemic inflammatory response and the incidence and severity of postoperative pancreatic fistula has not been reported.
+Moreover, the role postoperative CRP in predicting infective complications when these occur in association with postoperative pancreatic fistula has not been studied either.
+
+The aim of this study was to investigate the association between postoperative systemic inflammation, postoperative pancreatic fistula and infective complications in patients undergoing pancreaticoduodenectomy.
 
@@
 \todo{Check dates for this study}
-Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 were included in the study.
+Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 were included in this study.
 Patients who underwent only a trial dissection or trial dissection and surgical bypass during this period were excluded.
@@
 All patients had general anaesthesia supplemented either with patient-controlled epidural analgesia or spinal diamorphine and patient-controlled opiate analgesia for postoperative pain relief.
+Octreotide (200$\mu$g) was administered subcutaneously three times a day for 5 days in all patients and continued longer in patients diagnosed with a postoperative pancreatic fistula.
 Patients were routinely admitted to the surgical high dependency unit where a standardised regimen was followed for early mobilisation, chest physiotherapy and early enteral nutrition.
@@
 
-All other complications were graded using the Clavien-Dindo system as shown in Table \ref{table:clavien-dindo}.
-This allows the grading of the severity of the complications based on the magnitude of the intervention(s) required to treat them.
+All other complications were graded using the Clavien-Dindo classification system as shown in Table \ref{table:clavien-dindo}.
+This allowed the grading of the severity of the complications based on the magnitude of the intervention(s) required to treat them.
 Postoperative mortality was defined as death within 30-days of the operation or while still in hospital after the operation.
@@
 
-Receiver operating characteristic (ROC) analysis was used to identify the optimum thresholds for postoperative CRP for predicting infectious complications with a preference for thresholds that had a greater negative predictive value.
-Patients were categorised using these thresholds to analyse the relationship between CRP (as a categorical variable) and re-operation, hospital stay, critical care stay, number of critical care admissions as well as postoperative mortality using the Chi-square test of proportions.
+Receiver operating characteristic (ROC) analysis was used to identify the optimum thresholds for postoperative CRP for predicting infective complications with a preference for thresholds that had a greater negative predictive value.
+Patients were categorised using these thresholds to analyse the relationship between CRP (as a categorical variable) and re-operation, hospital stay, critical care stay, number of critical care admissions as well as postoperative mortality using the Chi-square test.
 
-SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis.
-Effects were considered significant at $\alpha \leq0.05$.
+SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. Effects were considered significant at $\alpha \leq0.05$.
 
@@
 \section{Results}
+\subsection{POPF, infective complications and other outcomes}
 Pancreaticoduodenectomy was performed in 188 patients (126 male, 67\%) during the study period. The median age was 63.5 years (IQR 54 - 70 years). 79 (42\%) patients were over the age of 65 years.
 
-Post-operative pancreatic fistula (POPF) as defined by the International Study Group for Pancreatic Fistula (ISGPF) occurred in 61 (33\%) patients.
-Of these patients, 18 (10.1\%) had a Grade A POPF, 25 (13.3\%) had a Grade B POPF and 18 (9.6\%) had a Grade C POPF.
-Significant infectious complications of Clavien-Dindo grade 3 or higher occurred in 84 patients.
+Post-operative pancreatic fistula (POPF) as defined by the International Study Group for Pancreatic Fistula (ISGPF) occurred in 62 (33\%) patients.
+Of these patients, 19 (10.1\%) had a Grade A POPF, 25 (13.3\%) had a Grade B POPF and 18 (9.6\%) had a Grade C POPF.
+Significant infective complications of Clavien-Dindo grade 3 or higher occurred in 84 patients.
 Infectious complications were more common in patients with age $>$65 years (35.6\% vs 50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 mls/kg/min (28.8\% vs.
-52.4\%, p=0.006) were significantly associated with infectious complications.
-Gender, body mass index, smoking status, modified Glasgow Prognostic Score, preoperative serum bilirubin and POSSUM physiology score were not associated with infectious complications.
+52.4\%, p=0.006) were significantly associated with infective complications.
+Gender, body mass index, smoking status, modified Glasgow Prognostic Score, preoperative serum bilirubin and POSSUM physiology score were not associated with infective complications.
 
@@
 %CRP trends and pancreatic fistula
+\subsection{Postoperative CRP and POPF}
 Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p<0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
-However, CRP levels in the first postoperative week were not associated with the clinical severity of the pancreatic fistula.
 These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig.
@@
 
-The role of early postoperative CRP in predicting other infectious complications in the 137 patients who did not develop a POPF was analysed.
-The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infectious complications in the absence of a POPF (Fig.
-\ref{fig:crp_comp_infectious_leak0}).
-Moreover, the median difference in CRP also increases from 56 mg/dl on the third day to 79 mg/dl on the seventh day.
-However, there was no such association in patients who did develop a POPF (Fig.
-\ref{fig:crp_comp_infectious_leak1}).
-For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infectious complication was 97 mg/dl (IQR 71 - 174) but was 173 mg/dl if they had an infectious complication.
-(p$<$0.001).
-However, in patients who had POPF, the CRP on day 4 was 214 mg/dl (IQR 151 - 250) in the absence of infectious complications and 190 mg/dl (IQR 137 - 252) in the presence of infectious complications.
-(p = 0.559)
-
-Therefore, CRP on or after the third postoperative day appears to be useful in predicting infectious complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_infectious_leak} on p\pageref{fig:crp_comp_infectious_leak}.
-
-Receiver operating characteristic (ROC) analysis was undertaken in patients without POPF to find the optimum threshold of CRP that was associated with infectious complications.
-The area under curve (AUC) was significantly higher than 0.5 for CRP on days 3 though 7.
-The optimum thresholds alongwith the corresponding sensitivity, specificity, negative predictive value, postive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
-\ref{fig:crp_comp_ROC_infection} on p\pageref{fig:crp_comp_ROC_infection}.
-
-The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infectious complication.
-A CRP level of 125 mg/dl had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infectious complications.
-The AUC improved to 0.80 on day 7, with corresponding sensitivity, specificity and negative predictive values of 65\%, 81\% and 87\% respectively.
-
-%SPSS results and graphs are stored in crp_complications_graphs_etc.spv file
 %========================CRP vs POPF============================================
@@
 %==============================================================================
+\clearpage
 
-%Ability of CRP to predict infectious complications in leak vs. no leak complications
-%Rationale - POPF is diagnosed with a separate definition - ISGPS on Day 3.
-Therefore, there is no role for CRP in diagnosing POPF.
-Above results show that it does not discriminate between the grades either.
-Therefore, the following analysis is to find out if the first week CRP has any role in predicting other outcome measures.
+\subsection{Postoperative CRP and infective complications}
+The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
+\ref{fig:crp_comp_infective_leak0}).
+Moreover, the median difference in CRP also increases from 56 mg/dl on the third day to 79 mg/dl on the seventh day.
+However, there was no such association in patients who did develop a POPF (Fig.
+\ref{fig:crp_comp_infective_leak1}).
+For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/dl (IQR 71 - 174) but was 173 mg/dl if they had an infective complication.
+(p$<$0.001).
+However, in patients who had POPF, the CRP on day 4 was 214 mg/dl (IQR 151 - 250) in the absence of infective complications and 190 mg/dl (IQR 137 - 252) in the presence of infective complications.
+(p = 0.559)
+
+Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_infective_leak} on p\pageref{fig:crp_comp_infective_leak}.
 
@@
 \begin{figure}[t]
-	\caption{Relationship between postoperative CRP and clinically significant infectious complications in patients without (A) and with (B) POPF.}
-	\label{fig:crp_comp_infectious_leak}
+	\caption{Relationship between postoperative CRP and clinically significant infective complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_infective_leak}
 	\centering
@@
 		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_infectious_leak0}
+		\includegraphics[width=\textwidth]{Figures/crp_comp_infective_leak0}
 		\caption{Patients with \textbf{\underline{no}} POPF}
-		\label{fig:crp_comp_infectious_leak0}
+		\label{fig:crp_comp_infective_leak0}
 	\end{subfigure}
@@
 		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_infectious_leak1}
+		\includegraphics[width=\textwidth]{Figures/crp_comp_infective_leak1}
 		\caption{Patients with POPF}
-		\label{fig:crp_comp_infectious_leak1}
+		\label{fig:crp_comp_infective_leak1}
 	\end{subfigure}
@@
 %==============================================================================
+\clearpage
+
+\subsection{Postoperative white cell counts, albumin and infective complications}
+
+Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically signficant.
+However, these differences were small with wide, overlapping 95\% confidence intervals (Fig.
+\ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant.
+For instance, the median white cell count on the fifth postoperative day in patients who did not have an infective complication was 8.9 (IQR 6.6 - 11) while in patients who had an infective complication it was 9.8 (IQR 7.9 - 13).
+While this difference was statistically significant (p = 0.028), both medians were within the normal limits for healthy adults.
+There was no association between white cell counts and infective complications in patients who had a POPF.
+(Table \ref{table:crp_comp_WCC_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_WCC_infective_leak} on p\pageref{fig:crp_comp_WCC_infective_leak})
+
+Similar statistically significant differences were noted in the postoperative neutrophil counts on postoperative days 5, 6 and 7 in patients without a POPF while there was no difference in patients with a POPF.
+(Table \ref{table:crp_comp_Neutrophil_vs_infections_popf_y1n0})	However, these differences were associated with a large overlap between the 95\% confidence intervals (error bars in Fig.
+\ref{fig:crp_comp_Neutrophil_infective_leak0}).
+
+Patients with infective complications in the absence of POPF had a lower median serum albumin  during the entire first postoperative week with statistically significant differences from days 3 through 7.
+There was no association between serum albumin and infective complications in patients who had a POPF.
+(Table \ref{table:crp_comp_Albumin_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_Albumin_infective_leak} on p\pageref{fig:crp_comp_Albumin_infective_leak})
+
+%========================WCC vs Infections====================================
+%WCC trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
+\clearpage
+\begin{figure}[t]
+	\caption{Relationship between postoperative white cell count and clinically significant infective complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_WCC_infective_leak}
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_WCC_infective_leak0}
+		\caption{Patients with \textbf{\underline{no}} POPF}
+		\label{fig:crp_comp_WCC_infective_leak0}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_WCC_infective_leak1}
+		\caption{Patients with POPF}
+		\label{fig:crp_comp_WCC_infective_leak1}
+	\end{subfigure}
+\end{figure}
+\vfill
+\input{Tables/crp_comp_WCC_vs_infections_popf_y1n0}
+%==============================================================================
+
+%==========Neutrophil-Neutrophil-Neutrophil-Neutrophil-Neutrophil-Neutrophil=====
+%Neutrophil trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
+\clearpage
+\begin{figure}[t]
+	\caption{Relationship between postoperative neutrophil count and clinically significant infective complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_Neutrophil_infective_leak}
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_Neutrophil_infective_leak0}
+		\caption{Patients with \textbf{\underline{no}} POPF}
+		\label{fig:crp_comp_Neutrophil_infective_leak0}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_Neutrophil_infective_leak1}
+		\caption{Patients with POPF}
+		\label{fig:crp_comp_Neutrophil_infective_leak1}
+	\end{subfigure}
+\end{figure}
+\vfill
+\input{Tables/crp_comp_Neutrophil_vs_infections_popf_y1n0}
+%==============================================================================
+
+%==========Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-=====
+%Albumin trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
+\clearpage
+\begin{figure}[t]
+	\caption{Relationship between postoperative serum albumin and clinically significant infective complications in patients without (A) and with (B) POPF.}
+	\label{fig:crp_comp_Albumin_infective_leak}
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_Albumin_infective_leak0}
+		\caption{Patients with \textbf{\underline{no}} POPF}
+		\label{fig:crp_comp_Albumin_infective_leak0}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/crp_comp_Albumin_infective_leak1}
+		\caption{Patients with POPF}
+		\label{fig:crp_comp_Albumin_infective_leak1}
+	\end{subfigure}
+\end{figure}
+\vfill
+\input{Tables/crp_comp_Albumin_vs_infections_popf_y1n0}
+%=============================================================================
+\clearpage
+
+\subsection{Receiver operating characteristic (ROC) analysis}
+Receiver operating characteristic (ROC) analysis was undertaken in order to find the optimum threshold for the relationship between CRP and infective complications in patients without POPF.
+ROC curves were plotted with CR as a continuous variable and infective complications as a dichotomous outcome variable for postoperative days 3 through 7.
+
+The area under curve (AUC) was significantly higher than 0.5 for CRP on days 3 though 7.
+The optimum thresholds along with the corresponding sensitivity, specificity, negative predictive value, postive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
+\ref{fig:crp_comp_ROC_infection} on p\pageref{fig:crp_comp_ROC_infection}.
+The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infective complication.
+A CRP level of 125 mg/dl on the fourth postoperative day had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infective complications (AUC 0.71, 95\%CI 0.61 - 0.80, p$<$0.001).
+The AUC improved to 0.80 on day 7 but only with a modest increase in the negative predictive value to 87\%, with corresponding sensitivity, specificity of 65\% and 81\% respectively (95\%CI 0.72 - 0.88, p$<$0.001).
+
+The relationship between CRP on the fourth day and other postoperative events is shown in Table \ref{table:crp_comp_CRP4_vs_LOS}.
+There were no deaths in this group of patients who did not have a POPF.
+
+%SPSS results and graphs are stored in crp_complications_graphs_etc.spv file
 
@@
 \input{Tables/crp_comp_ROC_infections_noPOPF}
+%=============================================================================
 
-%=============================================================================
+%===================CRP Day 4 vs LOS=========================================
+
+\input{Tables/crp_comp_CRP4_vs_LOS}
+
 %Quick analysis of CPET vs postop CRP - maybe one table
@@
 \section{Discussion}
+
+The results of the present study show that the severity of the postoperative systemic inflammatory response as measured by serial serum CRP levels in the first week is associated with postoperative complications after pancreaticoduodenectomy.
+While numerous studies have reported on a variety of methods to predict as well as diagnose POPF, few have reported on the role of CRP in predicting complications other than POPF.
+
+The introduction of the ISGPF definition for POPF has made the diagnosis of this complication simple and uniform.
+The sole criterion for the diagnosis of POPF is the drain amylase content on the third postoperative day being three times the upper limit of normal for serum amylase in the testing laboratory.
+
+In the present study, while CRP was significantly greater in patients with POPF, this difference did not occur until the second postoperative day and therefore had no further clinical use in the prediction of this complication.
+Moreover, the magnitude of increase in CRP during the first postoperative week was not associated the severity of POPF either.
+
+This lack of association appears to suggest that the determinants of the severity of POPF may occur later in the postoperative course or may not be related to the magnitude of the initial systemic inflammatory response.
+The presumed cascade of events following a POPF involves accumulation of amylase rich fluid in the peritoneal cavity around recently dissected tissues and ligated blood vessels.
+This can lead to localised collections that can become secondarily infected, result in erosion of blood vessels resulting in post-pancreatectomy haemorrhage, result in delayed gastric emptying or postoperative ileus as well as respiratory complications including sympathetic pleural effusions or pneumonic consolidation.
+In fact, most large series report a close association between POPF and PPH with the latter often following the former.
+The eventual outcome in patients with POPF is often determined by the cascade of complications associated with the POPF itself.
+In fact, all deaths in this study were in patients who developed POPF.
+
+However, in patients without a POPF, CRP in the first week after surgery was associated with infective complications.
+The postoperative course of patients who do not develop a POPF is not dissimilar to that in patients undergoing other major gastrointestinal resectional surgery such as colorectal or oesophago-gastric surgery.
+The impact of the systemic inflammatory response and the compensatory anti-inflammatory response in these patients is not confounded by the presence of POPF.
+Our results show that once the absence of POPF is confirmed by normal drain amylase content on the third postoperative day, serum CRP on the next (fourth) postoperative day less than 125 mg/dl had a negative predictive value of 83\% in excluding an infective complication.
+
+The results of this study are similar to the findings of Welsch and co-workers who analysed 688 patients undergoing a variety of pancreatic resections with pancreaticojejunal anastomosis.\parencite{welsch_persisting_2008} They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections.
+They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%).
+However, the definitions used for POPF in this study were different from the ISPGF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF.
+Moreover, Grade A and B POPF were not included in the analysis.
+Our results clearly demonstrate that early CRP elevation occurs in all grades of POPF with no difference between the different grades.
+
+In the present study, preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score was not associated with infective complications.
+The modified Glasgow Prognostic Score has been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections.
+The lack of association in the present study may have been due to the multiple factors that may have been responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
+These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery.\parencite{dutta_persistent_2011}
+
+The advent of enhanced recovery protocols in surgery including pancreatic surgery requires the early identification of patients not only at increased risk of complications but also those who are likely to have an uncomplicated course.
+This allows early identification of which patients who are expected to continue on the enhanced recovery pathway and therefore allows better discharge planning while in-hospital and critical care resources are utilised more efficiently in the management of the high risk patients.
+
+%=========Strengths and limitations==========================

### 2015-07-15 Tables/crp_comp_CRP4_vs_LOS.tex
CRP_Comp - Massive amount of work
- See individual files for changes
- Epic all-nighter while on-call

%12/07/15 - Started this table
\begin{table}[p]
	\centering
	\caption{The relationship between CRP on Day 4 and other postoperative events in patient with \textbf{no} POPF.}
	\label{table:crp_comp_CRP4_vs_LOS}
	\renewcommand{\arraystretch}{1.2} %Increases space between rows
	%\setlength{\tabcolsep}{9pt} %sets the space between columns
	\begin{tabular}{|l l | c c r |}
		\hline
		                      &          &  \multicolumn{3}{c|}{Day 4 CRP (mg/dl)}   \\
		                      &          & $\leq$125   & $>$125       & p            \\ \hline
		Postoperative stay    & Days     & 12 (9 - 16) & 15 (13 - 26) & $<$0.001$^a$ \\
		(days)                & $\leq$14 & 43 (66.2\%) & 30 (49.2\%)  & $<$0.054$^b$ \\
		                      & $>$14    & 22 (33.8\%) & 31 (50.8\%)  &  \\
		Critical care stay    & Days     & 6 (4 - 7)   & 7 (5 - 10)   & 0.001$^a$    \\
		(days)                & $\leq$7  & 57 (87.7\%) & 38 (62.3\%)  & 0.001$^b$    \\
		                      & $>$7     & 8 (12.3\%)  & 23 (37.7\%)  &  \\
		Crit. care admissions & 1        & 60 (92.3\%) & 48 (78.7\%)  & 0.029$^b$    \\
		                      & $>$1     & 5 (7.7\%)   & 13 (21.3\%)  &  \\
		Reoperation           & No       & 63 (96.9\%) & 57 (93.4\%)  & $<$0.359$^b$ \\
		                      & Yes      & 2 (3.1\%)   & 4 (6.6\%)    &  \\ \hline
		\multicolumn{5}{l}{a - Mann-Whitney U test; b - Chi-square test}
	\end{tabular}
\end{table}

### 2015-07-16 Chapters/CRP_Complications.tex
CRP_Comp - Completed Draft
- Completed draft chapter - dont think will need any major changes

@@
 \chapter{An investigation into the relationship between postoperative systemic inflammation and complications after pancreaticoduodenectomy.}
-\label{ch_survival}
+\label{ch_crp_comp}
 
-\lhead{Chapter \ref{ch_survival}. \emph{Postoperative CRP and complications}} % This is for the header on each page - perhaps a shortened title
+\lhead{Chapter \ref{ch_crp_comp}. \emph{Postoperative CRP and complications}} % This is for the header on each page - perhaps a shortened title
 
@@
 
-Postoperative CRP levels have been shown to predict infective complications after colorectal surgery as well as oesophago-gastric surgery.
-It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
+Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery.
+\parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008} It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
 
@@
 
-The aim of this study was to investigate the association between postoperative systemic inflammation, postoperative pancreatic fistula and infective complications in patients undergoing pancreaticoduodenectomy.
+The aim of this study was to investigate the association between systemic inflammation, postoperative pancreatic fistula and infective complications in patients undergoing pancreaticoduodenectomy.
 
@@
 
-Receiver operating characteristic (ROC) analysis was used to identify the optimum thresholds for postoperative CRP for predicting infective complications with a preference for thresholds that had a greater negative predictive value.
+Receiver operating characteristic (ROC) analysis \parencite{robertson_use_1981,zweig_receiver-operating_1993} was used to identify the optimum thresholds for postoperative CRP for predicting infective complications with a preference for thresholds that had a greater negative predictive value.
 Patients were categorised using these thresholds to analyse the relationship between CRP (as a categorical variable) and re-operation, hospital stay, critical care stay, number of critical care admissions as well as postoperative mortality using the Chi-square test.
@@
 Significant infective complications of Clavien-Dindo grade 3 or higher occurred in 84 patients.
-Infectious complications were more common in patients with age $>$65 years (35.6\% vs 50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 mls/kg/min (28.8\% vs.
+Infectious complications were more common in patients with age $>$65 years (35.6\% vs.
+50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 ml/kg/min (28.8\% vs.
 52.4\%, p=0.006) were significantly associated with infective complications.
@@
 
-Infectious complications were also significantly associated with other complications including post-operative pancreatic fistula, post-pancreatectomy haemorrhage, length of stay in hospital, critical care stay, number of critical care admissions, reoperation rates and in-hospital mortality.
+Infectious complications were also significantly associated with other complications including post-operative pancreatic fistula, post-pancreatectomy haemorrhage, length of stay in hospital, critical care stay, number of critical care admissions, re-operation rates and in-hospital mortality.
 (Table \ref{table:crp_comp_infect_vs_other_complications})
@@
 Fig.
-\ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not signficantly different between postoperative pancreatic fistulae of ISGPS Grade A, B and C.
+\ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not significantly different between postoperative pancreatic fistulae of ISGPF Grade A, B and C.
 
@@
 		\includegraphics[width=\textwidth]{Figures/crp_comp_crp_popf_isgps}
-		\caption{POPF ISGPS Grades}
+		\caption{POPF ISGPF Grades}
 		\label{fig:crp_comp_crp_popf_isgps}
@@
 \ref{fig:crp_comp_infective_leak0}).
-Moreover, the median difference in CRP also increases from 56 mg/dl on the third day to 79 mg/dl on the seventh day.
+Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day.
 However, there was no such association in patients who did develop a POPF (Fig.
 \ref{fig:crp_comp_infective_leak1}).
-For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/dl (IQR 71 - 174) but was 173 mg/dl if they had an infective complication.
+For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
 (p$<$0.001).
-However, in patients who had POPF, the CRP on day 4 was 214 mg/dl (IQR 151 - 250) in the absence of infective complications and 190 mg/dl (IQR 137 - 252) in the presence of infective complications.
+However, in patients who had POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
 (p = 0.559)
@@
 
-Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically signficant.
-However, these differences were small with wide, overlapping 95\% confidence intervals (Fig.
+Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically significant However, these differences were small with wide, overlapping 95\% confidence intervals (Fig.
 \ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant.
@@
 The area under curve (AUC) was significantly higher than 0.5 for CRP on days 3 though 7.
-The optimum thresholds along with the corresponding sensitivity, specificity, negative predictive value, postive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
+The optimum thresholds along with the corresponding sensitivity, specificity, negative predictive value, positive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
 \ref{fig:crp_comp_ROC_infection} on p\pageref{fig:crp_comp_ROC_infection}.
 The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infective complication.
-A CRP level of 125 mg/dl on the fourth postoperative day had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infective complications (AUC 0.71, 95\%CI 0.61 - 0.80, p$<$0.001).
+A CRP level of 125 mg/L on the fourth postoperative day had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infective complications (AUC 0.71, 95\%CI 0.61 - 0.80, p$<$0.001).
 The AUC improved to 0.80 on day 7 but only with a modest increase in the negative predictive value to 87\%, with corresponding sensitivity, specificity of 65\% and 81\% respectively (95\%CI 0.72 - 0.88, p$<$0.001).
@@
 The introduction of the ISGPF definition for POPF has made the diagnosis of this complication simple and uniform.
-The sole criterion for the diagnosis of POPF is the drain amylase content on the third postoperative day being three times the upper limit of normal for serum amylase in the testing laboratory.
+The rate of POPF before the standardised definition was introduced varied anywhere between 2\% and 40\%.
+However, in the post-ISGPF era, the rate of POPF in published literature has remained approximately 25\%.
+The sole criterion for the diagnosis of POPF is a drain amylase content on or after the third postoperative day three times the upper limit of normal for serum amylase in the testing laboratory.
 
 In the present study, while CRP was significantly greater in patients with POPF, this difference did not occur until the second postoperative day and therefore had no further clinical use in the prediction of this complication.
-Moreover, the magnitude of increase in CRP during the first postoperative week was not associated the severity of POPF either.
+The early rise in CRP in patients who developed a POPF is expected as the most common risk factors for POPF are encountered intra-operatively including soft pancreatic texture, small pancreatic duct size and operative blood loss.
+Soft pancreatic texture has in fact been reported to be associated with elevated CRP levels after pancreaticoduodenectomy.\parencite{murakami_soft_2008}
 
-This lack of association appears to suggest that the determinants of the severity of POPF may occur later in the postoperative course or may not be related to the magnitude of the initial systemic inflammatory response.
+However, the magnitude of increase in CRP during the first postoperative week was not associated with the severity of POPF.
+This lack of association appears to suggest that the determinants of the severity of POPF may occur later in the postoperative course and are not related to the magnitude of the initial systemic inflammatory response.
+
 The presumed cascade of events following a POPF involves accumulation of amylase rich fluid in the peritoneal cavity around recently dissected tissues and ligated blood vessels.
@@
 
-However, in patients without a POPF, CRP in the first week after surgery was associated with infective complications.
+However, in patients without a POPF, CRP in the first week after surgery was associated with infective complications and preceded the clinical manifestation of complications.
 The postoperative course of patients who do not develop a POPF is not dissimilar to that in patients undergoing other major gastrointestinal resectional surgery such as colorectal or oesophago-gastric surgery.
 The impact of the systemic inflammatory response and the compensatory anti-inflammatory response in these patients is not confounded by the presence of POPF.
-Our results show that once the absence of POPF is confirmed by normal drain amylase content on the third postoperative day, serum CRP on the next (fourth) postoperative day less than 125 mg/dl had a negative predictive value of 83\% in excluding an infective complication.
+Our results show that once the absence of POPF is confirmed by normal drain amylase content on the third postoperative day, serum CRP on the next (fourth) postoperative day less than 125 mg/L had a negative predictive value of 83\% in excluding an infective complication.
 
@@
 They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%).
-However, the definitions used for POPF in this study were different from the ISPGF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF.
+However, the definitions used for POPF in this study were different from the ISGPF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF.
 Moreover, Grade A and B POPF were not included in the analysis.
@@
 In the present study, preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score was not associated with infective complications.
-The modified Glasgow Prognostic Score has been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections.
-The lack of association in the present study may have been due to the multiple factors that may have been responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
+A preoperative acute phase response has been reported to be associated with an exaggerated postoperative systemic inflammatory response and septic complications.
+\parencite{haupt_association_1997} The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections.\parencite{moyes_preoperative_2009} The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
 These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery.\parencite{dutta_persistent_2011}
 
-The advent of enhanced recovery protocols in surgery including pancreatic surgery requires the early identification of patients not only at increased risk of complications but also those who are likely to have an uncomplicated course.
-This allows early identification of which patients who are expected to continue on the enhanced recovery pathway and therefore allows better discharge planning while in-hospital and critical care resources are utilised more efficiently in the management of the high risk patients.
+The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication.
+This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery.\parencite{mustard_c-reactive_1987} as well as in several subsequent reports.
+\parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011} The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation.
+This study was not designed to look at inflammatory markers later in the postoperative period.
+However, Mustard and co-workers monitored CRP, white cell count and temperature for 14 days after surgery and found that CRP was the only reliable marker of septic complications.
+
+Several mechanisms have been postulated to explain the association between postoperative CRP and complications.
+The most common explanation is that the CRP rise is consequent to tissue ischaemia, necrosis at the anastomotic site, bacterial translocation from the gut or bacterial infection of wounds.
+However, our findings and that of other authors show that the difference in CRP occurs very early in the postoperative period before infective complications become clinically manifest.
+
+More recently, the compensatory anti-inflammatory response with the associated immunosuppressive effects has been recognised as a possible factor in predisposing the patient to infective complications and delayed wound healing.
+Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications.
+Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer.
+They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day.
+T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6.\parencite{faist_immunosuppression_1997} The complex interactions between the pro and anti-inflammatory processes, neurohormonal factors of the stress response, the role of the blood-gut barrier in bacterial translocation, local factors including tissue ischaemia and damage, the immunological effects of anaesthesia, blood transfusion and hypothermia remain the subject of ongoing research.\parencite{buttenschoen_effect_2010}
 
 %=========Strengths and limitations==========================
+One of the most important strengths of this study is the separate analysis of infective complications in patients who did not develop a POPF.
+Other strengths of this study include the homogeneous cohort of patients all of who underwent a pancreaticoduodenectomy, the serial measurements of CRP on every day of the first postoperative week, prospective record of complications and use of the ISGPF definition for POPF.
+We did not measure levels of other inflammatory cytokines routinely in these patients.
+Future work should be aimed at characterising the inflammatory cytokine profile after pancreaticoduodenectomy to identify potential targets for immunomodulation in the perioperative period.
+
+In summary, the advent of enhanced recovery protocols in surgery including pancreatic surgery requires the early identification of patients not only at increased risk of complications but also those who are likely to have an uncomplicated course.
+This allows early identification of patients who are expected to continue on the enhanced recovery pathway and allocation of resources to improve the management of patients who are likely to develop complications.
+This study shows that CRP on the third postoperative day is associated with infective complications in patients undergoing pancreaticoduodenectomy who do not develop a POPF but not in patients with a POPF.

### 2015-07-16 Chapters/Introduction.tex
Numerous small formatting changes
- Linespacing set to 1.5
- Several tables placement changed from htbp to p
- Other minor changes
- Email this version to Donny

@@
 \subsection{Epidemiology of pancreatic cancer}
-\parencite{crozier_preoperative_2007}
 Tumours involving the head of the pancreas and the periampullary region account for a small proportion of gastrointestinal tumours.

### 2015-07-16 main.tex
CPET Survival Cox Univariate regression
- Complete CPET vs Survival Univariate table
- Create new SPSS dataset for this chapter -ds_at_pdac
- Remove redundant variables in this dataset
- Changes to main.tex are simply commenting out other chapters from output
Additions to results section of CPET Survival
- Add description of results of Cox regression to CPET Survival
- Shorten cpet_survival_cox_regression table by condensing rows into one per parameter
- Save long form of above table in separate file - not currently in use
- New table - cpet_survival_6months - needs fleshing out
- Add todo to

@@
 \usepackage{pdfpages} %for including pdf files or individual pages of pdf files
-\usepackage{easy-todo} %Adds todos to the document. Use [diable] as package option in final verion. Also see \listoftodos below.
+\usepackage{easy-todo} %Adds todos to the document. Use [disable] as package option in final verion. Also see \listoftodos below.
 \usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
@@
 
-\setstretch{2.0} % Line spacing of 2.0 although I have set \doublespace set somewhere else
+\setstretch{1.5} % Line spacing of 1.5 although I have set \doublespace set somewhere else - not sure where
 
@@
 
+\clearpage
 \lhead{\emph{List of Figures}} % Set the left side page header to "List of Figures"
@@
 
+\clearpage
 \lhead{\emph{List of Tables}} % Set the left side page header to "List of Tables"
@@
 % Include the chapters of the thesis as separate files from the Chapters folder
-\input{Chapters/Introduction}
-\input{Chapters/CPET_Outcomes} 
-\input{Chapters/CPET_Jaundice}
-\input{Chapters/BodyComp}
-\input{Chapters/CPET_Survival}
-\input{Chapters/Discussion}
+\input{Chapters/Introduction} 		%1
+\input{Chapters/CPET_Outcomes} 	%2
+\input{Chapters/CPET_Jaundice} 	%3
+\input{Chapters/BodyComp}			%4
+\input{Chapters/CPET_Survival}		%5
+\input{Chapters/CRP_Complications}	%6
+\input{Chapters/Discussion}		%7
 
@@
 
-%\input{Appendices/AppendixExcelMacros}
+\input{Appendices/AppendixExcelMacros}
 \input{Appendices/AppendixCPETRawData}
@@
 %\bibliographystyle{unsrtnat} % Use the "unsrtnat" BibTeX style for formatting the Bibliography
-\printbibliography % The references (bibliography) information are stored in the file named "Bibliography.bib"
+\printbibliography
 

### 2015-07-26 Chapters/Introduction.tex
Introduction - Multiple edits

@@
 
-Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection.
-Research is ongoing to assess the role of neoadjuvant therapy and newer treatment modalities such as electroporation in these patients to improve resectability.
+Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.
+The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced disease and in some patients with metastatic disease.
+Early reports suggest that some of these patients can undergo potentially curative surgery after such neo-adjuvant treatment.
 
 \subsubsection{Patient factors}
+
 \subsection{Operative technique}
@@
 Antibiotic prophylaxis is administered at induction.
-While the use of Octreotide, a somatostatin analogue, to reduce the risk of postoperative pancreatic fistula formation is still debated, it was routinely used in all patients at this centre.
-Octreotide was administered intra-operatively (200 mcg s.c.) and was continued for 5 days postoperatively (200 mcg s.c., t.d.s.).
+While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre.
+Octreotide was administered intra-operatively (200 micrograms subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms subcutaneous injection administered three times a day).
 
-A roof-top incision was used for access.
-After assessing the peritoneal cavity for absence of metastatic disease, an early assessment was made for local resectability.
+A roof-top incision was made to enter the abdominal cavity.
+After assessing the peritoneal cavity and liver for absence of metastatic disease, an early assessment was made for local resectability.
 This involved complete Kocherisation of the duodenum to assess the retroperitoneum.
@@
 The superior mesenteric vein was identified and a retro-pancreatic tunnel was created between the pancreatic neck and the portal vein.
-If less than half the circumference of the SMV or PV was involved, an en-bloc resection was performed with vein repair at the same time.
+If less than half the circumference of the superior mesenteric vein or portal vein was involved, an en-bloc resection was performed with vein repair at the same time.
 The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy.
-The gastro-duodenal artery was divided.
-Resection was then completed by dividing the stomach (classical Whipple procedure) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy, PPPD) and transecting the pancreatic neck.
+The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery should the patient require embolisation of this vessel for post-pancreatectomy haemorrhage.
+The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device.
+The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference.
+The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel.
+The retroportal lamina was divided preserving the superior mesenteric artery.
 
@@
 \subsection{Complications}
-The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006,emick_hospital_2006,yeo_six_1997, winter_1423_2006,teh_patient_2009,gouma_rates_2000}
+The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000} Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes.
+The consensus definitions of the International Study Group for Pancreatic Surgery (ISGPS) and the Clavien-Dindo classification systems were used during the course of the studies reported here.
 
@@
 \label{sec:ch_intro_POPF}
+
+\input{Tables/c1_isgps}
+
 Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability.
@@
 Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}.
-Grade B and C fistulae are considered to be clinically significant in that they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage, delayed gastric emptying as well as need for intervention (either radiological or operative) and/or prolonged critical care support.
+Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying.
+They may require further intervention (either radiological or operative) and may result in prolonged critical care support.
 
@@
 
-\input{Tables/c1_isgps}
-
 \subsubsection{Clavien-Dindo classification of complications}
-A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrythmias, pneumonia, pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal sepsis including intra-abdominal sepsis, leakage from the hepaticojejunostomy or the gastrojejunostomy, renal dysfunction, etc.
-The Clavien-Dindo method grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis.
-
 \input{Tables/c1_clavien_dindo}
+A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc.
+The Clavien-Dindo system grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis.
+Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.
 
@@
 
-\clearpage
+While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care.
+The European Study Group for Pancreatic Cancer (ESPAC) have conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect.
+\parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}
+
+The ESPAC-1 trial compared the use of chemotherapy with fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
+The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive demonstrating a clear survival benefit for adjuvant chemotherapy.
+However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting.
+The ESPAC-3 trial compared the efficacy of 5-FU and folinic acid with gemcitabine and reported that gemcitabine was not superior to 5-FU/folinic acid.
+
+However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma end up either not receiving or not completing adjuvant chemotherapy.
+\parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbdity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}
 
@@
 Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest.
-The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or chemoradiotherapy.\parencite{sandroussi_sociodemographics_2010}
+Major pancreatic surgery requires the patient to have adequate reserve to cope not only with the increased physiological demand during and immediately after surgery but also to cope with any adverse events that may occur.
 
-Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009}
+The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment.
+\parencite{sandroussi_sociodemographics_2010}
 
-Major pancreatic surgery requires the patient to have adequate physiological reserve to cope with the increased demand during and immediately after surgery.
-However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery.
+Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery.
 \parencite{shah_limitations_2012}
@@
 %Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
-Physiological fitness or reserve may be defined as the ability of the patient’s organ systems to respond appropriately and adequately to the stress of major surgery.
-Major surgery places a significant physiological stress on multiple organ systems, especially the cardiorespiratory system.
-The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and the postoperative recovery plays a major role in determining short-term outcomes.
-
-Accurate measurement of physiological fitness
+Physiological fitness or reserve may be defined as the ability of the patient's organ systems to respond appropriately and adequately to the stress of surgery.
+Major surgery has a significant physiological impact on multiple organ systems, especially the cardiorespiratory system.
+The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes.
+Therefore, accurate assessment of physiololgical fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions.
+A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness.
+Numerous methods including individual tests and composite scores derived from physiological and biochemical variables have been used for risk assessment.
 
@@
 \todo{Static Versus Dynamic Testing}
+
 Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase.
-Identification of such high-risk patients allows not only for improved patient selection, but also for risk-stratified, anaesthetic and postoperative critical care.
-Preoperative risk stratification will also allow for prehabilitation of these patients in an attempt to improve outcomes.
 
@@
 \todo{The physiological basis of CPET}
+
+\subsection{Origins and physiological basis of cardiopulmonary exercise testing}
+The measurement of aerobic fitness has been an important aspect of improving atheletic performance in competitive sports.
+Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes as well as offer customised training regimens aimed at improving performance at activities that place a signficant stress on the cardiovascular and respiratory systems.
 
@@
 \subsection{Measuring the Anaerobic Threshold}
+
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width=\textwidth]{Figures/cpet_9panel.pdf}
+	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
+	\label{fig:cpet_9panel}
+\end{figure}
+
 The anaerobic threshold (variously described as the lactate threshold or ventilatory threshold) is the point during exercise when oxygen demand by exercising skeletal muscle outstrips supply.
@@
 
-\begin{figure}[p]
+\subsubsection{V-slope method}
+
+\begin{figure}[htbp]
 	\centering
-	\includegraphics[width=0.9\textwidth]{Figures/cpet_9panel.pdf}
-	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
-	\label{fig:cpet_9panel}
+	\includegraphics[width=\textwidth]{Figures/cpet_vslope}
+	\caption{Determination of $\dot{V}_{O_2}$AT by the V-slope method.}
+	\label{fig:cpet_vslope}
 \end{figure}
 
-\subsubsection{V-slope method}
-
-During aerobic exercise, VO2 and VCO2 share a linear relationship as shown in segment A of the graph in figure 1.
-However, as anaerobic respiration starts to supplement aerobic respiration, VCO2 increases disproportionate to VO2 as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}.
-This results in a distinct difference in the slope of the initial part of the graph (seg A) and the later part (seg B).
-The point at which the two slopes intersect is the anaerobic threshold and the V02 at this point in exercise is commonly referred to as the anaerobic threshold, VO2at or simply AT.
+During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}.
+However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionately to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}.
+This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B).
+The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold, $\dot{V}_{O_2}$AT or simply AT.
 
@@
 \todo{Ventilatory equivalents method}
+The ventilatory equivalents method of determining $\dot{V}_{O_2}$AT involves plotting $\dot{V}_E/\dot{V}_{O_2}$
+and $\dot{V}_E/\dot{V}_{CO_2}$ against time, exercise load or $\dot{V}_{O_2}$.
+The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig.
+\ref{fig:cpet_9panel}F.
+
+\clearpage
 
@@
 Some of the factors are encompassed in the following formula for $\dot{V}_{O_2}$.
-\begin{equation} \label{eq:VO2=CaO2xCO}
+\begin{equation} \label{eq:oxygen_consumption}
 	\dot{V}_{O_2} = CaO_2 \times Cardiac\ Output
@@
 where $CaO_2$ is $O_2$ content per ml of blood and is defined by,
-\begin{equation} \label{eq:CaO2=Hbx1.34xSaO2}
+\begin{equation} \label{eq:CaO2}
 	CaO_2 = Haemoglobin \times 1.34 \times SaO_2
@@
 Oxygen pulse is defined as the oxygen uptake per heart beat.
-\begin{equation} \label{eq:O2Pulse=VO2/HR}
+\begin{equation} \label{eq:oxygen_pulse}
 	O_2Pulse = \frac{\dot{V}_{O_2}}{Heart\ rate}
@@
 \subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
-The change in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise.
+The changes in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise.
 Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise.
@@
 This change in direction is yet another method to confirm the anaerobic threshold.
-$\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as well as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.
+$\dot{V}_E/\dot{V}_{CO_2}$ eventually starts increasing as respiratory compensation of metabolic acidosis results in increased $\dot{V}_E$.
 
@@
 
-These reports in conjunction with the studies on preoperative inflammation, sepsis, SIRS and CARS emphasise the important role of peri;operative systemic inflammation as a causative factor in postoperative complications and the impact of the 'second-hit' of postoperative complications on long-term survival after curative cancer surgery.
+These reports in conjunction with the studies on preoperative inflammation, sepsis, SIRS and CARS emphasise the important role of perioperative systemic inflammation as a causative factor in postoperative complications and the impact of the 'second-hit' of postoperative complications on long-term survival after curative cancer surgery.
 

### 2015-08-04 Chapters/Introduction.tex
Introduction - OJ - immunity and others
- OJ and immunity
- Include CPET machine picture
- Update ventilatory equivalents method of measuring AT

@@
 \label{sec:cpx_method}
+
+\begin{figure}[htbp]
+	\centering
+	\includegraphics[width=\textwidth]{Figures/cpet-zan}
+	\caption{ZAN-600 Cycle Erogmeter and CPET suite. (www.nspirehealth.com)}
+	\label{fig:cpet-zan}
+\end{figure}
+
 CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.
@@
 \subsubsection{Ventilatory equivalents method}
-\todo{Ventilatory equivalents method}
 The ventilatory equivalents method of determining $\dot{V}_{O_2}$AT involves plotting $\dot{V}_E/\dot{V}_{O_2}$
@@
 \ref{fig:cpet_9panel}F.
-
-\clearpage
+Most respiratory physiologists including at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.
 
@@
 \subsection{Exercise Load}
-The most common form of cardiopulmonary exercise testing for clinical purposes involves a cycle ergometer with steadily increasing resistance delivered through electric braking allowing accurate measurement of work load in Watts.
+The most common form of cardiopulmonary exercise testing for clinical use involves a cycle ergometer with steadily increasing resistance delivered through electric braking.
+This allows accurate measurement of work load in Watts.
 The relationship between $\dot{V}_{O_2}$ and work rate is usually linear and the slope of this relationship is independent of sex, age or height.
@@
 
+Obstructive jaundice is associated with a wide range of deleterious effects on the immune system.
+The absence of bile in the gut increases the proliferation of Gram negative bacteria.
+This in association with increased bacterial translocation, increased gut mucosal permeability and reduced endotoxin elimination by Kuppfer cells can predispose the jaundiced patient to septic complications.
+Intestinal permeability is normalised within a few weeks of restoring biliary drainage.
+
+Increased serum bilirubin and bile salts directly affect lymphocyte function as well as the response to pro-inflammatory cytokines.
+Bile salts in higher than normal concentrations also induce apoptosis in hepatocytes.
+Biliary obstruction is also associated with significant alterations in hepatic macro and micro-circulation.
+This in turn can affect the phagocyte clearing function of the liver.
+Obstructive jaundice is also associated with abnormal cytokine homeostasis predisposing the patient to impaired wound healing, sepsis and increased risk of peri-operative mortality.
+For instance, concentrations of TNF-\alpha and IL-6 are significantly elevated in jaundiced patients while the concentrations of IL-1 and IL-2 are decreased.
+
+Biliary obstruction affects both cellular and humoral immunity.
+Endotoxins derived from the gut in bile duct ligated rats are thought to suppress cellular immunity.
+Moreover the clearance of immune-complexes may also be impaired in biliary obstruction.
+
 \subsection{Role of preoperative biliary drainage}

### 2015-08-05 Chapters/Introduction.tex
Introduction - Body Composition
- Completed obesity and surgical outcomes

@@
 Obstructive jaundice is also associated with abnormal cytokine homeostasis predisposing the patient to impaired wound healing, sepsis and increased risk of peri-operative mortality.
-For instance, concentrations of TNF-\alpha and IL-6 are significantly elevated in jaundiced patients while the concentrations of IL-1 and IL-2 are decreased.
+For instance, concentrations of TNF-$\alpha$ and IL-6 are significantly elevated in jaundiced patients while the concentrations of IL-1 and IL-2 are decreased.
 
@@
 \todo{Introduction - body composition}
+
+Obesity is an increasing problem in all developed nations with the proportion of obese patients undergoing surgery increasing every year.
+40\% of the adult population in the United Kingdom are expected to have a body mass index of over 30 kg/m$^2$ by 2025.
+Obesity, especially morbid obesity, is associated with a wide range of secondary disease states such as ischaemic heart disease, diabetes mellitus, hypertension and increased incidence of cancers.
+
+Obesity presents unique problems in surgical patients for several reasons.
+Surgery and anaesthesia in the morbidly obese patients can be technically challenging.
+The comorbidities associated with morbid obesity require evaluation and management in the perioperative period.
+Postoperative physiotherapy and mobilisation require particular attention in the morbidly obese patients.
+
+\subsection{Obesity and Surgical Outcomes}
+
+The impact of obesity on the rate of postoperative complications remains controversial.
+It has been a commonly held belief that obese patients had greater incidence of complications.
+Some of the adverse outcomes such as wound complications and respiratory dyfunction where conseidered to be the direct result of the increased body mass index while other complications where related to the associated comorbidities such as cardiovascular disease, obstructive sleep apnea or diabetes mellitus.
+
+Several studies have reported increased incidence of postoperative complications after cardiac, transplant, urologic and gynaecological surgery.
+\parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004} Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in pateints undergoing laparascopic sigmoid colectomy for cancer.
+\parencite{tsujinaka_visceral_2008}
+
+However, in a prospective, multi-institutional, risk-adjusted cohort study of 2258 patients including 699 patients who underwent pancreatectomy for cancer, Mullen and co-workers reported that obesity on its own was not a predictor of major postoperative complications or mortality.
+However, they noted that underweight patients had a five-fold increased risk of postoperative mortality.
+They also reported that obesity was an independent predictor of minor complications including wound infections.\parencite{mullen_impact_2008}
+
+Dindo and co-workers studied the impact of obesity on postoperative outcomes in a prospective study of 6336 patients undergoing elective general surgery.
+They reported that obesity was not an independent risk factor for morbidity or mortality.
+There was no difference in the median operating time, blood transfusions or lenght of hospital stay between obese and non-obese patients.
+The type and severity of complications were similar between obese and non-obese patients.
+\parencite{dindo_obesity_2003}
+
+In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
+This appears to confirm the existence of an 'obesity paradox' in surgical patients.
+\parencite{mullen_obesity_2009}

### 2015-08-05 Chapters/Introduction.tex
Introduction - Spelling and grammar

@@
 In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile ducts, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile ducts for cytological examination.
-The role of preoperative biliary drainage is discussed in more detail in section %\ref{sec:preoperative_biliary_drainage].
+The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.
 
@@
 Reconstruction was performed as follows: Either a pancreatico-jejunostomy was performed using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was performed using 3/0 Biosyn sutures placed in a similar manner.
-Hepaticojejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastrojeunonostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 PDS sutures in a 2-layers.
+Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers.
 One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.
@@
 Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion.
-All patients received proton pump inhibitors and octreotide.
-Patients were discharged to the general surgical ward as early as possible.
+All patients received proton pump inhibitors and Octreotide Patients were discharged to the general surgical ward as early as possible.
 
@@
 The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula.
-It can be a result of breakdown or poor healing at the pancreaticojejunostomy/pancreaticogastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
+It can be a result of breakdown or poor healing at the pancreatico-jejunostomy/pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
 It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.
@@
 
-Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodnenal artery.
+Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery.
 Post-pancreatectomy haemorrhage is often managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required.
@@
 
-The ESPAC-1 trial compared the use of chemotherapy with fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
+The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
 The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive demonstrating a clear survival benefit for adjuvant chemotherapy.
 However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting.
-The ESPAC-3 trial compared the efficacy of 5-FU and folinic acid with gemcitabine and reported that gemcitabine was not superior to 5-FU/folinic acid.
+The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid.
 
@@
 The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes.
-Therefore, accurate assessment of physiololgical fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions.
+Therefore, accurate assessment of physiological fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions.
 A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness.
@@
 \subsection{Origins and physiological basis of cardiopulmonary exercise testing}
-The measurement of aerobic fitness has been an important aspect of improving atheletic performance in competitive sports.
-Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes as well as offer customised training regimens aimed at improving performance at activities that place a signficant stress on the cardiovascular and respiratory systems.
+The measurement of aerobic fitness has been an important aspect of improving athletic performance in competitive sports.
+Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes as well as offer customised training regimens aimed at improving performance at activities that place a significant stress on the cardiovascular and respiratory systems.
 
@@
 This allowed breath-by-breath gas analysis thus allowing measurement of several respiratory parameters as listed in table [].
-12-lead electocardiogram was recorded at the same time.
+12-lead electrocardiogram was recorded at the same time.
 
 The test started with an initial 3-minute rest period to allow measurement of baseline parameters.
-This was followed by an incremental work-load test that involved the patient pedalling approximately at 60 revoultions per minute while the resistance to pedalling was gradually increased in preset increments.
+This was followed by an incremental work-load test that involved the patient pedalling approximately at 60 revolutions per minute while the resistance to pedalling was gradually increased in preset increments.
 The test was terminated when patients reached volitional fatigue (maximal exercise tolerance), significant ischaemic changes on ECG or for other safety reasons.
@@
 \subsection{End-tidal $O_2$ and $CO_2$, $P_{ET_{O_2}}$, $P{ET_{CO_2}}$}
-$PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to PaO2 and PaCO2 respectively.
+$PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to $P_aO_2$ and $P_aCO_2$ respectively.
 $P{ET_{CO_2}}$ is dependent on pulmonary gas-exchange which is in turn influenced by the right ventricular output, pulmonary blood flow and alveolar gas exchange.
@@
 In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake.
-However, a variety of respiratory disorders including obstructive disorders such as chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response.
+However, a variety of respiratory disorders including chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response.
 Morbidly obese patients may also have additional restrictions on respiratory response due to their body habitus.
@@
 Hightower and co-workers studied 32 patients over the age of 18 undergoing a variety of major elective abdominal surgery.
-In this heterogenous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic thresold were all independently associated with postoperative morbidity.
-While it is difficult to extrapolate results from this small heterogenous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients.\parencite{hightower_pilot_2010}
+In this heterogeneous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic threshold were all independently associated with postoperative morbidity.
+While it is difficult to extrapolate results from this small heterogeneous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients.\parencite{hightower_pilot_2010}
 
-Snowden and coworkers reported in a study of 171 patients of who 123 underwent surgery, that $\dot{V}_{O_2}AT<10.1$ ml/kg/min was associated with not only cardiovascular complications but also with other complications including pulmonary, renal, gastrointestinal, infective and haematological complications.
+Snowden and co-workers reported in a study of 171 patients of who 123 underwent surgery, that $\dot{V}_{O_2}AT<10.1$ ml/kg/min was associated with not only cardiovascular complications but also with other complications including pulmonary, renal, gastrointestinal, infective and haematological complications.
 One of the strengths of this study was the fact that the clinicians treating these patients were blinded to the results of preoperative cardiopulmonary exercise testing results thus avoiding management bias.
 They also included other cardiopulmonary exercise test parameters such as $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ at the anaerobic threshold in addition to $\dot{V}_{O_2}$AT in their analysis and showed that $\dot{V}_{O_2}$AT was more predictive of postoperative morbidity than POSSUM derived morbidity, cardiac risk scoring index or a validated activity questionnaire.
-However, this study also included a heterogenous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas.\parencite{snowden_submaximal_2010}
+However, this study also included a heterogeneous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas.\parencite{snowden_submaximal_2010}
 
@@
 \subsection{Systemic inflammation and long-term survival}
-Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable pancreatic cancer.\parencite{glen_evaluation_2006} Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated at over 10 mg/dl approximately a month after their operation.\parencite{jamieson_systemic_2005} Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR).
+Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable disease.\parencite{glen_evaluation_2006} Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated over 10 mg/dl approximately a month after their operation.\parencite{jamieson_systemic_2005} Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR).
 A selection of these studies are presented in Table %Create new table here for this.
+\todo{Table for systemic inflammation and long-term survival}
 
@@
 
-Preoperative systemic inflammation has been reported to be associated with infectious complications in patients undergoing potentially curative surgery for colorectal cancer.\parencite{moyes_preoperative_2009} In a study of 455 patients, Moyes and coworkers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery.
+Preoperative systemic inflammation has been reported to be associated with infectious complications in patients undergoing potentially curative surgery for colorectal cancer.\parencite{moyes_preoperative_2009} In a study of 455 patients, Moyes and co-workers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery.
 They postulated that several mechanisms may have a role including disregulation of cell-mediated immunity, impaired T-lymphocyte response, disorders in the complement pathway and possibly due to loss of lean tissue and protein as a consequence of systemic inflammation.
@@
 
-In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak.\parencite{welsch_c-reactive_2007} They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreaticojejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
+In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak.\parencite{welsch_c-reactive_2007} They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreatico-jejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
 
@@
 
-While CRP level between the third and fifth postoperative day has been reported to be most predictive of complications,  the complications themselves do not become clinically apparent until a later in the postperative course, ofter after the fifth postoperative period.
+While CRP level between the third and fifth postoperative day has been reported to be most predictive of complications,  the clinical signs do not become apparent until later in the postoperative course, often after the fifth day.
 This has led some authors to postulate that the elevated CRP levels may in fact be due to an abnormally modulated postoperative inflammatory response resulting in an initial exaggerated systemic inflammatory response syndrome (SIRS) followed by a compensatory anti-inflammatory response syndrome (CARS).
@@
 This form of 'immunoparalysis' was first described in patients after major trauma with tissue damage \parencite{abraham_effects_1985,bandyopadhyay_negative_2007} or after haemorrhage on its own without associated tissue trauma.
-\parencite{stephan_hemorrhage_1987} In a detailed review of the mechanisms underlying the compensatory anti-inflammatory response syndrome, Ward and coworkers describe SIRS and CARS to be mirror images suggesting that a disproportionately high SIRS is followed by a period of immunosuppression that leaves the patient prone to further complications.\parencite{ward_compensatory_2008}
+\parencite{stephan_hemorrhage_1987} In a detailed review of the mechanisms underlying the compensatory anti-inflammatory response syndrome, Ward and co-workers describe SIRS and CARS to be mirror images suggesting that a disproportionately high SIRS is followed by a period of immunosuppression that leaves the patient prone to further complications.\parencite{ward_compensatory_2008}
 
@@
 They reported that the initial inflammatory phase was followed by an immunosuppressive phase that started around the seventh postoperative day in patients undergoing oesophagectomy.
-However, patient who underwent underwent an open cholecystectomy did not experience this immunosuppressive phase, leading them to postulate that the degree of immunosuppression was directly proportional to the intial pro-inflammatory process.
+However, patient who underwent an open cholecystectomy did not experience this immunosuppressive phase, leading them to postulate that the degree of immunosuppression was directly proportional to the initial pro-inflammatory process.
 This in turn was related to the greater degree of surgical stress and tissue trauma that occurs with a trans-thoracic oesophagectomy.
@@
 
-The detrimental effects of jaundice on the heart and the circuilatory system has been recognised for over 150 years now.
-King and coworkers reported that the intravenous injection of bile caused bradycardia, hypotension and eventually death in dogs.\parencite{king_effect_1909} The concept of a \textit{'jaundiced heart'} was fist put forth by Green and coworkers in 1986.\parencite{green_jaundiced_1986} They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure.
-They reported that 'cholemia' was associated with impaired left ventricular function and blunted reponse to sympathomimmetic agents.
+The detrimental effects of jaundice on the heart and the circulatory system has been recognised for over 150 years now.
+King and co-workers reported that the intravenous injection of bile caused bradycardia, hypotension and eventually death in dogs.\parencite{king_effect_1909} The concept of a \textit{'jaundiced heart'} was fist put forth by Green and co-workers in 1986.\parencite{green_jaundiced_1986} They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure.
+They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimmetic agents.
 Similar findings have been reported by other authors in animal studies.
@@
 \subsection{Role of preoperative biliary drainage}
+\label{sec:preoperative_biliary_drainage}
 %fang_pre-operative_2012 - Cochrane review
@@
 
-Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated ith increased risk and in fact preoperative biliary drainage increased hospitalisation costs.
+Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated with greater risk and in fact preoperative biliary drainage increased hospitalisation costs.
 \parencite{snellen_influence_1985, bakkevold_morbidity_1993, pitt_does_1985} Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications.
@@
 However, the study does show that early surgery in the jaundiced patient with serum bilirubin $<$ 250 $\mu$mol/L is safe and not associated with increased morbidity or mortality as once feared.
-Mezhir and co-workers also reported a high (23\%) stent-related complication rate in their retrospective study of 340 patients where 96\% of 201 patients in the biliary drainage group had plastic stents placed.
+Mezhir and co-workers also reported a high (23\%) stent related complication rate in their retrospective study of 340 patients where 96\% of 201 patients in the biliary drainage group had plastic stents placed.
 However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage.
@@
 It has been a commonly held belief that obese patients had greater incidence of complications.
-Some of the adverse outcomes such as wound complications and respiratory dyfunction where conseidered to be the direct result of the increased body mass index while other complications where related to the associated comorbidities such as cardiovascular disease, obstructive sleep apnea or diabetes mellitus.
+Some of the adverse outcomes such as wound complications and respiratory dysfunction where considered to be the direct result of the increased body mass index while other complications where related to the associated comorbidities such as cardiovascular disease, obstructive sleep apnea or diabetes mellitus.
 

### 2015-08-08 Chapters/Introduction.tex
Introduction - Obesity and Survival

@@
 
-\section{Body Composition}
+\section{Body Composition, Obesity and Outcomes}
 \todo{Introduction - body composition}
 
+The Body Mass Index (BMI) is a measure of a person's weight relative to their height and corresponds well with their body composition, especially with the proportion that is body fat.
+It is the most widely accepted measure of body composition, especially of body fat in spite of criticisms that it may not apply to all populations.
+BMI was first devised by Adolphe Quetelet and was formerly known as the Quetelet index.\parencite{eknoyan_adolphe_2008} Quetelet was a Belgian sociologist and mathematician who played a major role in the application of statistical methods to the social sciences.
+In his attempts to describe the characteristics of 'normal man', Quetelet found that weight increased nearly as the square of the height except at times of growth spurts in the new born and during puberty.
+\parencite{quetelet_treatise_2013} The calculation of body mass index is simple and as follows:
+
+\begin{equation} \label{eq:bmi}
+Body\ Mass\ Index = \frac{Weight\ in\ Kilograms}{(Height\ in\ Metres)^2}
+\end{equation}
+
+The World Health Organisation uses thresholds to categorise adult populations based on body mass index.
+These thresholds are independent of age, gender and race.
+However, there has been increasing criticism of these thresholds and the World Health Organisation is undertaking a further review to assess the validity of these thresholds in different populations.
+For instance, there is increasing evidence in Asians, the risk of diabetes and cardiovascular disease is greater at BMI levels lower than the current WHO cut-off for being overweight.
+Nonetheless, the thresholds shown in Table \ref{table:bmi_who} are currently in universal use and have been used in this thesis.
+
+\begin{table}[h]
+	\centering
+	\caption{The WHO classification of Body Mass Index}
+	\label{table:bmi_who}
+	\begin{tabular}{ |l|l|}
+		\hline
+		BMI         & Category                  \\ \hline
+		Below 18.5  & Underweight               \\
+		18.5 – 24.9 & Normal range              \\
+		25.0 – 29.9 & Overweight or Pre-obesity \\
+		30.0 - 34.9 & Obesity class 1             \\
+		35.0 - 39.9 & Obesity class 2             \\
+		Above 40    & Obesity class 3             \\ \hline
+	\end{tabular}
+\end{table}
+
 Obesity is an increasing problem in all developed nations with the proportion of obese patients undergoing surgery increasing every year.
-40\% of the adult population in the United Kingdom are expected to have a body mass index of over 30 kg/m$^2$ by 2025.
+Among the European Member States, the proportion of adults  considered overweight or obese varied in 2008 between  37.0\% and 56.7\% for women and 51.0\% and 69.3\% for men.
+The United Kingdom had the highest proportion of obese women (23.9\%) and the second highest proportion of obese men (22.1\%).
+A greater proportion of men than women were obese and there was a marked increase in BMI with increasing age.
+However, as the proportion of overweight or obese decreased in women as their education level increased.
+
 Obesity, especially morbid obesity, is associated with a wide range of secondary disease states such as ischaemic heart disease, diabetes mellitus, hypertension and increased incidence of cancers.
@@
 
-\subsection{Obesity and Surgical Outcomes}
+\subsection{Obesity and Postoperative Complications}
 
@@
 Several studies have reported increased incidence of postoperative complications after cardiac, transplant, urologic and gynaecological surgery.
-\parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004} Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in pateints undergoing laparascopic sigmoid colectomy for cancer.
+\parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004} Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer.
 \parencite{tsujinaka_visceral_2008}
@@
 However, in a prospective, multi-institutional, risk-adjusted cohort study of 2258 patients including 699 patients who underwent pancreatectomy for cancer, Mullen and co-workers reported that obesity on its own was not a predictor of major postoperative complications or mortality.
+But,obesity was an independent predictor of minor complications including wound infections.
 However, they noted that underweight patients had a five-fold increased risk of postoperative mortality.
-They also reported that obesity was an independent predictor of minor complications including wound infections.\parencite{mullen_impact_2008}
+\parencite{mullen_impact_2008}
 
@@
 They reported that obesity was not an independent risk factor for morbidity or mortality.
-There was no difference in the median operating time, blood transfusions or lenght of hospital stay between obese and non-obese patients.
+There was no difference in the median operating time, blood transfusions or length of hospital stay between obese and non-obese patients.
 The type and severity of complications were similar between obese and non-obese patients.
-\parencite{dindo_obesity_2003}
-
-In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
+\parencite{dindo_obesity_2003} In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen and co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
 This appears to confirm the existence of an 'obesity paradox' in surgical patients.
 \parencite{mullen_obesity_2009}
+
+In a study of 356 consecutive patients undergoing pancreaticoduodenectomy at the Memorial Sloan-Kettering Cancer Center between 2000 and 2005, House and co-workers used preoperative cross-sectional imaging of the abdomen (computer tomography and/or magnetic resonance imaging) to measure a variety of indices of body composition including abdominal wall fat thickness, hip girdle fat thickness, retro-renal visceral fat and pancreatic duct diameter.
+\parencite{house_preoperative_2008} The incidence of wound infection was greater in obese patients with the risk increasing linearly with BMI.
+But, BMI was not associated with any other complications.
+They also reported that the thickness of retro-renal fat, a surrogate for visceral adipose tissue, was a significant risk factor for pancreatic fistula as well as wound infection.
+However, there was no correlation between complications and abdominal wall thickness or hip girdle fat.
+The importance of visceral fat in the incidence and severity of postoperative complications is further emphasised by the finding that excess visceral fat is associated with the the occurence of pancreatic fistula after radical total gastrectomy for gastric cancer.
+Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese pateints undergoing surgery compared to non-obese pateints.
+However, there was no difference in operative time or length of stay in hospital.
+\parencite{benns_impact_2009}
+
+On the contrary, other authors have reported that obesity is not associated with increased complications after pancreatic surgery.
+Khan and co-workers reported that while obesity was associated with increased incidence of intra-operative bleeding, there was no increase in complications.
+\parencite{khan_does_2010} Length of stay in hospital, overall morbidity and in-hospital mortality were not affected by BMI after pancreaticoduodenectomy \parencite{tsai_impact_2010} while neither BMI nor intra-abdominal visceral fat affected postoperative outcomes in another study of patients undergoing pancreaticoduodenectomy.
+\parencite{balentine_obesity_2011}
+
+In a recent meta-analysis of 17 studies, including some of those discussed above, Ramsey and co-workers reported that BMI was not associated with length of hospital stay, in-hospital mortality  or survival after pancreaticoduodenectomy for cancer.
+Obesity was associated with postoperative pancreatic fistula.
+However, the authors concluded that with aggressive perioperative care, postoperative morbidity in obese patients could be mitigated.
+They also noted several limitations to making meaningful comparisons between studies due to the lack of clear definitions for complications after pancreaticoduodenectomy.
+\parencite{ramsey_body_2011}
+
+%lin_risk_2004, lermite_risk_2007 - Coranary artery disease is associated with pancreatic fistula
+\subsection{Obesity and Cancer Survival}
+
+The effect of obesity and adipose tissue content on cancer-specific survival in patients with pancreatic cancer is not entirely clear.
+In a study involving 586 patients who underwent pancreaticoduodenectomy for pancreatic adenocarcinoma, Khan and co-workers reported that obesity was not associated with overall or disease-free survival.
+\parencite{khan_does_2010} These are similar to the findings reported by Benns and co-workers who found no difference in overall survival in 306 patients undergoing resections for pancreatic adenocarcinoma.
+\parencite{benns_impact_2009} Obesity during early adulthood is associated with increased risk of pancreatic cancer while obesity at an older age is associated with reduced overall survival in both resected and unresected patients.
+\parencite{li_body_2009} Morbidly obese patients (BMI $>$ 35) were found to be at 12-fold increased risk of lymph node metastases and associated decreased survival.
+\parencite{fleming_influence_2009}
+
+Tsai and co-workers reported that while obese patients had similar tumour characteristics to non-obese patients and comparable perioperative outomes, long-term survival was better in obese subjects.
+\parencite{tsai_impact_2010}  This study of 795 patients undergoing potentially curative surgery for pancreatic adenocarcinoma at the Johns Hopkins Medical Institutions in the United States is in contrast to other similar studies.
+The role of obesity on long-term survival after pancreaticoduodenectomy for pancreatic cancer is less clear with more recent studies suggesting that there was no link between body mass index and survival.
+\parencite{dandona_influence_2011, benns_impact_2009}

### 2015-08-09 Chapters/Introduction.tex
Major changes as below
- Thesis summary completed
- Many changes to thesis.cls and main.tex to format abstract page and university url, etc
- Update Intro-body comp - measuring body comp
- Zot.bin backed up and updated
- Conclusion chapter title renamed to Discussion
- It is 01:40 am - Have listened to Dr. Dre, Snoop Dogg and Eminem all day
- 2 shots of Jura and I am knackered yet sleepless after 2 strong coffees

@@
 12.4 months, p=0.002) and this was independent of other recognised risk factors such as tumour grade and lymph node status.
-\parencite{kamphues_postoperative_2011} Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007} and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head.
+\parencite{kamphues_postoperative_2012} Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007} and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head.
 \parencite{kang_detrimental_2009}
@@
 \todo{Introduction - body composition}
+\subsection{Body Mass Index}
 
@@
 In his attempts to describe the characteristics of 'normal man', Quetelet found that weight increased nearly as the square of the height except at times of growth spurts in the new born and during puberty.
-\parencite{quetelet_treatise_2013} The calculation of body mass index is simple and as follows:
+\parencite{quetelet_treatise_1842} The calculation of body mass index is simple and as follows:
 
@@
 	\label{table:bmi_who}
-	\begin{tabular}{ |l|l|}
+	\renewcommand{\arraystretch}{1.5} %Increases space between rows
+	\begin{tabular}{|l l|}
 		\hline
@@
 
-Several studies have reported increased incidence of postoperative complications after cardiac, transplant, urologic and gynaecological surgery.
+Several studies have reported increased incidence of postoperative complications in obese patients after cardiac, transplant, urologic and gynaecological surgery.
 \parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004} Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer.
@@
 However, there was no correlation between complications and abdominal wall thickness or hip girdle fat.
+
 The importance of visceral fat in the incidence and severity of postoperative complications is further emphasised by the finding that excess visceral fat is associated with the the occurence of pancreatic fistula after radical total gastrectomy for gastric cancer.
@@
 \parencite{dandona_influence_2011, benns_impact_2009}
+
+While sarcopenia or loss of skeletal muscle mass is associated with poor outcome in patients with pancreatic cancer, the association between adipose tissue (especially, visceral fat) and long-term outcomes is less clear.
+Sarcopenic-obesity where concurrent sarcopenia and high fat mass are present is associated with significantly worse survival when compared to sarcopenia or high adiposity on their own.
+\parencite{tan_sarcopenia_2009} It is clear that body mass index on its own is a poor predictor of outcome and body composition changes over time play a more important role in determining cancer survival.
+While obesity appears to have a protective role in a wide range of chronic disease states\parencite{kalantar-zadeh_risk_2007, amundson_obesity_2010, myers_obesity_2011} sarcopenic-obesity has a negative impact on survival.
+\parencite{tan_sarcopenia_2009}
+
+\subsection{Measuring Body Composition}
+While it is possible to estimate the amount of most tissue types, the ones that are of clinical interest and are related to surgical and cancer outcomes are adipose tissue and skeletal muscle.
+Numerous methods have been described for estimation of both absolute and proportional amount of these tissues with varying degrees of complexity and accuracy.
+Some of the simpler methods include measurement of subcutaneous fat using calipers or bioelectrical impedance, estimation of skeletal muscle from mid-arm circumference or creatinine height ratio.
+Such methods are valuable to public health and epidemiological studies of malnutrition.
+
+The gold-standard method for assessment of body composition is Dual Energy X-ray Absorptiometry (DEXA).\parencite{haarbo_validation_1991} While primarily designed for assessment of bone mineral density, DEXA scans have been adapted to measure fat, muscle and water content of subjects accurately and are used widely both in research and clinical settings.
+Other techniques such as whole body water or air displacement plethysmography are relatively cumbersome for routine clinical use.
+
+More recently, cross-sectional imaging including computed tomography (CT) and magnetic resonance imaging (MRI) have become widely available.
+Nearly all patients with gastrointestinal cancer undergo either CT or MRI or both before commencing treatment.
+Accurate measurements of subcutaneous adipose tissue, visceral adipose tissue and skeletal muscle can be made relatively easy either from the entire series of images or estimates can be made from measurements made at certain predetermined anatomical levels.
+Both specialised software or readily available generic image analysis tools have been used to make these measurements.
+
+The proportion of the total cross-sectional area that is made of adipose tissue or skeletal muscle at the level of the 3$^{rd}$  or the 4$^{th}$ lumbar vertebra is commonly used as a surrogate for total body measurements.
+Thickness of retro-renal visceral fat at the level of the renal vein has also been reported to be usable as a surrogate for total visceral adiposity.

### 2015-08-09 main.tex
Introduction - Obesity and Survival
Major changes as below
- Thesis summary completed
- Many changes to thesis.cls and main.tex to format abstract page and university url, etc
- Update Intro-body comp - measuring body comp
- Zot.bin backed up and updated
- Conclusion chapter title renamed to Discussion
- It is 01:40 am - Have listened to Dr. Dre, Snoop Dogg and Eminem all day
- 2 shots of Jura and I am knackered yet sleepless after 2 strong coffees

@@
 \abstract{\addtocontents{toc}{\vspace{1em}} % Add a gap in the Contents, for aesthetics
+Pancreaticoduodenectomy with or without adjuvant chemotherapy remains the only modality of possible cure in patients with cancer involving the head of the pancreas and the periampullary region.
+While mortality rates after pancreaticoduodenectomy have improved considerably over the course of the last century, morbidity remains high.
+Patient selection is of paramount importance in ensuring that major surgery is offered to individuals who will most benefit from a pancreaticoduodenectomy.
+Moreover, identifying preoperative risk factors provides potential targets for prehabilitation and optimisation of the patient's physiology before undertaking surgery.
+In addition to this, early identification of patients who are likely to develop postoperative complications allows for better allocation of critical care resources and more targeted treatment of the high risk patients.
 
-To be finalised...
+Cardiopulmonary exercise testing is becoming an increasingly popular tool in the preoperative risk assessment of the surgical patient.
+However, very little work has been done to investigate the role of cardiopulmonary exercise test in patients scheduled to undergo pancreaticoduodenectomy.
+The impact of jaundice, sytemic inflammation and other preoperative clinicopathological characteristics of the patient on their cardiopulmonary exercise physiology has not been studied in detail before in this cohort of patients.
+Moreover, the impact of these factors on long-term survival is also unknown.
+
+The overall aim of the thesis was to evaluate the relationship between cardiopulmonary exercise testing, perioperative clinico-pathological characteristics and postoperative outcomes as well as long-term survival in patients undergoing pancreaticoduodenectomy.
+
+\textbf{Chapter \ref{ch_intro}} reviews the existing literature on preoperative cardiopulmonary exercise testing, the impact of obstructive jaundice, perioperative systemic inflammation and the importance of body composition in determining outcomes in patients undergoing major surgery with particular reference to pancreatic surgery.
+
+\textbf{Chapter \ref{ch_cpet_outcomes}} reports on the role of cardiopulmonary exercise testing in predicting postoperative complications.
+The results demonstrate that patients with anaerobic threshold less than 10 mls/kg/min were more likely to develop a postoperative pancreatic fistula, stay longer in hospital and less likely to receive adjuvant therapy after pancreaticoduodenectomy.
+These results emphasise the importance of excellent aerobic fitness to improve postoperative outcomes and provide a potential target for future prehabilitation programmes.
+
+\textbf{Chapter \ref{ch_cpet_jaundice}} evaluates the relationship between cardiopulmonary exercise physiology and other clinicopathological characteristics of the patient.
+In a detailed analysis of cardiopulmonary exercise test parameters in jaundiced versus non-jaundiced patients, the results in this chapter demonstrate that obstructive jaundice does not impair cardiopulmonary exercise physiology.
+This further supports emerging evidence in contemporary literature that jaundiced patients can proceed directly to surgery without underoing preoperative biliary drainage.
+The results of this study also show an interesting inverse relationship between body mass index and anaerobic threshold which is analysed in more detail in Chapter 3.
+
+\textbf{Chapter \ref{ch_bodycomp}} examines the relationship between preoperative cardiopulmonary exercise physiology and body composition in depth.
+All parameters measured at cardiopulmonary exercise test are compared against body composition and body mass index.
+The results of this chapter report that the current method of reporting $\dot{V}_{O_2}$, both at peak exercise and anaerobic threshold, is biased against obese subjects and advises caution in the interpretation of cardiopulmonary exercise test results in patients with a high BMI.
+This is particularly important as current evidence in literature suggests that postoperative outcomes in obese subjects are comparable to non-obese subjects while cardiopulmonary exercise test results are also abnormally low in this very same cohort of patients.
+
+\textbf{Chapter \ref{ch_survival}} analyses the relationship between preoperative comorbidity including cardiopulmonary exercise test and long-term survival in patients with pancreatic ductal adenocarcinoma undergoing pancreaticoduodenectomy.
+The results of this chapter suggest that cardiopulmonary exercise testing parameters are not useful in predicting long-term survival.
+However, a composite score of preoperative physiology as measured by the POSSUM Physiology score was associated with survival.
+
+\textbf{Chapter \ref{ch_crp_comp}} evaluates the role of the early postoperative systemic inflammatory response in predicting complications after pancreaticoduodenectomy and aims to establish clinically relevant thresholds for C-Reactive Protein for the prediction of complications.
+The results of this chapter demonstrate that CRP levels as early as the second postoperative day are associated with complications.
+While post-operative CRP was useful in the prediction of infectious complications, this was only the case in patients who did not develop a post-operative pancreatic fistula.
+The predictive ability of inflammatory markers for infectious complications was blunted in patients with a pancreatic fistula.
+In these patients, outcome was largely determined by the severity of the pancreatic fistula.
+
+\textbf{Chapter \ref{ch_discussion}} summarises the findings of this thesis, their place in current literature and future directions.
+The contents of this thesis aim to add to the current body of knowledge regarding the complex pathophysiological alterations that occur in patients with lesions of the head of the pancreas with the associated obstructive jaundice, systemic inflammation and their impact on cardiopulmonary exercise physiology, postoperative complications and long-term outcomes.
+The results establish the value of cardiopulmonary exercise test as an objective measure of aerobic fitness while refuting a commonly held belief that obstructive jaundice significantly affected cardiovascular function.
+The current work paves the way for further study into evaluating the role of cardiopulmonary exercise test in prehabilitation regimens as well as the incorporation of preoperative $\dot{V}_{O_2}$AT and postoperative CRP into existing models for predicting complications after pancreaticoduodenectomy.
+These results will help improve patient selection and preparation for pancreaticoduodenectomy  and ultimately the short and long-term outcomes after this morbid yet necessary operation.
+
 }
@@
 \input{Chapters/Introduction} 		%1
-\input{Chapters/CPET_Outcomes} 	%2
-\input{Chapters/CPET_Jaundice} 	%3
-\input{Chapters/BodyComp}			%4
-\input{Chapters/CPET_Survival}		%5
-\input{Chapters/CRP_Complications}	%6
-\input{Chapters/Discussion}		%7
+%\input{Chapters/CPET_Outcomes} 	%2
+%\input{Chapters/CPET_Jaundice} 	%3
+%\input{Chapters/BodyComp}			%4
+%\input{Chapters/CPET_Survival}		%5
+%\input{Chapters/CRP_Complications}	%6
+%\input{Chapters/Discussion}		%7
 
@@
 
-\input{Appendices/AppendixExcelMacros}
-\input{Appendices/AppendixCPETRawData}
+%\input{Appendices/AppendixExcelMacros}
+%\input{Appendices/AppendixCPETRawData}
 %\input{Appendices/AppendixC}

### 2015-08-09 Chapters/Discussion.tex
Major changes as below
- Thesis summary completed
- Many changes to thesis.cls and main.tex to format abstract page and university url, etc
- Update Intro-body comp - measuring body comp
- Zot.bin backed up and updated
- Conclusion chapter title renamed to Discussion
- It is 01:40 am - Have listened to Dr. Dre, Snoop Dogg and Eminem all day
- 2 shots of Jura and I am knackered yet sleepless after 2 strong coffees
Discussion - 2/3 complete
- Minor changes in introduction
- Fixed effing error in zot.bib as a consequence of error in junejo abstract with Tex commands embedded - WHY???
- Fixed a fe

@@
 
-\chapter{Conclusion}
-\label{ch_conclusion}
+\chapter{Discussion}
+\label{ch_discussion}
 
-\lhead{Chapter \ref{ch_conclusion}. \emph{Conclusion}} % This is for the header on each page - perhaps a shortened title
+\lhead{Chapter \ref{ch_discussion}. \emph{Discussion}} % This is for the header on each page - perhaps a shortened title
 
@@
 
-This is the easy bit
+The overall aim of the thesis was to examine the relationships between cardiopulmonary exercise physiology, obstructive jaundice, systemic inflammation, body composition and short and long-term outcomes after major pancreatic surgery.
+
+A secondary aim of the thesis was to clarify the factors that affect cardiopulmonary exercise physiology and provide a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
+
+The results of the present work have demonstrated that cardiopulmonary exercise testing provides an objective, reproducible method of identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy where the underlying pathology is a malignancy.
+These results have since been replicated by other authors examining the role of cardiopulmonary exercise testing in pancreatic surgery.
+\parencite{ausania_effects_2012, ausania_double_2012, junejo_cardiopulmonary_2014}
+
+Equally important is the observation that preoperative obstructive jaundice did not impair aerobic capacity as measured by cardiopulmonary exercise testing.
+We are aware of only one other published study looking at cardiopulmonary exercise testing and peripheral oxygen extraction in jaundiced versus non-jaundiced patients undergoing pancreaticoduodenectomy.
+Junejo and co-workers reported that peripheral oxygen extraction was normal during exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014}
+Routine preoperative biliary drainage is no longer recommended before a pancreaticoduodenectomy.
+Several studies have shown that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
+\parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}
+The results reported in this thesis in conjunction with recently published clinical studies support the view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+
+We also observed that overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities.
+The analysis of the relationship between body composition and preoperative cardiopulmonary exercise testing has not been reported before in surgical patients.
+These results demonstrate that interpretation of cardiopulmonary exercise test parameters in patients with a high BMI must be made with caution, especially when such parameters have been corrected for the patient's total body weight.
+With the significant increase in the proportion of obese patients undergoing major surgery and the presence of sarcopenic-obesity in patients with pancreatic cancer, poor performance at cardiopulmonary exercise testing may not necessarily be related to cardiac or pulmonary function.
+The loss of skeletal mass and spurious correlations due to correction for total body weight may partly explain the low $\dot{V}_{O_2}$ in these patients.
+
+Recognition of these factors is important not only in patient selection and informed consent, but also in instituting the appropriate interventions as part of prehabilitation programmes.
+In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
+The authors report that major complications (31.3 vs.
+26.2\%) and mortality (4.1 vs.
+2.3\%) were greater in these patients.\parencite{tzeng_morbidity_2014}
+The authors recommend that surgeons identify these patients early, institute interventions to optimise their comorbidities and enrol these patients for prehabilitation.
+
+West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy.\parencite{west_effect_2015}
+Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery.\parencite{gillis_prehabilitation_2014}
+Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity measured objectively by cardiopulmonary exercise testing by upto 10\% which can potentially move approximately 30\% of high risk patients into the low risk category.\parencite{dunne_pmo-029_2012}
+
+Early recognition of patients at increased risk of postoperative complications facilitates prompt initiation of appropriate treatment and judicious allocation of critical care resources.
+Moreover, identification of patients who are unlikely to develop a major postoperative complication allows their progress on enhanced recovery pathways with the benefits of rapid recovery, reduced morbidity and early discharge.
+These patients are also more likely to receive adjuvant therapy without undue delay.
+Recognition of these low risk patients helps improve outcomes, reduces hospital costs and releases valuable resources for the care of high risk patients.
+The results reported in Chapter [ss] demonstrate the value of early post-operative C-reactive protein in identifying patients who are more likely to develop an infective complication.
+The results show that CRP is only useful in the absence of a postoperative pancreatic fistula.
+The combination of low drain amylase and low CRP on the third postoperative day identifies patients who do not have a pancreatic fistula and are less likely to develop an infective complication.

### 2015-08-09 main.tex
Discussion - 2/3 complete
- Minor changes in introduction
- Fixed effing error in zot.bib as a consequence of error in junejo abstract with Tex commands embedded - WHY???
- Fixed a few minor issues with main.tex

@@
 Cardiopulmonary exercise testing is becoming an increasingly popular tool in the preoperative risk assessment of the surgical patient.
-However, very little work has been done to investigate the role of cardiopulmonary exercise test in patients scheduled to undergo pancreaticoduodenectomy.
-The impact of jaundice, sytemic inflammation and other preoperative clinicopathological characteristics of the patient on their cardiopulmonary exercise physiology has not been studied in detail before in this cohort of patients.
-Moreover, the impact of these factors on long-term survival is also unknown.
+However, very little work has been done to investigate the role of cardiopulmonary exercise testing in predicting complications after pancreaticoduodenectomy.
+The impact of jaundice, systemic inflammation and other preoperative clinicopathological characteristics on cardiopulmonary exercise physiology has not been studied in detail before in this cohort of patients.
 
@@
 
-\textbf{Chapter \ref{ch_cpet_outcomes}} reports on the role of cardiopulmonary exercise testing in predicting postoperative complications.
-The results demonstrate that patients with anaerobic threshold less than 10 mls/kg/min were more likely to develop a postoperative pancreatic fistula, stay longer in hospital and less likely to receive adjuvant therapy after pancreaticoduodenectomy.
-These results emphasise the importance of excellent aerobic fitness to improve postoperative outcomes and provide a potential target for future prehabilitation programmes.
+\textbf{Chapter \ref{ch_cpet_outcomes}} reports on the role of cardiopulmonary exercise testing in predicting postoperative complications after pancreaticoduodenectomy.
+The results demonstrate that patients with $\dot{V}_{O_2}$AT less than 10 ml/kg/min are more likely to develop a postoperative pancreatic fistula, stay longer in hospital and less likely to receive adjuvant therapy.
+These results emphasise the importance of excellent aerobic fitness to recover from the operative stress of major surgery without significant morbidity.
+Cardiopulmonary exercise testing may prove useful in selecting patients for intensive prehabilitation programmes as well as for other optimisation measures to prepare them for major surgery.
 
@@
 In a detailed analysis of cardiopulmonary exercise test parameters in jaundiced versus non-jaundiced patients, the results in this chapter demonstrate that obstructive jaundice does not impair cardiopulmonary exercise physiology.
-This further supports emerging evidence in contemporary literature that jaundiced patients can proceed directly to surgery without underoing preoperative biliary drainage.
-The results of this study also show an interesting inverse relationship between body mass index and anaerobic threshold which is analysed in more detail in Chapter 3.
+This further supports emerging evidence in contemporary literature that jaundiced patients can proceed directly to surgery without preoperative biliary drainage.
+The results of this study also show an interesting inverse relationship between body mass index and anaerobic threshold which is analysed in more detail in Chapter \ref{ch_bodycomp}.
 
@@
 The results of this chapter suggest that cardiopulmonary exercise testing parameters are not useful in predicting long-term survival.
-However, a composite score of preoperative physiology as measured by the POSSUM Physiology score was associated with survival.
+However, a composite score of preoperative comorbidity as measured by the POSSUM Physiology score appears to predict long-term survival.
 
@@
 
-}
 \clearpage
@@
 \input{Chapters/Introduction} 		%1
-%\input{Chapters/CPET_Outcomes} 	%2
-%\input{Chapters/CPET_Jaundice} 	%3
-%\input{Chapters/BodyComp}			%4
-%\input{Chapters/CPET_Survival}		%5
-%\input{Chapters/CRP_Complications}	%6
-%\input{Chapters/Discussion}		%7
+\input{Chapters/CPET_Outcomes} 	%2
+\input{Chapters/CPET_Jaundice} 	%3
+\input{Chapters/BodyComp}			%4
+\input{Chapters/CPET_Survival}		%5
+\input{Chapters/CRP_Complications}	%6
+\input{Chapters/Discussion}		%7
 
@@
 
-%\input{Appendices/AppendixExcelMacros}
-%\input{Appendices/AppendixCPETRawData}
+\input{Appendices/AppendixExcelMacros}
+\input{Appendices/AppendixCPETRawData}
 %\input{Appendices/AppendixC}
@@
 \lhead{\emph{Bibliography}} % Change the page header to say "Bibliography"
-
 %\bibliographystyle{unsrtnat} % Use the "unsrtnat" BibTeX style for formatting the Bibliography
@@
 
+\clearpage
 \listoftodos

### 2015-08-10 Chapters/Discussion.tex
Discussion - Minor changes

@@
 A secondary aim of the thesis was to clarify the factors that affect cardiopulmonary exercise physiology and provide a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
-
+%################################################################################
+% Cardiopulmonary exercise testing
+%################################################################################
 The results of the present work have demonstrated that cardiopulmonary exercise testing provides an objective, reproducible method of identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy where the underlying pathology is a malignancy.
@@
 
+A low $\dot{V}_{O_2}$AT informs the patient, anaesthetist and surgeon of an increased perioperative risk.
+This allows everyone to make informed decisions on appropriate treatment strategies.
+For instance, a patient with excellent aerobic fitness may be able to proceed through surgery on an enhanced recovery pathway with early surgery and potentially early discharge from hospital.
+On the other hand, a patient with poor aerobic fitness as demonstrated by cardiopulmonary exercise testing will be better suited for a prehabilitation programme that involves measures such as nutritional supplementation, cessation of smoking, moderate tailored exercise, other cardiorespiratory optimisation measures including change in cardiac medication or bronchodilators.
+Some patients with extremely poor aerobic fitness in the presence of other known significant medical comorbidities may not be suitable for surgery.
+Cardiopulmonary exercise testing provides an objective, evidence-based measure of risk to rationalise treatment decisions in such high risk patients.
+%################################################################################
+% Obstructive Jaundice
+%################################################################################
 Equally important is the observation that preoperative obstructive jaundice did not impair aerobic capacity as measured by cardiopulmonary exercise testing.
@@
 The results reported in this thesis in conjunction with recently published clinical studies support the view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
-
+%################################################################################
+%Body composition
+%################################################################################
 We also observed that overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities.
@@
 
+%################################################################################
+%Prehabilitation
+%################################################################################
 Recognition of these factors is important not only in patient selection and informed consent, but also in instituting the appropriate interventions as part of prehabilitation programmes.

### 2015-08-12 Chapters/Introduction.tex
Discussion - 2/3 complete
- Minor changes in introduction
- Fixed effing error in zot.bib as a consequence of error in junejo abstract with Tex commands embedded - WHY???
- Fixed a few minor issues with main.tex
Introduction - revision to section 1.3
- Spelling, grammar, style corrections

@@
 \section{Pancreatic Neoplasia}
+%This is the only paper I need for this entire section - pancreatic_section_of_the_british_society_of_gastroenterology_pancreatic_society_of_great_britain_and_ireland_association_of_upper_gastrointestinal_surgeons_of_great_britain_and_ireland_royal_college_of_pathologists_special_interest_group_for_gastro-i_guidelines_2005
 
@@
 However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers.
-These include the distal common bile duct that can give rise to cholangiocarcinoma, the duodenum that can give rise to duodenal adenocarcinoma and the ampulla that can give rise to ampullary adenocarcinoma.
-The endocrine portion of the pancreas can give rise to a variety of tumours that are collectively called neuroendocrine tumours (NET).
-The milieu of tumours is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours.
+These include the distal common bile duct (cholangiocarcinoma), the duodenum (duodenal adenocarcinoma) and the ampulla (ampullary adenocarcinoma).
+Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas.
+This milieu is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours.
 Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.
 
-Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 21\% surviving beyond the first year and 3\% surviving beyond 5 years.\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}
+Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death.
+Only 21\% survive beyond the first year and 5-year survival is 3\%.
+\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}
 
-In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} remains the primary modality of cure.
+In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} offers the only chance of a cure.
 However, major pancreatic surgery places significant physiological stresses on multiple organ systems.
-The ability of the cardiac and respiratory systems, in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.
+The ability of the cardiac and respiratory systems in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.
 
@@
 
-The anatomical location of the pancreas, deep within the retroperitoneum surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours.
+The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours.
 Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis.
@@
 Other non-specific symptoms include weight loss, early satiety, vomiting, fatigue and pain in the epigastrium or the back.
+Occasionally, the first presentation of a pancreatic neoplasm may be in the form of an episode of acute pancreatitis.
 
@@
 
-Aside from a thorough history, clinical examination, blood tests including liver function tests, diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase).
-CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced MRI Liver in specific cases.
-CECT-pancreas is also useful for assessing local resectability with regards to vascular involvement.
-Endoscopic ultrasound (EUS) is also useful in assessing vascular involvement and for obtaining tissue samples for histological examination.
-In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile ducts, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile ducts for cytological examination.
+Initial assessment includes a thorough history, clinical examination and blood tests including full blood count, renal function tests, liver function tests and tumour markers, especially CA 19-9.
+However, definitive diagnosis requires cross-sectional imaging in the form of a contrast-enhanced computerised tomogram (CECT) of the abdomen using a pancreas-specific protocol (a modified form of the portal-venous phase).
+CECT of the pancreas when combined with CT Thorax also provides accurate information on staging of the disease with regards to metastasis and this can be supplemented by further imaging such as Positron Emission Tomography (PET-CT) or contrast-enhanced Magnetic Resonance Imaging (MRI) of the liver in specific cases.
+CECT-pancreas is also useful for assessing local resectability with accurate information regarding vascular involvement.
+Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination.
+In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination.
 The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.
@@
 In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival.
-Assessing the resectability is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
+Assessing the resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
 
@@
 Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description.
-The procedure was performed as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}.
+Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}.
 The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}
@@
 \subsubsection{Resectability criteria}
-Resectable pancreatic cancer is defined as a tumour that
-- does not involve the coeliac axis or the superior mesenteric artery
-- and is not associated with distant metastatic disease
+Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease.
+Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.\parencite{fuhrman_rationale_1996} The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation\parencite{bower_irreversible_2011}  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease.
+Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.
 
-Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.
-The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced disease and in some patients with metastatic disease.
-Early reports suggest that some of these patients can undergo potentially curative surgery after such neo-adjuvant treatment.
-
+%10/08/2015 - Review completete to here
 \subsubsection{Patient factors}
@@
 
-After a comprehensive preoperative work-up including both assessments of the tumour as well as patient fitness, informed consent was obtained.
+A comprehensive preoperative evaluation was performed including staging of the disease for local resectability and absence of metatstatic disease, assessment of patient fitness and anaesthetic review.
+All patients with suspected cancer were discussed at the pancreato-biliary multidisciplinary team meeting.
+A fully informed consent was obtained after discussion regarding treatment options, the details of a pancreaticoduodenectomy, expected perioperative course, potential risks and follow-up care.
+
 Patients received thrombo-prophylaxis on the night before surgery which was continued until discharge from hospital.
@@
 Epidural analgesia was used routinely in patients during the early part of the study period while all patients in the later half of the study period received spinal diamorphine.
-Antibiotic prophylaxis is administered at induction.
+Prophylactic antibiotics were administered at induction.
 While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre.
@@
 Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion.
-All patients received proton pump inhibitors and Octreotide Patients were discharged to the general surgical ward as early as possible.
+All patients received proton pump inhibitors and Octreotide.
+Patients were discharged to the general surgical ward as early as possible.
+Amylase content in the drain fluid was measured on the third post-opertive day to aid in the diagnosis of post-operative pancreatic fistula.
+Surgical drains and naso-gastric/naso-jeujunal tubes were removed at surgeon's discretion.
+Patients were discharged when medically fit.
+
+Histology of the removed specimen was discussed at the multi-disciplinary team meeting in cancer patients and suitable patients were referred for adjuvant chemotherapy.
 
@@
 Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability.
-The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.\parencite{yeo_six_1997,deoliveira_assessment_2006,bassi_postoperative_2005,winter_biochemical_2007,pratt_risk_2008}
-The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula.
-It can be a result of breakdown or poor healing at the pancreatico-jejunostomy/pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
+The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.
+\parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008} The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula.
+It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
 It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.
@@
 Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying.
-They may require further intervention (either radiological or operative) and may result in prolonged critical care support.
+They may require further intervention (either radiological or operative) or prolonged critical care support.
 
@@
 Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery.
-Post-pancreatectomy haemorrhage is often managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required.
+Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required.
 The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.
@@
 A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc.
-The Clavien-Dindo system grades the severity of complications based on the impact the complication has on the management of the patient and has been validated on large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} This is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and has been used to grade complications in this thesis.
+The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and was used to grade complications in this thesis.
 Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.
@@
 
-While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care.
-The European Study Group for Pancreatic Cancer (ESPAC) have conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect.
+While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma.
+The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect.
 \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}
 
-The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
-The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive demonstrating a clear survival benefit for adjuvant chemotherapy.
+The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
+\parencite{neoptolemos_randomized_2004} The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy.
 However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting.
 The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid.
+\parencite{neoptolemos_adjuvant_2010}
 
-However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma end up either not receiving or not completing adjuvant chemotherapy.
-\parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbdity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}
+However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy.
+\parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}
 
@@
 Both specialised software or readily available generic image analysis tools have been used to make these measurements.
-
 The proportion of the total cross-sectional area that is made of adipose tissue or skeletal muscle at the level of the 3$^{rd}$  or the 4$^{th}$ lumbar vertebra is commonly used as a surrogate for total body measurements.
 Thickness of retro-renal visceral fat at the level of the renal vein has also been reported to be usable as a surrogate for total visceral adiposity.
+Studies have also reported on the association between intra-pancreatic fat content and postoperative complications.

### 2015-08-16 Chapters/CPET_Jaundice.tex
CPET-Jaundice - Major changes
- Reformat all tables
- Change all values to median (IQR)
- Change captions of all tables to fit with Donny style
- Include new binary log regression for vo2peak similar to vo2at
- Change all < and vo2 to correct tex format
- Next commit should be changes within chapter to reflect the changes in tables and more

@@
 
-\lhead{Chapter \ref{ch_cpet_jaundice}. \emph{CPET and Obstructive Jaundice}} % This is for the header on each page - perhaps a shortened title
+\lhead{Chapter \ref{ch_cpet_jaundice}. \emph{Obstructive Jaundice}} % This is for the header on each page - perhaps a shortened title
 
@@
 \section{Patients and Methods}
-Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative workup at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
+Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
 Established criteria for resectability in patients with malignant disease were used as outlined in previous published work.
@@
 Serum bilirubin levels were measured in all patients on the day before surgery.
-Obstructive jaundice (OJ) was defined as bilirubin levels greater than 35 micromol/litre and severe obstructive jaundice (sOJ) was defined as bilirubin levels greater than 250 micromol/litre.
+Obstructive jaundice was defined as bilirubin levels greater than 35 micromol/litre and severe obstructive jaundice was defined as bilirubin levels greater than 250 micromol/litre.
 This threshold was selected because the DROP trial did not investigate patients with bilirubin levels greater than 250 micromol/litre and this study aimed to evaluate preoperative pathophysiology in this particular group.
-
-Data on PBD (PBD) was also recorded. Serum bilirubin levels before and after biliary stenting were recorded [I will expand this section when I get the updated stent data]
+\todo{A word about stents}
 
@@
 The test was stopped when patients achieved their maximum exercise tolerance, when significant ischaemic changes occurred on ECG or for other safety reasons.
-Peak oxygen consumption achieved at this stage was defined as VO$_2$Peak.
-The VO$_2$AT was calculated using the V-slope\parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents\parencite{society_ats/accp_2003} methods.
-VO$_2$AT less than 10 ml/kg/min was considered to be low based on previous work by us\parencite{chandrabalan_pre-operative_2013} as well as Ausania and co-workers\parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients.
-Oxygen consumption at peak exercise (VO$_2$Peak) was dichotomised using a cut-off of 16 ml/kg/min.
+Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak.
+The $\dot{V}_{O_2}$AT was calculated using the V-slope\parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents\parencite{society_ats/accp_2003} methods.
+$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us\parencite{chandrabalan_pre-operative_2013} as well as Ausania and co-workers\parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients.
+Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min.
 Detailed description of cardiopulmonary exercise testing as well as the physiological parameters described in this study are published elsewhere.\parencite{balady_clinicians_2010}
@@
 Non-parametric tests were used to analyse the association between categorical and continuous variables while Chi-square tests were used to analyse the association between categorical variables.
-Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and VO$_2$AT / VO$_2$Peak.
+Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and $\dot{V}_{O_2}$AT / $\dot{V}_{O_2}$Peak.
 SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.
@@
 Obstructive jaundice (serum bilirubin 35 – 250) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin > 250) was present in 19 (14\%) patients.
-The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_table1}.
-A larger proportion of jaundiced patients were females compared to the non-jaundiced cohort (p<0.05) and smokers (p<0.05).
-Patients with jaundice were more likely to have an elevated POSSUM Physiology Score (p<0.005).
-Patients with cancer were more likely to be jaundiced (p<0.001).
+The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_patient}.
+A larger proportion of jaundiced patients were females compared to the non-jaundiced cohort (p$<$0.05) and smokers (p$<$0.05).
+Patients with jaundice were more likely to have an elevated POSSUM Physiology Score (p$<$0.005).
+Patients with cancer were more likely to be jaundiced (p$<$0.001).
 However, there was no statistically significant difference in age, BMI, cardiovascular comorbidity, or respiratory comorbidity between the non-jaundiced and jaundiced patients.
 
-\input{Tables/c3_t1}
+\input{Tables/cpet_oj_patient}
 
-The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_table2}.
+The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_bloods}.
 While obstructive jaundice was statistically associated with multiple haematological and biochemical abnormalities, most of these did not appear to be of clinically significance.
 As expected, obstructive jaundice was associated with markedly elevated liver enzymes with severity of derangement associated with severity of jaundice.
-Obstructive jaundice and sOJ were associated with increasing CRP levels (p<0.001) and decreasing serum albumin levels (p<0.001).
+Obstructive jaundice and sOJ were associated with increasing CRP levels (p$<$0.001) and decreasing serum albumin levels (p$<$0.001).
 Obstructive jaundice was not associated with deranged renal function with both urea and creatinine remaining similar across all cohorts (p=0.09 and p=0.22 respectively).
 
-\input{Tables/c3_t2}
+\input{Tables/cpet_oj_bloods}
 
-There was no association between obstructive jaundice and preoperative pulmonary function tests (Table \ref{table:cpet_oj_table3}).
+There was no association between obstructive jaundice and preoperative pulmonary function tests (Table \ref{table:cpet_oj_pft}).
 
-\input{Tables/c3_t3}
+\input{Tables/cpet_oj_pft}
 
 \subsubsection{Univariate analysis of obstructive jaundice versus CPET}
-The relationship between obstructive jaundice and multiple physiological parameters measured at cardiopulmonary exercise testing is shown in Table \ref{table:cpet_oj_table4}.
-There was an inverse relationship between oxygen consumption at the anaerobic threshold (VO$_2$AT) and increasing severity of jaundice (p<0.05).
+The relationship between obstructive jaundice and multiple physiological parameters measured at cardiopulmonary exercise testing is shown in Table \ref{table:cpet_oj_anaerobic}.
+There was an inverse relationship between oxygen consumption at the anaerobic threshold ($\dot{V}_{O_2}$AT) and increasing severity of jaundice (p$<$0.05).
 However, no such linear relationship was noted between any of the other parameters measured both at anaerobic threshold and at peak exercise in spite of apparent statistically significant associations.
 
-\input{Tables/c3_t4}
+\input{Tables/cpet_oj_anaerobic}
+\input{Tables/cpet_oj_peak}
 
-\subsubsection{Association between preoperative clinico-pathological factors and VO$_2$AT}
-On multivariate analysis female sex (HR 3.75 CI 1.57-8.95 p<0.005), high BMI (HR 3.65 CI 1.61-8.26 p<0.005), presence of cancer (HR 4.02 CI 1.33-12.16 p<0.05) and raised CRP (HR 2.98 CI 1.29-6.86 p<0.05) were independently associated with low VO$_2$AT (<10mls/kg/min).
-However, jaundice was not associated with low VO$_2$AT.
-These results are shown in Table \ref{table:cpet_oj_table5}
+\subsubsection{Association between preoperative clinico-pathological factors and $\dot{V}_{O_2}$AT}
+On multivariate analysis female sex (HR 3.75 CI 1.57-8.95 p$<$0.005), high BMI (HR 3.65 CI 1.61-8.26 p$<$0.005), presence of cancer (HR 4.02 CI 1.33-12.16 p$<$0.05) and raised CRP (HR 2.98 CI 1.29-6.86 p$<$0.05) were independently associated with low $\dot{V}_{O_2}$AT ($<$10mls/kg/min).
+However, jaundice was not associated with low $\dot{V}_{O_2}$AT.
+These results are shown in Table \ref{table:cpet_oj_at_regression}
 
-\input{Tables/c3_t5}
+\input{Tables/cpet_oj_at_regression}
+\input{Tables/cpet_oj_peak_regression}
 
 \subsubsection{Scatter-plot analysis}
-Scatter-plot analysis comparing serum bilirubin and VO$_2$AT as continuous variables is depicted in Figure 1.
+Scatter-plot analysis comparing serum bilirubin and $\dot{V}_{O_2}$AT as continuous variables is depicted in Figure 1.
 This shows that the relationship between serum bilirubin and AT is weak with an r2 value of only 0.04 (I will have to confirm this but it is not more than 0.1).
@@
 While the aforementioned studies have undermined the role of PBD in jaundiced patients undergoing pancreaticoduodenectomy, the results of the present study show for the first time that the premise for performing PBD, namely the adverse effect of jaundice on cardiopulmonary physiology may itself be flawed in patients undergoing pancreaticoduodenectomy.
-In our study, obstructive jaundice including severe obstructive jaundice did not affect cardiopulmonary exercise capacity as measured by VO$_2$AT or the peak oxygen consumption.
+In our study, obstructive jaundice including severe obstructive jaundice did not affect cardiopulmonary exercise capacity as measured by $\dot{V}_{O_2}$AT or the peak oxygen consumption.
 These findings taken together with previously published findings of adverse effects of PBD further support the fact that major surgery may be safe in jaundiced patients without subjecting them to preoperative biliary drainage.
-The basis of the relationship between low VO$_2$AT and raised BMI is not clear.
+The basis of the relationship between low $\dot{V}_{O_2}$AT and raised BMI is not clear.
 
-However, such an association has been previously reported.\parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate VO$_2$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
+However, such an association has been previously reported.\parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate $\dot{V}_{O_2}$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
 Other authors have suggested that different thresholds for CPET parameters may have to be considered in obese patients to improve risk-prediction.\parencite{donnelly_criteria_1990,hulens_exercise_2001} Cardiopulmonary exercise testing measures oxygen delivery to skeletal muscle.
@@
 However, this does not account for the disproportionately higher amount of adipose tissue in overweight/obese patients resulting in a spuriously low AT (in mls/kg/min).
-The present study found no association between cardiorespiratory comorbidity and VO$_2$AT.
-Low VO$_2$AT in female patients and overweight/obese patients should be interpreted with caution as this may not be due to true poor aerobic capacity.
+The present study found no association between cardiorespiratory comorbidity and $\dot{V}_{O_2}$AT.
+Low $\dot{V}_{O_2}$AT in female patients and overweight/obese patients should be interpreted with caution as this may not be due to true poor aerobic capacity.
 

### 2015-08-16 Tables/cpet_oj_pft.tex
CPET-Jaundice - Major changes
- Reformat all tables
- Change all values to median (IQR)
- Change captions of all tables to fit with Donny style
- Include new binary log regression for vo2peak similar to vo2at
- Change all < and vo2 to correct tex format
- Next commit should be changes within chapter to reflect the changes in tables and more

%OJ - PFT
\begin{sidewaystable}[p]
	\caption{The relationship  between obstructive jaundice and pulmonary function tests in patients undergoing pancreaticoduodenectomy.}
	\label{table:cpet_oj_pft}
	\centering
	\renewcommand{\arraystretch}{1.4} %Increases space between rows
	\setlength{\tabcolsep}{9pt} %sets the space between columns

	\begin{tabular}{|l| c c c c|c|}
		\hline
		                    &              \multicolumn{4}{c|}{Preoperative Serum Bilirubin}              &  \\
		                    & $\leq$ 17        & 18-35            & 35-250            & $>$ 250          & \textit{p} \\ \hline
		FVC                 & 4.09 (3.49-4.69) & 3.76 (3.38-4.59) & 3.76 (3.16-4.05)  & 3.35 (2.85-4.38) & 0.092      \\
		FEV1                & 2.95 (2.39-3.51) & 2.90 (2.12-3.34) & 2.68 (2.37-3.07)  & 2.72 (2.2-3.28)  & 0.556      \\
		PREDICTED FEV1 (\%) & 105.0 (91-116)   & 98.50 (87-114)   & 103.0 (95-111.5)  & 101.0 (94-116)   & 0.761      \\
		FEV1/FVC            & 72.0 (65-77)     & 73.0 (65-78)     & 75.50 (70.5-79.5) & 78.0 (70-82)     & 0.115      \\
		PREDICTED FEV1/FVC  & 94.0 (87-102)    & 96.0 (86-100)    & 99.0 (93-103)     & 102.0 (88-108)   & 0.107      \\ \hline
		\multicolumn{6}{l}{Values are median (inter-quartile range); \textit{p} - Kruskal-Wallis test}
	\end{tabular}
\end{sidewaystable}

### 2015-08-16 Chapters/Introduction.tex
Introduction - Minor changes
- Minor changes to introduction

@@
 
-\subsection{Patient selection}
-\subsubsection{Resectability criteria}
+\subsection{Resectability criteria}
+\label{sec:resectability_criteria}
 Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease.
@@
 %10/08/2015 - Review completete to here
-\subsubsection{Patient factors}
 
@@
 It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
+
 It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.
+
 Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}.
@@
 \todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
+Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness.
+Major pancreatic surgery is associated with significant morbidity.
+Subjecting patients with poor physiological reserve to a pancreaticoduodenectomy is fraught with risk.
+It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in longevity as compared to palliation.
+Patient factors that affect postoperative recovery and long-term survival are therefore important in determining suitability for surgery.
+
 \subsection{Comorbidity}
@@
 
-The presence of comorbid conditions is associated with adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010} and often limits therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment.
-\parencite{sandroussi_sociodemographics_2010}
+The presence of comorbid conditions is associated with increased incidence of adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010}.
+It can also limit the therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment.
+\parencite{sandroussi_sociodemographics_2010} Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.
+\parencite{schneider_patient_2012}
 
-Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.\parencite{schneider_patient_2012} DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy\parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery.
+DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy \parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery.
 \parencite{shah_limitations_2012}
@@
 
-Objective measurement of oxygen delivery at the tissue level at times of physiological stress allows for identification of patients who may struggle during the perioperative phase.
+Conventional tests of cardiac and respiratory function assess either performance at rest or only one component of the cardiorespiratory system.
+These tests include electrocardiography, echocardiography, exercise tolerance testing, myocardial perfusion scan, pulmonary function tests including spirometry.
+However, none of these tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand.
+Patients with subclinical cardiopulmonary dysfunction or limitation cannot be identified on these tests alone and it is this group of patients who are most likely to be at greater risk of complications and who will most benefit from targeted prehabilitation.
+Cardiopulmonary exercise testing overcomes the limitations of the conventional static tests by providing a global assessment of the patient's oxygen delivery mechanisms at times of increased physiological stress and demand.
 
-Several tests have been used for preoperative assessment of cardiac function. These include
-- electrocardiography
-- echocardiography
-- exercise tolerance testing
-- myocardial perfusion scans
-
-Tests of respiratory function that are commonly performed in selected patients undergoing major surgery include
-- pulmonary function tests including forced expiratory volume and forced vital capacity
-- spirometry
-
-However, neither of the above cardiac or respiratory function tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand.
+One of the earliest attempts to develop a dynamic test in surgical patients was reported by Starr and co-workers in 1945 who tried to devise a test that could assess respiratory function and oxygen utilisation in surgical patients before, immediately after and a week after surgery.
+They reported on the difficulties in obtaining reproducible results but suggested that a slow return to baseline of heart rate, respiratory rate and oxygen utilisation after surgery may indicate patients who are slow to recover.
+\parencite{starr_convalescence_1945}
 
 \section{Cardiopulmonary Exercise Testing}
+%Not sure I need these two sections - delete once done vacillating
 \todo{A brief history of CPET in general}
@@
 The measurement of aerobic fitness has been an important aspect of improving athletic performance in competitive sports.
-Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes as well as offer customised training regimens aimed at improving performance at activities that place a significant stress on the cardiovascular and respiratory systems.
+Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes.
+It has also been used to design customised training regimens aimed at improving performance at activities that place a significant stress on the cardiovascular and respiratory systems by allowing the optimisation of the individual components of the oxygen delivery system.
 

### 2015-08-17 Chapters/CPET_Jaundice.tex
CPET Jaundice - Major update
- Next few commits will be following Donny chat this morning

@@
 \section{Results}
-One-hundred and thirty eight patients had undergone pancreaticoduodenectomy (n=138), with preoperative cardiopulmonary exercise testing during the study period.
-Over half the patients were male (n=93, 67\%).
+One-hundred and thirty eight patients underwent pancreaticoduodenectomy with preoperative cardiopulmonary exercise testing during the study period.
+Over half of the patients were male (n=93, 67\%).
 Approximately half the number of patients were over the age of 65 (n=68, 49\%) and overweight or obese (n=69, 50\%).
@@
 The POSSUM Physiology Score was greater than 14 in 61 patients (44\%).
-Obstructive jaundice (serum bilirubin 35 – 250) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin > 250) was present in 19 (14\%) patients.
+Obstructive jaundice (serum bilirubin 35 – 250) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin $>$ 250) was present in 19 (14\%) patients.
+
 The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_patient}.
-A larger proportion of jaundiced patients were females compared to the non-jaundiced cohort (p$<$0.05) and smokers (p$<$0.05).
-Patients with jaundice were more likely to have an elevated POSSUM Physiology Score (p$<$0.005).
-Patients with cancer were more likely to be jaundiced (p$<$0.001).
-However, there was no statistically significant difference in age, BMI, cardiovascular comorbidity, or respiratory comorbidity between the non-jaundiced and jaundiced patients.
+A greater proportion of jaundiced patients were females compared to the non-jaundiced cohort (p=0.028) and smokers (p=0.038).
+Elevated POSSUM Physiology Score (p=0.004) and malignancy (p$<$0.001) were significantly associated with the presence of jaundice.
+However, there was no statistically significant difference in age, body mass index, cardiovascular comorbidity, respiratory comorbidity or preoperative biliary drainage between the non-jaundiced and jaundiced patients.
 
@@
 The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_bloods}.
-While obstructive jaundice was statistically associated with multiple haematological and biochemical abnormalities, most of these did not appear to be of clinically significance.
-As expected, obstructive jaundice was associated with markedly elevated liver enzymes with severity of derangement associated with severity of jaundice.
-Obstructive jaundice and sOJ were associated with increasing CRP levels (p$<$0.001) and decreasing serum albumin levels (p$<$0.001).
-Obstructive jaundice was not associated with deranged renal function with both urea and creatinine remaining similar across all cohorts (p=0.09 and p=0.22 respectively).
+Obstructive jaundice was associated with the presence of a systemic inflammatory response as characterised by raised serum C-reactive protein levels and decrease in serum albumin levels (p$<$0.001).
+The degree of systemic inflammation was proportionate to the severity of jaundice.
+Non-jaundiced patients had a median CRP of 3.6 mg/dl (inter-quartile range 0.3 - 89) and serum albumin of 37 g/dl (IQR 18 - 46) while patients with severe obstructive jaundice had a median CRP of 13 (IQR 1.7 - 51) and a median serum albumin of 25 g/dl (IQR 18 - 33).
+Obstructive jaundice was also associated with electrolyte abnormalities including serum sodium, serum potassium and serum chloride levels.
+Jaundiced patients were more likely to be anaemic (p$<$0.001) with lower haematocrit (p$<$0.001) and lower mean corpuscular volume (p=0.001).
+However, there was no significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.
 
@@
 
-There was no association between obstructive jaundice and preoperative pulmonary function tests (Table \ref{table:cpet_oj_pft}).
+Pulmonary function tests including forced vital capacity (FVC), forced expiratory volume in one second (FEV1) and their derived parameters (predicted FEV1, FEV1/FVC, predicted FEV1/FVC) were compared between jaundiced and non-jaundiced patients.
+There was a trend towards lower forced vital capacity with increasing severity of jaundice but this did not reach statistical significance (p=0.092).
+There was no association between the other pulmonary function tests and jaundice.
+(Table \ref{table:cpet_oj_pft}).
 
@@
 
-\subsubsection{Univariate analysis of obstructive jaundice versus CPET}
-The relationship between obstructive jaundice and multiple physiological parameters measured at cardiopulmonary exercise testing is shown in Table \ref{table:cpet_oj_anaerobic}.
-There was an inverse relationship between oxygen consumption at the anaerobic threshold ($\dot{V}_{O_2}$AT) and increasing severity of jaundice (p$<$0.05).
-However, no such linear relationship was noted between any of the other parameters measured both at anaerobic threshold and at peak exercise in spite of apparent statistically significant associations.
+Cardiopulmonary exercise test parameters measured at the anaerobic threshold and at peak exercise were compared between jaundiced and non-jaundiced patients using the Kruskal-Wallis test.
+Obstructive jaundice was associated with lower tidal volume (p=0.017), lower absolute $\dot{V}_{O_2}$ (p=0.003), lower corrected $\dot{V}_{O_2}$ (p=0.029), lower oxygen pulse (p=0.037) and lower respiratory rate (p=0.022).
+At peak exercise, jaundice was associated with lower tidal volume (p=0.028), lower absolute $\dot{V}_{O_2}$Peak (p=0.007), lower absolute $\dot{V}_{CO_2}$ (p=0.016) and lower end-tidal $O_2$ or $PET_{O_2}$ (p=0.026).
+However, as shown in Tables \ref{table:cpet_oj_anaerobic} and \ref{table:cpet_oj_peak}, several of these associations were not linear with a trend towards poorer results between non-jaundiced and mildly jaundiced patients and better values in the severely jaundiced cohort.
 
@@
 
-\subsubsection{Association between preoperative clinico-pathological factors and $\dot{V}_{O_2}$AT}
-On multivariate analysis female sex (HR 3.75 CI 1.57-8.95 p$<$0.005), high BMI (HR 3.65 CI 1.61-8.26 p$<$0.005), presence of cancer (HR 4.02 CI 1.33-12.16 p$<$0.05) and raised CRP (HR 2.98 CI 1.29-6.86 p$<$0.05) were independently associated with low $\dot{V}_{O_2}$AT ($<$10mls/kg/min).
-However, jaundice was not associated with low $\dot{V}_{O_2}$AT.
-These results are shown in Table \ref{table:cpet_oj_at_regression}
+Binary logistic regression analysis was undertaken to assess preoperative clinico-pathological patient factors associated with a low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak.
+On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/dl (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min.
+Age, cardiovascular and respiratory comorbidity, preoperative biliary drainage, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.
+
+The statistically significant independent variables with p$<$0.05 were entered into a backward step-wise regression model with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min as a categorical, binomial, dependent variable.
+Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/dl (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT.
+Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis.
+(Table \ref{table:cpet_oj_at_regression})
+
+Binary logistic regression analysis of the relationship between patient factors and low $\dot{V}_{O_2}$Peak ($<$ 16 ml/kg/min) is shown in Table \ref{table:cpet_oj_peak_regression}.
+On both univariate and multivariate analysis, female sex, body mass index $>$ 25 and haemoglobin $<$ 12 g/dl were significantly associated with a low $\dot{V}_{O_2}$Peak.
 
@@
 
-\subsubsection{Scatter-plot analysis}
-Scatter-plot analysis comparing serum bilirubin and $\dot{V}_{O_2}$AT as continuous variables is depicted in Figure 1.
-This shows that the relationship between serum bilirubin and AT is weak with an r2 value of only 0.04 (I will have to confirm this but it is not more than 0.1).
+\clearpage
+
+Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak as continuous variables is depicted in Figure \ref{fig:cpet_oj_scatter}.
+This shows that the relationship between serum bilirubin and $\dot{V}_{O_2}$AT is weak with a $\rho^2$ value of 0.035 Pearson's $\rho$ = - 0.187, p = 0.028.
+There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)
+
+\begin{figure}[p]
+	\centering
+	\begin{subfigure}{0.6\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_at_bil}
+		\caption{Pearson's $\rho$ = - 0.187, $\rho^2$ = 0.035, p = 0.028}
+		\label{fig:cpet_oj_scatter_at_bil}
+	\end{subfigure}
+	\begin{subfigure}{0.6\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_peak_bil}
+		\caption{$\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123}
+		\label{fig:cpet_oj_scatter_peak_bil}
+	\end{subfigure}
+
+	\caption{Pearson's correlation and scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients undergoing pancreaticoduodenectomy.}
+	\label{fig:cpet_oj_scatter}
+
+\end{figure}
 

### 2015-08-18 Chapters/BodyComp.tex
Major changes as below
- Thesis summary completed
- Many changes to thesis.cls and main.tex to format abstract page and university url, etc
- Update Intro-body comp - measuring body comp
- Zot.bin backed up and updated
- Conclusion chapter title renamed to Discussion
- It is 01:40 am - Have listened to Dr. Dre, Snoop Dogg and Eminem all day
- 2 shots of Jura and I am knackered yet sleepless after 2 strong coffees
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published

@@
 Major abdominal surgery especially for pancreatic disease is associated with significant morbidity and mortality.
-Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes.
-\parencite{balthazar_acute_2002}
+Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes \parencite{balthazar_acute_2002}.
 
@@
 The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established.
-A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery.\parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011} CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
+A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes}], liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery \parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011}.
+CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
 Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.
@@
 Conventionally these have been reported as per weight ratios (ml/kg/min) to allow comparison between patients.
-However, numerous studies on cardiorespiratory exercise physiology have reported that normalising $\dot{V}_{O_2}$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects.\parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}
+However, numerous studies on cardiorespiratory exercise physiology have reported that normalising $\dot{V}_{O_2}$ using total body weight leads to spurious correlation errors unfairly penalising obese subjects \parencite{seltzer_body_1940, tanner_fallacy_1949, toth_examination_1993, batterham_modeling_1999, goran_total_2000, krachler_cardiopulmonary_2014}
 
@@
 \subsection{Calculation of body composition}
-Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods.\parencite{bredella_comparison_2010,shen_total_2004}
+Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods \parencite{bredella_comparison_2010,shen_total_2004}.
 
 %http://regionstraumapro.com/post/16349545265 - Excellent description of CT window width(W) and level/center(C)
-The coronal and saggital reconstructions were used to accurately identify the L3 and L4 vertebrae.
+The coronal and sagittal reconstructions were used to accurately identify the L3 and L4 vertebrae.
 The CT window width (W) was set at 400 and the level/center (C) was set at 40.
@@
 1.The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org).
-The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us.
-\parencite{anblagan_measurement_2013}
+The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us \parencite{anblagan_measurement_2013}.
 
@@
 The average of each parameter measured between the first and second minute (6 readings) was treated as the rest value.
-Anaerobic threshold was identified using previously established methods and all corresponding parameters at this point were recorded.\parencite{beaver_new_1986,sue_metabolic_1988} Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.
+Anaerobic threshold was identified using previously established methods and all corresponding parameters at this point were recorded \parencite{beaver_new_1986,sue_metabolic_1988}.
+Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.
 
@@
 There were several significant associations between clinico-pathological variables and body composition as depicted in this table.
-
+	\input{Tables/bc_sex_specific_distrib}
+	\input{Tables/bc_clinical}
 	\input{Tables/c4_t1}
@@
 
-Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{"Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation"}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass.
-In a detailed anaysis comparing $\dot{V}_{O_2}$ and body build, he concludes that \textit{"as the index wt./stature increases, O2/wt.
-must be expected to decrease purely as a result of the method used for representing the data." }
+Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight
 
-Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas.\parencite{batterham_modeling_1999} The authors report that as body mass increased, the proportion composed of fat-free mass decreased.
+Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}.
+The authors report that as body mass increased, the proportion composed of fat-free mass decreased.
 They also found that fat-free mass had a linear relationship with oxygen consumption while total body mass did not.
@@
 
-Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998.\parencite{janz_longitudinal_1998} Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
+Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}.
+Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
 Body composition changes were also tracked over this period.
@@
 
-Goran et al reported that total body fat did not affect maximal aerobic capacity.\parencite{goran_total_2000} They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss.
+Goran et al reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}.
+They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss.
 $\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss.
@@
 
-These findings have been replicated by several other authors in different subject groups.\parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014} Several of the above studies also recommend using allometric scaling to avoid the confounding effects of total body weight.
+These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014}.
+Several of the above studies also recommend using allometric scaling to avoid the confounding effects of total body weight.
 However, this has not gained widespread clinical use.
 
-In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage.\parencite{maciejczyk_influence_2014} They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
+In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
+They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
 However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups.
@@
 \subsection{Clinical implications of spurious correlation}
-Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993} While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
+Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
+While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
 However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure.
-This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes.
-\parencite{older_cardiopulmonary_1999} The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}
+This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
+The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.
 
-Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes.\parencite{swart_case-controlled_2012} However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%).
+Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}.
+However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%).
 The average $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
-In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT.\parencite{wilson_impaired_2010} There was no data presented on body mass index in this study.
+In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}.
+There was no data presented on body mass index in this study.
 
@@
 Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery and mitigating the effects of neoadjuvant oncological therapy.
-Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients.\parencite{west_effect_2015}
+Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients \parencite{west_effect_2015}.
 

### 2015-08-18 Chapters/CPET_Jaundice.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

@@
 However, major pancreatic surgery is associated with significant morbidity and mortality and is only undertaken in specialist centres.
-Patient selection, preoperative optimisation, good surgical technique and improvements in postoperative care have all contributed to a reduction in mortality\parencite{winter_1423_2006} but morbidity remains high.
+Patient selection, preoperative optimisation, good surgical technique and improvements in postoperative care have all contributed to a reduction in mortality \parencite{winter_1423_2006} but morbidity remains high.
 While several technical strategies have been described in recent years to minimise morbidity, these strategies are not necessarily based on a better understanding of the physiological basis of postoperative complications in these patients.
@@
 Obstructive jaundice has been reported to be associated with abnormal cardiovascular physiology in several animal and human studies.
-Surgery in the jaundiced patient has been reported to be associated with adverse postoperative haemodynamic events and renal dysfunction.\parencite{pain_perioperative_1985,green_systemic_1995} The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death.\parencite{king_effect_1909} Green and co-workers (1986) described the effects of ‘cholemia' in dogs that were subjected to choledochocaval anastomosis.
-The resultant myocardial depression was described by them as the ‘jaundiced heart'\parencite{green_jaundiced_1986} and has been reported to be associated with poor myocardial response to inotropic stimulation in dogs\parencite{binah_obstructive_1985, bomzon_systemic_1986} as well as humans.\parencite{lumlertgul_jaundiced_1991}
+Surgery in the jaundiced patient has been reported to be associated with adverse postoperative haemodynamic events and renal dysfunction \parencite{pain_perioperative_1985,green_systemic_1995}.
+The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death \parencite{king_effect_1909}.
+Green and co-workers (1986) described the effects of ‘cholemia' in dogs that were subjected to choledochocaval anastomosis.
+The resultant myocardial depression was described by them as the ‘jaundiced heart'\parencite{green_jaundiced_1986} and has been reported to be associated with poor myocardial response to inotropic stimulation in dogs \parencite{binah_obstructive_1985, bomzon_systemic_1986} as well as humans \parencite{lumlertgul_jaundiced_1991}.
 
@@
 Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak.
-The $\dot{V}_{O_2}$AT was calculated using the V-slope\parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents\parencite{society_ats/accp_2003} methods.
-$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us\parencite{chandrabalan_pre-operative_2013} as well as Ausania and co-workers\parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients.
+The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
+$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients.
 Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min.
-Detailed description of cardiopulmonary exercise testing as well as the physiological parameters described in this study are published elsewhere.\parencite{balady_clinicians_2010}
+Detailed description of cardiopulmonary exercise testing as well as the physiological parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
 
@@
 
-The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers.\parencite{older_preoperative_1993} Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general\parencite{snowden_submaximal_2010}, pancreatic\parencite{chandrabalan_pre-operative_2013,ausania_effects_2012}, oesophagogastric\parencite{nagamatsu_preoperative_2001} as well as vascular\parencite{carlisle_mid-term_2007} surgery.
+The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers \parencite{older_preoperative_1993}.
+Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012},[Chapter \ref{ch_cpet_outcomes}], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery.
 Cardiopulmonary exercise testing has been reported to be superior to conventional measures of comorbidity chiefly due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress.
@@
 
-The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death.\parencite{king_effect_1909}
+The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death \parencite{king_effect_1909}.
 
-Jaundice has been reported to be associated with myocardial depression\parencite{green_jaundiced_1986}, poor myocardial response to inotropic stimulation\parencite{lumlertgul_jaundiced_1991}, impaired sympathetic baroreflex sensitivity\parencite{song_baroreflex_2009}, deranged atrial natriuretic peptide levels\parencite{pereira_increased_1994,gallardo_increased_1998} as well as multiple other bile-acid receptor mediated effects on the cardiovascular system.\parencite{khurana_bile_2011} Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and coworkers.\parencite{padillo_improved_2001}
+Jaundice has been reported to be associated with myocardial depression \parencite{green_jaundiced_1986}, poor myocardial response to inotropic stimulation \parencite{lumlertgul_jaundiced_1991}, impaired sympathetic baroreflex sensitivity \parencite{song_baroreflex_2009}, deranged atrial natriuretic peptide levels \parencite{pereira_increased_1994,gallardo_increased_1998} as well as multiple other bile-acid receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}.
+Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and coworkers \parencite{padillo_improved_2001}.
 
@@
 
-Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications.\parencite{pitt_does_1985} But, this study looked at a heterogenous group of patients of which only 7 underwent pancreaticoduodenectomy.
+Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
+But, this study looked at a heterogenous group of patients of which only 7 underwent pancreaticoduodenectomy.
 
-A recent meta-analysis\parencite{sewnath_meta-analysis_2002} analysed data from 5 randomised controlled trials comparing surgery with PBD versus surgery without PBD and concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself.
+A recent meta-analysis \parencite{sewnath_meta-analysis_2002} analysed data from 5 randomised controlled trials comparing surgery with PBD versus surgery without PBD and concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself.
 All five RCTs included in this meta-analysis included a heterogenous group of operations with only a few undergoing pancreaticoduodenectomy while more than 50\% of patients underwent palliative bypass or exploratory laparotomy making comparison of outcomes difficult.
-A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings.\parencite{wang_preoperative_2008}
+A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings \parencite{wang_preoperative_2008}.
 
-The DROP trial sought to clarify the role of PBD in patients undergoing pancreaticoduodenectomy.\parencite{van_der_gaag_preoperative_2010} It randomised patients with bilirubin levels between 40 and 250 either to undergo surgery without PBD or to undergo PBD followed by surgery after 4 - 6 weeks.
+The DROP trial sought to clarify the role of PBD in patients undergoing pancreaticoduodenectomy \parencite{van_der_gaag_preoperative_2010}.
+It randomised patients with bilirubin levels between 40 and 250 either to undergo surgery without PBD or to undergo PBD followed by surgery after 4 - 6 weeks.
 The authors reported that PBD resulted in an increase in incidence of complications of which the majority were related to the drainage procedure itself.
@@
 
-However, such an association has been previously reported.\parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate $\dot{V}_{O_2}$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
-Other authors have suggested that different thresholds for CPET parameters may have to be considered in obese patients to improve risk-prediction.\parencite{donnelly_criteria_1990,hulens_exercise_2001} Cardiopulmonary exercise testing measures oxygen delivery to skeletal muscle.
+However, such an association has been previously reported \parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate $\dot{V}_{O_2}$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
+Other authors have suggested that different thresholds for CPET parameters may have to be considered in obese patients to improve risk-prediction \parencite{donnelly_criteria_1990,hulens_exercise_2001}.
+Cardiopulmonary exercise testing measures oxygen delivery to skeletal muscle.
 Adipose tissue, however, does not contribute to the metabolic activity that is measured during CPET.

### 2015-08-18 Chapters/CPET_Outcomes.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

@@
 Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years.
-\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999} In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy\parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.
+\parencite{cancerresearchuk_cancer_2014}.
+The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}.
+In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.
 
-The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors.\parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010} Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery.
-\parencite{mann_review_2010, mayo_management_2012} However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy.\parencite{teh_patient_2009}
+The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}.
+Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery.
+\parencite{mann_review_2010, mayo_management_2012}.
+However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy \parencite{teh_patient_2009}.
 
 There have been a number of attempts to objectively define patient fitness and its relationship with postoperative outcome.
-Copeland and co-workers (1991) reported that the Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) criteria, in particular the POSSUM physiology score (PPS) could be used to quantify the risk of postoperative morbidity and mortality.\parencite{copeland_possum:_1991} However, the role of POSSUM in predicting postoperative outcome after surgery for pancreatic cancer is not entirely clear.\parencite{de_castro_evaluation_2009, khan_evaluation_2003, kocher_risk-adjustment_2005, pratt_possum_2008, tamijmarane_application_2008} The physiological component of POSSUM as well as other similar risk scoring systems such as E-PASS (Estimation of Physiologic Ability and Surgical Stress)\parencite{haga_estimation_1999} are calculated based on known comorbidities, clinically evident abnormalities in patient physiology or blood tests.
+Copeland and co-workers (1991) reported that the Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) criteria, in particular the POSSUM physiology score (PPS) could be used to quantify the risk of postoperative morbidity and mortality \parencite{copeland_possum:_1991}.
+However, the role of POSSUM in predicting postoperative outcome after surgery for pancreatic cancer is not entirely clear \parencite{de_castro_evaluation_2009, khan_evaluation_2003, kocher_risk-adjustment_2005, pratt_possum_2008, tamijmarane_application_2008}.
+The physiological component of POSSUM as well as other similar risk scoring systems such as E-PASS (Estimation of Physiologic Ability and Surgical Stress)\parencite{haga_estimation_1999} are calculated based on known comorbidities, clinically evident abnormalities in patient physiology or blood tests.
 
-More recently, there has been some evidence that the presence of an ongoing systemic inflammatory response before surgery is associated with the development of postoperative complications in patients undergoing surgery for colorectal cancer\parencite{moyes_preoperative_2009}, oesophageal cancer\parencite{vashist_glasgow_2010} as well as pancreatic cancer.\parencite{knight_evaluation_2010}
+More recently, there has been some evidence that the presence of an ongoing systemic inflammatory response before surgery is associated with the development of postoperative complications in patients undergoing surgery for colorectal cancer \parencite{moyes_preoperative_2009}, oesophageal cancer \parencite{vashist_glasgow_2010} as well as pancreatic cancer \parencite{knight_evaluation_2010}.
 
-Older and co-workers (1993) reported that cardiopulmonary exercise testing (CPET) was an objective evaluation of the response of the cardiovascular and respiratory systems to an increase in oxygen demand during exercise and was useful in predicting perioperative morbidity and mortality in patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}
+Older and co-workers (1993) reported that cardiopulmonary exercise testing (CPET) was an objective evaluation of the response of the cardiovascular and respiratory systems to an increase in oxygen demand during exercise and was useful in predicting perioperative morbidity and mortality in patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
 
@@
 The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons.
-The VO$_2$AT was calculated using the V-slope\parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents\parencite{sue_metabolic_1988} methods.
-Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers\parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
+The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods.
+Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
 
@@
 Postoperative adverse events were recorded using internationally recognised definitions.
-The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae\parencite{bassi_postoperative_2005} and post-operative haemorrhage\parencite{wente_postpancreatectomy_2007}.
-The Clavien-Dindo (CD) classification\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major.
+The International Study Group for Pancreatic Surgery (ISGPS) definitions were used to classify pancreatic fistulae \parencite{bassi_postoperative_2005} and post-operative haemorrhage \parencite{wente_postpancreatectomy_2007}.
+The Clavien-Dindo (CD) classification \parencite{clavien_clavien-dindo_2009, dindo_classification_2004} was used to grade other complications and CD grades III-V were considered major.
 Multiple admissions to critical care as well as re-operations were recorded.
@@
 The incidence of pancreatic fistula was greater in patients with a low VO$_2$AT.
-This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss.\parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006} It is possible that local or operative factors may be compounded by poor oxygen delivery and organ perfusion as measured by cardiopulmonary exercise testing.
+This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss \parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006}.
+It is possible that local or operative factors may be compounded by poor oxygen delivery and organ perfusion as measured by cardiopulmonary exercise testing.
 There was a non-significant trend towards clinically relevant pancreatic fistulae (ISGPS Grades B and C) as well as a significant association with major intra-abdominal abscesses (Clavien-Dindo Grades 3-5 i.e., requiring intervention, associated with organ dysfunction requiring intensive care or resulting in mortality).
@@
 The results of this study also show that patients with a low VO$_2$AT were less likely to receive adjuvant therapy.
-Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly.\parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
-While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade\parencite{winter_1423_2006} even in elderly patients\parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high.\parencite{mann_review_2010} The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.
+Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
+While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade \parencite{winter_1423_2006} even in elderly patients \parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high \parencite{mann_review_2010}.
+The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.
 
@@
 The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.
-While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this.\parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010} Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery.\parencite{older_preoperative_1993}
+While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.
+Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
 
-However, Snowden and co-workers\parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers\parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy.
+However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy.
 The lack of association between low VO$_2$AT and cardiopulmonary complications in this study may have been due to two reasons.
-Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy.\parencite{braga_prognostic_2011} Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.
+Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy \parencite{braga_prognostic_2011}.
+Moreover, the stringent fitness criteria for undergoing pancreaticoduodenectomy may have excluded patients with known co-morbid cardiorespiratory diseases such as severe chronic obstructive pulmonary disease or cardiac failure.
 
-The results of this study are consistent with the results of the study by Ausania and co-workers\parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min.
+The results of this study are consistent with the results of the study by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min.
 However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.
@@
 
-This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’.\parencite{topp_effect_2002} It has been suggested that prehabilitation not only improves aerobic capacity\parencite{jones_effects_2007} but may also improve postoperative recovery.\parencite{mayo_impact_2011, pehlivan_effects_2011} The results of this study show that impaired aerobic capacity is associated with postoperative adverse events.
+This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’ \parencite{topp_effect_2002}.
+It has been suggested that prehabilitation not only improves aerobic capacity \parencite{jones_effects_2007} but may also improve postoperative recovery \parencite{mayo_impact_2011, pehlivan_effects_2011}.
+The results of this study show that impaired aerobic capacity is associated with postoperative adverse events.
 Therefore, it would appear that prehabilitation using interventions such as exercise and nutrition, by improving physiological fitness, may have a role in improving postoperative outcomes after major pancreatic surgery and may improve the proportion of patients receiving adjuvant therapy.

### 2015-08-18 Chapters/CRP_Complications.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

@@
 
-Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery.
-\parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008} It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
+Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery \parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008}.
+It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
 
@@
 The early rise in CRP in patients who developed a POPF is expected as the most common risk factors for POPF are encountered intra-operatively including soft pancreatic texture, small pancreatic duct size and operative blood loss.
-Soft pancreatic texture has in fact been reported to be associated with elevated CRP levels after pancreaticoduodenectomy.\parencite{murakami_soft_2008}
+Soft pancreatic texture has in fact been reported to be associated with elevated CRP levels after pancreaticoduodenectomy \parencite{murakami_soft_2008}.
 
@@
 
-The results of this study are similar to the findings of Welsch and co-workers who analysed 688 patients undergoing a variety of pancreatic resections with pancreaticojejunal anastomosis.\parencite{welsch_persisting_2008} They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections.
+The results of this study are similar to the findings of Welsch and co-workers who analysed 688 patients undergoing a variety of pancreatic resections with pancreaticojejunal anastomosis \parencite{welsch_persisting_2008}.
+They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections.
 They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%).
@@
 In the present study, preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score was not associated with infective complications.
-A preoperative acute phase response has been reported to be associated with an exaggerated postoperative systemic inflammatory response and septic complications.
-\parencite{haupt_association_1997} The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections.\parencite{moyes_preoperative_2009} The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
-These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery.\parencite{dutta_persistent_2011}
+A preoperative acute phase response has been reported to be associated with an exaggerated postoperative systemic inflammatory response and septic complications \parencite{haupt_association_1997}.
+The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections \parencite{moyes_preoperative_2009}.
+The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
+These findings are similar to those reported by Dutta and co-workers who did not find any relationship between preoperative systemic inflammation and postoperative complications in patients undergoing oesophago-gastric surgery \parencite{dutta_persistent_2011}.
 
 The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication.
-This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery.\parencite{mustard_c-reactive_1987} as well as in several subsequent reports.
-\parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011} The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation.
+This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery \parencite{mustard_c-reactive_1987} as well as in several subsequent reports \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011}.
+The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation.
 This study was not designed to look at inflammatory markers later in the postoperative period.
@@
 They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day.
-T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6.\parencite{faist_immunosuppression_1997} The complex interactions between the pro and anti-inflammatory processes, neurohormonal factors of the stress response, the role of the blood-gut barrier in bacterial translocation, local factors including tissue ischaemia and damage, the immunological effects of anaesthesia, blood transfusion and hypothermia remain the subject of ongoing research.\parencite{buttenschoen_effect_2010}
+T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6 \parencite{faist_immunosuppression_1997}.
+The complex interactions between the pro and anti-inflammatory processes, neurohormonal factors of the stress response, the role of the blood-gut barrier in bacterial translocation, local factors including tissue ischaemia and damage, the immunological effects of anaesthesia, blood transfusion and hypothermia remain the subject of ongoing research \parencite{buttenschoen_effect_2010}.
 

### 2015-08-18 Chapters/Introduction.tex
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for body comp chapter based on Colin Richards paper - these dont work

@@
 Neuroendocrine tumours (NET) are a collection of a variety of neoplasia that originate from the endocrine portion of the pancreas.
-This milieu is complicated by other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours.
+Other neoplasia such as intra-ductal papillary neoplasms (IPMN) as well as rare stromal tumours may also occur in this region.
 Occasionally, chronic pancreatitis may present with features similar to pancreatic cancer and can be morphologically, radiologically and histologically difficult to differentiate from cancer.
@@
 Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death.
-Only 21\% survive beyond the first year and 5-year survival is 3\%.
-\parencite{cancerresearchuk_cancer_2014} The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease.\parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}
+Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2014}.
+The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}.
 
@@
 Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination.
-In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination.
+In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination.
 The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.
@@
 In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival.
-Assessing the resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
+Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
 
 \section{Surgical treatment of pancreatic cancer}
-Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure over a hundred years after its description.
-Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg.\parencite{kausch_carcinom_1912}.
-The operation was further popularised initially as a two-stage procedure by Whipple\parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s.\parencite{whipple_rationale_1941,whipple_radical_1950}
+Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure.% over a hundred years after its description. 
+Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}.
+The operation was further popularised initially as a two-stage procedure by Whipple \parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.
 
@@
 Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease.
-Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases.\parencite{fuhrman_rationale_1996} The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation\parencite{bower_irreversible_2011}  and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease.
+Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}.
+The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease.
 Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.
@@
 Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach).
-The rest of the procedure was performed as described extensively elsewhere.
 The gastrocolic omentum was divided to enter the lesser sac.
@@
 
-Reconstruction was performed as follows: Either a pancreatico-jejunostomy was performed using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was performed using 3/0 Biosyn sutures placed in a similar manner.
+Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner.
 Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers.
@@
 \subsection{Complications}
-The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world.\parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000} Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes.
+The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world \parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000}.
+Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes.
 The consensus definitions of the International Study Group for Pancreatic Surgery (ISGPS) and the Clavien-Dindo classification systems were used during the course of the studies reported here.
@@
 Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can be associated with significant short-term morbidity as well as long-term disability.
-The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy.
-\parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008} The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula.
+The reported incidence of postoperative pancreatic fistula varies from 2\% to 30\% after pancreaticoduodenectomy \parencite{yeo_six_1997, deoliveira_assessment_2006, bassi_postoperative_2005, winter_biochemical_2007, pratt_risk_2008}.
+The variation in reported incidence has been largely due to lack of clear definition of what constituted a postoperative pancreatic fistula.
 It can be the result of breakdown or poor healing at the pancreatico-jejunostomy/ pancreatico-gastrostomy or may be the result of direct parenchymal leak unrelated to the anastomosis.
 
-It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula.\parencite{bassi_postoperative_2005} A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.
+It is now generally accepted that 1 in 4 patients will develop a pancreatic fistula as defined by the International Study Group for Pancreatic Fistula (ISGPF) which has published a consensus statement on the definition and grading of postoperative pancreatic fistula \parencite{bassi_postoperative_2005}.
+A postoperative pancreatic fistula is defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at the laboratory used for testing.
 
@@
 
-Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak.\parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996} This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery.
+Haemorrhage is often secondary to non-healing of the pancreatico-jejunal anastomosis leading to leakage of amylase-rich pancreatic juices into the retroperitoneum or secondary to intra-abdominal sepsis or bile leak \parencite{tien_risk_2005, koukoutsis_haemorrhage_2006, choi_delayed_2004, balladur_bleeding_1996}.
+This can then lead to erosion of ligated blood vessels, most commonly the stump of the gastro-duodenal artery.
 Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required.
-The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery\parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.
+The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery \parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.
+.
 
@@
 A number of other adverse events may occur following pancreaticoduodenectomy including cardiopulmonary complications such as myocardial infarction, cardiac arrhythmias, pneumonia and pleural effusions, wound complications such as wound sepsis and dehiscence, intra-abdominal problems including intra-abdominal sepsis, leakage from the hepatico-jejunostomy or the gastro-jejunostomy, infected fluid collections, renal dysfunction, deep vein thrombosis, pulmonary embolism, etc.
-The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients.\parencite{clavien_clavien-dindo_2009, dindo_classification_2004} The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo} and was used to grade complications in this thesis.
+The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients \parencite{clavien_clavien-dindo_2009, dindo_classification_2004}.
+The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo}.
+and was used to grade complications in this thesis.
 Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.
@@
 While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma.
-The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect.
-\parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}
+The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}.
 
-The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma.
-\parencite{neoptolemos_randomized_2004} The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy.
+The ESPAC-1 trial compared the use of chemotherapy with Fluorouracil and Folinic acid or chemoradiotherapy with 20Gy radiotherapy and 5-Fluorouracil (5-FU) versus observation only in patients who underwent potentially curative surgery for pancreatic ductal adenocarcinoma \parencite{neoptolemos_randomized_2004}.
+The 5-year survival rate in patients who received chemotherapy was 20\% while it was 8\% in patients who did not receive chemotherapy, demonstrating a clear survival benefit for adjuvant chemotherapy.
 However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting.
@@
 
-However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy.
-\parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007} A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features.\parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}
+However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007}.
+A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features \parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}.
 
@@
 
-The presence of comorbid conditions is associated with increased incidence of adverse outcomes in patients undergoing treatment for pancreatic cancer\parencite{mann_review_2010}.
-It can also limit the therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment.
-\parencite{sandroussi_sociodemographics_2010} Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy.
-\parencite{schneider_patient_2012}
+The presence of comorbid conditions is associated with increased incidence of adverse outcomes in patients undergoing treatment for pancreatic cancer \parencite{mann_review_2010}.
+It can also limit the therapeutic options available due to the associated complications or side effects of surgery or adjuvant treatment \parencite{sandroussi_sociodemographics_2010}.
+Patients with multiple comorbidities are more likely to have higher readmission rates, morbidity and mortality following discharge after pancreaticoduodenectomy \parencite{teh_patient_2009}.
 
-DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy.\parencite{deoliveira_assessment_2006} Cancer cachexia is associated with increased incidence of complications and mortality after pancreaticoduodenectomy \parencite{pausch_cachexia_2012} while obesity is known to be associated with greater incidence and severity of postoperative complications.\parencite{benns_impact_2009} However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery.
-\parencite{shah_limitations_2012}
+DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy \parencite{deoliveira_assessment_2006}.
+Obesity is known to be associated with greater incidence and severity of postoperative complications \parencite{benns_impact_2009}.
+However, existing methods of measuring the impact of comorbidity on physiological fitness are limited and do not adequately predict outcomes after major pancreatic surgery \parencite{de_castro_evaluation_2009}.
 
@@
 The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing.
-This transition from aerobic to anaerobic respiration may be determined using the V-slope method\parencite{sue_metabolic_1988} or the ventilatory equivalents method.\parencite{beaver_new_1986} Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing.\parencite{society_ats/accp_2003}
+This transition from aerobic to anaerobic respiration may be determined using the V-slope method \parencite{sue_metabolic_1988} or the ventilatory equivalents method \parencite{beaver_new_1986}.
+Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.
 
@@
 \ref{fig:cpet_9panel}F.
-Most respiratory physiologists including at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.
+Most respiratory physiologists including those at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.
 
@@
 They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$).
-Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$).\parencite{older_preoperative_1993}
+Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$) \parencite{older_preoperative_1993}.
 
@@
 Moderate risk group ($\dot{V}_{O_2}AT>11$ml/kg/min and either $\dot{V}_E/\dot{V}_{CO_2}>35$ or evidence of myocardial ischaemia, n=115) was managed in the high dependency unit after surgery and had a cardiovascular mortality of 1.7\%.
-Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality.\parencite{older_cardiopulmonary_1999} Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient.
-\parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}
+Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality \parencite{older_cardiopulmonary_1999}.
+Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient.
+\parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}.
 
@@
 In this heterogeneous cohort, they reported that the percent predicted anaerobic threshold achieved $<$ 75\%, heart rate at the anaerobic threshold and the heart rate response from rest to anaerobic threshold were all independently associated with postoperative morbidity.
-While it is difficult to extrapolate results from this small heterogeneous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients.\parencite{hightower_pilot_2010}
+While it is difficult to extrapolate results from this small heterogeneous cohort to the general surgical population, it would appear that other cardiopulmonary exercise test parameters may have a role in predicting risk in these patients \parencite{hightower_pilot_2010}.
 
@@
 They also included other cardiopulmonary exercise test parameters such as $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ at the anaerobic threshold in addition to $\dot{V}_{O_2}$AT in their analysis and showed that $\dot{V}_{O_2}$AT was more predictive of postoperative morbidity than POSSUM derived morbidity, cardiac risk scoring index or a validated activity questionnaire.
-However, this study also included a heterogeneous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas.\parencite{snowden_submaximal_2010}
+However, this study also included a heterogeneous cohort of diseases including vascular, pancreatic and hepatobiliary disorders and sarcomas \parencite{snowden_submaximal_2010}.
 
-In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9).\parencite{wilson_impaired_2010} This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
-Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) {a}nd in patients without ischaemic heart disease.
+In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9) \parencite{wilson_impaired_2010}.
+This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease.
+Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) {a}.
+nd in patients without ischaemic heart disease.
 It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.
@@
 In a retrospective study involving 91 patients who underwent curative oesophagectomy with 3-field lymphadenectomy via a
-right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$ predicted cardiopulmonary complications.\parencite{nagamatsu_preoperative_2001} They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications.
-They recommended that in patients with $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$, surgical treatment be modified either into a 2-stage procedure or a trans-hiatal procedure avoiding a thoracotomy where possible.
+right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$ predicted cardiopulmonary complications \parencite{nagamatsu_preoperative_2001}.
+They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications.
+They recommended that in patients with $\dot{V}_{O_2}.
+max/m^2<800$ml/min/m$^2$, surgical treatment be modified either into a 2-stage procedure or a trans-hiatal procedure avoiding a thoracotomy where possible.
 
@@
 They also observed a significant negative relationship between BMI and $\dot{V}_{O_2}$ ($p<0.0001$) and age($p<0.0001$).
-Poor aerobic capacity was also associated with the presence of diabetes and hypertension suggesting further evidence of the metabolic cost of obesity in these patients.\parencite{mccullough_cardiorespiratory_2006}
+Poor aerobic capacity was also associated with the presence of diabetes and hypertension suggesting further evidence of the metabolic cost of obesity in these patients \parencite{mccullough_cardiorespiratory_2006}.
 
-However, Forshaw and co-workers reported that cardiopulmonary exercise testing was of limited use in predicting complications after oesophagectomy.\parencite{forshaw_is_2008} They studied 78 patients undergoing either trans-hiatal (n=39) or trans-thoracic (n=39) oesophagectomy.
+However, Forshaw and co-workers reported that cardiopulmonary exercise testing was of limited use in predicting complications after oesophagectomy \parencite{forshaw_is_2008}.
+They studied 78 patients undergoing either trans-hiatal (n=39) or trans-thoracic (n=39) oesophagectomy.
 They evaluated several thresholds for both $\dot{V}_{O_2}$Peak and $\dot{V}_{O_2}$AT and found no useful correlation with cardiopulmonary complications, non-cardiopulmonary complications, length of stay in critical care or in hospital.
@@
 Older and Hall, the pioneers of cardiopulmonary exercise test in surgical patients, also raise this issue in their letter to the authors of the above study.
-\parencite{hall_cardiopulmonary_2009}
+\parencite{hall_cardiopulmonary_2009}.
 
@@
 They found no difference in the $\dot{V}_{O_2}$Peak in patients who developed complications versus those who did not (18.6 vs 2.8 ml/kg/min, p$>$0.05).
-However, they reported that in the 4 patient who were denied surgery as they were deemed medically unfit, $\dot{V}_{O_2}$Peak was less than 20 ml/kg/min.
-However, this threshold did not discriminate patients who developed complications and the authors concluded that cardiopulmonary exercise testing should not be used on its own to guide clinical management of these patients.\parencite{nugent_cardiopulmonary_1998}
+However, they reported that in the 4 patients who were denied surgery as they were deemed medically unfit, $\dot{V}_{O_2}$Peak was less than 20 ml/kg/min.
+However, this threshold did not discriminate patients who developed complications and the authors concluded that cardiopulmonary exercise testing should not be used on its own to guide clinical management of these patients \parencite{nugent_cardiopulmonary_1998}.
 
@@
 Carlisle and Swart reported that a Revised Cardiac Risk Index (RCRI) score $>1$ in combination with a $\dot{V}_E/\dot{V}_{CO_2}>42$ was associated with a 2-year survival rate of 55\% while the 2-year survival rate was 97\% if neither risk factor was present.
-$\dot{V}_{O_2}$AT was not a predictor of survival in multi-variate analysis in this study.\parencite{carlisle_mid-term_2007}
+$\dot{V}_{O_2}$AT was not a predictor of survival in multi-variate analysis in this study \parencite{carlisle_mid-term_2007}.
 
@@
 In patients with cirrhosis awaiting hepatic transplantation, reduced aerobic capacity may not only be due to primary cardiorespiratory insufficiency but may also be due to the secondary effects of hepatic dysfunction itself.
-These include cirrhotic cardiomyopathy, hepatopulmonary syndrome and decreased peripheral oxygen utilisation due to cirrhotic myopathy.\parencite{epstein_aerobic_2004} This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery.
-\parencite{iscar_functional_2009} However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
+These include cirrhotic cardiomyopathy, hepatopulmonary syndrome and decreased peripheral oxygen utilisation due to cirrhotic myopathy \parencite{epstein_aerobic_2004}.
+This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery \parencite{iscar_functional_2009}.
+.
+However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
 
@@
 They also found that the mean $\dot{V}_{O_2}$Max in patients who did not develop complications was approximately 20 ml/kg/min and the mean $\dot{V}_{O_2}$Max in patients who developed complications was 15 ml/kg/min across all studies.
-This would suggest that cardiopulmonary exercise testing can offer clinically useful thresholds across different populations that can be used to risk stratify patients.\parencite{benzo_complications_2007}
+This would suggest that cardiopulmonary exercise testing can offer clinically useful thresholds across different populations that can be used to risk stratify patients \parencite{benzo_complications_2007}.
 
 The 2$^nd$ edition of the American College of Chest Physicians' Evidence-Based Clinical Practice Guidelines for the 'Physiologic evaluation of the patient with lung cancer being considered for resectional surgery' recommends that patients with a low $\dot{V}_{O_2}$Max ($<10$ ml/kg/min) or with low $\dot{V}_{O_2}$Max ($<15$ ml/kg/min) associated with other risk factors on spirometry and cardiopulmonary exercise testing should be counselled about non-operative treatment options and non-standard surgery as the risk of perioperative death and cardiopulmonary complications was very high.
-The 1C grading of this recommendation not only represents the observational nature of most of the studies it is based on but also that such observational evidence is sufficiently consistent\parencite{benzo_complications_2007} to make a strong recommendation.\parencite{colice_physiologic_2007}
+The 1C grading of this recommendation not only represents the observational nature of most of the studies it is based on but also that such observational evidence is sufficiently consistent \parencite{benzo_complications_2007} to make a strong recommendation \parencite{colice_physiologic_2007}.
 
@@
 
-One of the earliest reports on the use of CRP to predict cancer-specific survival was by McMillan and co-workers in 1995 when they reported that an elevated CRP 4 months after curative resection for colorectal cancer was associated with earlier recurrence.\parencite{mcmillan_prospective_1995} The modified Glasgow Prognostic Score (mGPS)\parencite{elahi_score_2004} is based on a combination of C-reactive protein and serum albumin and is outlined in Table \ref{table:mGPS}.
+One of the earliest reports on the use of CRP to predict cancer-specific survival was by McMillan and co-workers in 1995 when they reported that an elevated CRP 4 months after curative resection for colorectal cancer was associated with earlier recurrence \parencite{mcmillan_prospective_1995}.
+The modified Glasgow Prognostic Score (mGPS)\parencite{elahi_score_2004} is based on a combination of C-reactive protein and serum albumin and is outlined in Table \ref{table:mGPS}.
+.
 Since its introduction, mGPS has been validated in over a hundred studies looking at several thousand patients with a wide-range of cancers and an increasing score is associated with poorer long-term survival in patients with operable as well as inoperable cancers.
@@
 \subsection{Systemic inflammation and long-term survival}
-Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable disease.\parencite{glen_evaluation_2006} Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated over 10 mg/dl approximately a month after their operation.\parencite{jamieson_systemic_2005} Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR).
+Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable disease \parencite{glen_evaluation_2006}.
+Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated over 10 mg/dl approximately a month after their operation \parencite{jamieson_systemic_2005}.
+Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR).
 A selection of these studies are presented in Table %Create new table here for this.
@@
 \subsubsection{Preoperative systemic inflammation}
-Elevated levels of interleukin-6, alpha-1 antitrypsin and CRP and decreased levels of albumin and prealbumin before surgery have been reported to be associated with a more exaggerated postoperative systemic inflammatory response and infectious complications after major abdominal surgery.\parencite{haupt_association_1997}
+Elevated levels of interleukin-6, alpha-1 antitrypsin and CRP and decreased levels of albumin and prealbumin before surgery have been reported to be associated with a more exaggerated postoperative systemic inflammatory response and infectious complications after major abdominal surgery \parencite{haupt_association_1997}.
 
-Preoperative systemic inflammation has been reported to be associated with infectious complications in patients undergoing potentially curative surgery for colorectal cancer.\parencite{moyes_preoperative_2009} In a study of 455 patients, Moyes and co-workers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery.
+Preoperative systemic inflammation has been reported to be associated with infectious complications in patients undergoing potentially curative surgery for colorectal cancer \parencite{moyes_preoperative_2009}.
+In a study of 455 patients, Moyes and co-workers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery.
 They postulated that several mechanisms may have a role including disregulation of cell-mediated immunity, impaired T-lymphocyte response, disorders in the complement pathway and possibly due to loss of lean tissue and protein as a consequence of systemic inflammation.
-Preoperative mGPS has also been shown to predict postoperative morbidity in patients undergoing oesophageal resection for cancer.\parencite{vashist_glasgow_2010}
+Preoperative mGPS has also been shown to predict postoperative morbidity in patients undergoing oesophageal resection for cancer \parencite{vashist_glasgow_2010}.
 
@@
 An exaggerated and persistent systemic inflammatory response in the early postoperative period is associated with an increased incidence of complications.
-One of the earliest studies comparing several 'acute-phase proteins' and their role in predicting postoperative complications reported that in patients who developed surgical inflammatory complications, CRP remained elevated after the third postoperative day while other acute-phase proteins such as ceruloplasmin and alpha-1 antitrypsin where not useful in monitoring the postoperative course.\parencite{fischer_quantitation_1976}
+One of the earliest studies comparing several 'acute-phase proteins' and their role in predicting postoperative complications reported that in patients who developed surgical inflammatory complications, CRP remained elevated after the third postoperative day while other acute-phase proteins such as ceruloplasmin and alpha-1 antitrypsin where not useful in monitoring the postoperative course \parencite{fischer_quantitation_1976}.
 
-Further studies have established the value of monitoring trends in serum CRP levels in predicting complications after both elective and emergency surgery.\parencite{mustard_c-reactive_1987}
+Further studies have established the value of monitoring trends in serum CRP levels in predicting complications after both elective and emergency surgery \parencite{mustard_c-reactive_1987}.
 
-In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak.\parencite{welsch_c-reactive_2007} They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreatico-jejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
+In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak \parencite{welsch_c-reactive_2007}.
+They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreatico-jejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
 
-Similar findings have been reported after elective colorectal surgery\parencite{ortega-deballon_c-reactive_2010, woeste_increased_2010}, oesophago-gastric surgery\parencite{dutta_persistent_2011}, spinal surgery\parencite{meyer_c-reactive_1995,mok_use_2008}, neurosurgery\parencite{al-jabi_value_2010}, simultaneous pancreas-kidney transplantation\parencite{wullstein_high_2004}, stem-cell transplantation\parencite{mcneer_early_2010} and paediatric surgery\parencite{laporta_baez_c-reactive_2011}.
+Similar findings have been reported after elective colorectal surgery \parencite{ortega-deballon_c-reactive_2010, woeste_increased_2010}, oesophago-gastric surgery \parencite{dutta_persistent_2011}, spinal surgery \parencite{meyer_c-reactive_1995,mok_use_2008}, neurosurgery \parencite{al-jabi_value_2010}, simultaneous pancreas-kidney transplantation \parencite{wullstein_high_2004}, stem-cell transplantation \parencite{mcneer_early_2010} and paediatric surgery\parencite{laporta_baez_c-reactive_2011}.
 
@@
 
-In their seminal work on the role of SIRS and CARS in the pathogenesis of sepsis and organ dysfunction, Bone and co-workers described a state of 'immunologic dissonance' where a 'pre-primed' immune system may result in an inappropriate, out-of-balance massive pro-inflammatory response which is followed by a proportionately large compensatory anti-inflammatory response that leaves the patient immunosuppressed and prone to further organ dysfunction, infections and death.
-\parencite{bone_sepsis:_1997, bone_immunologic_1996} It is very likely that similar mechanisms are involved in surgical patients except that the initial stressor in this case is surgical trauma rather than a bacterial infection as in sepsis.
+In their seminal work on the role of SIRS and CARS in the pathogenesis of sepsis and organ dysfunction, Bone and co-workers described a state of \textit{`immunologic dissonance'} where a \textit{`pre-primed'} immune system may result in an inappropriate, out-of-balance massive pro-inflammatory response which is followed by a proportionately large compensatory anti-inflammatory response that leaves the patient immunosuppressed and prone to further organ dysfunction, infections and death \parencite{bone_sepsis:_1997, bone_immunologic_1996}.
+It is very likely that similar mechanisms are involved in surgical patients except that the initial stressor in this case is surgical trauma rather than a bacterial infection as in sepsis.
 
-This form of 'immunoparalysis' was first described in patients after major trauma with tissue damage \parencite{abraham_effects_1985,bandyopadhyay_negative_2007} or after haemorrhage on its own without associated tissue trauma.
-\parencite{stephan_hemorrhage_1987} In a detailed review of the mechanisms underlying the compensatory anti-inflammatory response syndrome, Ward and co-workers describe SIRS and CARS to be mirror images suggesting that a disproportionately high SIRS is followed by a period of immunosuppression that leaves the patient prone to further complications.\parencite{ward_compensatory_2008}
+This form of \textit{`immunoparalysis'} was first described in patients after major trauma with tissue damage \parencite{abraham_effects_1985,bandyopadhyay_negative_2007} or after haemorrhage on its own without associated tissue trauma \parencite{stephan_hemorrhage_1987}.
+In a detailed review of the mechanisms underlying the compensatory anti-inflammatory response syndrome, Ward and co-workers describe SIRS and CARS to be mirror images suggesting that a disproportionately high SIRS is followed by a period of immunosuppression that leaves the patient prone to further complications \parencite{ward_compensatory_2008}.
 
-Patients who developed infectious complications after major cancer surgery had higher levels of interleukin-10 (IL-10), an anti-inflammatory cytokine and marker of the compensatory anti-inflammatory process.\parencite{mokart_early_2002} Major surgery and the associated surgical trauma is associated with elevated levels of IL-10 which in turn is associated with increase in lymphocyte apoptosis \parencite{delogu_interleukin-10_2001}, reduced monocyte expression of HLA-DR antigens \parencite{klava_interleukin-10._1997} and a blunted response to endotoxins \parencite{ogata_role_2000, kawasaki_surgical_2001}, all considered to be key features of a compensatory anti-inflammatory response syndrome.
+Patients who developed infectious complications after major cancer surgery had higher levels of interleukin-10 (IL-10), an anti-inflammatory cytokine and marker of the compensatory anti-inflammatory process \parencite{mokart_early_2002}.
+Major surgery and the associated surgical trauma is associated with elevated levels of IL-10 which in turn is associated with increase in lymphocyte apoptosis \parencite{delogu_interleukin-10_2001}, reduced monocyte expression of HLA-DR antigens \parencite{klava_interleukin-10._1997} and a blunted response to endotoxins \parencite{ogata_role_2000, kawasaki_surgical_2001}.
+, all considered to be key features of a compensatory anti-inflammatory response syndrome.
 
@@
 This in turn was related to the greater degree of surgical stress and tissue trauma that occurs with a trans-thoracic oesophagectomy.
-They also reported that in a randomised cohort that received an infusion of lymphokine-activated natural killer cells immediately after oesophagectomy, there was a trend towards fewer infectious complications.\parencite{yamaguchi_postoperative_2006}
+They also reported that in a randomised cohort that received an infusion of lymphokine-activated natural killer cells immediately after oesophagectomy, there was a trend towards fewer infectious complications \parencite{yamaguchi_postoperative_2006}.
 
@@
 There has been increasing evidence that postoperative complications not only have an impact on the short-term outcomes but also on long-term survival after major cancer surgery.
-A recent meta-analysis of 21 studies including 21,902 patients found that anastomotic leakage was associated with earlier local recurrence after rectal cancer surgery, a trend towards early local recurrence in other colonic cancer surgery and a significant reduction in overall survival.\parencite{mirnezami_increased_2011} The reviewers suggested that several mechanisms may be involved in early recurrence including local spillage of cancer cells from within the bowel lumen.
+A recent meta-analysis of 21 studies including 21,902 patients found that anastomotic leakage was associated with earlier local recurrence after rectal cancer surgery, a trend towards early local recurrence in other colonic cancer surgery and a significant reduction in overall survival \parencite{mirnezami_increased_2011}.
+The reviewers suggested that several mechanisms may be involved in early recurrence including local spillage of cancer cells from within the bowel lumen.
 However, the role of the local inflammatory processes that occur as a consequence of anastomotic leakage may play a more important role.
@@
 McArdle and co-workers reported in their study of 2235 patients undergoing colorectal cancer surgery that anastomotic leakage was associated with early local recurrence and reduced survival.
-They suggested that the 'double-hit' of surgery followed by anastomotic leak may result in an inflammatory response that is greater and more protracted and that this may explain the poorer cancer outcomes in these patients.\parencite{mcardle_impact_2005} In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP 2 weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications.\parencite{katoh_anastomotic_2011} Wound infections and intra-abdominal infections have also been associated with poorer survival in colorectal cancer patients.\parencite{nespoli_impact_2006} Similar findings have been reported after curative surgery for advanced gastric cancer with patients who develop an anastomotic leak surviving for 30.5 months while patients who did not have a leak survived for a median of 96.2 months (p$<$0.001).
-\parencite{yoo_negative_2011}
+They suggested that the 'double-hit' of surgery followed by anastomotic leak may result in an inflammatory response that is greater and more protracted and that this may explain the poorer cancer outcomes in these patients \parencite{mcardle_impact_2005}.
+In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP 2 weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications \parencite{katoh_anastomotic_2011}.
+Wound infections and intra-abdominal infections have also been associated with poorer survival in colorectal cancer patients \parencite{nespoli_impact_2006}.
+Similar findings have been reported after curative surgery for advanced gastric cancer with patients who develop an anastomotic leak surviving for 30.5 months while patients who did not have a leak survived for a median of 96.2 months (p$<$0.001) \parencite{yoo_negative_2011}.
+.
 
 Patients who develop severe postoperative complications after pancreaticoduodenectomy for cancer had significantly shortened survival in a study involving 428 patients (16.5 vs.
-12.4 months, p=0.002) and this was independent of other recognised risk factors such as tumour grade and lymph node status.
-\parencite{kamphues_postoperative_2012} Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007} and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head.
+12.4 months, p=0.002) and this was independent of other recognised risk factors such as tumour grade and lymph node status \parencite{kamphues_postoperative_2012}.
+Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007}.
+and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head.
 \parencite{kang_detrimental_2009}
@@
 The detrimental effects of jaundice on the heart and the circulatory system has been recognised for over 150 years now.
-King and co-workers reported that the intravenous injection of bile caused bradycardia, hypotension and eventually death in dogs.\parencite{king_effect_1909} The concept of a \textit{'jaundiced heart'} was fist put forth by Green and co-workers in 1986.\parencite{green_jaundiced_1986} They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure.
+King and co-workers reported that the intravenous injection of bile caused bradycardia, hypotension and eventually death in dogs \parencite{king_effect_1909}.
+The concept of a \textit{'jaundiced heart'} was fist put forth by Green and co-workers in 1986 \parencite{green_jaundiced_1986}.
+They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure.
 They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimmetic agents.
 Similar findings have been reported by other authors in animal studies.
-\parencite{binah_obstructive_1985,bomzon_systemic_1986}
+\parencite{binah_obstructive_1985,bomzon_systemic_1986}.
 
 The role of atrial natriuretic peptide (ANP) has also been investigated.
-Obstructive jaundice is associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits.\parencite{pereira_increased_1994} Similar findings have been reported in humans as well.
-\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}
+Obstructive jaundice is associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits \parencite{pereira_increased_1994}.
+Similar findings have been reported in humans as well.
+\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}.
 
-Moreover, relief of obstructive jaundice is associated with improvement in endocrine markers of fluid homeostasis as well as cardiac function.
-\parencite{padillo_improved_2001, gallardo_increased_1998} Padillo and co-workers reported that there was a negative correlation between serum bilirubin levels and left ventricular systolic work and this was associated with elevated ANP and BNP (brain natriuretic peptide) levels.
-However, both ANP and BNP levels decreased after biliary drainage and there was a significant improvement in cardiac output, cardiac index, systolic volume and left ventricular systolic work.\parencite{padillo_improved_2001}
+Moreover, relief of obstructive jaundice is associated with improvement in endocrine markers of fluid homeostasis as well as cardiac function \parencite{padillo_improved_2001, gallardo_increased_1998}.
+Padillo and co-workers reported that there was a negative correlation between serum bilirubin levels and left ventricular systolic work and this was associated with elevated ANP and BNP (brain natriuretic peptide) levels.
+However, both ANP and BNP levels decreased after biliary drainage and there was a significant improvement in cardiac output, cardiac index, systolic volume and left ventricular systolic work \parencite{padillo_improved_2001}.
 
-More recently, TNF-$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF-$\alpha$ agent restored myocardial contractility.\parencite{yang_mechanisms_2010} Obstructive jaundice is also associated with systemic hypotension and there has been increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree.\parencite{green_systemic_1995, lefebvre_role_2009} Bile acids can thus cause vasodilation by decreasing arterial tone and this may partly explain some of the haemodynamic adverse events that occur after surgery in jaundiced patients.
+More recently, TNF-$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF-$\alpha$ agent restored myocardial contractility \parencite{yang_mechanisms_2010}.
+Obstructive jaundice is also associated with systemic hypotension and there has been increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree \parencite{green_systemic_1995, lefebvre_role_2009}.
+Bile acids can thus cause vasodilation by decreasing arterial tone and this may partly explain some of the haemodynamic adverse events that occur after surgery in jaundiced patients.
 
@@
 Several studies have reported that obstructive jaundice is associated with significant abnormalities in fluid homeostasis.
-Obstructive jaundice is associated with reduction in the interstitial volume as well as the circulating plasma volume.\parencite{sitges-serra_body_1992, padillo_preoperative_1999} In a study of 63 patients with obstructive jaundice, Padillo and co-workers reported that severity of jaundice, age of patient and reduced urinary sodium excretion were independently related to postoperative renal dysfunction.
-They also reported that these variables were related to abnormalities in the levels of hormones responsible for sodium and water homeostasis including atrial natriuretic peptide.\parencite{padillo_multivariate_2005} Endotoxemia consequent to lack of bile salts in the gut has also been postulated as a possible mechanism for renal dysfunction in patients with severe jaundice.\parencite{bailey_endotoxin_1976} Jaundice is one of the leading causes of acute renal failure in tertiary-care hospitals.\parencite{liano_epidemiology_1996}
+Obstructive jaundice is associated with reduction in the interstitial volume as well as the circulating plasma volume \parencite{sitges-serra_body_1992, padillo_preoperative_1999}.
+In a study of 63 patients with obstructive jaundice, Padillo and co-workers reported that severity of jaundice, age of patient and reduced urinary sodium excretion were independently related to postoperative renal dysfunction.
+They also reported that these variables were related to abnormalities in the levels of hormones responsible for sodium and water homeostasis including atrial natriuretic peptide \parencite{padillo_multivariate_2005}.
+Endotoxemia consequent to lack of bile salts in the gut has also been postulated as a possible mechanism for renal dysfunction in patients with severe jaundice \parencite{bailey_endotoxin_1976}.
+Jaundice is one of the leading causes of acute renal failure in tertiary-care hospitals \parencite{liano_epidemiology_1996}.
 
-While expansion of the intravascular volume and avoiding dehydration by the judicious use of intravenous fluids has been recommended in avoiding renal dysfunction in jaundiced patients \parencite{parks_prospective_1994}, others have reported that relief of obstructive jaundice by restoring bile flow improves renal function independent of fluid therapy.\parencite{padillo_randomized_2005} Nonetheless, adequate perioperative fluid management, avoidance of hypotension, sepsis control and relief of obstructive jaundice play important roles in the prevention of renal failure in jaundiced patients.
+While expansion of the intravascular volume and avoiding dehydration by the judicious use of intravenous fluids has been recommended in avoiding renal dysfunction in jaundiced patients \parencite{parks_prospective_1994}, others have reported that relief of obstructive jaundice by restoring bile flow improves renal function independent of fluid therapy \parencite{padillo_randomized_2005}.
+Nonetheless, adequate perioperative fluid management, avoidance of hypotension, sepsis control and relief of obstructive jaundice play important roles in the prevention of renal failure in jaundiced patients.
 
@@
 
-Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated with greater risk and in fact preoperative biliary drainage increased hospitalisation costs.
-\parencite{snellen_influence_1985, bakkevold_morbidity_1993, pitt_does_1985} Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications.
+Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated with greater risk and in fact preoperative biliary drainage increased hospitalisation costs \parencite{snellen_influence_1985, bakkevold_morbidity_1993, pitt_does_1985}.
+Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications.
 \parencite{hatfield_preoperative_1982}
 
-Similar findings have been reported by other authors as well.
-\parencite{lai_preoperative_1994, lai_preoperative_1994, jagannath_effect_2005} However, some authors have recommended that expandable metal stents were associated with fewer stent related complications than plastic stents and may be considered in jaundiced patients where surgery is delayed for other reasons.
+Similar findings have been reported by other authors as well \parencite{lai_preoperative_1994, lai_preoperative_1994, jagannath_effect_2005}.
+However, some authors have recommended that expandable metal stents were associated with fewer stent related complications than plastic stents and may be considered in jaundiced patients where surgery is delayed for other reasons.
 \parencite{wasan_use_2005, mullen_pancreaticoduodenectomy_2005}
 
-A recent randomised controlled trial of 202 patients reported that there was no significant difference in post-operative morbidity (47\% vs, 37\%, p=0.14), mortality or length of stay between the group that underwent preoperative biliary drainage and the group that underwent early surgery.
+A recent randomised controlled trial of 202 patients reported that there was no significant difference in post-operative morbidity (47\% vs, 37\%, p=0.14), mortality or length of stay between the group that underwent preoperative biliary drainage and the group that underwent early surgery \parencite{van_der_gaag_preoperative_2010}.
 There was a high incidence of procedure related complications in the drainage group (46\%).
@@
 It is the most widely accepted measure of body composition, especially of body fat in spite of criticisms that it may not apply to all populations.
-BMI was first devised by Adolphe Quetelet and was formerly known as the Quetelet index.\parencite{eknoyan_adolphe_2008} Quetelet was a Belgian sociologist and mathematician who played a major role in the application of statistical methods to the social sciences.
-In his attempts to describe the characteristics of 'normal man', Quetelet found that weight increased nearly as the square of the height except at times of growth spurts in the new born and during puberty.
-\parencite{quetelet_treatise_1842} The calculation of body mass index is simple and as follows:
+BMI was first devised by Adolphe Quetelet and was formerly known as the Quetelet index \parencite{eknoyan_adolphe_2008}.
+Quetelet was a Belgian sociologist and mathematician who played a major role in the application of statistical methods to the social sciences.
+In his attempts to describe the characteristics of 'normal man', Quetelet found that weight increased nearly as the square of the height except at times of growth spurts in the new born and during puberty \parencite{quetelet_treatise_1842}.
+.
+The calculation of body mass index is simple and as follows:
 
@@
 
-Several studies have reported increased incidence of postoperative complications in obese patients after cardiac, transplant, urologic and gynaecological surgery.
-\parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004} Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer.
+Several studies have reported increased incidence of postoperative complications in obese patients after cardiac, transplant, urologic and gynaecological surgery \parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004}.
+Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer.
 \parencite{tsujinaka_visceral_2008}
@@
 However, in a prospective, multi-institutional, risk-adjusted cohort study of 2258 patients including 699 patients who underwent pancreatectomy for cancer, Mullen and co-workers reported that obesity on its own was not a predictor of major postoperative complications or mortality.
-But,obesity was an independent predictor of minor complications including wound infections.
+However, obesity was an independent predictor of minor complications including wound infections.
 However, they noted that underweight patients had a five-fold increased risk of postoperative mortality.
@@
 There was no difference in the median operating time, blood transfusions or length of hospital stay between obese and non-obese patients.
-The type and severity of complications were similar between obese and non-obese patients.
-\parencite{dindo_obesity_2003} In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen and co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
+The type and severity of complications were similar between obese and non-obese patients \parencite{dindo_obesity_2003}.
+In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen and co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
 This appears to confirm the existence of an 'obesity paradox' in surgical patients.
@@
 
-In a study of 356 consecutive patients undergoing pancreaticoduodenectomy at the Memorial Sloan-Kettering Cancer Center between 2000 and 2005, House and co-workers used preoperative cross-sectional imaging of the abdomen (computer tomography and/or magnetic resonance imaging) to measure a variety of indices of body composition including abdominal wall fat thickness, hip girdle fat thickness, retro-renal visceral fat and pancreatic duct diameter.
-\parencite{house_preoperative_2008} The incidence of wound infection was greater in obese patients with the risk increasing linearly with BMI.
+In a study of 356 consecutive patients undergoing pancreaticoduodenectomy at the Memorial Sloan-Kettering Cancer Center between 2000 and 2005, House and co-workers used preoperative cross-sectional imaging of the abdomen (computer tomography and/or magnetic resonance imaging) to measure a variety of indices of body composition including abdominal wall fat thickness, hip girdle fat thickness, retro-renal visceral fat and pancreatic duct diameter \parencite{house_preoperative_2008}.
+The incidence of wound infection was greater in obese patients with the risk increasing linearly with BMI.
 But, BMI was not associated with any other complications.
@@
 The importance of visceral fat in the incidence and severity of postoperative complications is further emphasised by the finding that excess visceral fat is associated with the the occurence of pancreatic fistula after radical total gastrectomy for gastric cancer.
-Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese pateints undergoing surgery compared to non-obese pateints.
+Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese patients undergoing surgery compared to non-obese pateints.
 However, there was no difference in operative time or length of stay in hospital.
@@
 On the contrary, other authors have reported that obesity is not associated with increased complications after pancreatic surgery.
-Khan and co-workers reported that while obesity was associated with increased incidence of intra-operative bleeding, there was no increase in complications.
-\parencite{khan_does_2010} Length of stay in hospital, overall morbidity and in-hospital mortality were not affected by BMI after pancreaticoduodenectomy \parencite{tsai_impact_2010} while neither BMI nor intra-abdominal visceral fat affected postoperative outcomes in another study of patients undergoing pancreaticoduodenectomy.
-\parencite{balentine_obesity_2011}
+Khan and co-workers reported that while obesity was associated with increased incidence of intra-operative bleeding, there was no increase in complications \parencite{khan_does_2010}.
+Length of stay in hospital, overall morbidity and in-hospital mortality were not affected by BMI after pancreaticoduodenectomy \parencite{tsai_impact_2010}.
+while neither BMI nor intra-abdominal visceral fat affected postoperative outcomes in another study of patients undergoing pancreaticoduodenectomy \parencite{balentine_obesity_2011}.
 
@@
 However, the authors concluded that with aggressive perioperative care, postoperative morbidity in obese patients could be mitigated.
-They also noted several limitations to making meaningful comparisons between studies due to the lack of clear definitions for complications after pancreaticoduodenectomy.
-\parencite{ramsey_body_2011}
+They also noted several limitations to making meaningful comparisons between studies due to the lack of clear definitions for complications after pancreaticoduodenectomy \parencite{ramsey_body_2011}.
 
@@
 The effect of obesity and adipose tissue content on cancer-specific survival in patients with pancreatic cancer is not entirely clear.
-In a study involving 586 patients who underwent pancreaticoduodenectomy for pancreatic adenocarcinoma, Khan and co-workers reported that obesity was not associated with overall or disease-free survival.
-\parencite{khan_does_2010} These are similar to the findings reported by Benns and co-workers who found no difference in overall survival in 306 patients undergoing resections for pancreatic adenocarcinoma.
-\parencite{benns_impact_2009} Obesity during early adulthood is associated with increased risk of pancreatic cancer while obesity at an older age is associated with reduced overall survival in both resected and unresected patients.
-\parencite{li_body_2009} Morbidly obese patients (BMI $>$ 35) were found to be at 12-fold increased risk of lymph node metastases and associated decreased survival.
-\parencite{fleming_influence_2009}
+In a study involving 586 patients who underwent pancreaticoduodenectomy for pancreatic adenocarcinoma, Khan and co-workers reported that obesity was not associated with overall or disease-free survival \parencite{khan_does_2010}.
+These are similar to the findings reported by Benns and co-workers who found no difference in overall survival in 306 patients undergoing resections for pancreatic adenocarcinoma \parencite{benns_impact_2009}.
+Obesity during early adulthood is associated with increased risk of pancreatic cancer while obesity at an older age is associated with reduced overall survival in both resected and unresected patients \parencite{li_body_2009}.
+Morbidly obese patients (BMI $>$ 35) were found to be at 12-fold increased risk of lymph node metastases and associated decreased survival \parencite{fleming_influence_2009}.
 
-Tsai and co-workers reported that while obese patients had similar tumour characteristics to non-obese patients and comparable perioperative outomes, long-term survival was better in obese subjects.
-\parencite{tsai_impact_2010}  This study of 795 patients undergoing potentially curative surgery for pancreatic adenocarcinoma at the Johns Hopkins Medical Institutions in the United States is in contrast to other similar studies.
-The role of obesity on long-term survival after pancreaticoduodenectomy for pancreatic cancer is less clear with more recent studies suggesting that there was no link between body mass index and survival.
-\parencite{dandona_influence_2011, benns_impact_2009}
+Tsai and co-workers reported that while obese patients had similar tumour characteristics to non-obese patients and comparable perioperative outomes, long-term survival was better in obese subjects \parencite{tsai_impact_2010}.
+This study of 795 patients undergoing potentially curative surgery for pancreatic adenocarcinoma at the Johns Hopkins Medical Institutions in the United States is in contrast to other similar studies.
+The role of obesity on long-term survival after pancreaticoduodenectomy for pancreatic cancer is less clear with more recent studies suggesting that there was no link between body mass index and survival \parencite{dandona_influence_2011, benns_impact_2009}.
 
 While sarcopenia or loss of skeletal muscle mass is associated with poor outcome in patients with pancreatic cancer, the association between adipose tissue (especially, visceral fat) and long-term outcomes is less clear.
-Sarcopenic-obesity where concurrent sarcopenia and high fat mass are present is associated with significantly worse survival when compared to sarcopenia or high adiposity on their own.
-\parencite{tan_sarcopenia_2009} It is clear that body mass index on its own is a poor predictor of outcome and body composition changes over time play a more important role in determining cancer survival.
-While obesity appears to have a protective role in a wide range of chronic disease states\parencite{kalantar-zadeh_risk_2007, amundson_obesity_2010, myers_obesity_2011} sarcopenic-obesity has a negative impact on survival.
-\parencite{tan_sarcopenia_2009}
+Sarcopenic-obesity where concurrent sarcopenia and high fat mass are present is associated with significantly worse survival when compared to sarcopenia or high adiposity on their own \parencite{tan_sarcopenia_2009}.
+It is clear that body mass index on its own is a poor predictor of outcome and body composition changes over time play a more important role in determining cancer survival.
+While obesity appears to have a protective role in a wide range of chronic disease states \parencite{kalantar-zadeh_risk_2007, amundson_obesity_2010, myers_obesity_2011}, sarcopenic-obesity has a negative impact on survival \parencite{tan_sarcopenia_2009}.
 
@@
 
-The gold-standard method for assessment of body composition is Dual Energy X-ray Absorptiometry (DEXA).\parencite{haarbo_validation_1991} While primarily designed for assessment of bone mineral density, DEXA scans have been adapted to measure fat, muscle and water content of subjects accurately and are used widely both in research and clinical settings.
+The gold-standard method for assessment of body composition is Dual Energy X-ray Absorptiometry (DEXA) \parencite{haarbo_validation_1991}.
+While primarily designed for assessment of bone mineral density, DEXA scans have been adapted to measure fat, muscle and water content of subjects accurately and are used widely both in research and clinical settings.
 Other techniques such as whole body water or air displacement plethysmography are relatively cumbersome for routine clinical use.

### 2015-08-18 main.tex
CPET Jaundice - Major update
- Next few commits will be following Donny chat this morning
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 2013 to Chapter 2
- Create some new tables for b

@@
 The results demonstrate that patients with $\dot{V}_{O_2}$AT less than 10 ml/kg/min are more likely to develop a postoperative pancreatic fistula, stay longer in hospital and less likely to receive adjuvant therapy.
-These results emphasise the importance of excellent aerobic fitness to recover from the operative stress of major surgery without significant morbidity.
+These results emphasise the importance of aerobic fitness to recover from the operative stress of major surgery without significant morbidity.
 Cardiopulmonary exercise testing may prove useful in selecting patients for intensive prehabilitation programmes as well as for other optimisation measures to prepare them for major surgery.
@@
 University Department of Surgery, Glasgow Royal Infirmary
+
+Mr. Euan J Dickson,
+West of Scotland Pancreatic Unit, Glasgow Royal Infirmary
 }
@@
 % Include the chapters of the thesis as separate files from the Chapters folder
-\input{Chapters/Introduction} 		%1
-\input{Chapters/CPET_Outcomes} 	%2
-\input{Chapters/CPET_Jaundice} 	%3
+%\input{Chapters/Introduction} 		%1
+%\input{Chapters/CPET_Outcomes} 	%2
+%\input{Chapters/CPET_Jaundice} 	%3
 \input{Chapters/BodyComp}			%4
-\input{Chapters/CPET_Survival}		%5
-\input{Chapters/CRP_Complications}	%6
-\input{Chapters/Discussion}		%7
+%\input{Chapters/CPET_Survival}		%5
+%\input{Chapters/CRP_Complications}	%6
+%\input{Chapters/Discussion}		%7
 

### 2015-08-18 Chapters/BodyComp.tex
Body Composition - Major Changes
- Tables changed to include SAT/VAT instead of CSA
- Text changed in most places accordingly
- New tables and figures to reflect new vo2 scalings
- Next commit should finalise this chapter with some luck

@@
 \subsection{Body composition and Clinico-pathological characteristics}
-Eighty-two patients (35 male) were included in the study.
+Eighty-two patients (52 male) were included in the study. 
 The clinico-pathological characteristics of the study patients and their relationship to body composition is shown in Table \ref{table:bc_clinical} on page \pageref{table:bc_clinical}. 
-There were several significant associations between clinico-pathological variables and body composition as depicted in this table.
-	\input{Tables/bc_sex_specific_distrib}
-	\input{Tables/bc_clinical}
+There were several significant associations between clinico-pathological variables and body composition as shown in this table. 
+Females had a greater area of subcutaneous adipose tissue (223.9 versus 141.6 cm$^2$, p$<$0.001) but less visceral adipose tissue (79.1 versus 174.9 cm$^2$, p$<$0.001).
+However, there was no difference in the total adipose tissue area between men and women (316.6 versus 303.0 cm$^2$, p=0.665).
+Skeletal muscle area was greater in males than in females (141.3 versus 99.7 cm$^2$).
+
+Patients from more deprived areas of Scotland as defined by the Scottish Index of Multiple Deprivation (SIMD) had a greater amount of subcutaneous adipose tissue (p=0.041) but no difference in the other body composition parameters.
+Hypoalbuminemia was associated with less visceral adipose tissue (p=0.013) and less skeletal muscle area that did not reach statistical significance (p=0.054). 
+Anaemia (haemoglobin $<$ 12 g/dl) was associated with increased adiposity in the subcutaneous plane and reduced skeletal muscle area. 
+	%\input{Tables/bc_sex_specific_distrib}
+	%\input{Tables/bc_clinical}
 	\input{Tables/c4_t1}
 
-\subsection{Body Composition in Normal BMI vs Overweight/Obese Patients}
-The body composition differences between patients with a normal BMI and patients who are overweight or obese is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}.
+\subsection{Body composition differences between the sexes with increasing BMI}
+The body composition differences between male and female patients with increasing body mass index is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}.
 There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females.
-Men had generally larger cross-sectional area, less SAT but greater VAT and SM areas.
-However, the proportion of skeletal muscle in both males and females decreased significantly with increasing BMI.
+Men generally had less subcutaneous fat but more visceral fat and skeletal muscle areas.
+However, the proportion of skeletal muscle in both males and females decreased significantly with increasing body mass index.
 
@@
 	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
-	\caption{Differences in body composition according to gender and body mass index.}
+	\caption{Differences in body composition according to sex and body mass index.}
 	\label{fig:bc_gender_bmi}
@@
 
-The proportion of skeletal muscle area at L3/L4 decreases from 38\% in male patients with normal BMI to 22\% in males who are obese.
-There was a greater decrease in the proportion of skeletal muscle area in females with normal BMI (32\%) and obese females (14\%).
+The proportion of skeletal muscle area decreases from 38\% in male patients with normal BMI to 22\% in males who are obese.
+There was a greater decrease in the proportion of skeletal muscle in females with increasing BMI, with skeletal muscle area decreasing from 32\% in females with normal BMI to 14\% in obese females.
 The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle.
 Moreover, the distribution of the adipose tissue differed between males and females with visceral adipose tissue contributing more to weight in obese males (43\% VAT vs.
-35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (68\% SAT vs.
-18\% VAT)
+35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (18\% VAT vs.
+68\% SAT).
 
@@
 FEV1/FVC was not correlated with any of the body composition components.
+\todo{move this to discussion}
 This would indicate that pulmonary function was dependent on skeletal muscle area while FEV1/FVC, a calculated index to quantify restrictive or obstructive lung disease, was not associated with skeletal muscle area.
@@
 \subsection{Correlation with Exercise Load}
-
-\input{Tables/c4_t2}
-
-Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area and subcutaneous adipose tissue area measured at L3/L4 to create scatter-plots(Fig.
-\ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}.
-Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig.
-\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig.
-\ref{fig:bc_pk_load_vs_sm}).
-However, no correlation was identified between exercise loads achieved and subcutaneous adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$).
 
@@
 
+Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area to create scatter-plots(Fig.
+\ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}.
+Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig.
+\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig.
+\ref{fig:bc_pk_load_vs_sm}).
+However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on p\pageref{table:bc_cpet}.
+
 %This would suggest as one would expect that exercise capacity is mostly related to skeletal muscle volume rather than subcutaneous adipose tissue. - Move this to discussion
@@
 
-The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender.
+\input{Tables/bc_cpet}
+
+Partial correlations between cardiopulmonary exercise parameters at rest, anaerobic threshold and peak exercise versus body composition were adjusted for gender.
 Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age.
@@
 
-% Absolute oxygen consumption at the anaerobic threshold in litres per minute was plotted against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area.
-This was repeated by plotting the oxygen consumption at anaerobic threshold corrected for total body weight (in millilitres/kg/min) against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area.
-These scatter-plots are shown in FigureX1.
-The same was repeated for oxygen consumption at peak exercise and at rest.
-
 Tidal volume ($\dot{V}_T$, litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise.
 There was a statistically significant but weak positive correlation between minute ventilation ($\dot{V}_E$, litres) and skeletal muscle at anaerobic threshold and peak exercise but not at rest.
-There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.
+There was no correlation between either of these measures of pulmonary function and adipose tissue area at any phase of exercise.
 
@@
 
-Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$).
+Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.353, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p = 0.001$).
 However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.
@@
 
+$O_2$Pulse at the anaerobic thresold and peak exercise was strongly associated with skeletal muscle area with weaker positive correlations with visceral fat and total fat content.
+There was no correlation between $O_2$Pulse and subcutaneous adipose tissue at any phase of exercise.
+
 %$\dot{V}_{O_2}$ corrected to height seems to correlate best with BC and everything else
+
+\subsection{Scaling $\dot{V}_{O_2}$ to body composition and other factors}
+
+Absolute $\dot{V}_{O_2}$ at the anaerobic threshold was corrected for weight, the square of height, body mass index, skeletal muscle area and estimated lean body mass.
+Estimated lean body mass (eLBM) was calculated using the Boer formula \parencite{boer_estimated_1984} as shown below.
+
+Men:\ eLBM = 0.407*weight(kg) + 0.267*height(cm) - 19.2
+
+Women:\ eLBM = 0.252*weight(kg) + 0.473*height(cm) - 48.3
+
+The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex.
+These results are shown in Table \ref{table:bc_at_new_indices} on p\pageref{table:bc_at_new_indices}.
+The results from Table \ref{table:bc_cpet} comparing absolute $\dot{V}_{O_2}$ and corrected $\dot{V}_{O_2}$ at the anaerobic threshold with body composition are reproduced in the first two columns of this table to allow comparison with the new indices.
+$\dot{V}_{O_2}$ scaled for skeletal muscle area had no significant correlation with adipose tissue, height, weight or BMI but had a negative correlation with skeletal muscle area.
+However, $\dot{V}_{O_2}$ corrected for the square of patient's height had no correlation with adipose tissue but retained a postive correlation with skeletal muscle area.
+However, there was also a positive correlation with weight and BMI.
+
+However, when absolute $\dot{V}_{O_2}$ was corrected for lean body mass, there was no correlation with any of the body composition indices or with height, weight or body mass index.
+The negative correlation that occurs between adipose tissue and $\dot{V}_{O_2}$ corrected for total body weight does not occur when $\dot{V}_{O_2}$ is corrected for lean body mass (Fig.
+\ref{fig:bc_scatter_VO2_TAT_elbm}).
+Moreover, the correlation with skeletal muscle area remains unchanged regardless of whether $\dot{V}_{O_2}$ is corrected for total body weight or lean body mass (Fig.
+\ref{fig:bc_scatter_VO2_skeletal_elbm}).
+
+%----------------------------------------------------------------------------------------------
+% Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
+\begin{figure}[htb]
+	\centering
+	\begin{subfigure}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal_elbm}
+		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
+		\label{fig:bc_scatter_VO2_skeletal_elbm}
+	\end{subfigure}
+	\hfill
+	\begin{subfigure}[b]{0.45\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT_elbm}
+		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
+		\label{fig:bc_scatter_VO2_TAT_elbm}
+	\end{subfigure}
+	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.}
+	\label{fig:bc_scatter_VO2_bodycomp_elbm}
+\end{figure}
+%----------------------------------------------------------------------------------------------
+
+\input{Tables/bc_at_new_indices}
+
 \clearpage
@@
 \subsection{Clinical implications of spurious correlation}
+
+\textit{`Spurious correlation'} is a tern introduced by English mathematician and biometrician, Karl Pearson in 1896.
+He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied\parencite{pearson_mathematical_1896}
 Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.

### 2015-08-18 Tables/bc_at_new_indices.tex
Body Composition - Major Changes
- Tables changed to include SAT/VAT instead of CSA
- Text changed in most places accordingly
- New tables and figures to reflect new vo2 scalings
- Next commit should finalise this chapter with some luck

\begin{sidewaystable}[p]
	\caption{The relationship between body composition, body habitus and $\dot{V}_{O_2}$AT scaled using different factors in patients undergoing major pancreatic surgery.}
	\label{table:bc_at_new_indices}
	\footnotesize
	\centering
	\renewcommand{\arraystretch}{1.5} %Increases space between rows
	%\setlength{\tabcolsep}{9pt} %sets the space between columns
	%\hline\noalign{\smallskip} %increases space after the line
	\begin{tabular}{|l|cc|cc|cc|cc|cc|cc|}
		\hline
		& \multicolumn{2}{c|}{Absolute $\dot{V}_{O_2}$} & \multicolumn{2}{c|}{$\frac{\dot{V}_{O_2}}{Weight}$} & \multicolumn{2}{c|}{$\frac{\dot{V}_{O_2}}{Height^2}$} & \multicolumn{2}{c|}{$\frac{\dot{V}_{O_2}}{BMI}$} & \multicolumn{2}{c|}{$\frac{\dot{V}_{O_2}}{Skeletal\ Muscle}$}  & \multicolumn{2}{c|}{$\frac{\dot{V}_{O_2}}{Lean\ Body\ Mass}$}\\

		Variable              & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p    & $\rho$ &    p \\ \hline
		Subcut. fat (cm2)     & 0.191  &  0.089   & -0.326 &  0.003   & 0.169  &  0.132   & -0.217 &  0.051   & 0.018  &  0.872  & -0.041 & 0.716 \\
		Visceral fat (cm2)    & 0.184  &  0.103   & -0.365 &  0.001   & 0.154  &  0.168   & -0.279 &  0.012   & -0.065 &  0.563   & -0.108 & 0.339\\
		Total fat (cm2)       & 0.231  &  0.038   & -0.400 & $<$0.001 & 0.191  &  0.088   & -0.286 &  0.010   & -0.021 &  0.851   & -0.082 & 0.466\\
		Skeletal Muscle (cm2) & 0.463  & $<$0.001 & -0.078 &  0.487   & 0.329  &  0.003   & 0.105  &  0.353   & -0.429 & $<$0.001 & 0.133 & 0.237 \\
		Height                & 0.446  & $<$0.001 & 0.019  &  0.870   & 0.001  &  0.993   & 0.475  & $<$0.001 & 0.105  &  0.350   & 0.019 & 0.865\\
		Weight                & 0.545  & $<$0.001 & -0.302 &  0.006   & 0.358  &  0.001   & -0.043 &  0.703   & 0.001  &  0.995   & 0.033 & 0.770\\
		Body Mass Index       & 0.357  &  0.001   & -0.363 &  0.001   & 0.425  & $<$0.001 & -0.345 &  0.002   & -0.059 &  0.603   & 0.035 & 0.756\\ \hline
		\multicolumn{11}{l}{$\rho$ - Pearson's r adjusted for sex.}
	\end{tabular}
\end{sidewaystable}

### 2015-08-18 Chapters/BodyComp.tex
Body Composition - Nearly there
- Almost done
- Just need to check minor things
- LEAVE THIS BE NOW

@@
 
-The results of this study show that the most important cardiopulmonary exercise test parameters as used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition.
+The results of this study demonstrate that the most important cardiopulmonary exercise test parameters used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition.
+Both $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak are significantly affected by patient's body habitus when scaled using total body weight.
+Furthermore, the results suggest that there may be alternate methods of scaling of $\dot{V}_{O_2}$ that removes the effect of body composition on these parameters. 
+The results also demonstrate the disproportionate decrease in skeletal muscle area with associated increased in adipose tissue with increasing body mass index. 
+Taken together, these results would suggest that caution should be exercised in interpreting $\dot{V}_{O_2}$ (corrected for total body weight, mls/kg/min) in patients with a raised BMI as abnormally low values in these patients may not necessarily be due to poor aerobic fitness or lack of cardiopulmonary reserve.
 
@@
 
-The positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area is easily explained by the physiology of aerobic exercise.
-During periods of increased physical activity, the greater oxygen demand is primarily due to increased metabolic activity within the skeletal muscle.
+Increased physical activity such as during cardiopulmonary exercise testing results in greater oxygen consumption ($\dot{V}_{O_2}$).
+The increased demand for oxygen is primarily due to increased metabolic activity within the exercising skeletal muscle.
+Therefore, the positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area as has been found here is easily explained by the physiology of aerobic exercise.
 
-Current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according the following formula: 
+However, current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according to the following formula: 
 \[Corrected\ \dot{V}_{O_2} (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
 
-In a previous analysis (refer to chapter and table), we reported that there was a significant negative correlation between $\dot{V}_{O_2}$ at anaerobic threshold and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease.
+In chapter \ref{ch_cpet_jaundice}, we reported that there was a significant negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min) at the anaerobic threshold as well as at peak exercise and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease (Tables \ref{table:cpet_oj_at_regression} and \ref{table:cpet_oj_peak_regression}).
+The results of the present study demonstrate that this negative correlation is consequent to the scaling convention used rather than due to any pathophysiological effect of obesity.
 
-The results of the present study suggest that the negative correlation between corrected $VO_{2}$ (mls/kg/min) and BMI is consequent to the reporting convention rather than due to any pathophysiological effect of obesity.
-
-The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for body weight further supports the argument that the corrected value under-reports aerobic capacity in obese patients.
-Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between O2Pulse and adipose tissue area appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.
+The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for total body weight further supports the argument that the corrected value for $\dot{V}_{O_2}$ under-reports aerobic capacity in obese patients.
+Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.
 
@@
 
-Seltzer reported in his 1940 study of 34 subjects, that the individuals who were more "lateral" than "linear" had lower oxygen intakes per kilo body weight
+Seltzer reported in his 1940 study of 34 subjects, that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{`Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation'}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass.
+In a detailed analysis comparing $\dot{V}_{O_2}$ and body build, he concluded that \textit{'as the index wt./stature increases, O2/wt.
+must be expected to decrease purely as a result of the method used for representing the data.'}
 
 Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}.
-The authors report that as body mass increased, the proportion composed of fat-free mass decreased.
+The authors reported that as body mass increased, the proportion composed of fat-free mass decreased.
 They also found that fat-free mass had a linear relationship with oxygen consumption while total body mass did not.
 They suggest that ideally estimates of fat-free mass should be used in the representation of oxygen consumption to allow more reliable comparison between subjects.
+These findings are similar to our results where the proportion of skeletal muscle area decreasing significantly as body mass index increased with a greater decrease in females than males.
+Moreover, the linear relationship between skeletal muscle area and absolute $\dot{V}_{O_2}$ has also been clearly demonstrated in our findings.
 
-Janz et al studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}.
+We used the Boer formula \parencite{boer_estimated_1984}to estimate lean body mass and used this to normalise $\dot{V}_{O_2}$.
+The resulting value for $\dot{V}_{O_2}$ showed no `spurious' correlations with body composition, height, total body weight or body mass index.
+These findings are similar to the recommendations made by Janz et al who studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}.
 Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
@@
 
+Our results demonstrate that there is no correlation between absolute oxygen consumption at rest or during exercise and the amount of adipose tissue.
+This further emphasises the fact that adipose tissue is metabolically inactive during exercise although it contributes to increased work load.
+Therefore, dividing absolute $\dot{V}_{O_2}$ by total body weight in overweight/obese subjects whose extra weight is largely due to adipose tissue results in lower corrected $\dot{V}_{O_2}$ as we have demonstrated here.
 Goran et al reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}.
@@
 These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014}.
-Several of the above studies also recommend using allometric scaling to avoid the confounding effects of total body weight.
+Several of the above studies also recommended using allometric scaling to avoid the confounding effects of total body weight.
 However, this has not gained widespread clinical use.
 
-In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in a two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
+In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
 They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
@@
 
-\textit{`Spurious correlation'} is a tern introduced by English mathematician and biometrician, Karl Pearson in 1896.
-He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied\parencite{pearson_mathematical_1896}
+The term \textit{`spurious correlation'} was first introduced by English mathematician and biometrician, Karl Pearson in 1896.
+He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied \parencite{pearson_mathematical_1896}.
+
 Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
 While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
-However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure.
+However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tool concluded that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure.
 This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
@@
 However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%).
-The average $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
+The average corrected $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
 In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}.
 There was no data presented on body mass index in this study.
+We reported in Chapter \ref{ch_cpet_jaundice} that females were more likely to have a lower $\dot{V}_{O_2}$AT (Table \ref{table:cpet_oj_at_regression}) and lower $\dot{V}_{O_2}$Peak (Table \ref{table:cpet_oj_peak_regression}).
+However, all the results in the present study were controlled for the effect of sex.
 
-This is similar to the findings in our cohort of patients.
-This may have been due to the increased incidence of obesity especially in the subcutaneous plane as we have found in our cohort of patients as shown in Fig.
-\ref{fig:bc_figure1}.
+The `obesity paradox' would suggest that, in fact, obese patients are not necessarily at an increased risk of postoperative complications.
+Several authors have reported that obesity is not a risk factor for major complications after pancreaticoduodenectomy \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}.
+In a large study of 118,707 patients undergoing non-bariatric general surgery, the authors reported that the risk of death was `paradoxically' low in the overweight and obese group of patients \parencite{mullen_obesity_2009}.
+The results of the present study show that $\dot{V}_{O_2}$ as is currently reported for clinical use will also be low in this very same cohort of patients.
+This obesity paradox is particularly apparent in patients with heart failure where in spite of a low $\dot{V}_{O_2}$ (corrected for total body weight), obese patients have better survival rates.
+Horwich and co-workers reported in a study of 2331 patients that body mass index was associated with a low $\dot{V}_{O_2}$Peak (ml/kg/min) independent of all other factors including age, diabetes, cardiac disease, New York Heart Association Class of cardiac failure and race \parencite{horwich_relationship_2009}.
 
-It is clear from the review presented in Chapter 1, that cardiopulmonary exercise testing is useful in predicting risk after major surgery.
-Cardiopulmonary exercise testing has become ubiquitous in the preoperative workup of complex surgical patients.
-However, the results of the present study suggest that the results especially in the obese, female patient must be interpreted with caution, especially when used to select patients who may be declined surgery based on their cardiopulmonary exercise test results.
+It is clear from the review presented in Chapter \ref{ch_intro} as well as the results presented in Chapter \ref{ch_cpet_outcomes}, that cardiopulmonary exercise testing is useful in predicting risk after major surgery.
+Cardiopulmonary exercise testing has become ubiquitous in the preoperative assessment of complex surgical patients.
+However, the present study demonstrates that cardiopulmonary exercise test parameters in the overweight/obese patient must be interpreted with caution, especially when used to select patients who may be declined surgery based on these results.
 
-\subsection{Measuring impact of Prehabilitation}
+\subsection{Measuring impact of prehabilitation}
 
-Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery and mitigating the effects of neoadjuvant oncological therapy.
+Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery, reducing complications \parencite{jones_effects_2007, dronkers_preoperative_2010, valkenet_effects_2010, gillis_prehabilitation_2014} and mitigating the effects of neoadjuvant oncological therapy\parencite{west_effect_2015}.
 Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients \parencite{west_effect_2015}.
 
-The design of such prehabilitation programs must not depend solely on body weight adjusted parameters of cardiopulmonary exercise testing when assessing the success of the interventions in these programs.
-Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with other parameters that are not affected by body composition such as O2Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.
+The success of such prehabilitation programs must not be determined solely by $\dot{V}_{O_2}$ corrected for total body weight.
+Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O^2$Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.
+The combination of nutritional assessment and advice, home-based exercise programs and weight loss will help improve aerobic capacity and eventually postoperative outcomes.
+
+Further studies are required to validate the findings presented here as well as to identify optimal scaling methods for cardiopulmonary exercise test parameters.
+This will improve the clinical utility of cardiopulmonary exercise testing in the preoperative risk assessment as well as optimisation of patients undergoing major surgery.
 

### 2015-08-20 Chapters/Introduction.tex
PrePost SIRS
- All tables and figures
- Tables ending with means have means (sd)
- Included tables have medians (IQR)
- Text includes tables and figures
- Start writing the bloody chapter now!

@@
 Khan and co-workers reported that while obesity was associated with increased incidence of intra-operative bleeding, there was no increase in complications \parencite{khan_does_2010}.
-Length of stay in hospital, overall morbidity and in-hospital mortality were not affected by BMI after pancreaticoduodenectomy \parencite{tsai_impact_2010}.
-while neither BMI nor intra-abdominal visceral fat affected postoperative outcomes in another study of patients undergoing pancreaticoduodenectomy \parencite{balentine_obesity_2011}.
+Length of stay in hospital, overall morbidity and in-hospital mortality were not affected by BMI after pancreaticoduodenectomy \parencite{tsai_impact_2010} while neither BMI nor intra-abdominal visceral fat affected postoperative outcomes in another study of patients undergoing pancreaticoduodenectomy \parencite{balentine_obesity_2011}.
 

### 2015-08-20 Chapters/Pre_Post_SIRS.tex
PrePost SIRS
- All tables and figures
- Tables ending with means have means (sd)
- Included tables have medians (IQR)
- Text includes tables and figures
- Start writing the bloody chapter now!

\chapter{An investigation into the relationship between preoperative clinico-pathological characteristics and post-operative systemic inflammatory response in patients undergoing pancreaticoduodenectomy.}
\label{ch_bodycomp}

\lhead{Chapter \ref{ch_bodycomp}. \emph{Factors affecting post-operative systemic inflammation}} % This is for the header on each page - perhaps a shortened title

\clearpage
%----------------------------------------------------------------------------------------

\section{Introduction}

\section{Methods}

\section{Results}

\input{Tables/sirs_crp}
\input{Tables/sirs_albumin}
\input{Tables/sirs_crp_pvalues}
\input{Tables/sirs_wcc}
\input{Tables/sirs_neut}
\input{Tables/sirs_wcc_neut_pvalues}

%========================CRP vs Post-op SIRS============================================
\clearpage
\begin{figure}[p]
	\caption{Relationship between preoperative CRP levels and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
	\label{fig:sirs_crp}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_crp_crp}
		\caption{Preop. CRP vs. postop. CRP}
		\label{fig:sirs_crp_crp}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_crp_alb}
		\caption{Preop. CRP vs. postop. Albumin}
		\label{fig:sirs_crp_alb}
	\end{subfigure}

	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_crp_wcc}
		\caption{Preop. CRP vs. postop. White Cell Count}
		\label{fig:sirs_crp_wcc}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_crp_neut}
		\caption{Preop. CRP vs. postop. Neutrophil Count}
		\label{fig:sirs_crp_neut}
	\end{subfigure}	
\end{figure}
%==============================================================================

%========================Albumin vs Post-op SIRS============================================
\clearpage
\begin{figure}[p]
	\caption{Relationship between preoperative Albumin levels and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
	\label{fig:sirs_alb}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_alb_crp}
		\caption{Preop. Albumin vs. postop. CRP}
		\label{fig:sirs_alb_crp}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_alb_alb}
		\caption{Preop. Albumin vs. postop. Albumin}
		\label{fig:sirs_alb_alb}
	\end{subfigure}

	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_alb_wcc}
		\caption{Preop. Albumin vs. postop. White Cell Count}
		\label{fig:sirs_alb_wcc}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_alb_neut}
		\caption{Preop. Albumin vs. postop. Neutrophil Count}
		\label{fig:sirs_alb_neut}
	\end{subfigure}	
\end{figure}
%==============================================================================

%========================Bilirubin vs Post-op SIRS============================================
\clearpage
\begin{figure}[p]
	\caption{Relationship between preoperative obstructive jaundice and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
	\label{fig:sirs_bilirubin}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_bil_crp}
		\caption{Preop. Bilirubin vs. postop. CRP}
		\label{fig:sirs_bil_crp}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_bil_alb}
		\caption{Preop. Bilirubin vs. postop. Albumin}
		\label{fig:sirs_bil_alb}
	\end{subfigure}

	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_bil_wcc}
		\caption{Preop. Bilirubin vs. postop. White Cell Count}
		\label{fig:sirs_bil_wcc}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_bil_neut}
		\caption{Preop. Bilirubin vs. postop. Neutrophil Count}
		\label{fig:sirs_bil_neut}
	\end{subfigure}	
\end{figure}
%==============================================================================
%\input{Tables/sirs_crp_bodycomp}

%========================Bilirubin vs Post-op SIRS============================================
\clearpage
\begin{figure}[p]
	\caption{Relationship between preoperative $\dot{V}_{O_2}$AT and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
	\label{fig:sirs_at}
	\centering
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_at_crp}
		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. CRP}
		\label{fig:sirs_at_crp}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_at_alb}
		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Albumin}
		\label{fig:sirs_at_alb}
	\end{subfigure}

	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_at_wcc}
		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. White Cell Count}
		\label{fig:sirs_at_wcc}
	\end{subfigure}
	\hfill
	\begin{subfigure}{0.48\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/sirs_at_neut}
		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Neutrophil Count}
		\label{fig:sirs_at_neut}
	\end{subfigure}	
\end{figure}
%==============================================================================

\section{Discussion}

### 2015-08-20 Chapters/Pre_Post_SIRS.tex
Pre_Post_SIRS - Introduction complete
- Zot backed up and updated

@@
 \section{Introduction}
+The perioperative systemic inflammatory response has a significant role in determining short-term and long-term outcomes following potentially curative surgery for a wide variety of cancers.
+Systemic inflammation both before and after major surgery has been reported to be associated with significant morbidity.
+
+An exaggerated preoperative systemic inflammatory response syndrome is associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
+The modified Glasgow Prognostic Score in particular, which uses the combination of C-reactive protein and serum albumin, has been reported to be associated with increased incidence of complications \parencite{moyes_preoperative_2009, mohri_correlation_2014, vashist_glasgow_2010}.
+Preoperative CRP levels have also been reported to be associated with increased incidence of complications including infections and renal dysfunction as well as increased in-hospital mortality after cardiac surgery \parencite{lorenzo_increased_2012, mezzomo_preoperative_2011, kim_predictive_2009, biancari_preoperative_2003, boeken_increased_1998}.
+
+Moreover, an exaggerated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+The magnitude of this postoperative inflammatory response has also been reported to be associated with the severity of the complications \parencite{mcsorley_postoperative_2015}.
+
+In patients undergoing pancreaticoduodenectomy, preoperative systemic inflammation may be affected by several factors including the presence of obstructive jaundice with or without cholangitis, preoperative biliary intervention including endoscopic retrograde cholangio-pancreatography for diagnosis or biliary drainage and in some patients acute or chronic pancreatitis either due to obstruction of the main pancreatic duct or due to other causes.
+The effect this `priming' of the immune system on postoperative outcomes is poorly understood in this cohort of patients.
+
+Chronic inflammation is a recognised feature of obesity which is increasingly common in patients undergoing major surgery for pancreatic and other gastro-intestinal cancers. 
+The impact of obesity, especially visceral obesity, on complications after pancreaticoduodenectomy remains controversial with some authors reporting that obesity is associated with increased incidence of complications \parencite{house_preoperative_2008, ramsey_body_2011} while others reporting similar outcomes in obese and non-obese patients \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}.
+
+More recently, levels of adipocytokines, inflammatory mediators produced exclusively in adipose tissue, have been reported to be associated with postoperative surgical site infections after colorectal \parencite{ortega-deballon_preoperative_2013, matsuda_preoperative_2009} and gastric cancer surgery \parencite{yamamoto_association_2013}.
+This emphasises the importance of adipose tissue metabolism and body composition in the preoperative systemic inflammatory status of the patient undergoing surgery.
+
+To our knowledge, the relationship between the preoperative systemic inflammatory response and the magnitude of the postoperative systemic inflammatory response has not been examined before. 
+While obstructive jaundice in itself has recently been reported to have no effect on postoperative complications, the impact of preoperative obstructive jaundice on postoperative systemic inflammation has not been reported before.
+Moreover, the relationship between comorbidity, body composition and aerobic capacity as measured by cardiopulmonary exercise testing and postoperative systemic inflammation has not been studied.
+
+The aim of this study was to examine the relationship between patient factors including preoperative systemic inflammation, obstructive jaundice, cardiopulmonary exercise test parameters and body composition and the magnitude of the postoperative systemic inflammation during the first week after a pancreaticoduodenectomy.
+
+%Post-operative CRP levels have been reported to be associated with the magnitude of surgery as well as to predict infectious complications after neurosurgery \parencite{al-jabi_value_2010}.
 

### 2015-08-20 Chapters/Pre_Post_SIRS.tex
Pre_PostSirs - Results nearly complete
- Results nearly done
- Body comp data yet to be finalised

@@
 \chapter{An investigation into the relationship between preoperative clinico-pathological characteristics and post-operative systemic inflammatory response in patients undergoing pancreaticoduodenectomy.}
-\label{ch_bodycomp}
+\label{ch_pre_post_sirs}
 
-\lhead{Chapter \ref{ch_bodycomp}. \emph{Factors affecting post-operative systemic inflammation}} % This is for the header on each page - perhaps a shortened title
+\lhead{Chapter \ref{ch_pre_post_sirs}. \emph{Factors affecting post-operative systemic inflammation}} % This is for the header on each page - perhaps a shortened title
 
@@
 
-In patients undergoing pancreaticoduodenectomy, preoperative systemic inflammation may be affected by several factors including the presence of obstructive jaundice with or without cholangitis, preoperative biliary intervention including endoscopic retrograde cholangio-pancreatography for diagnosis or biliary drainage and in some patients acute or chronic pancreatitis either due to obstruction of the main pancreatic duct or due to other causes.
+In patients undergoing pancreaticoduodenectomy, preoperative systemic inflammation may be affected by several factors. 
+These include the presence of obstructive jaundice with or without cholangitis, preoperative biliary intervention including endoscopic retrograde cholangio-pancreatography for diagnosis or biliary drainage and in some patients acute or chronic pancreatitis either due to obstruction of the main pancreatic duct or due to other causes.
 The effect this `priming' of the immune system on postoperative outcomes is poorly understood in this cohort of patients.
@@
 \section{Methods}
+Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
+Patients who underwent only a trial dissection or palliative surgical bypass for unresectable disease during this period were excluded.
+
+Routine preoperative blood tests including full blood count, liver function tests and serum C-reactive protein were performed in all patients on the day before surgery. 
+These blood tests were also repeated every day for at least the first postoperative week. 
+These results were collected from the hospital laboratory database using an automated MS Access application as outlined in Appendix \ref{AppendixAccessDatabase}. 
+The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on p\pageref{table:mGPS} in Chapter \ref{ch_intro}. 
+The neutrophil-lymphocyte ratio was calculated by dividing the preoperative neutrophil count by the preoperative lymphocyte count and a threshold of 5.0 was used to dichotomise this variable. 
+Standard thresholds were used to categorise other biochemical parameters. 
+Obstructive jaundice was defined as serum bilirubin $>$35 $\mu$mol/L while severe obstructive jaundice was defined as serum bilirubin $>$250 $\mu$mol/L.
+
+Body Mass Index (BMI) was categorised using the World Health Organisation thresholds as shown in Table \ref{table:bmi_who} on p\ref{table:bmi_who} in Chapter \ref{ch_intro}. 
+Only one patient had a BMI less than 18.5 kg/m$2$ and was excluded from analysis involving BMI.
+The Scottish Index of Multiple Deprivation (SIMD) Quintile Scores were calculated from the post-code of the patient's primary residence and dichotomised into two groups with scores of 1-3 and 4-5 respectively.
+The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described in Table \todo{Create a table for PPS in introduction and reference it wherever necessary}.
+
+In patients who underwent cardiopulmonary exercise testing, $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak were compared against postoperative systemic inflammation. 
+$\dot{V}_{O_2}$AT was dichotomised using a value of 10 mls/kg/min while $\dot{V}_{O_2}$Peak was dichotomised using a value of 16 mls/kg/min. 
+Cardiopulmonary exercise testing methodology has been described in Section \ref{sec:cpx_method}.
+
+Body composition was calculated in a subset of these patients using preoperative computed tomography of the abdomen. 
+The methodology used in the calculation of the individual components of body composition including visceral fat, subcutaneous fat and skeletal muscle is described in detail in Section \ref{sec:bodycomp_calculation}.
+%Check this with Donny
+Continuous data were converted into categorical data using tertiles.
+
+\subsection{Statistics}
+
+Continuous variables are reported as median (inter-quartile range).
+Non-parametric tests were used to compare postoperative inflammatory markers (continuous data) with preoperative clinico-pathological characteristics (categorical data). 
+Mann-Whitney U test was used when two categories were present and Kruskal-Wallis test was used when more than two categories were present.
+
+Line-plots were created comparing the trend of inflammatory markers during the first postoperative week with preoperative systemic inflammation, obstructive jaundice and $\dot{V}_{O_2}$AT with error bars representing 95\% confidence intervals.
+
+SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. Effects were considered significant at $\alpha \leq0.05$.
 
 \section{Results}
+
+Pancreaticoduodenectomy was performed in 188 patients (126 male, 67\%) during the study period.
+Preoperative C-reactive protein was elevated in 70 (37.6\%) patients while just over half the patients had a low preoperative serum albumin (96, 51.1\%). 
+The modified Glasgow Prognostic Score revealed normal preoperative systemic inflammatory status (mGPS = 0) in 116 (62.4\%) patients, mildly systemic inflammation in 16 (8.6\%) and severe inflammation in 54 (29.0\%) patients.
+Obstructive jaundice was present in 44 (23.4\%) patients and severe obstructive jaundice was present in 31 (16.5\%) of patients. 
+BMI data was available in 167 patients. More than half of these patients were overweight or obese with a BMI $>$ 25 in 87 (52.1\%) patients.
+
+Cardiopulmonary exercise testing was performed in 130 patients. 
+$\dot{V}_{O_2}$AT could not be estimated in one patient.
+$\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
+All the variables necessary for calculation of the POSSUM Physiology Score, a composite score of comorbidity and preoperative biochemistry, were available in 180 patients and it was elevated in 87 (48.3\%) patients.
+
+The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (median, IQR in Table \ref{table:sirs_crp}, p-values in Table \ref{table:sirs_crp_pvalues}), postoperative serum albumin (median, IQR in Table \ref{table:sirs_alb}, p-values in Table \ref{table:sirs_alb_pvalues}), postoperative total white cell count (median, IQR in Table \ref{table:sirs_wcc}, p-values in Table \ref{table:sirs_wcc_pvalues}) and postoperative differential neutrophil count (median, IQR in Table \ref{table:sirs_neut}, p-values in Table \ref{table:sirs_neut_pvalues}) are presented.
+
+The relationship between preoperative CRP and postoperative systemic inflammation is also depicted graphically in Fig. \ref{fig:sirs_crp}. 
+The median CRP levels on the day of surgery and the first postoperative day (POD) were significantly higher in patients who had an elevated preoperative CRP. 
+However, this association was not present after the first postoperative day.
+Elevated preoperative CRP was associated with a persistently lower serum albumin level for the entire duration of the first postoperative week (p$<$0.001 to p$<$0.002).
+It was also associated with a higher total white cell count (p$<$0.01) and higher differential neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
+
+Preoperative hypoalbuminemia on the other hand was associated with lower median postoperative CRP levels starting with POD 3. 
+The median difference in postoperative CRP increased from 30 mg/dl on POD 3 (p$<$0.045) to 53 mg/dl by POD 6 (p=0.002) with patients with hypoalbuminemia having a lower CRP. 
+This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
+Preoperative hypoalbuminemia persisted postoperatively throughout the first week (p$<$0.001, Fig. \ref{fig:sirs_alb_alb}).
+There was no relationship between preoperative serum albumin levels and postoperative white cell count (Fig. \ref{fig:sirs_alb_wcc}) or neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
+
+Obstructive jaundice was associated with significant differences in the postoperative systemic inflammatory response. 
+Postoperative serum albumin levels were lower in jaundiced patients and with the severely jaundiced patients having the lowest levels (p$<$0.001, Fig. \ref{fig:sirs_bil_alb}). 
+There were no significant changes in the trends of albumin levels and it is likely that preoperative hypoalbuminemia persisted postoperatively in these patients.
+
+However, the trends in postoperative CRP were significantly different between the jaundiced and non-jaundiced patients. 
+Obstructive jaundice was associated with a significantly lower peak CRP on POD 2 (p=0.001) and this persisted until POD 6 (p=0.005, Table \ref{table:sirs_crp}, \ref{table:sirs_crp_pvalues}).
+Median postoperative CRP had an inverse relationship with severity of preoperative obstructive jaundice (Fig. \ref{fig:sirs_bil_crp}).
+Obstructive jaundice was also associated with a delayed rise in the white cell count and neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although these relationships were less significant (Fig.
+\ref{fig:sirs_bil_wcc} and Fig.
+\ref{fig:sirs_bil_neut}).
+
+There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP, white cell count or neutrophil count. 
+Serum albumin levels were lower in patients with $\dot{V}_{O_2}$AT$<$10 mls/kg/min (Fig. \ref{fig:sirs_at}).
+
+There was no significant and persistent relationship between postoperative CRP, white cell count and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
+However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Table \ref{table:sirs_alb} and Table \ref{table:sirs_alb_pvalues}).
 

### 2015-08-20 Appendices/AppendixAccessDatabase.tex
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database

\chapter{Visual Basic for Applications For Data Collection} % Main appendix title
\label{AppendixAccessDatabase} % For referencing this appendix elsewhere, use \ref{AppendixA}
\lhead{Appendix \ref{AppendixAccessDatabase}. \emph{MS Access VBA For Data Collection}} % This is for the header on each page - perhaps a shortened title

\setstretch{1.0}
%------------------------------------------

### 2015-08-20 Appendices/AppendixExcelMacros.tex
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database

@@
 
-\lhead{Appendix B. \emph{Excel VBA For CPET Analysis}} % This is for the header on each page - perhaps a shortened title
+\lhead{Appendix \ref{AppendixExcelMacros}. \emph{Excel VBA For CPET Analysis}} % This is for the header on each page - perhaps a shortened title
 
 \setstretch{1.0}
+
+%------------------------------------------
+
 The following VBA code was written by me to facilitate detailed analysis of large volumes of raw CPET data that would not have been otherwise available from the final report generated for clinical use.

### 2015-08-20 Chapters/BodyComp.tex
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database

@@
 \subsection{Calculation of body composition}
+\label{sec:bodycomp_calculation}
 Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods \parencite{bredella_comparison_2010,shen_total_2004}.

### 2015-08-20 Chapters/CRP_Complications.tex
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database

@@
 Data are presented as median (inter-quartile range, IQR) unless otherwise specified.
-Line plots with error bars depicting 95\% confidence intervals are used to depict the trends in CRP over time across different patient groups.
+Line plots with error bars depicting 95\% confidence intervals were used to depict the trends in CRP over time across different patient groups.
 
@@
 This study shows that CRP on the third postoperative day is associated with infective complications in patients undergoing pancreaticoduodenectomy who do not develop a POPF but not in patients with a POPF.
+
+\todo{straatman_predictive_2015 - This is hot-off the press - reference this}

### 2015-08-20 Chapters/Introduction.tex
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database

@@
 
-The importance of visceral fat in the incidence and severity of postoperative complications is further emphasised by the finding that excess visceral fat is associated with the the occurence of pancreatic fistula after radical total gastrectomy for gastric cancer.
-Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese patients undergoing surgery compared to non-obese pateints.
+The importance of visceral fat in the incidence and severity of postoperative complications is further emphasised by the finding that excess visceral fat is associated with the the occurrence of pancreatic fistula after radical total gastrectomy for gastric cancer.
+Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese patients undergoing surgery compared to non-obese patients.
 However, there was no difference in operative time or length of stay in hospital.

### 2015-08-20 Chapters/Pre_Post_SIRS.tex
Pre_Post_SIRS - Donny comments
- Remove all references to white cell count
- Remove tables with WCC
- Place p-value tables immediately after the results tables
- Remove white cell count graphs

@@
 
-An exaggerated preoperative systemic inflammatory response syndrome is associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
+For example, an elevated preoperative systemic inflammatory response is associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
 The modified Glasgow Prognostic Score in particular, which uses the combination of C-reactive protein and serum albumin, has been reported to be associated with increased incidence of complications \parencite{moyes_preoperative_2009, mohri_correlation_2014, vashist_glasgow_2010}.
@@
 
-Moreover, an exaggerated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+%Include some recent reviews pp singh, bjs as well as check Donny's email.
 The magnitude of this postoperative inflammatory response has also been reported to be associated with the severity of the complications \parencite{mcsorley_postoperative_2015}.
@@
 Chronic inflammation is a recognised feature of obesity which is increasingly common in patients undergoing major surgery for pancreatic and other gastro-intestinal cancers. 
-The impact of obesity, especially visceral obesity, on complications after pancreaticoduodenectomy remains controversial with some authors reporting that obesity is associated with increased incidence of complications \parencite{house_preoperative_2008, ramsey_body_2011} while others reporting similar outcomes in obese and non-obese patients \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}.
+The impact of obesity, especially visceral obesity, on complications after pancreaticoduodenectomy remains controversial with some authors reporting that obesity was associated with increased incidence of complications \parencite{house_preoperative_2008, ramsey_body_2011} while others reporting similar outcomes in obese and non-obese patients \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}.
 
 More recently, levels of adipocytokines, inflammatory mediators produced exclusively in adipose tissue, have been reported to be associated with postoperative surgical site infections after colorectal \parencite{ortega-deballon_preoperative_2013, matsuda_preoperative_2009} and gastric cancer surgery \parencite{yamamoto_association_2013}.
-This emphasises the importance of adipose tissue metabolism and body composition in the preoperative systemic inflammatory status of the patient undergoing surgery.
+This emphasises the importance of adipose tissue metabolism and body composition in the preoperative systemic inflammatory status of surgical patients.
 
 To our knowledge, the relationship between the preoperative systemic inflammatory response and the magnitude of the postoperative systemic inflammatory response has not been examined before. 
-While obstructive jaundice in itself has recently been reported to have no effect on postoperative complications, the impact of preoperative obstructive jaundice on postoperative systemic inflammation has not been reported before.
+While obstructive jaundice in itself has recently been reported to have no effect on postoperative complications, the impact of preoperative obstructive jaundice on postoperative systemic inflammation has not been previously reported.
 Moreover, the relationship between comorbidity, body composition and aerobic capacity as measured by cardiopulmonary exercise testing and postoperative systemic inflammation has not been studied.
@@
 
-The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (median, IQR in Table \ref{table:sirs_crp}, p-values in Table \ref{table:sirs_crp_pvalues}), postoperative serum albumin (median, IQR in Table \ref{table:sirs_alb}, p-values in Table \ref{table:sirs_alb_pvalues}), postoperative total white cell count (median, IQR in Table \ref{table:sirs_wcc}, p-values in Table \ref{table:sirs_wcc_pvalues}) and postoperative differential neutrophil count (median, IQR in Table \ref{table:sirs_neut}, p-values in Table \ref{table:sirs_neut_pvalues}) are presented.
+The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (median, IQR in Table \ref{table:sirs_crp}, p-values in Table \ref{table:sirs_crp_pvalues}), postoperative serum albumin (median, IQR in Table \ref{table:sirs_alb}, p-values in Table \ref{table:sirs_alb_pvalues}) and postoperative neutrophil count (median, IQR in Table \ref{table:sirs_neut}, p-values in Table \ref{table:sirs_neut_pvalues}) are presented.
 
@@
 Elevated preoperative CRP was associated with a persistently lower serum albumin level for the entire duration of the first postoperative week (p$<$0.001 to p$<$0.002).
-It was also associated with a higher total white cell count (p$<$0.01) and higher differential neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
+It was also associated with a higher neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
 
@@
 Preoperative hypoalbuminemia persisted postoperatively throughout the first week (p$<$0.001, Fig. \ref{fig:sirs_alb_alb}).
-There was no relationship between preoperative serum albumin levels and postoperative white cell count (Fig. \ref{fig:sirs_alb_wcc}) or neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
+There was no relationship between preoperative serum albumin levels and postoperative neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
 
@@
 Median postoperative CRP had an inverse relationship with severity of preoperative obstructive jaundice (Fig. \ref{fig:sirs_bil_crp}).
-Obstructive jaundice was also associated with a delayed rise in the white cell count and neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although these relationships were less significant (Fig.
-\ref{fig:sirs_bil_wcc} and Fig.
+Obstructive jaundice was also associated with a delayed rise in the neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although this relationship was less significant (Fig.
 \ref{fig:sirs_bil_neut}).
 
-There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP, white cell count or neutrophil count. 
+There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP or neutrophil count. 
 Serum albumin levels were lower in patients with $\dot{V}_{O_2}$AT$<$10 mls/kg/min (Fig. \ref{fig:sirs_at}).
 
-There was no significant and persistent relationship between postoperative CRP, white cell count and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
+There was no significant and persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
 However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Table \ref{table:sirs_alb} and Table \ref{table:sirs_alb_pvalues}).
@@
 \input{Tables/sirs_crp}
+\input{Tables/sirs_crp_pvalues}
+
 \input{Tables/sirs_albumin}
-\input{Tables/sirs_crp_pvalues}
-\input{Tables/sirs_wcc}
+\input{Tables/sirs_alb_pvalues}
+
 \input{Tables/sirs_neut}
-\input{Tables/sirs_wcc_neut_pvalues}
+\input{Tables/sirs_neut_pvalues}
 
@@
 
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_crp_wcc}
-		\caption{Preop. CRP vs. postop. White Cell Count}
-		\label{fig:sirs_crp_wcc}
-	\end{subfigure}
-	\hfill
+	\vspace{1cm}
+
 	\begin{subfigure}{0.48\textwidth}
@@
 
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_alb_wcc}
-		\caption{Preop. Albumin vs. postop. White Cell Count}
-		\label{fig:sirs_alb_wcc}
-	\end{subfigure}
-	\hfill
+	\vspace{1cm}
+
 	\begin{subfigure}{0.48\textwidth}
@@
 
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_bil_wcc}
-		\caption{Preop. Bilirubin vs. postop. White Cell Count}
-		\label{fig:sirs_bil_wcc}
-	\end{subfigure}
-	\hfill
+	\vspace{1cm}
+
 	\begin{subfigure}{0.48\textwidth}
@@
 
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_at_wcc}
-		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. White Cell Count}
-		\label{fig:sirs_at_wcc}
-	\end{subfigure}
-	\hfill
+	\vspace{1cm}
+
 	\begin{subfigure}{0.48\textwidth}
@@
 \section{Discussion}
+
+The results of the present study demonstrate that preoperative systemic inflammation is associated with the magnitude of the postoperative systemic inflammatory response  in patients undergoing pancreaticoduodenectomy.
+The results also appear to show that preoperative obstructive jaundice is associated with a dampened systemic inflammatory response as measured by CRP levels.
+However, neither aerobic capacity or the POSSUM Physiology Score were related to postoperative systemic inflammation.
+
+Taken together, these results appear to suggest that the postoperative systemic inflammatory response may not simply be due to the effect of surgery and its sequelae

### 2015-08-21 Chapters/Pre_Post_SIRS.tex
Chapter  Pre_Post_SIRS - Complete
- Complete discussion
- Complete bodycomp tables
- Update zot.bib

@@
 
-\clearpage
 %----------------------------------------------------------------------------------------
@@
 
-Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{singh_systematic_2014, platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
 %Include some recent reviews pp singh, bjs as well as check Donny's email.
@@
 More recently, levels of adipocytokines, inflammatory mediators produced exclusively in adipose tissue, have been reported to be associated with postoperative surgical site infections after colorectal \parencite{ortega-deballon_preoperative_2013, matsuda_preoperative_2009} and gastric cancer surgery \parencite{yamamoto_association_2013}.
+Sarcopenia has been reported to be associated with  elevated postoperative systemic inflammation after colorectal surgery \parencite{reisinger_sarcopenia_2015}.
 This emphasises the importance of adipose tissue metabolism and body composition in the preoperative systemic inflammatory status of surgical patients.
@@
 BMI data was available in 167 patients. More than half of these patients were overweight or obese with a BMI $>$ 25 in 87 (52.1\%) patients.
-
-Cardiopulmonary exercise testing was performed in 130 patients. 
-$\dot{V}_{O_2}$AT could not be estimated in one patient.
-$\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
 All the variables necessary for calculation of the POSSUM Physiology Score, a composite score of comorbidity and preoperative biochemistry, were available in 180 patients and it was elevated in 87 (48.3\%) patients.
 
-The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (median, IQR in Table \ref{table:sirs_crp}, p-values in Table \ref{table:sirs_crp_pvalues}), postoperative serum albumin (median, IQR in Table \ref{table:sirs_alb}, p-values in Table \ref{table:sirs_alb_pvalues}) and postoperative neutrophil count (median, IQR in Table \ref{table:sirs_neut}, p-values in Table \ref{table:sirs_neut_pvalues}) are presented.
+The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (Tables \ref{table:sirs_crp} and \ref{table:sirs_crp_pvalues}), postoperative serum albumin (Tables \ref{table:sirs_alb} and \ref{table:sirs_alb_pvalues}) and postoperative neutrophil count (Tables \ref{table:sirs_neut} and \ref{table:sirs_neut_pvalues}) are presented.
 
@@
 
+The trend in postoperative CRP appeared to differ depending on whether preoperative CRP and albumin were independently deranged or not.
+Fig. \ref{fig:sirs_crp_crp_alb} and Table \ref{table:sirs_crp_with_alb} show the results of this analysis.
+When both preoperative CRP and preoperative albumin were normal, the trend in postoperative CRP is shown by the solid line (Line 0 in Fig. \ref{fig:sirs_crp_crp_alb}). 
+When preoperative CRP was elevated in the presence of a normal preoperative albumin, postoperative CRP levels were significantly elevated and remained elevated during the first postoperative week (Line 1 in Fig.
+\ref{fig:sirs_crp_crp_alb}).
+However, hypoalbuminemia was associated with lower postoperative CRP levels regardless of whether preoperative CRP was normal (Line 2 in Fig.
+\ref{fig:sirs_crp_crp_alb}) or elevated (Line 3 in Fig.
+\ref{fig:sirs_crp_crp_alb}).
+
 Obstructive jaundice was associated with significant differences in the postoperative systemic inflammatory response. 
@@
 
+Cardiopulmonary exercise testing was performed in 130 patients. 
+$\dot{V}_{O_2}$AT could not be estimated in one patient.
+$\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
 There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP or neutrophil count. 
@@
 
-There was no significant and persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
-However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Table \ref{table:sirs_alb} and Table \ref{table:sirs_alb_pvalues}).
+There was no significant, persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
+However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Table \ref{table:sirs_alb}, \ref{table:sirs_alb_pvalues}).
+
+Body composition data was available for 90 patients.
+Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, 128.9 cm$^2$, 119.0-150.8)and high (n=30, 241.2 cm$^2$, 221.9-313.1). 
+Subcutaneous adipose tissue area was divided into tertiles as low (n=30, median 94.1 cm$^2$, IQR 63.2-102.5), moderate (n=30, median 152.5 cm$^2$, IQR 141.3-175.3) and high (n=30, median 253.9 cm$^2$, IQR 203.2-290.3).
+Skeletal muscle area was also divided into tertiles as low (n=30, median 96.7 cm$^2$, IQR 91.9-102.5), moderate (n=30, median 121.7 cm$^2$, IQR 116.4-130.9) and high (n=30, median 153.0 cm$^2$, IQR 145.3-167.6).
+
+The relationship between body composition and postoperative CRP, albumin and neutrophil count is shown in Tables \ref{table:sirs_bodycomp_crp}, \ref{table:sirs_bodycomp_alb} and \ref{table:sirs_bodycomp_neut} respectively.
+Postoperative CRP and neutrophil count were not related to any of the components of body composition aside from a weak linear association between neutrophil count on POD 2 and visceral fat area (p=0.046).
+However, lower preoperative skeletal muscle area was associated with lower postoperative serum albumin on days 1 to 5 (all p<0.05).
 
@@
 
+\clearpage
 %========================CRP vs Post-op SIRS============================================
-\clearpage
+
 \begin{figure}[p]
@@
 %========================Albumin vs Post-op SIRS============================================
-\clearpage
+
 \begin{figure}[p]
@@
 
+%==============================================================================
+\clearpage
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=\textwidth]{Figures/sirs_crp_crp_alb}
+	\caption{Relationship between preoperative CRP and postoperative CRP in the presence or absence of preoperative hypoalbuminemia.}
+	\label{fig:sirs_crp_crp_alb}
+\end{figure}
+\input{Tables/sirs_crp_with_alb}
+\clearpage
+%==============================================================================
+
 %========================Bilirubin vs Post-op SIRS============================================
-\clearpage
+
 \begin{figure}[p]
@@
 %==============================================================================
-%\input{Tables/sirs_crp_bodycomp}
 
-%========================Bilirubin vs Post-op SIRS============================================
-\clearpage
+%========================AT vs Post-op SIRS============================================
+
 \begin{figure}[p]
@@
 
+\input{Tables/sirs_bodycomp_crp}
+\input{Tables/sirs_bodycomp_alb}
+\input{Tables/sirs_bodycomp_neut}
+
+\clearpage
 \section{Discussion}
@@
 The results also appear to show that preoperative obstructive jaundice is associated with a dampened systemic inflammatory response as measured by CRP levels. 
-However, neither aerobic capacity or the POSSUM Physiology Score were related to postoperative systemic inflammation.
+%However, neither aerobic capacity or the POSSUM Physiology Score were related to postoperative systemic inflammation. 
+Taken together, these results appear to suggest that the preoperative inflammatory status of the patient plays an important role in modulating postoperative systemic inflammation which may not simply be due to the effect of surgery and its sequelae.
 
-Taken together, these results appear to suggest that the postoperative systemic inflammatory response may not simply be due to the effect of surgery and its sequelae
+The role of postoperative CRP in predicting complications after surgery has been well established. 
+A recent meta-analysis of six studies involving 1832 patients \parencite{warschkow_safe_2012} reported that CRP level less than 135 mg/dl on the fourth postoperative day had a high negative predictive value of 89\% for infectious complications.
+Postoperative CRP has been shown to predict complications after pancreaticoduodenectomy by us (\ref{ch_crp_comp}) as well as other authors (\parencite{welsch_persisting_2008, hiyoshi_usefulness_2013, kosaka_multivariate_2014}).
+The role of preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score or the neutrophil-lymphocyte ratio has also been reported to be adversely affect long-term survival in patients undergoing potentially curative surgery in over a hundred studies involving a variety of gastrointestinal and non-gastrointestinal cancers.
+
+However, preoperative systemic inflammation as measured by elevated C-reactive protein or an abnormal modified Glasgow Prognostic Score is increasingly recognised as an independent risk factor for postoperative infectious complications \parencite{mohri_correlation_2014, kubo_elevated_2013, moyes_preoperative_2009}.
+
+We hypothesized that preoperative systemic inflammation may adversely affect postoperative immuno-modulation resulting in an abnormal response in patients whose immune systems have been `primed' before the surgical insult.
+This may affect the healing processes as well as the patient's ability to clear endotoxins and may predispose the patient to infective and other complications.
+Our results appear to support this hypothesis.
+
+In this study, raised preoperative CRP in the absence of hypoalbuminemia was associated with elevated postoperative systemic inflammation.
+However, preoperative hypoalbuminemia was associated with significantly lower levels of postoperative CRP.
+These apparently opposite effects of preoperative CRP and albumin on the postoperative systemic inflammatory response is further evidenced by the fact that the effect of preoperative CRP on postoperative inflammation is attenuated by the presence of preoperative hypoalbuminemia.
+These results appear to suggest when hypoalbuminemia is present as part of preoperative systemic inflammation, not only is the exaggerated postoperative response lost but the response appears to be dampened.
+
+%Hypoalbuminemia
+The impact of hypoalbuminemia on the acute phase response was noted by Christou and co-workers who performed a delayed type hypersensitivity (DTH) skin test in addition to measuring acute phase markers including CRP, albumin, white cell count, haemoglobin and immunoglobulins in 245 patients prior to undergoing gastrointestinal surgery \parencite{christou_estimating_1989}.
+They noted that hypoalbuminemia and anergy to the DTH skin test were the only variables independently associated with postoperative sepsis-related mortality.
+Moreover, both low albumin and elevated CRP were significantly associated with cutaneous anergy.
+Cutaneous anergy is one of the hallmarks of the compensatory anti-inflammatory response (CARS) \parencite{ward_compensatory_2008}.
+
+Preoperative hypoalbuminemia has been reported to be a risk factor for postoperative complications independent of both inflammation and nutritional status \parencite{gibbs_preoperative_1999, don_poor_2004, hennessey_preoperative_2010}.
+Albumin is a negative acute phase protein and elevated CRP levels are often associated with low serum albumin levels \parencite{margarson_serum_1998}. 
+Hypoalbuminemia in patients scheduled to undergo pancreaticoduodenectomy is likely to be multi-factorial in origin. 
+Systemic inflammation, obstructive jaundice and the associated complex physiological, biochemical and immunological abnormalities, malnutrition, cancer cachexia, changes in body composition including sarcopenic obesity may all contribute to low serum albumin levels.
+Contrary to our findings, hypoalbuminemia and poor preoperative nutritional status were found to be associated with higher postoperative levels of pro-inflammatory cytokines and CRP in surgical patients \parencite{nakamura_influence_1999}.
+The association of hypoalbuminemia with obstructive jaundice may have influenced our findings.
+
+%Obstructive jaundice
+In patients with pancreatic cancer, malignant obstructive jaundice further complicates the preoperative inflammatory status and the behaviour of the immune system \parencite{nehez_compromise_2002}. 
+We reported in Chapter \ref{ch_cpet_jaundice} that obstructive jaundice was associated with elevated preoperative CRP and low serum albumin on the day before surgery with a linear relationship with severity of jaundice (Table \ref{table:cpet_oj_bloods} on p\pageref{table:cpet_oj_bloods}).
+Hypoalbuminemia in jaundiced patients persisted after surgery in the present study.
+However, the CRP response appears to have reversed in jaundiced patients.
+While preoperative CRP was elevated in jaundiced patients, CRP levels after surgery were significantly lower in jaundiced patients with a negative linear relationship with severity of jaundice.
+
+While preoperative biliary intervention and resolution of jaundice may be offered as an explanation for the priming of the immune system in the non-jaundiced patients, this does not explain the initially low preoperative CRP in these patients.
+Padillo and co-workers reported that malignant obstructive jaundice was associated with elevated CRP levels which only improved transiently following biliary drainage.
+CRP returned to pre-drainage levels presumably due to bacterial colonisation of the biliary tree  \parencite{padillo_effect_2002, padillo_cytokines_2001}.
+It would therefore appear that obstructive jaundice adversely affects the ability to mount an adequate postoperative inflammatory response and this may be further compounded by the presence of hypoalbuminemia in these patients.
+
+Immunosuppression and immune dysregulation have been recognised in patients with obstructive jaundice\parencite{scott-conner_pathophysiology_1994}.
+Our results are similar to those reported by Mackenzie and Woodhouse who studied the CRP response to bacteraemia in 126 patients critically ill patients with or without liver dysfunction \parencite{mackenzie_c-reactive_2006}.
+Liver dysfunction was defined as serum bilirubin $>$ 20 $\mu$mol/l and prothrombin time longer than 18 seconds. 
+The CRP response to bacteraemia was significantly lower in patients with liver dysfunction than in those with normal liver function (146 vs. 103 mg/dl, p=0.03). 
+These findings have been corroborated by other authors in patients with cirrhosis\parencite{pieri_c-reactive_2014, janum_c-reactive_2011}.
+
+%There appears to be some evidence from animal studies that the immunosuppression due to obstructive jaundice may be a therapeutic target \parencite{padillo_anti-tnf-alpha_2007}.
+
+%Aerobic capacity/comorbidity
+Although CRP on POD 0 was significantly different between patients with a normal $\dot{V}_{O_2}$AT/$\dot{V}_{O_2}$Peak and those with low aerobic capacity, this difference was not present on other days.
+The lack of association between aerobic capacity, POSSUM Physiology Score and postoperative CRP in this study may have been due to the dominant effect of preoperative systemic inflammation and obstructive jaundice on the postoperative inflammatory response or simply due to any influence of aerobic capacity on postoperative inflammation.
+However, the association of postoperative hypoalbuminemia with low aerobic capacity, POSSUM Physiology Score and anaemia appears to suggest a link between comorbidity and systemic inflammation and/or nutritional status.
+Bernal and co-workers studied 399 patients undergoing liver transplantation who had cardiopulmonary exercise testing as part of their preoperative work-up\parencite{bernal_aerobic_2014} and reported that hypoalbuminemia was associated with both low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak.
+
+%Body Composition
+The association between hypoalbuminemia and low aerobic capacity may be explained by lower skeletal muscle and greater subcutaneous fat in patients with persistent postoperative hypoalbuminemia.
+Alternate explanations 
+Richards and co-workers reported that elevated preoperative systemic inflammation was related to low skeletal muscle index \parencite{hassen_preoperative_2007}.
+Hassen and co-workers reported that fat-free mass and skeletal muscle were inversely related to the severity of the postoperative inflammatory response following endo-vascular surgery for abdominal aortic aneurysm \parencite{hassen_preoperative_2007}.
+Elevated systemic inflammation is related to both obesity as well as sarcopenia. 
+`Sarcopenic obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival in patients with pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012}.
+
+\subsection{Limitations of the study}
+
+The results in this study report the relationship between preoperative clinico-pathological factors and postoperative systemic inflammation in patients undergoing pancreaticoduodenectomy. 
+The impact of complications on the systemic inflammatory response was not studied here but has been reported separately in Chapter \ref{ch_crp_comp}. 
+Moreover, it is not clear how the interaction between the various preoperative factors, especially obstructive jaundice and systemic inflammation affect postoperative inflammation or which variables are independently associated with the postoperative inflammatory response.
+Postoperative systemic inflammation is likely to be determined by multiple factors including preoperative factors, intra-operative factors including magnitude of the surgical insult, blood loss, anaesthetic factors as well as post-operative complications.
+Further study of the inter-play between these factors extending from the preoperative period into the early postoperative period will help elucidate the complex interaction between preoperative inflammation, aerobic capacity and obstructive jaundice on postoperative outcomes.
+
+\subsection{Conclusion}
+The results presented in this study emphasise the importance of the preoperative inflammatory status of the patient in influencing postoperative systemic inflammation.
+The results also demonstrate the immunosuppressive effects of obstructive jaundice in these patients.
+
+%Limitations
+%no multivariate analysys
+%potential confounding factors
+%impact of surgical factors, complicaitons, etc

### 2015-08-21 Chapters/CPET_Outcomes.tex
Review 1
- Every sentence on a new line
- Fits my OCD

@@
 \section{Introduction}
-Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years.
-\parencite{cancerresearchuk_cancer_2014}.
+Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years \parencite{cancerresearchuk_cancer_2014}.
 The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}. 
@@
 The decision to operate on these patients depends not only on preoperative tumour stage but also on patient factors \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}. 
-Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery.
-\parencite{mann_review_2010, mayo_management_2012}.
+Patient factors, in particular those that affect fitness, are also important in determining short term outcome in those that do undergo potentially curative surgery \parencite{mann_review_2010, mayo_management_2012}.
 However, major pancreatic surgery is associated with significant morbidity and mortality and patients who have postoperative complications are less likely to get adjuvant therapy \parencite{teh_patient_2009}.
@@
 
+\subsection{Aims}
 The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.
@@
 
-\section{Methods}
+\section{Patients and Methods}
 Patients who underwent pancreaticoduodenectomy or total pancreatectomy for pancreatic head lesions between August 2008, when cardiopulmonary exercise testing was first used for fitness assessment at our hospital, and January 2012 were considered for this retrospective study.
@@
 Table \ref{table:cpet_outcomes_table1} shows the clinico-pathological characteristics of patients included in the study compared to the excluded patients. 
-The median age in the study cohort was higher than in the excluded cohort (66 vs.
-54 years, p=0.001).
+The median age in the study cohort was higher than in the excluded cohort (66 vs. 54 years, p=0.001). 
 However, there was no difference in gender, body mass index, preoperative biliary drainage, jaundice at the time of surgery, modified Glasgow Prognostic Score, POSSUM physiology score, preoperative blood tests including haemoglobin and liver function tests and length of critical care/hospital stay.

### 2015-08-21 main.tex
PrePost SIRS
- All tables and figures
- Tables ending with means have means (sd)
- Included tables have medians (IQR)
- Text includes tables and figures
- Start writing the bloody chapter now!
Minor changes - Minor edits to Bodycomp, intro, main and appendix - New appendix file for Access database
Review 1
- Every sentence on a new line
- Fits my OCD

--- 2015-08-09 main.tex
@@
 University Department of Surgery, Glasgow Royal Infirmary
+
+%Mr. Euan J Dickson,
+%West of Scotland Pancreatic Unit, Glasgow Royal Infirmary
 }
@@
 \input{Chapters/BodyComp}			%4
-\input{Chapters/CPET_Survival}		%5
+%\input{Chapters/CPET_Survival}		%5
+\input{Chapters/Pre_Post_SIRS}
 \input{Chapters/CRP_Complications}	%6
@@
 \input{Appendices/AppendixCPETRawData}
-%\input{Appendices/AppendixC}
+\input{Appendices/AppendixAccessDatabase}
 

### 2015-08-21 Chapters/CPET_Outcomes.tex
Review 1 - CPET_Outcomes
- Complete

@@
 
-\subsection{Aims}
+\subsection{Aim}
 The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.
@@
 \section{Patients and Methods}
+
+\subsection{Patients}
 Patients who underwent pancreaticoduodenectomy or total pancreatectomy for pancreatic head lesions between August 2008, when cardiopulmonary exercise testing was first used for fitness assessment at our hospital, and January 2012 were considered for this retrospective study.
@@
 
+\subsection{Preoperative data}
 Data on patient demographics, comorbidity including cardiovascular and respiratory disease, preoperative blood tests, chest x-ray and cardiopulmonary exercise tests were collected from prospectively maintained databases (march 2009 - January 2012) and case note review (August 2008 - March 2009).
@@
 
+\subsection{Cardiopulmonary exercise test}
 Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
@@
 The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. 
-Low VO$_2$AT was defined as oxygen consumption less than 10ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
+Low VO$_2$AT was defined as oxygen consumption less than 10 ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
+A detailed description of the cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
 
-The decision to operate was based on overall preoperative evaluation of the patient’s comorbid conditions and performance status and not exclusively on the result of cardiopulmonary exercise testing. 
+\subsection{Perioperative care}
+The decision to operate was based on overall preoperative evaluation of the patient's comorbid conditions and performance status and not exclusively on the result of cardiopulmonary exercise testing. 
 Whilst the results of cardiopulmonary exercise tests were available to the clinicians before surgery, no specific changes were made to perioperative management based exclusively on these results. 
@@
 
+\subsection{Outcome measures}
 Postoperative adverse events were recorded using internationally recognised definitions. 
@@
 \section{Results}
+
+\subsection{Clinico-pathological characteristics}
 One hundred and twenty-nine patients had undergone pancreaticoduodenectomy (n=127), sub-total pancreatectomy (n=1) or total pancreatectomy (n=1) during the study period. 
@@
 Of these, 100 patients (pancreaticoduodenectomy - 98, sub-total/total pancreatectomy - 2) had undergone cardiopulmonary exercise testing as part of their preoperative assessment and were included in the study.
-Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intraductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).
+Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).
 
@@
 The median VO$_2$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). 
-The VO$_2$AT was less than 10ml/kg/min in 49 patients. 
+The VO$_2$AT was less than 10 ml/kg/min in 49 patients. 
 The distribution of VO$_2$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.
@@
 
+\subsection{Anaerobic threshold vs. complications}
 The relationship between VO$_2$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. 
-Patients with VO$_2$AT less than 10ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042).
+Patients with VO$_2$AT less than 10 ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042).
 While there was an association between low VO$_2$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). 
@@
 
+\subsection{Anaerobic threshold vs. length of stay}
 The median length of postoperative stay was 17 days (IQR 13 - 26). 
@@
 On univariate analysis, age over 65 years (p=0.072) and low VO$_2$AT (p=0.010) were associated with prolonged postoperative stay. 
-On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay.
+On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10 ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay.
 A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low and normal VO$_2$ATs is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. 
@@
 \includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
-\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10ml/kg/min versus $<$ 10ml/kg/min.}
+\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
 \label{fig:cpet_outcomes_km_at_los}
@@
 
+\subsection{Anaerobic threshold vs. receipt of adjuvant therapy}
 The relationship between clinico-pathological patient factors and receipt of adjuvant therapy is shown in Table \ref{table:cpet_outcomes_table4}. 
@@
 Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6).
-On binary logistic regression analysis, VO$_2$AT less than 10ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).
+On binary logistic regression analysis, VO$_2$AT less than 10 ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).
 
@@
 
-In the present study, VO$_2$AT was less than 10ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
+In the present study, VO$_2$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
 The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.
 While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.
-Older and co-workers reported in 1993 that low VO$_2$AT less than 11ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
+Older and co-workers reported in 1993 that low VO$_2$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
 
@@
 
-The results of this study are consistent with the results of the study by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min.
+The results of this study are consistent with the findings by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min.
 However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.
@@
 
-This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of ‘prehabilitation’ \parencite{topp_effect_2002}. 
+This functional capacity of patients to withstand the physiological burden of major surgery can be improved by the process of `prehabilitation' \parencite{topp_effect_2002}. 
 It has been suggested that prehabilitation not only improves aerobic capacity \parencite{jones_effects_2007} but may also improve postoperative recovery \parencite{mayo_impact_2011, pehlivan_effects_2011}.

### 2015-08-22 Chapters/CPET_Jaundice.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - CPET_Jaundice
- Major rewrite of introduction and discussion
- Reads a lot better now - lets see what Donny says
- Time to sleep ?!

@@
 
-\chapter{An investigation into the relationship between obstructive jaundice and preoperative pathophysiology in patients undergoing major pancreatic surgery.}
+\chapter{An investigation into the relationship between cardiopulmonary exercise testing, obstructive jaundice and preoperative pathophysiology in patients undergoing major pancreatic surgery.}
 
 \label{ch_cpet_jaundice}
-
-\lhead{Chapter \ref{ch_cpet_jaundice}. \emph{Obstructive Jaundice}} % This is for the header on each page - perhaps a shortened title
-
+\lhead{Chapter \ref{ch_cpet_jaundice}. \emph{CPET, Jaundice and Preoperative Pathophysiology}}
 \clearpage
@@
 However, major pancreatic surgery is associated with significant morbidity and mortality and is only undertaken in specialist centres. 
-Patient selection, preoperative optimisation, good surgical technique and improvements in postoperative care have all contributed to a reduction in mortality \parencite{winter_1423_2006} but morbidity remains high.
+Patient selection, preoperative optimisation, good surgical technique and improvements in postoperative care have all contributed to a reduction in mortality \parencite{winter_1423_2006} but morbidity remains high \parencite{mann_review_2010}.
 While several technical strategies have been described in recent years to minimise morbidity, these strategies are not necessarily based on a better understanding of the physiological basis of postoperative complications in these patients.
 
-The anatomical relationship between the distal bile duct, distal pancreatic duct, head of the pancreas and the duodenum is responsible for obstructive jaundice being the most common presenting symptom in patients with tumours affecting this region.
-Distal bile duct strictures also occur in a small proportion of patients with severe chronic pancreatitis involving the pancreatic head.
-The perioperative management of the patient with obstructive jaundice is complex and management algorithms are still evolving.
+The close anatomical relationship between the distal bile duct, distal pancreatic duct, head of the pancreas and the duodenum is responsible for obstructive jaundice being the most common presenting symptom in patients with tumours affecting this region.
+Distal bile duct strictures also occur in a proportion of patients with severe chronic pancreatitis involving the pancreatic head.
+In such patients, it may be impossible to distinguish inflammation from malignant disease on preoperative imaging and a proportion of these patients will require a pancreaticoduodenectomy \parencite{abraham_pancreaticoduodenectomy_2003}.
+The perioperative management of the patient with obstructive jaundice is complex and management algorithms are still evolving \parencite{wang_preoperative_2008}.
 
-Obstructive jaundice has been reported to be associated with abnormal cardiovascular physiology in several animal and human studies.
+Obstructive jaundice has been shown to be associated with abnormal cardiovascular physiology in several animal and human studies. 
 Surgery in the jaundiced patient has been reported to be associated with adverse postoperative haemodynamic events and renal dysfunction \parencite{pain_perioperative_1985,green_systemic_1995}. 
 The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death \parencite{king_effect_1909}.
-Green and co-workers (1986) described the effects of ‘cholemia' in dogs that were subjected to choledochocaval anastomosis.
-The resultant myocardial depression was described by them as the ‘jaundiced heart'\parencite{green_jaundiced_1986} and has been reported to be associated with poor myocardial response to inotropic stimulation in dogs \parencite{binah_obstructive_1985, bomzon_systemic_1986} as well as humans \parencite{lumlertgul_jaundiced_1991}.
+Green and co-workers (1986) described the effects of `cholemia' in dogs that were subjected to choledochocaval anastomosis. 
+The resultant myocardial depression was described by them as the `jaundiced heart' \parencite{green_jaundiced_1986} and was associated with poor myocardial response to inotropic stimulation in dogs \parencite{binah_obstructive_1985, bomzon_systemic_1986} as well as humans \parencite{lumlertgul_jaundiced_1991}.
 
-Preoperative biliary drainage used to be advocated before subjecting a patient to pancreaticoduodenectomy with the intention of reducing postoperative morbidity.
-However, several recent studies have reported that routine PBD is associated with increased complication rates as a consequence of the drainage procedure itself as well as increased incidence of postoperative complications.
-The DROP trial reported that PBD was associated with drainage related complication as well as postoperative infectious complications.
-However, this trial excluded patients with a bilirubin levels greater than 250 mg/dl from the study.
+As a result, preoperative biliary drainage (PBD) used to be advocated routinely before subjecting a patient to pancreaticoduodenectomy with the intention of reducing postoperative morbidity. 
+A recent multi-center, randomised trial found that PBD was associated with increased incidence of complications after pancreaticoduodenectomy \parencite{van_der_gaag_preoperative_2010}.
+However, this trial excluded patients with a bilirubin levels greater than 250 $\mu$mol/l from the study. 
+A Cochrane meta-analysis of 6 studies including 520 patients also concluded that pre-operative biliary drainage was associated with increased incidence of complications and should not be used routinely in patients scheduled to undergo pancreaticoduodenectomy \parencite{fang_pre-operative_2012}.
 
-We have recently reported that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy resulting in an increased incidence of POPF and prolonged hospital stay.
-However, the effects of 'severe jaundice' where bilirubin levels exceed 250 on preoperative patient physiology have not been studied adequately.
-
-The aim of the present study was to evaluate the relationship between obstructive jaundice and preoperative pathophysiology including cardiopulmonary exercise physiology in patients undergoing pancreaticoduodenectomy.
+\subsection{Aim}
+The results presented in Chapter \ref{ch_cpet_outcomes} demonstrated that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy resulting in an increased incidence of POPF and prolonged hospital stay.
+Obstructive jaundice was present in a large proportion of these patients.
+The aim of the present study was to evaluate the relationship between cardiopulmonary exercise testing, obstructive jaundice (especially severe obstructive jaundice with serum bilirubin $>$ 250 $\mu$mol/l) and preoperative pathophysiology in patients undergoing major pancreatic surgery.
 
@@
 \section{Patients and Methods}
+
+\subsection{Patients}
 Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
@@
 
-\subsection{Preoperative Data}
+\subsection{Preoperative data}
 Patient demographics, preoperative clinico-pathological characteristics including cardiorespiratory comorbidity, results of preoperative blood tests, chest x-ray, ECG and cardiopulmonary exercise tests were collected from prospectively held databases.
@@
 
-\subsection{Obstructive Jaundice}
-Serum bilirubin levels were measured in all patients on the day before surgery.
-Obstructive jaundice was defined as bilirubin levels greater than 35 micromol/litre and severe obstructive jaundice was defined as bilirubin levels greater than 250 micromol/litre.
-This threshold was selected because the DROP trial did not investigate patients with bilirubin levels greater than 250 micromol/litre and this study aimed to evaluate preoperative pathophysiology in this particular group.
-\todo{A word about stents}
+\subsection{Obstructive jaundice}
+Serum bilirubin levels and liver function tests were measured in all patients on the day before surgery. 
+Obstructive jaundice was defined as serum bilirubin $>$ 35 $\mu$mol/litre and severe obstructive jaundice was defined as serum bilirubin $>$ 250 $\mu$mol/litre. 
+These thresholds were chosen as the recent randomised controlled trial that demonstrated increased complications in patients who underwent PBD excluded patients with serum bilirubin $>$ 250 $\mu$mol/litre from the study \parencite{van_der_gaag_preoperative_2010}.
+The present study aimed to evaluate preoperative pathophysiology in this particular group of patients with `severe obstructive jaundice'.
+Due to referral practices at the time of this study, a proportion of patients underwent preoperative biliary interventions and drainage before referral to the West of Scotland Pancreatic Unit. 
+Data regarding biliary intervention and stenting was collected where feasible.
 
-\subsection{Cardiopulmonary Exercise Test}
-Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA) (9).
+\subsection{Cardiopulmonary exercise test}
+Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
 All patients underwent standard pulmonary function tests and spirometry prior to cardiopulmonary exercise testing. 
@@
 The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
-$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in these patients.
+$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshol.
 Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
-Detailed description of cardiopulmonary exercise testing as well as the physiological parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
+Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
+A mored detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
 
@@
 \section{Results}
+
+\subsection{Clinico-pathological characteristics}
 One-hundred and thirty eight patients underwent pancreaticoduodenectomy with preoperative cardiopulmonary exercise testing during the study period. 
@@
 The POSSUM Physiology Score was greater than 14 in 61 patients (44\%). 
-Obstructive jaundice (serum bilirubin 35 – 250) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin $>$ 250) was present in 19 (14\%) patients.
+Obstructive jaundice (serum bilirubin 35-250 $\mu$mol/l) was present in 32 (23\%) patients while severe obstructive jaundice (serum bilirubin $>$ 250 $\mu$mol/l) was present in 19 (14\%) patients.
 
@@
 Elevated POSSUM Physiology Score (p=0.004) and malignancy (p$<$0.001) were significantly associated with the presence of jaundice. 
-However, there was no statistically significant difference in age, body mass index, cardiovascular comorbidity, respiratory comorbidity or preoperative biliary drainage between the non-jaundiced and jaundiced patients.
+However, there was no statistically significant difference in age, body mass index, cardiovascular comorbidity, respiratory comorbidity or PBD between the non-jaundiced and jaundiced patients.
 
@@
 
+\subsection{Obstructive jaundice vs. preoperative blood tests}
 The relationship between obstructive jaundice and preoperative blood tests is shown in Table \ref{table:cpet_oj_bloods}. 
@@
 
+\subsection{Obstructive jaundice vs. pulmonary function tests}
 Pulmonary function tests including forced vital capacity (FVC), forced expiratory volume in one second (FEV1) and their derived parameters (predicted FEV1, FEV1/FVC, predicted FEV1/FVC) were compared between jaundiced and non-jaundiced patients.
 There was a trend towards lower forced vital capacity with increasing severity of jaundice but this did not reach statistical significance (p=0.092). 
-There was no association between the other pulmonary function tests and jaundice.
-(Table \ref{table:cpet_oj_pft}).
+There was no association between the other pulmonary function tests and jaundice. (Table \ref{table:cpet_oj_pft}).
 
@@
 
+\subsection{Obstructive jaundice vs. CPET}
 Cardiopulmonary exercise test parameters measured at the anaerobic threshold and at peak exercise were compared between jaundiced and non-jaundiced patients using the Kruskal-Wallis test. 
@@
 
+Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak as continuous variables is depicted in Figure \ref{fig:cpet_oj_scatter}. 
+This shows that the relationship between serum bilirubin and $\dot{V}_{O_2}$AT is weak with a $\rho^2$ value of 0.035 Pearson's $\rho$ = - 0.187, p = 0.028. 
+There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)
+
+\begin{figure}[htbp]
+	\centering
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_at_bil}
+		\caption{$\dot{V}_{O_2}$AT versus serum bilirubin}
+		\label{fig:cpet_oj_scatter_at_bil}
+	\end{subfigure}
+	\begin{subfigure}{0.48\textwidth}
+		\centering
+		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_peak_bil}
+		\caption{$\dot{V}_{O_2}$Peak versus serum bilirubin}
+		\label{fig:cpet_oj_scatter_peak_bil}
+	\end{subfigure}
+
+	\caption{Pearson's correlation and scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients undergoing pancreaticoduodenectomy.}
+	\label{fig:cpet_oj_scatter}
+
+\end{figure}
+
+\subsection{Preoperative factors related to CPET}
 Binary logistic regression analysis was undertaken to assess preoperative clinico-pathological patient factors associated with a low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak. 
 On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/dl (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min.
-Age, cardiovascular and respiratory comorbidity, preoperative biliary drainage, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.
+Age, cardiovascular and respiratory comorbidity, PBD, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.
 
@@
 Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/dl (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT.
-Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis.
-(Table \ref{table:cpet_oj_at_regression})
+Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis. (Table \ref{table:cpet_oj_at_regression})
 
@@
 
-Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak as continuous variables is depicted in Figure \ref{fig:cpet_oj_scatter}.
-This shows that the relationship between serum bilirubin and $\dot{V}_{O_2}$AT is weak with a $\rho^2$ value of 0.035 Pearson's $\rho$ = - 0.187, p = 0.028.
-There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)
+\section{Discussion}
 
-\begin{figure}[p]
-	\centering
-	\begin{subfigure}{0.6\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_at_bil}
-		\caption{Pearson's $\rho$ = - 0.187, $\rho^2$ = 0.035, p = 0.028}
-		\label{fig:cpet_oj_scatter_at_bil}
-	\end{subfigure}
-	\begin{subfigure}{0.6\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/cpet_oj_scatter_peak_bil}
-		\caption{$\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123}
-		\label{fig:cpet_oj_scatter_peak_bil}
-	\end{subfigure}
+The results of the present study demonstrate that obstructive jaundice, even when severe, is not an independent risk factor for poor preoperative cardiopulmonary exercise physiology in patients undergoing pancreaticoduodenectomy.
+The results also demonstrate that obstructive jaundice is associated with elevated preoperative systemic inflammation with the severity of systemic inflammation related to the severity of the jaundice.
 
-	\caption{Pearson's correlation and scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients undergoing pancreaticoduodenectomy.}
-	\label{fig:cpet_oj_scatter}
-
-\end{figure}
-
-\clearpage
-
-\section{Discussion}
-The optimal preoperative management of obstructive jaundice, especially with extremely high serum bilirubin levels, in the patient with periampullary cancer requiring pancreaticoduodenectomy is still unclear.
-The results of the present study also show for the first time that while obstructive jaundice is associated with a range of biochemical and haematological abnormalities, it does not affect cardiopulmonary physiology as measured by cardiopulmonary exercise testing.
-
+%History of CPET
 The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers \parencite{older_preoperative_1993}. 
 Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012},[Chapter \ref{ch_cpet_outcomes}], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery.
-Cardiopulmonary exercise testing has been reported to be superior to conventional measures of comorbidity chiefly due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress.
-However, the factors responsible for poor aerobic capacity in preoperative patients have not been adequately studied.
+Cardiopulmonary exercise testing is superior to conventional measures of fitness due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress.
+However, the factors responsible for poor aerobic capacity in surgical patients have not been adequately studied. 
+The present study aimed to address this in patients undergoing major pancreatic surgery.
 
-The association between jaundice and cardiovascular physiology was reported over a hundred years ago by King and co-workers who found that injection of porcine bile pigment into dogs resulted in bradycardia, hypotension and eventually death \parencite{king_effect_1909}.
-
-Jaundice has been reported to be associated with myocardial depression \parencite{green_jaundiced_1986}, poor myocardial response to inotropic stimulation \parencite{lumlertgul_jaundiced_1991}, impaired sympathetic baroreflex sensitivity \parencite{song_baroreflex_2009}, deranged atrial natriuretic peptide levels \parencite{pereira_increased_1994,gallardo_increased_1998} as well as multiple other bile-acid receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}.
-Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and coworkers \parencite{padillo_improved_2001}.
+%History of jaundice - reason for PBD
+Experimental and animal studies have shown that obstructive jaundice was associated with myocardial depression \parencite{green_jaundiced_1986}, poor myocardial response to inotropic stimulation \parencite{lumlertgul_jaundiced_1991}, impaired sympathetic baroreflex sensitivity \parencite{song_baroreflex_2009} and deranged atrial natriuretic peptide levels \parencite{pereira_increased_1994,gallardo_increased_1998}.
 
 Historically, obstructive jaundice has also been reported to be associated with adverse haemodynamic events in patients undergoing major surgery. 
-Intraopertive blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were all more common in patients with obstructive jaundice.
-This increased incidence of complications as a consequence of obstructive jaundice resulted in routine PBD being recommended in these patients in order to alleviate their jaundice before undertaking major surgery.
-In fact, Whipple described his earliest pancreaticoduodenectomy as a two-stage operation, with the first stage aimed at performing a biliary bypass to reduce jaundice levels before undertaking the resection at a later second operation.
+Intra-operative blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were all more common in patients with obstructive jaundice \parencite{dixon_factors_1983, pain_perioperative_1985, green_systemic_1995}.
+A recent review noted that bile acids have a complex range of receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}. 
+Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and co-workers \parencite{padillo_improved_2001}.
+This increased incidence of complications resulted in routine pre-operative biliary drainage being recommended in jaundiced patients before undertaking major surgery. 
+Pancreaticoduodenectomy was described by Whipple initially as a two-stage operation, with the first stage involving a cholecysto-gastrostomy aimed at relieving biliary obstruction before undertaking the resection at a later second operation \parencite{whipple_treatment_1935}.
 
-However, more recently, there has been increasing evidence that such routine PBD may itself be associated with increased complications both associated with the drainage procedure itself as well as the effects of PBD on surgical outcomes.
-
+%Reasons for not doing PBD
+However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
 Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
-But, this study looked at a heterogenous group of patients of which only 7 underwent pancreaticoduodenectomy.
-
-A recent meta-analysis \parencite{sewnath_meta-analysis_2002} analysed data from 5 randomised controlled trials comparing surgery with PBD versus surgery without PBD and concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself.
-All five RCTs included in this meta-analysis included a heterogenous group of operations with only a few undergoing pancreaticoduodenectomy while more than 50\% of patients underwent palliative bypass or exploratory laparotomy making comparison of outcomes difficult.
+A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
+The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself.
+These findings have been reproduced in a more recent multi-center, randomised trial \parencite{van_der_gaag_preoperative_2010}.
 A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings \parencite{wang_preoperative_2008}.
 
-The DROP trial sought to clarify the role of PBD in patients undergoing pancreaticoduodenectomy \parencite{van_der_gaag_preoperative_2010}.
-It randomised patients with bilirubin levels between 40 and 250 either to undergo surgery without PBD or to undergo PBD followed by surgery after 4 - 6 weeks.
-The authors reported that PBD resulted in an increase in incidence of complications of which the majority were related to the drainage procedure itself.
-However, this trial excluded patients with bilirubin levels over 250.
+%CPET vs Jaundice
+In the present study, obstructive jaundice was not independently related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak.
+There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
+These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02}) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
+Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
+They reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}
 
-While the aforementioned studies have undermined the role of PBD in jaundiced patients undergoing pancreaticoduodenectomy, the results of the present study show for the first time that the premise for performing PBD, namely the adverse effect of jaundice on cardiopulmonary physiology may itself be flawed in patients undergoing pancreaticoduodenectomy.
-In our study, obstructive jaundice including severe obstructive jaundice did not affect cardiopulmonary exercise capacity as measured by $\dot{V}_{O_2}$AT or the peak oxygen consumption.
-These findings taken together with previously published findings of adverse effects of PBD further support the fact that major surgery may be safe in jaundiced patients without subjecting them to preoperative biliary drainage.
-The basis of the relationship between low $\dot{V}_{O_2}$AT and raised BMI is not clear.
+Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.
 
-However, such an association has been previously reported \parencite{horwich_relationship_2009}  This may reflect the difficulty in obtaining accurate $\dot{V}_{O_2}$AT values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
+%Obstructive jaundice and systemic inflammation
+However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
+Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002}. 
+Both cellular immunity and humoral immunity are adversely affected by obstructive jaundice \parencite{padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
+It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
+However, the immunological abnormalities associated with obstructive jaundice do not appear to improve after biliary drainage \parencite{kimmings_endotoxin_2000}.
+
+Elevated preoperative CRP was also independently associated with a low $\dot{V}_{O_2}$AT suggesting that impaired aerobic capacity may be associated with a systemic inflammatory response.
+This finding is similar to that reported by Sultan and co-workers who found that low $\dot{V}_{O_2}$AT was independently associated with an elevated preoperative neutrophil-lymphocyte ratio in patients undergoing colorectal surgery \parencite{sultan_cardiopulmonary_2014}.
+
+%CPET vs Hemoglobin
+The association between low haemoglobin and low $\dot{V}_{O_2}$Peak would appear to be more easily explained since the oxygen carrying capacity of blood depends on haemoglobin concentration.  
+Indeed, the linear relationship between oxygen consumption during exercise and haemoglobin concentration has been shown both in disease states \parencite{agostoni_relationship_2010} as well as in healthy volunteers \parencite{dellweg_cardiopulmonary_2008}.
+This would suggest that maintaining optimal haemoglobin levels in the perioperative period will improve oxygen delivery and perhaps outcome as well. 
+Excessive intra-operative blood loss has been reported to be associated with postoperative complications \parencite{pratt_risk_2008} and this may in part be due impaired oxygen delivery at a time of maximal physiological burden.
+
+%CPET vs BMI, Sex
+In the present study, it was of interest that $\dot{V}_{O_2}$, both at anaerobic threshold and at peak exercise, was significantly lower in females as well as in patients with  a high body mass index.
+The basis of this relationship is not clear and may be related to body composition differences between these groups.
+However, such an association has been previously reported \parencite{horwich_relationship_2009}.
+This may reflect the difficulty in obtaining accurate $\dot{V}_{O_2}$ values in obese patients as a result of the calculations involved rather than due to true cardiopulmonary dysfunction. 
 Other authors have suggested that different thresholds for CPET parameters may have to be considered in obese patients to improve risk-prediction \parencite{donnelly_criteria_1990,hulens_exercise_2001}.
-Cardiopulmonary exercise testing measures oxygen delivery to skeletal muscle.
-Adipose tissue, however, does not contribute to the metabolic activity that is measured during CPET.
-However, AT as normally reported, is calculated by dividing the oxygen consumption per minute at the ‘anaerobic threshold' into the weight of the patient.
-However, this does not account for the disproportionately higher amount of adipose tissue in overweight/obese patients resulting in a spuriously low AT (in mls/kg/min).
-The present study found no association between cardiorespiratory comorbidity and $\dot{V}_{O_2}$AT.
-Low $\dot{V}_{O_2}$AT in female patients and overweight/obese patients should be interpreted with caution as this may not be due to true poor aerobic capacity.
+Presence of cancer was also an independent predictor of low $\dot{V}_{O_2}$AT.
+Sarcopenia is associated with reduced aerobic capacity \parencite{evans_sarcopenia_1993} and is common in pancreatic cancer \parencite{joglekar_sarcopenia_2015}.
+This is the subject of further study in Chapter \ref{ch_bodycomp}.
 
-\section{Conclusions}
-Obstructive jaundice, including severe obstructive jaundice (serum bilirubin>250 mg/dl) does not affect preoperative cardiopulmonary exercise physiology.
-Reduction of cardiovascular adverse events can no longer be the rationale for preoperative biliary drainage even in patients with severe obstructive jaundice.
-Future studies must evaluate the safety of elective surgery in patients with severe jaundice and show comparable outcomes to non-jaundiced patients before PBD can be completely abandoned except in special circumstances.
+%\subsection{Conclusion}
+Obstructive jaundice, including severe obstructive jaundice did not affect preoperative cardiopulmonary exercise physiology. 
+Therefore, preoperative biliary drainage cannot be expected to improve aerobic capacity in patients undergoing pancreaticoduodenectomy.
+Both obstructive jaundice and low $\dot{V}_{O_2}$AT were associated with elevated preoperative systemic inflammation and provide further evidence of the role of jaundice and impaired aerobic capacity in immune dysregulation in these patients.
+High BMI and female sex were independently associated with low $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak and these parameters must be interpreted with caution in these subgroups.

### 2015-08-22 Tables/cpet_oj_pft.tex
Review 1 - CPET_Jaundice
- Major rewrite of introduction and discussion
- Reads a lot better now - lets see what Donny says
- Time to sleep ?!

@@
 
-	\begin{tabular}{|l| c c c c|c|}
+	\begin{tabular}{|l| c c c c c|}
 		\hline
-		                    &              \multicolumn{4}{c|}{Preoperative Serum Bilirubin}              &  \\
+		                    &              \multicolumn{4}{c}{Preoperative Serum Bilirubin}              &  \\
 		                    & $\leq$ 17        & 18-35            & 35-250            & $>$ 250          & \textit{p} \\ \hline

### 2015-08-22 Chapters/Introduction.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - Introduction
- Fixed nearly all incomplete sections
- All todos sorted
- Multiple tables amended for formatting
- More references for jaundice, immunity
- Added table for POSSUM
- Fixed several overfull box issues with TOC

@@
 
-\section{Pancreatic Neoplasia}
+\section{Pancreatic neoplasia}
 %This is the only paper I need for this entire section - pancreatic_section_of_the_british_society_of_gastroenterology_pancreatic_society_of_great_britain_and_ireland_association_of_upper_gastrointestinal_surgeons_of_great_britain_and_ireland_royal_college_of_pathologists_special_interest_group_for_gastro-i_guidelines_2005
@@
 
-%10/08/2015 - Review completete to here
-
 \subsection{Operative technique}
@@
 Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner.
-Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures while the gastro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers.
+Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures.
+Gstro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
 One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.
@@
 A standardised regimen of intravenous fluids, naso-jejunal feeding, mobilisation and physiotherapy was implemented in all patients. 
-Standard physiological parameters including haemodynamic parameters, renal function and arterial blood gases were used to monitor adequate end organ perfusion.
+Standard physiological parameters including haemodynamic measurements, renal function and arterial blood gases were used to monitor adequate end organ perfusion. 
 All patients received proton pump inhibitors and Octreotide. 
 Patients were discharged to the general surgical ward as early as possible. 
-Amylase content in the drain fluid was measured on the third post-opertive day to aid in the diagnosis of post-operative pancreatic fistula.
-Surgical drains and naso-gastric/naso-jeujunal tubes were removed at surgeon's discretion.
+Amylase content in the drain fluid was measured on the third post-operative day to aid in the diagnosis of post-operative pancreatic fistula. 
+Surgical drains and naso-gastric/naso-jeujunal tubes were removed at the surgeon's discretion. 
 Patients were discharged when medically fit.
@@
 
-\section{Adjuvant and Neoadjuvant treatment}
+\section{Adjuvant and neoadjuvant treatment}
 
@@
 However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. 
-The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid.
-\parencite{neoptolemos_adjuvant_2010}
+The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid. \parencite{neoptolemos_adjuvant_2010}
 
@@
 
-\section{Comorbidity and Risk Stratification}
+\section{Preoperative risk stratification}
 \label{sec:comorbidity_risk_stratification}
-\todo{Comorbidity and Risk Stratification - needs major tidying up - will take several days}
-Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness.
+
+Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}.
 Major pancreatic surgery is associated with significant morbidity. 
 Subjecting patients with poor physiological reserve to a pancreaticoduodenectomy is fraught with risk. 
-It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in longevity as compared to palliation.
+It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in survival when compared to palliation.
 Patient factors that affect postoperative recovery and long-term survival are therefore important in determining suitability for surgery.
@@
 
-\subsection{Risk Stratification}
-\todo{Intro - risk stratification}
-%Include other measures - discuss POSSUM, APACHE, mGPS, EPASS and everything else here.
+\subsection{Risk stratification}
 Physiological fitness or reserve may be defined as the ability of the patient's organ systems to respond appropriately and adequately to the stress of surgery. 
 Major surgery has a significant physiological impact on multiple organ systems, especially the cardiorespiratory system. 
-The ability of the cardiorespiratory system as well as other physiological systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes.
+The ability of the cardiorespiratory system as well as other organ systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes.
 Therefore, accurate assessment of physiological fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions.
@@
 
-\subsection{Static Versus Dynamic Testing}
-\todo{Static Versus Dynamic Testing}
+The Physiological and Operative Severity Score for the enumeration of Mortality and Morbidity (POSSUM) is one of the most used scores for risk-prediction in general surgery \parencite{copeland_possum:_1991}.
+It was devised by Copeland and co-workers as a method of predicting risk-adjusted mortality and morbidity rates for comparative audit between different surgical departments \parencite{copeland_comparative_1993} and individual surgeons \parencite{copeland_risk-adjusted_1995}.
+While some authors have reported it to be useful in predicting postoperative outcomes after pancreatic surgery \parencite{pratt_possum_2008} others have found that POSSUM was not a good risk prediction model in this group of patients \parencite{khan_evaluation_2003, kocher_risk-adjustment_2005, tamijmarane_application_2008, de_castro_evaluation_2009}.
+The physiology component of the POSSUM score (POSSUM Physiology Score, PPS) which takes into account both comorbidity and specific preoperative biochemical parameters was used as an objective static measure of comorbidity in this thesis.
+The POSSUM Physiology Score has been reported to be associated with preoperative systemic inflammation and cancer-specific survival in patients who underwent potentially curative surgery for colorectal cancer \parencite{richards_relationship_2010}.
+The variables used in the calculation of the POSSUM Physiology Score are shown in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
+As the Glasgow Coma Scale in all patients undergoing pancreaticoduodenectomy was 15, this variable was excluded in the calculation of the score in this thesis.
+
+\input{Tables/c1_possum}
+
+More recently, elevated preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score (mGPS) has been reported to be associated with increased incidence of infective complications after colorectal surgery \parencite{moyes_preoperative_2009} as well as oesophageal resections \parencite{vashist_glasgow_2010}.
+
+POSSUM and other similar composite scores such as the APACHE II (Acute Physiology and Chronic Health Evaluation II) \parencite{knaus_apache_1985} and Estimation of Physiologic Ability and Surgical Stress (E-PASS)  \parencite{haga_estimation_1999, hashimoto_is_2010} have not found widespread use for risk prediction in pancreatic surgery.
+While these scores are either useful for comparative audit between patient populations or in the case of APACHE II, for monitoring response to therapy in intensive care, they are of limited value when applied to individual patients.
+
+\subsection{Static versus dynamic testing}
 
@@
 However, none of these tests adequately measure the ability of the cardiopulmonary and circulatory systems to deliver oxygen to the tissues at times of increased demand. 
-Patients with subclinical cardiopulmonary dysfunction or limitation cannot be identified on these tests alone and it is this group of patients who are most likely to be at greater risk of complications and who will most benefit from targeted prehabilitation.
-Cardiopulmonary exercise testing overcomes the limitations of the conventional static tests by providing a global assessment of the patient's oxygen delivery mechanisms at times of increased physiological stress and demand.
+Patients with sub-clinical cardiopulmonary dysfunction or limitation cannot be identified on these tests alone. 
+It is this group of patients who are most likely to be at greater risk of complications and who will most benefit from targeted prehabilitation. 
+Moreover, composite score such as POSSUM do not provide information on the ability of the patient to cope with physiological stress or on how their risk may be mitigated before surgery.
+Cardiopulmonary exercise testing overcomes the limitations of the conventional static tests by providing a global assessment of the patient's oxygen delivery mechanisms at times of increased physiological demand.
 
-One of the earliest attempts to develop a dynamic test in surgical patients was reported by Starr and co-workers in 1945 who tried to devise a test that could assess respiratory function and oxygen utilisation in surgical patients before, immediately after and a week after surgery.
-They reported on the difficulties in obtaining reproducible results but suggested that a slow return to baseline of heart rate, respiratory rate and oxygen utilisation after surgery may indicate patients who are slow to recover.
-\parencite{starr_convalescence_1945}
+\subsection{Pancreas-specific risk factors}
+In addition to all of the above risk factors, post pancreatectomy complications are significantly associated with factors unique to pancreatic disease.
+Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and is discussed in more detail in Section \ref{sec:ch_intro_POPF}.
+Soft pancreas texture, narrow pancreatic duct, pathology that does not obstruct the pancreatic duct such as ampullary or duodenal cancers are independent risk factors for pancreatic fistula \parencite{pratt_risk_2008}.
+These organ-specific risk factors are independent of comorbidity and patient fitness \parencite{deoliveira_assessment_2006}.
+Lin and co-workers reported that in a study of 1891 patients undergoing pancreaticoduodenectomy, only soft pancreatic gland texture and coronary disease were related to postoperative pancreatic fistula on multivariate analysis \parencite{lin_risk_2004}.
+These pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.
 
-\section{Cardiopulmonary Exercise Testing}
-%Not sure I need these two sections - delete once done vacillating
-\todo{A brief history of CPET in general}
-\todo{The physiological basis of CPET}
+\section{Cardiopulmonary exercise testing}
 
-\subsection{Origins and physiological basis of cardiopulmonary exercise testing}
-The measurement of aerobic fitness has been an important aspect of improving athletic performance in competitive sports.
-Cardiopulmonary exercise testing has been used for a long time for detailed measurement of aerobic fitness in athletes.
-It has also been used to design customised training regimens aimed at improving performance at activities that place a significant stress on the cardiovascular and respiratory systems by allowing the optimisation of the individual components of the oxygen delivery system.
+\subsection{Origins of preoperative cardiopulmonary exercise testing}
 
-\subsection{Cardiopulmonary Exercise Test Methodology}
+The idea of measuring $\dot{V}_{O_2}$ (oxygen consumption) and v$\dot{V}_{CO_2}$ (expired carbon-dioxide) as a measure of cardiopulmonary response to exercise was described in the early twentieth century.
+In 1924, Hill and co-workers reported on adaptations made to an existing apparatus to allow measurement of oxygen consumption during exercise \parencite{hill_muscular_1924}. 
+Their method involved the subject running while connected to a large bag slung on their back to collect expired gases. 
+The subjects then returned to the laboratory where they were connected to a fairly large apparatus involving several more large bags and other equipment that filled the most part of a large room. 
+But the premise of their method remains unchanged to this date; to examine the response of the cardiopulmonary system to exercise.
+Such methods were refined by other physiologists to reduce the chance of errors and to devise more practical methods suitable for medical use \parencite{katz_ln_metabolic_1934, sutton_estimation_1940}.
+
+One of the earliest attempts to apply these dynamic tests in surgical patients was reported by Starr and co-workers in 1945 \parencite{starr_convalescence_1945}.
+They tried to devise a test that could assess respiratory function and oxygen utilisation in surgical patients before, immediately after and a week after surgery. 
+They reported on the difficulties in obtaining reproducible results but suggested that a slow return to baseline of heart rate, respiratory rate and oxygen utilisation after surgery could identify patients who are slow to recover.
+
+However, cardiopulmonary exercise testing in its current form where breath-by-breath analysis of gas exchange is combined with real-time electrocardiography during incremental exercise was devised and popularised by Karlman Wasserman and his team of physiologists.
+Wassermann and his team reported on the ventilatory response to exercise and how it may be used to identify cardiorespiratory disease in 1964 \parencite{naimark_continuous_1964}.
+The pioneering work on the detection of anaerobic threshold was also done by his team and applied in the assessment of patients with cardiac disease \parencite{wasserman_detecting_1964, wasserman_lactate_1967, wasserma.k_anaerobic_1973}.
+
+The use of cardiopulmonary exercise testing as part of preoperative assessment was first reported by Older and co-workers.
+They conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery  \parencite{older_preoperative_1993}. 
+All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. 
+The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. 
+There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. 
+They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55).
+Only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, p$<$0.001). 
+Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia (p$<$0.01).
+
+\subsection{Cardiopulmonary exercise test methodology}
 \label{sec:cpx_method}
@@
 	\includegraphics[width=\textwidth]{Figures/cpet-zan}
-	\caption{ZAN-600 Cycle Erogmeter and CPET suite. (www.nspirehealth.com)}
+	\caption{ZAN-600 Cycle Erogmeter and CPET suite.}
+	(Image from www.nspirehealth.com)
 	\label{fig:cpet-zan}
@@
 
-CPET is composed of several components that involve measuring not only the response of the cardiac and respiratory system to exercise but the test also helps establish the adequacy of this response to sustain oxygen delivery to skeletal muscle as demand increases with increasing exercise.
-
 Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
-The equipment was calibrated regularly to the standards set by the manufacturer and currently published guidelines[].
 All tests were performed by specialist respiratory physiologists. 
@@
 
-Spirometry was performed in all patients prior to CPET.
-Capillary blood gases were measured in all patients after CPET.
+Spirometry was performed in all patients prior to cardiopulmonary exercise testing. 
+Capillary blood gases were measured in all patients after the test. 
+The gas and flow analysis equipment were calibrated prior to every test.
 An electronically braked cycle ergometer was used to increase resistance to pedalling in preset increments. 
 A tight-fitting face mask was placed on the patient covering the nose and the mouth. 
-This allowed breath-by-breath gas analysis thus allowing measurement of several respiratory parameters as listed in table [].
-12-lead electrocardiogram was recorded at the same time.
+Breath-by-breath gas analysis and 12-lead electrocardiography were displayed in real-time to the investigator allowing measurement of several cardiac and respiratory parameters as they changed during exercise.
 
 The test started with an initial 3-minute rest period to allow measurement of baseline parameters. 
-This was followed by an incremental work-load test that involved the patient pedalling approximately at 60 revolutions per minute while the resistance to pedalling was gradually increased in preset increments.
+This was followed by an incremental work-load test that involved the patient pedalling at approximately 60 revolutions per minute while the resistance to pedalling was gradually increased in preset increments.
 The test was terminated when patients reached volitional fatigue (maximal exercise tolerance), significant ischaemic changes on ECG or for other safety reasons.
 
-The parameters measured at spirometry are shown in Table \ref{table:spirometry} and those measured during cardiopulmonary exercise testing are shown in Table \ref{table:cpet_parameters} on p\pageref{table:cpet_parameters}.
+The oxygen consumption at this point (considered `peak' exercise capacity) is called $\dot{V}_{O_2}$Peak.
+This is different from $\dot{V}_{O_2}$Max which is identified by the plateauing of the increasing $\dot{V}_{O_2}$ when the physiological limits of oxygen consumption are met in an individual.
+$\dot{V}_{O_2}$Max is only observed in very fit individuals or well-trained athletes.
+This seldom happens in surgical patients and $\dot{V}_{O_2}$Peak is used as a surrogate for $\dot{V}_{O_2}$Max.
 
-\subsection{Measuring the Anaerobic Threshold}
+The list of parameters measured is provided in Tables \ref{table:spirometry} and \ref{table:cpet_parameters} on p\pageref{table:spirometry}.
+A detailed description of these parameters is provided in Section \ref{sec:cpx_parameters}.
+
+\subsection{Measuring the anaerobic threshold}
+\label{sec:cpx_at_method}
 
@@
 	\includegraphics[width=\textwidth]{Figures/cpet_9panel.pdf}
-	\caption{9-panel view of trending parameters during incremental cardiopulmonary exercise.}
+	\caption{9-panel view of trending parameters during cardiopulmonary exercise.}
 	\label{fig:cpet_9panel}
@@
 \end{equation}
-The resulting excess $CO_2$ is exhaled and is one of the many parameters measured during cardiopulmonary exercise testing.
+The resulting excess $CO_2$ is exhaled and is one of many parameters measured during cardiopulmonary exercise testing. 
 This transition from aerobic to anaerobic respiration may be determined using the V-slope method \parencite{sue_metabolic_1988} or the ventilatory equivalents method \parencite{beaver_new_1986}. 
-Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold as recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.
+Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold.
+This approach is also recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.
 
@@
 During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}. 
-However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionately to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}.
+However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionate to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}.
 This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B). 
-The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold, $\dot{V}_{O_2}$AT or simply AT.
+The point at which the two slopes intersect is the anaerobic threshold and the $\dot{V}_{O_2}$ at this point in exercise is commonly referred to as the anaerobic threshold or $\dot{V}_{O_2}$AT.
 
@@
 and $\dot{V}_E/\dot{V}_{CO_2}$ against time, exercise load or $\dot{V}_{O_2}$. 
-The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig.
-\ref{fig:cpet_9panel}F.
+The point at which $\dot{V}_E/\dot{V}_{O_2}$ increases while $\dot{V}_E/\dot{V}_{CO_2}$ stays the same or decreases is the anaerobic threshold and this can be seen in Fig. \ref{fig:cpet_9panel}F. 
 Most respiratory physiologists including those at our centre use both methods as well as other trends during exercise to determine the anaerobic threshold as well as its validity.
 
-\section{Description of CPET Parameters}
+\section{Description of CPET parameters}
 \label{sec:cpx_parameters}
@@
 
-\subsection{Exercise Load}
+\subsection{Exercise load}
 The most common form of cardiopulmonary exercise testing for clinical use involves a cycle ergometer with steadily increasing resistance delivered through electric braking. 
@@
 
-\subsection{Minute Ventilation, $\dot{V}_E$}
-Minute ventilation or respiratory minute volume is the volume of air that is inhaled/expired in a minute.
+\subsection{Minute ventilation}
+Minute ventilation or respiratory minute volume ($\dot{V}_E$) is the volume of air that is inhaled/expired in a minute.
 \begin{equation} \label{eq:VE=VTxBf}
@@
 
-Increasing $\dot{V}_E$ is one of the main mechanisms involved in increasing oxygen delivery during exercise. It is also an important factor in clearing $CO_2$ from the blood.
+Increasing $\dot{V}_E$ is one of the main mechanisms involved in increasing oxygen delivery during exercise. 
+It is also an important factor in clearing $CO_2$ from the blood.
 
-\subsection{Oxygen Uptake, $\dot{V}_{O_2}$}
+\subsection{Oxygen uptake}
 
@@
 \end{equation}
-Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.
+Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. 
+The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.
 
@@
 
-\subsection{Oxygen Pulse, $O_2Pulse$}
-Oxygen pulse is defined as the oxygen uptake per heart beat.
+\subsection{Oxygen pulse}
+$O_2Pulse$ is defined as the oxygen uptake per heart beat.
 \begin{equation} \label{eq:oxygen_pulse}
@@
 \end{equation}
-While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. The clinical application of oxygen pulse in surgical patients remains unclear.
+While some authors have suggested that oxygen pulse may be a surrogate for stroke volume others disagree. 
+The clinical application of oxygen pulse in surgical patients remains unclear.
 
-\subsection{Respiratory Exchange Ratio, RER}
-The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio.
+\subsection{Respiratory exchange ratio}
+The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio (RER). 
 An RER greater that 1.0 may be caused either by lactic acidosis or due to hyperventilation. 
@@
 
-\subsection{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
-The changes in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise.
+\subsection{Ventilatory equivalents}
+The changes in the ventilatory equivalents for $O_2$ ($\dot{V}_E/\dot{V}_{O_2}$) and $CO_2$ ($\dot{V}_E/\dot{V}_{CO_2}$) during exercise provide valuable information regarding the ventilatory response to exercise.
 Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. 
@@
 
-\subsection{End-tidal $O_2$ and $CO_2$, $P_{ET_{O_2}}$, $P{ET_{CO_2}}$}
+\subsection{End-tidal oxygen and CO2} 
 $PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to $P_aO_2$ and $P_aCO_2$ respectively.
@@
 
-\subsection{Heart Rate, HR}
+\subsection{Heart rate}
 \label{sec:heart_rate}
@@
 
-\subsection{Breathing frequency, $B_f$}
-In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake.
+\subsection{Breathing frequency}
+In healthy individuals, breathing frequency ($B_f$) is one of the most important respiratory responses to exercise in order to increase oxygen uptake. 
 However, a variety of respiratory disorders including chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response. 
@@
 \label{sec:cpx_roleInAssessment}
-\subsection{General Surgery}
+\subsection{General surgery}
 
 Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery.
-Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery.
-All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring.
-The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg.
-There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes.
-They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$).
-Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$) \parencite{older_preoperative_1993}.
 
+Older and co-workers first reported on the value of cardiopulmonary exercise testing for preoperative risk assessment in a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery  \parencite{older_preoperative_1993}.
 In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise.
@@
 McCullough and co-workers defined a composite primary outcome measure that included myocardial infarction, unstable angina, deep vein thrombosis, pulmonary embolism, renal failure, stroke and death and applied this in a cohort of 109 consecutive patients who underwent Roux-en-Y gastric bypass surgery.
-This composite adverse outcome was more likely in patients with high BMI ($>45$) and low $\dot{V}_{O_2}$Peak ($<15.9 ml/kg/min$).
+This composite adverse outcome was more likely in patients with high BMI ($>$45) and low $\dot{V}_{O_2}$Peak ($<$15.9 ml/kg/min). 
 In fact, patients with neither of these two risk factors had no complications after surgery. 
-They also observed a significant negative relationship between BMI and $\dot{V}_{O_2}$ ($p<0.0001$) and age($p<0.0001$).
+They also observed a significant negative relationship between BMI and $\dot{V}_{O_2}$ (p$<$0.0001) and age(p$<$0.0001). 
 Poor aerobic capacity was also associated with the presence of diabetes and hypertension suggesting further evidence of the metabolic cost of obesity in these patients \parencite{mccullough_cardiorespiratory_2006}.
@@
 They postulated that the fact that post-oesophagectomy complications such as anastomotic leak or sepsis often happen within the thorax and are not necessarily due to cardiopulmonary dysfunction, preoperative cardiopulmonary exercise testing may not have identified these patients.
-Older and Hall, the pioneers of cardiopulmonary exercise test in surgical patients, also raise this issue in their letter to the authors of the above study.
-\parencite{hall_cardiopulmonary_2009}.
+Older and Hall, the pioneers of cardiopulmonary exercise test in surgical patients, also raise this issue in their letter to the authors of the above study. \parencite{hall_cardiopulmonary_2009}.
 
-\subsection{Vascular Surgery}
+\subsection{Vascular surgery}
 The earliest report of the application of cardiopulmonary exercise test in elective abdominal aortic aneurysm surgery was by Nugent and co-workers in 1998. 
@@
 
-\subsection{Liver Transplantation}
+\subsection{Liver transplantation}
 Peak $\dot{V}_{O_2}<60\%$ predicted and $\dot{V}_{O_2}AT<50\%$ predicted have been reported to be associated with increased 100-day mortality in patients undergoing liver transplantation. 
@@
 These include cirrhotic cardiomyopathy, hepatopulmonary syndrome and decreased peripheral oxygen utilisation due to cirrhotic myopathy \parencite{epstein_aerobic_2004}. 
-This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery \parencite{iscar_functional_2009}.
-.
+This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery \parencite{iscar_functional_2009}. . 
 However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
 
-\subsection{Thoracic Surgery}
+\subsection{Thoracic surgery}
 The role of cardiopulmonary exercise testing in patient undergoing non-cardiac thoracic surgery has been well established for over a decade. 
@@
 
-The 2$^nd$ edition of the American College of Chest Physicians' Evidence-Based Clinical Practice Guidelines for the 'Physiologic evaluation of the patient with lung cancer being considered for resectional surgery' recommends that patients with a low $\dot{V}_{O_2}$Max ($<10$ ml/kg/min) or with low $\dot{V}_{O_2}$Max ($<15$ ml/kg/min) associated with other risk factors on spirometry and cardiopulmonary exercise testing should be counselled about non-operative treatment options and non-standard surgery as the risk of perioperative death and cardiopulmonary complications was very high.
+The 2$^{nd}$ edition of the American College of Chest Physicians' Evidence-Based Clinical Practice Guidelines for the 'Physiologic evaluation of the patient with lung cancer being considered for resectional surgery' recommends that patients with a low $\dot{V}_{O_2}$Max ($<$10 ml/kg/min) or with low $\dot{V}_{O_2}$Max ($<$15 ml/kg/min) associated with other risk factors on spirometry and cardiopulmonary exercise testing should be counselled about non-operative treatment options and non-standard surgery as the risk of perioperative death and cardiopulmonary complications was very high.
 The 1C grading of this recommendation not only represents the observational nature of most of the studies it is based on but also that such observational evidence is sufficiently consistent \parencite{benzo_complications_2007} to make a strong recommendation \parencite{colice_physiologic_2007}.
@@
 
-%Preop inflammation - short-term and long-term outcomes
-%Postop inflammation -> complications -> long-term ouctomes
-%Possible mechanisms behind poor outcomes
-
-The host inflammatory response to cancer, comorbidity and surgical trauma has been known to influence both short-term and long-term outcomes after major cancer surgery.
-Moreover, postoperative complications have been reported to be associated with poorer oncologic outcomes and cancer-specific survival in patients undergoing potentially curative surgery for cancer.
+The host inflammatory response to cancer, comorbidity and surgical trauma has been known to influence both short-term and long-term outcomes after major cancer surgery \parencite{mcmillan_measurement_2001, read_differential_2004, roxburgh_relationship_2010}.
+Moreover, postoperative complications have been reported to be associated with poorer oncologic outcomes and cancer-specific survival in patients undergoing potentially curative surgery for cancer \parencite{mcardle_impact_2005}.
 The complex interactions between pro-inflammatory cytokines and anti-inflammatory cytokines at different phases during the perioperative period further impact upon the incidence of complications as well as survival.
@@
 One of the earliest reports on the use of CRP to predict cancer-specific survival was by McMillan and co-workers in 1995 when they reported that an elevated CRP 4 months after curative resection for colorectal cancer was associated with earlier recurrence \parencite{mcmillan_prospective_1995}.
-The modified Glasgow Prognostic Score (mGPS)\parencite{elahi_score_2004} is based on a combination of C-reactive protein and serum albumin and is outlined in Table \ref{table:mGPS}.
-.
+The modified Glasgow Prognostic Score (mGPS) \parencite{elahi_score_2004} is based on a combination of C-reactive protein and serum albumin and is outlined in Table \ref{table:mGPS}.
 Since its introduction, mGPS has been validated in over a hundred studies looking at several thousand patients with a wide-range of cancers and an increasing score is associated with poorer long-term survival in patients with operable as well as inoperable cancers.
@@
 	\label{table:mGPS}
-	\begin{tabular}{c c c}
-		mGPS & CRP (mg/dL) & Albumin (mg/dL) \\ \hline
+	\renewcommand{\arraystretch}{1.5} %Increases space between rows
+	\setlength{\tabcolsep}{12pt} %sets the space between columns
+	\begin{tabular}{|ccc|}
+		\hline
+		mGPS & CRP (mg/l) & Albumin (g/l) \\
 		 0   & $\leq 10$  & $\geq 35$     \\
 		 1   & $> 10$     & $\geq 35$     \\
-		 2  & $> 10$   & $< 35$
+		 2   & $> 10$     & $< 35$        \\ \hline
 	\end{tabular}
@@
 
-\subsection{Systemic inflammation and long-term survival}
-Systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005,clark_preoperative_2007,bhatti_preoperative_2010} as well as in patients with inoperable disease \parencite{glen_evaluation_2006}.
-Patients with ductal adenocarcinoma of the head of the pancreas undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated over 10 mg/dl approximately a month after their operation \parencite{jamieson_systemic_2005}.
-Similar findings have been reported in cancers involving other organs using both the mGPS and other scores such as the neutrophil-lymphocyte ratio (NLR).
-A selection of these studies are presented in Table %Create new table here for this.
-\todo{Table for systemic inflammation and long-term survival}
+\subsection{Inflammation and long-term survival}
+Elevated preoperative systemic inflammation is associated with poorer survival in patients undergoing potentially curative surgery for pancreatic cancer \parencite{jamieson_systemic_2005, clark_preoperative_2007, bhatti_preoperative_2010} as well as in patients with advanced pancreatic cancer \parencite{glen_evaluation_2006}.
+Patients with pancreatic ductal adenocarcinoma undergoing potentially curative resection survived for a median of 21.5 months if their CRP was $\leq$ 10 mg/dl a month after their surgery but only 8.4 months if their CRP remained persistently elevated over 10 mg/dl approximately a month after their operation \parencite{jamieson_systemic_2005}.
 
-\subsection{Systemic inflammation and postoperative complications}
+Similar findings have been reported in cancers involving other organs using both the modified Glasgow Prognostic Score and other scores such as the neutrophil-lymphocyte ratio (NLR). 
+The modified Glasgow Prognostic Score was associated with survival independent of tumour stage in patients with colorectal \parencite{mcmillan_evaluation_2007, leitch_comparison_2007}, breast \parencite{al_murri_evaluation_2006, al_murri_evaluation_2007}, lung \parencite{forrest_comparison_2004}, oesophago-gastric \parencite{crumley_elevated_2006}, bladder \parencite{hilmy_relationship_2006} and prostate \parencite{mcardle_systemic_2006} cancer.
 
+\subsection{Inflammation and postoperative complications}
 Abnormalities of systemic inflammatory processes present as a continuum that starts in the preoperative phase possibly as a consequence of underlying comorbid illnesses, presence of cancer, or an abnormality of the immune system or a due to a combination of all of these factors.
@@
 \subsubsection{Preoperative systemic inflammation}
-Elevated levels of interleukin-6, alpha-1 antitrypsin and CRP and decreased levels of albumin and prealbumin before surgery have been reported to be associated with a more exaggerated postoperative systemic inflammatory response and infectious complications after major abdominal surgery \parencite{haupt_association_1997}.
+Elevated levels of interleukin-6, $\alpha$-1 antitrypsin and CRP and decreased levels of albumin and prealbumin before surgery have been reported to be associated with a more exaggerated postoperative systemic inflammatory response and infectious complications after major abdominal surgery \parencite{haupt_association_1997}.
 
@@
 In a study of 455 patients, Moyes and co-workers reported that an elevated preoperative modified Glasgow Prognostic Score (\ref{table:mGPS}) was associated with increased incidence of infectious complications in patients undergoing elective as well emergency colorectal cancer surgery.
-They postulated that several mechanisms may have a role including disregulation of cell-mediated immunity, impaired T-lymphocyte response, disorders in the complement pathway and possibly due to loss of lean tissue and protein as a consequence of systemic inflammation.
-Preoperative mGPS has also been shown to predict postoperative morbidity in patients undergoing oesophageal resection for cancer \parencite{vashist_glasgow_2010}.
+They postulated that several mechanisms may play a role including dysregulation of cell-mediated immunity, impaired T-lymphocyte response, disorders in the complement pathway and possibly due to loss of lean tissue and protein as a consequence of systemic inflammation.
+Preoperative mGPS was also associated with increased incidence of postoperative complications in patients undergoing oesophageal resection for cancer \parencite{vashist_glasgow_2010}.
 
@@
 An exaggerated and persistent systemic inflammatory response in the early postoperative period is associated with an increased incidence of complications. 
-One of the earliest studies comparing several 'acute-phase proteins' and their role in predicting postoperative complications reported that in patients who developed surgical inflammatory complications, CRP remained elevated after the third postoperative day while other acute-phase proteins such as ceruloplasmin and alpha-1 antitrypsin where not useful in monitoring the postoperative course \parencite{fischer_quantitation_1976}.
+One of the earliest studies comparing several 'acute-phase proteins' and their role in predicting postoperative complications reported that in patients who developed surgical inflammatory complications, CRP remained significantly elevated after the third postoperative day \parencite{fischer_quantitation_1976}.
+But,  other acute-phase proteins such as ceruloplasmin and  $\alpha$-1 antitrypsin where not useful in monitoring the postoperative course.
 
@@
 
-\subsubsection{Compensatory Anti-inflammatory Response Syndrome (CARS)}
-The compensatory anti-inflammatory response syndrome is characterised by several features including reduction in lymphocyte numbers by apoptosis, decreased responsiveness of monocytes to cytokines, reduced number of human leukocyte antigen presenting receptors on monocytes, expression of cytokines that suppress Tumour Necrosis Factor (TNF) and clonal anergy.
+\subsubsection{Compensatory anti-inflammatory response}
+The compensatory anti-inflammatory response syndrome (CARS) is characterised by several features including reduction in lymphocyte numbers by apoptosis, decreased responsiveness of monocytes to cytokines, reduced number of human leukocyte antigen presenting receptors on monocytes, expression of cytokines that suppress Tumour Necrosis Factor (TNF) and clonal anergy.
 
@@
 They suggested that the 'double-hit' of surgery followed by anastomotic leak may result in an inflammatory response that is greater and more protracted and that this may explain the poorer cancer outcomes in these patients \parencite{mcardle_impact_2005}.
+
 In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP 2 weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications \parencite{katoh_anastomotic_2011}.
@@
 Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007}. 
-and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head.
-\parencite{kang_detrimental_2009}
+and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head. \parencite{kang_detrimental_2009}
 
@@
 
-\subsection{Relationship between systemic inflammation and comorbidity}
-\todo{Relationship between systemic inflammation and comorbidity}
+%\subsection{Relationship between systemic inflammation and comorbidity}
 
-\section{The Jaundiced Patient}
+\section{The jaundiced patient}
 
@@
 While this concept has been recently challenged, surgeons remain wary of operating on the severely jaundiced patient. 
-In fact, pancreaticoduodenectomy was initially described as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out.
+In fact, pancreaticoduodenectomy was initially described by Whipple as a two-stage procedure where the first stage involved a biliary bypass aimed at relieving obstructive jaundice before the second stage of resection was carried out \parencite{whipple_treatment_1935}.
 
@@
 They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimmetic agents. 
-Similar findings have been reported by other authors in animal studies.
-\parencite{binah_obstructive_1985,bomzon_systemic_1986}.
+Similar findings have been reported by other authors in animal studies. \parencite{binah_obstructive_1985,bomzon_systemic_1986}.
 
 The role of atrial natriuretic peptide (ANP) has also been investigated. 
-Obstructive jaundice is associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits \parencite{pereira_increased_1994}.
-Similar findings have been reported in humans as well.
-\parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}.
+Obstructive jaundice was associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits \parencite{pereira_increased_1994}. 
+Similar findings have been reported in humans as well. \parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}.
 
@@
 More recently, TNF-$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF-$\alpha$ agent restored myocardial contractility \parencite{yang_mechanisms_2010}.
-Obstructive jaundice is also associated with systemic hypotension and there has been increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree \parencite{green_systemic_1995, lefebvre_role_2009}.
-Bile acids can thus cause vasodilation by decreasing arterial tone and this may partly explain some of the haemodynamic adverse events that occur after surgery in jaundiced patients.
+Obstructive jaundice was also associated with systemic hypotension and there is increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree \parencite{green_systemic_1995, lefebvre_role_2009}.
+Bile acids can cause vasodilation by decreasing arterial tone and this may partly explain some of the haemodynamic adverse events that occur after surgery in jaundiced patients.
 
@@
 While expansion of the intravascular volume and avoiding dehydration by the judicious use of intravenous fluids has been recommended in avoiding renal dysfunction in jaundiced patients \parencite{parks_prospective_1994}, others have reported that relief of obstructive jaundice by restoring bile flow improves renal function independent of fluid therapy \parencite{padillo_randomized_2005}.
-Nonetheless, adequate perioperative fluid management, avoidance of hypotension, sepsis control and relief of obstructive jaundice play important roles in the prevention of renal failure in jaundiced patients.
+Perioperative fluid management, avoidance of hypotension and sepsis control is important in the prevention of renal failure in jaundiced patients.
 
@@
 %nehez_compromise_2002 - get other citations from this one
-\todo{Impact of jaundice on the immune system}
 
 Obstructive jaundice is associated with a wide range of deleterious effects on the immune system. 
-The absence of bile in the gut increases the proliferation of Gram negative bacteria.
-This in association with increased bacterial translocation, increased gut mucosal permeability and reduced endotoxin elimination by Kuppfer cells can predispose the jaundiced patient to septic complications.
+The absence of bile in the gut increases the proliferation of gram negative bacteria. 
+The absence of bile in the gut of bile duct ligated mice was associated with intestinal bacterial overgrowth and bacterial translocation \parencite{deitch_obstructive_1990}
+
+This in association with increased bacterial translocation, increased gut mucosal permeability and reduced endotoxin elimination by Kuppfer cells within the liver can predispose the jaundiced patient to septic complications \parencite{nehez_compromise_2002}.
 Intestinal permeability is normalised within a few weeks of restoring biliary drainage.
 
-Increased serum bilirubin and bile salts directly affect lymphocyte function as well as the response to pro-inflammatory cytokines.
+Increased serum bilirubin and bile salts directly affect lymphocyte function as well as the response to pro-inflammatory cytokines \parencite{scott-conner_pathophysiology_1994}
 Bile salts in higher than normal concentrations also induce apoptosis in hepatocytes. 
@@
 Obstructive jaundice is also associated with abnormal cytokine homeostasis predisposing the patient to impaired wound healing, sepsis and increased risk of peri-operative mortality. 
-For instance, concentrations of TNF-$\alpha$ and IL-6 are significantly elevated in jaundiced patients while the concentrations of IL-1 and IL-2 are decreased.
+For instance, concentrations of TNF-$\alpha$ and IL-6 are significantly elevated in jaundiced patients while the concentrations of IL-1 and IL-2 are decreased \parencite{padillo_cytokines_2001}.
 
-Biliary obstruction affects both cellular and humoral immunity.
-Endotoxins derived from the gut in bile duct ligated rats are thought to suppress cellular immunity.
-Moreover the clearance of immune-complexes may also be impaired in biliary obstruction.
+However, unlike in animal models, relief of obstructive jaundice by endoscopic biliary drainage is not associated with an improvement in immune function.
+On the contrary, it results in colonisation of the biliary tree by gut bacteria in nearly all instances of biliary stenting \parencite{kimmings_endotoxin_2000}.
 
@@
 However, the study does show that early surgery in the jaundiced patient with serum bilirubin $<$ 250 $\mu$mol/L is safe and not associated with increased morbidity or mortality as once feared.
+
 Mezhir and co-workers also reported a high (23\%) stent related complication rate in their retrospective study of 340 patients where 96\% of 201 patients in the biliary drainage group had plastic stents placed.
-However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage.
+However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage \parencite{mezhir_matched_2009}
 
@@
 
-\section{Body Composition, Obesity and Outcomes}
-\todo{Introduction - body composition}
-\subsection{Body Mass Index}
+\section{Body composition, obesity and outcomes}
+
+\subsection{Body mass index}
 
@@
 It is the most widely accepted measure of body composition, especially of body fat in spite of criticisms that it may not apply to all populations.
+
 BMI was first devised by Adolphe Quetelet and was formerly known as the Quetelet index \parencite{eknoyan_adolphe_2008}. 
@@
 
-\subsection{Obesity and Postoperative Complications}
+\subsection{Obesity and postoperative complications}
 
@@
 However, obesity was an independent predictor of minor complications including wound infections. 
-However, they noted that underweight patients had a five-fold increased risk of postoperative mortality.
-\parencite{mullen_impact_2008}
+However, they noted that underweight patients had a five-fold increased risk of postoperative mortality. \parencite{mullen_impact_2008}
 
@@
 In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen and co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
-This appears to confirm the existence of an 'obesity paradox' in surgical patients.
-\parencite{mullen_obesity_2009}
+This appears to confirm the existence of an 'obesity paradox' in surgical patients. \parencite{mullen_obesity_2009}
 
@@
 Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese patients undergoing surgery compared to non-obese patients. 
-However, there was no difference in operative time or length of stay in hospital.
-\parencite{benns_impact_2009}
+However, there was no difference in operative time or length of stay in hospital. \parencite{benns_impact_2009}
 
@@
 %lin_risk_2004, lermite_risk_2007 - Coranary artery disease is associated with pancreatic fistula
-\subsection{Obesity and Cancer Survival}
+\subsection{Obesity and cancer survival}
 
@@
 
-\subsection{Measuring Body Composition}
+\subsection{Measuring body composition}
 While it is possible to estimate the amount of most tissue types, the ones that are of clinical interest and are related to surgical and cancer outcomes are adipose tissue and skeletal muscle. 
@@
 Studies have also reported on the association between intra-pancreatic fat content and postoperative complications.
+
+\clearpage
+\section{Aims of the thesis}
+\vfill
+The overall aim of the thesis was to examine the inter-relationships between preoperative clinico-pathological characteristics including cardiopulmonary exercise physiology, obstructive jaundice, body composition and preoperative systemic inflammation and post-operative complications and the post-surgical systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
+\vfill
+More specifically the aims of the thesis were:
+\vfill
+To evaluate the clinical utility of preoperative cardiopulmonary exercise testing in predicting postoperative adverse events after pancreaticoduodenectomy.
+\vfill
+To examine the patient factors that are related to cardiopulmonary exercise physiology with particular attention to the effect of obstructive jaundice and body composition on aerobic capacity in patients undergoing pancreaticoduodenectomy.
+\vfill
+To examine the value of serial daily postoperative markers of systemic inflammatory response like white cell count, albumin 
+and C-reactive protein concentrations in the prediction of post-operative complications after pancreaticoduodenectomy.
+\vfill
+To examine the effect of preoperative systemic inflammation and poor aerobic capacity on the magnitude of the post-operative systemic inflammatory response after pancreaticoduodenectomy.
+\vfill

### 2015-08-22 Tables/c1_clavien_dindo.tex
Introduction - revision to section 1.3
- Spelling, grammar, style corrections
Review 1 - Introduction
- Fixed nearly all incomplete sections
- All todos sorted
- Multiple tables amended for formatting
- More references for jaundice, immunity
- Added table for POSSUM
- Fixed several overfull box issues with TOC

@@
 \begin{sidewaystable}[htbp]
-\caption{The Clavien-Dindo Classification of Surgical Complications}
+	\centering
+	\caption{The Clavien-Dindo classification of surgical complications.}
 	\label{table:clavien-dindo}	
-\centering
+	\renewcommand{\arraystretch}{1.7} %Increases space between rows
+	\setlength{\tabcolsep}{14pt} %sets the space between columns
 	\begin{tabular}{|l m{15cm}|}
@@
 		Grade       & Description \\ \hline
-
 		Grade I     & Any deviation from the normal postoperative course without the need for pharmacological treatment or surgical, endoscopic and radiological interventions.  \\
-
-Grade II      &   Requiring pharmacological treatment with drugs other than such allowed for grade I complications. Blood transfusions and total parenteral nutrition are also included.\\ 
-Grade III     &   Requiring surgical, endoscopic or radiological intervention\\
-                &  Grade III-a: - intervention not under general anesthesia\\
-                &  Grade III-b: - intervention under general anesthesia\\ 
-Grade IV       &  Life-threatening complication (including CNS complications)‡ requiring IC/ICU-management\\
-                &   Grade IV-a: - single organ dysfunction (including dialysis)\\
-                &   Grade IV-b: - multi organ dysfunction\\ 
-Grade V        &   Death of a patient\\ \hline
+		Grade II    & Requiring pharmacological treatment with drugs other than such allowed for grade I complications. Blood transfusions and total parenteral nutrition are also included.  \\
+		Grade III   & Requiring surgical, endoscopic or radiological intervention  \\
+		            & Grade III-a: - intervention not under general anaesthesia \\
+		            & Grade III-b: - intervention under general anaesthesia  \\
+		Grade IV    & Life-threatening complication (including CNS complications)‡ requiring  HDU/ICU-management  \\
+		            & Grade IV-a: - single organ dysfunction (including dialysis)     \\
+		            & Grade IV-b: - multi organ dysfunction \\
+		Grade V     & Death of a patient  \\ \hline
 		Suffix 'd': & If the patients suffers from a complication at the time of discharge,  the suffix  “d”  (for ‘disability’) is added to the respective grade of complication.
-This label indicates the need for a follow-up to fully evaluate the complication.\\ \hline
-
+This label indicates the need for a follow-up to fully evaluate the complication.
+\\ \hline
 	\end{tabular}
-
 \end{sidewaystable}

### 2015-08-22 Tables/c1_cpet_parameters.tex
Review 1 - Introduction
- Fixed nearly all incomplete sections
- All todos sorted
- Multiple tables amended for formatting
- More references for jaundice, immunity
- Added table for POSSUM
- Fixed several overfull box issues with TOC

@@
 	\label{table:spirometry}
-\begin{tabular}{| m{2cm}  m{2cm}  m{7cm} |}
+	\renewcommand{\arraystretch}{1.5} %Increases space between rows
+	\setlength{\tabcolsep}{10pt} %sets the space between columns
+	\begin{tabular}{| m{2cm}  m{2cm}  m{7.7cm} |}
 		\hline
@@
 		FEV1      & litres & Forced Expiratory Volume in 1 second \\
-	FEV1/FVC  & \%     & Tiffeneau-Pinelli[1] index           \\ \hline
+		FEV1/FVC  & \%     & Tiffeneau-Pinelli index              \\ \hline
 	\end{tabular}
@@
 	\label{table:cpet_parameters}
-\begin{tabular}{| m{2cm}  m{2cm}  m{7cm} |}
+	\begin{tabular}{| m{2cm}  m{2cm}  m{7.7cm} |}
 		\hline
 		Parameter                  & Units      & Description                         \\ \hline
-	\%peakVO2 & \%          & VO2 as a \% of predicted VO2Peak             \\
 		Load                       & Watts      & Exercise Workload                   \\
-	VE        & litres/min  & Ventilatory Equivalent                       \\
+		$\dot{V}_E$                & litres/min & Ventilatory Equivalent              \\
 		Vt                         & litres     & Tidal volume                        \\
-	VO2       & litres/min  & Absolute Oxygen uptake/consumption           \\
-	VO2/kg    & ml/(kg*min) & Corrected Oxygen uptake/consumption          \\
-	VE/VO2    &             & Ventilatory Equivalent for O$_2$             \\
-	VCO2      & litres/min  & Carbon-dioxide output                        \\
-	VE/VCO2   &             & Ventilatory Equivalent for CO$_2$            \\
+		$\dot{V}_{O_2}$            & litres/min & Absolute Oxygen uptake/consumption  \\
+		$\dot{V}_{O_2}$/kg         & ml//kg/min & Corrected Oxygen uptake/consumption \\
+		$\dot{V}_E/\dot{V}_{O_2}$  &            & Ventilatory Equivalent for $O_2$    \\
+		$\dot{V}_{CO_2}$           & litres/min & Carbon-dioxide output               \\
+		$\dot{V}_E/\dot{V}_{CO_2}$ &            & Ventilatory Equivalent for $CO_2$   \\
 		RER                        &            & Respiratory Exchange Ratio          \\
-	PETO2     & mmHg        & End Tidal O2                                 \\
-	PETCO2    & mmHg        & End Tidal CO2                                \\
-	O2Pulse   & ml/beat     & Oxygen pulse                                 \\
+		$PET_{O_2}$                & mmHg       & End Tidal $O_2$                     \\
+		$PET_{CO_2}$               & mmHg       & End Tidal $CO_2$                    \\
+		${O_2}$Pulse               & ml/beat    & Oxygen pulse                        \\
 		HR                         & beats/min  & Heart Rate                          \\
-	Bf        & /min        & Breathing Frequency                          \\
-	P(A-a)O2  & mmHg        & Alveolar-arterial PO2 difference             \\
-	Vd/Vt     &             & Physiologic dead space-to-tidal volume ratio \\
-	SBP       & mmHg        & Systolic blood pressure                      \\
-	DBP       & mmHg        & Diastolic blood pressure                     \\
-	O2sat     & \%          & Oxygen saturation                            \\ \hline
+		$B_f$                      & resps/min  & Breathing Frequency                 \\ \hline
 	\end{tabular}

### 2015-08-22 Tables/c1_isgps.tex
Minor changes to Introduction
- Include section labels for POPF, PPH
- vfill/hfill in ISGPS tables
Review 1 - Introduction
- Fixed nearly all incomplete sections
- All todos sorted
- Multiple tables amended for formatting
- More references for jaundice, immunity
- Added table for POSSUM
- Fixed several overfull box issues with TOC

@@
 \centering
+\footnotesize
 \caption{Postoperative pancreatic fistula: ISGPF definition.}
 \label{table:isgps_popf}
-\begin{tabular}{|m{4cm}|m{4cm}|m{5cm}|m{5cm}|}
+\renewcommand{\arraystretch}{1.2} %Increases space between rows
+%\setlength{\tabcolsep}{10pt} %sets the space between columns
+\begin{tabular}{|p{4.5cm}|p{4cm}|p{5cm}|p{5cm}|}
 	\hline
@@
 
+\vspace{0.7cm}
+
 \caption{Postpancreatectomy haemorrhage: ISGPS definition.}
 \label{table:isgps_pph}
-\begin{tabular}{|m{4cm}|m{4cm}|m{5cm}|m{5cm}|}
+\begin{tabular}{|m{4.5cm}|m{4cm}|m{5cm}|m{5cm}|}
 	\hline
@@
 	Time of onset, location, severity and clinical impact of bleeding & Early, intra- or extraluminal, mild                                              & Early, intra- or extraluminal, severe	or Late, intra- or extraluminal, mild                                                   & Late, intra- or extraluminal, severe                                                      \\
-	Clinical condition                                                & Well                                                                             & Often well/ intermediate, very rarely life-threatening                                                                        & Severely impaired, life-threatening                                                       \\
-	Diagnostic consequence                                            & Observation, blood count, ultrasonography and, if necessary, computed tomography & Observation, blood count, ultrasonography, computed tomography, angiography, endoscopy                                        & Angiography, computed tomography, endoscopy                                               \\
+	Clinical condition                                                & Well                                                                             & Often well/ intermediate, very rarely life-threatening                                                                        & Life-threatening, severely impaired                                                      \\
+	Diagnostic consequence                                            & Observation, blood count, ultrasonography and, if necessary, computed tomography & Observation, blood count, ultrasonography, endoscopy, computed tomography, angiography                                      & Angiography, endoscopy, computed tomography                                              \\
 	Therapeutic consequence                                           & No                                                                               & Transfusion of fluid/blood, intermediate care unit (or ICU), therapeutic endoscopy,† embolization, relaparotomy for early PPH & Localization of bleeding, angiography and embolization, (endoscopy†) or relaparotomy, ICU \\ \hline

### 2015-08-22 Chapters/CPET_Outcomes.tex
Review 2 - CPET Outcomes
- Changed all vo2 to correct format
- Fixed some captions to avoid toc problems

@@
 The test was stopped at maximum exercise tolerance, significant ischaemic changes on ECG or for other safety reasons. 
-The VO$_2$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. 
-Low VO$_2$AT was defined as oxygen consumption less than 10 ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that VO$_2$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
+The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986, sue_metabolic_1988} and ventilatory equivalents \parencite{sue_metabolic_1988} methods. 
+Low $\dot{V}_{O_2}$AT was defined as oxygen consumption less than 10 ml/kg/min based on work by Snowden and co-workers \parencite{snowden_submaximal_2010} who reported that $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min was associated with an increase in postoperative complications after major abdominal surgery.
 A detailed description of the cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
@@
 Cox proportional hazards regression analysis was used to study the relationship between preoperative risk factors and length of hospital stay. 
-Chi-square test was used to examine the relationship between complications and VO$_2$AT as a categorical variable. 
+Chi-square test was used to examine the relationship between complications and $\dot{V}_{O_2}$AT as a categorical variable. 
 Univariate binary logistic regression analysis with calculation of hazard ratios (HR) and 95\% confidence intervals was used to explore the association between perioperative clinico-pathological factors and receipt of adjuvant therapy.
@@
 
-The median VO$_2$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). 
-The VO$_2$AT was less than 10 ml/kg/min in 49 patients. 
-The distribution of VO$_2$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.
+The median $\dot{V}_{O_2}$AT was 10.3 ml/kg/min (inter-quartile range, IQR 8.8 - 11.6). 
+The $\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 49 patients. 
+The distribution of $\dot{V}_{O_2}$AT across the study cohort is shown in Figure \ref{fig:cpet_outcomes_dist_of_AT}.
 
@@
 	\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_dist_of_AT}
-	\caption{Distribution of VO$_2$AT across the study population.}
+	\caption{Distribution of $\dot{V}_{O_2}$AT across the study population.}
 	\label{fig:cpet_outcomes_dist_of_AT}
@@
 \subsection{Anaerobic threshold vs. complications}
-The relationship between VO$_2$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. 
-Patients with VO$_2$AT less than 10 ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042).
-While there was an association between low VO$_2$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). 
-There was no association between low VO$_2$AT and cardiopulmonary complications or postoperative mortality. 
+The relationship between $\dot{V}_{O_2}$AT and major postoperative adverse events including mortality is shown in Table \ref{table:cpet_outcomes_table2}. 
+Patients with $\dot{V}_{O_2}$AT less than 10 ml/kg/min had significantly greater incidence of postoperative pancreatic fistula (35.4\% vs.16\%, p=0.028) as well as major intra-abdominal abscesses (Clavien-Dindo Grade III - V, 22.4\% vs.7.8\%, p=0.042).
+While there was an association between low $\dot{V}_{O_2}$AT and grade of pancreatic fistula, this was not statistically significant (p=0.091). 
+There was no association between low $\dot{V}_{O_2}$AT and cardiopulmonary complications or postoperative mortality. 
 Major cardiopulmonary complications occurred more often in patients with major intra-abdominal adverse events including major intra-abdominal abscesses or Grade B and C pancreatic fistulae or haemorrhage than in patients who did not have these complications (5/31,16.1\% vs.
 2/69,2.9\%, p=0.017).
-Postoperative mortality was not associated with VO$_2$AT (HR 0.77, 95\% CI 0.16-3.61, p 0.737) or the POSSUM Physiology Score (HR 0.39, 95\% CI 0.07-2.12, p 0.277). 
+Postoperative mortality was not associated with $\dot{V}_{O_2}$AT (HR 0.77, 95\% CI 0.16-3.61, p 0.737) or the POSSUM Physiology Score (HR 0.39, 95\% CI 0.07-2.12, p 0.277). 
 Postoperative mortality was associated with postoperative pancreatic fistula (n=5), post-pancreatectomy haemorrhage (n=3), major intra-abdominal sepsis (n=6) and major cardiorespiratory complications (n=4) with 6 patients requiring radiological or operative intervention.
@@
 The relationship between preoperative clinico-pathological characteristics and length of postoperative stay in patients who were discharged from hospital (n=93) is shown in Table \ref{table:cpet_outcomes_table3}.
-On univariate analysis, age over 65 years (p=0.072) and low VO$_2$AT (p=0.010) were associated with prolonged postoperative stay. 
-On multivariate Cox proportional hazards regression analysis, VO$_2$AT less than 10 ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay.
-A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low and normal VO$_2$ATs is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. 
-Patients with a low VO$_2$AT stayed a median 6 days longer in hospital (14 versus 20 days, Mann-Whitney Test p=0.001). 
-There was no significant association between any of the preoperative factors including VO$_2$AT and length of critical care stay or number of critical care admissions.
+On univariate analysis, age over 65 years (p=0.072) and low $\dot{V}_{O_2}$AT (p=0.010) were associated with prolonged postoperative stay. 
+On multivariate Cox proportional hazards regression analysis, $\dot{V}_{O_2}$AT less than 10 ml/kg/min (hazard ratio 1.74, 95\% confidence intervals 1.14-2.65, p=0.010) was the only significant factor associated with prolonged postoperative stay.
+A Kaplan-Meier plot for the probability of remaining in hospital over time for patients with low or normal $\dot{V}_{O_2}$AT is shown in Figure \ref{fig:cpet_outcomes_km_at_los}. 
+Patients with a low $\dot{V}_{O_2}$AT stayed a median 6 days longer in hospital (14 versus 20 days, Mann-Whitney Test p=0.001). 
+There was no significant association between any of the preoperative factors including $\dot{V}_{O_2}$AT and length of critical care stay or number of critical care admissions.
 
@@
 \includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
-\caption{Kaplan-Meier Plot of postoperative length of stay in patients with VO$_2$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
+\caption{Kaplan-Meier Plot of postoperative length of stay in patients with $\dot{V}_{O_2}$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
 \label{fig:cpet_outcomes_km_at_los}
@@
 Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6).
-On binary logistic regression analysis, VO$_2$AT less than 10 ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).
+On binary logistic regression analysis, $\dot{V}_{O_2}$AT less than 10 ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).
 
@@
 \section{Discussion}
-The results of the present study show that a low VO$_2$AT is associated with prolonged postoperative stay in hospital, postoperative pancreatic fistula and intra-abdominal abscesses in patients undergoing major resections for pancreatic head lesions.
-The results of this study also show that patients with low VO$_2$AT are less likely to receive adjuvant therapy.
+The results of the present study show that a low $\dot{V}_{O_2}$AT is associated with prolonged postoperative stay in hospital, postoperative pancreatic fistula and intra-abdominal abscesses in patients undergoing major resections for pancreatic head lesions.
+The results of this study also show that patients with low $\dot{V}_{O_2}$AT are less likely to receive adjuvant therapy.
 
@@
 
-Patients with a low VO$_2$AT stayed longer in hospital after their operation. 
-While length of stay in hospital is influenced by multiple factors including postoperative complications, it would appear that patients with a low VO$_2$AT take longer to recover from the physiological stress placed by major pancreatic surgery and its sequelae.
+Patients with a low $\dot{V}_{O_2}$AT stayed longer in hospital after their operation. 
+While length of stay in hospital is influenced by multiple factors including postoperative complications, it would appear that patients with a low $\dot{V}_{O_2}$AT take longer to recover from the physiological stress placed by major pancreatic surgery and its sequelae.
 
-The incidence of pancreatic fistula was greater in patients with a low VO$_2$AT. 
+The incidence of pancreatic fistula was greater in patients with a low $\dot{V}_{O_2}$AT. 
 This association needs further evaluation taking into consideration other well-recognised risk factors for pancreatic fistula such as pancreatic texture, pancreatic duct size and intra-operative blood loss \parencite{braga_prognostic_2011, pratt_possum_2008, winter_1423_2006}.
@@
 There was a non-significant trend towards clinically relevant pancreatic fistulae (ISGPS Grades B and C) as well as a significant association with major intra-abdominal abscesses (Clavien-Dindo Grades 3-5 i.e., requiring intervention, associated with organ dysfunction requiring intensive care or resulting in mortality).
-This would suggest that complications in patients with low VO$_2$AT are more likely to be severe than in patients with normal VO$_2$AT. 
-However, there was no difference in mortality between patients with normal or low VO$_2$AT, indicating that multiple factors including preoperative patient fitness, local and operative factors, systemic inflammatory response, number of complications as well as perioperative critical care all play a role.
+This would suggest that complications in patients with low $\dot{V}_{O_2}$AT are more likely to be severe than in patients with normal $\dot{V}_{O_2}$AT. 
+However, there was no difference in mortality between patients with normal or low $\dot{V}_{O_2}$AT, indicating that multiple factors including preoperative patient fitness, local and operative factors, systemic inflammatory response, number of complications as well as perioperative critical care all play a role.
 
-The results of this study also show that patients with a low VO$_2$AT were less likely to receive adjuvant therapy. 
+The results of this study also show that patients with a low $\dot{V}_{O_2}$AT were less likely to receive adjuvant therapy. 
 Adjuvant therapy in patients undergoing pancreatic resections for cancer has been shown in multiple randomised trials to improve survival significantly \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009}
@@
 
-In the present study, VO$_2$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
-The proportion of patients with VO$_2$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.
-While several studies have shown that low VO$_2$AT and/or low VO$_2$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.
-Older and co-workers reported in 1993 that low VO$_2$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
+In the present study, $\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 49\% of patients and less than 11 ml/kg/min in 64\% of patients. 
+The proportion of patients with $\dot{V}_{O_2}$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.
+While several studies have shown that low $\dot{V}_{O_2}$AT and/or low $\dot{V}_{O_2}$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.
+Older and co-workers reported in 1993 that low $\dot{V}_{O_2}$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
 
-However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an VO$_2$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the VO$_2$AT did not predict postoperative adverse events less after oesophagectomy.
-The lack of association between low VO$_2$AT and cardiopulmonary complications in this study may have been due to two reasons. 
+However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the $\dot{V}_{O_2}$AT did not predict postoperative adverse events less after oesophagectomy.
+The lack of association between low $\dot{V}_{O_2}$AT and cardiopulmonary complications in this study may have been due to two reasons. 
 Major cardiopulmonary complications occurred more often in association with major intra-abdominal adverse events which are determined largely by pancreatic morphology and local anatomy \parencite{braga_prognostic_2011}.
@@
 
-The results of this study are consistent with the findings by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with VO$_2$AT less than 10.1 ml/kg/min.
-However, this study did not report the association between VO$_2$AT and receipt of adjuvant therapy.
+The results of this study are consistent with the findings by Ausania and co-workers \parencite{ausania_effects_2012} who reported increased incidence of pancreatic fistula and prolonged postoperative stay in patients with $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min.
+However, this study did not report the association between $\dot{V}_{O_2}$AT and receipt of adjuvant therapy.
 

### 2015-08-23 Chapters/BodyComp.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - Body Comp
- Major changes to text
- Spelling checked
- References checked
- No compile erros
- Couple of overfull boxes that cant be fixed

@@
-% Chapter 04
 \chapter{An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing major pancreatic surgery.}
 \label{ch_bodycomp}
-
 \lhead{Chapter \ref{ch_bodycomp}. \emph{CPET and Body Composition}} % This is for the header on each page - perhaps a shortened title
-
 \clearpage
 %----------------------------------------------------------------------------------------
-\todo{Consider including relationship with systemic inflammation}
-
-zotero - richards-relationships-2012
 
@@
 \subsection{Role of preoperative CPET}
-The role of cardiopulmonary exercise testing in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery has become well established.
-A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both as measured at cardiopulmonary exercise testing is associated with increased morbidity and mortality after major surgery including bariartic\parencite{mccullough_cardiorespiratory_2006}, pancreatic\parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes}], liver \parencite{epstein_aerobic_2004}, cardiothoracic\parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery \parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011}.
-CPET is now routinely used as part of the preoperative processes used to select patients for surgery as well as to help in decision making regarding preoperative care including the need for additional tests, preoperative and intraoperative optimisation, admission to critical care and postoperative care.
-Patients are sometimes denied surgery if their performance at cardiopulmonary exercise testing is felt to be poor based on currently available evidence.
+Cardiopulmonary exercise testing has become a useful tool in the preoperative evaluation and risk assessment/stratification of patients undergoing major thoracic and abdominal surgery. 
+A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both was associated with increased morbidity and mortality after major surgery including bariatric \parencite{mccullough_cardiorespiratory_2006}, pancreatic \parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes}], liver \parencite{epstein_aerobic_2004}, cardiothoracic \parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery \parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011}.
+It is now routinely used as part of the preoperative work-up to aid in risk assessment and patient selection.
+Cardiopulmonary exercise tests also help in identifying patients who may benefit from optimisation and more focussed care before, during and after major pancreatic surgery.
+Patients are sometimes denied surgery if their aerobic fitness as determined by cardiopulmonary exercise testing is felt to be too poor based on currently available evidence.
 
-\subsection{The pathophysiological basis of CPET}
-Aerobic fitness, as defined by the ability to perform physical exercise, is dependant on and often limited by the ability of the cardiorespiratory and circulatory systems (henceforth simply the cardiorespiratory system) to supply O2 to skeletal muscles at times of increased demand as well as remove the main end product of aerobic metabolism, namely CO2.
-Several factors play an important role in this increased response of the cardiorespiratory system.
-The most important factor is an increase in cardiac output which in healthy adults can increase by upto six-fold during exercise.
-Aside from increased stroke volume and heart rate, the redistribution of blood volume from the splanchnic circulation increases venous return to the heart.
-A consequent increase in pulmonary blood flow and skeletal blood flow occurs which in turn is assisted by vasodilation in these circulatory beds.
+\subsection{Factors affecting aerobic fitness}
+Aerobic fitness, as defined by the ability to perform physical exercise, is dependant on and often limited by the ability of the cardiorespiratory and circulatory systems to supply oxygen to skeletal muscles at times of increased demand as well as remove the main end product of aerobic metabolism, $CO_2$.
 
-Oxygenation of the increased pulmonary blood flow and removal of the excess CO2 generated by aerobic exercise is effected by increased minute ventilation as a result of increase in its constituent factors namely respiratory rate and tidal volume.
-Oxygenation of skeletal muscle is further dependant on numerous other factors including the oxygen carrying capacity of blood (primary determinant being haemoglobin), adequate peripheral circulation and the ability of the mitochondria within the skeletal muscle to utilise the oxygen that is being delivered to them.
+Several factors play an important role in meeting this increased metabolic demand. 
+These include a normal cardiac and respiratory response to exercise and the oxygen carrying capacity of blood primarily determined by haemoglobin levels. 
+Peripherally, an adequate circulatory response to local humoral factors and central autonomic control, capillary density within the skeletal muscle and the ability of skeletal muscle to utilise oxygen, which is in turn determined by mitochondrial mass and function are important factors.
 
-It is clear that limitations in the patient's physiology resulting in inadequate or inappropriate response in any of the above mentioned factors will result in overall limitation of their aerobic fitness.
-Cardiopulmonary exercise testing allows the accurate measurement of most of these factors either directly or indirectly during dynamic exercise thus allowing identifying not only limitations in aerobic fitness but also the cause for such limitation.
+Inadequate or inappropriate response of any of the above mentioned factors will result in overall limitation of oxygen delivery or utilisation and aerobic fitness. 
+Cardiopulmonary exercise testing allows the accurate measurement of most of these factors either directly or indirectly during dynamic exercise.
+This allows identifying not only limitations in aerobic fitness but also the cause for such limitation.
 
-\subsection{Factors influencing aerobic fitness}
-A low $\dot{V}_{O_2}$ has universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems to increased oxygen demand during exercise.
+A low $\dot{V}_{O_2}$ has universally been attributed to low aerobic fitness due to an inadequate response of the cardiovascular and respiratory systems. 
 This is often thought to be due to cardiorespiratory disease, overt or sub-clinical. 
-Occasionally, other factors such as anaemia, peripheral vascular disease and rarely, mitochondrial diseases have been recognised as factors contributing to low $\dot{V}_{O_2}$ or abnormalities in other parameters measured at cardiopulmonary exercise testing but these causes are less common in patients undergoing major abdominal surgery.
+Occasionally, other factors such as anaemia \parencite{agostoni_relationship_2010}, peripheral vascular disease and rarely, mitochondrial diseases \parencite{fernandez_use_2000} have been identified as contributing to low aerobic capacity.
 
@@
 In chapter \ref{ch_cpet_outcomes}, we reported that low $\dot{V}_{O_2}$AT in patients undergoing pancreaticoduodenectomy was associated with increased incidence of postoperative pancreatic fistula and prolonged hospital stay.
-We also reported that patients with a $\dot{V}_{O_2}$AT less than 10 ml/kg/min were less likely to receive postoperative adjuvant chemotherapy as a result of postoperative complications, prolonged hospital stay and possibly due to lack of physiological reserve after surgery.
-However, we also observed that high BMI was associated with a low $\dot{V}_{O_2}$AT independent of all other clinicopathological characteristics.
-Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain the very low levels of $\dot{V}_{O_2}$AT.
+We also reported that patients with a $\dot{V}_{O_2}$AT less than 10 ml/kg/min were less likely to receive postoperative adjuvant chemotherapy.
+In chapter \ref{ch_cpet_jaundice}, while examining the relationship between preoperative clinico-pathological characteristics and cardiopulmonary exercise physiology, we observed that a high body mass index was associated with a low $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak independent of all other factors.
+Moreover, most of our patients did not have overt cardiac or respiratory comorbidity to explain these findings.
 
-The aim of the present study was to explore the association between body composition, total body weight and the physiological parameters measured at cardiopulmonary exercise testing.
+We hypothesised at the end of chapter \ref{ch_cpet_jaundice} that low $\dot{V}_{O_2}$ values in obese patients may be a result of the calculations involved rather than due to true cardiopulmonary dysfunction.
+The aim of the present study was to examine the relationship between cardiopulmonary exercise testing and body composition in patients undergoing pancreaticoduodenectomy in an attempt to explain the strong negative correlation between low $\dot{V}_{O_2}$ and body mass index.
 
 \clearpage
-\section{Methods}
+\section{Patients and methods}
 
@@
 All data were recorded in a prospectively maintained database. 
-Data was collected on demographics, preoperative clinicopathological characteristics including blood tests, body mass index, weight, height and the underlying surgical pathology.
+Data was collected on demographics, preoperative clinico-pathological characteristics including blood tests, body mass index, weight, height and the underlying surgical pathology. 
 Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. 
@@
 \label{sec:bodycomp_calculation}
-Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published and well established methods \parencite{bredella_comparison_2010,shen_total_2004}.
+Computed tomography of the abdomen, performed as part of the routine preoperative staging, was used to calculate body composition based on previously published methods \parencite{bredella_comparison_2010,shen_total_2004}.
 
@@
 The coronal and sagittal reconstructions were used to accurately identify the L3 and L4 vertebrae. 
-The CT window width (W) was set at 400 and the level/center (C) was set at 40.
-This allowed tissue between -160 HU and +240 HU to be represented in grayscale with adequate contrast between tissues of interest.
-The cross-sectional images at these levels where then exported as gray-scale bitmap images.
-The scale in millimeters was included with every image.
-A representative image is shown in Fig.
-1.The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org).
+The CT window width (W) was set at 400 and the level/centre (C) was set at 40. 
+This allowed tissue between -160 Hounsfield units and +240 Hounsfield units to be represented in grayscale with adequate contrast between tissues of interest. 
+The cross-sectional images at these levels where then exported as grayscale bitmap images. 
+The scale in millimetres was included with every image. 
+A representative image is shown in Fig. \ref{fig:bc_ct_csa}.
+The GNU Image Manipulation Program (GIMP), an advanced, free, open-source, raster graphics editor was used for analysis of all images (www.gimp.org). 
 The use of GIMP to analyse cross-sectional imaging for body composition has been described previously although by using a different technique to what has been employed by us \parencite{anblagan_measurement_2013}.
@@
 
-The next step involved standardising the scale of all images by dividing the length of the scale on every image by the number of pixels along the scale thus providing a length in millimetres for each pixel in each image.
+The next step involved standardising the scale of all images by dividing the length of the scale on every image by the number of pixels along the scale. 
+This resulted in a length in millimetres for each pixel in each image. 
 As pixels on a CT image are square, the area of each pixel was calculated as a square of its length.
 
-The Fuzzy Select (Magic Wand) tool was used to select contiguous areas of similar colour while simultaneously using visual confirmation that the correct anatomical structures had been selected without overspill into unwanted areas.
-The number of pixels within the selection was obtained using the 'Histogram' dialog window and entered into an excel spreadsheet against the selected area of interest.
-The area in mm2 was calculated by multiplying the number of pixels by the area of each pixel.
+The Fuzzy Select (Magic Wand) tool was used to select contiguous areas of similar colour while simultaneously confirming visually that the correct anatomical structures had been selected without overspill into adjoining unwanted tissue areas.
+The number of pixels within the selection was obtained using the 'Histogram' dialog window and entered into a Microsoft Excel spreadsheet.
+The area was calculated by multiplying the number of pixels by the area of each pixel.
 
@@
 
-The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig.
-\ref{fig:bc_ct_gimp} on p \pageref{fig:bc_ct_gimp}.
+The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig. \ref{fig:bc_ct_gimp} on p\pageref{fig:bc_ct_gimp}. 
 The total cross-sectional area of the abdomen at the level of L3/L4 was calculated by first selecting all the empty space outside the image followed by inverting this selection. 
-This is depicted in Fig.
-\ref{fig:bc_ct_csa}.
-Subcutaneous fat in the image was selected using the Fuzzy Select tool (if necessary by choosing multiple times and removing any unnecessary areas) as depicted in Fig.
-\ref{fig:bc_ct_sat}.
-The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig.
-\ref{fig:bc_ct_vat} and Fig.
-\ref{fig:bc_ct_sm} respectively.
+This is depicted in Fig. \ref{fig:bc_ct_csa}. 
+Subcutaneous fat in the image was selected using the Fuzzy Select tool (if necessary by choosing multiple times and removing any unnecessary areas) as depicted in Fig. \ref{fig:bc_ct_sat}. 
+The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig. \ref{fig:bc_ct_vat} and Fig. \ref{fig:bc_ct_sm} respectively. 
 Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the area under selection as shown in the insets in each of the images.
@@
 
-	\caption{Selection of components of body composition from CT images using GIMP.}($^*$ The selected area has been removed for representation purposes. The inset confirms the area selected.)
+	\caption{Selection of components of body composition from CT images using GIMP.}($^*$ The selected area has been removed for representation purposes. 
+The inset confirms the area selected.)
 	\label{fig:bc_ct_gimp}
@@
 The average of each parameter measured between the first and second minute (6 readings) was treated as the rest value. 
-Anaerobic threshold was identified using previously established methods and all corresponding parameters at this point were recorded \parencite{beaver_new_1986,sue_metabolic_1988}.
+Anaerobic threshold was identified using previously established methods \parencite{beaver_new_1986,sue_metabolic_1988} and all corresponding parameters at this point were recorded. 
 Peak exercise was identified by the maximum oxygen consumption recorded towards the end of the exercise period and all other parameters recorded at this point were considered as peak exercise values.
 
+In order to distinguish between absolute values and values scaled to total body weight, the terms `absolute $\dot{V}_{O_2}$' and `corrected $\dot{V}_{O_2}$' are used throughout this chapter. 
+When $\dot{V}_{O_2}$ has been scaled to an alternate factor, this is stated clearly.
+
+\subsection{Estimation of lean body mass}
+\label{sec:elbm}
+Estimated lean body mass (eLBM) was calculated using the Boer formula \parencite{boer_estimated_1984} as shown below.
+\begin{equation} \label{eq:elbm_men}
+	Men:\ eLBM = 0.407 \times weight(kg) + 0.267 \times height(cm) - 19
+\end{equation}
+\begin{equation} \label{eq:elbm_women}
+	Women:\ eLBM = 0.252 \times weight(kg) + 0.473 \times height(cm) - 48.3
+\end{equation}
+
 \subsection{Statistics} 
-
-All analyses were performed using the SPSS statistical package for Microsoft Windows (version 22 ).
-Comparison between body composition and cardiopulmonary exercise testing parameters was done using Partial correlations controlling for the effect of gender (and/or age).
-All p-values reported are two-sided.
-The relationship between body composition and preoperative clinico-pathological characteristics (as categorical variables) was analysed using the Mann-Whitney U test for variables with two categories or the Kruskal-Wallis test for variables with more than two categories.
-Previously established cut-offs were used for categorising continuous variables where applicable.
+Comparison between body composition and cardiopulmonary exercise testing parameters was done using Partial correlations controlling for the effect of sex (and/or age). 
+The relationship between body composition and preoperative clinico-pathological characteristics (as categorical variables) was analysed using non-parametric tests. 
+The Mann-Whitney U test for variables with two categories or the Kruskal-Wallis test for variables with more than two categories. 
+Previously established thresholds were used for categorising continuous variables where applicable. 
+All reported p-values are two-sided. 
 The level of significance was set at $p<0.05$.
+SPSS software (Version 22.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.
 
@@
 
-\subsection{Body composition and Clinico-pathological characteristics}
+\subsection{Body composition and clinico-pathological characteristics}
 Eighty-two patients (52 male) were included in the study. 
@@
 
-\subsection{Body composition differences between the sexes with increasing BMI}
+\subsection{Body composition vs. BMI/sex}
 The body composition differences between male and female patients with increasing body mass index is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. 
@@
 The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. 
-Moreover, the distribution of the adipose tissue differed between males and females with visceral adipose tissue contributing more to weight in obese males (43\% VAT vs.
+Moreover, the distribution of the adipose tissue differed between males and females.
+Visceral adipose tissue contributed more to weight in obese males (43\% VAT vs.
 35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (18\% VAT vs.
@@
 
-\subsection{Correlation with Pulmonary Function Tests}
+\subsection{Body composition vs. pulmonary function tests}
 
@@
 FEV1/FVC was not correlated with any of the body composition components.
-\todo{move this to discussion}
+
 This would indicate that pulmonary function was dependent on skeletal muscle area while FEV1/FVC, a calculated index to quantify restrictive or obstructive lung disease, was not associated with skeletal muscle area.
@@
 
-\subsection{Correlation with Exercise Load}
+\subsection{Body composition vs. exercise load}
 
@@
 
-\subsection{Correlation with Oxygen consumption}
+\subsection{Body composition vs. oxygen consumption}
 
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal}
-		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
+		\caption{$\dot{V}_{O_2}$AT vs. 
+Skeletal Muscle}
 		\label{fig:bc_scatter_VO2_skeletal}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
-		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
+		\caption{$\dot{V}_{O_2}$AT vs. 
+Total Adipose Tissue}
 		\label{fig:bc_scatter_VO2_TAT}
@@
 Absolute $\dot{V}_{O_2}$ (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold. 
-However, when it was corrected for total body weight, there was a strong correlation between corrected $\dot{V}_{O_2}$ (mls/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
+However, corrected for total body weight, there was a strong negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
 
@@
 
-$O_2$Pulse at the anaerobic thresold and peak exercise was strongly associated with skeletal muscle area with weaker positive correlations with visceral fat and total fat content.
+$O_2$Pulse at the anaerobic threshold and peak exercise was strongly associated with skeletal muscle area with weaker positive correlations with visceral fat and total fat content. 
 There was no correlation between $O_2$Pulse and subcutaneous adipose tissue at any phase of exercise.
@@
 
-\subsection{Scaling $\dot{V}_{O_2}$ to body composition and other factors}
+\subsection{Scaling oxygen consumption to body composition and other factors}
+Absolute $\dot{V}_{O_2}$ at the anaerobic threshold was corrected for weight, the square of height, body mass index, skeletal muscle area and estimated lean body mass.
+Calculation of estimated lean body mass is shown in Section \ref{sec:elbm}.
+The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex (Table \ref{table:bc_at_new_indices} on p\pageref{table:bc_at_new_indices}).
+The results from Table \ref{table:bc_cpet} comparing absolute $\dot{V}_{O_2}$ and corrected $\dot{V}_{O_2}$ at the anaerobic threshold with body composition are reproduced in the first two columns of this table to allow comparison with the new indices.
 
-Absolute $\dot{V}_{O_2}$ at the anaerobic threshold was corrected for weight, the square of height, body mass index, skeletal muscle area and estimated lean body mass.
-Estimated lean body mass (eLBM) was calculated using the Boer formula \parencite{boer_estimated_1984} as shown below.
-
-Men:\ eLBM = 0.407*weight(kg) + 0.267*height(cm) - 19.2
-
-Women:\ eLBM = 0.252*weight(kg) + 0.473*height(cm) - 48.3
-
-The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex.
-These results are shown in Table \ref{table:bc_at_new_indices} on p\pageref{table:bc_at_new_indices}.
-The results from Table \ref{table:bc_cpet} comparing absolute $\dot{V}_{O_2}$ and corrected $\dot{V}_{O_2}$ at the anaerobic threshold with body composition are reproduced in the first two columns of this table to allow comparison with the new indices.
 $\dot{V}_{O_2}$ scaled for skeletal muscle area had no significant correlation with adipose tissue, height, weight or BMI but had a negative correlation with skeletal muscle area. 
-However, $\dot{V}_{O_2}$ corrected for the square of patient's height had no correlation with adipose tissue but retained a postive correlation with skeletal muscle area.
-However, there was also a positive correlation with weight and BMI.
+$\dot{V}_{O_2}$ corrected for the square of patient's height had no correlation with adipose tissue but retained a positive correlation with skeletal muscle area as well as a positive correlation with total body weight and BMI.
 
 However, when absolute $\dot{V}_{O_2}$ was corrected for lean body mass, there was no correlation with any of the body composition indices or with height, weight or body mass index. 
-The negative correlation that occurs between adipose tissue and $\dot{V}_{O_2}$ corrected for total body weight does not occur when $\dot{V}_{O_2}$ is corrected for lean body mass (Fig.
+The negative correlation between adipose tissue and $\dot{V}_{O_2}$ corrected for total body weight did not occur when $\dot{V}_{O_2}$ was corrected for lean body mass (Fig.
 \ref{fig:bc_scatter_VO2_TAT_elbm}).
-Moreover, the correlation with skeletal muscle area remains unchanged regardless of whether $\dot{V}_{O_2}$ is corrected for total body weight or lean body mass (Fig.
+Moreover, the correlation with skeletal muscle area remained unchanged regardless of whether $\dot{V}_{O_2}$ was corrected for total body weight or lean body mass (Fig.
 \ref{fig:bc_scatter_VO2_skeletal_elbm}).
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal_elbm}
-		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
+		\caption{$\dot{V}_{O_2}$AT vs. 
+Skeletal Muscle}
 		\label{fig:bc_scatter_VO2_skeletal_elbm}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT_elbm}
-		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
+		\caption{$\dot{V}_{O_2}$AT vs. 
+Total Adipose Tissue}
 		\label{fig:bc_scatter_VO2_TAT_elbm}
@@
 The results of this study demonstrate that the most important cardiopulmonary exercise test parameters used for preoperative risk evaluation in surgery are influenced significantly by the patient's body composition.
-Both $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak are significantly affected by patient's body habitus when scaled using total body weight.
+Both $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak were significantly affected by patient's body habitus when scaled using total body weight. 
 Furthermore, the results suggest that there may be alternate methods of scaling of $\dot{V}_{O_2}$ that removes the effect of body composition on these parameters. 
-The results also demonstrate the disproportionate decrease in skeletal muscle area with associated increased in adipose tissue with increasing body mass index. 
-Taken together, these results would suggest that caution should be exercised in interpreting $\dot{V}_{O_2}$ (corrected for total body weight, mls/kg/min) in patients with a raised BMI as abnormally low values in these patients may not necessarily be due to poor aerobic fitness or lack of cardiopulmonary reserve.
+The results also demonstrate the disproportionate decrease in skeletal muscle area with associated increase in adipose tissue with increasing body mass index.
+
+Taken together, these results would suggest that caution should be exercised in interpreting $\dot{V}_{O_2}$ (corrected for total body weight, ml/kg/min) in patients with a raised BMI.
+Abnormally low values in these patients may not necessarily be due to poor aerobic fitness or lack of cardiopulmonary reserve and may simply be due to current convention used for scaling/representing $\dot{V}_{O_2}$.
 
@@
 
-Increased physical activity such as during cardiopulmonary exercise testing results in greater oxygen consumption ($\dot{V}_{O_2}$).
+Increased physical activity such as during cardiopulmonary exercise testing or during and after surgery results in greater oxygen consumption ($\dot{V}_{O_2}$). 
 The increased demand for oxygen is primarily due to increased metabolic activity within the exercising skeletal muscle.
@@
 However, current convention is to report $\dot{V}_{O_2}$ measured at cardiopulmonary exercise testing according to the following formula: 
-\[Corrected\ \dot{V}_{O_2} (mls.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
+\[Corrected\ \dot{V}_{O_2} (ml.kg^{-1}.min^{-1}) = \frac{Absolute\ \dot{V}_{O_2}\ (litres.min^{-1}) * 1000}{Total\ body\ weight\ (kg)}\]
 
-In chapter \ref{ch_cpet_jaundice}, we reported that there was a significant negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min) at the anaerobic threshold as well as at peak exercise and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease (Tables \ref{table:cpet_oj_at_regression} and \ref{table:cpet_oj_peak_regression}).
+In chapter \ref{ch_cpet_jaundice}, we reported that there was a significant negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min) and the patient's body mass index in spite of no observable cardiopulmonary comorbid disease (Tables \ref{table:cpet_oj_at_regression} and \ref{table:cpet_oj_peak_regression}).
+This negative correlation existed both at anaerobic threshold and peak exercise.
 The results of the present study demonstrate that this negative correlation is consequent to the scaling convention used rather than due to any pathophysiological effect of obesity.
@@
 The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for total body weight further supports the argument that the corrected value for $\dot{V}_{O_2}$ under-reports aerobic capacity in obese patients.
-Moreover, the lack of correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight but statistically significant positive correlation between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest appear to suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.
+There was no correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight.
+However, a statistically significant, positive correlation was present between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest appear.
+These findings suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.
 
@@
 
-Seltzer reported in his 1940 study of 34 subjects, that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight.\parencite{seltzer_body_1940} Tanner in his article titled \textit{`Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation'}\parencite{tanner_fallacy_1949} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass.
+Seltzer reported in his 1940 study of 34 subjects, that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight \parencite{seltzer_body_1940}.
+Tanner in his article titled \textit{`Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation'} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass \parencite{tanner_fallacy_1949}.
 In a detailed analysis comparing $\dot{V}_{O_2}$ and body build, he concluded that \textit{'as the index wt./stature increases, O2/wt.
@@
 
-Batterham et al studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}.
+Batterham and co-workers studied 1314 apparently healthy men employed at the National Aeronautics and Space Administration Johnson Space Center in Houston, Texas \parencite{batterham_modeling_1999}. 
 The authors reported that as body mass increased, the proportion composed of fat-free mass decreased. 
@@
 They suggest that ideally estimates of fat-free mass should be used in the representation of oxygen consumption to allow more reliable comparison between subjects. 
-These findings are similar to our results where the proportion of skeletal muscle area decreasing significantly as body mass index increased with a greater decrease in females than males.
+These findings are similar to our results where the proportion of skeletal muscle area decreased significantly as body mass index increased with a greater decrease in females than males. 
 Moreover, the linear relationship between skeletal muscle area and absolute $\dot{V}_{O_2}$ has also been clearly demonstrated in our findings.
 
-We used the Boer formula \parencite{boer_estimated_1984}to estimate lean body mass and used this to normalise $\dot{V}_{O_2}$.
+We used the Boer formula \parencite{boer_estimated_1984} to estimate lean body mass and used this to normalise $\dot{V}_{O_2}$. 
 The resulting value for $\dot{V}_{O_2}$ showed no `spurious' correlations with body composition, height, total body weight or body mass index. 
-These findings are similar to the recommendations made by Janz et al who studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997\parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}.
-Aerobic capacity in the form of $\dot{V}_{O_2}$peak was evaluated annually in 126 children (mean age 10.3 years) for five years.
+These findings are similar to the recommendations made by Janz and co-workers who studied oxygen consumption and aerobic capacity in adolescents over several years as part of the Muscatine study and reported their findings in 1997 \parencite{janz_three-year_1997} and 1998 \parencite{janz_longitudinal_1998}.
+Aerobic capacity in the form of $\dot{V}_{O_2}$Peak was evaluated annually in 126 children (mean age 10.3 years) for five years. 
 Body composition changes were also tracked over this period. 
 They reported on the changes in body composition that occur over time and the differences in these changes between circum-pubertal boys and girls. 
-They reported on the significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$.
+They demonstrate significant difficulties in normalising $\dot{V}_{O_2}$ using total body mass and suggested that fat-free mass was the most appropriate variable for normalising $\dot{V}_{O_2}$. 
 They found that $\dot{V}_{O_2}$ normalised using total body mass underestimated aerobic fitness levels of heavier boys and girls. 
@@
 
+In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$Max, Maciejczyk and co-workers analysed the differing influence of body fat and lean body mass on aerobic performance in two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
+High body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ (ml/kg/min) when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
+However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. 
+They, similar to Goran and co-workers \parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.
+
 Our results demonstrate that there is no correlation between absolute oxygen consumption at rest or during exercise and the amount of adipose tissue. 
 This further emphasises the fact that adipose tissue is metabolically inactive during exercise although it contributes to increased work load. 
-Therefore, dividing absolute $\dot{V}_{O_2}$ by total body weight in overweight/obese subjects whose extra weight is largely due to adipose tissue results in lower corrected $\dot{V}_{O_2}$ as we have demonstrated here.
-Goran et al reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}.
-They reported on $\dot{V}_{O_2}$max in obese women before and after weight loss.
-$\dot{V}_{O_2}$max corrected for total body weight was significantly lower in the obese state while $\dot{V}_{O_2}$max corrected for fat-free mass did not change significantly after weight loss.
-They also reported that the limiting factor in the obese state was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities.
-This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may increase the exercise load.
+Therefore, dividing absolute $\dot{V}_{O_2}$ (l/min) by total body weight in overweight/obese subjects whose extra weight is largely due to adipose tissue would result in lower corrected $\dot{V}_{O_2}$ (ml/kg/min) as we have demonstrated here.
 
-These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001,  lemaitre_maximum_2006,savonen_current_2012, krachler_cardiopulmonary_2014}.
+Goran and co-workers reported that total body fat did not affect maximal aerobic capacity \parencite{goran_total_2000}. 
+They reported on $\dot{V}_{O_2}$Max in obese women before and after weight loss. 
+$\dot{V}_{O_2}$Max corrected for total body weight was significantly lower in obese subjects while $\dot{V}_{O_2}$Max corrected for fat-free mass did not change significantly after weight loss. 
+They also reported that the limiting factor in obese patients was not the cardio-respiratory system but the fact that it was more difficult for obese individuals to do the same amount of work as a normal weight person in weight-bearing activities.
+This is likely due to the extra fat mass in these individuals that did not contribute to aerobic capacity but instead may have increased the exercise load in addition to that set by the investigator during the exercise.
+
+These findings have been replicated by several other authors in different subject groups \parencite{loftin_scaling_2001, lemaitre_maximum_2006, savonen_current_2012, krachler_cardiopulmonary_2014}. 
 Several of the above studies also recommended using allometric scaling to avoid the confounding effects of total body weight. 
 However, this has not gained widespread clinical use.
-
-In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$max, Maciejczyk and coworkers analysed the differing influence of body fat and lean body mass on aerobic performance in two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
-They reported that high body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
-However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups.
-They, similar to Goran et al\parencite{goran_total_2000}, recommend that $\dot{V}_{O_2}$ be normalised to lean body mass rather than total body weight.
 
@@
 
-Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
-While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome.
-However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tool concluded that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure.
-This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
+Older and co-workers in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
+While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome and that a low $\dot{V}_{O_2}$AT represented cardiac failure.
+This association is repeated in their later work on 548 patients which also showed a clear association between low $\dot{V}_{O_2}$AT (ml/kg/min) and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
 The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.
 
-Swart and Carlisle reported that $\dot{V}_{O_2}$AT < 11mls/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}.
-However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT groups (24\% vs 51\%).
-The average corrected $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 mls/kg/min while in women it was 9.81 mls/kg/min.
-In a study by Wilson et al that reported cardiopulmonary exercise testing predicted outcome in major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}.
+Swart and Carlisle als reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. 
+However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT group (24\% vs 51\%). 
+The average corrected $\dot{V}_{O_2}$AT in men calculated from the data presented in their paper was 11.02 ml/kg/min while in women it was 9.81 ml/kg/min.
+
+In a study by Wilson and co-workers which reported that cardiopulmonary exercise testing predicted outcome after major elective intra-abdominal surgery, the proportion of females in the low $\dot{V}_{O_2}$AT group was 51\% while it was 28\% in the group with normal AT \parencite{wilson_impaired_2010}.
 There was no data presented on body mass index in this study. 
@@
 
-Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery, reducing complications \parencite{jones_effects_2007, dronkers_preoperative_2010, valkenet_effects_2010, gillis_prehabilitation_2014} and mitigating the effects of neoadjuvant oncological therapy\parencite{west_effect_2015}.
-Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients \parencite{west_effect_2015}.
+Where time to surgery is not critical, prehabilitation has gained an increasingly important role in optimising patients for surgery.
+Prehabilitation has been defined as the `process of enhancing functional capacity of the individual to enable him or her to withstand the stressor of inactivity' \parencite{topp_effect_2002}. 
+Jones and co-workers reported an increase in $\dot{V}_{O_2}$Peak by an average 2.4 ml/kg/min when patients were placed on an aerobic exercise regimen before undergoing surgery for lung cancer \parencite{jones_effects_2007}.
+Kim and co-workers also demonstrated a significant improvement in exercise loads achieved as well as sub-maximal oxygen consumption in patients who undertook a 4-week exercise program before major bowel surgery \parencite{kim_responsive_2009}.
+Moreover, deterioration in aerobic capacity during prehabilitation was reported to be associated with increased incidence of postoperative complications \parencite{mayo_impact_2011}.
 
-The success of such prehabilitation programs must not be determined solely by $\dot{V}_{O_2}$ corrected for total body weight.
-Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O^2$Pulse, tidal volume\parencite{jones_effects_2007} or maximal exercise load may provide more reliable evidence of improvement in aerobic capacity.
-The combination of nutritional assessment and advice, home-based exercise programs and weight loss will help improve aerobic capacity and eventually postoperative outcomes.
+However, other authors have not been able to reproduce these findings. 
+Patient compliance and difficulties in trying to modify lifestyle prior to surgery were considered important factors for failure of prehabilitation \parencite{carli_randomized_2010, dronkers_preoperative_2010}.
 
+Cardiopulmonary exercise testing has been reported to be a useful objective measure of the impact of prehabilitation in surgical patients. 
+It was used to demonstrate the decline in $\dot{V}_{O_2}$AT associated with neo-adjuvant chemoradiotherapy in patients with rectal cancer who did not undertake an exercise program.
+However, $\dot{V}_{O_2}$AT improved in the intervention group who undertook a 6 week exercise program \parencite{west_effect_2015}.
+
+However, the results presented in this study demonstrate that the success of such prehabilitation programs must not be determined solely by $\dot{V}_{O_2}$ corrected for total body weight. 
+Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O^2$Pulse, tidal volume \parencite{jones_effects_2007} or maximal exercise load \parencite{kim_responsive_2009} may provide more reliable evidence of improvement in aerobic capacity.
+
+The combination of nutritional assessment and advice, home-based exercise programs and weight loss may help improve aerobic capacity and eventually postoperative outcomes. 
+Such preoperative interventions may also be associated with improvements in body composition with increase in skeletal muscle mass and reduction in adipose tissue.
+This may result in apparent improvements in cardiopulmonary exercise test parameters that are scaled using total body weight.
 Further studies are required to validate the findings presented here as well as to identify optimal scaling methods for cardiopulmonary exercise test parameters. 
 This will improve the clinical utility of cardiopulmonary exercise testing in the preoperative risk assessment as well as optimisation of patients undergoing major surgery.
-
-%This needs to be expanded significantly.
-
-%Mention studies that have used VE/VCO2. 
-%I think there is a study in bariatrics where bmi and at were independent predictors. 
-%Mention studies that used predicted VO2 or absolute $\dot{V}_{O_2}$ where calculated AT was not useful. 
-%This should give me at least another 2 paragraphs

### 2015-08-23 Tables/bc_at_new_indices.tex
Review 1 - Body Comp
- Major changes to text
- Spelling checked
- References checked
- No compile erros
- Couple of overfull boxes that cant be fixed

@@
 \begin{sidewaystable}[p]
-	\caption{The relationship between body composition, body habitus and $\dot{V}_{O_2}$AT scaled using different factors in patients undergoing major pancreatic surgery.}
+	\caption{The relationship between body composition, body habitus and $\dot{V}_{O_2}$AT scaled using different factors (controlled for sex) in patients undergoing major pancreatic surgery.}
 	\label{table:bc_at_new_indices}

### 2015-08-23 Chapters/Discussion.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review - Discussion
- Committing small changes
- Added sections to discussion

@@
 
-The overall aim of the thesis was to examine the relationships between cardiopulmonary exercise physiology, obstructive jaundice, systemic inflammation, body composition and short and long-term outcomes after major pancreatic surgery.
+The overall aim of the thesis was to examine the inter-relationships between preoperative clinico-pathological characteristics including cardiopulmonary exercise physiology, obstructive jaundice, body composition and preoperative systemic inflammation and post-operative complications and the post-surgical systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
 
-A secondary aim of the thesis was to clarify the factors that affect cardiopulmonary exercise physiology and provide a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
+The present work also aims to examine the factors that affect cardiopulmonary exercise physiology in order to enable a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
+
 %################################################################################
-% Cardiopulmonary exercise testing
+\section{Clinical utility of CPET in predicting postoperative complications}
 %################################################################################
@@
 %################################################################################
-% Obstructive Jaundice
+\section{Determinants of aerobic capacity}
 %################################################################################
@@
 %################################################################################
+%Benefits of specialisation
+%################################################################################
+
+%Better rescue rates after complications gouma_rates_2000
+
+\section{Improving outomes after pancreaticoduodenectomy}
+
+\section{Prehabilitation}
+
+\section{Future directions}
+%################################################################################
 %Prehabilitation
@@
 In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
-The authors report that major complications (31.3 vs.
-26.2\%) and mortality (4.1 vs.
-2.3\%) were greater in these patients.\parencite{tzeng_morbidity_2014}
+The authors report that major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in these patients.\parencite{tzeng_morbidity_2014} 
 The authors recommend that surgeons identify these patients early, institute interventions to optimise their comorbidities and enrol these patients for prehabilitation.

### 2015-08-23 Appendices/AppendixAccessDatabase.tex
Review - Minor changes to Appendices
- Modify short title
- insert \clearpage

@@
-\chapter{Visual Basic for Applications For Data Collection} % Main appendix title
-\label{AppendixAccessDatabase} % For referencing this appendix elsewhere, use \ref{AppendixA}
-\lhead{Appendix \ref{AppendixAccessDatabase}. \emph{MS Access VBA For Data Collection}} % This is for the header on each page - perhaps a shortened title
+\chapter{Visual Basic for Applications For Data Collection}
+\label{AppendixAccessDatabase}
+\lhead{Appendix \ref{AppendixAccessDatabase}. \emph{MS Access VBA For Data Collection}}
+\clearpage
+%------------------------------------------
 
 \setstretch{1.0}
-%------------------------------------------

### 2015-08-23 Chapters/CPET_Jaundice.tex
Bug Hunt - CPET Jaundice
- an extra curly }) f%$£ it all up. An hour wasted hunting it down
- This is the actual commit for the error trap
- Not sure why the change didn;'t commit last time
- Maybe because I discarded the change and couldnt stage it again

@@
 
-\chapter{An investigation into the relationship between cardiopulmonary exercise testing, obstructive jaundice and preoperative pathophysiology in patients undergoing major pancreatic surgery.}
+\chapter{An investigation into the relationship between cardiopulmonary exercise testing, preoperative pathophysiology and obstructive jaundice in patients undergoing pancreaticoduodenectomy.}
 
@@
 
-\section{Patients and Methods}
+\section{Patients and methods}
 
@@
 Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
-Established criteria for resectability in patients with malignant disease were used as outlined in previous published work. 
+Established criteria for resectability in patients with malignant disease were used as outlined in Section \ref{sec:resectability_criteria}. 
 Segmental or wedge resection of the portal vein or superior mesenteric vein was carried out if the lesion was otherwise resectable.
@@
 The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
-$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshol.
+$\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshold.
 Oxygen consumption at peak exercise ($\dot{V}_{O_2}$Peak) was dichotomised using a cut-off of 16 ml/kg/min. 
 Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
-A mored detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
+A more detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
 
@@
 The degree of systemic inflammation was proportionate to the severity of jaundice. 
-Non-jaundiced patients had a median CRP of 3.6 mg/dl (inter-quartile range 0.3 - 89) and serum albumin of 37 g/dl (IQR 18 - 46) while patients with severe obstructive jaundice had a median CRP of 13 (IQR 1.7 - 51) and a median serum albumin of 25 g/dl (IQR 18 - 33).
+Non-jaundiced patients had a median CRP of 3.6 mg/l (inter-quartile range 0.3 - 89) and serum albumin of 37 g/l (IQR 18 - 46) while patients with severe obstructive jaundice had a median CRP of 13 (IQR 1.7 - 51) and a median serum albumin of 25 g/l (IQR 18 - 33).
 Obstructive jaundice was also associated with electrolyte abnormalities including serum sodium, serum potassium and serum chloride levels. 
@@
 Binary logistic regression analysis was undertaken to assess preoperative clinico-pathological patient factors associated with a low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak. 
-On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/dl (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min.
+On univariate analysis, female sex (HR 2.74, 95\% CI 1.30-5.74, p=0.008), body mass index $>$ 25 (HR 3.09, 95\% CI 1.51-6.32, p=0.002), presence of malignancy (HR 3.59, 95\% CI 1.36-9.43, p=0.010), POSSUM Physiology Score $>$ 14 (HR 2.06, 95\% CI 1.02-4.17, p= 0.044), serum bilirubin $>$ 250 $\mu$mol/l (HR 5.66, 95\% CI 1.87-17.16, p=0.002), haemoglobin $<$ 12 g/dl (HR 2.74, 95\% CI 1.30-5.74, p=0.008) and serum C-reactive protein $>$ 10 mg/l (HR 2.18, 95\% CI 1.06-4.51, p=0.035) were associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min.
 Age, cardiovascular and respiratory comorbidity, PBD, mild jaundice and serum albumin were not significantly related to a low $\dot{V}_{O_2}$AT.
@@
 The statistically significant independent variables with p$<$0.05 were entered into a backward step-wise regression model with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min as a categorical, binomial, dependent variable.
-Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/dl (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT.
+Female sex (HR 3.75, 95\% CI 1.57-8.95 p$<$0.005), body mass index $>$ 25 (HR 3.65, 95\% CI 1.61-8.26, p$<$0.005), malignancy (HR 4.02, 95\% CI 1.33-12.16, p$<$0.05) and C-reactive protein $>$ 10 mg/l (HR 2.98, 95\% CI 1.29-6.86, p$<$0.05) were independently related to low $\dot{V}_{O_2}$AT.
 Obstructive jaundice was not related to low $\dot{V}_{O_2}$AT on multivariate analysis. (Table \ref{table:cpet_oj_at_regression})
@@
 However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
-Pitt and coworkers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
+Pitt and co-workers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
 A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
 The authors concluded that PBD not only did not improve postoperative complication rates or mortality but resulted in a higher overall complication rate due to the morbidity associated with the procedure itself.
-These findings have been reproduced in a more recent multi-center, randomised trial \parencite{van_der_gaag_preoperative_2010}.
-A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outwith trial settings \parencite{wang_preoperative_2008}.
+These findings have been reproduced in a more recent multicenter, randomised trial \parencite{van_der_gaag_preoperative_2010}.
+A recent Cochrane Collaboration review of six trials including 520 patients concluded that PBD may be associated with serious adverse events and must not be performed routinely outside trial settings \parencite{wang_preoperative_2008}.
 
@@
 There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
-These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02}) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
+These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02} and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
 Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
@@
 However, jaundice was associated with elevated preoperative systemic inflammation as measured by C-reactive protein levels and serum albumin. 
-Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002}. 
-Both cellular immunity and humoral immunity are adversely affected by obstructive jaundice \parencite{padillo_cytokines_2001, scott-conner_pathophysiology_1994}. 
+Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002, padillo_cytokines_2001, scott-conner_pathophysiology_1994}.
 It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
@@
 
-%CPET vs Hemoglobin
+%CPET vs Haemoglobin
 The association between low haemoglobin and low $\dot{V}_{O_2}$Peak would appear to be more easily explained since the oxygen carrying capacity of blood depends on haemoglobin concentration.  

### 2015-08-23 Chapters/CPET_Jaundice.tex
Bug Hunt Complete
- Actually, it is this commit that has the solution
- See the single extra } that f$%£ed up the entire show?!

@@
 There was only a weak negative correlation between serum bilirubin and $\dot{V}_{O_2}$AT in our cohort.
-These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02} and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
+These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
 Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 

### 2015-08-23 Chapters/CRP_Complications.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - CRP_Complications
- Major tidying up
- References and spell checked
- Compiles ok
- HAVE NOT CHECKED BAD BOXES, TITLES, ETC.

@@
 
+\clearpage
 \section{Methods}
-\todo{Check dates for this study}
+
 Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 were included in this study. 
-Patients who underwent only a trial dissection or trial dissection and surgical bypass during this period were excluded.
+Patients who underwent only a trial dissection or trial dissection and palliative surgical bypass during this period were excluded. 
 All patients were given antibiotic prophylaxis at induction and this was continued for 24 hours after surgery. 
@@
 
-SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. Effects were considered significant at $\alpha \leq0.05$.
+SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. 
+Effects were considered significant at $\alpha \leq0.05$.
 
-\todo{Complete this section - most of this can be modified from the stats sections in other chapters}
-
+\clearpage
 \section{Results}
 \subsection{POPF, infective complications and other outcomes}
-Pancreaticoduodenectomy was performed in 188 patients (126 male, 67\%) during the study period. The median age was 63.5 years (IQR 54 - 70 years). 79 (42\%) patients were over the age of 65 years.
+Pancreaticoduodenectomy was performed in 188 patients (126 male, 67\%) during the study period. 
+The median age was 63.5 years (IQR 54 - 70 years). 79 (42\%) patients were over the age of 65 years.
 
@@
 Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p<0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
-These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig.
-\ref{fig:crp_comp_crp_popf} on p\pageref{fig:crp_comp_crp_popf}.
-Fig.
-\ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not significantly different between postoperative pancreatic fistulae of ISGPF Grade A, B and C.
+These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig. \ref{fig:crp_comp_crp_popf} on p\pageref{fig:crp_comp_crp_popf}. 
+Fig. \ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not significantly different between postoperative pancreatic fistulae of ISGPF Grade A, B and C.
 
@@
 		\includegraphics[width=\textwidth]{Figures/crp_comp_crp_popf_yes_no}
-		\caption{POPF Absent vs. Any Grade}
+		\caption{POPF Absent vs. 
+Any Grade}
 		\label{fig:crp_comp_crp_popf_yes_no}
@@
 
-%Quick analysis of CPET vs postop CRP - maybe one table
-%ROC analysis of CRP vs LOS
-%Multivariate analysis of CRP as continuous or categorical vs complications, 
 \clearpage
@@
 The results of the present study show that the severity of the postoperative systemic inflammatory response as measured by serial serum CRP levels in the first week is associated with postoperative complications after pancreaticoduodenectomy.
+The results also demonstrate that postoperative CRP trends are more useful in predicting infective complications in patients who did not develop a post-operative pancreatic fistula.
+
 While numerous studies have reported on a variety of methods to predict as well as diagnose POPF, few have reported on the role of CRP in predicting complications other than POPF.
 
-The introduction of the ISGPF definition for POPF has made the diagnosis of this complication simple and uniform.
+The introduction of the ISGPF definition for POPF has made the diagnosis of this complication simple and uniform \parencite{bassi_postoperative_2005}.
 The rate of POPF before the standardised definition was introduced varied anywhere between 2\% and 40\%. 
-However, in the post-ISGPF era, the rate of POPF in published literature has remained approximately 25\%.
-The sole criterion for the diagnosis of POPF is a drain amylase content on or after the third postoperative day three times the upper limit of normal for serum amylase in the testing laboratory.
+However, in the post-ISGPF era, the rate of POPF in published literature has remained approximately 25\% \parencite{knight_evaluation_2010}. 
+The sole criterion for the diagnosis of POPF is a drain amylase content on or after the third postoperative day three times the upper limit of normal for serum amylase in the testing laboratory \parencite{bassi_postoperative_2005} .
 
@@
 
-\todo{straatman_predictive_2015 - This is hot-off the press - reference this}
+\todo{his is hot-off the press - reference this}%straatman_predictive_2015 - T

### 2015-08-23 Chapters/Pre_Post_SIRS.tex
Review 1
- Every sentence on a new line
- Fits my OCD
Review 1 - Pre-Post-Sirs
- Major cleaning up

@@
+%Pre - Post SIRS
+
 \chapter{An investigation into the relationship between preoperative clinico-pathological characteristics and post-operative systemic inflammatory response in patients undergoing pancreaticoduodenectomy.}
@@
 \lhead{Chapter \ref{ch_pre_post_sirs}. \emph{Factors affecting post-operative systemic inflammation}} % This is for the header on each page - perhaps a shortened title
+\clearpage
 
@@
 
-For example, an elevated preoperative systemic inflammatory response is associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
+An elevated preoperative systemic inflammatory response was associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
 The modified Glasgow Prognostic Score in particular, which uses the combination of C-reactive protein and serum albumin, has been reported to be associated with increased incidence of complications \parencite{moyes_preoperative_2009, mohri_correlation_2014, vashist_glasgow_2010}.
-Preoperative CRP levels have also been reported to be associated with increased incidence of complications including infections and renal dysfunction as well as increased in-hospital mortality after cardiac surgery \parencite{lorenzo_increased_2012, mezzomo_preoperative_2011, kim_predictive_2009, biancari_preoperative_2003, boeken_increased_1998}.
+Elevated preoperative CRP levels have also been reported to be associated with increased incidence of complications including infections and renal dysfunction as well as increased in-hospital mortality after cardiac surgery \parencite{lorenzo_increased_2012, mezzomo_preoperative_2011, kim_predictive_2009, biancari_preoperative_2003, boeken_increased_1998}.
 
-Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures\parencite{singh_systematic_2014, platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery\parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures \parencite{singh_systematic_2014, platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery \parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
 %Include some recent reviews pp singh, bjs as well as check Donny's email.
@@
 
-To our knowledge, the relationship between the preoperative systemic inflammatory response and the magnitude of the postoperative systemic inflammatory response has not been examined before. 
+To our knowledge, the relationship between the preoperative systemic inflammatory response and the magnitude of the postoperative systemic inflammatory response after pancreaticoduodenectomy has not been examined before.
 While obstructive jaundice in itself has recently been reported to have no effect on postoperative complications, the impact of preoperative obstructive jaundice on postoperative systemic inflammation has not been previously reported.
@@
 
+\subsection{Aim}
 The aim of this study was to examine the relationship between patient factors including preoperative systemic inflammation, obstructive jaundice, cardiopulmonary exercise test parameters and body composition and the magnitude of the postoperative systemic inflammation during the first week after a pancreaticoduodenectomy.
@@
 
-\section{Methods}
+\clearpage
+\section{Patients and methods}
+
+\subsection{Patients}
 Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
@@
 
+\subsection{Preoperative data}
 Routine preoperative blood tests including full blood count, liver function tests and serum C-reactive protein were performed in all patients on the day before surgery. 
@@
 These results were collected from the hospital laboratory database using an automated MS Access application as outlined in Appendix \ref{AppendixAccessDatabase}.
+
 The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on p\pageref{table:mGPS} in Chapter \ref{ch_intro}. 
@@
 
+\subsection{BMI and body composition}
 Body Mass Index (BMI) was categorised using the World Health Organisation thresholds as shown in Table \ref{table:bmi_who} on p\ref{table:bmi_who} in Chapter \ref{ch_intro}. 
 Only one patient had a BMI less than 18.5 kg/m$2$ and was excluded from analysis involving BMI.
+Body composition was calculated in a subset of these patients using preoperative computed tomography of the abdomen. 
+The methodology used in the calculation of the individual components of body composition including visceral fat, subcutaneous fat and skeletal muscle is described in detail in Section \ref{sec:bodycomp_calculation}.
+Continuous data were converted into categorical data using tertiles.
+
+\subsection{Comorbidity and CPET}
 The Scottish Index of Multiple Deprivation (SIMD) Quintile Scores were calculated from the post-code of the patient's primary residence and dichotomised into two groups with scores of 1-3 and 4-5 respectively.
-The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described in Table \todo{Create a table for PPS in introduction and reference it wherever necessary}.
+The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
 
 In patients who underwent cardiopulmonary exercise testing, $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak were compared against postoperative systemic inflammation. 
-$\dot{V}_{O_2}$AT was dichotomised using a value of 10 mls/kg/min while $\dot{V}_{O_2}$Peak was dichotomised using a value of 16 mls/kg/min. 
+$\dot{V}_{O_2}$AT was dichotomised using a value of 10 ml/kg/min while $\dot{V}_{O_2}$Peak was dichotomised using a value of 16 ml/kg/min. 
 Cardiopulmonary exercise testing methodology has been described in Section \ref{sec:cpx_method}.
-
-Body composition was calculated in a subset of these patients using preoperative computed tomography of the abdomen. 
-The methodology used in the calculation of the individual components of body composition including visceral fat, subcutaneous fat and skeletal muscle is described in detail in Section \ref{sec:bodycomp_calculation}.
-%Check this with Donny
-Continuous data were converted into categorical data using tertiles.
 
@@
 
-SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. Effects were considered significant at $\alpha \leq0.05$.
+SPSS software (Version 22.0; IBM, USA) was used to perform statistical analysis. 
+Effects were considered significant at $\alpha \leq0.05$.
 
+\clearpage
 \section{Results}
+
+\subsection{Clinico-pathological characteristics}
 
@@
 Obstructive jaundice was present in 44 (23.4\%) patients and severe obstructive jaundice was present in 31 (16.5\%) of patients. 
-BMI data was available in 167 patients. More than half of these patients were overweight or obese with a BMI $>$ 25 in 87 (52.1\%) patients.
+BMI data was available in 167 patients. 
+More than half of these patients were overweight or obese with a BMI $>$ 25 in 87 (52.1\%) patients.
 All the variables necessary for calculation of the POSSUM Physiology Score, a composite score of comorbidity and preoperative biochemistry, were available in 180 patients and it was elevated in 87 (48.3\%) patients.
@@
 
+\subsection{Preoperative vs. postoperative systemic inflammation}
 The relationship between preoperative CRP and postoperative systemic inflammation is also depicted graphically in Fig. \ref{fig:sirs_crp}. 
@@
 Preoperative hypoalbuminemia on the other hand was associated with lower median postoperative CRP levels starting with POD 3. 
-The median difference in postoperative CRP increased from 30 mg/dl on POD 3 (p$<$0.045) to 53 mg/dl by POD 6 (p=0.002) with patients with hypoalbuminemia having a lower CRP. 
+The median difference in postoperative CRP increased from 30 mg/l on POD 3 (p$<$0.045) to 53 mg/l by POD 6 (p=0.002) with patients with hypoalbuminemia having a lower CRP. 
 This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
@@
 
+\subsection{Obstructive jaundice vs. postoperative inflammation}
 Obstructive jaundice was associated with significant differences in the postoperative systemic inflammatory response. 
@@
 
+\subsection{CPET, comorbidity vs. postoperative inflammation}
 Cardiopulmonary exercise testing was performed in 130 patients. 
@@
 
+\subsection{Body composition vs. postoperative inflammation}
 Body composition data was available for 90 patients.
@@
 The role of postoperative CRP in predicting complications after surgery has been well established. 
-A recent meta-analysis of six studies involving 1832 patients \parencite{warschkow_safe_2012} reported that CRP level less than 135 mg/dl on the fourth postoperative day had a high negative predictive value of 89\% for infectious complications.
-Postoperative CRP has been shown to predict complications after pancreaticoduodenectomy by us (\ref{ch_crp_comp}) as well as other authors (\parencite{welsch_persisting_2008, hiyoshi_usefulness_2013, kosaka_multivariate_2014}).
+A recent meta-analysis of six studies involving 1832 patients reported that CRP level less than 135 mg/l on the fourth postoperative day had a high negative predictive value of 89\% for infectious complications \parencite{warschkow_safe_2012}.
+Postoperative CRP has been shown to predict complications after pancreaticoduodenectomy by us (\ref{ch_crp_comp}) as well as other authors \parencite{welsch_persisting_2008, hiyoshi_usefulness_2013, kosaka_multivariate_2014}.
 The role of preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score or the neutrophil-lymphocyte ratio has also been reported to be adversely affect long-term survival in patients undergoing potentially curative surgery in over a hundred studies involving a variety of gastrointestinal and non-gastrointestinal cancers.
@@
 Hypoalbuminemia in patients scheduled to undergo pancreaticoduodenectomy is likely to be multi-factorial in origin. 
-Systemic inflammation, obstructive jaundice and the associated complex physiological, biochemical and immunological abnormalities, malnutrition, cancer cachexia, changes in body composition including sarcopenic obesity may all contribute to low serum albumin levels.
+Systemic inflammation, obstructive jaundice and the associated complex physiological, biochemical and immunological abnormalities, malnutrition, cancer cachexia, changes in body composition including sarcopenic-obesity may all contribute to low serum albumin levels.
 Contrary to our findings, hypoalbuminemia and poor preoperative nutritional status were found to be associated with higher postoperative levels of pro-inflammatory cytokines and CRP in surgical patients \parencite{nakamura_influence_1999}.
@@
 
-Immunosuppression and immune dysregulation have been recognised in patients with obstructive jaundice\parencite{scott-conner_pathophysiology_1994}.
+Immunosuppression and immune dysregulation have been recognised in patients with obstructive jaundice \parencite{scott-conner_pathophysiology_1994}.
 Our results are similar to those reported by Mackenzie and Woodhouse who studied the CRP response to bacteraemia in 126 patients critically ill patients with or without liver dysfunction \parencite{mackenzie_c-reactive_2006}.
 Liver dysfunction was defined as serum bilirubin $>$ 20 $\mu$mol/l and prothrombin time longer than 18 seconds. 
-The CRP response to bacteraemia was significantly lower in patients with liver dysfunction than in those with normal liver function (146 vs. 103 mg/dl, p=0.03). 
-These findings have been corroborated by other authors in patients with cirrhosis\parencite{pieri_c-reactive_2014, janum_c-reactive_2011}.
+The CRP response to bacteraemia was significantly lower in patients with liver dysfunction than in those with normal liver function (146 vs. 103 mg/l, p=0.03). 
+These findings have been corroborated by other authors in patients with cirrhosis \parencite{pieri_c-reactive_2014, janum_c-reactive_2011}.
 
@@
 However, the association of postoperative hypoalbuminemia with low aerobic capacity, POSSUM Physiology Score and anaemia appears to suggest a link between comorbidity and systemic inflammation and/or nutritional status.
-Bernal and co-workers studied 399 patients undergoing liver transplantation who had cardiopulmonary exercise testing as part of their preoperative work-up\parencite{bernal_aerobic_2014} and reported that hypoalbuminemia was associated with both low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak.
+Bernal and co-workers studied 399 patients undergoing liver transplantation who had cardiopulmonary exercise testing as part of their preoperative work-up and reported that hypoalbuminemia was associated with both low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak \parencite{bernal_aerobic_2014}.
 
@@
 Elevated systemic inflammation is related to both obesity as well as sarcopenia. 
-`Sarcopenic obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival in patients with pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012}.
+`Sarcopenic-obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival in patients with pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012}.
 
@@
 The results presented in this study emphasise the importance of the preoperative inflammatory status of the patient in influencing postoperative systemic inflammation. 
-The results also demonstrate the immunosuppressive effects of obstructive jaundice in these patients.
-
-%Limitations
-%no multivariate analysys
-%potential confounding factors
-%impact of surgical factors, complicaitons, etc
+The results also demonstrate the immunosuppressive effects of obstructive jaundice in patients undergoing pancreaticoduodenectomy.

### 2015-08-23 Chapters/Introduction.tex
Review - Introduction
- Minor tidying up of CPET parameter titles
- Other small changes

@@
 The variables used in the calculation of the POSSUM Physiology Score are shown in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
-As the Glasgow Coma Scale in all patients undergoing pancreaticoduodenectomy was 15, this variable was excluded in the calculation of the score in this thesis.
 
@@
 
-\subsection{Minute ventilation}
+\subsection[Minute ventilation]{Minute ventilation, $\dot{V}_E$}
 Minute ventilation or respiratory minute volume ($\dot{V}_E$) is the volume of air that is inhaled/expired in a minute.
@@
 
-\subsection{Oxygen uptake}
+\subsection[Oxygen Uptake]{Oxygen Uptake, $\dot{V}_{O_2}$}
 
@@
 
-\subsection{Oxygen pulse}
+\subsection[Oxygen pulse]{Oxygen pulse, $O_2Pulse$}
 $O_2Pulse$ is defined as the oxygen uptake per heart beat.
@@
 
-\subsection{Respiratory exchange ratio}
-The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio (RER). 
+\subsection[Respiratory Exchange Ratio]{Respiratory Exchange Ratio, RER}
+The ratio of $\dot{V}_{CO_2}/\dot{V}_{O_2}$ is called the Respiratory Exchange Ratio. 
 An RER greater that 1.0 may be caused either by lactic acidosis or due to hyperventilation. 
@@
 
-\subsection{Ventilatory equivalents}
-The changes in the ventilatory equivalents for $O_2$ ($\dot{V}_E/\dot{V}_{O_2}$) and $CO_2$ ($\dot{V}_E/\dot{V}_{CO_2}$) during exercise provide valuable information regarding the ventilatory response to exercise.
+\subsection[Ventilatory Equivalent for O2 and CO2]{Ventilatory Equivalent for $O_2$ and $CO_2$, $\dot{V}_E/\dot{V}_{O_2}$, $\dot{V}_E/\dot{V}_{CO_2}$}
+The changes in $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ during exercise provide valuable information regarding the ventilatory response to exercise. 
 Both $\dot{V}_E/\dot{V}_{O_2}$ and $\dot{V}_E/\dot{V}_{CO_2}$ tend to decrease initially during exercise. 
@@
 
-\subsection{End-tidal oxygen and CO2} 
+\subsection[End-tidal O2, CO2]{End-tidal $O_2$ and $CO_2$, $P_{ET_{O_2}}$, $P{ET_{CO_2}}$}
 $PET_{O_2}$ and $PET_{CO_2}$ are the partial pressures of $O_2$ and $CO_2$ at the end of an exhaled breath and are closely related to $P_aO_2$ and $P_aCO_2$ respectively.
@@
 
-\subsection{Heart rate}
+\subsection[Heart Rate]{Heart Rate, HR}
 \label{sec:heart_rate}
@@
 
-\subsection{Breathing frequency}
-In healthy individuals, breathing frequency ($B_f$) is one of the most important respiratory responses to exercise in order to increase oxygen uptake. 
+\subsection[Breathing frequency]{Breathing frequency, $B_f$}
+In healthy individuals, breathing frequency is one of the most important respiratory responses to exercise in order to increase oxygen uptake. 
 However, a variety of respiratory disorders including chronic obstructive pulmonary disease or restrictive disorders such as interstitial lung disease can impair this response. 
@@
 \label{sec:cpx_roleInAssessment}
-\subsection{General surgery}
+\subsection{General Surgery}
 
 Several studies over the past 2 decades have established the value of cardiopulmonary exercise testing in patients, especially elderly, undergoing major abdominal surgery. 
+Older and co-workers conducted a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery. 
+All patients underwent a symptom-limited exercise test on a cycle ergometer with real-time 12-lead ECG monitoring. 
+The average $\dot{V}_{O_2}$ across all patients was 12.4 ml/min/kg. 
+There were a total of 11 deaths related to cardiovascular causes (5.9\%) and 3 deaths due to non-cardiovascular causes. 
+They reported that 10 out of the 11 deaths due to cardiovascular causes occurred in patients with $\dot{V}_{O_2}AT<11$ml/kg/min (n=55) while only one patient in the group with $\dot{V}_{O_2}AT\geq11$ml/kg/min (n=132) died of cardiovascular complications (18\% vs 0.8\%, $p<0.001$).
+Eight out of the 10 patients in the low $\dot{V}_{O_2}$AT group who died also had evidence of myocardial ischaemia ($p<0.01$) \parencite{older_preoperative_1993}.
 
-Older and co-workers first reported on the value of cardiopulmonary exercise testing for preoperative risk assessment in a prospective study of 187 patients over the age of 60 undergoing major abdominal surgery  \parencite{older_preoperative_1993}.
 In a later, larger prospective study of 548 patients who underwent major abdominal surgery including colorectal and abdominal aneurysm surgery, they used a risk stratification system that combined $\dot{V}_{O_2}AT<11$ml/kg/min, $\dot{V}_E/\dot{V}_{CO_2}>35$ and evidence of myocardial ischaemia during exercise.

### 2015-08-24 Chapters/Discussion.tex
Discussion - Major rewrite - phase 1
- Will need another evening to tidy this up

--- 2015-08-09 Chapters/Discussion.tex
@@
-% Chapter 06 - Conclusion
+% Discussion
 
@@
 %----------------------------------------------------------------------------------------
+\section{Aims of thesis}
 
-The overall aim of the thesis was to examine the relationships between cardiopulmonary exercise physiology, obstructive jaundice, systemic inflammation, body composition and short and long-term outcomes after major pancreatic surgery.
+The overall aim of the thesis was to examine the inter-relationships between preoperative clinico-pathological characteristics including cardiopulmonary exercise physiology, obstructive jaundice, body composition and preoperative systemic inflammation and post-operative complications and the post-surgical systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
 
-A secondary aim of the thesis was to clarify the factors that affect cardiopulmonary exercise physiology and provide a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
+The present work also examined the factors that affect cardiopulmonary exercise physiology in order to enable a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
 
-The results of the present work have demonstrated that cardiopulmonary exercise testing provides an objective, reproducible method of identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy where the underlying pathology is a malignancy.
-These results have since been replicated by other authors examining the role of cardiopulmonary exercise testing in pancreatic surgery.
-\parencite{ausania_effects_2012, ausania_double_2012, junejo_cardiopulmonary_2014}
+%################################################################################
+\section{Clinical utility of CPET in predicting postoperative complications}
+%################################################################################
+The results presented in chapter \ref{ch_cpet_outcomes} demonstrated that cardiopulmonary exercise testing has a role in identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy after pancreaticoduodenectomy.
+These findings have since been replicated by other authors examining the role of cardiopulmonary exercise testing in pancreatic surgery.
 
-Equally important is the observation that preoperative obstructive jaundice did not impair aerobic capacity as measured by cardiopulmonary exercise testing.
+Ausania and co-workers evaluated the role of cardiopulmonary exercise testing in 122 patients who underwent a pancreaticoduodenectomy \parencite{ausania_effects_2012}
+Low $\dot{V}_{O_2}$AT ($<$10.1 ml/kg/min) was the only independent predictor of a postoperative pancreatic fistula.
+The incidence of a postoperative pancreatic fistula was 45\% in patients with a low $\dot{V}_{O_2}$AT while it was only 19\% in patients with a normal $\dot{V}_{O_2}$AT (p=0.02).
+The association between $\dot{V}_{O_2}$AT and postoperative pancreatic fistula was independent of pancreatic duct size, body mass index, obstructive jaundice or preoperative biliary drainage.
+They also reported in a separate study of 50 patients that complication  rates were higher in patients with a low $\dot{V}_{O_2}$AT undergoing palliative surgical bypass for advanced pancreatic cancer \parencite{ausania_double_2012}.
+
+Junejo and co-workers identified elevated $\dot{V}_E/\dot{V}_{CO_2}$ to be an independent predictor of 30-day mortality after pancreaticoduodenectomy in their study that included 64 patients who had undergone cardiopulmonary exercise testing \parencite{junejo_cardiopulmonary_2014}.
+$\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak  were not related to mortality or complications in this study.
+However, they also noted that a high $\dot{V}_E/\dot{V}_{CO_2}$ was also associated with poor long-term survival.
+
+The results of our work as well as that of several others over the past 2 decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
+However, as presented in the subsequent chapters of this thesis, a better understanding of the determinants of aerobic capacity as well as the perioperative systemic inflammatory response will enable clinicians to identify high risk patients and optimise their perioperative care.
+
+%################################################################################
+\section{Determinants of aerobic capacity}
+%################################################################################
+In Chapters \ref{ch_cpet_jaundice} and \ref{ch_bodycomp} we examined the preoperative patient factors that had an adverse effect on aerobic capacity.
+A better understanding of the factors that affect $\dot{V}_{O_2}$ may identify potential therapeutic targets or modifiable risk factors that may be used to improve patient fitness and mitigate the risks of surgery.
+We found that female sex, high body mass index, anaemia, presence of cancer and elevated CRP were all associated with either a low $\dot{V}_{O_2}$AT or a low $\dot{V}_{O_2}$Peak.
+
+In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} is the lack of association between obstructive jaundice and poor aerobic fitness. 
 We are aware of only one other published study looking at cardiopulmonary exercise testing and peripheral oxygen extraction in jaundiced versus non-jaundiced patients undergoing pancreaticoduodenectomy.
 Junejo and co-workers reported that peripheral oxygen extraction was normal during exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014}
-Routine preoperative biliary drainage is no longer recommended before a pancreaticoduodenectomy.
-Several studies have shown that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
+
+Preoperative biliary drainage is unlikely to improve aerobic fitness or modify their cardiopulmonary response to exercise/surgery.
+Moreover, several studies have shown that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
 \parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}
-The results reported in this thesis in conjunction with recently published clinical studies support the view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+Our results therefore support the view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
 
-We also observed that overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities.
-The analysis of the relationship between body composition and preoperative cardiopulmonary exercise testing has not been reported before in surgical patients.
-These results demonstrate that interpretation of cardiopulmonary exercise test parameters in patients with a high BMI must be made with caution, especially when such parameters have been corrected for the patient's total body weight.
+Overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities. 
+We present a detailed analysis of the relationship between body composition and preoperative cardiopulmonary exercise testing in chapter \ref{ch_bodycomp}.
+This has not been reported before in surgical patients to the best of out knowledge.
+
+The loss of skeletal muscle mass in patients with pancreatic cancer and spurious correlation with obesity due to correction for total body weight may partly explain the low $\dot{V}_{O_2}$ in these patients.
+Low aerobic capacity as measured by cardiopulmonary exercise testing must be interpreted with caution in overweight/obese patients and in patients with pancreatic cancer. 
+This is especially true for $\dot{V}_{O_2}$ that has been scaled for total body weight.
+Further studies must evaluate other parameters such as $O_2$Pulse or $\dot{V}_E/\dot{V}_{O_2}$ for their ability to predict postoperative outcomes. 
 With the significant increase in the proportion of obese patients undergoing major surgery and the presence of sarcopenic-obesity in patients with pancreatic cancer, poor performance at cardiopulmonary exercise testing may not necessarily be related to cardiac or pulmonary function.
-The loss of skeletal mass and spurious correlations due to correction for total body weight may partly explain the low $\dot{V}_{O_2}$ in these patients.
+
+Moreover, low aerobic capacity was not associated with documented cardiovascular or respiratory comorbidity or with an elevated POSSUM physiology score.
+It is possible that most patients with significant cardiac or respiratory comorbidity were identified as unsuitable for surgery and therefore were never subject to cardiopulmonary exercise testing. 
+Alternatively, it is possible that these patients did very poorly at cardiopulmonary exercise testing and this, in addition to their cardiorespiratory medical history precluded them from surgery.
+We did not include any patients who did not have an operation and this may have introduced a selection bias.
+
+%################################################################################
+\section{The role of systemic inflammation}
+%################################################################################
+We reported on the preoperative factors affecting postoperative systemic inflammation in chapter \ref{ch_pre_post_sirs} and the clinical utility of monitoring trends in the early postoperative systemic inflammatory response in predicting complications in chapter \ref{ch_crp_comp}.
+
+Systemic inflammation is increasingly recognised as playing an important role in determining not only long-term outcomes but also short-term outcomes after cancer surgery.
+Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy.
+These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}
+
+\bigskip \hrule
+
+ANYTHING AFTER THIS IS IN EARLY DRAFT STAGE
+
+\bigskip \hrule
+
+\section{Improving outomes after pancreaticoduodenectomy}
+
+- centralisation
+- specialisation
+- early rescue and aggressive management of complications \parencite{gouma_rates_2000}
+
+- better understanding of physiology
+- role of sytemic inflammation \parencite{van_heek_hospital_2005, ho_effect_2003, birkmeyer_surgeon_2003, halm_is_2002}
+- impact of sarcopenic-obesity \parencite{joglekar_sarcopenia_2015, reisinger_sarcopenia_2015, gonzalez_obesity_2014}
+- role of tumour related factors \parencite{williams_surgical_2014}
+
+\section{Prehabilitation}
+
+%using cpet to assess impact of neo-adjuvant treatment
+%recommend prehab for neoadj in pancreatic cancer
 
@@
 
-Early recognition of patients at increased risk of postoperative complications facilitates prompt initiation of appropriate treatment and judicious allocation of critical care resources.
-Moreover, identification of patients who are unlikely to develop a major postoperative complication allows their progress on enhanced recovery pathways with the benefits of rapid recovery, reduced morbidity and early discharge.
-These patients are also more likely to receive adjuvant therapy without undue delay.
-Recognition of these low risk patients helps improve outcomes, reduces hospital costs and releases valuable resources for the care of high risk patients.
-The results reported in Chapter [ss] demonstrate the value of early post-operative C-reactive protein in identifying patients who are more likely to develop an infective complication.
-The results show that CRP is only useful in the absence of a postoperative pancreatic fistula.
-The combination of low drain amylase and low CRP on the third postoperative day identifies patients who do not have a pancreatic fistula and are less likely to develop an infective complication.
+\section{Future directions}
+
+%Early recognition of patients at increased risk of postoperative complications facilitates prompt initiation of appropriate treatment and judicious allocation of critical care resources. 
+%Moreover, identification of patients who are unlikely to develop a major postoperative complication allows their progress on enhanced recovery pathways with the benefits of rapid recovery, reduced morbidity and early discharge.
+%These patients are also more likely to receive adjuvant therapy without undue delay. 
+%Recognition of these low risk patients helps improve outcomes, reduces hospital costs and releases valuable resources for the care of high risk patients. 
+%The results reported in Chapter [ss] demonstrate the value of early post-operative C-reactive protein in identifying patients who are more likely to develop an infective complication. 
+%The results show that CRP is only useful in the absence of a postoperative pancreatic fistula. 
+%The combination of low drain amylase and low CRP on the third postoperative day identifies patients who do not have a pancreatic fistula and are less likely to develop an infective complication.

### 2015-08-24 Chapters/Discussion.tex
Discussion - Struggling a bit
- Can barely keep my eyes open
- But can't sleep either
- Need to pull this fucking thing together by tomorrow latest
- Take wednesday off to tidy up the whole bloody lot
- Then get Mo to do his MD as well

@@
 However, as presented in the subsequent chapters of this thesis, a better understanding of the determinants of aerobic capacity as well as the perioperative systemic inflammatory response will enable clinicians to identify high risk patients and optimise their perioperative care.
+
+% Different surgeries and patient groups may need different thresholds for risk prediction
+% CPET and long-term survival in PDAC
+% Why is low $\dot{V}_{O_2}$ likely to leak - pancreatic perfusion 
+high lactate - popf de_schryver_early_2015
+
+karoliska - ansorge_early_2012
 
@@
 Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy.
-These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}
+These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
 
@@
 
-\section{Improving outomes after pancreaticoduodenectomy}
+\section{Improving outcomes after pancreaticoduodenectomy}
 
@@
 
-\section{Prehabilitation}
+\subsection{Prehabilitation}
 
@@
 %recommend prehab for neoadj in pancreatic cancer
+%different types of surgery may need different levels of preparation
 
@@
 In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
-The authors report that major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in these patients.\parencite{tzeng_morbidity_2014} 
+The authors report that major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in these patients \parencite{tzeng_morbidity_2014}.
 The authors recommend that surgeons identify these patients early, institute interventions to optimise their comorbidities and enrol these patients for prehabilitation.
 
-West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy.\parencite{west_effect_2015}
-Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery.\parencite{gillis_prehabilitation_2014}
-Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity measured objectively by cardiopulmonary exercise testing by upto 10\% which can potentially move approximately 30\% of high risk patients into the low risk category.\parencite{dunne_pmo-029_2012}
+West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
+Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery \parencite{gillis_prehabilitation_2014}.
+Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity measured objectively by cardiopulmonary exercise testing by upto 10\% which can potentially move approximately 30\% of high risk patients into the low risk category \parencite{dunne_pmo-029_2012}.
+
+\subsection{Intra-operative care}
+
+The results of chapter \ref{ch_cpet_outcomes} and the work of Ausania and co-workers appeared to suggest a link between low aerobic capacity and increased incidence of pancreatic fistula.
+We hypothesised that tissue hypoperfusion may have played a role in the poor healing of the pancreatico-jejunal anastomosis.
+Reyad and co-workers performed a randomised trial to examine the effect of intra-operative dobutamine infusion during pancreaticoduodenectomy on splanchnic perfusion, hemodynamics, and overall postoperative outcome \parencite{reyad_effect_2013}.
+They reported that intra-operative dobutamine use was associated with improved global oxygen delivery as measured by arterial and venous blood gases, splanchnic perfusion as measured by gastric tonometry and postoperative complications.
+It was interesting to note that the incidence of anastomotic leak was 30\% in the control group, 15\% in the group that received 3$\mu$g/kg/min of dobutamine and 5\% in the group that received 5$\mu$g/kg/min of dobutamine (p$<$0.05).
+The overall complication rate also decreased from 70\% in the control group to 40\% and 20\% in the dobutamine groups (p$<0.05$).
+
+However, a recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
+The study did show that patients whose postoperative oxygen delivery was similar to their preoperative values had fewer complications.
+However, oxygen delivery was not influenced by goal-directed therapy as any beneficial effect of the intervention was lost with the autonomic nervous system changes that accompanied the increased intravenous fluids and inotropes in the treatment cohort.
+
+The OPTIMISE randomised controlled trial also did not show any benefit of goal-directed therapy in high risk patients undergoing major gastrointestinal surgery \parencite{pearse_effect_2014}.
+
+\subsection{Post-operative care}
 

### 2015-08-25 Chapters/Discussion.tex
Discussion - More updates
- See text
- This is a partial updata
- Proabably 2 or 3 more updates

@@
 %################################################################################
+
+\subsection{Risk assessment and predicting complications}
+
 The results presented in chapter \ref{ch_cpet_outcomes} demonstrated that cardiopulmonary exercise testing has a role in identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy after pancreaticoduodenectomy.
@@
 
+\subsection{Assessing impact of prehabilitation or neoadjuvant treatment}
 The results of our work as well as that of several others over the past 2 decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
@@
 
+In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
+Major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in patients with borderline fitness \parencite{tzeng_morbidity_2014}.
+The authors recommended that surgeons identify these patients early, institute interventions to optimise their comorbidities and enrol these patients for prehabilitation.
+
+Neoadjuvant treatments were associated with significant reduction in $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients with rectal cancer \parencite{west_effects_2014} and oesphago-gastric cancer \parencite{jack_effect_2014}.
+West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy.
+They used cardiopulmonary exercise testing to monitor the impact of neo-adjuvant treatment and prehabilitation \parencite{west_effect_2015}.
+Neoadjuvant therapy in borderline resectable pancreatic cancer is the subject of several trials with generally promising results \parencite{evans_preoperative_2008, gillen_preoperative/neoadjuvant_2010, golcher_neoadjuvant_2015, kim_multi-institutional_2013, sahora_phase_2014}.
+
+It is likely that the impact of such treatment on fitness in these patients will be similar to that observed in patients with colorectal and oesophago-gastric cancer.
+Prehabilitation would play an important role in improving outcomes in these patients while cardiopulmonary exercise testing will serve as an important clinical tool to objectively measure the impact of prehabilitation.
+
+\subsection{Assessing impact on long-term survival}
+
+Poor aerobic capacity is associated with reduced long-term survival after abdominal aortic aneurysm surgery \parencite{tang_cardiopulmonary_2012, grant_cardiopulmonary_2015}.
+There is some evidence that cardiopulmonary exercise testing may predict long-term survival after pancreaticoduodenectomy for pancreatic cancer \parencite{junejo_cardiopulmonary_2014}.
+Reduced aerobic capacity was also associated with poor survival in patients with lung cancer \parencite{jones_peak_2010}.
+
+While the an elevated preoperative systemic inflammatory response has been shown to have an adverse impact on long-term survival in patients with pancreatic cancer both in those who underwent potentially curative surgery or palliation, the impact of aerobic capacity and comorbidity on long-term survival in patients with pancreatic cancer is less clear.
+Some have suggested that comorbidity was associated with shorter survival in patients with pancreatic cancer \parencite{dias-santos_charlson_2015}.
+However, other authors have reported that the relatively short median survival after pancreatic cancer is not affected by comorbidity in patients who underwent potentially curative surgery \parencite{kos_evaluation_2014}
+
+The relationship between aerobic capacity and cancer-survival is important as it provides for a relatively easy therapeutic target.
+
 % Different surgeries and patient groups may need different thresholds for risk prediction
-% CPET and long-term survival in PDAC
+% Done - CPET and long-term survival in PDAC
 % Why is low $\dot{V}_{O_2}$ likely to leak - pancreatic perfusion 
-high lactate - popf de_schryver_early_2015
+%high lactate - popf de_schryver_early_2015
 
-karoliska - ansorge_early_2012
+%karoliska - ansorge_early_2012
+- better understanding of physiology
+- role of sytemic inflammation \parencite{van_heek_hospital_2005, ho_effect_2003, birkmeyer_surgeon_2003, halm_is_2002}
+- impact of sarcopenic-obesity \parencite{joglekar_sarcopenia_2015, reisinger_sarcopenia_2015, gonzalez_obesity_2014}
+- role of tumour related factors \parencite{williams_surgical_2014}
 
 %################################################################################
-\section{Determinants of aerobic capacity}
+\section{Prehabilitation,improving aerobic capacity and mitigating risk}
 %################################################################################
@@
 
+%################################################################################
+\subsection{Obstructive jaundice}
+%################################################################################
 In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} is the lack of association between obstructive jaundice and poor aerobic fitness. 
+In chapter \ref{ch_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay.
+
 We are aware of only one other published study looking at cardiopulmonary exercise testing and peripheral oxygen extraction in jaundiced versus non-jaundiced patients undergoing pancreaticoduodenectomy.
 Junejo and co-workers reported that peripheral oxygen extraction was normal during exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014} 
+Preoperative biliary drainage is unlikely to improve aerobic fitness or modify their cardiopulmonary response to exercise/surgery \parencite{parker_serum_2014}.
 
-Preoperative biliary drainage is unlikely to improve aerobic fitness or modify their cardiopulmonary response to exercise/surgery.
-Moreover, several studies have shown that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
-\parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}
-Our results therefore support the view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+We also demonstrated the immunosuppressive effects of obstructive jaundice on the postoperative systemic inflammatory response in chapter \ref{ch_pre_post_sirs}. 
+To our knowledge, this is the first detailed study of the postoperative systemic inflammatory response in jaundiced versus non-jaundiced patients undergoing pancreaticoduodenectomy.
 
+In their, appropriately titled paper \textit{`The Quandary of Preresection Biliary Drainage for Pancreatic Cancer'}, the authors discuss the complex physiological effects of obstructive jaundice, its impact on immune function and the differences between findings in animal studies and humans \parencite{tol_quandary_2012}.
+However, they also review recent literature where there appears to be overwhelming evidence that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
+\parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}.
+
+Our results therefore support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+
+%################################################################################
+\subsection{Systemic inflammation}
+%################################################################################
+We reported on the preoperative factors affecting postoperative systemic inflammation in chapter \ref{ch_pre_post_sirs} and the clinical utility of monitoring trends in the early postoperative systemic inflammatory response in predicting complications in chapter \ref{ch_crp_comp}.
+
+Preoperative systemic inflammation is increasingly recognised as playing an important role in determining not only long-term outcomes but also short-term outcomes after cancer surgery \parencite{kubo_elevated_2013, mohri_correlation_2014, moyes_preoperative_2009, neal_preoperative_2011, vashist_glasgow_2010,  haupt_association_1997}.
+While we did not find an association between elevated preoperative systemic inflammation in our cohort of patients, we demonstrated in chapter \ref{ch_pre_post_sirs} that preoperative systemic inflammation had a significant impact on  the magnitude of the postoperative inflammatory response.
+This supports the theory that abnormal systemic inflammation is a continuum that starts in the preoperative phase and continues into the postoperative phase.
+
+It would appear that preoperative patient factors affect the postoperative systemic inflammatory response independent of intra-operative factors or postoperative complications. 
+The mechanisms underlying the association between pre- and post-operative systemic inflammation require further study.
+
+Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy.
+These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
+
+Our approach to interpreting the postoperative systemic inflammatory reponse has been different from the above mentioned studies. 
+We demonstrated in chapter \ref{ch_crp_comp} that postoperative CRP did not predict the severity of a pancreatic fistula and did not have any clinical utility in the prediction of a POPF (section \ref{Postoperative CRP vs.
+POPF}).
+However, we demonstrated a clear role for the use of postoperative CRP in predicting infective complications in the absence of a POPF.
+
+In summary the combination of a low CRP and absence of POPF reduces the likelihood of any other infective complications and identifies patients who are suitable for early discharge.
+Such patients will also be more likely to benefit from a full course of adjuvant chemotherapy although this requires further study.
+
+Postoperative complications and the associated systemic inflammatory response are associated with early recurrence and poor-survival after colorectal \parencite{artinyan_infectious_2015, mcardle_impact_2005}, gastric \parencite{hayashi_impact_2015,kubota_prognostic_2014}  and pancreatic cancer surgery \parencite{aoyama_impact_2015, kamphues_postoperative_2012}.
+
+Early recognitioon and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation.
+
+%################################################################################
+\subsection{Body composition}
+%################################################################################
 Overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities. 
@@
 
+%################################################################################
+\subsection{Aerobic exercise}
+%################################################################################
+
+West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
+Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery \parencite{gillis_prehabilitation_2014}.
+Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity by upto 10\%  measured objectively by cardiopulmonary exercise testing.
+This could potentially move approximately 30\% of high risk patients into the low risk category \parencite{dunne_pmo-029_2012}.
+
+A recent consensus statement by the `Exercise Prehabilitation in Colorectal Cancer Delphi Study Group' in the United Kingdom showed agreement between colorectal surgeons that prehabilitation must form an important part of the preparation of elective patients for major surgery \parencite{boereboom_forming_2015}.
+The surgeons also agreed that improving aerobic capacity is likely to improve outcomes but recommended that further studies were needed before this became established clinical practice.
+
+%################################################################################
+\subsection{Anaemia}
+%################################################################################
+
+Measures other than exercise may also improve aerobic capacity. 
+Correction of anaemia was associated with an improvement in $\dot{V}_{O_2}$AT by 0.39 ml/kg/min for every 1 gm/dl improvement in haemoglobin \parencite{wright_cardiopulmonary_2014}. 
+While anaemia did not change the effect of aerobic exercise in improving physical performance, the overall aerobic fitness was significantly lower in anaemic patients than in patients with normal haemoglobin \parencite{bellotto_anemia_2011}.
+However, perioperative blood transfusion in pancreatic cancer patients was associated with poorer long-term outcomes \parencite{kneuertz_effects_2011, sutton_perioperative_2014}.
+
+While iron supplementation is associated with correction of anaemia and reduction in the requirement for blood transfusion, the role of preoperative iron supplementation in improving outcomes is unclear.
+Further randomised controlled trials investigating the short-term and long-term consequences of preoperative iron supplementation in patients undergoing cancer surgery are required before this becomes accepted clinical practice \parencite{beris_perioperative_2008,hallet_impact_2014}.
+
+%################################################################################
 Moreover, low aerobic capacity was not associated with documented cardiovascular or respiratory comorbidity or with an elevated POSSUM physiology score.
@@
 We did not include any patients who did not have an operation and this may have introduced a selection bias.
+%################################################################################
 
 %################################################################################
-\section{The role of systemic inflammation}
+\section{Improving outcomes after pancreaticoduodenectomy}
 %################################################################################
-We reported on the preoperative factors affecting postoperative systemic inflammation in chapter \ref{ch_pre_post_sirs} and the clinical utility of monitoring trends in the early postoperative systemic inflammatory response in predicting complications in chapter \ref{ch_crp_comp}.
 
-Systemic inflammation is increasingly recognised as playing an important role in determining not only long-term outcomes but also short-term outcomes after cancer surgery.
-Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy.
-These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
+\subsection{Centralisation and Specialisation}
 
-\bigskip \hrule
+\subsection{Early and aggressive management of complications}
 
-ANYTHING AFTER THIS IS IN EARLY DRAFT STAGE
-
-\bigskip \hrule
-
-\section{Improving outcomes after pancreaticoduodenectomy}
+\subsection{Enhanced Recovery Programmes}
 
@@
 - early rescue and aggressive management of complications \parencite{gouma_rates_2000}
-
-- better understanding of physiology
-- role of sytemic inflammation \parencite{van_heek_hospital_2005, ho_effect_2003, birkmeyer_surgeon_2003, halm_is_2002}
-- impact of sarcopenic-obesity \parencite{joglekar_sarcopenia_2015, reisinger_sarcopenia_2015, gonzalez_obesity_2014}
-- role of tumour related factors \parencite{williams_surgical_2014}
 
@@
 
-Recognition of these factors is important not only in patient selection and informed consent, but also in instituting the appropriate interventions as part of prehabilitation programmes. 
-In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
-The authors report that major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in these patients \parencite{tzeng_morbidity_2014}.
-The authors recommend that surgeons identify these patients early, institute interventions to optimise their comorbidities and enrol these patients for prehabilitation.
-
-West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
-Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery \parencite{gillis_prehabilitation_2014}.
-Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity measured objectively by cardiopulmonary exercise testing by upto 10\% which can potentially move approximately 30\% of high risk patients into the low risk category \parencite{dunne_pmo-029_2012}.
+%boereboom_forming_2015
 
@@
 
-The results of chapter \ref{ch_cpet_outcomes} and the work of Ausania and co-workers appeared to suggest a link between low aerobic capacity and increased incidence of pancreatic fistula.
+The results of chapter \ref{ch_cpet_outcomes} and the work of Ausania and co-workers appeared to suggest a link between reduced oxygen delivery and increased incidence of pancreatic fistula. 
 We hypothesised that tissue hypoperfusion may have played a role in the poor healing of the pancreatico-jejunal anastomosis.
+
 Reyad and co-workers performed a randomised trial to examine the effect of intra-operative dobutamine infusion during pancreaticoduodenectomy on splanchnic perfusion, hemodynamics, and overall postoperative outcome \parencite{reyad_effect_2013}.
@@
 
-However, a recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
+However, the benefit of goal-directed therapy to improve organ perfusion and oxygen delivery in improving postoperative outcomes remains unclear \parencite{grocott_perioperative_2013, pearse_effect_2014}.
+A recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
 The study did show that patients whose postoperative oxygen delivery was similar to their preoperative values had fewer complications.
 However, oxygen delivery was not influenced by goal-directed therapy as any beneficial effect of the intervention was lost with the autonomic nervous system changes that accompanied the increased intravenous fluids and inotropes in the treatment cohort.
-
-The OPTIMISE randomised controlled trial also did not show any benefit of goal-directed therapy in high risk patients undergoing major gastrointestinal surgery \parencite{pearse_effect_2014}.
 

### 2015-08-25 Chapters/Discussion.tex
Discussion - more updates

@@
 %################################################################################
-\section{Prehabilitation,improving aerobic capacity and mitigating risk}
+\section{Prehabilitation, improving aerobic capacity and mitigating risk}
 %################################################################################
@@
 
-Our results therefore support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+Our results support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
 
@@
 The mechanisms underlying the association between pre- and post-operative systemic inflammation require further study.
+
+%################################################################################
+\subsection{Body composition}
+%################################################################################
+Overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities. 
+We present a detailed analysis of the relationship between body composition and preoperative cardiopulmonary exercise testing in chapter \ref{ch_bodycomp}.
+This has not been reported before in surgical patients to the best of our knowledge.
+
+The loss of skeletal muscle mass in patients with pancreatic cancer and spurious correlation with obesity due to correction for total body weight may partly explain the low $\dot{V}_{O_2}$ in these patients.
+Low aerobic capacity as measured by cardiopulmonary exercise testing must be interpreted with caution in overweight/obese patients and in patients with pancreatic cancer. 
+This is especially true for $\dot{V}_{O_2}$ that has been scaled for total body weight.
+Further studies must evaluate other parameters such as $O_2$Pulse or $\dot{V}_E/\dot{V}_{O_2}$ for their ability to predict postoperative outcomes. 
+The proportion of obese patients undergoing major surgery is increasing. 
+Moreover, sarcopenia and sarcopenic-obesity are recognised as an important adverse prognostic feature in pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012} and is associated with increased postoperative complications after pancreaticoduodenectomy \parencite{joglekar_sarcopenia_2015, pausch_cachexia_2012}.
+Sarcopenia was also an independent risk factor for an elevated postoperative systemic inflammatory response in patients with colorectal cancer \parencite{reisinger_sarcopenia_2015}.
+
+Our results suggest that poor performance at cardiopulmonary exercise testing may not necessarily be related to only cardiac or pulmonary function. 
+Body composition appears to not only affect aerobic capacity but also systemic inflammation. 
+Improvements in body composition are achievable through nutritional supplementation \parencite{machado_whey_2015} as well as increased physical activity.
+
+%################################################################################
+\subsection{Anaemia}
+%################################################################################
+
+Measures other than exercise may also improve aerobic capacity. 
+Correction of anaemia was associated with an improvement in $\dot{V}_{O_2}$AT by 0.39 ml/kg/min for every 1 gm/dl improvement in haemoglobin \parencite{wright_cardiopulmonary_2014}. 
+While anaemia did not change the effect of aerobic exercise in improving physical performance, the overall aerobic fitness was significantly lower in anaemic patients than in patients with normal haemoglobin \parencite{bellotto_anemia_2011}.
+However, perioperative blood transfusion in pancreatic cancer patients was associated with poorer long-term outcomes \parencite{kneuertz_effects_2011, sutton_perioperative_2014}.
+
+While iron supplementation is associated with correction of anaemia and reduction in the requirement for blood transfusion, the role of preoperative iron supplementation in improving outcomes is unclear.
+Further randomised controlled trials investigating the short-term and long-term consequences of preoperative iron supplementation in patients undergoing cancer surgery are required before this becomes accepted clinical practice \parencite{beris_perioperative_2008,hallet_impact_2014}.
+
+%################################################################################
+\subsection{Aerobic exercise}
+%################################################################################
+
+West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
+Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery \parencite{gillis_prehabilitation_2014}.
+Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity by upto 10\%  measured objectively by cardiopulmonary exercise testing.
+This could potentially move approximately 30\% of high risk patients into the low risk category \parencite{dunne_pmo-029_2012}.
+
+A recent consensus statement by the `Exercise Prehabilitation in Colorectal Cancer Delphi Study Group' in the United Kingdom showed agreement between colorectal surgeons that prehabilitation must form an important part of the preparation of elective patients for major surgery \parencite{boereboom_forming_2015}.
+The surgeons also agreed that improving aerobic capacity is likely to improve outcomes but recommended that further studies were needed before this became established clinical practice.
+
+%%################################################################################
+%Moreover, low aerobic capacity was not associated with documented cardiovascular or respiratory comorbidity or with an elevated POSSUM physiology score.
+%It is possible that most patients with significant cardiac or respiratory comorbidity were identified as unsuitable for surgery and therefore were never subject to cardiopulmonary exercise testing. 
+%Alternatively, it is possible that these patients did very poorly at cardiopulmonary exercise testing and this, in addition to their cardiorespiratory medical history precluded them from surgery.
+%We did not include any patients who did not have an operation and this may have introduced a selection bias.
+%%################################################################################
+
+%################################################################################
+\section{Improving outcomes after pancreaticoduodenectomy}
+%################################################################################
+
+\subsection{Centralisation and Specialisation}
+
+One of the biggest factors in improving both short-term and long-term outcomes after cancer surgery has been the centralisation of cancer cancer and the associated specialisation of surgeons.
+
+Birkmeyer and co-workers reported in 1999 that hospital volume was not only related to postoperative mortality and morbidity but also to long-term cancer survival in patients undergoing pancreaticoduodenectomy.
+
+\subsection{Early and aggressive management of complications}
 
@@
 
-Early recognitioon and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation.
-
-%################################################################################
-\subsection{Body composition}
-%################################################################################
-Overweight and obese patients had significantly lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in spite of no significant increase in known cardiac or respiratory comorbidities. 
-We present a detailed analysis of the relationship between body composition and preoperative cardiopulmonary exercise testing in chapter \ref{ch_bodycomp}.
-This has not been reported before in surgical patients to the best of out knowledge.
-
-The loss of skeletal muscle mass in patients with pancreatic cancer and spurious correlation with obesity due to correction for total body weight may partly explain the low $\dot{V}_{O_2}$ in these patients.
-Low aerobic capacity as measured by cardiopulmonary exercise testing must be interpreted with caution in overweight/obese patients and in patients with pancreatic cancer. 
-This is especially true for $\dot{V}_{O_2}$ that has been scaled for total body weight.
-Further studies must evaluate other parameters such as $O_2$Pulse or $\dot{V}_E/\dot{V}_{O_2}$ for their ability to predict postoperative outcomes. 
-With the significant increase in the proportion of obese patients undergoing major surgery and the presence of sarcopenic-obesity in patients with pancreatic cancer, poor performance at cardiopulmonary exercise testing may not necessarily be related to cardiac or pulmonary function.
-
-%################################################################################
-\subsection{Aerobic exercise}
-%################################################################################
-
-West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
-Moderate aerobic and resistance exercises and protein supplementation started 4 weeks before surgery was noted to improve postoperative functional exercise capacity in patients undergoing colorectal cancer surgery \parencite{gillis_prehabilitation_2014}.
-Exercise programs as short as 4 weeks in duration have been reported to improve aerobic capacity by upto 10\%  measured objectively by cardiopulmonary exercise testing.
-This could potentially move approximately 30\% of high risk patients into the low risk category \parencite{dunne_pmo-029_2012}.
-
-A recent consensus statement by the `Exercise Prehabilitation in Colorectal Cancer Delphi Study Group' in the United Kingdom showed agreement between colorectal surgeons that prehabilitation must form an important part of the preparation of elective patients for major surgery \parencite{boereboom_forming_2015}.
-The surgeons also agreed that improving aerobic capacity is likely to improve outcomes but recommended that further studies were needed before this became established clinical practice.
-
-%################################################################################
-\subsection{Anaemia}
-%################################################################################
-
-Measures other than exercise may also improve aerobic capacity. 
-Correction of anaemia was associated with an improvement in $\dot{V}_{O_2}$AT by 0.39 ml/kg/min for every 1 gm/dl improvement in haemoglobin \parencite{wright_cardiopulmonary_2014}. 
-While anaemia did not change the effect of aerobic exercise in improving physical performance, the overall aerobic fitness was significantly lower in anaemic patients than in patients with normal haemoglobin \parencite{bellotto_anemia_2011}.
-However, perioperative blood transfusion in pancreatic cancer patients was associated with poorer long-term outcomes \parencite{kneuertz_effects_2011, sutton_perioperative_2014}.
-
-While iron supplementation is associated with correction of anaemia and reduction in the requirement for blood transfusion, the role of preoperative iron supplementation in improving outcomes is unclear.
-Further randomised controlled trials investigating the short-term and long-term consequences of preoperative iron supplementation in patients undergoing cancer surgery are required before this becomes accepted clinical practice \parencite{beris_perioperative_2008,hallet_impact_2014}.
-
-%################################################################################
-Moreover, low aerobic capacity was not associated with documented cardiovascular or respiratory comorbidity or with an elevated POSSUM physiology score.
-It is possible that most patients with significant cardiac or respiratory comorbidity were identified as unsuitable for surgery and therefore were never subject to cardiopulmonary exercise testing. 
-Alternatively, it is possible that these patients did very poorly at cardiopulmonary exercise testing and this, in addition to their cardiorespiratory medical history precluded them from surgery.
-We did not include any patients who did not have an operation and this may have introduced a selection bias.
-%################################################################################
-
-%################################################################################
-\section{Improving outcomes after pancreaticoduodenectomy}
-%################################################################################
-
-\subsection{Centralisation and Specialisation}
-
-\subsection{Early and aggressive management of complications}
+Early recognition and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation.
 
@@
 - early rescue and aggressive management of complications \parencite{gouma_rates_2000}
-
-\subsection{Prehabilitation}
-
-%using cpet to assess impact of neo-adjuvant treatment
-%recommend prehab for neoadj in pancreatic cancer
-%different types of surgery may need different levels of preparation
-
-%boereboom_forming_2015
 

### 2015-08-25 Chapters/Discussion.tex
Discussion - More updates
- Nearly there

@@
 
-One of the biggest factors in improving both short-term and long-term outcomes after cancer surgery has been the centralisation of cancer cancer and the associated specialisation of surgeons.
+One of the biggest factors in improving both short-term and long-term outcomes after cancer surgery has been the centralisation of cancer treatment and the associated specialisation of surgeons.
 
-Birkmeyer and co-workers reported in 1999 that hospital volume was not only related to postoperative mortality and morbidity but also to long-term cancer survival in patients undergoing pancreaticoduodenectomy.
+Birkmeyer and co-workers were the first to report in 1999 that hospital volume was not only related to postoperative mortality and morbidity but also to long-term cancer survival in patients undergoing pancreaticoduodenectomy in the United States\parencite{birkmeyer_relationship_1999}.
+Better overall quality of care, better cancer operations based on oncological principles, better rates of adjuvant therapy may have all contributed to better outcomes in high-volume centres. 
+They also reported in 2003 that operative mortality was inversely related to surgeon volume emphasising the importance of surgical expertise in improving outcomes \parencite{birkmeyer_surgeon_2003}.
+Gouma and co-workers reported very similar findings in the Netherlands were morbidity and mortality after pancreaticoduodenectomy were linked to hospital volume\parencite{gouma_rates_2000}.
 
-\subsection{Early and aggressive management of complications}
+The importance of multi-disciplinary teams in the administration of complex perioperative care in these patients by Ho and Heslin who reported that the overall duration with pancreaticoduodenectomy in a hospital was associated with better postoperative outcomes \parencite{ho_effect_2003}.
+This finding stresses the importance of organisational processes and allied specialists including anaesthetists, intensivists, diagnostic/interventional radiologists, nurses, physiotherapists and dietitians.
 
-Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy.
-These studies performed at large, specialist centres, have used a combination of postoperative CRP and drain amylase levels to identify patients who are at the least risk of developing a Grade B/C pancreatic fistula \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
+However, in spite of these advances, significant variation in practice remains an obstacle in improving outcomes after pancreaticoduodenectomy for pancreatic cancer \parencite{cyr_canadian_2015}.
 
-Our approach to interpreting the postoperative systemic inflammatory reponse has been different from the above mentioned studies. 
+\subsection{Early recognition and aggressive management of complications}
+
+Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
+
+Our approach to interpreting the postoperative systemic inflammatory response has been different from these studies. 
 We demonstrated in chapter \ref{ch_crp_comp} that postoperative CRP did not predict the severity of a pancreatic fistula and did not have any clinical utility in the prediction of a POPF (section \ref{Postoperative CRP vs.
@@
 However, we demonstrated a clear role for the use of postoperative CRP in predicting infective complications in the absence of a POPF. 
-
-In summary the combination of a low CRP and absence of POPF reduces the likelihood of any other infective complications and identifies patients who are suitable for early discharge.
-Such patients will also be more likely to benefit from a full course of adjuvant chemotherapy although this requires further study.
+The combination of a low CRP and absence of POPF reduced the likelihood of any other infective complications and identified patients who may be suitable for early discharge.
 
@@
 Early recognition and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation. 
+`Failure to rescue' after postoperative complications is one of the factors responsible for the discrepancy between morbidity and mortality in low-volume versus high-volume centres.
+Mortality rates in patients undergoing a variety of complex surgical procedures including pancreatectomy was 2.5 times higher in low-volume centres than high-volume centres. 
+However, complication rates were similar between these hospitals. 
+Failure to identify and manage complications aggressively may have resulted in the increased mortality in the low-volume hospitals \parencite{ghaferi_complications_2009}.
+In a recent large study involving nearly 2 million surgical patients, Ferraris VA and co-workers studied the outcome of over two-hundred thousand patients who had complications. 
+They noted that 20\% of high-risk patients accounted for 90\% of deaths due to postoperative complications. 
+They also reported that a simple risk-scoring system based on preoperative characteristics was able to identify these patients. 
+Early recognition of complications in these patients and aggressive management may result in improved outcomes in these high risk patients.
+
+These findings are supported by other studies involving emergency abdominal surgery \parencite{sheetz_improving_2013} and patients at major trauma centres \parencite{haas_prevention_2011}
+
+The results presented in this thesis not only allow identification of high risk patients before surgery but also help in the early identification of postoperative complications.
 

### 2015-08-25 Chapters/Introduction.tex
Introduction - Appa Changes  - Appa has printed out the entire introduction and spent several hours going through it - Has picked out a huge number of mistakes - Back to school - Fixed all of them - Thanks

@@
 
-In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} offers the only chance of a cure.
+In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000,geer_prognostic_1993} followed by adjuvant chemotherapy in selected patients \parencite{neoptolemos_randomized_2004, neoptolemos_adjuvant_2009} offers the only chance of a cure.
 However, major pancreatic surgery places significant physiological stresses on multiple organ systems. 
@@
 
-The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon result in early involvement of these structures even by relatively small tumours.
+The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon results in early involvement of these structures even by relatively small tumours.
 Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. 
@@
 In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. 
-Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of patient fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
+Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of physiological fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
 
@@
 Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}.
-The operation was further popularised initially as a two-stage procedure by Whipple \parencite{whipple_treatment_1935} before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.
+The operation was further popularised initially as a two-stage procedure by Allen Whipple \parencite{whipple_treatment_1935}, an American surgeon, before evolving into the current single stage operation by the 1950s \parencite{whipple_rationale_1941, whipple_radical_1950}.
 
@@
 Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. 
-Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely (R0) with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}.
-The role of neoadjuvant therapy such as chemotherapy with or without local radiotherapy and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy is the subject of ongoing clinical research in patients with locally advanced and/or metastatic disease.
-Early reports suggest that some of these patients can undergo potentially curative surgery after neo-adjuvant treatment.
+Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}.
+Neoadjuvant therapies such as chemotherapy with or without local radiotherapy \parencite{gillen_preoperative/neoadjuvant_2010, evans_preoperative_2008} and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy are the subject of ongoing clinical research.
+There is increasing evidence that some of these patients may be able to undergo potentially curative surgery after neo-adjuvant treatment.
 
@@
 
-A comprehensive preoperative evaluation was performed including staging of the disease for local resectability and absence of metatstatic disease, assessment of patient fitness and anaesthetic review.
+A comprehensive preoperative evaluation was performed including staging of the disease for local resectability and absence of metastatic disease, assessment of patient fitness and anaesthetic review. 
 All patients with suspected cancer were discussed at the pancreato-biliary multidisciplinary team meeting. 
-A fully informed consent was obtained after discussion regarding treatment options, the details of a pancreaticoduodenectomy, expected perioperative course, potential risks and follow-up care.
+Fully informed consent was obtained after discussion regarding treatment options, the details of a pancreaticoduodenectomy, expected perioperative course, potential risks and follow-up care.
 
@@
 Prophylactic antibiotics were administered at induction. 
-While the use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate, it was routinely used in all patients at this centre. 
-Octreotide was administered intra-operatively (200 micrograms subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms subcutaneous injection administered three times a day).
+The use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate.
+However, it was routinely used in all patients at our centre. 
+Octreotide was administered intra-operatively (200 micrograms by subcutaneous injection at induction and repeated after 4 hours) and was continued for 5 days postoperatively (200 micrograms by subcutaneous injection administered three times a day).
 
@@
 The hepatoduodenal ligament was dissected after a fundus-first cholecystectomy to isolate the common bile duct which was transected after ascertaining the hepatic artery anatomy. 
-The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery should the patient require embolisation of this vessel for post-pancreatectomy haemorrhage. 
+The gastro-duodenal artery was divided leaving a short segment on the common hepatic artery. 
+This was done to facilitate mesenteric angiography and embolisation of this vessel in case of post-pancreatectomy haemorrhage. 
 The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. 
@@
 Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures.
-Gstro-jejunostomy or duodenojejunostomy (in PPPD) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
+Gastro-jejunostomy or duodenojejunostomy (in pylorus-preserving pancreaticoduodenectomy) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
 One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.
@@
 
-Histology of the removed specimen was discussed at the multi-disciplinary team meeting in cancer patients and suitable patients were referred for adjuvant chemotherapy.
+Histopathology reports were discussed at the multi-disciplinary team meeting for cancer patients and suitable patients were referred for adjuvant chemotherapy.
 
 \subsection{Complications}
-The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950's to less than 5\% in most large volume centres around the world \parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000}.
+The incidence of complications after pancreaticoduodenectomy remains high in spite of a steady decline in postoperative mortality from over 40\% in the 1950s to less than 5\% in most large volume centres around the world \parencite{deoliveira_assessment_2006, emick_hospital_2006, yeo_six_1997, winter_1423_2006, teh_patient_2009, gouma_rates_2000}.
 Accurate definitions and clinical severity grading are useful in comparing outcomes between centres as well as for research, audit and quality improvement purposes. 
@@
 Post-pancreatectomy haemorrhage may either be intra-luminal into the gastrointestinal tract or intra-abdominal into the peritoneal/retro-peritoneal space. 
-Post-pancreatectomy haemorrhage may be from any of a number of potential sources although bleeding from the stump of the gastroduodenal artery is the most common cause. 
+Post-pancreatectomy haemorrhage may be from a number of potential sources although bleeding from the stump of the gastroduodenal artery is the most common cause. 
 Other potential sources include suture lines at the anastomoses, gastric/duodenal ulcers or diffuse gastritis, pseudoaneurysms of the gastro-duodenal, splenic or rarely the hepatic artery or rarely, haemobilia.
@@
 and was used to grade complications in this thesis. 
-Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death while grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.
+Complications of grade 3 or greater are considered clinically significant as they require further procedures either under local or general anaesthesia, single or multiple organ support or result in death.
+However, grades 1 and 2 are managed with ward-based care including antibiotics, wound management, supplemental enteral or parenteral nutrition and similar interventions.
 
@@
 While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma.
-The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery and have reported that chemotherapy prolonged survival after surgery while chemoradiotherapy had a harmful effect \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}.
+The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery.
+They have reported that chemotherapy after surgery prolonged survival while chemoradiotherapy had a harmful effect \parencite{neoptolemos_adjuvant_2001, neoptolemos_randomized_2004, neoptolemos_adjuvant_2009, neoptolemos_adjuvant_2010}.
 
@@
 However, adjuvant chemoradiotherapy resulted in 5-year survival dropping from 20\% to 10\% and has largely been abandoned in the adjuvant setting. 
-The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid. \parencite{neoptolemos_adjuvant_2010}
+The ESPAC-3 trial compared the efficacy of 5-FU and Folinic acid with Gemcitabine and reported that Gemcitabine was not superior to 5-FU/Folinic acid \parencite{neoptolemos_adjuvant_2010}.
 
@@
 
-Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on patient fitness \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}.
+Selection of patients suitable for potentially curative surgery not only depends on the resectability criteria that were discussed in section \ref{sec:resectability_criteria} but also on physiological fitness \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}.
 Major pancreatic surgery is associated with significant morbidity. 
 Subjecting patients with poor physiological reserve to a pancreaticoduodenectomy is fraught with risk. 
-It can result in a worse outcome for the patient due to either post-operative mortality or prolonged hospitalisation due to complications resulting in poor quality of life with little difference in survival when compared to palliation.
-Patient factors that affect postoperative recovery and long-term survival are therefore important in determining suitability for surgery.
+High risk patients undergoing surgery are more likely to develop postoperative complications resulting in prolonged hospitalisation and poor quality of life with little difference in survival when compared to palliative treatment only.
+The risk of post-operative mortality is also increased.
 
@@
 Comorbidity is defined as the presence of or the effect of other diseases that a patient has in addition to the primary disease of interest. 
-Major pancreatic surgery requires the patient to have adequate reserve to cope not only with the increased physiological demand during and immediately after surgery but also to cope with any adverse events that may occur.
+Major pancreatic surgery requires the patient to have adequate reserve to cope not only with the increased perioperative physiological demand but also with any adverse events that may occur.
 
@@
 
-DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also complication severity after pancreaticoduodenectomy \parencite{deoliveira_assessment_2006}.
+DeOliveira and co-workers reported that cardiovascular disease was a risk factor not only for overall morbidity but also for severity of complications after pancreaticoduodenectomy \parencite{deoliveira_assessment_2006}.
 Obesity is known to be associated with greater incidence and severity of postoperative complications \parencite{benns_impact_2009}. 
@@
 The ability of the cardiorespiratory system as well as other organ systems including renal, gastrointestinal, hepatic, coagulatory and immunological systems to cope with major surgery and its sequelae plays a major role in determining postoperative outcomes.
-Therefore, accurate assessment of physiological fitness is important in selecting patients who will benefit from surgery, advising patients of the risks of surgery as it applies to them, preoperative optimisation as well as in perioperative management decisions.
+Therefore, accurate assessment of physiological fitness is important in patient selection, individualised risk prediction and consent, preoperative optimisation as well as perioperative management. 
 A thorough clinical history of co-existing illnesses, general examination and review of patient's medications is followed by more specialist testing to assess fitness. 
@@
 While some authors have reported it to be useful in predicting postoperative outcomes after pancreatic surgery \parencite{pratt_possum_2008} others have found that POSSUM was not a good risk prediction model in this group of patients \parencite{khan_evaluation_2003, kocher_risk-adjustment_2005, tamijmarane_application_2008, de_castro_evaluation_2009}.
-The physiology component of the POSSUM score (POSSUM Physiology Score, PPS) which takes into account both comorbidity and specific preoperative biochemical parameters was used as an objective static measure of comorbidity in this thesis.
+
+The physiology component of the POSSUM score (POSSUM Physiology Score, PPS) which takes into account comorbidities and specific preoperative biochemical parameters was used as an objective static measure of comorbidity in this thesis.
 The POSSUM Physiology Score has been reported to be associated with preoperative systemic inflammation and cancer-specific survival in patients who underwent potentially curative surgery for colorectal cancer \parencite{richards_relationship_2010}.
@@
 Cardiopulmonary exercise tests were performed in the Department of Respiratory Medicine at the Glasgow Royal Infirmary using the ZAN-600 CPET suite (nSpire Health, Longmont, CO 80501, USA). 
-All tests were performed by specialist respiratory physiologists. 
+All tests were performed by specialised respiratory physiologists. 
 Suitable equipment for cardiopulmonary resuscitation were available in the department in case of unexpected problems. 
@@
 Low risk patients (with none of the risk factors mentioned above, n=280) were managed on the general surgical ward with no cardiovascular mortality \parencite{older_cardiopulmonary_1999}. 
-Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient.
-\parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}.
+Subsequent literature reviews by Older and co-workers have emphasised the value of cardiopulmonary exercise testing in risk assessment as well as optimising perioperative care in the high-risk surgical patient \parencite{older_preoperative_2000, older_clinical_2004, older_preoperative_2005}.
 
@@
 Snowden and co-workers reported in a study of 171 patients of who 123 underwent surgery, that $\dot{V}_{O_2}AT<10.1$ ml/kg/min was associated with not only cardiovascular complications but also with other complications including pulmonary, renal, gastrointestinal, infective and haematological complications.
-One of the strengths of this study was the fact that the clinicians treating these patients were blinded to the results of preoperative cardiopulmonary exercise testing results thus avoiding management bias.
+One of the strengths of this study was the fact that clinicians were blinded to the results of preoperative cardiopulmonary exercise testing, thus avoiding management bias. 
 They also included other cardiopulmonary exercise test parameters such as $\dot{V}_{O_2}$Peak and $\dot{V}_E/\dot{V}_{CO_2}$ at the anaerobic threshold in addition to $\dot{V}_{O_2}$AT in their analysis and showed that $\dot{V}_{O_2}$AT was more predictive of postoperative morbidity than POSSUM derived morbidity, cardiac risk scoring index or a validated activity questionnaire.
@@
 
-In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in patients with no documented history of cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9) \parencite{wilson_impaired_2010}.
+In a study of 847 patients undergoing elective abdominal surgery for colorectal disease, bladder or renal cancer, Wilson and co-workers found that $\dot{V}_{O_2}AT<11$ ml/kg/min in the absence of documented cardiac risk factors was associated with a relative risk of mortality of 10.0 (95\% CI 1.7-61.9) \parencite{wilson_impaired_2010}.
 This emphasises the value of cardiopulmonary exercise testing in diagnosing sub-clinical or previously undiagnosed (and untreated) cardiovascular and respiratory disease. 
-Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) {a}. 
-nd in patients without ischaemic heart disease. 
-It would appear therefore that This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium term survival after patients left hospital.
+Moreover, 90-day survival was also better in patients with better aerobic capacity ($\dot{V}_{O_2}AT>11$ ml/kg/min and $\dot{V}_E/\dot{V}_{CO_2}<34$) and in patients without ischaemic heart disease. 
+This appears to suggest that aerobic capacity influenced not only postoperative outcomes but also medium-term survival after patients left hospital.
 
@@
 This composite adverse outcome was more likely in patients with high BMI ($>$45) and low $\dot{V}_{O_2}$Peak ($<$15.9 ml/kg/min). 
-In fact, patients with neither of these two risk factors had no complications after surgery. 
+Patients in whom both these risk factors were absent did not develop postoperative complications. 
 They also observed a significant negative relationship between BMI and $\dot{V}_{O_2}$ (p$<$0.0001) and age(p$<$0.0001). 
@@
 While $\dot{V}_{O_2}$Peak was significantly lower in patients who developed postoperative cardiopulmonary complications, receiver-operator characteristics (ROC) analysis did not identify a clinically useful threshold that would stratify patients into different risk groups.
-They postulated that the fact that post-oesophagectomy complications such as anastomotic leak or sepsis often happen within the thorax and are not necessarily due to cardiopulmonary dysfunction, preoperative cardiopulmonary exercise testing may not have identified these patients.
-Older and Hall, the pioneers of cardiopulmonary exercise test in surgical patients, also raise this issue in their letter to the authors of the above study. \parencite{hall_cardiopulmonary_2009}.
+Post-oesophagectomy complications such as anastomotic leak or sepsis often happen within the thorax and are not necessarily due to cardiopulmonary dysfunction.
+They postulated that this may have reduced the predictive ability of preoperative cardiopulmonary exercise testing.
+Older and Hall, the pioneers of cardiopulmonary exercise testing in surgical patients, also raise this issue in their letter to the authors of the above study \parencite{hall_cardiopulmonary_2009}.
 
@@
 \subsection{Inflammation and postoperative complications}
-Abnormalities of systemic inflammatory processes present as a continuum that starts in the preoperative phase possibly as a consequence of underlying comorbid illnesses, presence of cancer, or an abnormality of the immune system or a due to a combination of all of these factors.
+Abnormalities of systemic inflammatory processes present as a continuum that starts in the preoperative phase possibly as a consequence of underlying comorbid illnesses, presence of cancer, an abnormality of the immune system or due to a combination of all of these factors.
 Surgical trauma in such 'primed' patients results in a cascade of events that trigger several inflammatory pathways that have now shown to have a direct impact not only on the incidence of postoperative complications but also on cancer recurrence and long-term survival.
@@
 In a study of 383 patients undergoing elective rectal cancer surgery with primary anastomosis, Welsch and co-workers reported that persistently raised CRP level over 140 mg/L after the third/fourth postoperative day was associated with anastomotic leak \parencite{welsch_c-reactive_2007}.
-They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreatico-jejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP levels greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
+They also reported in a separate study of 688 patients undergoing pancreatic resection with pancreatico-jejunostomy for neoplastic disease or chronic pancreatitis, that persistently elevated CRP greater than 140 mg/L on the fourth postoperative day was associated with increased incidence of complications.
 
@@
 While CRP level between the third and fifth postoperative day has been reported to be most predictive of complications, the clinical signs do not become apparent until later in the postoperative course, often after the fifth day.
-This has led some authors to postulate that the elevated CRP levels may in fact be due to an abnormally modulated postoperative inflammatory response resulting in an initial exaggerated systemic inflammatory response syndrome (SIRS) followed by a compensatory anti-inflammatory response syndrome (CARS).
+This has led some authors to postulate that elevated CRP may in fact be due to an abnormally modulated postoperative inflammatory response resulting in an initial exaggerated systemic inflammatory response syndrome (SIRS) followed by a compensatory anti-inflammatory response syndrome (CARS).
 
@@
 
-In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP 2 weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications \parencite{katoh_anastomotic_2011}.
+In a study of 207 patients undergoing surgery for Duke's B colorectal cancer, Katoh and co-workers reported that anastomotic leakage and persistently elevated CRP two weeks after surgery were independent risk factors for systemic recurrence, further emphasising the important role of inflammation in cancer recurrence as a consequence of complications \parencite{katoh_anastomotic_2011}.
 Wound infections and intra-abdominal infections have also been associated with poorer survival in colorectal cancer patients \parencite{nespoli_impact_2006}. 
@@
 Similar finding were reported by Raut and co-workers in their study of 360 patients who underwent pancreaticoduodenectomy for pancreatic ductal adenocarcinoma \parencite{raut_impact_2007}. 
-and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head. \parencite{kang_detrimental_2009}
+and by Kang and co-workers in their report on 103 patients undergoing R0 resections for cancer of the pancreatic head \parencite{kang_detrimental_2009}.
 
@@
 They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimmetic agents. 
-Similar findings have been reported by other authors in animal studies. \parencite{binah_obstructive_1985,bomzon_systemic_1986}.
+Similar findings have been reported by other authors in animal studies \parencite{binah_obstructive_1985,bomzon_systemic_1986}.
 
@@
 Obstructive jaundice was associated with increased levels of ANP as a result of increased cardiac endocrine activity in bile duct ligated rabbits \parencite{pereira_increased_1994}. 
-Similar findings have been reported in humans as well. \parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}.
+Similar findings have been reported in humans as well \parencite{gallardo_increased_1998, martinez-rodenas_circulating_1998}.
 
@@
 Early observational studies of small cohorts of patients challenged this paradigm reporting that surgery in jaundiced patients was not associated with greater risk and in fact preoperative biliary drainage increased hospitalisation costs \parencite{snellen_influence_1985, bakkevold_morbidity_1993, pitt_does_1985}.
-Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications.
-\parencite{hatfield_preoperative_1982}
+Hatfield and co-workers reported in one of the earliest randomised controlled trial comparing pancreaticoduodenectomy with or without preoperative biliary drainage in 57 patients that there was no difference in postoperative complications or mortality but increased biliary drainage associated complications \parencite{hatfield_preoperative_1982}.
 
 Similar findings have been reported by other authors as well \parencite{lai_preoperative_1994, lai_preoperative_1994, jagannath_effect_2005}. 
-However, some authors have recommended that expandable metal stents were associated with fewer stent related complications than plastic stents and may be considered in jaundiced patients where surgery is delayed for other reasons.
-\parencite{wasan_use_2005, mullen_pancreaticoduodenectomy_2005}
+However, some authors have recommended that expandable metal stents were associated with fewer stent related complications than plastic stents and may be considered in jaundiced patients where surgery is delayed for other reasons \parencite{wasan_use_2005, mullen_pancreaticoduodenectomy_2005}.
 
@@
 The high incidence of procedure related complications, the use of plastic stents, the 25\% failure to drain at first attempt and exclusion of patients with serum bilirubin $>$ 250 $\mu$mol/L were significant limitations of the study.
-However, the study does show that early surgery in the jaundiced patient with serum bilirubin $<$ 250 $\mu$mol/L is safe and not associated with increased morbidity or mortality as once feared.
+However, the study does show that early surgery in the jaundiced patient with serum bilirubin $<$ 250 $\mu$mol/L is safe and not associated with increased morbidity or mortality as was once feared.
 
@@
 A greater proportion of men than women were obese and there was a marked increase in BMI with increasing age. 
-However, as the proportion of overweight or obese decreased in women as their education level increased.
+There was also an inverse relationship between education level and incidence of obesity \parencite{who_obesity:_2000}.
 
@@
 Several studies have reported increased incidence of postoperative complications in obese patients after cardiac, transplant, urologic and gynaecological surgery \parencite{prem_operative_1965, holley_obesity_1990, fasol_influence_1992, gruberg_impact_2002, lee_impact_2004}.
-Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer.
-\parencite{tsujinaka_visceral_2008}
+Visceral obesity rather than body mass index has been reported to be associated with greater incidence of wound complications, overall complication rate and prolonged hospital stay in patients undergoing laparascopic sigmoid colectomy for cancer  \parencite{tsujinaka_visceral_2008}.
 
@@
 However, obesity was an independent predictor of minor complications including wound infections. 
-However, they noted that underweight patients had a five-fold increased risk of postoperative mortality. \parencite{mullen_impact_2008}
+However, they noted that underweight patients had a five-fold increased risk of postoperative mortality \parencite{mullen_impact_2008}.
 
@@
 In a more recent prospective, multi-institutional, risk-adjusted cohort study of 118,707 patients undergoing non-bariatric general surgery, Mullen and co-workers reported that the risk of mortality was lowest in the overweight (BMI 25 - 30) and moderately obese (BMI 30 - 35) patients and highest in the underweight and morbidly obese patients.
-This appears to confirm the existence of an 'obesity paradox' in surgical patients. \parencite{mullen_obesity_2009}
+This appears to confirm the existence of an 'obesity paradox' in surgical patients \parencite{mullen_obesity_2009}.
 
@@
 Benns and co-workers reported greater blood loss and increased incidence and severity of postoperative complications in obese patients undergoing surgery compared to non-obese patients. 
-However, there was no difference in operative time or length of stay in hospital. \parencite{benns_impact_2009}
+However, there was no difference in operative time or length of stay in hospital \parencite{benns_impact_2009}.
 
@@
 \vfill
-To examine the value of serial daily postoperative markers of systemic inflammatory response like white cell count, albumin 
-and C-reactive protein concentrations in the prediction of post-operative complications after pancreaticoduodenectomy.
+To examine the value of serial daily postoperative markers of systemic inflammatory response like white cell count, albumin and C-reactive protein concentrations in the prediction of post-operative complications after pancreaticoduodenectomy.
 \vfill

### 2015-08-25 Chapters/Discussion.tex
Discussion - Complete
- Final big commit
- Next few commits will all be just minor fixes and formatting

@@
 %----------------------------------------------------------------------------------------
-\section{Aims of thesis}
+\section{Summary of Aims}
 
@@
 
+\clearpage
+
 %################################################################################
-\section{Clinical utility of CPET in predicting postoperative complications}
+\section{Clinical utility of CPET}
 %################################################################################
@@
 Junejo and co-workers identified elevated $\dot{V}_E/\dot{V}_{CO_2}$ to be an independent predictor of 30-day mortality after pancreaticoduodenectomy in their study that included 64 patients who had undergone cardiopulmonary exercise testing \parencite{junejo_cardiopulmonary_2014}.
-$\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak  were not related to mortality or complications in this study.
+%$\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak  were not related to mortality or complications in this study.
 However, they also noted that a high $\dot{V}_E/\dot{V}_{CO_2}$ was also associated with poor long-term survival.
 
-\subsection{Assessing impact of prehabilitation or neoadjuvant treatment}
-The results of our work as well as that of several others over the past 2 decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
-However, as presented in the subsequent chapters of this thesis, a better understanding of the determinants of aerobic capacity as well as the perioperative systemic inflammatory response will enable clinicians to identify high risk patients and optimise their perioperative care.
+The results of our work in chapter \ref{ch_cpet_outcomes} as well as that of several others over the past 2 decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
 
-In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
+\subsection{Assessing impact of prehabilitation and neoadjuvant treatment}
+
+Our results also suggest that a better understanding of the determinants of aerobic capacity as well as the perioperative systemic inflammatory response will enable clinicians to identify high risk patients and optimise their perioperative care.
+
+In a recent study of 8266 patients undergoing pancreaticoduodenectomy in the United States, Tzeng and co-workers reported that a third (3033) were of borderline fitness as a consequence of advanced age ($>$ 80 years), poor performance status, weight loss $>$ 10\%, pulmonary disease, recent myocardial infarction/angina, stroke history, and/or preoperative sepsis.
 Major complications (31.3 vs. 26.2\%) and mortality (4.1 vs. 2.3\%) were greater in patients with borderline fitness \parencite{tzeng_morbidity_2014}.
@@
 Neoadjuvant treatments were associated with significant reduction in $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients with rectal cancer \parencite{west_effects_2014} and oesphago-gastric cancer \parencite{jack_effect_2014}.
-West and co-workers used prehabilitation to return aerobic capacity to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy.
-They used cardiopulmonary exercise testing to monitor the impact of neo-adjuvant treatment and prehabilitation \parencite{west_effect_2015}.
+West and co-workers used prehabilitation to return aerobic capacity, as measured by cardiopulmonary exercise testing, to baseline levels in patients with locally advanced rectal cancer undergoing neo-adjuvant chemoradiotherapy \parencite{west_effect_2015}.
 Neoadjuvant therapy in borderline resectable pancreatic cancer is the subject of several trials with generally promising results \parencite{evans_preoperative_2008, gillen_preoperative/neoadjuvant_2010, golcher_neoadjuvant_2015, kim_multi-institutional_2013, sahora_phase_2014}.
 
-It is likely that the impact of such treatment on fitness in these patients will be similar to that observed in patients with colorectal and oesophago-gastric cancer.
-Prehabilitation would play an important role in improving outcomes in these patients while cardiopulmonary exercise testing will serve as an important clinical tool to objectively measure the impact of prehabilitation.
+The impact of neoadjuvant treatment on fitness in patients with pancreatic cancer are likely to be similar or worse to that observed in other cancer patients.
+Prehabilitation can play an important role in improving outcomes in these patients while cardiopulmonary exercise testing serves as an objective clinical tool to quantify the impact of prehabilitation.
 
@@
 
-Poor aerobic capacity is associated with reduced long-term survival after abdominal aortic aneurysm surgery \parencite{tang_cardiopulmonary_2012, grant_cardiopulmonary_2015}.
 There is some evidence that cardiopulmonary exercise testing may predict long-term survival after pancreaticoduodenectomy for pancreatic cancer \parencite{junejo_cardiopulmonary_2014}.
-Reduced aerobic capacity was also associated with poor survival in patients with lung cancer \parencite{jones_peak_2010}.
+Reduced aerobic capacity was also associated with poor survival in patients with lung cancer \parencite{jones_peak_2010} and reduced long-term survival after abdominal aortic aneurysm surgery \parencite{tang_cardiopulmonary_2012, grant_cardiopulmonary_2015}.
 
-While the an elevated preoperative systemic inflammatory response has been shown to have an adverse impact on long-term survival in patients with pancreatic cancer both in those who underwent potentially curative surgery or palliation, the impact of aerobic capacity and comorbidity on long-term survival in patients with pancreatic cancer is less clear.
+While an elevated preoperative systemic inflammatory response was associated with shorter survival in patients with pancreatic cancer, both after potentially curative surgery \parencite{jamieson_systemic_2005, torre_glasgow_2012} or palliation for advanced disease \parencite{glen_evaluation_2006}, the impact of aerobic capacity on long-term survival in patients with pancreatic cancer is less clear.
 Some have suggested that comorbidity was associated with shorter survival in patients with pancreatic cancer \parencite{dias-santos_charlson_2015}.
-However, other authors have reported that the relatively short median survival after pancreatic cancer is not affected by comorbidity in patients who underwent potentially curative surgery \parencite{kos_evaluation_2014}
+However, other authors have reported that the relatively short median survival after pancreatic cancer surgery is not affected by comorbidity whose effects are more long-term \parencite{kos_evaluation_2014}
 
-The relationship between aerobic capacity and cancer-survival is important as it provides for a relatively easy therapeutic target.
-
-% Different surgeries and patient groups may need different thresholds for risk prediction
-% Done - CPET and long-term survival in PDAC
-% Why is low $\dot{V}_{O_2}$ likely to leak - pancreatic perfusion 
-%high lactate - popf de_schryver_early_2015
-
-%karoliska - ansorge_early_2012
-- better understanding of physiology
-- role of sytemic inflammation \parencite{van_heek_hospital_2005, ho_effect_2003, birkmeyer_surgeon_2003, halm_is_2002}
-- impact of sarcopenic-obesity \parencite{joglekar_sarcopenia_2015, reisinger_sarcopenia_2015, gonzalez_obesity_2014}
-- role of tumour related factors \parencite{williams_surgical_2014}
+Further studies examining the relationship between aerobic capacity measured using objective methods and long-term survival are important. 
+Aerobic capacity is a relatively easy therapeutic target that may have the potential to improve not only short-term but also long-term outcomes.
 
@@
 \section{Prehabilitation, improving aerobic capacity and mitigating risk}
+\label{disc_prehabilitation}
 %################################################################################
@@
 %################################################################################
-In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} is the lack of association between obstructive jaundice and poor aerobic fitness. 
-In chapter \ref{ch_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay.
-
-We are aware of only one other published study looking at cardiopulmonary exercise testing and peripheral oxygen extraction in jaundiced versus non-jaundiced patients undergoing pancreaticoduodenectomy.
-Junejo and co-workers reported that peripheral oxygen extraction was normal during exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014} 
+In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} was the lack of association between obstructive jaundice and poor cardiopulmonary function. 
+We are aware of only one other similar study by Junejo and co-workers.
+They reported that peripheral oxygen extraction was normal during cardiopulmonary exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014}
 Preoperative biliary drainage is unlikely to improve aerobic fitness or modify their cardiopulmonary response to exercise/surgery \parencite{parker_serum_2014}.
@@
 
-Our results support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+In chapter \ref{ch_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay. 
+Taken together, our results support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
 
@@
 Preoperative systemic inflammation is increasingly recognised as playing an important role in determining not only long-term outcomes but also short-term outcomes after cancer surgery \parencite{kubo_elevated_2013, mohri_correlation_2014, moyes_preoperative_2009, neal_preoperative_2011, vashist_glasgow_2010,  haupt_association_1997}.
-While we did not find an association between elevated preoperative systemic inflammation in our cohort of patients, we demonstrated in chapter \ref{ch_pre_post_sirs} that preoperative systemic inflammation had a significant impact on  the magnitude of the postoperative inflammatory response.
+While we did not find an association between elevated preoperative systemic inflammation and complications in our cohort of patients, we demonstrated in chapter \ref{ch_pre_post_sirs} that preoperative systemic inflammation had a significant impact on  the magnitude of the postoperative inflammatory response.
 This supports the theory that abnormal systemic inflammation is a continuum that starts in the preoperative phase and continues into the postoperative phase.
@@
 Body composition appears to not only affect aerobic capacity but also systemic inflammation.
-Improvements in body composition are achievable through nutritional supplementation \parencite{machado_whey_2015} as well as increased physical activity.
+Improvements in body composition in cancer patients are achievable through nutritional supplementation \parencite{machado_whey_2015} as well as increased physical activity.
+Improvement in our understanding of cancer cachexia and the role of inflammation will also help optimise patients for treatment and prolong survival.
 
@@
 
+\subsection{Goal-directed therapy}
+
+The results of chapter \ref{ch_cpet_outcomes} and the work of Ausania and co-workers \parencite{ausania_effects_2012} suggest a link between reduced oxygen delivery and increased incidence of pancreatic fistula.
+We hypothesised that tissue hypoperfusion may have contributed to poor healing of the pancreatico-jejunal anastomosis.
+
+Reyad and co-workers performed a randomised trial of the effect of dobutamine infusion during pancreaticoduodenectomy on splanchnic perfusion, hemodynamics, and overall postoperative outcome \parencite{reyad_effect_2013}.
+They reported that intra-operative dobutamine use was associated with improved global oxygen delivery as measured by arterial and venous blood gases. 
+Splanchnic perfusion as measured by gastric tonometry also improved.
+It was interesting to note that the incidence of anastomotic leak was 30\% in the control group, 15\% in the group that received 3$\mu$g/kg/min of dobutamine and 5\% in the group that received 5$\mu$g/kg/min of dobutamine (p$<$0.05).
+The overall complication rate also decreased from 70\% in the control group to 40\% and 20\% in the dobutamine groups (p$<0.05$).
+
+However, the benefit of goal-directed therapy to improve organ perfusion and oxygen delivery in improving postoperative outcomes remains unclear \parencite{grocott_perioperative_2013, pearse_effect_2014}.
+A recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
+The study did show that patients whose postoperative oxygen delivery was similar to their preoperative values had fewer complications.
+However, oxygen delivery was not influenced by goal-directed therapy as any beneficial effect of the intervention was lost with the autonomic nervous system changes that accompanied the increased intravenous fluids and inotropes in the treatment cohort.
+
+Improving oxygen delivery and organ perfusion in surgical patients remain the subjects of intense research.
+However, the results presented in this thesis support the argument for therapeutic interventions aimed at improving oxygen delivery and organ perfusion before the patient with pancreatic cancer reaches the operating room as outlined in section \ref{disc_prehabilitation}.
+
 \subsection{Early recognition and aggressive management of complications}
 
+Postoperative complications and the associated systemic inflammatory response are associated with early recurrence and poor-survival after colorectal \parencite{artinyan_infectious_2015, mcardle_impact_2005}, gastric \parencite{hayashi_impact_2015,kubota_prognostic_2014}  and pancreatic cancer surgery \parencite{aoyama_impact_2015, kamphues_postoperative_2012}.
+
+Early recognition and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation. 
 Several recent studies have demonstrated the value of monitoring trends in postoperative CRP levels in predicting complications after pancreaticoduodenectomy \parencite{hiyoshi_usefulness_2013, ansorge_diagnostic_2014, kosaka_multivariate_2014}.
@@
 
-Postoperative complications and the associated systemic inflammatory response are associated with early recurrence and poor-survival after colorectal \parencite{artinyan_infectious_2015, mcardle_impact_2005}, gastric \parencite{hayashi_impact_2015,kubota_prognostic_2014}  and pancreatic cancer surgery \parencite{aoyama_impact_2015, kamphues_postoperative_2012}.
-
-Early recognition and aggressive management of complications is therefore important in mitigating the harmful long-term effects of such adverse events and the associated inflammation. 
 `Failure to rescue' after postoperative complications is one of the factors responsible for the discrepancy between morbidity and mortality in low-volume versus high-volume centres.
@@
 Early recognition of complications in these patients and aggressive management may result in improved outcomes in these high risk patients.
-
 These findings are supported by other studies involving emergency abdominal surgery \parencite{sheetz_improving_2013} and patients at major trauma centres \parencite{haas_prevention_2011}
@@
 
-\subsection{Enhanced Recovery Programmes}
-
-- centralisation
-- specialisation
-- early rescue and aggressive management of complications \parencite{gouma_rates_2000}
-
-\subsection{Intra-operative care}
-
-The results of chapter \ref{ch_cpet_outcomes} and the work of Ausania and co-workers appeared to suggest a link between reduced oxygen delivery and increased incidence of pancreatic fistula. 
-We hypothesised that tissue hypoperfusion may have played a role in the poor healing of the pancreatico-jejunal anastomosis.
-
-Reyad and co-workers performed a randomised trial to examine the effect of intra-operative dobutamine infusion during pancreaticoduodenectomy on splanchnic perfusion, hemodynamics, and overall postoperative outcome \parencite{reyad_effect_2013}.
-They reported that intra-operative dobutamine use was associated with improved global oxygen delivery as measured by arterial and venous blood gases, splanchnic perfusion as measured by gastric tonometry and postoperative complications.
-It was interesting to note that the incidence of anastomotic leak was 30\% in the control group, 15\% in the group that received 3$\mu$g/kg/min of dobutamine and 5\% in the group that received 5$\mu$g/kg/min of dobutamine (p$<$0.05).
-The overall complication rate also decreased from 70\% in the control group to 40\% and 20\% in the dobutamine groups (p$<0.05$).
-
-However, the benefit of goal-directed therapy to improve organ perfusion and oxygen delivery in improving postoperative outcomes remains unclear \parencite{grocott_perioperative_2013, pearse_effect_2014}.
-A recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
-The study did show that patients whose postoperative oxygen delivery was similar to their preoperative values had fewer complications.
-However, oxygen delivery was not influenced by goal-directed therapy as any beneficial effect of the intervention was lost with the autonomic nervous system changes that accompanied the increased intravenous fluids and inotropes in the treatment cohort.
-
-\subsection{Post-operative care}
-
 \section{Future directions}
 
-%Early recognition of patients at increased risk of postoperative complications facilitates prompt initiation of appropriate treatment and judicious allocation of critical care resources. 
-%Moreover, identification of patients who are unlikely to develop a major postoperative complication allows their progress on enhanced recovery pathways with the benefits of rapid recovery, reduced morbidity and early discharge.
-%These patients are also more likely to receive adjuvant therapy without undue delay. 
-%Recognition of these low risk patients helps improve outcomes, reduces hospital costs and releases valuable resources for the care of high risk patients. 
-%The results reported in Chapter [ss] demonstrate the value of early post-operative C-reactive protein in identifying patients who are more likely to develop an infective complication. 
-%The results show that CRP is only useful in the absence of a postoperative pancreatic fistula. 
-%The combination of low drain amylase and low CRP on the third postoperative day identifies patients who do not have a pancreatic fistula and are less likely to develop an infective complication.
+The work done in this thesis raises a number of important clinical questions. 
+Further studies on the inter-relationships between aerobic fitness, comorbidity, systemic inflammation at the metabolic level in patients with cancer will allow better understanding of their impact of postoperative outcomes.
+Objective measurement of aerobic fitness and systemic inflammation will allow for individualised perioperative care that starts at the time of diagnosis and continues in the postoperative period. 
+Further studies will have to focus on improving modifiable risk factors such as $\dot{V}_{O_2}$ starting with community/home-based interventions and continuing with individualised goal-directed therapy during and after surgery.
+Aggressive protocol-driven early rescue strategies that include identification of high risk patients before surgery, early recognition of complications and multi-disciplinary management may help improve postoperative outcomes and attenuate the adverse effects of complications on long-term survival.
+This in combination with individualised cancer treatments both in the neo-adjuvant and adjuvant settings will help improve outcomes in patients with pancreatic cancer.

### 2015-08-25 Tables/c1_clavien_dindo.tex
Minor commit following chat with Appa

@@
 		Grade V     & Death of a patient  \\ \hline
-		Suffix 'd': & If the patients suffers from a complication at the time of discharge,  the suffix  “d”  (for ‘disability’) is added to the respective grade of complication.
+		Suffix 'd': & If the patient suffers from a complication at the time of discharge,  the suffix  “d”  (for ‘disability’) is added to the respective grade of complication.
 This label indicates the need for a follow-up to fully evaluate the complication.

### 2015-08-25 Chapters/Discussion.tex
Discussion - Everything is in the right place.

@@
 They reported that intra-operative dobutamine use was associated with improved global oxygen delivery as measured by arterial and venous blood gases. 
-Splanchnic perfusion as measured by gastric tonometry also improved.
+Splanchnic perfusion measured by gastric tonometry also improved.
 It was interesting to note that the incidence of anastomotic leak was 30\% in the control group, 15\% in the group that received 3$\mu$g/kg/min of dobutamine and 5\% in the group that received 5$\mu$g/kg/min of dobutamine (p$<$0.05).
@@
 
-However, the benefit of goal-directed therapy to improve organ perfusion and oxygen delivery in improving postoperative outcomes remains unclear \parencite{grocott_perioperative_2013, pearse_effect_2014}.
+However, the benefit of such goal-directed therapy to improve organ perfusion and oxygen delivery in improving postoperative outcomes remains unclear \parencite{grocott_perioperative_2013, pearse_effect_2014}.
 A recent randomised controlled trial evaluated the role of goal-directed therapy in high-risk patients undergoing major elective surgery that targeted individualised oxygen delivery \parencite{ackland_individualised_2015}.
@@
 
-Improving oxygen delivery and organ perfusion in surgical patients remain the subjects of intense research.
-However, the results presented in this thesis support the argument for therapeutic interventions aimed at improving oxygen delivery and organ perfusion before the patient with pancreatic cancer reaches the operating room as outlined in section \ref{disc_prehabilitation}.
+%Improving oxygen delivery and organ perfusion in surgical patients remain the subjects of intense research.
+However, the results presented in this thesis support the argument for therapeutic interventions aimed at improving oxygen delivery and organ perfusion before the patient reaches the operating room as outlined in section \ref{disc_prehabilitation}.
 
@@
 `Failure to rescue' after postoperative complications is one of the factors responsible for the discrepancy between morbidity and mortality in low-volume versus high-volume centres.
-Mortality rates in patients undergoing a variety of complex surgical procedures including pancreatectomy was 2.5 times higher in low-volume centres than high-volume centres. 
+Mortality rates in patients undergoing a variety of complex surgical procedures including pancreatectomy was 2.5 times higher in low-volume centres than high-volume centres \parencite{ghaferi_complications_2009}.
 However, complication rates were similar between these hospitals. 
-Failure to identify and manage complications aggressively may have resulted in the increased mortality in the low-volume hospitals \parencite{ghaferi_complications_2009}.
-In a recent large study involving nearly 2 million surgical patients, Ferraris VA and co-workers studied the outcome of over two-hundred thousand patients who had complications. 
+Failure to identify and manage complications aggressively may have resulted in the increased mortality in the low-volume hospitals.
+In a recent large study involving nearly 2 million surgical patients, Ferraris VA and co-workers studied the outcome of over two-hundred thousand patients who had complications \parencite{ferraris_va_identification_2014}.
 They noted that 20\% of high-risk patients accounted for 90\% of deaths due to postoperative complications. 
-They also reported that a simple risk-scoring system based on preoperative characteristics was able to identify these patients. 
-Early recognition of complications in these patients and aggressive management may result in improved outcomes in these high risk patients.
-These findings are supported by other studies involving emergency abdominal surgery \parencite{sheetz_improving_2013} and patients at major trauma centres \parencite{haas_prevention_2011}
+They also reported that a simple risk-score based on preoperative characteristics was able to identify these patients. 
+Early recognition of complications in these high patients and aggressive management will reduce mortality.
+These findings are supported by other studies involving emergency abdominal surgery \parencite{sheetz_improving_2013} and major trauma \parencite{haas_prevention_2011}
 

### 2015-08-26 Chapters/Discussion.tex
Discussion - Mo Comments

@@
 %----------------------------------------------------------------------------------------
-\section{Summary of Aims}
+\section{Summary}
 
@@
 The present work also examined the factors that affect cardiopulmonary exercise physiology in order to enable a better understanding of the complex pathophysiology in these patients that is a consequence of the interaction of the patient's chronic conditions with the acute derangements brought on by pancreatic disease, including obstructive jaundice.
+
+This was one of the largest studies involving patients who underwent cardiopulmonary exercise testing before undergoing pancreaticoduodenectomy in a tertiary specialist pancreato-biliary unit in the West of Scotland.
+The results presented here demonstrate the clinical utility of cardiopulmonary exercise testing in major pancreatic surgery in the largest cohort of patients in published literature.
+
+The thesis also examined the effect of obstructive jaundice including severe obstructive jaundice on cardiopulmonary exercise physiology and postoperative systemic inflammation in this cohort. 
+We are not aware of any other published reports on this subject.
+
+This work also demonstrated a novel application of the trends in postoperative CRP in identifying patients at risk of infective complications which is different from the accepted paradigm for the use of postoperative CRP in published literature.
+
+The results presented here provide the foundations for future work on prehabilitation, mitigating perioperative risks and improving outcomes after pancreaticoduodenectomy.
 
@@
 
-The results presented in chapter \ref{ch_cpet_outcomes} demonstrated that cardiopulmonary exercise testing has a role in identifying high risk patients who are more likely to develop complications, stay longer in hospital and are less likely to receive adjuvant therapy after pancreaticoduodenectomy.
+The results presented in chapter \ref{ch_cpet_outcomes} demonstrated that cardiopulmonary exercise testing had a role in identifying high risk patients who were more likely to develop complications, stay longer in hospital and were less likely to receive adjuvant therapy after pancreaticoduodenectomy.
 These findings have since been replicated by other authors examining the role of cardiopulmonary exercise testing in pancreatic surgery.
@@
 
-The results of our work in chapter \ref{ch_cpet_outcomes} as well as that of several others over the past 2 decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
+The results of our work in chapter \ref{ch_cpet_outcomes} as well as that of several others over the past two decades support the use of cardiopulmonary exercise testing as a clinical risk assessment tool in patients undergoing major surgery.
 
@@
 We are aware of only one other similar study by Junejo and co-workers.
-They reported that peripheral oxygen extraction was normal during cardiopulmonary exercise in patients with malignant obstructive jaundice.\parencite{junejo_peripheral_2014}
+They reported that peripheral oxygen extraction was normal during cardiopulmonary exercise in patients with malignant obstructive jaundice \parencite{junejo_peripheral_2014}.
 Preoperative biliary drainage is unlikely to improve aerobic fitness or modify their cardiopulmonary response to exercise/surgery \parencite{parker_serum_2014}.
@@
 In their, appropriately titled paper \textit{`The Quandary of Preresection Biliary Drainage for Pancreatic Cancer'}, the authors discuss the complex physiological effects of obstructive jaundice, its impact on immune function and the differences between findings in animal studies and humans \parencite{tol_quandary_2012}.
-However, they also review recent literature where there appears to be overwhelming evidence that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications.
-\parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}.
+However, they also review recent literature where there appears to be overwhelming evidence that preoperative biliary drainage increases the incidence of postoperative complications, especially infective complications \parencite{van_der_gaag_preoperative_2010, arkadopoulos_preoperative_2014, fujii_preoperative_2015, furukawa_negative_2015}.
 
 In chapter \ref{ch_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay. 
-Taken together, our results support the emerging view that routine preoperative biliary drainage should no longer be performed in all patients scheduled to undergo a pancreaticoduodenectomy.
+Taken together, our results support the emerging view that preoperative biliary drainage should no longer be performed routinely in all patients undergoing a pancreaticoduodenectomy unless specifically indicated such as in cholangitis.
 
@@
 Improvements in body composition in cancer patients are achievable through nutritional supplementation \parencite{machado_whey_2015} as well as increased physical activity.
-Improvement in our understanding of cancer cachexia and the role of inflammation will also help optimise patients for treatment and prolong survival.
+Perioperative nutritional supplementation has been reported to be associated with reduced postoperative complications and improved outcomes \parencite{kabata_preoperative_2015, zhang_perioperative_2012}.
 

### 2015-08-26 main.tex
Main Summary
- Heavily edited summary to fit into 3 pages - dont change

@@
 \usepackage{easy-todo} %Adds todos to the document. Use [disable] as package option in final verion. Also see \listoftodos below.
-\usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
-\linenumbers
+%\usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
+%\linenumbers
 
@@
 Moreover, identifying preoperative risk factors provides potential targets for prehabilitation and optimisation of the patient's physiology before undertaking surgery. 
-In addition to this, early identification of patients who are likely to develop postoperative complications allows for better allocation of critical care resources and more targeted treatment of the high risk patients.
+In addition to this, early identification of patients who are likely to develop postoperative complications allows for better allocation of critical care resources and more aggressive management high risk patients.
 
@@
 
-The overall aim of the thesis was to evaluate the relationship between cardiopulmonary exercise testing, perioperative clinico-pathological characteristics and postoperative outcomes as well as long-term survival in patients undergoing pancreaticoduodenectomy.
+%The overall aim of the thesis was to evaluate the inter-relationship between cardiopulmonary exercise testing, perioperative clinico-pathological characteristics and postoperative outcomes as well as long-term survival in patients undergoing pancreaticoduodenectomy.
+
+The overall aim of the thesis was to examine the relationships between preoperative clinico-pathological characteristics including cardiopulmonary exercise physiology, obstructive jaundice, body composition and systemic inflammation and complications and the post-surgical systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
 
 \textbf{Chapter \ref{ch_intro}} reviews the existing literature on preoperative cardiopulmonary exercise testing, the impact of obstructive jaundice, perioperative systemic inflammation and the importance of body composition in determining outcomes in patients undergoing major surgery with particular reference to pancreatic surgery.
+
+\clearpage
 
@@
 \textbf{Chapter \ref{ch_cpet_jaundice}} evaluates the relationship between cardiopulmonary exercise physiology and other clinicopathological characteristics of the patient. 
-In a detailed analysis of cardiopulmonary exercise test parameters in jaundiced versus non-jaundiced patients, the results in this chapter demonstrate that obstructive jaundice does not impair cardiopulmonary exercise physiology.
+A detailed analysis of cardiopulmonary exercise test parameters in jaundiced versus non-jaundiced patients demonstrates that obstructive jaundice does not impair cardiopulmonary exercise physiology. 
 This further supports emerging evidence in contemporary literature that jaundiced patients can proceed directly to surgery without preoperative biliary drainage. 
@@
 
-\textbf{Chapter \ref{ch_survival}} analyses the relationship between preoperative comorbidity including cardiopulmonary exercise test and long-term survival in patients with pancreatic ductal adenocarcinoma undergoing pancreaticoduodenectomy.
+\clearpage
+
+\textbf{Chapter \ref{ch_pre_post_sirs}} analyses the relationship between preoperative clinico-pathological characteristics including systemic inflammation and the magnitude of the postoperative systemic inflammatory response.
+%The results of this chapter suggest that both preoperative systemic inflammation and obstructive jaundice have significant impact on postoperative inflammation. 
+Obstructive jaundice appears to have an immunosuppressive effect while elevated preoperative CRP and hypoalbuminemia appear to have opposite effects with hypoalbuminemia resulting in a lower response while elevated CRP in the absence of hypoalbuminemia resulted in a greater postoperative systemic inflammatory response.
+
+%\textbf{Chapter \ref{cf_survival}} analyses the relationship between preoperative comorbidity including cardiopulmonary exercise test and long-term survival in patients with pancreatic ductal adenocarcinoma undergoing pancreaticoduodenectomy.
 The results of this chapter suggest that cardiopulmonary exercise testing parameters are not useful in predicting long-term survival.
@@
 The results of this chapter demonstrate that CRP levels as early as the second postoperative day are associated with complications. 
-While post-operative CRP was useful in the prediction of infectious complications, this was only the case in patients who did not develop a post-operative pancreatic fistula.
+While post-operative CRP was useful in the prediction of infective complications, this was the case only in patients who did not develop a post-operative pancreatic fistula. 
 The predictive ability of inflammatory markers for infectious complications was blunted in patients with a pancreatic fistula. 
-In these patients, outcome was largely determined by the severity of the pancreatic fistula.
+%In these patients, outcome was largely determined by the severity of the pancreatic fistula.
 
 \textbf{Chapter \ref{ch_discussion}} summarises the findings of this thesis, their place in current literature and future directions. 
-The contents of this thesis aim to add to the current body of knowledge regarding the complex pathophysiological alterations that occur in patients with lesions of the head of the pancreas with the associated obstructive jaundice, systemic inflammation and their impact on cardiopulmonary exercise physiology, postoperative complications and long-term outcomes.
-The results establish the value of cardiopulmonary exercise test as an objective measure of aerobic fitness while refuting a commonly held belief that obstructive jaundice significantly affected cardiovascular function.
-The current work paves the way for further study into evaluating the role of cardiopulmonary exercise test in prehabilitation regimens as well as the incorporation of preoperative $\dot{V}_{O_2}$AT and postoperative CRP into existing models for predicting complications after pancreaticoduodenectomy.
-These results will help improve patient selection and preparation for pancreaticoduodenectomy  and ultimately the short and long-term outcomes after this morbid yet necessary operation.
+The results of this thesis add to the current knowledge regarding the complex pathophysiological abnormalities in patients undergoing pancreaticoduodenectomy, with specific emphasis on the interaction between cardiopulmonary exercise physiology, obstructive jaundice, systemic inflammation and postoperative outcomes.
+%The results establish the value of cardiopulmonary exercise test while refuting a commonly held belief that obstructive jaundice significantly affected cardiovascular function. 
+%The current work paves the way for further study into evaluating the role of cardiopulmonary exercise test in prehabilitation regimens as well as the incorporation of preoperative $\dot{V}_{O_2}$AT and postoperative CRP into existing models for predicting complications after pancreaticoduodenectomy.
+%These results will help improve patient selection and preparation for pancreaticoduodenectomy and ultimately improve the short and long-term outcomes after this morbid yet necessary operation.
+
+The work presented in this thesis lays the foundations for further studies aimed at improving outcomes after pancreaticoduodenectomy through the development of individualised, goal-directed therapies that are initiated well before this morbid yet necessary operation is performed.
 
@@
 
-%Mr. Euan J Dickson,
-%West of Scotland Pancreatic Unit, Glasgow Royal Infirmary
+Mr. Euan J Dickson,
+West of Scotland Pancreatic Unit, Glasgow Royal Infirmary
 }
@@
 
-\input{Chapters/Abbreviations}
+%\input{Chapters/Abbreviations}
 
@@
 \input{Chapters/BodyComp}			%4
-%\input{Chapters/CPET_Survival}		%5
-\input{Chapters/Pre_Post_SIRS}
+\input{Chapters/Pre_Post_SIRS}		%5
 \input{Chapters/CRP_Complications}	%6
@@
 
+%\input{Chapters/CPET_Survival}
 %----------------------------------------------------------------------------------------

### 2015-08-26 Chapters/CPET_Jaundice.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday

@@
 
-	\caption{Pearson's correlation and scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in patients undergoing pancreaticoduodenectomy.}
+	\caption{Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak.}
 	\label{fig:cpet_oj_scatter}
@@
 Sarcopenia is associated with reduced aerobic capacity \parencite{evans_sarcopenia_1993} and is common in pancreatic cancer \parencite{joglekar_sarcopenia_2015}.
+$\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak must be interpreted with caution in these patient groups.
 This is the subject of further study in Chapter \ref{ch_bodycomp}.
@@
 Both obstructive jaundice and low $\dot{V}_{O_2}$AT were associated with elevated preoperative systemic inflammation and provide further evidence of the role of jaundice and impaired aerobic capacity in immune dysregulation in these patients.
-High BMI and female sex were independently associated with low $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak and these parameters must be interpreted with caution in these subgroups.

### 2015-08-26 Chapters/CPET_Outcomes.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday

@@
 The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}. 
-In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.
+In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy in selected patients \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.
 

### 2015-08-26 Chapters/CRP_Complications.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday

@@
 \section{Results}
+
 \subsection{POPF, infective complications and other outcomes}
@@
 %CRP trends and pancreatic fistula
-\subsection{Postoperative CRP and POPF}
+\subsection{Postoperative CRP vs. POPF}
+\label{Postoperative CRP vs. POPF}
 Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p<0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
@@
 
-\subsection{Postoperative CRP and infective complications}
+\subsection{Postoperative CRP vs. infective complications}
 The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
@@
 
-\subsection{Postoperative white cell counts, albumin and infective complications}
+\subsection{Postoperative WCC, albumin vs. infective complications}
 

### 2015-08-26 Chapters/Discussion.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday

@@
 
-In chapter \ref{ch_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay. 
+In chapter \ref{ch_cpet_outcomes} we reported that obstructive jaundice was not associated with increased incidence of postoperative complications or prolonged hospital stay. 
 Taken together, our results support the emerging view that preoperative biliary drainage should no longer be performed routinely in all patients undergoing a pancreaticoduodenectomy unless specifically indicated such as in cholangitis.

### 2015-08-26 Chapters/Pre_Post_SIRS.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday

@@
 \begin{figure}[p]
-	\caption{Relationship between preoperative obstructive jaundice and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
+	\caption{Relationship between obstructive jaundice and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
 	\label{fig:sirs_bilirubin}
@@
 \begin{figure}[p]
-	\caption{Relationship between preoperative $\dot{V}_{O_2}$AT and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
+	\caption{Relationship between $\dot{V}_{O_2}$AT and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
 	\label{fig:sirs_at}

### 2015-08-26 Chapters/CRP_Complications.tex
Final Review - CRP Complications
- Numerous minor changes to text and to multiple tables

@@
 
-The aim of this study was to investigate the association between systemic inflammation, postoperative pancreatic fistula and infective complications in patients undergoing pancreaticoduodenectomy.
+\subsection{Aim}
+The aim of this study was to investigate the association between trends in postoperative systemic inflammation, postoperative pancreatic fistula and infective complications in patients undergoing pancreaticoduodenectomy.
 
@@
 All other complications were graded using the Clavien-Dindo classification system as shown in Table \ref{table:clavien-dindo}. 
-This allowed the grading of the severity of the complications based on the magnitude of the intervention(s) required to treat them. 
-Postoperative mortality was defined as death within 30-days of the operation or while still in hospital after the operation. 
+This allowed grading of the severity of complications based on the magnitude of intervention(s) required to treat them. 
+Postoperative mortality was defined as death within 30-days of surgery or while still in hospital after the operation. 
 Re-intervention in the form of radiological, endoscopic or surgical procedures was recorded prospectively.
@@
 Receiver operating characteristic (ROC) analysis \parencite{robertson_use_1981,zweig_receiver-operating_1993} was used to identify the optimum thresholds for postoperative CRP for predicting infective complications with a preference for thresholds that had a greater negative predictive value.
+
 Patients were categorised using these thresholds to analyse the relationship between CRP (as a categorical variable) and re-operation, hospital stay, critical care stay, number of critical care admissions as well as postoperative mortality using the Chi-square test.
@@
 Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
-However, there was no such association in patients who did develop a POPF (Fig.
-\ref{fig:crp_comp_infective_leak1}).
+However, there was no such association in patients who did develop a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
+
 For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
@@
 
-Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically significant However, these differences were small with wide, overlapping 95\% confidence intervals (Fig.
-\ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant.
+Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically significant.
+However, these differences were small with wide, overlapping 95\% confidence intervals (Fig. \ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant. 
 For instance, the median white cell count on the fifth postoperative day in patients who did not have an infective complication was 8.9 (IQR 6.6 - 11) while in patients who had an infective complication it was 9.8 (IQR 7.9 - 13).
+
 While this difference was statistically significant (p = 0.028), both medians were within the normal limits for healthy adults. 
@@
 The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infective complication.
+
 A CRP level of 125 mg/L on the fourth postoperative day had a sensitivity of 74\% and specificity of 64\% with a negative predictive value of 83\% for predicting infective complications (AUC 0.71, 95\%CI 0.61 - 0.80, p$<$0.001).
@@
 The modified Glasgow Prognostic Score, a validated measure of preoperative systemic inflammation, has also been reported to be associated with postoperative infective complications in patients undergoing curative colorectal resections \parencite{moyes_preoperative_2009}.
+
 The lack of association between the modified Glasgow Prognostic Score and complications in the present study may have been due to the multiple factors that are responsible for preoperative systemic inflammation in these patients including obstructive jaundice, cholangitis, biliary drainage procedures and pancreatitis consequent to intervention or pancreatic duct obstruction by tumours.
@@
 Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications.
+
 Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer. 
-They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day.
+They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day \parencite{mokart_early_2002}.
 T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6 \parencite{faist_immunosuppression_1997}. 
@@
 This allows early identification of patients who are expected to continue on the enhanced recovery pathway and allocation of resources to improve the management of patients who are likely to develop complications.
+A recent meta-analysis of seven studies including 1427 patients established the value of CRP on the third postoperative day in identifying patients who are safe for early discharge on an enhanced recovery programme \parencite{straatman_predictive_2015}.
 This study shows that CRP on the third postoperative day is associated with infective complications in patients undergoing pancreaticoduodenectomy who do not develop a POPF but not in patients with a POPF.
-
-\todo{his is hot-off the press - reference this}%straatman_predictive_2015 - T

### 2015-08-26 Tables/crp_comp_CRP4_vs_LOS.tex
Table reformatting
- Reformat all tables for borders
- Check most captions
- Must check captions, decimals, etc. again for each chapter
Huge commit following on from Donny conversation
- Change all references to before period
- Remove post 2011/2012 references from introduction - one 2012 reference left was published online in Aug 2011
- Rename title - 'investigation into clinical utility of CPET'
- Remove hyphenation of words
- Left justify with ragged right edge
- Remove bold formatting in some tables
- Minor grammar and style corrections in multiple places
- Change reference Chandrabalan 20

@@
 	\centering
-	\caption{The relationship between CRP on Day 4 and other postoperative events in patient with \textbf{no} POPF.}
+	\caption{The relationship between CRP on Day 4 and other postoperative events in patients with no POPF.}
 	\label{table:crp_comp_CRP4_vs_LOS}
-	\renewcommand{\arraystretch}{1.2} %Increases space between rows
-	%\setlength{\tabcolsep}{9pt} %sets the space between columns
-	\begin{tabular}{|l l | c c r |}
+	\renewcommand{\arraystretch}{1.4} %Increases space between rows
+	\setlength{\tabcolsep}{9pt} %sets the space between columns
+	\begin{tabular}{|l l  c c r |}
 		\hline
-		                      &          &  \multicolumn{3}{c|}{Day 4 CRP (mg/dl)}   \\
+		                      &          &  \multicolumn{2}{c|}{Day 4 CRP (mg/dl)}   \\
 		                      &          & $\leq$125   & $>$125       & p            \\ \hline

### 2015-08-26 Chapters/Introduction.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday
Introduction - Rajeev comments and a few other minor edits

@@
 The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery \parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.
-.
 
@@
 \subsection{Oesophago-gastric and bariatric surgery}
-In a retrospective study involving 91 patients who underwent curative oesophagectomy with 3-field lymphadenectomy via a
-right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$ predicted cardiopulmonary complications \parencite{nagamatsu_preoperative_2001}.
+In a retrospective study involving 91 patients who underwent curative oesophagectomy with 3-field lymphadenectomy via a right thoracotomy for squamous cell carcinoma of the thoracic oesophagus between 1991 and 1995, Nagamatsu and co-workers reported that $\dot{V}_{O_2}max/m^2<800$ml/min/m$^2$ predicted cardiopulmonary complications \parencite{nagamatsu_preoperative_2001}.
 They also reported that $\dot{V}_{O_2}AT/m^2$ and routine spirometry did not predict cardiopulmonary complications. 
@@
 These include cirrhotic cardiomyopathy, hepatopulmonary syndrome and decreased peripheral oxygen utilisation due to cirrhotic myopathy \parencite{epstein_aerobic_2004}. 
-This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery \parencite{iscar_functional_2009}. . 
+This is further supported by the fact that patients who undergo liver transplantation have been shown to have improved aerobic capacity a year after surgery \parencite{iscar_functional_2009}. 
 However, there is very little other evidence of the application of cardiopulmonary exercise testing in patients undergoing major hepato-pancreato-biliary surgery or liver transplantation.
@@
 Similar findings have been reported after curative surgery for advanced gastric cancer with patients who develop an anastomotic leak surviving for 30.5 months while patients who did not have a leak survived for a median of 96.2 months (p$<$0.001) \parencite{yoo_negative_2011}.
-.
 
@@
 The absence of bile in the gut increases the proliferation of gram negative bacteria. 
-The absence of bile in the gut of bile duct ligated mice was associated with intestinal bacterial overgrowth and bacterial translocation \parencite{deitch_obstructive_1990}
+The absence of bile in the gut of bile duct ligated mice was associated with intestinal bacterial overgrowth and bacterial translocation \parencite{deitch_obstructive_1990}.
 
@@
 
-Increased serum bilirubin and bile salts directly affect lymphocyte function as well as the response to pro-inflammatory cytokines \parencite{scott-conner_pathophysiology_1994}
+Increased serum bilirubin and bile salts directly affect lymphocyte function as well as the response to pro-inflammatory cytokines \parencite{scott-conner_pathophysiology_1994}.
 Bile salts in higher than normal concentrations also induce apoptosis in hepatocytes. 
@@
 Mezhir and co-workers also reported a high (23\%) stent related complication rate in their retrospective study of 340 patients where 96\% of 201 patients in the biliary drainage group had plastic stents placed.
-However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage \parencite{mezhir_matched_2009}
+However, they also reported a greater incidence of wound infections (7\% vs 20\%, p=0.01) and intra-abdominal abscesses (3\% vs 12\%, p=0.03) in patients who underwent biliary drainage \parencite{mezhir_matched_2009}.
 
@@
 In his attempts to describe the characteristics of 'normal man', Quetelet found that weight increased nearly as the square of the height except at times of growth spurts in the new born and during puberty \parencite{quetelet_treatise_1842}.
-.
 The calculation of body mass index is simple and as follows:

### 2015-08-26 Chapters/BodyComp.tex
3 days to submission
- Had a long chat with dad
- Had a long chat with Mo - thanks man
- Final review tomorrow
- Aim to submit on Thursday
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

@@
 Major abdominal surgery especially for pancreatic disease is associated with significant morbidity and mortality. 
-Patient selection is as important as identifying surgical treatable pathology in ensuring optimal outcomes \parencite{balthazar_acute_2002}.
+Patient selection is as important as identifying surgically treatable pathology in ensuring optimal outcomes \parencite{bilimoria_national_2007, sandroussi_sociodemographics_2010}.
 
@@
 A number of studies have shown that poor aerobic fitness demonstrated by a low $\dot{V}_{O_2}$AT or low $\dot{V}_{O_2}$Peak or both was associated with increased morbidity and mortality after major surgery including bariatric \parencite{mccullough_cardiorespiratory_2006}, pancreatic \parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes}], liver \parencite{epstein_aerobic_2004}, cardiothoracic \parencite{brunelli_risk_2010, campione_oxygen_2010,torchio_exercise_2010} and abdominal aortic aneurysm surgery \parencite{carlisle_mid-term_2007,thompson_cardiopulmonary_2011}.
-It is now routinely used as part of the preoperative work-up to aid in risk assessment and patient selection.
-Cardiopulmonary exercise tests also help in identifying patients who may benefit from optimisation and more focussed care before, during and after major pancreatic surgery.
-Patients are sometimes denied surgery if their aerobic fitness as determined by cardiopulmonary exercise testing is felt to be too poor based on currently available evidence.
+It is now routinely used in preoperative assessment to aid in risk assessment and patient selection.
+Cardiopulmonary exercise tests also help in identifying patients who may benefit from perioperative optimisation and more focussed care before, during and after major pancreatic surgery.
+Occasionally, patients may be denied surgery if their aerobic fitness as determined by cardiopulmonary exercise testing is felt to be too poor based on currently available evidence.
 
@@
 Cardiopulmonary exercise testing allows the accurate measurement of most of these factors either directly or indirectly during dynamic exercise.
-This allows identifying not only limitations in aerobic fitness but also the cause for such limitation.
+This allows identifying not only limitations in aerobic fitness but also the possible causes for such limitation.
 
@@
 Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (p\pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (p\pageref{sec:cpx_parameters}).
+A complete dataset for one sample patient is shown in Appendix \ref{AppendixCPETRawData} on p\pageref{AppendixCPETRawData}.
 
@@
 This resulted in a length in millimetres for each pixel in each image. 
-As pixels on a CT image are square, the area of each pixel was calculated as a square of its length.
+As pixels on a CT image are square, the area of each pixel was calculated as the square of its length.
 
@@
 The same process was repeated for visceral adipose tissue and skeletal muscle as depicted in Fig. \ref{fig:bc_ct_vat} and Fig. \ref{fig:bc_ct_sm} respectively. 
-Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the area under selection as shown in the insets in each of the images.
+Every selection was visually confirmed for anatomical accuracy by using the layer selection tool to inspect the selected area as shown in the insets in each of the images.
 %CT slices images x 4
@@
 \begin{equation} \label{eq:elbm_men}
-	Men:\ eLBM = 0.407 \times weight(kg) + 0.267 \times height(cm) - 19
+	Men:\ eLBM = 0.407 \times weight(kg) + 0.267 \times height(cm) - 19.2
 \end{equation}
@@
 The relationship between body composition and preoperative clinico-pathological characteristics (as categorical variables) was analysed using non-parametric tests. 
-The Mann-Whitney U test for variables with two categories or the Kruskal-Wallis test for variables with more than two categories. 
+The Mann-Whitney U test was used for variables with two categories while the Kruskal-Wallis test was used for variables with more than two categories. 
 Previously established thresholds were used for categorising continuous variables where applicable. 
@@
 
-Patients from more deprived areas of Scotland as defined by the Scottish Index of Multiple Deprivation (SIMD) had a greater amount of subcutaneous adipose tissue (p=0.041) but no difference in the other body composition parameters.
+There was no association between deprivation as defined by the Scottish Index of Multiple Deprivation (SIMD) and body composition.
 Hypoalbuminemia was associated with less visceral adipose tissue (p=0.013) and less skeletal muscle area that did not reach statistical significance (p=0.054). 

### 2015-08-26 Chapters/CPET_Jaundice.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

@@
 \subsection{Aim}
-The results presented in Chapter \ref{ch_cpet_outcomes} demonstrated that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy resulting in an increased incidence of POPF and prolonged hospital stay.
-Obstructive jaundice was present in a large proportion of these patients.
-The aim of the present study was to evaluate the relationship between cardiopulmonary exercise testing, obstructive jaundice (especially severe obstructive jaundice with serum bilirubin $>$ 250 $\mu$mol/l) and preoperative pathophysiology in patients undergoing major pancreatic surgery.
+The results presented in Chapter \ref{ch_cpet_outcomes} demonstrated that poor performance at cardiopulmonary exercise testing (CPET) was associated with adverse outcomes after pancreaticoduodenectomy with an increased incidence of POPF and prolonged hospital stay.
+Obstructive jaundice was present in a large proportion of these patients but not associated with prolonged hospitalisation (Table \ref{table:cpet_outcomes_table3}).
+The aim of the present study was to evaluate the relationship between cardiopulmonary exercise testing, obstructive jaundice (especially severe obstructive jaundice with serum bilirubin $>$ 250 $\mu$mol/l) and preoperative pathophysiology in patients undergoing pancreaticoduodenectomy.
 
@@
 \subsection{Patients}
-Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and April 2013 and had undergone cardiopulmonary exercise testing as part of their preoperative work-up at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
+Patients who underwent classical or pylorus-preserving pancreaticoduodenectomy for periampullary lesions (both benign and malignant) between August 2008 and December 2012 and had undergone cardiopulmonary exercise testing as part of their preoperative assessment at the West of Scotland Pancreatic Unit, Glasgow Royal Infirmary, Glasgow were included in the study.
 Established criteria for resectability in patients with malignant disease were used as outlined in Section \ref{sec:resectability_criteria}. 
@@
 Patient demographics, preoperative clinico-pathological characteristics including cardiorespiratory comorbidity, results of preoperative blood tests, chest x-ray, ECG and cardiopulmonary exercise tests were collected from prospectively held databases.
-The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease, respiratory disease, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) and was used as an objective score of comorbidity.
+The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease, respiratory disease, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) and was used as an objective score of comorbidity (Table \ref{table:intro_possum} on p\pageref{table:intro_possum}).
 Cardiovascular comorbidity was defined as a score of 2 or more for either the cardiac disease or ECG component of the POSSUM score. 
@@
 Obstructive jaundice was defined as serum bilirubin $>$ 35 $\mu$mol/litre and severe obstructive jaundice was defined as serum bilirubin $>$ 250 $\mu$mol/litre. 
-These thresholds were chosen as the recent randomised controlled trial that demonstrated increased complications in patients who underwent PBD excluded patients with serum bilirubin $>$ 250 $\mu$mol/litre from the study \parencite{van_der_gaag_preoperative_2010}.
+These thresholds were chosen as a recent randomised controlled trial that demonstrated increased complications in patients who underwent PBD excluded patients with serum bilirubin $>$ 250 $\mu$mol/litre from the study \parencite{van_der_gaag_preoperative_2010}.
 The present study aimed to evaluate preoperative pathophysiology in this particular group of patients with `severe obstructive jaundice'.
@@
 Cardiopulmonary exercise testing methodology and calculation of $\dot{V}_{O_2}$AT is described in detail in Sections \ref{sec:cpx_method} and \ref{sec:cpx_at_method}.
-A more detailed description of cardiopulmonary exercise testing parameters described in this study are published elsewhere \parencite{balady_clinicians_2010}.
+The cardiopulmonary exercise testing parameters described in this study are explained in detail in Section \ref{sec:cpx_parameters} on p\pageref{sec:cpx_parameters} and in published literature \parencite{balady_clinicians_2010}.
 
@@
 Univariate and multivariate binary logistic regression analysis was used to study the relationship between preoperative patient characteristics and $\dot{V}_{O_2}$AT / $\dot{V}_{O_2}$Peak. 
+Scatter-plots were used for visual representation of the relationship between serum bilirubin and $\dot{V}_{O_2}$.
+
+The level of significance was set at $p<0.05$.
 SPSS software (Version 17.0; SPSS Inc., Chicago, IL, USA) was used to perform statistical analysis.
@@
 The baseline demographic and clinical characteristics of non-jaundiced and jaundiced patients are shown in Table \ref{table:cpet_oj_patient}. 
-A greater proportion of jaundiced patients were females compared to the non-jaundiced cohort (p=0.028) and smokers (p=0.038). 
+A greater proportion of jaundiced patients were females (p=0.028) and smokers (p=0.038) compared to the non-jaundiced cohort. 
 Elevated POSSUM Physiology Score (p=0.004) and malignancy (p$<$0.001) were significantly associated with the presence of jaundice. 
@@
 There was a trend towards lower forced vital capacity with increasing severity of jaundice but this did not reach statistical significance (p=0.092). 
-There was no association between the other pulmonary function tests and jaundice. (Table \ref{table:cpet_oj_pft}).
+There was no association between the other pulmonary function tests and jaundice (Table \ref{table:cpet_oj_pft}).
 
@@
 Cardiopulmonary exercise test parameters measured at the anaerobic threshold and at peak exercise were compared between jaundiced and non-jaundiced patients using the Kruskal-Wallis test. 
-Obstructive jaundice was associated with lower tidal volume (p=0.017), lower absolute $\dot{V}_{O_2}$ (p=0.003), lower corrected $\dot{V}_{O_2}$ (p=0.029), lower oxygen pulse (p=0.037) and lower respiratory rate (p=0.022).
+Obstructive jaundice was associated with lower tidal volume (p=0.017), lower absolute $\dot{V}_{O_2}$ (p=0.003), lower corrected $\dot{V}_{O_2}$ (p=0.029), lower oxygen pulse (p=0.037) and lower respiratory rate (p=0.022) at the anaerobic threshold.
 At peak exercise, jaundice was associated with lower tidal volume (p=0.028), lower absolute $\dot{V}_{O_2}$Peak (p=0.007), lower absolute $\dot{V}_{CO_2}$ (p=0.016) and lower end-tidal $O_2$ or $PET_{O_2}$ (p=0.026).
@@
 The results also demonstrate that obstructive jaundice is associated with elevated preoperative systemic inflammation with the severity of systemic inflammation related to the severity of the jaundice.
+The results also appear to suggest that there may be multiple other determinants of aerobic capacity including female sex, body mass index, haemoglobin, presence of cancer and systemic inflammation.
 
@@
 The use of CPET in preoperative risk prediction was first made popular over two decades ago by Older and co-workers \parencite{older_preoperative_1993}. 
-Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012},[Chapter \ref{ch_cpet_outcomes}], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery.
+Since then cardiopulmonary exercise testing has been reported to be useful in identifying high risk patients prior to major general \parencite{snowden_submaximal_2010}, pancreatic \parencite{ausania_effects_2012}[Chapter \ref{ch_cpet_outcomes} of this thesis], oesophagogastric \parencite{nagamatsu_preoperative_2001} as well as vascular \parencite{carlisle_mid-term_2007} surgery.
 Cardiopulmonary exercise testing is superior to conventional measures of fitness due to the dynamic nature of the test that evaluates the adequacy of oxygen delivery to tissues under physiological stress.
@@
 Historically, obstructive jaundice has also been reported to be associated with adverse haemodynamic events in patients undergoing major surgery. 
-Intra-operative blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were all more common in patients with obstructive jaundice \parencite{dixon_factors_1983, pain_perioperative_1985, green_systemic_1995}.
-A recent review noted that bile acids have a complex range of receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}. 
+Intra-operative blood loss, postoperative hypotension, increased susceptibility to shock and renal dysfunction were more common in patients with obstructive jaundice \parencite{dixon_factors_1983, pain_perioperative_1985, green_systemic_1995}.
+A recent review noted that bile acids had a complex range of receptor mediated effects on the cardiovascular system \parencite{khurana_bile_2011}. 
 Moreover, some of these effects appear to be partly reversible by biliary drainage as demonstrated by Padillo and co-workers \parencite{padillo_improved_2001}.
-This increased incidence of complications resulted in routine pre-operative biliary drainage being recommended in jaundiced patients before undertaking major surgery. 
+
+These observations led to routine pre-operative biliary drainage being recommended in all jaundiced patients before undertaking major surgery. 
 Pancreaticoduodenectomy was described by Whipple initially as a two-stage operation, with the first stage involving a cholecysto-gastrostomy aimed at relieving biliary obstruction before undertaking the resection at a later second operation \parencite{whipple_treatment_1935}.
@@
 However, several recent studies have reported that PBD is associated with increased incidence of complications and that surgery in the jaundiced patients is safe.
-Pitt and co-workers in a prospective randomised trial comparing outcomes in jaundiced patients undergoing surgery with or without PBD reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
+Pitt and co-workers in a prospective randomised trial compared outcomes in jaundiced patients undergoing surgery with or without PBD.
+They reported that PBD was associated with increased cost without any decrease in postoperative complications \parencite{pitt_does_1985}.
 A recent meta-analysis \parencite{sewnath_meta-analysis_2002} pooled data from 5 randomised controlled trials comparing surgery with PBD versus surgery without drainage.
@@
 These findings are similar to those of Parker and co-workers who reported that there was only a weak negative correlation between serum bilirubin and  $\dot{V}_{O_2}$Peak (Pearson's $\rho$ -0.21, p=0.02) and no correlation with $\dot{V}_{O_2}$AT ($\rho$ = -0.15, p=NS) in patients undergoing pancreaticoduodenectomy \parencite{parker_serum_2014}.
-Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein. 
-They reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}
+Junejo and co-workers studied oxygen extraction in 9 jaundiced patients during cardiopulmonary exercise test by placing catheters within the femoral vein and reported that oxygen extraction was normal during cardiopulmonary exercise even in the presence of jaundice \parencite{junejo_peripheral_2014}
 
@@
 Obstructive jaundice is known to be associated with wide ranging abnormalities of both humoral and innate immune systems \parencite{nehez_compromise_2002, padillo_cytokines_2001, scott-conner_pathophysiology_1994}.
-It is unclear if this immune dysfunction before surgery is associated with an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
-However, the immunological abnormalities associated with obstructive jaundice do not appear to improve after biliary drainage \parencite{kimmings_endotoxin_2000}.
+It is unclear if this preoperative immune dysfunction continues after surgery as an abnormal postoperative systemic inflammatory response and this is explored further in Chapter \ref{ch_pre_post_sirs}.
+However, studies suggest that the immunological abnormalities associated with obstructive jaundice did not improve after biliary drainage \parencite{kimmings_endotoxin_2000}.
 
@@
 %\subsection{Conclusion}
-Obstructive jaundice, including severe obstructive jaundice did not affect preoperative cardiopulmonary exercise physiology. 
+In conclusion, obstructive jaundice, including severe obstructive jaundice did not affect preoperative cardiopulmonary exercise physiology. 
 Therefore, preoperative biliary drainage cannot be expected to improve aerobic capacity in patients undergoing pancreaticoduodenectomy.

### 2015-08-26 Chapters/CPET_Outcomes.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

@@
 \section{Introduction}
-Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 16-17\% surviving beyond the first year and 3\% surviving beyond 5 years \parencite{cancerresearchuk_cancer_2014}.
-The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014, sener_pancreatic_1999}. 
+Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death with only 21\% surviving beyond the first year and 3\% surviving beyond 5 years \parencite{cancerresearchuk_cancer_2011}.
+The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2011, sener_pancreatic_1999}. 
 In patients with resectable disease, surgery \parencite{sener_pancreatic_1999, sohn_resected_2000, geer_prognostic_1993} followed by adjuvant chemotherapy in selected patients \parencite{neoptolemos_randomized_2004,neoptolemos_adjuvant_2009} remains the primary modality of cure.
@@
 %Figure 1
-\begin{figure}[htbp]
+\begin{figure}[p]
 	\centering
-	\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_dist_of_AT}
+	\includegraphics[width=\textwidth]{Figures/cpet_outcomes_dist_of_AT}
 	\caption{Distribution of $\dot{V}_{O_2}$AT across the study population.}
@@
 % Figure 2
-\begin{figure}[h]
+\begin{figure}[p]
 \centering
-\includegraphics[width=0.8\linewidth]{Figures/cpet_outcomes_km_at_los}
-\includegraphics[width=1\linewidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
+\includegraphics[width=\textwidth]{Figures/cpet_outcomes_km_at_los}
+\includegraphics[width=\textwidth]{Figures/cpet_outcomes_km_at_los_table} %Table below Kaplan Meier curve
 \caption{Kaplan-Meier Plot of postoperative length of stay in patients with $\dot{V}_{O_2}$AT $>=$ 10 ml/kg/min versus $<$ 10 ml/kg/min.}
@@
 The proportion of patients with $\dot{V}_{O_2}$AT less than 11 ml/kg/min in this study was much greater than reported in studies involving patients undergoing oesophageal surgery (16\%),\parencite{forshaw_is_2008} liver transplantation (39\%)\parencite{epstein_aerobic_2004} or other major abdominal surgery (29\%)\parencite{older_preoperative_1993} and may indicate the poor preoperative fitness levels of patients undergoing major pancreatic surgery at our unit.
+
 While several studies have shown that low $\dot{V}_{O_2}$AT and/or low $\dot{V}_{O_2}$peak are associated with postoperative complications or prolonged hospital stay following major abdominal surgery as well as non-abdominal surgery,\parencite{older_preoperative_1993, epstein_aerobic_2004, mccullough_cardiorespiratory_2006, nagamatsu_preoperative_2001, older_cardiopulmonary_1999, older_clinical_2004} others have disputed this \parencite{forshaw_is_2008, clayton_cardiopulmonary_2011, hightower_pilot_2010}.
+
 Older and co-workers reported in 1993 that low $\dot{V}_{O_2}$AT less than 11 ml/kg/min was associated with a significantly higher risk of postoperative mortality from cardiovascular causes in a series of 187 elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}.
+However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the $\dot{V}_{O_2}$AT did not predict postoperative adverse events less after oesophagectomy.
 
-However, Snowden and co-workers \parencite{snowden_submaximal_2010} reported that patients with an $\dot{V}_{O_2}$AT less than 10.1 ml/kg/min had significantly greater cardiopulmonary complications as well as non-cardiopulmonary and infectious complications while Forshaw and co-workers \parencite{forshaw_is_2008} reported that using a cut-off of 11 ml/kg/min for the $\dot{V}_{O_2}$AT did not predict postoperative adverse events less after oesophagectomy.
 The lack of association between low $\dot{V}_{O_2}$AT and cardiopulmonary complications in this study may have been due to two reasons. 

### 2015-08-26 Chapters/Introduction.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

@@
 They may be broadly classified as benign and malignant. 
-Most pancreatic neoplasia are malignant and arise from the exocrine component of the gland, the ductal epithelium.
+Most pancreatic neoplasia are malignant \parencite{li_pancreatic_2004}.
 
-Pancreatic ductal adenocarcinoma is the most common cancer of the pancreas. 
+Pancreatic ductal adenocarcinoma arising from the ductal epithelium of the exocrine part of the pancreas is the most common cancer of the pancreas. 
 However, the head of the pancreas is anatomically related to several other epithelium lined structures that can also give rise to cancers. 
@@
 Pancreatic cancer is the tenth most common cancer in the UK but the fifth most common cause of cancer death. 
-Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2014}. 
-The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2014,sener_pancreatic_1999}.
+Only 21\% survive beyond the first year and 5-year survival is 3\% \parencite{cancerresearchuk_cancer_2011}. 
+The majority of patients (80-85\%) with pancreatic cancer present with inoperable disease \parencite{cancerresearchuk_cancer_2011,sener_pancreatic_1999}.
 
@@
 
-The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon results in early involvement of these structures even by relatively small tumours.
+The anatomical location of the pancreas, deep within the retroperitoneum, surrounded by numerous vital blood vessels including the coeliac trunk and its branches, the superior mesenteric artery, portal vein and superior mesenteric vein as well as proximity to other viscera such as the stomach, duodenum, transverse colon results in early involvement of these structures even by relatively small tumours \parencite{li_pancreatic_2004}.
 Moreover, symptoms are often absent in the early stages and when present are too non-specific to help with diagnosis. 
@@
 Endoscopic Ultrasound (EUS) is used in selected patients to assess vascular involvement and for obtaining tissue samples for histological examination. 
-In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) plays an important role in the alleviation of jaundice by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination.
-The role of preoperative biliary drainage is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.
+In jaundiced patients, endoscopic retrograde cholangio pancreatography (ERCP) may play a role in the alleviation of jaundice in patients with cholangitis by placing stents across the obstructed bile duct, obtaining accurate visualisation of the biliary anatomy as well as obtaining brushings from within the bile duct for cytological examination.
+However, the role routine preoperative biliary drainage is controversial and is discussed in more detail in Section \ref{sec:preoperative_biliary_drainage}.
 
 \subsection{Treatment of pancreatic cancer}
-Pancreaticoduodenectomy followed by adjuvant chemotherapy offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery. 
+Pancreaticoduodenectomy (followed by adjuvant chemotherapy in some cancers) offers the only chance of cure in patients with resectable pancreatic cancer who are fit enough to undergo surgery \parencite{li_pancreatic_2004}
 In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. 
@@
 \section{Surgical treatment of pancreatic cancer}
-Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure.% over a hundred years after its description. 
+Pancreaticoduodenectomy remains a technically challenging and complex surgical procedure. 
 Excision of cancer of the head of the pancreas was first described as a two-stage operation by a German surgeon, Walther Kausch in 1909 at Augusta-Viktoria-Krankenhaus in Berlin-Schöneberg \parencite{kausch_carcinom_1912}.
@@
 \label{sec:resectability_criteria}
-Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease. 
+Resectable pancreatic cancer is defined as a tumour that does not involve the coeliac axis or the superior mesenteric artery and is not associated with distant metastatic disease \parencite{li_pancreatic_2004}.
 Tumours involving the portal vein or superior mesenteric vein are considered borderline resectable and can still be resected completely with en-bloc venous resection in selected cases \parencite{fuhrman_rationale_1996}.
-Neoadjuvant therapies such as chemotherapy with or without local radiotherapy \parencite{gillen_preoperative/neoadjuvant_2010, evans_preoperative_2008} and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy are the subject of ongoing clinical research.
-There is increasing evidence that some of these patients may be able to undergo potentially curative surgery after neo-adjuvant treatment.
+Neoadjuvant therapies such as chemotherapy with or without local radiotherapy \parencite{gillen_preoperative/neoadjuvant_2010, evans_preoperative_2008} and newer treatment modalities such as electroporation \parencite{bower_irreversible_2011} and intra-peritoneal chemotherapy \parencite{kamath_rationale_2009} are the subject of ongoing clinical research.
+There is increasing evidence that some of these patients may be able to undergo potentially curative surgery after neo-adjuvant treatment \parencite{gillen_preoperative/neoadjuvant_2010}.
 
@@
 Prophylactic antibiotics were administered at induction. 
-The use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate.
+The use of Octreotide, a somatostatin analogue, to reduce the incidence of postoperative pancreatic fistula is still under debate \parencite{halloran_complications_2002, li_pancreatic_2004} .
 However, it was routinely used in all patients at our centre. 
@@
 This involved complete Kocherisation of the duodenum to assess the retroperitoneum. 
-Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach). 
+Both the superior mesenteric artery and coeliac axis were assessed early for tumour involvement (‘artery-first’ approach) \parencite{weitz_artery_2010}. 
 The gastrocolic omentum was divided to enter the lesser sac. 
@@
 The stomach (classical Whipple) or the first part of the duodenum (pylorus-preserving pancreaticoduodenectomy) were divided using a stapling device. 
-The pancreatic neck was transected using Harmonic scalpel, diathermy or cold blade depending on surgeon preference. 
+The pancreatic neck was transected using Harmonic\texttrademark\ scalpel, diathermy or cold blade depending on surgeon's preference. 
 The jejunum was divided at the first jejunal arcade and the jejunal mesentery was divided using the Harmonic ultrasonic scalpel. 
@@
 
-Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn sutures placed in a similar manner.
-Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn sutures.
-Gastro-jejunostomy or duodenojejunostomy (in pylorus-preserving pancreaticoduodenectomy) was performed using continuous 3/0 Polydioxanone (PDS) sutures in 2-layers. 
+Reconstruction was performed as follows: either a pancreatico-jejunostomy was fashioned using 4-0 Biosyn\texttrademark\ sutures in a two-layer duct-to-mucosa technique or a pancreatico-gastrostomy was created using 3/0 Biosyn\texttrademark\ sutures placed in a similar manner.
+Hepatico-jejunostomy was performed using interrupted 4/0 Biosyn\texttrademark\ sutures.
+Gastro-jejunostomy or duodenojejunostomy (in pylorus-preserving pancreaticoduodenectomy) was performed using continuous 3/0 Polydioxanone (PDS\texttrademark) sutures in 2-layers. 
 One or two surgical drains were placed and the abdomen was closed after ensuring haemostasis.
@@
 Post-pancreatectomy haemorrhage is reported to occur in 1 to 8\% of patients undergoing pancreaticoduodenectomy. 
-However, it accounts for 11\% to 38\% of mortality after pancreaticoduodenectomy. 
+However, it accounts for 11\% to 38\% of mortality after pancreaticoduodenectomy \parencite{wente_postpancreatectomy_2007}. 
 Post-pancreatectomy haemorrhage may either be intra-luminal into the gastrointestinal tract or intra-abdominal into the peritoneal/retro-peritoneal space. 

### 2015-08-26 Chapters/Pre_Post_SIRS.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes

@@
 \subsection{Patients}
-Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
+Patients who underwent elective pancreaticoduodenectomy between January 2008 and December 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
 Patients who underwent only a trial dissection or palliative surgical bypass for unresectable disease during this period were excluded.

### 2015-08-26 Chapters/BodyComp.tex
Body comp - Paper review complete
- All is well.....I think
- Await final comments from Donny

@@
 Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. 
-Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (p\pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (p\pageref{sec:cpx_parameters}).
-A complete dataset for one sample patient is shown in Appendix \ref{AppendixCPETRawData} on p\pageref{AppendixCPETRawData}.
+Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (page \pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (page \pageref{sec:cpx_parameters}).
+A complete dataset for one sample patient is shown in Appendix \ref{AppendixCPETRawData} on page \pageref{AppendixCPETRawData}.
 
@@
 
-The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig. \ref{fig:bc_ct_gimp} on p\pageref{fig:bc_ct_gimp}. 
+The sequence of steps followed to calculate the area of each tissue compartment is depicted in Fig. \ref{fig:bc_ct_gimp} on page \pageref{fig:bc_ct_gimp}. 
 The total cross-sectional area of the abdomen at the level of L3/L4 was calculated by first selecting all the empty space outside the image followed by inverting this selection. 
@@
 
-The proportion of skeletal muscle area decreases from 38\% in male patients with normal BMI to 22\% in males who are obese. 
+The proportion of skeletal muscle area decreased from 38\% in male patients with normal BMI to 22\% in obese males. 
 There was a greater decrease in the proportion of skeletal muscle in females with increasing BMI, with skeletal muscle area decreasing from 32\% in females with normal BMI to 14\% in obese females. 
-The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. 
+The higher weight in the overweight/obese patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. 
 Moreover, the distribution of the adipose tissue differed between males and females.
@@
 Partial correlation analysis was performed to study the relationship between pulmonary function tests and body composition. 
-It has been well-established in previous studies that pulmonary function tests are correlated with age and gender and the analysis was therefore adjusted for these two variables. 
+It has been well-established in previous studies that pulmonary function tests are correlated with age and sex and the analysis was therefore adjusted for these two variables. 
 Forced Vital Capacity (FVC, litres), Forced Expiratory Volume in 1 second (FEV1, litres) and the ratio FEV1/FVC (Tiffeneau-Pinelli index,\%) were compared against the various components of body composition.
-Both FVC and FEV1 were positively correlated with skeletal muscle area but not with adipose tissue area or total cross-sectional area. 
-FEV1/FVC was not correlated with any of the body composition components.
+Both FVC and FEV1 were positively correlated with skeletal muscle but not with adipose tissue area. 
+FEV1/FVC was not associated with any of the body composition components.
 
@@
 
-Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area to create scatter-plots(Fig.
-\ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}.
-Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig.
-\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig.
+Exercise loads achieved at anaerobic threshold and at peak exercise capacity were plotted against skeletal muscle area to create scatter-plots(Fig.
+\ref{fig:bc_load_vs_sm} on page \pageref{fig:bc_load_vs_sm}).
+Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($\rho^2 = 0.284, p < 0.001$, Fig.
+\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($\rho^2 = 0.350, p < 0.001$, Fig.
 \ref{fig:bc_pk_load_vs_sm}).
-However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on p\pageref{table:bc_cpet}.
+However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($\rho^2 = 0.004, p = 0.587$) or peak exercise ($\rho^2 = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.
 
@@
 
-Partial correlations between cardiopulmonary exercise parameters at rest, anaerobic threshold and peak exercise versus body composition were adjusted for gender. 
-Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age. 
-The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).
+Partial correlations between cardiopulmonary exercise parameters at rest, anaerobic threshold and peak exercise versus body composition were adjusted for sex. 
+Our own findings (Tables \ref{table:cpet_oj_at_regression} and \ref{table:cpet_oj_peak_regression} on pages \pageref{table:cpet_oj_at_regression}, \pageref{table:cpet_oj_peak_regression}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age.
+The results of this analysis are shown in Table \ref{table:bc_cpet} (page \pageref{table:bc_cpet}).
 
@@
 Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.353, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p = 0.001$).
-However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.
+However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and there was a non-significant change in the direction of correlation to the negative.
 
 Absolute $\dot{V}_{O_2}$ (litres/min) had no correlation with total adipose tissue at rest or at peak exercise and only a weak correlation at anaerobic threshold. 
-However, corrected for total body weight, there was a strong negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
+However, when corrected for total body weight, there was a strong negative correlation between corrected $\dot{V}_{O_2}$ (ml/kg/min)and total adipose tissue at rest ($\rho = -0.482, p<0.001$), anaerobic threshold ($\rho = -0.400, p<0.001$) and peak exercise ($\rho = -0.374, p = 0.001$).
 
-The loss of the physiological relationship between $\dot{V}_{O_2}$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig.
+The loss of physiological relationship between $\dot{V}_{O_2}$ and skeletal muscle after correcting for total body weight is shown in Fig.\ref{fig:bc_scatter_VO2_skeletal} and the creation of a spurious relationship with total adipose tissue after correction for total body weight is shown in Fig.
 \ref{fig:bc_scatter_VO2_TAT}.
@@
 Calculation of estimated lean body mass is shown in Section \ref{sec:elbm}.
-The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex (Table \ref{table:bc_at_new_indices} on p\pageref{table:bc_at_new_indices}).
+The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex (Table \ref{table:bc_at_new_indices} on page \pageref{table:bc_at_new_indices}).
 The results from Table \ref{table:bc_cpet} comparing absolute $\dot{V}_{O_2}$ and corrected $\dot{V}_{O_2}$ at the anaerobic threshold with body composition are reproduced in the first two columns of this table to allow comparison with the new indices.
@@
 The loss of the strong positive correlation between absolute $VO_{2}$ (litres/min) and skeletal muscle area after correcting for total body weight further supports the argument that the corrected value for $\dot{V}_{O_2}$ under-reports aerobic capacity in obese patients.
-There was no correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area as well as the slight.
-However, a statistically significant, positive correlation was present between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest appear.
+There was no correlation between pulmonary function tests, tidal volume and minute ventilation and adipose tissue area. 
+However, a statistically significant, positive correlation was present between adipose tissue and $O_2$Pulse at anaerobic threshold as well as peak exercise but not at rest.
 These findings suggest that adiposity did not contribute to poor cardiopulmonary exercise performance in this cohort of patients.
@@
 
-Seltzer reported in his 1940 study of 34 subjects, that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight \parencite{seltzer_body_1940}.
+Seltzer reported in his 1940 study of 34 subjects that individuals who were more `lateral' than `linear' had lower oxygen intakes per kilo body weight \parencite{seltzer_body_1940}.
 Tanner in his article titled \textit{`Fallacy of per-weight and per-surface area standards, and their relation to spurious correlation'} in the Journal of Applied Physiology in 1947 recognised the dangers of expressing physiological variables as a function of total body mass \parencite{tanner_fallacy_1949}.
@@
 In a study aimed at determining the optimal method of expressing $\dot{V}_{O_2}$Max, Maciejczyk and co-workers analysed the differing influence of body fat and lean body mass on aerobic performance in two groups of physically fit men categorised based on their body fat percentage \parencite{maciejczyk_influence_2014}.
-High body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ (ml/kg/min) when it was corrected for total body weight penalising otherwise fit men purely based on the proportion of body weight that was contributed by body fat.
+High body mass regardless of composition was correlated negatively with $\dot{V}_{O_2}$ (ml/kg/min) when it was corrected for total body weight.
+This penalised otherwise fit men purely based on the proportion of body weight that was contributed by body fat. 
 However, when $\dot{V}_{O_2}$ was corrected for lean body mass, they found that the results were similar between the low body fat and high fat body groups. 
@@
 
-The term \textit{`spurious correlation'} was first introduced by English mathematician and biometrician, Karl Pearson in 1896. 
+The term \textit{`spurious correlation'} was first introduced by English mathematician and biometrician, Karl Pearson, in 1896. 
 He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied \parencite{pearson_mathematical_1896}.
@@
 While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome and that a low $\dot{V}_{O_2}$AT represented cardiac failure.
-This association is repeated in their later work on 548 patients which also showed a clear association between low $\dot{V}_{O_2}$AT (ml/kg/min) and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
+This assertion is repeated in their later work on 548 patients which also showed a clear association between low $\dot{V}_{O_2}$AT (ml/kg/min) and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
 The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.
 
-Swart and Carlisle als reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. 
+Swart and Carlisle als0 reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. 
 However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT group (24\% vs 51\%). 
@@
 The `obesity paradox' would suggest that, in fact, obese patients are not necessarily at an increased risk of postoperative complications. 
-Several authors have reported that obesity is not a risk factor for major complications after pancreaticoduodenectomy \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}. 
+Several authors have reported that obesity was not a risk factor for major complications after pancreaticoduodenectomy \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}. 
 In a large study of 118,707 patients undergoing non-bariatric general surgery, the authors reported that the risk of death was `paradoxically' low in the overweight and obese group of patients \parencite{mullen_obesity_2009}.
@@
 However, the results presented in this study demonstrate that the success of such prehabilitation programs must not be determined solely by $\dot{V}_{O_2}$ corrected for total body weight. 
-Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O^2$Pulse, tidal volume \parencite{jones_effects_2007} or maximal exercise load \parencite{kim_responsive_2009} may provide more reliable evidence of improvement in aerobic capacity.
+Instead, improvement in the absolute values of $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak in conjunction with improvement in other parameters that are not affected by body composition such as $O_2$Pulse, tidal volume \parencite{jones_effects_2007} or maximal exercise load \parencite{kim_responsive_2009} may provide more reliable evidence of improvement in aerobic capacity.
 
@@
 Such preoperative interventions may also be associated with improvements in body composition with increase in skeletal muscle mass and reduction in adipose tissue.
-This may result in apparent improvements in cardiopulmonary exercise test parameters that are scaled using total body weight.
+This may result in apparent improvements in cardiopulmonary exercise test parameters that are scaled using total body weight as well as improvements in true aerobic fitness.
 Further studies are required to validate the findings presented here as well as to identify optimal scaling methods for cardiopulmonary exercise test parameters. 

### 2015-08-26 Chapters/Pre_Post_SIRS.tex
Pre_Post_SIRS - Paper review complete
- Look at the number of bloody mistakes

@@
 
-An elevated preoperative systemic inflammatory response was associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
+Elevated preoperative systemic inflammation was associated with increased complications after colorectal surgery \parencite{moyes_preoperative_2009, kubo_elevated_2013}, oesophagectomy \parencite{vashist_glasgow_2010} as well as liver surgery for colorectal metastases \parencite{neal_preoperative_2011}.
 The modified Glasgow Prognostic Score in particular, which uses the combination of C-reactive protein and serum albumin, has been reported to be associated with increased incidence of complications \parencite{moyes_preoperative_2009, mohri_correlation_2014, vashist_glasgow_2010}.
@@
 
-Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery is associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures \parencite{singh_systematic_2014, platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery \parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
+Moreover, an elevated postoperative systemic inflammatory response in the first few days after surgery was associated with increased incidence of infective complications after a wide variety of thoraco-abdominal procedures \parencite{singh_systematic_2014, platt_c-reactive_2012, dutta_persistent_2011, welsch_persisting_2008} as well as other types of surgery \parencite{mcneer_early_2010, laporta_baez_c-reactive_2011}.
 %Include some recent reviews pp singh, bjs as well as check Donny's email.
@@
 These include the presence of obstructive jaundice with or without cholangitis, preoperative biliary intervention including endoscopic retrograde cholangio-pancreatography for diagnosis or biliary drainage and in some patients acute or chronic pancreatitis either due to obstruction of the main pancreatic duct or due to other causes.
-The effect this `priming' of the immune system on postoperative outcomes is poorly understood in this cohort of patients.
+The effect of this `priming' of the immune system on postoperative outcomes is poorly understood in this cohort of patients.
 
-Chronic inflammation is a recognised feature of obesity which is increasingly common in patients undergoing major surgery for pancreatic and other gastro-intestinal cancers. 
+Chronic inflammation is also a recognised feature of obesity which is increasingly common in patients undergoing major surgery for pancreatic and other gastro-intestinal cancers. 
 The impact of obesity, especially visceral obesity, on complications after pancreaticoduodenectomy remains controversial with some authors reporting that obesity was associated with increased incidence of complications \parencite{house_preoperative_2008, ramsey_body_2011} while others reporting similar outcomes in obese and non-obese patients \parencite{khan_does_2010, tsai_impact_2010, balentine_obesity_2011}.
@@
 More recently, levels of adipocytokines, inflammatory mediators produced exclusively in adipose tissue, have been reported to be associated with postoperative surgical site infections after colorectal \parencite{ortega-deballon_preoperative_2013, matsuda_preoperative_2009} and gastric cancer surgery \parencite{yamamoto_association_2013}.
-Sarcopenia has been reported to be associated with  elevated postoperative systemic inflammation after colorectal surgery \parencite{reisinger_sarcopenia_2015}.
+Sarcopenia has also been reported to be associated with  elevated postoperative systemic inflammation after colorectal surgery \parencite{reisinger_sarcopenia_2015}.
 This emphasises the importance of adipose tissue metabolism and body composition in the preoperative systemic inflammatory status of surgical patients.
@@
 
-The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on p\pageref{table:mGPS} in Chapter \ref{ch_intro}. 
+The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on page \pageref{table:mGPS} in Chapter \ref{ch_intro}. 
 The neutrophil-lymphocyte ratio was calculated by dividing the preoperative neutrophil count by the preoperative lymphocyte count and a threshold of 5.0 was used to dichotomise this variable. 
@@
 \subsection{BMI and body composition}
-Body Mass Index (BMI) was categorised using the World Health Organisation thresholds as shown in Table \ref{table:bmi_who} on p\ref{table:bmi_who} in Chapter \ref{ch_intro}. 
+Body Mass Index (BMI) was categorised using the World Health Organisation thresholds as shown in Table \ref{table:bmi_who} on p \pageref{table:bmi_who} in Chapter \ref{ch_intro}. 
 Only one patient had a BMI less than 18.5 kg/m$2$ and was excluded from analysis involving BMI.
 Body composition was calculated in a subset of these patients using preoperative computed tomography of the abdomen. 
-The methodology used in the calculation of the individual components of body composition including visceral fat, subcutaneous fat and skeletal muscle is described in detail in Section \ref{sec:bodycomp_calculation}.
+The methodology used in the calculation of the individual components of body composition including visceral fat, subcutaneous fat and skeletal muscle is described in detail in Section \ref{sec:bodycomp_calculation} on page \pageref{sec:bodycomp_calculation}.
 Continuous data were converted into categorical data using tertiles.
@@
 The Scottish Index of Multiple Deprivation (SIMD) Quintile Scores were calculated from the post-code of the patient's primary residence and dichotomised into two groups with scores of 1-3 and 4-5 respectively.
-The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
+Lower scores indicated greater deprivation.
+The POSSUM Physiology Score was calculated based on 11 physiological parameters (cardiac disease including hypertension, ischaemic heart disease and heart failure, respiratory disease causing breathlessness on exertion and COPD, ECG changes, pulse rate, blood pressure, haemoglobin, white cell count, serum sodium, serum potassium, serum urea and Glasgow Coma Scale) as described in Table \ref{table:intro_possum} on page \pageref{table:intro_possum}.
 
@@
 $\dot{V}_{O_2}$AT was dichotomised using a value of 10 ml/kg/min while $\dot{V}_{O_2}$Peak was dichotomised using a value of 16 ml/kg/min. 
-Cardiopulmonary exercise testing methodology has been described in Section \ref{sec:cpx_method}.
+Cardiopulmonary exercise testing methodology has been described in Section \ref{sec:cpx_method} on page \pageref{sec:cpx_method}.
 
@@
 Preoperative C-reactive protein was elevated in 70 (37.6\%) patients while just over half the patients had a low preoperative serum albumin (96, 51.1\%). 
-The modified Glasgow Prognostic Score revealed normal preoperative systemic inflammatory status (mGPS = 0) in 116 (62.4\%) patients, mildly systemic inflammation in 16 (8.6\%) and severe inflammation in 54 (29.0\%) patients.
-Obstructive jaundice was present in 44 (23.4\%) patients and severe obstructive jaundice was present in 31 (16.5\%) of patients. 
+The modified Glasgow Prognostic Score revealed normal preoperative systemic inflammatory status (mGPS = 0) in 116 (62.4\%) patients, mild systemic inflammation in 16 (8.6\%) and severe inflammation in 54 (29.0\%) patients.
+Obstructive jaundice was present in 44 (23.4\%) patients and severe obstructive jaundice was present in 31 (16.5\%) patients. 
 BMI data was available in 167 patients. 
@@
 Preoperative hypoalbuminemia on the other hand was associated with lower median postoperative CRP levels starting with POD 3. 
-The median difference in postoperative CRP increased from 30 mg/l on POD 3 (p$<$0.045) to 53 mg/l by POD 6 (p=0.002) with patients with hypoalbuminemia having a lower CRP. 
+The median difference in postoperative CRP between patients with or without preoperative hypoalbuminemia was 30 mg/l on POD 3 (p$<$0.045) and 53 mg/l on POD 6 (p=0.002) with lower CRP levels in patients with hypoalbuminemia.
 This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
@@
 
-The trend in postoperative CRP appeared to differ depending on whether preoperative CRP and albumin were independently deranged or not.
-Fig. \ref{fig:sirs_crp_crp_alb} and Table \ref{table:sirs_crp_with_alb} show the results of this analysis.
+The trend in postoperative CRP appeared to differ depending on whether preoperative CRP and albumin were independently deranged.
+Fig. \ref{fig:sirs_crp_crp_alb} and Table \ref{table:sirs_crp_with_alb} on page \pageref{fig:sirs_crp_crp_alb} show the results of this analysis.
 When both preoperative CRP and preoperative albumin were normal, the trend in postoperative CRP is shown by the solid line (Line 0 in Fig. \ref{fig:sirs_crp_crp_alb}). 
@@
 Obstructive jaundice was associated with significant differences in the postoperative systemic inflammatory response. 
-Postoperative serum albumin levels were lower in jaundiced patients and with the severely jaundiced patients having the lowest levels (p$<$0.001, Fig. \ref{fig:sirs_bil_alb}). 
+Postoperative serum albumin levels were lower in jaundiced patients with severely jaundiced patients having the lowest levels (p$<$0.001, Fig. \ref{fig:sirs_bil_alb}). 
 There were no significant changes in the trends of albumin levels and it is likely that preoperative hypoalbuminemia persisted postoperatively in these patients.
@@
 Median postoperative CRP had an inverse relationship with severity of preoperative obstructive jaundice (Fig. \ref{fig:sirs_bil_crp}).
-Obstructive jaundice was also associated with a delayed rise in the neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although this relationship was less significant (Fig.
-\ref{fig:sirs_bil_neut}).
+Obstructive jaundice was associated with a delayed rise in the neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although this relationship was less significant (Fig. \ref{fig:sirs_bil_neut}).
 
@@
 There was no significant, persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
-However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Table \ref{table:sirs_alb}, \ref{table:sirs_alb_pvalues}).
+However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Tables \ref{table:sirs_alb} and \ref{table:sirs_alb_pvalues}).
 
@@
 Body composition data was available for 90 patients.
-Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, 128.9 cm$^2$, 119.0-150.8)and high (n=30, 241.2 cm$^2$, 221.9-313.1). 
+Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, 128.9 cm$^2$, 119.0-150.8) and high (n=30, 241.2 cm$^2$, 221.9-313.1).
 Subcutaneous adipose tissue area was divided into tertiles as low (n=30, median 94.1 cm$^2$, IQR 63.2-102.5), moderate (n=30, median 152.5 cm$^2$, IQR 141.3-175.3) and high (n=30, median 253.9 cm$^2$, IQR 203.2-290.3).
@@
 Postoperative CRP and neutrophil count were not related to any of the components of body composition aside from a weak linear association between neutrophil count on POD 2 and visceral fat area (p=0.046).
-However, lower preoperative skeletal muscle area was associated with lower postoperative serum albumin on days 1 to 5 (all p<0.05).
+However, lower preoperative skeletal muscle area was associated with lower postoperative serum albumin on days 1 to 5 (all p$<$0.05).
 
@@
 		\includegraphics[width=\textwidth]{Figures/sirs_crp_neut}
-		\caption{Preop. CRP vs. postop. Neutrophil Count}
+		\caption{Preop. CRP vs. postop. Neutrophil count}
 		\label{fig:sirs_crp_neut}
@@
 		\includegraphics[width=\textwidth]{Figures/sirs_alb_neut}
-		\caption{Preop. Albumin vs. postop. Neutrophil Count}
+		\caption{Preop. Albumin vs. postop. Neutrophil count}
 		\label{fig:sirs_alb_neut}
@@
 		\includegraphics[width=\textwidth]{Figures/sirs_bil_neut}
-		\caption{Preop. Bilirubin vs. postop. Neutrophil Count}
+		\caption{Preop. Bilirubin vs. postop. Neutrophil count}
 		\label{fig:sirs_bil_neut}
@@
 		\includegraphics[width=\textwidth]{Figures/sirs_at_neut}
-		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Neutrophil Count}
+		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Neutrophil count}
 		\label{fig:sirs_at_neut}
@@
 A recent meta-analysis of six studies involving 1832 patients reported that CRP level less than 135 mg/l on the fourth postoperative day had a high negative predictive value of 89\% for infectious complications \parencite{warschkow_safe_2012}.
-Postoperative CRP has been shown to predict complications after pancreaticoduodenectomy by us (\ref{ch_crp_comp}) as well as other authors \parencite{welsch_persisting_2008, hiyoshi_usefulness_2013, kosaka_multivariate_2014}.
+Postoperative CRP has been shown to predict complications after pancreaticoduodenectomy by us (chapter \ref{ch_crp_comp}) as well as other authors \parencite{welsch_persisting_2008, hiyoshi_usefulness_2013, kosaka_multivariate_2014}.
 The role of preoperative systemic inflammation as measured by the modified Glasgow Prognostic Score or the neutrophil-lymphocyte ratio has also been reported to be adversely affect long-term survival in patients undergoing potentially curative surgery in over a hundred studies involving a variety of gastrointestinal and non-gastrointestinal cancers.
@@
 This may affect the healing processes as well as the patient's ability to clear endotoxins and may predispose the patient to infective and other complications.
-Our results appear to support this hypothesis.
+The results of the present study appear to support this hypothesis.
 
@@
 However, preoperative hypoalbuminemia was associated with significantly lower levels of postoperative CRP.
-These apparently opposite effects of preoperative CRP and albumin on the postoperative systemic inflammatory response is further evidenced by the fact that the effect of preoperative CRP on postoperative inflammation is attenuated by the presence of preoperative hypoalbuminemia.
-These results appear to suggest when hypoalbuminemia is present as part of preoperative systemic inflammation, not only is the exaggerated postoperative response lost but the response appears to be dampened.
+These apparently opposite effects of preoperative CRP and albumin on the postoperative systemic inflammatory response is further evidenced by the fact that the effect of elevated preoperative CRP on postoperative inflammation is attenuated by the presence of preoperative hypoalbuminemia.
+These results appear to suggest when hypoalbuminemia is present as part of preoperative systemic inflammation, not only is the exaggerated postoperative response due to elevated CRP lost but the response appears to be dampened.
 
@@
 Moreover, both low albumin and elevated CRP were significantly associated with cutaneous anergy.
-Cutaneous anergy is one of the hallmarks of the compensatory anti-inflammatory response (CARS) \parencite{ward_compensatory_2008}.
+Cutaneous anergy is one of the hallmarks of the compensatory anti-inflammatory response syndrome (CARS) \parencite{ward_compensatory_2008}.
 
@@
 In patients with pancreatic cancer, malignant obstructive jaundice further complicates the preoperative inflammatory status and the behaviour of the immune system \parencite{nehez_compromise_2002}. 
-We reported in Chapter \ref{ch_cpet_jaundice} that obstructive jaundice was associated with elevated preoperative CRP and low serum albumin on the day before surgery with a linear relationship with severity of jaundice (Table \ref{table:cpet_oj_bloods} on p\pageref{table:cpet_oj_bloods}).
+We reported in Chapter \ref{ch_cpet_jaundice} that obstructive jaundice was associated with elevated preoperative CRP and low serum albumin on the day before surgery with a linear relationship with severity of jaundice (Table \ref{table:cpet_oj_bloods} on page \pageref{table:cpet_oj_bloods}).
 Hypoalbuminemia in jaundiced patients persisted after surgery in the present study.
 However, the CRP response appears to have reversed in jaundiced patients.
-While preoperative CRP was elevated in jaundiced patients, CRP levels after surgery were significantly lower in jaundiced patients with a negative linear relationship with severity of jaundice.
+While preoperative CRP was elevated in jaundiced patients, CRP levels after surgery were significantly lower in jaundiced patients with an inverse relationship with severity of jaundice.
 
@@
 Immunosuppression and immune dysregulation have been recognised in patients with obstructive jaundice \parencite{scott-conner_pathophysiology_1994}.
-Our results are similar to those reported by Mackenzie and Woodhouse who studied the CRP response to bacteraemia in 126 patients critically ill patients with or without liver dysfunction \parencite{mackenzie_c-reactive_2006}.
+Our results are similar to those reported by Mackenzie and Woodhouse who studied the CRP response to bacteraemia in 126 critically ill patients with or without liver dysfunction \parencite{mackenzie_c-reactive_2006}.
 Liver dysfunction was defined as serum bilirubin $>$ 20 $\mu$mol/l and prothrombin time longer than 18 seconds. 
-The CRP response to bacteraemia was significantly lower in patients with liver dysfunction than in those with normal liver function (146 vs. 103 mg/l, p=0.03). 
+The CRP response to bacteraemia was significantly lower in patients with liver dysfunction than in those with normal liver function (103 vs. 146 mg/l, p=0.03). 
 These findings have been corroborated by other authors in patients with cirrhosis \parencite{pieri_c-reactive_2014, janum_c-reactive_2011}.
-
-%There appears to be some evidence from animal studies that the immunosuppression due to obstructive jaundice may be a therapeutic target \parencite{padillo_anti-tnf-alpha_2007}.
 
@@
 Although CRP on POD 0 was significantly different between patients with a normal $\dot{V}_{O_2}$AT/$\dot{V}_{O_2}$Peak and those with low aerobic capacity, this difference was not present on other days.
-The lack of association between aerobic capacity, POSSUM Physiology Score and postoperative CRP in this study may have been due to the dominant effect of preoperative systemic inflammation and obstructive jaundice on the postoperative inflammatory response or simply due to any influence of aerobic capacity on postoperative inflammation.
+The lack of association between aerobic capacity, POSSUM Physiology Score and postoperative CRP in this study may have been due to the dominant effect of preoperative systemic inflammation and obstructive jaundice on the postoperative inflammatory response.
 However, the association of postoperative hypoalbuminemia with low aerobic capacity, POSSUM Physiology Score and anaemia appears to suggest a link between comorbidity and systemic inflammation and/or nutritional status.
@@
 The association between hypoalbuminemia and low aerobic capacity may be explained by lower skeletal muscle and greater subcutaneous fat in patients with persistent postoperative hypoalbuminemia.
-Alternate explanations 
-Richards and co-workers reported that elevated preoperative systemic inflammation was related to low skeletal muscle index \parencite{hassen_preoperative_2007}.
+Richards and co-workers reported that elevated preoperative systemic inflammation was related to low skeletal muscle index \parencite{richards_relationships_2012}.
 Hassen and co-workers reported that fat-free mass and skeletal muscle were inversely related to the severity of the postoperative inflammatory response following endo-vascular surgery for abdominal aortic aneurysm \parencite{hassen_preoperative_2007}.
-Elevated systemic inflammation is related to both obesity as well as sarcopenia. 
-`Sarcopenic-obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival in patients with pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012}.
+
+Elevated systemic inflammation is related to obesity as well as sarcopenia. 
+`Sarcopenic-obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival \parencite{tan_sarcopenia_2009, peng_impact_2012}.
 

### 2015-08-27 Chapters/CRP_Complications.tex
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes
CRP Complications -  Paper review complete

@@
 Pancreaticoduodenectomy is associated with significant morbidity in spite of advances in patient selection, perioperative care and surgical technique. 
-Early identification of complications can help improve outcomes by better allocation of critical care resources as well as goal-directed therapy. 
+Early identification of complications can help improve outcomes by better allocation of critical care resources and aggressive early-rescue strategies \parencite{ferraris_va_identification_2014}. 
 Postoperative pancreatic fistula is one of the most dreaded complications after a pancreaticoduodenectomy and can lead to a cascade of other complications including delayed haemorrhage, infected intra-abdominal collections, delayed gastric emptying, prolonged hospitalisation and in some cases, death.
-The International Study Group for Pancreatic Fistula (ISGPF) have not only defined what constitutes a post-operative pancreatic fistula but have also graded the severity of this complication based on its impact on the management of the patient.
-However, these definitions are applied after the event and there is no clear way of predicting the severity of complications.
+The International Study Group for Pancreatic Fistula (ISGPF) have not only defined what constitutes a post-operative pancreatic fistula but have also graded the severity of this complication based on its impact on the management of the patient \parencite{bassi_postoperative_2005}.
+However, these definitions are applied after the event and there are no clear strategies to predict the severity of complications.
 
 Postoperative CRP levels have been shown to predict infective complications after major surgery including colorectal,  oesophago-gastric and pancreatic surgery \parencite{mustard_c-reactive_1987, matthiessen_increase_2008, mcneer_early_2010, dutta_persistent_2011, mackay_c-reactive_2011, murakami_soft_2008, welsch_persisting_2008}.
-It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing.
+It has been postulated that an unmitigated and exaggerated systemic inflammatory response in the early postoperative period is followed by a compensatory anti-inflammatory response that predisposes the patient to sepsis and impaired healing \parencite{bone_sepsis:_1997, bone_immunologic_1996}.
 
-But, the mechanisms underlying the pathogenesis of post-operative pancreatic fistula are largely attributed to anatomical factors including soft texture of the pancreas and small pancreatic duct diameter as well as surgical technique with increased intra-operative blood loss and pancreatico-jejunostomy rather than pancreatico-gastrostomy being associated with increased incidence of anastomotic leakage.
-The association between the early postoperative systemic inflammatory response and the incidence and severity of postoperative pancreatic fistula has not been reported. 
+However, the mechanisms underlying the pathogenesis of post-operative pancreatic fistula are largely attributed to anatomical factors including soft texture of the pancreas and small pancreatic duct diameter as well as surgical technique with increased intra-operative blood loss and pancreatico-jejunostomy rather than pancreatico-gastrostomy being associated with increased incidence of anastomotic leakage \parencite{pratt_risk_2008, xiong_meta-analysis_2014}.
+The association between the early postoperative systemic inflammatory response and the severity of postoperative pancreatic fistula has not been reported. 
 Moreover, the role postoperative CRP in predicting infective complications when these occur in association with postoperative pancreatic fistula has not been studied either.
@@
 \clearpage
-\section{Methods}
+\section{Patients and methods}
 
-Patients who underwent elective pancreaticoduodenectomy between January 2008 and July 2012 were included in this study. 
-Patients who underwent only a trial dissection or trial dissection and palliative surgical bypass during this period were excluded. 
+\subsection{Patients}
+Patients who underwent elective pancreaticoduodenectomy between January 2008 and December 2012 at the West of Scotland Pancreatic Unit at the Glasgow Royal Infirmary were included in this study. 
+Patients who underwent only a trial dissection or palliative surgical bypass for unresectable disease during this period were excluded.
+
+\subsection{Perioperative care}
 All patients were given antibiotic prophylaxis at induction and this was continued for 24 hours after surgery. 
-All patients had general anaesthesia supplemented either with patient-controlled epidural analgesia or spinal diamorphine and patient-controlled opiate analgesia for postoperative pain relief. 
+All patients had general anaesthesia supplemented either with epidural analgesia or spinal diamorphine and patient-controlled opiate analgesia for postoperative pain relief. 
 Octreotide (200$\mu$g) was administered subcutaneously three times a day for 5 days in all patients and continued longer in patients diagnosed with a postoperative pancreatic fistula. 
@@
 All patients had one or two surgical drain(s) placed at the time of surgery which were removed at the clinician's discretion during the postoperative period based on the presence or absence of postoperative pancreatic fistula or other intra-abdominal collections.
+Further details regarding operative technique and postoperative care are provided in Chapter \ref{ch_intro}.
+
+\subsection{Biochemical parameters}
 
 Patients had routine measurements of inflammatory markers including C-reactive protein (CRP) and differential white cell counts from the day before surgery and every day for at least a week after surgery.
-These results including the preoperative blood tests and the postoperative blood tests for the first 7 days after surgery were collected from the hospital laboratory databases.
+These results were collected from the hospital laboratory database using an automated MS Access application as outlined in Appendix \ref{AppendixAccessDatabase}.
 
+\subsection{Complications and severity grading}
 All complications were discussed at a weekly morbidity meeting and prospectively recorded in an electronic database.
-The diagnosis and grading of pancreas-specific complications including postoperative pancreatic fistula (POPF, Section \ref{sec:ch_intro_POPF}) and post-pancreatectomy haemorrhage (PPH, Section \ref{sec:ch_intro_PPH}) were made according to the International Study Group classifications as shown in Table \ref{table:isgps_popf} and Table \ref{table:isgps_pph} on p\pageref{table:isgps_popf} respectively.
+The diagnosis and grading of pancreas-specific complications including postoperative pancreatic fistula (POPF, Section \ref{sec:ch_intro_POPF}) were made according to the International Study Group classifications \parencite{bassi_postoperative_2005} as shown in Table \ref{table:isgps_popf} on page \pageref{table:isgps_popf}.
+A postoperative pancreatic fistula was defined as drain output of any measurable quantity after the third postoperative day with amylase content greater than three times the upper limit of the normal serum amylase value at our laboratory.
 
-All other complications were graded using the Clavien-Dindo classification system as shown in Table \ref{table:clavien-dindo}. 
+All other complications were graded using the Clavien-Dindo classification system \parencite{clavien_clavien-dindo_2009, dindo_classification_2004} as shown in Table \ref{table:clavien-dindo}. 
 This allowed grading of the severity of complications based on the magnitude of intervention(s) required to treat them. 
@@
 \subsection{Statistics}
+Continuous data are presented as median (inter-quartile range, IQR) unless otherwise specified. 
 Mann-Whitney U test was used to compare the distribution of postoperative CRP (as a continuous variable) in patients who had a complication and those who did not. 
-Data are presented as median (inter-quartile range, IQR) unless otherwise specified. 
 Line plots with error bars depicting 95\% confidence intervals were used to depict the trends in CRP over time across different patient groups.
@@
 Significant infective complications of Clavien-Dindo grade 3 or higher occurred in 84 patients. 
-Infectious complications were more common in patients with age $>$65 years (35.6\% vs.
-50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 ml/kg/min (28.8\% vs.
-52.4\%, p=0.006) were significantly associated with infective complications.
-Gender, body mass index, smoking status, modified Glasgow Prognostic Score, preoperative serum bilirubin and POSSUM physiology score were not associated with infective complications.
+Infective complications were more common in patients with age $>$65 years (35.6\% vs. 50.0\%, p=0.046) and $\dot{V}_{O_2}$AT$<$10 ml/kg/min (28.8\% vs. 52.4\%, p=0.006).
+Sex, body mass index, smoking status, modified Glasgow Prognostic Score, preoperative serum bilirubin and POSSUM physiology score were not associated with infective complications.
 
-Infectious complications were also significantly associated with other complications including post-operative pancreatic fistula, post-pancreatectomy haemorrhage, length of stay in hospital, critical care stay, number of critical care admissions, re-operation rates and in-hospital mortality.
-(Table \ref{table:crp_comp_infect_vs_other_complications})
+Infective complications were also significantly associated with other complications including post-operative pancreatic fistula, post-pancreatectomy haemorrhage as well as length of stay in hospital, critical care stay, number of critical care admissions, re-operation rates and in-hospital mortality (Table \ref{table:crp_comp_infect_vs_other_complications}).
 
@@
 \label{Postoperative CRP vs. POPF}
-Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p<0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
-These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig. \ref{fig:crp_comp_crp_popf} on p\pageref{fig:crp_comp_crp_popf}. 
+Postoperative CRP levels on days 2 through 7 were significantly higher in patients who developed a postoperative pancreatic fistula (p = 0.002 for CRP on Day 2 and p$<$0.001 for CRP on Days 3 through 7, Mann-Whitney U test).
+These results are presented in Table \ref{table:crp_comp_vs_POPF_ISGPS_p_values_only} and Fig. \ref{fig:crp_comp_crp_popf} on page \pageref{fig:crp_comp_crp_popf}. 
 Fig. \ref{fig:crp_comp_crp_popf_isgps} shows that CRP in the first postoperative week was not significantly different between postoperative pancreatic fistulae of ISGPF Grade A, B and C.
 
-However, there is no useful clinical application for the association between CRP in the early postoperative period and  pancreatic fistula as the diagnosis of POPF is based on the the amylase content in surgical drains on the third postoperative day.
+However, there is no useful clinical application for the association between CRP in the early postoperative period and  pancreatic fistula as the diagnosis of POPF is based on the amylase content in surgical drains on the third postoperative day.
 Moreover, early postoperative CRP did not predict the severity of the pancreatic fistula either.
@@
 Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
-However, there was no such association in patients who did develop a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
+However, there was no such association in patients who developed a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
 
-For instance, difference in median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
+For instance, the median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
 (p$<$0.001).
-However, in patients who had POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
+However, in patients who had a POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
 (p = 0.559)
@@
 Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_infective_leak} on p\pageref{fig:crp_comp_infective_leak}.
+\ref{fig:crp_comp_infective_leak} on page \pageref{fig:crp_comp_infective_leak}.
 
@@
 
-Patients who developed infective complications in the absence of POPF had a raised total white cell counts on postoperative days 5, 6 and 7 which were statistically significant.
+Patients who developed infective complications in the absence of POPF had a raised total white cell count on postoperative days 5, 6 and 7 which were statistically significant.
 However, these differences were small with wide, overlapping 95\% confidence intervals (Fig. \ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant.
+
 For instance, the median white cell count on the fifth postoperative day in patients who did not have an infective complication was 8.9 (IQR 6.6 - 11) while in patients who had an infective complication it was 9.8 (IQR 7.9 - 13).
-
 While this difference was statistically significant (p = 0.028), both medians were within the normal limits for healthy adults. 
@@
 (Table \ref{table:crp_comp_WCC_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_WCC_infective_leak} on p\pageref{fig:crp_comp_WCC_infective_leak})
+\ref{fig:crp_comp_WCC_infective_leak} on page \pageref{fig:crp_comp_WCC_infective_leak})
 
-Similar statistically significant differences were noted in the postoperative neutrophil counts on postoperative days 5, 6 and 7 in patients without a POPF while there was no difference in patients with a POPF.
-(Table \ref{table:crp_comp_Neutrophil_vs_infections_popf_y1n0})	However, these differences were associated with a large overlap between the 95\% confidence intervals (error bars in Fig.
-\ref{fig:crp_comp_Neutrophil_infective_leak0}).
+Similar statistically significant differences were noted in the postoperative neutrophil counts on postoperative days 5, 6 and 7 in patients without a POPF while there was no difference in patients with a POPF (Table \ref{table:crp_comp_Neutrophil_vs_infections_popf_y1n0}).
+However, these differences were associated with a large overlap between the 95\% confidence intervals (error bars in Fig. \ref{fig:crp_comp_Neutrophil_infective_leak0}).
 
@@
 (Table \ref{table:crp_comp_Albumin_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_Albumin_infective_leak} on p\pageref{fig:crp_comp_Albumin_infective_leak})
+\ref{fig:crp_comp_Albumin_infective_leak} on page \pageref{fig:crp_comp_Albumin_infective_leak})
 
@@
 \subsection{Receiver operating characteristic (ROC) analysis}
-Receiver operating characteristic (ROC) analysis was undertaken in order to find the optimum threshold for the relationship between CRP and infective complications in patients without POPF. 
-ROC curves were plotted with CR as a continuous variable and infective complications as a dichotomous outcome variable for postoperative days 3 through 7.
+Receiver operating characteristic (ROC) analysis was undertaken to find the optimum threshold for the relationship between CRP and infective complications in patients without POPF. 
+ROC curves were plotted with CRP as a continuous variable and infective complication as a dichotomous outcome variable for postoperative days 3 through 7.
 
@@
 The optimum thresholds along with the corresponding sensitivity, specificity, negative predictive value, positive predictive value, 95\% confidence intervals (CI) and p-values are shown in Table \ref{table:crp_comp_ROC_infections_noPOPF} and the corresponding ROC curves are shown in Fig.
-\ref{fig:crp_comp_ROC_infection} on p\pageref{fig:crp_comp_ROC_infection}.
+\ref{fig:crp_comp_ROC_infection} on page \pageref{fig:crp_comp_ROC_infection}.
 The thresholds were identified with a bias towards a higher negative predictive value to allow identification of patients who did not develop an infective complication.
@@
 \begin{figure}[t]
-	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications.}
+	\caption{Receiver operating characteristics analysis of C-reactive protein as a marker of postoperative infective complications.}
 	\label{fig:crp_comp_ROC_infection}
@@
 This can lead to localised collections that can become secondarily infected, result in erosion of blood vessels resulting in post-pancreatectomy haemorrhage, result in delayed gastric emptying or postoperative ileus as well as respiratory complications including sympathetic pleural effusions or pneumonic consolidation.
-In fact, most large series report a close association between POPF and PPH with the latter often following the former. 
+In fact, most large series report a close association between POPF and PPH with the latter often following the former \parencite{choi_delayed_2004, koukoutsis_haemorrhage_2006, tien_risk_2005}. 
 The eventual outcome in patients with POPF is often determined by the cascade of complications associated with the POPF itself. 
@@
 They compared 91 patients who developed an 'inflammatory postoperative complication' with a random subset of 60 consecutive pancreatic resections. 
-They reported that a threshold of 140 mg/L serum CRP level on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%).
+They reported that a threshold of 140 mg/L for serum CRP on the fourth postoperative day had the best diagnostic accuracy for the prediction of inflammatory postoperative complications (sensitivity 69.5\%, specificity 87.1\%, positive predictive value 48.7\% for an estimated prevalence of 15\%).
 However, the definitions used for POPF in this study were different from the ISGPF definition and the authors did not report on the predictive ability of CRP in the presence or absence of a POPF. 
@@
 The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication. 
-This finding is similar to one of the earliest reports on the role serial CRP measurements in predicting septic complications after surgery \parencite{mustard_c-reactive_1987} as well as in several subsequent reports \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011}.
+This finding is similar to one of the earliest reports on the role of serial CRP measurements in predicting septic complications after surgery \parencite{mustard_c-reactive_1987} as well as in several subsequent reports \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011}.
 The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation. 
 This study was not designed to look at inflammatory markers later in the postoperative period. 
-However, Mustard and co-workers monitored CRP, white cell count and temperature for 14 days after surgery and found that CRP was the only reliable marker of septic complications.
+However, Mustard and co-workers monitored CRP, white cell count and temperature for 14 days after surgery and found that CRP was the only reliable marker of septic complications \parencite{mustard_c-reactive_1987}.
 
@@
 More recently, the compensatory anti-inflammatory response with the associated immunosuppressive effects has been recognised as a possible factor in predisposing the patient to infective complications and delayed wound healing.
-Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications.
+Such an anti-inflammatory response is proportional to the initial systemic inflammatory response and therefore patients with an exaggerated inflammatory response in the immediate aftermath of surgery may be at increased risk of developing complications \parencite{ward_compensatory_2008}.
 
-Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer. 
-They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day \parencite{mokart_early_2002}.
+Mokart and co-workers studied trends in inflammatory cytokine levels in the serum of 30 patients undergoing major abdominal surgery for cancer \parencite{mokart_early_2002}. 
+They reported that the levels of anti-inflammatory cytokines IL-10 and IL-1ra were significantly greater in patients who developed infective complications and were correlated with IL-6 levels on the first postoperative day.
 T-cell function is also depressed after surgical trauma and the degree of suppression is related to the initial pro-inflammatory response and IL-6 \parencite{faist_immunosuppression_1997}. 

### 2015-08-27 Tables/crp_comp_CRP4_vs_LOS.tex
CRP Complications -  Paper review complete

@@
 		\hline
-		                      &          &  \multicolumn{2}{c|}{Day 4 CRP (mg/dl)}   \\
-		                      &          & $\leq$125   & $>$125       & p            \\ \hline
+		                      &          & \multicolumn{2}{c}{Day 4 CRP (mg/l)} &  \\
+		                      &          & $\leq$125   & $>$125                   & \textit{p}   \\ \hline
 		Postoperative stay    & Days     & 12 (9 - 16) & 15 (13 - 26)             & $<$0.001$^a$ \\

### 2015-08-27 Tables/crp_comp_ROC_infections_noPOPF.tex
CRP_Complications - Minor Changes
- Multiple minor cosmetic changes
Final Review - CRP Complications
- Numerous minor changes to text and to multiple tables
CRP Complications -  Paper review complete

@@
 	\centering
-	\caption{Receiver operating characteristics curve analysis of C-reactive protein as a marker of postoperative infective complications in patients who did not develop a POPF}
+	\caption{Receiver operating characteristics analysis of C-reactive protein as a marker of postoperative infective complications in patients who did not develop a POPF}
 	\label{table:crp_comp_ROC_infections_noPOPF}
-	\renewcommand{\arraystretch}{1.2} %Increases space between rows
+	\renewcommand{\arraystretch}{1.4} %Increases space between rows
 	%\setlength{\tabcolsep}{9pt} %sets the space between columns
@@
 		Day & AUC  & p        & 95\% CI     & CRP & Spec. & Sens. & NPV  & PPV               \\ \hline
-		3   & 0.64 & 0.011    & 0.54 - .74 & 177.5 & 0.66  & 0.64  & 0.79 & 0.48             \\
-		4   & 0.71 & $<$0.001 & 0.61 - .80 & 125   & 0.64  & 0.74  & 0.83 & 0.5              \\
-		5   & 0.70 & $<$0.001 & 0.61 - .80 & 106.5 & 0.64  & 0.63  & 0.78 & 0.45             \\
-		6   & 0.73 & $<$0.001 & 0.65 - .82 & 91.5  & 0.68  & 0.73  & 0.84 & 0.53             \\
-		7   & 0.80 & $<$0.001 & 0.72 - .88 & 87    & 0.65  & 0.81  & 0.87 & 0.52             \\ \hline
-		\multicolumn{9}{l}{AUC - Area under curve; Spec. - Specificity, Sens. - Sensitivity} \\
-		\multicolumn{9}{l}{NPV - Negative Predictive Value; PPV - Positive Predictive Value}
+		3   & 0.64 & 0.011    & 0.54 - 0.74 & 178 & 0.66  & 0.64  & 0.79 & 0.48              \\
+		4   & 0.71 & $<$0.001 & 0.61 - 0.80 & 125 & 0.64  & 0.74  & 0.83 & 0.50              \\
+		5   & 0.70 & $<$0.001 & 0.61 - 0.80 & 107 & 0.64  & 0.63  & 0.78 & 0.45              \\
+		6   & 0.73 & $<$0.001 & 0.65 - 0.82 & 92  & 0.68  & 0.73  & 0.84 & 0.53              \\
+		7   & 0.80 & $<$0.001 & 0.72 - 0.88 & 87  & 0.65  & 0.81  & 0.87 & 0.52              \\ \hline
+		\multicolumn{9}{l}{AUC - Area under curve, Spec. - Specificity, Sens. - Sensitivity} \\
+		\multicolumn{9}{l}{NPV - Negative Predictive Value, PPV - Positive Predictive Value}
 	\end{tabular}

### 2015-08-27 Chapters/BodyComp.tex
Final commit before Access VBA

@@
-\chapter{An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing major pancreatic surgery.}
+\chapter{An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing pancreaticoduodenectomy.}
 \label{ch_bodycomp}

### 2015-08-27 Chapters/Discussion.tex
Final commit before Access VBA

@@
 Some have suggested that comorbidity was associated with shorter survival in patients with pancreatic cancer \parencite{dias-santos_charlson_2015}.
-However, other authors have reported that the relatively short median survival after pancreatic cancer surgery is not affected by comorbidity whose effects are more long-term \parencite{kos_evaluation_2014}
+However, other authors have reported that the relatively short median survival after pancreatic cancer surgery is not affected by comorbidity whose effects are more long-term \parencite{kos_evaluation_2014}.
 
@@
 %################################################################################
-In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} was the lack of association between obstructive jaundice and poor cardiopulmonary function. 
+In our opinion, the most important observation made in chapter \ref{ch_cpet_jaundice} was the lack of association between obstructive jaundice and cardiopulmonary function. 
 We are aware of only one other similar study by Junejo and co-workers.
@@
 This is especially true for $\dot{V}_{O_2}$ that has been scaled for total body weight.
-Further studies must evaluate other parameters such as $O_2$Pulse or $\dot{V}_E/\dot{V}_{O_2}$ for their ability to predict postoperative outcomes. 
+Further studies must evaluate other parameters such as $O_2$Pulse or $\dot{V}_E/\dot{V}_{CO_2}$ for their ability to predict postoperative outcomes. 
 The proportion of obese patients undergoing major surgery is increasing. 
-Moreover, sarcopenia and sarcopenic-obesity are recognised as an important adverse prognostic feature in pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012} and is associated with increased postoperative complications after pancreaticoduodenectomy \parencite{joglekar_sarcopenia_2015, pausch_cachexia_2012}.
+Moreover, sarcopenia and sarcopenic-obesity are recognised as important adverse prognostic features in pancreatic cancer \parencite{tan_sarcopenia_2009, peng_impact_2012} and were associated with increased postoperative complications after pancreaticoduodenectomy \parencite{joglekar_sarcopenia_2015, pausch_cachexia_2012}.
 Sarcopenia was also an independent risk factor for an elevated postoperative systemic inflammatory response in patients with colorectal cancer \parencite{reisinger_sarcopenia_2015}.
@@
 Our results suggest that poor performance at cardiopulmonary exercise testing may not necessarily be related to only cardiac or pulmonary function. 
-Body composition appears to not only affect aerobic capacity but also systemic inflammation.
+Body composition appears to affect not only aerobic capacity but also systemic inflammation.
 Improvements in body composition in cancer patients are achievable through nutritional supplementation \parencite{machado_whey_2015} as well as increased physical activity.
@@
 
-Birkmeyer and co-workers were the first to report in 1999 that hospital volume was not only related to postoperative mortality and morbidity but also to long-term cancer survival in patients undergoing pancreaticoduodenectomy in the United States\parencite{birkmeyer_relationship_1999}.
+Birkmeyer and co-workers were the first to report in 1999 that hospital volume was not only related to postoperative mortality and morbidity but also to long-term cancer survival in patients undergoing pancreaticoduodenectomy in the United States \parencite{birkmeyer_relationship_1999}.
 Better overall quality of care, better cancer operations based on oncological principles, better rates of adjuvant therapy may have all contributed to better outcomes in high-volume centres. 
 They also reported in 2003 that operative mortality was inversely related to surgeon volume emphasising the importance of surgical expertise in improving outcomes \parencite{birkmeyer_surgeon_2003}.
-Gouma and co-workers reported very similar findings in the Netherlands were morbidity and mortality after pancreaticoduodenectomy were linked to hospital volume\parencite{gouma_rates_2000}.
+Gouma and co-workers reported very similar findings in the Netherlands were morbidity and mortality after pancreaticoduodenectomy were linked to hospital volume \parencite{gouma_rates_2000}.
 

### 2015-08-27 Chapters/Introduction.tex
Final commit before Access VBA

@@
 In patients with unresectable disease or who are not fit to undergo surgery, palliative chemotherapy plays a limited role in prolonging survival. 
-Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of physiological fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on p\pageref{sec:comorbidity_risk_stratification}.
+Assessment of resectability of lesions of the head of the pancreas is discussed in the next section while the assessment of physiological fitness and the impact of comorbidity are discussed in detail in section \ref{sec:comorbidity_risk_stratification} on page \pageref{sec:comorbidity_risk_stratification}.
 
@@
 
-Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on p\pageref{table:isgps_popf}. 
+Three grades of postoperative pancreatic fistula have been defined based on clinical severity as described in Table \ref{table:isgps_popf} on page \pageref{table:isgps_popf}. 
 Grade B and C fistulae are considered to be clinically significant as they alter patient management and are often associated with other secondary complications such as intra-abdominal sepsis, post-pancreatectomy haemorrhage or delayed gastric emptying.
@@
 Post-pancreatectomy haemorrhage is usually managed with angiographic embolisation of the bleeding vessel and surgical intervention is only rarely required. 
-The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery \parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on p\pageref{table:isgps_pph}.
+The grading of severity of post-pancreatectomy haemorrhage as described by the International Study Group of Pancreatic Surgery \parencite{wente_postpancreatectomy_2007} is shown in Table \ref{table:isgps_pph} on page \pageref{table:isgps_pph}.
 
@@
 The Clavien-Dindo system grades the severity of complications based on the impact of the complication on the management of the patient and has been validated in large numbers of surgical patients \parencite{clavien_clavien-dindo_2009, dindo_classification_2004}.
-The classification is summarised in Table \ref{table:clavien-dindo} on p\pageref{table:clavien-dindo}. 
+The classification is summarised in Table \ref{table:clavien-dindo} on page \pageref{table:clavien-dindo}. 
 and was used to grade complications in this thesis. 
@@
 The POSSUM Physiology Score has been reported to be associated with preoperative systemic inflammation and cancer-specific survival in patients who underwent potentially curative surgery for colorectal cancer \parencite{richards_relationship_2010}.
-The variables used in the calculation of the POSSUM Physiology Score are shown in Table \ref{table:intro_possum} on p\pageref{table:intro_possum}.
+The variables used in the calculation of the POSSUM Physiology Score are shown in Table \ref{table:intro_possum} on page \pageref{table:intro_possum}.
 
@@
 
-The list of parameters measured is provided in Tables \ref{table:spirometry} and \ref{table:cpet_parameters} on p\pageref{table:spirometry}.
+The list of parameters measured is provided in Tables \ref{table:spirometry} and \ref{table:cpet_parameters} on page \pageref{table:spirometry}.
 A detailed description of these parameters is provided in Section \ref{sec:cpx_parameters}.
@@
 
-During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on p\pageref{fig:cpet_vslope}. 
-However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionate to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on p\pageref{eq:bicarb_buffer}.
+During aerobic phase of exercise, $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ share a linear relationship as shown in the left half of Figure \ref{fig:cpet_vslope} on page \pageref{fig:cpet_vslope}. 
+However, as anaerobic respiration starts to supplement aerobic respiration, $\dot{V}_{CO_2}$ increases disproportionate to $\dot{V}_{O_2}$ as a direct result of the respiratory buffer described in equation \ref{eq:bicarb_buffer} on page \pageref{eq:bicarb_buffer}.
 This results in a distinct difference in the slope of the relationship between $\dot{V}_{O_2}$ and $\dot{V}_{CO_2}$ during aerobic exercise (Line A) versus during anaerobic exercise (Line B). 
@@
 Stroke volume is in turn influenced by ventricular function and end-diastolic volumes. 
-The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on p\pageref{sec:heart_rate}.
+The heart rate response to exercise is discussed in section \ref{sec:heart_rate} on page \pageref{sec:heart_rate}.
 

### 2015-08-27 main.tex
Grrrr...
- EJD ack
- Comment out line numbers
- Comment out todos
Paper Review - Half-way point
- Big error in SIMD - got it wrong way round and bad numbers in two chapters, multiple tables affected - fixed
- minor edit to main.tex - new package
- Other minor style and referencing changes
Final commit before Access VBA

@@
 \usepackage{pdfpages} %for including pdf files or individual pages of pdf files
-\usepackage{easy-todo} %Adds todos to the document. Use [disable] as package option in final verion. Also see \listoftodos below.
+\usepackage{textcomp}
+%\usepackage{easy-todo} %Adds todos to the document. Use [disable] as package option in final verion. Also see \listoftodos below.
 %\usepackage[pagewise]{lineno} %To help Donny review. Remove for final version
@@
 
-Mr. Euan J Dickson,
+Mr. Euan J Dickson,\\
 West of Scotland Pancreatic Unit, Glasgow Royal Infirmary
@@
 
-	Measurement of biochemical and haematological data was performed by the hospital laboratory service.
+	Measurement of biochemical and haematological parameters was performed by the hospital laboratory service.
 
-	Statistical analysis was performed with the assistance of Prof Donald C McMillan, University Dept of Surgery, Royal Infirmary, Glasgow.
+	Cardiopulmonary exercise tests were performed by the Department of Respiratory Medicine at the Glasgow Royal Infirmary.
+
+	Statistical analysis was performed with the assistance of Prof Donald C McMillan, Academic Unit of Surgery, Royal Infirmary, Glasgow.
 
@@
 
-\clearpage
-\listoftodos
+%\clearpage
+%\listoftodos
 

### 2015-08-27 Appendices/AppendixAccessDatabase.tex
Access VBA code added
- I think this officially completes this thesis

@@
 \clearpage
-%------------------------------------------
+%-------------------------------------------------------------
 
 \setstretch{1.0}
+
+The following VBA code was written by me to facilitate collection of laboratory data from the existing user interface that did not allow easy automation.
+It forms part of a Microsoft Access database.
+Some parts of the code has been removed or reformatted to be accommodated within the constraints of this thesis.
+
+The RealLink for Windows (RFW, http://www.nps-reality.com/reallink) application used at Glasgow Royal Infirmary allowed access form VBA via Dynamic Data Exchange.
+This was used to collect and parse laboratory results for preoperative and postoperative blood results efficiently and without error.
+Where necessary, I have used information from the MSDN knowledgebase and internet fora to supplement my knowledge of coding.
+
+\begin{lstlisting}
+'#########################################################################
+'Dynamic Data Exchange with RealLink application
+'#########################################################################
+Option Compare Database
+Function RLChannel() As Long
+  Dim AppName As String
+  Dim Topic As String
+  AppName = "******"
+  Topic = "******"
+  DDETerminateAll
+
+  On Error GoTo ErrorHandler
+  RLChannel = DDEInitiate(AppName, Topic)
+  Exit Function
+  ErrorHandler:
+  MsgBox "Error initiating connection to RealLink   Error ID: " & _
+  Err.Description, vbOKOnly, "Error"
+End Function
+Sub RLSendData(strdata As String, WaitSecs As Integer)
+  On Error GoTo ErrorHandler
+  Dim starttime As Double
+  AppActivate "RealLink   NGTBIO", Wait
+  starttime = Timer
+  SendKeys strdata, True
+  starttime = Timer
+  While Timer < starttime + WaitSecs: Wend
+  Exit Sub
+  ErrorHandler:
+'Error gets dealt with by procedure calling this function
+End Sub
+
+Function RLGetData(ChannelNum As Long) As String
+  On Error GoTo ErrorHandler
+  Dim RLItem As String
+  RLItem = "Table 0,0,80,24,Comma"
+  RLGetData = DDERequest(ChannelNum, RLItem)
+  Exit Function
+  ErrorHandler:
+  MsgBox "Error capturing data from RealLink   : " & _ 
+    Err.Description, vbOKOnly, "Error"
+End Function
+
+Sub WaitFor(intSeconds As Integer)
+  Dim starttime As Single
+  starttime = Timer
+  While Timer < starttime + intSeconds: Wend
+End Sub
+Function ChkLabScreen(strLabScreen As String, _
+ strSurname As String, strLab As String) As Integer
+  If InStr(1, strLabScreen, strSurname) = 0 Then 
+    ChkLabScreen = ChkLabScreen + 1
+  If InStr(1, strLabScreen, strLab) Then 
+    ChkLabScreen = ChkLabScreen + 2
+End Function
+Sub RunRealLink()
+'Wait for application to start then wait 5 seconds
+'Shell runs commands asynchronously
+'DDEPoke does not work   use windows shell and processID instead
+  Dim dReturnValue As Single
+  dReturnValue = Shell("C:\RFW\RFW.EXE /LC:\RFW\RFWDDE.DLL NGTBIO",_
+   vbMaximizedFocus)
+  While dReturnValue < 0: Wend
+  Call WaitFor(5)
+End Sub
+Sub Login(LabNum As String, Username As String, Password As String)
+' Login to Biochem/Haem
+  Dim strSend As String
+
+  strSend = "***" & vbCr
+  Call RLSendData(strSend, 2)
+
+  strSend = LabNum & vbCr
+  Call RLSendData(strSend, 2)
+
+  strSend = Username & vbCr
+  Call RLSendData(strSend, 1)
+
+  strSend = Password & vbCr
+  Call RLSendData(strSend, 1)
+End Sub
+Sub SearchPatient(Surname As String, HospitalNumber As String, _
+ Optional EarliestDate As Date, Optional LatestDate As Date)
+'Enter patient information and date range(optional)
+  Dim strSend As String
+  strSend = ""
+  Call RLSendData(strSend, 1)
+  strSend = HospitalNumber & vbCr
+  Call RLSendData(strSend, 1)
+  strSend = Left(Surname, 2) & vbCr
+  Call RLSendData(strSend, 1)
+  strSend = Format(EarliestDate, "dd.mm.yy") & vbCr
+  Call RLSendData(strSend, 0) 'Earliest date to search
+  strSend = Format(LatestDate, "dd.mm.yy") & vbCr
+  Call RLSendData(strSend, 0) 'Latest date to search
+End Sub
+
+'#########################################################  
+'Parser for results scraped from RealLink screengrab
+'#########################################################
+Option Compare Database
+'Results parsers
+Public Sub ParseFBC(strHD As String)
+On Error GoTo ErrorHandler
+Dim strA() As String 'string to hold rows of data
+Dim intRows As Long
+Dim intTests As Integer
+Dim intPosition As Integer
+Dim strTest() As Variant 'name of test eg WBC
+Dim strResult(10, 1) As String 
+'change first dimension if changing number of tests
+strTest = Array("WBC", "Hb", "Hct", "MCV", "Plts", _
+"NEUT", "LYMPH", "MONO", "EOS", "BASO")
+intTests = UBound(strTest)
+strA = Split(strHD, vbCr)
+intRows = UBound(strA)
+Dim tempRow As String
+Dim rowdate As Integer
+For rowdate = 0 To 10
+  tempRow = strA(rowdate)
+  If InStr(1, tempRow, "Collected") > 0 Then
+   Dim strDateTime() As String
+   Dim strDate As String
+   Dim strTime As String
+   strDateTime = Split(tempRow, "Collected")
+   strDateTime = Split(strDateTime(1), ",")
+   If InStr(1, strDateTime(1), "*") > 0 Then
+   strDate = Trim(Replace(strDateTime(1), ".", "/"))
+   End If
+   If UBound(strDateTime) > 2 Then
+   strDate = Trim(Replace(strDateTime(1), ".", "/"))
+   strTime = Trim(strDateTime(2))
+   End If
+  End If
+Next rowdate
+For t = 0 To intTests 'cycle through list of tests   strTest
+  For i = 0 To intRows 'cycle through rows   strA
+   tempRow = strA(i)
+   tempRow = Replace(tempRow, vbCrLf, "")
+   tempRow = Replace(tempRow, vbLf, "")
+   tempRow = Replace(tempRow, vbCr, "")
+   temptest = strTest(t)
+   intPosition = InStr(tempRow, temptest)
+   If intPosition > 0 Then
+    Dim strTemp() As String
+    strTemp = Split(tempRow, ",")
+
+    For j = 0 To UBound(strTemp) 
+    'cycle through values in a single row   strTemp
+     If Trim(strTemp(j)) = strTest(t) Then
+      strResult(t, 0) = strTest(t)
+
+      If Trim(strTemp(j + 1)) = "+" Or Trim(strTemp(j + 1)) = " " Then
+       strResult(t, 1) = Trim(strTemp(j + 2))
+      Else
+       strResult(t, 1) = Trim(strTemp(j + 1))
+      End If
+     End If
+    Next j
+   End If
+  Next i
+Next t
+'Fill in results into form
+  Call SaveHaem(strDate, strTime, strResult)
+  Exit Sub
+  ErrorHandler:
+End Sub
+Sub ParseCoag(strCoag As String)
+  On Error GoTo ErrorHandler
+  Dim strA() As String 'string to hold rows of data
+  Dim intRows As Long
+  Dim intTests As Integer
+  Dim intPosition As Integer
+  Dim strTest() As Variant 'name of test eg WBC
+  Dim strResult(23, 1) As String 
+  'change first dimension if changing number of tests
+  strTest = Array("PT", "PT C Ratio", "APTT", "APTT Ratio", "TCT", "TCT Ratio")
+  intTests = UBound(strTest)
+  strA = Split(strCoag, vbCr)
+  intRows = UBound(strA)
+  Dim tempRow As String
+  Dim rowdate As Integer
+  For rowdate = 0 To 10
+   tempRow = strA(rowdate)
+   If InStr(1, tempRow, "Collected") > 0 Then
+    Dim strDateTime() As String
+    Dim strDate As String
+    Dim strTime As String
+    strDateTime = Split(tempRow, "Collected")
+    strDateTime = Split(strDateTime(1), ",")
+    If InStr(1, strDateTime(1), "*") > 0 Then
+    strDate = Trim(Replace(strDateTime(1), ".", "/"))
+    End If
+    If UBound(strDateTime) > 2 Then
+    strDate = Trim(Replace(strDateTime(1), ".", "/"))
+    strTime = Trim(strDateTime(2))
+    End If
+   End If
+  Next rowdate
+  For t = 0 To intTests 'cycle through list of tests   strTest
+   For i = 0 To intRows 'cycle through rows   strA
+   tempRow = strA(i)
+   tempRow = Replace(tempRow, ",+", "")
+   tempRow = Replace(tempRow, ", ", "")
+   tempRow = Replace(tempRow, "+,", ",")
+   tempRow = Replace(tempRow, " ,", ",")
+   tempRow = Replace(tempRow, "|", ",")
+   temptest = strTest(t)
+   intPosition = InStr(tempRow, temptest)
+   If intPosition > 0 Then
+    Dim strTemp() As String
+    strTemp = Split(tempRow, ",")
+    For j = 0 To UBound(strTemp) 
+    'cycle through values in a single row   strTemp
+     If Trim(strTemp(j)) = strTest(t) Then
+      strResult(t, 0) = strTest(t)
+      If Trim(strTemp(j + 1)) = "+" Or Trim(strTemp(j + 1)) = " " _
+        Or Trim(strTemp(j + 1)) = "" Then
+       strResult(t, 1) = Trim(strTemp(j + 2))
+      Else
+       strResult(t, 1) = Trim(strTemp(j + 1))
+      End If
+     End If
+    Next j
+   End If
+    Next i
+  Next t
+'Fill in results into form
+  Call SaveCoag(strDate, strTime, strResult)
+  Exit Sub
+  ErrorHandler:
+End Sub
+Sub ParseBio(strBD As String)
+  Dim strSurname As String
+  Dim strHNum As String
+  Dim strA() As String 'string to hold rows of data
+  Dim intRows As Long
+  Dim intTests As Integer
+  Dim intPosition As Integer
+  Dim strTest() As Variant 'name of test eg Sodium
+  strTest = Array("Sodium", "Potassium", "Urea", "Creat", "eGFR", "Phos",_
+   "Mg++", "Chloride", "Bili", "G GT", "A Phos", "ALT", "AST",_
+   "T Prot", "Alb", "Globulin", "Glucose", "CRP", "Trop I", _
+   "Calcium", "Ad Cal", "Amylase")
+  Dim strResult(21, 1) As String 
+  'change first dimension if changing number of tests
+  intTests = UBound(strTest)
+  strBD = Replace(strBD, ".Amylase.", "..")
+  strBD = Replace(strBD, vbCrLf, "")
+  strBD = Replace(strBD, vbLf, "")
+  strBD = Replace(strBD, vbCr, "")
+  strA = Split(strBD, "|")
+  Dim strTempDate As String
+  strTempDate = strA(0)
+  If InStr(strTempDate, "Collected") Then
+   Dim strDateTime() As String
+   Dim strDate As String
+   Dim strTime As String
+   strDateTime = Split(strTempDate, "Collected")
+   strDateTime = Split(strDateTime(1), ",")
+   strDate = strDateTime(1)
+   strDate = Replace(strDate, ".", "/")
+   strTime = strDateTime(2)
+  End If
+  intRows = UBound(strA)
+  For t = 0 To intTests 'cycle through list of tests   strTest
+   For i = 0 To intRows 'cycle through rows   strA
+   tempRow = strA(i)
+   temptest = strTest(t)
+   intPosition = InStr(tempRow, temptest)
+    If intPosition > 0 Then
+     Dim strTemp() As String
+     Dim intDots As Integer
+
+    'Routine to remove dots word dots
+     intDots = InStr(strA(i), "..")
+     If intDots > 0 Then
+      strTemp = Split(NoDots(strA(i)), ",")
+     Else
+      strTemp = Split(strA(i), ",")
+     End If
+    'End of Routine to remove dots word dots
+
+     For j = 0 To UBound(strTemp) 
+     'cycle through values in a single row   strTemp
+      If Trim(strTemp(j)) = strTest(t) Then
+       strResult(t, 0) = strTest(t)
+
+       If Trim(strTemp(j + 1)) = "+" Or Trim(strTemp(j + 1)) = " " Then
+        strResult(t, 1) = Trim(strTemp(j + 2))
+       Else
+        strResult(t, 1) = Trim(strTemp(j + 1))
+       End If
+      End If
+     Next j
+    End If
+   Next i
+  Next t
+  'Fill in results into form
+   Call SaveBio(strDate, strTime, strResult)
+End Sub
+Function NoDots(strWithDots As String) As String
+  Dim intDotEndPos As Integer
+  Dim intDotStartPos As Integer
+  Dim intRowLen As Integer
+  intDotEndPos = InStrRev(strWithDots, "..") 
+  'find the first occurrence of .. from the end
+  intDotStartPos = InStr(strWithDots, "..") 
+  'find the first occurrence of .. from the end
+  intRowLen = Len(strWithDots) 
+  'get the total length of the string with dots
+'get the substring on the right and left of the dots based on above values
+  NoDots = Left(strWithDots, intDotStartPos) & "," & _
+  Right(strWithDots, intRowLen   intDotEndPos   1)
+'Debug.Print NoDots
+End Function
+
+'#########################################################
+'Code to save results to respective tables
+'#########################################################
+Option Compare Database
+Sub SaveBio(strDate, strTime, strResult)
+Dim tf_B As SubForm
+Set tf_B = Forms!frm_Demo!frm_Bloods
+'create a new record
+tf_B.SetFocus
+DoCmd.GoToRecord , , acNewRec
+tf_B!CHI = Forms!frm_Demo!CHI
+tf_B!Surname = Forms!frm_Demo!Surname
+tf_B!Date = strDate
+tf_B!Time = strTime
+intTests = UBound(strResult, 1)
+For i = 0 To intTests
+  Select Case strResult(i, 0)
+  Case "Potassium"
+   tf_B!K = strResult(i, 1)
+  Case "Urea"
+   tf_B!Urea = strResult(i, 1)
+  Case "Creat"
+   tf_B!Creatine = strResult(i, 1)
+'Case "eGFR"
+  Case "Phos"
+   tf_B!PO4 = strResult(i, 1)
+  Case "Mg++"
+   tf_B!Magnesium = strResult(i, 1)
+  Case "Chloride"
+   tf_B!Cl = strResult(i, 1)
+  Case "Bili"
+   tf_B!Bilirubin = strResult(i, 1)
+  Case "G GT"
+   tf_B!GGT = strResult(i, 1)
+  Case "A Phos"
+   tf_B!Al_Phos = strResult(i, 1)
+  Case "ALT"
+   tf_B!ALT = strResult(i, 1)
+  Case "AST"
+   tf_B!AST = strResult(i, 1)
+  Case "T Prot"
+   tf_B!Protein = strResult(i, 1)
+  Case "Alb"
+   tf_B!Albumin = strResult(i, 1)
+  Case "Globulin"
+   tf_B!Globulin = strResult(i, 1)
+  Case "Glucose"
+   tf_B!Glucose = strResult(i, 1)
+  Case "CRP"
+   tf_B!CRP = strResult(i, 1)
+  Case "Trop I"
+   tf_B!Troponin = strResult(i, 1)
+  Case "Calcium"
+   tf_B!Ca = strResult(i, 1)
+  Case "Ad Cal"
+   tf_B!Adj_Ca = strResult(i, 1)
+  Case "Amylase"
+   tf_B!Amylase = strResult(i, 1)
+  Case "Sodium"
+   tf_B!Na = strResult(i, 1)
+  End Select
+  Next i
+End Sub
+Sub SaveCoag(TestDate As String, TestTime As String, Results() As String)
+Dim tf_C As SubForm
+Set tf_C = Forms!frm_Demo!frm_Coag
+'create a new record
+tf_C.SetFocus
+DoCmd.GoToRecord , , acNewRec
+tf_C!CHI = Forms!frm_Demo!CHI
+tf_C!Surname = Forms!frm_Demo!Surname
+tf_C!Date = TestDate
+If TestTime <> "*" And TestTime <> "NK" Then tf_C!Time = TestTime
+intTests = UBound(Results, 1)
+For i = 0 To intTests
+  Select Case Results(i, 0)
+  Case "PT"
+   tf_C!PT = Results(i, 1)
+  Case "PT C Ratio"
+   tf_C![PT C Ratio] = Results(i, 1)
+  Case "APTT"
+   tf_C!APTT = Results(i, 1)
+  Case "APTT Ratio"
+   tf_C![APTT Ratio] = Results(i, 1)
+  Case "TCT"
+   tf_C!TCT = Results(i, 1)
+  Case "TCT Ratio"
+   tf_C![TCT Ratio] = Results(i, 1)
+  End Select
+  Next i
+End Sub
+Sub SaveHaem(TestDate As String, TestTime As String, Results() As String)
+Dim tf_H As SubForm
+Set tf_H = Forms!frm_Demo!frm_Haematology
+'create a new record
+tf_H.SetFocus
+DoCmd.GoToRecord , , acNewRec
+tf_H!CHI = Forms!frm_Demo!CHI
+tf_H!Surname = Forms!frm_Demo!Surname
+tf_H!Date = TestDate
+If TestTime <> "*" Then tf_H!Time = TestTime
+intTests = UBound(Results, 1)
+For i = 0 To intTests
+'frmHaem.strResult(i, 0) = strResult(i, 1)
+  Select Case Results(i, 0)
+  Case "WBC"
+   tf_H!WBC = Results(i, 1)
+  Case "Plts"
+   tf_H!Plts = Results(i, 1)
+  Case "Hb"
+   tf_H!HB = Results(i, 1)
+  Case "MCV"
+   tf_H!MCV = Results(i, 1)
+  Case "MCHC"
+   tf_H!MCHC = Results(i, 1)
+  Case "RDW"
+   tf_H!RDW = Results(i, 1)
+  Case "RBC"
+   tf_H!RBC = Results(i, 1)
+  Case "NEUT"
+   tf_H!NEUT = Results(i, 1)
+  Case "LYMPH"
+   tf_H!LYMPH = Results(i, 1)
+  Case "MONO"
+   tf_H!MONO = Results(i, 1)
+  Case "EOS"
+   tf_H!EOS = Results(i, 1)
+  Case "BASO"
+   tf_H!BASO = Results(i, 1)
+  Case "Hct"
+   tf_H!Hct = Results(i, 1)
+  Case "MCH"
+   tf_H!MCH = Results(i, 1)
+  Case "MPV"
+   tf_H!MPV = Results(i, 1)
+  Case "PDW"
+   tf_H!PDW = Results(i, 1)
+  Case "PCT"
+   tf_H!PCT = Results(i, 1)
+  End Select
+  Next i
+End Sub
+
+'#########################################################
+'User interface functions
+'#########################################################
+
+Option Compare Database
+Private Sub Form_Load()
+  Me.RecordSource = "tb_Demo"
+End Sub
+Private Sub cmbAllPatients_AfterUpdate()
+  ' Find the record that matches the control.
+  Dim rs As Object
+  Set rs = Me.Recordset.Clone
+  rs.FindFirst "[IPID] = " & Str(Nz(Me![cmbAllPatients], 0))
+  If Not rs.EOF Then Me.Bookmark = rs.Bookmark
+End Sub
+Private Sub cmbChooseProcedure_AfterUpdate()
+  'Filter by operation name
+  Dim strSQL As String
+  Dim strFilter As String
+  If Me.cmbChooseProcedure = "All" Or IsNull(Me.cmbChooseProcedure) Then
+   ' If the combo is Null, use the whole table as the RecordSource.
+   Me.RecordSource = "tb_Demo"
+  Else
+   strFilter = Me.cmbChooseProcedure
+   strSQL = "SELECT tb_Demo.* FROM tb_Demo LEFT JOIN tb_Operation ON _
+    tb_Demo.IPID = tb_Operation.IPID WHERE ((([tb_Operation]![Primary _
+     Procedure]) LIKE '*" & strFilter & "*'))ORDER BY _
+      [tb_Operation]![Operation Date];"
+   Me.RecordSource = strSQL
+  End If
+End Sub
+Private Sub cmbChooseOpDate_AfterUpdate()
+  'Filter by operation DATE
+  Dim strSQL As String
+  Dim strFilter As Date
+  If IsNull(Me.cmbChooseOpDate) Then
+   ' If the combo is Null, use the whole table as the RecordSource.
+   Me.RecordSource = "tb_Demo"
+  Else
+   strFilter = CDate(Me.cmbChooseOpDate)
+   strSQL = "SELECT tb_Demo.* FROM tb_Demo LEFT JOIN tb_Operation ON _
+    tb_Demo.IPID = tb_Operation.IPID WHERE _
+    ((([tb_Operation]![Operation Date]) > #" & strFilter & "#))_
+    ORDER BY [tb_Operation]![Operation Date];"
+   Me.RecordSource = strSQL
+  End If
+End Sub
+Private Sub cmbChooseSpecialty_AfterUpdate()
+  'Filter by Specialty
+  Dim strSQL As String
+  Dim strFilter As String
+  strFilter = Me.cmbChooseSpecialty
+  If strFilter = "All" Or IsNull(Me.cmbChooseSpecialty) Then
+   ' If the combo is Null, use the whole table as the RecordSource.
+   Me.RecordSource = "tb_Demo"
+  Else
+   strSQL = "SELECT tb_Demo.* FROM tb_Demo LEFT JOIN tb_Event_Admission _
+   ON tb_Demo.IPID = tb_Event_Admission.IPID WHERE _
+    ((([tb_Event_Admission]![Specialty] _
+    LIKE '*" & strFilter & "*')));"
+   Me.RecordSource = strSQL
+  End If
+End Sub
+Private Sub cmdNewPatient_Click()
+  On Error GoTo Err_cmdNewPatient_Click
+  DoCmd.DoMenuItem acFormBar, acRecordsMenu, 5, , acMenuVer70
+  DoCmd.GoToRecord , , acNewRec
+  Exit Sub
+  Err_cmdNewPatient_Click:
+  MsgBox Err.Description
+End Sub
+Private Sub listCurrentPatients_AfterUpdate()
+  ' Find the record that matches the control.
+  Dim rs As Object
+  Set rs = Me.Recordset.Clone
+  rs.FindFirst "[IPID] = " & Str(Nz(Me![listCurrentPatients], 0))
+  If Not rs.EOF Then Me.Bookmark = rs.Bookmark
+  Me.frm_Physiology.Form.Refresh
+End Sub
+Private Sub cmdSendDetails_Click()
+  'checked and works
+  Dim Surname As String 'Surname from frm_demo
+  Dim HospitalNum As String 'Hospital number from frm_demo
+  Dim OpDate As Date 'Date of operation from frm_operation
+  Dim Edate As Date 'Earliest date of bloods to return
+  Dim Ldate As Date 'Latest date of bloods to return
+  On Error GoTo ErrorHandler:
+  Surname = Me.Surname
+  HospitalNum = Me.HospitalNum
+  OpDate = Me.frm_Operation.Form![Operation Date]
+  'Set search date range
+  Edate = OpDate   10
+  Ldate = OpDate + 5
+  If Me.chkResuse = 0 Then 'Is Reuse box not checked then start a new RL session
+   Call RunRealLink
+   Select Case Me.frameChooseLab.Value
+    Case 1
+     Call Login("3", "******", "******")
+    Case 2
+     Call Login("2", "******", "******")
+   End Select
+  End If
+  Call SearchPatient(Surname, HospitalNum, Edate, Ldate)
+  Exit Sub
+  ErrorHandler:
+MsgBox "cmdHaem Error" & Err.Description
+End Sub
+Private Sub cmdImport_Click()
+  On Error GoTo ErrorHandler
+  Dim intRepeat As Integer
+  KillSwitch = False
+  If CInt(Me.txtAutoRepeat) > 0 Then
+   intRepeat = CInt(Me.txtAutoRepeat)
+  Else
+   intRepeat = 1
+  End If
+  If Me.chkAutoUpdate.Value =  1 Then
+   For i = 1 To intRepeat
+    Call ImportData
+    Me.Refresh
+    Call cmdLater_Click
+    If KillSwitch = True Then Exit Sub
+   Next
+  End If
+  ErrorHandler:
+  Exit Sub
+End Sub
+Private Sub ImportData()
+  Dim LabChannel As Long
+  LabChannel = RLChannel
+  'Grab screen text from existing RL window
+  Dim strResultScreen As String
+  Call WaitFor(3)
+  If LabChannel > 0 Then
+   strResultScreen = RLGetData(LabChannel)
+  Else
+   MsgBox "RealLink connection could not be established", vbOKOnly,_
+    "Connection Error"
+   Exit Sub
+  End If
+  'Send to either FBC/Coag parser
+  If InStr(1, strResultScreen, "Press Esc") Then KillSwitch = True
+  If InStr(1, strResultScreen, "..FBC..") Then
+   Call ParseFBC(strResultScreen)
+  End If
+  If InStr(1, strResultScreen, "..C/Screen..") Then
+   Call ParseCoag(strResultScreen)
+  End If
+  If InStr(1, strResultScreen, "Biochem") Then
+   Call ParseBio(strResultScreen)
+  End If
+  'Terminate all DDE connections
+DDETerminate LabChannel
+End Sub
+Private Sub cmdEarlier_Click()
+  Dim strTemp As String
+  strTemp = "E" & vbCr
+  Call RLSendData(strTemp, 0)
+End Sub
+Private Sub cmdLater_Click()
+  Dim strTemp As String
+  strTemp = "L" & vbCr
+  Call RLSendData(strTemp, 0)
+End Sub
+Private Sub tb_ABG_ALL_subform_Enter()
+  Me.Refresh
+End Sub
+Private Sub tgDemoDupes_Click()
+  Select Case Me.tgDemoDupes.Value
+   Case  1
+    Me.tgDemoDupes.SetFocus
+    Me.tgDemoDupes.Caption = "DemoDupes ON"
+    Me.RecordSource = "qs_DemoDupes"
+   Case 0
+    Me.tgDemoDupes.SetFocus
+    Me.tgDemoDupes.Caption = "DemoDupes OFF"
+    Me.RecordSource = "tb_Demo"
+  End Select
+End Sub
+Private Sub cmdRefresh_Click()
+  On Error GoTo Err_cmdRefresh_Click
+  DoCmd.DoMenuItem acFormBar, acRecordsMenu, 5, , acMenuVer70
+  Exit_cmdRefresh_Click:
+  Exit Sub
+  Err_cmdRefresh_Click:
+  MsgBox Err.Description
+  Resume Exit_cmdRefresh_Click
+End Sub
+
+'#########################################################
+'VBA code to import CPET results into database
+'#########################################################
+Option Compare Database
+Private Sub cmdImport_Click()
+  Dim strEmptyline As String
+  On Error Resume Next
+  Dim strResult As String
+  Dim strResult2 As String
+  Dim strTemp() As String
+  Dim channel As Long
+  channel = TermChannel
+  strResult = TermGetData(channel)
+  If strResult = "" Then Exit Sub
+  strResult = Replace(strResult, "qq", "")
+  'strTemp = Split(strResult, "COMMENT SCREEN")
+  ' strResult = strTemp(1)
+  'strResult = strTemp(0)
+  strTemp = Split(strResult, vbCrLf)
+  For j = 0 To UBound(strTemp)
+  If Trim(strTemp(j)) <> "" Then
+   Me.txtFull.SetFocus
+   Me.txtFull.Value = Me.txtFull.Value & Trim(strTemp(j)) & vbCrLf
+  'Debug.Print Trim(strTemp(j))
+  End If
+  Next j
+  Me.Application.DDEPoke channel, "qvt_term_item", "1"
+  WaitFor (1)
+  DDETerminateAll
+End Sub
+Function TermGetData(ChannelNum As Long) As String
+  On Error GoTo ErrorHandler
+  Dim RLItem As String
+  RLItem = "Table 0,0,80,24,Comma"
+  TermGetData = DDERequest(ChannelNum, RLItem)
+  Exit Function
+  ErrorHandler:
+  MsgBox "Error capturing data from RealLink   : " & _
+  Err.Description, vbOKOnly, "Error"
+End Function
+Function TermChannel() As Long
+  Dim AppName As String
+  Dim Topic As String
+  AppName = "qvt_term"
+  Topic = "Respiratory"
+  DDETerminateAll
+  On Error GoTo ErrorHandler
+  TermChannel = DDEInitiate(AppName, Topic)
+  Exit Function
+  ErrorHandler:
+  MsgBox "Error initiating connection to Term   Error ID: " & _
+  Err.Description, vbOKOnly, "Error"
+End Function
+Sub TermSendData(strdata As String, WaitSecs As Integer)
+  On Error GoTo ErrorHandler
+  Dim starttime As Double
+  AppActivate "RealLink   NGTBIO", Wait
+  SendKeys strdata, True
+  starttime = Timer
+  While Timer < starttime + WaitSecs: Wend
+  Exit Sub
+  ErrorHandler:
+'Error gets dealt with by procedure calling this function
+End Sub
+Private Sub cmdSendPatient_Click()
+  Dim channel As Long
+  Dim strName As String
+  strName = "ret" & vbCr & "7" & Me.Parent!Surname & " " _
+   & Me.Parent!Forename & vbCr
+  channel = TermChannel
+  Me.Application.DDEPoke channel, "qvt_term_item", strName
+  DDETerminateAll
+End Sub
+\end{lstlisting}

### 2015-08-27 Appendices/AppendixExcelMacros.tex
Review - Minor changes to Appendices
- Modify short title
- insert \clearpage
Access VBA code added
- I think this officially completes this thesis

--- 2015-06-02 Appendices/AppendixExcelMacros.tex
@@
-% Appendix A
-
 \chapter{Visual Basic for Applications For CPET Analysis} % Main appendix title
-
 \label{AppendixExcelMacros} % For referencing this appendix elsewhere, use \ref{AppendixA}
-
-\lhead{Appendix B. \emph{Excel VBA For CPET Analysis}} % This is for the header on each page - perhaps a shortened title
+\lhead{Appendix \ref{AppendixExcelMacros}. \emph{Excel VBA For CPET Analysis}} % This is for the header on each page - perhaps a shortened title
+\clearpage
+%------------------------------------------
 
@@
    For Each ws In Worksheets
-     If ws.Range("$B$1").Value = "" Then
+     If ws.Range("$B$1").Value="" Then
        Exit Sub
@@
      If ws.Name <> "aaa_main" Then
-       strName = Split(CStr(ws.Range("$B$1").Value), "(")
-       ws.Name = LCase(strName(0))
+       strName=Split(CStr(ws.Range("$B$1").Value), "(")
+       ws.Name=LCase(strName(0))
        ws.Activate
        ws.Range("B4").Select
-       ActiveWindow.FreezePanes = True
+       ActiveWindow.FreezePanes=True
      End If
@@
 
-   SortDescending = False
+   SortDescending=False
 
-   If ActiveWindow.SelectedSheets.Count = 1 Then
+   If ActiveWindow.SelectedSheets.Count=1 Then
      'Change the 1 to the worksheet you want sorted first
-     FirstWSToSort = 1
-     LastWSToSort = Worksheets.Count
+     FirstWSToSort=1
+     LastWSToSort=Worksheets.Count
     Else
      With ActiveWindow.SelectedSheets
-       For N = 2 To .Count
+       For N=2 To .Count
          If .Item(N - 1).Index <> .Item(N).Index - 1 Then
@@
 
-       FirstWSToSort = .Item(1).Index
-       LastWSToSort = .Item(.Count).Index
+       FirstWSToSort=.Item(1).Index
+       LastWSToSort=.Item(.Count).Index
      End With
@@
 
-   For M = FirstWSToSort To LastWSToSort
-     For N = M To LastWSToSort
-       If SortDescending = True Then
+   For M=FirstWSToSort To LastWSToSort
+     For N=M To LastWSToSort
+       If SortDescending=True Then
          If UCase(Worksheets(N).Name) > UCase(Worksheets(M).Name) Then
@@
    Dim x As Integer
-   x = 2
+   x=2
    Sheets("aaa_main").Range("A:A").Clear
-   Sheets("aaa_main").Cells(1, 1) = "SheetName"
+   Sheets("aaa_main").Cells(1, 1)="SheetName"
    For Each ws In Worksheets
-     Sheets("aaa_main").Cells(x, 1) = ws.Name
-     x = x + 1
+     Sheets("aaa_main").Cells(x, 1)=ws.Name
+     x=x + 1
    Next ws
@@
 Sub GotoSheet()
-  If ActiveWindow.ActiveSheet.Name = "aaa_main" Then
+  If ActiveWindow.ActiveSheet.Name="aaa_main" Then
     Dim selName As String
-    selName = ActiveWindow.ActiveCell.Value
+    selName=ActiveWindow.ActiveCell.Value
     Worksheets(selName).Activate
@@
     Dim x As Integer
-    x = 4
+    x=4
     While ws.Cells(x, 2) > 0
       If ws.Cells(x, 9) <> "-" And ws.Cells(x, 6) <> "-" Then
-        ws.Cells(x, 22) = CDbl(ws.Cells(x, 9)) / CDbl(ws.Cells(x, 6))
+        ws.Cells(x, 22)=CDbl(ws.Cells(x, 9)) / CDbl(ws.Cells(x, 6))
       End If
-      x = x + 1
+      x=x + 1
     Wend
@@
   For Each ws In Worksheets
-    ws.Range("AD9") = ws.Range("A1")
+    ws.Range("AD9")=ws.Range("A1")
     ws.Range("A1").Clear
@@
   Dim x As Integer
-  For x = 3 To 88
-    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD8") = _
+  For x=3 To 88
+    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD8")=_
       Sheets("aaa_main").Cells(x, 27).Value
-    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD9") = _
+    Sheets(Sheets("aaa_main").Cells(x, 1).Value).Range("AD9")=_
       Sheets("aaa_main").Cells(x, 28).Value
@@
   Dim selrng As String  
-  selrow = ActiveWindow.ActiveCell.Row
-  selrng = "AF" & CStr(selrow) & ":BX" & CStr(selrow)
+  selrow=ActiveWindow.ActiveCell.Row
+  selrng="AF" & CStr(selrow) & ":BX" & CStr(selrow)
   Range(selrng).Select
@@
   Dim r As Integer
-  r = 3 - Selection.Cells(1, 1).Row
-  ActiveSheet.Range("X29") = Selection.Cells(r, 1)
-  ActiveSheet.Range("Y29") = Selection.Cells(r, 3)
-  ActiveSheet.Range("Z29") = Selection.Cells(r, 4) 'VE
-  ActiveSheet.Range("AA29") = Selection.Cells(r, 5) 'VT
-  ActiveSheet.Range("AB29") = Selection.Cells(r, 6) 'VO2
-  ActiveSheet.Range("AC29") = Selection.Cells(r, 7) 'VO2/KG
-  ActiveSheet.Range("AD29") = Selection.Cells(r, 8) 'VE/VO2
-  ActiveSheet.Range("AE29") = Selection.Cells(r, 9) 'VCO2
-  ActiveSheet.Range("AF29") = Selection.Cells(r, 10) 'VE/VCO2
-  ActiveSheet.Range("AG29") = Selection.Cells(r, 11) 'RER
-  ActiveSheet.Range("AH29") = Selection.Cells(r, 12) 'PETO2
-  ActiveSheet.Range("AI29") = Selection.Cells(r, 13) 'PETCO2
-  ActiveSheet.Range("AJ29") = Selection.Cells(r, 14) 'O2PULSE
-  ActiveSheet.Range("AK29") = Selection.Cells(r, 15) 'HR
-  ActiveSheet.Range("AL29") = Selection.Cells(r, 16) 'Bf
+  r=3 - Selection.Cells(1, 1).Row
+  ActiveSheet.Range("X29")=Selection.Cells(r, 1)
+  ActiveSheet.Range("Y29")=Selection.Cells(r, 3)
+  ActiveSheet.Range("Z29")=Selection.Cells(r, 4)'VE
+  ActiveSheet.Range("AA29")=Selection.Cells(r, 5)'VT
+  ActiveSheet.Range("AB29")=Selection.Cells(r, 6)'VO2
+  ActiveSheet.Range("AC29")=Selection.Cells(r, 7)'VO2/KG
+  ActiveSheet.Range("AD29")=Selection.Cells(r, 8)'VE/VO2
+  ActiveSheet.Range("AE29")=Selection.Cells(r, 9)'VCO2
+  ActiveSheet.Range("AF29")=Selection.Cells(r, 10)'VE/VCO2
+  ActiveSheet.Range("AG29")=Selection.Cells(r, 11)'RER
+  ActiveSheet.Range("AH29")=Selection.Cells(r, 12)'PETO2
+  ActiveSheet.Range("AI29")=Selection.Cells(r, 13)'PETCO2
+  ActiveSheet.Range("AJ29")=Selection.Cells(r, 14)'O2PULSE
+  ActiveSheet.Range("AK29")=Selection.Cells(r, 15)'HR
+  ActiveSheet.Range("AL29")=Selection.Cells(r, 16)'Bf
 
-  ActiveSheet.Range("W30") = "AT"
-  ActiveSheet.Range("X30") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
-  ActiveSheet.Range("Y30") = WorksheetFunction.Average(Selection.Columns(3)) 'LOAD
-  ActiveSheet.Range("Z30") = WorksheetFunction.Average(Selection.Columns(4)) 'VE
-  ActiveSheet.Range("AA30") = WorksheetFunction.Average(Selection.Columns(5)) 'VT
-  ActiveSheet.Range("AB30") = WorksheetFunction.Average(Selection.Columns(6)) 'VO2
-  ActiveSheet.Range("AC30") = WorksheetFunction.Average(Selection.Columns(7)) 'VO2/KG
-  ActiveSheet.Range("AD30") = WorksheetFunction.Average(Selection.Columns(8)) 'VE/VO2
-  ActiveSheet.Range("AE30") = WorksheetFunction.Average(Selection.Columns(9)) 'VCO2
-  ActiveSheet.Range("AF30") = WorksheetFunction.Average(Selection.Columns(10)) 'VE/VCO2
-  ActiveSheet.Range("AG30") = WorksheetFunction.Average(Selection.Columns(11)) 'RER
-  ActiveSheet.Range("AH30") = WorksheetFunction.Average(Selection.Columns(12)) 'PETO2
-  ActiveSheet.Range("AI30") = WorksheetFunction.Average(Selection.Columns(13)) 'PETCO2
-  ActiveSheet.Range("AJ30") = WorksheetFunction.Average(Selection.Columns(14)) 'O2PULSE
-  ActiveSheet.Range("AK30") = WorksheetFunction.Average(Selection.Columns(15)) 'HR
-  ActiveSheet.Range("AL30") = WorksheetFunction.Average(Selection.Columns(16)) 'Bf
+  ActiveSheet.Range("W30")="AT"
+  ActiveSheet.Range("X30")=WorksheetFunction.Average(Selection.Columns(1))'TIME
+  ActiveSheet.Range("Y30")=WorksheetFunction.Average(Selection.Columns(3))'LOAD
+  ActiveSheet.Range("Z30")=WorksheetFunction.Average(Selection.Columns(4))'VE
+  ActiveSheet.Range("AA30")=WorksheetFunction.Average(Selection.Columns(5))'VT
+  ActiveSheet.Range("AB30")=WorksheetFunction.Average(Selection.Columns(6))'VO2
+  ActiveSheet.Range("AC30")=WorksheetFunction.Average(Selection.Columns(7))'VO2/KG
+  ActiveSheet.Range("AD30")=WorksheetFunction.Average(Selection.Columns(8))'VE/VO2
+  ActiveSheet.Range("AE30")=WorksheetFunction.Average(Selection.Columns(9))'VCO2
+  ActiveSheet.Range("AF30")=WorksheetFunction.Average(Selection.Columns(10))'VE/VCO2
+  ActiveSheet.Range("AG30")=WorksheetFunction.Average(Selection.Columns(11))'RER
+  ActiveSheet.Range("AH30")=WorksheetFunction.Average(Selection.Columns(12))'PETO2
+  ActiveSheet.Range("AI30")=WorksheetFunction.Average(Selection.Columns(13))'PETCO2
+  ActiveSheet.Range("AJ30")=WorksheetFunction.Average(Selection.Columns(14))'O2PULSE
+  ActiveSheet.Range("AK30")=WorksheetFunction.Average(Selection.Columns(15))'HR
+  ActiveSheet.Range("AL30")=WorksheetFunction.Average(Selection.Columns(16))'Bf
 End Sub
@@
 Sub ValuesPeak()
-  ActiveSheet.Range("W31") = "Peak"
-  ActiveSheet.Range("X31") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
-  ActiveSheet.Range("Y31") = WorksheetFunction.Max(Selection.Columns(3)) 'LOAD
-  ActiveSheet.Range("Z31") = WorksheetFunction.Max(Selection.Columns(4)) 'VE
-  ActiveSheet.Range("AA31") = WorksheetFunction.Max(Selection.Columns(5)) 'VT
-  ActiveSheet.Range("AB31") = WorksheetFunction.Max(Selection.Columns(6)) 'VO2
-  ActiveSheet.Range("AC31") = WorksheetFunction.Max(Selection.Columns(7)) 'VO2/KG
-  ActiveSheet.Range("AD31") = WorksheetFunction.Max(Selection.Columns(8)) 'VE/VO2
-  ActiveSheet.Range("AE31") = WorksheetFunction.Max(Selection.Columns(9)) 'VCO2
-  ActiveSheet.Range("AF31") = WorksheetFunction.Max(Selection.Columns(10)) 'VE/VCO2
-  ActiveSheet.Range("AG31") = WorksheetFunction.Max(Selection.Columns(11)) 'RER
-  ActiveSheet.Range("AH31") = WorksheetFunction.Max(Selection.Columns(12)) 'PETO2
-  ActiveSheet.Range("AI31") = WorksheetFunction.Max(Selection.Columns(13)) 'PETCO2
-  ActiveSheet.Range("AJ31") = WorksheetFunction.Max(Selection.Columns(14)) 'O2PULSE
-  ActiveSheet.Range("AK31") = WorksheetFunction.Max(Selection.Columns(15)) 'HR
-  ActiveSheet.Range("AL31") = WorksheetFunction.Max(Selection.Columns(16)) 'Bf
+  ActiveSheet.Range("W31")="Peak"
+  ActiveSheet.Range("X31")=WorksheetFunction.Average(Selection.Columns(1))'TIME
+  ActiveSheet.Range("Y31")=WorksheetFunction.Max(Selection.Columns(3))'LOAD
+  ActiveSheet.Range("Z31")=WorksheetFunction.Max(Selection.Columns(4))'VE
+  ActiveSheet.Range("AA31")=WorksheetFunction.Max(Selection.Columns(5))'VT
+  ActiveSheet.Range("AB31")=WorksheetFunction.Max(Selection.Columns(6))'VO2
+  ActiveSheet.Range("AC31")=WorksheetFunction.Max(Selection.Columns(7))'VO2/KG
+  ActiveSheet.Range("AD31")=WorksheetFunction.Max(Selection.Columns(8))'VE/VO2
+  ActiveSheet.Range("AE31")=WorksheetFunction.Max(Selection.Columns(9))'VCO2
+  ActiveSheet.Range("AF31")=WorksheetFunction.Max(Selection.Columns(10))'VE/VCO2
+  ActiveSheet.Range("AG31")=WorksheetFunction.Max(Selection.Columns(11))'RER
+  ActiveSheet.Range("AH31")=WorksheetFunction.Max(Selection.Columns(12))'PETO2
+  ActiveSheet.Range("AI31")=WorksheetFunction.Max(Selection.Columns(13))'PETCO2
+  ActiveSheet.Range("AJ31")=WorksheetFunction.Max(Selection.Columns(14))'O2PULSE
+  ActiveSheet.Range("AK31")=WorksheetFunction.Max(Selection.Columns(15))'HR
+  ActiveSheet.Range("AL31")=WorksheetFunction.Max(Selection.Columns(16))'Bf
 End Sub
@@
 Sub ValuesOther()
-  ActiveSheet.Range("W32") = "Other"
-  ActiveSheet.Range("X32") = WorksheetFunction.Average(Selection.Columns(1)) 'TIME
-  ActiveSheet.Range("Y32") = WorksheetFunction.Average(Selection.Columns(3)) 'LOAD
-  ActiveSheet.Range("Z32") = WorksheetFunction.Average(Selection.Columns(4)) 'VE
-  ActiveSheet.Range("AA32") = WorksheetFunction.Average(Selection.Columns(5)) 'VT
-  ActiveSheet.Range("AB32") = WorksheetFunction.Average(Selection.Columns(6)) 'VO2
-  ActiveSheet.Range("AC32") = WorksheetFunction.Average(Selection.Columns(7)) 'VO2/KG
-  ActiveSheet.Range("AD32") = WorksheetFunction.Average(Selection.Columns(8)) 'VE/VO2
-  ActiveSheet.Range("AE32") = WorksheetFunction.Average(Selection.Columns(9)) 'VCO2
-  ActiveSheet.Range("AF32") = WorksheetFunction.Average(Selection.Columns(10)) 'VE/VCO2
-  ActiveSheet.Range("AG32") = WorksheetFunction.Average(Selection.Columns(11)) 'RER
-  ActiveSheet.Range("AH32") = WorksheetFunction.Average(Selection.Columns(12)) 'PETO2
-  ActiveSheet.Range("AI32") = WorksheetFunction.Average(Selection.Columns(13)) 'PETCO2
-  ActiveSheet.Range("AJ32") = WorksheetFunction.Average(Selection.Columns(14)) 'O2PULSE
-  ActiveSheet.Range("AK32") = WorksheetFunction.Average(Selection.Columns(15)) 'HR
-  ActiveSheet.Range("AL32") = WorksheetFunction.Average(Selection.Columns(16)) 'Bf
+  ActiveSheet.Range("W32")="Other"
+  ActiveSheet.Range("X32")=WorksheetFunction.Average(Selection.Columns(1))'TIME
+  ActiveSheet.Range("Y32")=WorksheetFunction.Average(Selection.Columns(3))'LOAD
+  ActiveSheet.Range("Z32")=WorksheetFunction.Average(Selection.Columns(4))'VE
+  ActiveSheet.Range("AA32")=WorksheetFunction.Average(Selection.Columns(5))'VT
+  ActiveSheet.Range("AB32")=WorksheetFunction.Average(Selection.Columns(6))'VO2
+  ActiveSheet.Range("AC32")=WorksheetFunction.Average(Selection.Columns(7))'VO2/KG
+  ActiveSheet.Range("AD32")=WorksheetFunction.Average(Selection.Columns(8))'VE/VO2
+  ActiveSheet.Range("AE32")=WorksheetFunction.Average(Selection.Columns(9))'VCO2
+  ActiveSheet.Range("AF32")=WorksheetFunction.Average(Selection.Columns(10))'VE/VCO2
+  ActiveSheet.Range("AG32")=WorksheetFunction.Average(Selection.Columns(11))'RER
+  ActiveSheet.Range("AH32")=WorksheetFunction.Average(Selection.Columns(12))'PETO2
+  ActiveSheet.Range("AI32")=WorksheetFunction.Average(Selection.Columns(13))'PETCO2
+  ActiveSheet.Range("AJ32")=WorksheetFunction.Average(Selection.Columns(14))'O2PULSE
+  ActiveSheet.Range("AK32")=WorksheetFunction.Average(Selection.Columns(15))'HR
+  ActiveSheet.Range("AL32")=WorksheetFunction.Average(Selection.Columns(16))'Bf
 End Sub
@@
       Dim fdrng As Range
-      Set rng = ws.Range("C:C")
-      Set fdrng = rng.Find(30, LookIn:=xlValues)
+      Set rng=ws.Range("C:C")
+      Set fdrng=rng.Find(30, LookIn:=xlValues)
       If Not fdrng Is Nothing Then
-        topRow = fdrng(0, 1).Row
-        bottomRow = fdrng(2, 0).Row
-        Set rng = ws.Range(topRow & ":" & bottomRow)
+        topRow=fdrng(0, 1).Row
+        bottomRow=fdrng(2, 0).Row
+        Set rng=ws.Range(topRow & ":" & bottomRow)
         rng.Select
@@
       Dim maxVO2
-      maxVO2 = WorksheetFunction.Max(ws.Range("G:G"))
-      Set rng = ws.Range("G:G")
-      Set fdrng = rng.Find(maxVO2)
+      maxVO2=WorksheetFunction.Max(ws.Range("G:G"))
+      Set rng=ws.Range("G:G")
+      Set fdrng=rng.Find(maxVO2)
         If Not fdrng Is Nothing Then
-          topRow = fdrng(-1, 1).Row
-          bottomRow = fdrng(3, 0).Row
-        If ws.Range("W31").Value = "" Then
-          Set rng = ws.Range(topRow & ":" & bottomRow)
+          topRow=fdrng(-1, 1).Row
+          bottomRow=fdrng(3, 0).Row
+        If ws.Range("W31").Value="" Then
+          Set rng=ws.Range(topRow & ":" & bottomRow)
           rng.Select
@@
       (Left:=800, Top:=725, Width:=400, Height:=225)
-      .chart.ChartType = xlLine
+      .chart.ChartType=xlLine
       .chart.SetSourceData Source:=ws.Range("A:A,H:H,J:J") _
@@
 
-      .chart.HasAxis(xlCategory, xlPrimary) = True
-      .chart.HasAxis(xlValue, xlPrimary) = True
+      .chart.HasAxis(xlCategory, xlPrimary)=True
+      .chart.HasAxis(xlValue, xlPrimary)=True
 
-      .chart.Axes(xlCategory, xlPrimary).CategoryType = xlAutomatic
+      .chart.Axes(xlCategory, xlPrimary).CategoryType=xlAutomatic
     End With
@@
   Dim x As Integer
-  For x = 3 To 99
+  For x=3 To 99
     Dim sheetname As String
     Sheets("aaa_main").Activate
-    sheetname = ActiveSheet.Cells(x, 1)
+    sheetname=ActiveSheet.Cells(x, 1)
     ActiveSheet.Cells(x, 32).Select
@@
   Dim x As Integer
-  For x = 3 To 99
+  For x=3 To 99
     Dim sheetname As String
     Sheets("aaa_main").Activate
-    sheetname = ActiveSheet.Cells(x, 1)
+    sheetname=ActiveSheet.Cells(x, 1)
     ActiveSheet.Cells(x, 79).Select
@@
   Dim sheetname As String
-  For x = 3 To 99
+  For x=3 To 99
     Sheets("aaa_main").Activate
-    sheetname = ActiveSheet.Cells(x, 1)
+    sheetname=ActiveSheet.Cells(x, 1)
     Sheets(sheetname).Select
-    y = 15
-    While ActiveSheet.Cells(y, 3) = "-"
-      y = y + 1
+    y=15
+    While ActiveSheet.Cells(y, 3)="-"
+      y=y + 1
     Wend
-    starttime = ActiveSheet.Cells(y, 1)
-    y = y + 5
+    starttime=ActiveSheet.Cells(y, 1)
+    y=y + 5
     While ActiveSheet.Cells(y, 3) > "0"
-      y = y + 1
+      y=y + 1
     Wend
-    endtime = ActiveSheet.Cells(y - 1, 1)
+    endtime=ActiveSheet.Cells(y - 1, 1)
     Sheets("aaa_main").Activate
-    ActiveSheet.Cells(x, 77) = starttime
-    ActiveSheet.Cells(x, 78) = endtime
+    ActiveSheet.Cells(x, 77)=starttime
+    ActiveSheet.Cells(x, 78)=endtime
   Next x

### 2015-08-27 Appendices/AppendixAccessDatabase.tex
Off to the printers now - Single sided - Black and white - 3 copies - soft comb binding

@@
 
-The following VBA code was written by me to facilitate collection of laboratory data from the existing user interface that did not allow easy automation.
+The following VBA code was written to facilitate collection of laboratory data from the existing user interface that did not allow easy automation.
 It forms part of a Microsoft Access database.
@@
 This was used to collect and parse laboratory results for preoperative and postoperative blood results efficiently and without error.
-Where necessary, I have used information from the MSDN knowledgebase and internet fora to supplement my knowledge of coding.
 

### 2015-08-27 Appendices/AppendixExcelMacros.tex
Off to the printers now - Single sided - Black and white - 3 copies - soft comb binding

@@
 \setstretch{1.0}
-The following VBA code was written by me to facilitate detailed analysis of large volumes of raw CPET data that would not have been otherwise available from the final report generated for clinical use.
-Where necessary, I have used information from the MSDN knowledgebase and internet fora to supplement my knowledge of coding.
+The following VBA code was written to facilitate detailed analysis of large volumes of raw CPET data that would not have been otherwise available from the final report generated for clinical use.
+This allowed analysis of all parameters measured at rest, anaerobic threshold and peak exercise rather than just $\dot{V}_{O_2}$.
 
@@
    Dim strName() As String
-
    For Each ws In Worksheets
@@
    Next ws
-
    Dim N As Integer
@@
    Dim SortDescending As Boolean
-
    SortDescending=False
-
    If ActiveWindow.SelectedSheets.Count=1 Then
@@
        Next N 
-
        FirstWSToSort=.Item(1).Index
@@
    End If
-
    For M=FirstWSToSort To LastWSToSort
@@
    Next M
-
    Dim x As Integer
@@
 End Sub
-
 Sub GotoSheet()
@@
 End Sub
-
 Sub VCO2VO2()
@@
 End Sub
-
 Sub ClearColumn()
@@
 End Sub
-
 Sub DeleteAllCharts()
@@
 End Sub
-
 Sub AddRCValues()
@@
 End Sub
-
 Sub CopyLineForSPSS()

### 2016-02-10 Presentations/viva_presentation.tex
Viva Presentation - Introduction slides

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%----------------------------------------------------------------------------------------

\documentclass{beamer}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.

%\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

% As well as themes, the Beamer class has a number of color themes
% for any slide theme. Uncomment each of these in turn to see how it
% changes the colors of your current slide theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%\setbeamertemplate{footline} % To remove the footer line in all slides uncomment this line
%\setbeamertemplate{footline}[page number] % To replace the footer line in all slides with a simple slide count uncomment this line

%\setbeamertemplate{navigation symbols}{} % To remove the navigation symbols from the bottom of all slides uncomment this line
}

\usepackage{graphicx} % Allows including images
\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title[Improving Outcomes in Pancreatic Surgery]{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.} % The short title appears at the bottom of every slide, the full title is only on the title page

\author{Vishnu V Chandrabalan} % Your name
\institute[UoG] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
University of Glasgow \\ % Your institution for the title page
\medskip
% \textit{john@smith.com} % Your email address
}
\date{\today} % Date, can be changed to a custom date

\begin{document}

\begin{frame}
\titlepage % Print the title page as the first slide
\end{frame}

\begin{frame}
\frametitle{Overview} % Table of contents slide, comment this block out to remove it
\tableofcontents % Throughout your presentation, if you choose to use \section{} and \subsection{} commands, these will automatically be printed on this slide as an overview of your presentation
\end{frame}

%----------------------------------------------------------------------------------------
%	PRESENTATION SLIDES
%----------------------------------------------------------------------------------------

%------------------------------------------------
\section{Pancreatic cancer}
\begin{frame}
	\frametitle{Pancreatic cancer} %Epidemiology/outcomes
	\begin{itemize}
		\item 10th most common cancer
		\item 5th most common cause of cancer death
		\item 1-year survival	: 20.8\%
		\item 5-year survival	: 3.3\%
		\item 10-year survival	: 1.1\%	
		\medskip
		\item Operable disease : ~ 15\%
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{Management of Pancreatic Cancer} 
	\begin{itemize}
		\item Surgery is the primary modality of cure
		\item Adjuvant chemotherapy has an important role in improving survival
		\item Pancreaticoduodenectomy (Whipple's procedure)
		\begin{itemize}
			\item Complex surgery with 50\% morbidity
			\item Patient fitness as important as tumour factors
		\end{itemize}
	\end{itemize}

\end{frame}

\begin{frame}
	\frametitle{Outcomes after major pancreatic surgery}

\end{frame}

%------------------------------------------------

\subsection{Subsection Example} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks

\begin{frame}
\frametitle{Paragraphs of Text}
Sed iaculis dapibus gravida. Morbi sed tortor erat, nec interdum arcu. Sed id lorem lectus. Quisque viverra augue id sem ornare non aliquam nibh tristique. Aenean in ligula nisl. Nulla sed tellus ipsum. Donec vestibulum ligula non lorem vulputate fermentum accumsan neque mollis.\\~\\

Sed diam enim, sagittis nec condimentum sit amet, ullamcorper sit amet libero. Aliquam vel dui orci, a porta odio. Nullam id suscipit ipsum. Aenean lobortis commodo sem, ut commodo leo gravida vitae. Pellentesque vehicula ante iaculis arcu pretium rutrum eget sit amet purus. Integer ornare nulla quis neque ultrices lobortis. Vestibulum ultrices tincidunt libero, quis commodo erat ullamcorper id.
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{Bullet Points}
\begin{itemize}
\item Lorem ipsum dolor sit amet, consectetur adipiscing elit
\item Aliquam blandit faucibus nisi, sit amet dapibus enim tempus eu
\item Nulla commodo, erat quis gravida posuere, elit lacus lobortis est, quis porttitor odio mauris at libero
\item Nam cursus est eget velit posuere pellentesque
\item Vestibulum faucibus velit a augue condimentum quis convallis nulla gravida
\end{itemize}
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{Blocks of Highlighted Text}
\begin{block}{Block 1}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue. Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan dolor.
\end{block}

\begin{block}{Block 2}
Pellentesque sed tellus purus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis magna at risus dictum tempor eu vitae velit.
\end{block}

\begin{block}{Block 3}
Suspendisse tincidunt sagittis gravida. Curabitur condimentum, enim sed venenatis rutrum, ipsum neque consectetur orci, sed blandit justo nisi ac lacus.
\end{block}
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{Multiple Columns}
\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment

\column{.45\textwidth} % Left column and width
\textbf{Heading}
\begin{enumerate}
\item Statement
\item Explanation
\item Example
\end{enumerate}

\column{.5\textwidth} % Right column and width
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue. Aliquam ut tortor mauris. Sed volutpat ante purus, quis accumsan dolor.

\end{columns}
\end{frame}

%------------------------------------------------
\section{Second Section}
%------------------------------------------------

\begin{frame}
\frametitle{Table}
\begin{table}
\begin{tabular}{l l l}
\toprule
\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
\midrule
Treatment 1 & 0.0003262 & 0.562 \\
Treatment 2 & 0.0015681 & 0.910 \\
Treatment 3 & 0.0009271 & 0.296 \\
\bottomrule
\end{tabular}
\caption{Table caption}
\end{table}
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{Theorem}
\begin{theorem}[Mass--energy equivalence]
$E = mc^2$
\end{theorem}
\end{frame}

%------------------------------------------------

\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
\frametitle{Verbatim}
\begin{example}[Theorem Slide Code]
\begin{verbatim}
\begin{frame}
\frametitle{Theorem}
\begin{theorem}[Mass--energy equivalence]
$E = mc^2$
\end{theorem}
\end{frame}\end{verbatim}
\end{example}
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{Figure}
Uncomment the code on this slide to include your own image from the same directory as the template .TeX file.
%\begin{figure}
%\includegraphics[width=0.8\linewidth]{test}
%\end{figure}
\end{frame}

%------------------------------------------------

\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
\frametitle{Citation}
An example of the \verb|\cite| command to cite within the presentation:\\~

This statement requires citation \cite{p1}.
\end{frame}

%------------------------------------------------

\begin{frame}
\frametitle{References}
\footnotesize{
\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below
\bibitem[Smith, 2012]{p1} John Smith (2012)
\newblock Title of the publication
\newblock \emph{Journal Name} 12(3), 45 -- 678.
\end{thebibliography}
}
\end{frame}

%------------------------------------------------

\begin{frame}
\Huge{\centerline{The End}}
\end{frame}

%----------------------------------------------------------------------------------------

\end{document}

### 2016-02-11 Presentations/viva_presentation.tex
Viva Presentation - Basic structure
- Set theme and color
- Create basic framework slides for the presentation

@@
 %\usetheme{Bergen}
-%\usetheme{Berkeley}
+%\usetheme{Berkeley} % Neat Contents on left
 %\usetheme{Berlin}
 %\usetheme{Boadilla}
-%\usetheme{CambridgeUS}
+%\usetheme{CambridgeUS} % Neat but breadcrumbs at top
 %\usetheme{Copenhagen}
@@
 %\usetheme{Frankfurt}
-%\usetheme{Goettingen}
-%\usetheme{Hannover}
+%\usetheme{Goettingen} %Neat contents on right
+\usetheme{Hannover} %Neat contents on left
 %\usetheme{Ilmenau}
@@
 %\usetheme{Luebeck}
-\usetheme{Madrid}
+%\usetheme{Madrid}
 %\usetheme{Malmoe}
-%\usetheme{Marburg}
+%\usetheme{Marburg} % Contents on right in contrast color
 %\usetheme{Montpellier}
-%\usetheme{PaloAlto}
+%\usetheme{PaloAlto} % Contents on left
 %\usetheme{Pittsburgh}
@@
 %\usecolortheme{seagull}
-%\usecolortheme{seahorse}
+\usecolortheme{seahorse}
 %\usecolortheme{whale}
@@
 %------------------------------------------------
-\section{Pancreatic cancer}
+\section{Introduction}
+\subsection{Pancreatic cancer}
 \begin{frame}
 	\frametitle{Pancreatic cancer} %Epidemiology/outcomes
+	\begin{block}{The Problem}
 		\begin{itemize}
-		\item 10th most common cancer
-		\item 5th most common cause of cancer death
-		\item 1-year survival	: 20.8\%
-		\item 5-year survival	: 3.3\%
-		\item 10-year survival	: 1.1\%	
-		\medskip
-		\item Operable disease : ~ 15\%
+			\item 10th common cancer but 5th common cause of cancer death
+			\item Survival	1-yr: 20.8\%, 5-yr: 3.3\%, 10yr: 1.1\%
+			\item Operable disease : approx 15\%
 		\end{itemize}
+	\end{block}
+	\begin{block}{Treatment}
+		\begin{itemize}
+			\item Surgery is the primary modality of cure
+			\item Adjuvant chemotherapy has an important role in improving survival
+		\end{itemize}
+	\end{block}
+	\begin{block}{Pancreaticoduodenectomy}
+		\begin{itemize}
+			\item Complex surgery with 50\% morbidity
+			\item Performed only in specialist centres
+			\item Patient fitness is as important as tumour factors
+		\end{itemize}
+	\end{block}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Management of Pancreatic Cancer} 
-	\begin{itemize}
-		\item Surgery is the primary modality of cure
-		\item Adjuvant chemotherapy has an important role in improving survival
-		\item Pancreaticoduodenectomy (Whipple's procedure)
-		\begin{itemize}
-			\item Complex surgery with 50\% morbidity
-			\item Patient fitness as important as tumour factors
-		\end{itemize}
-	\end{itemize}
+	\frametitle{Preoperative Risk Stratification}
 
@@
 
+\subsection{CPET}
+
 \begin{frame}
-	\frametitle{Outcomes after major pancreatic surgery}
-
+	\frametitle{Basis of CPET - Cellular Respiration} 
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.7\linewidth]{CellRespiration}
+		\caption[Cellular Respiration]{Cellular Respiration}
+		\label{fig:CellRespiration}
+	\end{figure}
 \end{frame}
 
-%------------------------------------------------
+\begin{frame}
+	\frametitle{CPET and Body Composition}
+\end{frame}
 
-\subsection{Subsection Example} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
+\subsection{Jaundice}
 
 \begin{frame}
-\frametitle{Paragraphs of Text}
-Sed iaculis dapibus gravida.
-Morbi sed tortor erat, nec interdum arcu.
-Sed id lorem lectus.
-Quisque viverra augue id sem ornare non aliquam nibh tristique.
-Aenean in ligula nisl.
-Nulla sed tellus ipsum.
-Donec vestibulum ligula non lorem vulputate fermentum accumsan neque mollis.\\~\\
-
-Sed diam enim, sagittis nec condimentum sit amet, ullamcorper sit amet libero.
-Aliquam vel dui orci, a porta odio.
-Nullam id suscipit ipsum.
-Aenean lobortis commodo sem, ut commodo leo gravida vitae.
-Pellentesque vehicula ante iaculis arcu pretium rutrum eget sit amet purus.
-Integer ornare nulla quis neque ultrices lobortis.
-Vestibulum ultrices tincidunt libero, quis commodo erat ullamcorper id.
+	\frametitle{The Jaundiced Patient}
 \end{frame}
 
-%------------------------------------------------
+\subsection{Inflammation}
+
+\section{Results}
+\subsection{CPET vs Outcomes}
+\begin{frame}
+	\frametitle{CPET and Postoperative Outcomes}
+\end{frame}
+
+\subsection{CPET vs Jaundice}
+\begin{frame}
+	\frametitle{CPET and Jaundice}
+\end{frame}
+
+\subsection{CPET vs Body Composition}
+\begin{frame}
+	\frametitle{CPET and Body Composition}
+\end{frame}
+
+\subsection{Systemic inflammation}
 
 \begin{frame}
-\frametitle{Bullet Points}
-\begin{itemize}
-\item Lorem ipsum dolor sit amet, consectetur adipiscing elit
-\item Aliquam blandit faucibus nisi, sit amet dapibus enim tempus eu
-\item Nulla commodo, erat quis gravida posuere, elit lacus lobortis est, quis porttitor odio mauris at libero
-\item Nam cursus est eget velit posuere pellentesque
-\item Vestibulum faucibus velit a augue condimentum quis convallis nulla gravida
-\end{itemize}
-\end{frame}
-
-%------------------------------------------------
-
-\begin{frame}
-\frametitle{Blocks of Highlighted Text}
-\begin{block}{Block 1}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue.
-Aliquam ut tortor mauris.
-Sed volutpat ante purus, quis accumsan dolor.
-\end{block}
-
-\begin{block}{Block 2}
-Pellentesque sed tellus purus. Class aptent taciti sociosqu ad litora torquent per conubia nostra, per inceptos himenaeos. Vestibulum quis magna at risus dictum tempor eu vitae velit.
-\end{block}
-
-\begin{block}{Block 3}
-Suspendisse tincidunt sagittis gravida. Curabitur condimentum, enim sed venenatis rutrum, ipsum neque consectetur orci, sed blandit justo nisi ac lacus.
+	\frametitle{Preop Pathophysiology and Postoperative Systemic Inflammation}
+	\begin{block}{title}
+		Preoperative inflammatory status of the patient and obstructive jaundice play an important role in modulating postoperative systemic inflammation which may not simply be due to the effect of surgery and its sequelae.
 	\end{block}
@@
 
-%------------------------------------------------
+\begin{frame}
+	\frametitle{Postoperative Systemic Inflammation and Outcomes}
+\end{frame}
 
@@
 
-%------------------------------------------------
-\section{Second Section}
-%------------------------------------------------
-
 \begin{frame}
@@
 \begin{frame}
-\frametitle{Theorem}
+\frametitle{Oxygen Delivery}
 \begin{theorem}[Mass--energy equivalence]

### 2016-02-13 Presentations/viva_presentation.tex
Viva Presentation - More changes, more slides

@@
 		\begin{itemize}
-			\item 10th common cancer but 5th common cause of cancer death
+			\item 10th common but 5th common cause of cancer death
 			\item Survival	1-yr: 20.8\%, 5-yr: 3.3\%, 10yr: 1.1\%
@@
 		\begin{itemize}
-			\item Surgery is the primary modality of cure
-			\item Adjuvant chemotherapy has an important role in improving survival
+			\item Surgery + adjuvant chemotherapy only chance of cure
 		\end{itemize}
@@
 \begin{frame}
-	\frametitle{Preoperative Risk Stratification}
+	\frametitle{Pancreaticoduodenectomy}
+	\begin{columns}[t]
+		\column{0.45\textwidth}
+		\begin{figure}
+			\centering
+			\includegraphics[width=0.8\linewidth]{whipple_schematic_pre}
+			\label{fig:whipple_schematic_pre}
+		\end{figure}
+		{\scriptsize
+			\textbf{Structures removed:}
+			\begin{itemize}
+			\item Head of pancreas
+			\item Distal stomach, duodenum, proximal jejunum
+			\item  Common bile duct, gallbladder
+			\item Lymphadenectomy
+			\item Venous resection in some
+		\end{itemize}}
+
+		\column{0.45\textwidth}
+		\begin{figure}
+			\centering
+			\includegraphics[width=0.8\linewidth]{whipple_schematic_post}
+			\label{fig:whipple_schematic_post}
+		\end{figure}
+		{\scriptsize
+		\textbf{Reconstruction:}
+		\begin{itemize}
+			\item Pancreaticojejunostomy (25\% leak rate)
+			\item Hepaticojejunostomy
+			\item Gastrojejunostomy
+		\end{itemize}
+		}
+	\end{columns}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Patient selection}
 
@@
 \begin{frame}
-	\frametitle{Basis of CPET - Cellular Respiration} 
+	\frametitle{Respiratory compensation of lactic acidosis} 
 	\begin{figure}
@@
 		\includegraphics[width=0.7\linewidth]{CellRespiration}
-		\caption[Cellular Respiration]{Cellular Respiration}
 		\label{fig:CellRespiration}
+	\end{figure}
+	\centering
+	$H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2$
+
+\end{frame}
+
+\begin{frame}
+	\frametitle{Determination of Anaerobic Threshold}
+	\begin{figure}[htbp]
+		\centering
+		\includegraphics[width=0.7\linewidth]{../Figures/cpet_vslope}
+		%\caption{Determination of $\dot{V}_{O_2}$AT by the V-slope method.}
+		\label{fig:cpet_vslope}
 	\end{figure}
@@
 \begin{frame}
-	\frametitle{CPET and Body Composition}
+	\frametitle{Role of CPET in preoperative assessment}
+	\begin{itemize}
+		\item General Surgery
+		\item Oesophago-gastric surgery
+		\item Bariatric surgery
+		\item Liver transplantation
+		\item Thoracic surgery
+		\medskip
+		\item Little evidence in pancreatic surgery in 2010
+
+	\end{itemize}
 \end{frame}
 
-\subsection{Jaundice}
+\begin{frame}
+	\frametitle{Aims of Thesis}
+	\begin{enumerate}
+		\item To evaluate the clinical utility of preoperative CPET in predicting postoperative adverse events after pancreaticoduodenectomy.
+		\item To examine the patient factors that are related to cardiopulmonary exercise physiology with particular attention to the effect of obstructive jaundice and body composition.
+		\item To examine the effect of preoperative systemic inflammation and poor aerobic capacity on the magnitude of the post-operative systemic inflammatory response.
+		\item To examine the value of serial daily postoperative markers of systemic inflammatory response in the prediction of post-operative complications.
+	\end{enumerate}
+\end{frame}
+
+\section{Methods}
+\begin{frame}
+	\frametitle{Patients and Methods}
+	\begin{enumerate}
+		\item Patients undergoing pancreaticoduodenectomy at Glasgow Royal Infirmary
+		\item August 2008 - August 2012
+		\item Prospectively maintained MS Access database
+		\item Preoperative data - Bloods, co-morbidity including POSSUM Physiology Score, stent information
+		\item Cardiopulmonary exercise testing
+		\item Prospective recording of complications using ISGPS or Clavie-Dindo classification
+		\item Serial measurements of biochemical parameters during first postoperative week
+	\end{enumerate}
+\end{frame}
+
+% Discuss the results in the context of a patient.
+% 63 yr old female with BMI of 35 presents with painless OJ
+% PMH - HTN, T2DM, smoked 10/day for 40 yrs
+% Bloods: Bil 180, CRP 22, Alb 30
+% Investigations reveal an operable tumour of the head of the pancreas
+% How would this patient have been managed 10 years ago?
+\section{Case study}
+\begin{frame}
+	\frametitle{A typical patient}
+	\begin{itemize}
+		\item 63 yr old female with BMI of 35 presents with painless OJ
+		\item PMH - HTN, T2DM, smoked 10/day for 40 yrs
+		\item Bloods: Bil 180, CRP 22, Alb 30
+		\item Imaging: Resectable tumour of head of pancreas
+		\item How would this patient have been managed 10 years ago?
+	\end{itemize}
+\end{frame}
 
 \begin{frame}
-	\frametitle{The Jaundiced Patient}
+	\frametitle{10 years ago...}
+	\begin{itemize}
+		\item Non-dynamic measures of fitness - ECG, PFTs, ASA, ECHO
+		\item Routine biliary drainage
+		\item No objective methods to predict complications other than pancreatic leak
+		\item Impact of postoperative SIRS on complications - no clinical utility
+		\item Limited ability to identifying patients at risk of complications
+		\item ERAS principles not universally applied
+		\item No prehabilitation
+	\end{itemize}
 \end{frame}
 
-\subsection{Inflammation}
+\begin{frame}
+	\frametitle{Now... }
+	\begin{itemize}
+		\item CPET in all or selected cases (evidence from other patient groups)
+		\item Routine biliary drainage no longer indicated
+		\item Enhanced recovery more widespread
+		\medskip
+		\hrule
+		\medskip
+		\item Reluctance to perform CPET in jaundiced patients
+		\item No recognised thresholds for CPET parameters
+		\item No clear thresholds for post-op inflammatory markers in predicting complications
+		\item Poor understanding of impact of obesity on preop function and postop outcomes
+		\item No prehabilitation
+	\end{itemize}
+\end{frame}
 
@@
 	\frametitle{CPET and Postoperative Outcomes}
+	\begin{block}{\textbf{Complications}}
+		Low $\dot{V}_{O_2}$AT associated with pancreatic fistula (16\% vs 35\%) \\
+		Not associated with cardio-respiratory complications or mortality
+	\end{block}
+
+	\begin{block}{\textbf{Length of stay}}
+		Low $\dot{V}_{O_2}$AT associated with longer hospital stay (14 vs 20 days)\\
+		Not associated with critical care stay or admissions
+	\end{block}
+
+	\begin{block}{\textbf{Adjuvant Therapy}}
+		Low $\dot{V}_{O_2}$AT associated with non-progression to adjuvant therapy after surgery.
+	\end{block}
+
 \end{frame}
@@
 	\frametitle{CPET and Jaundice}
+	\begin{columns}[t]
+		\column{0.5\textwidth}
+			\begin{figure}
+				\centering
+				\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_at_bil}
+				\caption{$\dot{V}_{O_2}$AT versus serum bilirubin}
+				\label{fig:cpet_oj_scatter_at_bil}
+			\end{figure}
+
+		\column{0.5\textwidth}
+			\begin{figure}
+				\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_peak_bil}
+				\caption{$\dot{V}_{O_2}$Peak versus serum bilirubin}
+				\label{fig:cpet_oj_scatter_peak_bil}
+			\end{figure}
+	\end{columns}
+
 \end{frame}
@@
 
-\begin{frame}
-\frametitle{Oxygen Delivery}
-\begin{theorem}[Mass--energy equivalence]
-$E = mc^2$
-\end{theorem}
-\end{frame}
-
 %------------------------------------------------

### 2016-02-13 Presentations/viva_presentation.tex
Viva Presentation - More work

@@
 %\usetheme{Goettingen} %Neat contents on right
-\usetheme{Hannover} %Neat contents on left
+%\usetheme{Hannover} %Neat contents on left
 %\usetheme{Ilmenau}
@@
 %\usetheme{Montpellier}
-%\usetheme{PaloAlto} % Contents on left
+\usetheme{PaloAlto} % Contents on left
 %\usetheme{Pittsburgh}
@@
 %\usecolortheme{rose}
-%\usecolortheme{seagull}
-\usecolortheme{seahorse}
+\usecolortheme{seagull}
+%\usecolortheme{seahorse}
 %\usecolortheme{whale}
@@
 \begin{frame}
-	\frametitle{Patient selection}
+	\frametitle{Complications of Surgery}
+	\framesubtitle{Section 1.2.4 Pages 8-11}
+	\begin{itemize}
+		\item Significant improvements in mortality ($<$5\% from $>$40\%)
+		\item Morbidity remains high at over 40\%
+		\item Postoperative pancreatic fistula (POPF) is serious
+		\item POPF can lead to other complications: Sepsis, PPH, DGE
+		\item International Study Group for Pancreatic Surgery
+		\item Clavien-Dindo classification of complications
+	\end{itemize}
+\end{frame}
 
+\begin{frame}
+	\frametitle{Adjuvant therapy}
+	\framesubtitle{Section 1.3 Pages 11,13 }
+	\begin{itemize}
+		\item Adjuvant therapy improves survival
+		\item Currently standard of care and subject of multiple trials
+		\item 25\% - 50\% of patients do not receive adjuvant treatment
+		\item Multiple factors including:
+		\begin{itemize}
+			\item Comorbidity
+			\item Complications of surgery
+			\item Favourable pathological features
+		\end{itemize}
+	\end{itemize}
 \end{frame}
@@
 \subsection{CPET}
+\begin{frame}
+	\frametitle{Risk Stratification }
+	\framesubtitle{Section 1.4 Pages 13-18 }
+	\begin{itemize}
+		\item Patient fitness is important determinant of outcome
+		\item Effect of comorbidity on function is difficult to assess
+		\item POSSUM, APACHE, mGPS, E-PASS, etc.
+		\item Static measures do not predict performance under stress
+		\item Do not provide targets for prehabilitation
+		\item Pancreas specific risk factors of POPF
+		\begin{itemize}
+			\item Soft pancreas
+			\item Narrow pancreatic duct
+			\item Pathology not obstructing PD
+			\item These factors are not modifiable
+			\item Coronary artery disease! (Lin 2004)
+		\end{itemize}
+	\end{itemize}
+\end{frame}
 
 \begin{frame}
-	\frametitle{Respiratory compensation of lactic acidosis} 
+	\frametitle{What is CPET?}
+	\framesubtitle{Section 1.5 Page 18}
+	\begin{itemize}
+		\item Objective measure of cardiopulmonary response to exercise
+		\item Dynamic test measuring numerous parameters
+		\item Cardiac, respiratory, vascular and mitochondrial function
+		\item First described for preoperative assessment by Older and co-workers
+		\item Widely used in general, oesophago-gastric, vascular and thoracic surgery
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{CPET Methodology}
+	\framesubtitle{Section 1.5.2 Page 20 }
+	\begin{itemize}
+		\item Dynamic exercise test
+		\item Cycle ergometer with electronic braking and incremental workload
+		\item Breath-by-breath gas analysis using a tight-fitting face mask
+		\item Continuous monitoring of 12-lead ECG
+		\item Initial 3-minute rest period on bike for baseline parameters
+		\item Incremental work-load test to volitional fatigue
+		%include picture here
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Physiological basis of CPET} 
 	\begin{figure}
 		\centering
-		\includegraphics[width=0.7\linewidth]{CellRespiration}
+		\includegraphics[width=0.5\linewidth]{CellRespiration}
 		\label{fig:CellRespiration}
 	\end{figure}
+	\begin{block}
 	\centering
 	$H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2$
+	\end{block}
 
@@
 	\frametitle{Determination of Anaerobic Threshold}
+	\framesubtitle{Section 1.5.3 Page 22}
 	\begin{figure}[htbp]
@@
 	\end{figure}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Parameters measured during CPET }
+	\framesubtitle{ Section 1.6 Page 25}
+	\begin{itemize}
+		\item
+		\item
+		\item
+	\end{itemize}
 \end{frame}
@@
 	\frametitle{Patients and Methods}
-	\begin{enumerate}
+	\begin{itemize}
 		\item Patients undergoing pancreaticoduodenectomy at Glasgow Royal Infirmary
@@
 		\item Serial measurements of biochemical parameters during first postoperative week
-	\end{enumerate}
+	\end{itemize}
 \end{frame}
@@
 \section{Results}
+
+%--CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2--
 \subsection{CPET vs Outcomes}
@@
 	\frametitle{CPET and Postoperative Outcomes}
+	\framesubtitle{Chapter 2 Page 56; (n=100)}
 	\begin{block}{\textbf{Complications}}
@@
 	\end{block}
-
 \end{frame}
 
-\subsection{CPET vs Jaundice}
+\begin{frame}
+	\frametitle{CPET and Postoperative Outcomes}
+	\framesubtitle{The Glasgow Cohort}
+	\begin{figure}
+		\centering
+		\includegraphics[height=0.35\textheight]{../Figures/cpet_outcomes_dist_of_AT}
+	\end{figure}
+
+	\begin{itemize}
+		\item Median AT = 10.3 ml/kg/min (IQR 8.8 -11.6)
+		\item $\dot{V}_{O_2}$AT$<$10 ml/kg/min in 49\% of patients
+		\item Different from other cohorts
+		\item $\dot{V}_{O_2}AT<11$ ml/kg/min in
+		\begin{itemize}
+			\item 16\% undergoing oesophageal surgery [Forshaw 2008]
+			\item 39\% undergoing liver transplantation [Epstein 2004]
+			\item 29\% undergoing abdominal surgery [Older 1993]
+		\end{itemize}
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{CPET and Postoperative Outcomes}
+	\framesubtitle{Citations}
+	\begin{columns}[c]
+	\column{0.4\textwidth}
+		\begin{figure}
+			\centering
+			\includegraphics[width=\textwidth]{cpet_citations}
+		\end{figure}
+
+	\column{0.6\textwidth}
+	Why is $\dot{V}_{O_2}$AT so low in patients undergoing pancreatic surgery?\\
+	\medskip
+	What factors are associated with a low $\dot{V}_{O_2}$AT?\\
+	\medskip
+	Why was $\dot{V}_{O_2}$AT not associated with cardio-respiratory complications?\\
+	\medskip
+	Does $\dot{V}_{O_2}$AT truely reflect cardiopulmonary function in this cohort of patients?\\	
+	\end{columns}
+\end{frame}
+
+%--CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3--
+\subsection{CPET vs Preop factors}
+
+\begin{frame}
+	\frametitle{Chapter 3}
+	\framesubtitle{CPET, Jaundice and Preoperative Pathophysiology - Page 76-96 }
+	\begin{block}{Title}
+	An investigation into the relationship between cardiopulmonary exercise testing, preoperative pathophysiology and obstructive jaundice in patients undergoing pancreaticoduodenectomy.
+	\end{block}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Distribution of Jaundice in the study cohort }
+	\framesubtitle{n=138}
+
+	\begin{figure}
+		\centering
+		\includegraphics[width=0.7\linewidth]{jaundice_distribution}
+		\caption{}
+		\label{fig:jaundice_distribution}
+	\end{figure}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Jaundice and patient characteristics}
+	\framesubtitle{Table 3.1 Page 83, Table 3.2 Page 85 }
+	\begin{figure}
+		\includegraphics[height=0.4\textheight]{jaundice_vs_patient_factors}
+	\end{figure}
+	{\scriptsize			
+	Severity of systemic inflammation proportionate to severity of jaundice. \\
+	\medskip
+
+	Obstructive jaundice associated with electrolyte abnormalities.\\
+	\medskip
+
+	Jaundiced patients were more likely to be anaemic (p$<$0.001) with lower haematocrit (p$<$0.001) and lower mean corpuscular volume (p=0.001).\\
+	\medskip
+
+	No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
+	}
+\end{frame}
+
 \begin{frame}
 	\frametitle{CPET and Jaundice}
+	\framesubtitle{Figure 3.1 Page 87}
 	\begin{columns}[t]
@@
 	\end{columns}
+\end{frame}
+
+\begin{frame}
+	\frametitle{$\dot{V}_{O_2}$ vs Patient Factors }
+	\framesubtitle{Table 3.6 Page 91 - Multivariate binary logistic regression}
+
+	Factors associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min
+	\begin{table}
+		\begin{tabular}{l l l l}
+			           & HR   & 95\% CI    & p     \\ \hline
+			Female sex & 3.75 & 1.57-8.95  & 0.003 \\
+			BMI$>$25   & 3.65 & 1.61-8.26  & 0.002 \\
+			Cancer     & 4.02 & 1.33-12.16 & 0.014 \\
+			CRP$>$10   & 2.98 & 1.29-6.86  & 0.010
+		\end{tabular}
+	\end{table}
+
+	Factors associated with $\dot{V}_{O_2}$Peak $<$ 16ml/kg/min
+	\begin{table}
+		\begin{tabular}{l l l l}
+			            & HR   & 95\% CI   & p        \\ \hline
+			Female sex  & 7.57 & 3.09-18.5 & $<$0.001 \\
+			BMI$>$25    & 2.57 & 1.18-5.63 & 0.018    \\
+			Haemoglobin & 2.43 & 1.05-5.63 & 0.038
+		\end{tabular}
+	\end{table}
+\end{frame}
+
+%--CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4--
+\subsection{Chapter 4}
+
+\begin{frame}
+	\frametitle{Chapter 4}
+	\framesubtitle{CPET and Body Composition - Page 97-122 }
+	\begin{block}{Title}
+		An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing pancreaticoduodenectomy.
+	\end{block}
+\end{frame}
+
+\begin{frame}
+	\frametitle{What causes a low $\dot{V}_{O_2}$? }
+	\framesubtitle{Section 4.1.2 Page 98-99 }
+	\begin{itemize}
+		\item Low $\dot{V}_{O_2}$ has been universally ascribed to low aerobic fitness due to inadequate cardio-pulmonary reserve
+		\item Occassionally anaemia, peripheral vascular disease or even mitochondrial diseases may play a role
+		\item $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak are most commonly used parameters
+		\item Both are normalised by dividing using patient's total body weight
+		\item But, during exercise $\dot{V}_{O_2}$ is a function of skeletal muscle mass
+		\item This can result in 'spurious correlation' unfairly penalising obese subjects
+		\item Most patients in this cohort did not have overt cardiac/respiratory disease that could explain the low $\dot{V}_{O_2}$.
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Corrected $\dot{V}_{O_2}$ vs. Absolute $\dot{V}_{O_2}$}
+	\framesubtitle{Could this be a source of spurious correlation?}
+\[Corrected\ \dot{V}_{O_2} (ml/kg/min) = \frac{Absolute\ \dot{V}_{O_2}\ (litres/min) * 1000}{Total\ body\ weight\ (kg)}\]
+\end{frame}
+
+\begin{frame}
+	\frametitle{Body composition estimation}
+	\framesubtitle{Figure 4.1 Page 103}
+	\begin{figure}
+		\centering
+		\includegraphics[width=\linewidth]{bc_gimp}
+		\label{fig:bc_gimp}
+	\end{figure}
+\end{frame}
+
+\begin{frame}
+	\frametitle{$\dot{V}_{O_2}$AT vs Skeletal Muscle}
+	\framesubtitle{Figure 4.4(A) Page 111, Figure 4.5(A) on Page 114}
+
+	\begin{figure}
+		\includegraphics[height=0.4\textheight]{../Figures/bc_scatter_VO2_skeletal}
+	\end{figure}
+
+	\begin{figure}
+		\includegraphics[height=0.45\textheight]{../Figures/bc_scatter_VO2_skeletal_elbm}
+	\end{figure}
 
@@
 
-\subsection{CPET vs Body Composition}
 \begin{frame}
-	\frametitle{CPET and Body Composition}
+	\frametitle{$\dot{V}_{O_2}$AT vs Total Adipose Tissue}
+	\framesubtitle{Figure 4.4(B) Page 111, Figure 4.5(B) on Page 114}
+
+	\begin{figure}
+		\includegraphics[height=0.4\textheight]{../Figures/bc_scatter_VO2_TAT}
+	\end{figure}
+	\begin{figure}
+		\includegraphics[height=0.45\textheight]{../Figures/bc_scatter_VO2_TAT_elbm}
+	\end{figure}
+
 \end{frame}

### 2016-02-13 Presentations/viva_presentation.tex
Viva presentation - More changes

@@
 
-\documentclass{beamer}
+\documentclass[10pt]{beamer}
 
@@
 
-\section{Results}
-
 %--CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2--
-\subsection{CPET vs Outcomes}
+\section{Chapter 2}
 \begin{frame}
@@
 %--CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3--
-\subsection{CPET vs Preop factors}
+\section{Chapter 3}
 
@@
 %--CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4--
-\subsection{Chapter 4}
+\section{Chapter 4}
 
@@
 	\[Corrected\ \dot{V}_{O_2} (ml/kg/min) = \frac{Absolute\ \dot{V}_{O_2}\ (litres/min) * 1000}{Total\ body\ weight\ (kg)}\]
+
 \end{frame}
@@
 
-\subsection{Systemic inflammation}
+\section{Chapter 5}
 
 \begin{frame}
-	\frametitle{Preop Pathophysiology and Postoperative Systemic Inflammation}
-	\begin{block}{title}
-		Preoperative inflammatory status of the patient and obstructive jaundice play an important role in modulating postoperative systemic inflammation which may not simply be due to the effect of surgery and its sequelae.
+	\frametitle{Chapter 5}
+	\begin{block}{Title}
+		An investigation into the relationship between preoperative clinico-pathological characteristics and post-operative systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
 	\end{block}
@@
 \begin{frame}
-	\frametitle{Postoperative Systemic Inflammation and Outcomes}
+	\frametitle{Background}
+	\framesubtitle{Section 5.1 Page 124}
+	\begin{itemize}
+		\item Perioperative systemic inflammation influences short and long-term outcomes
+		\item Elevated preoperative systemic inflammation is associated with postoperative complications in colorectal surgery, after oesophagectomy, liver resections and cardiac surgery. 
+		\item Early exaggerated postoperative systemic inflammatory response is also associated with increased infective complications
+		\item In patients undergoing pancreatic surgery, preoperative inflammatory state is further influenced by:
+		\begin{itemize}
+			\item Obstructive jaundice
+			\item Cholangitis
+			\item ERCP and stenting
+			\item Post-ERCP / EUS related complications
+		\end{itemize}
+		\item Obesity is associated with chronic inflammation
+		\item Impact of aerobic capacity on postoperative inflammation not studied before
+	\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Multiple Columns}
-\begin{columns}[c] % The "c" option specifies centered vertical alignment while the "t" option is used for top vertical alignment
-
-\column{.45\textwidth} % Left column and width
-\textbf{Heading}
-\begin{enumerate}
-\item Statement
-\item Explanation
-\item Example
-\end{enumerate}
-
-\column{.5\textwidth} % Right column and width
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Integer lectus nisl, ultricies in feugiat rutrum, porttitor sit amet augue.
-Aliquam ut tortor mauris.
-Sed volutpat ante purus, quis accumsan dolor.
-
-\end{columns}
+	\frametitle{Preop inflammation vs. Postop CRP}
+	\framesubtitle{Figure 5.3 Page 142}
+	\begin{figure}
+		\includegraphics[width=\textwidth]{../Figures/sirs_crp_crp_alb}
+	\end{figure}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Table}
-\begin{table}
-\begin{tabular}{l l l}
-\toprule
-\textbf{Treatments} & \textbf{Response 1} & \textbf{Response 2}\\
-\midrule
-Treatment 1 & 0.0003262 & 0.562 \\
-Treatment 2 & 0.0015681 & 0.910 \\
-Treatment 3 & 0.0009271 & 0.296 \\
-\bottomrule
-\end{tabular}
-\caption{Table caption}
-\end{table}
+	\frametitle{Obstructive jaundice vs. Postop CRP}
+	\framesubtitle{ }
+	\begin{figure}
+		\includegraphics[width=0.7\textwidth]{../Figures/sirs_bil_crp}
+	\end{figure}
+
 \end{frame}
 
-%------------------------------------------------
+\begin{frame}
+	\frametitle{CPET, Body composition vs. Postop Inflammation }
+	\framesubtitle{Section 5.3.4, 5.3.5 Pages 132-133 }
+	\begin{itemize}
+		\item No relationship between $\dot{V}_{O_2}$ and post op. CRP or neutrophil count
+		\item Serum albumin levels lower in patients with low VO2
+		\begin{itemize}
+			\item Probably related to preop hypoalbuminemia
+			\item This in turn related to low skeletal muscle and multiple other factors
+		\end{itemize}
+		\item No relationship between body composition and postoperative CRP/neutrophil counts
+	\end{itemize}
+\end{frame}
+
+\section{Chapter 6}
+\begin{frame}
+	\frametitle{Postoperative Systemic Inflammation and Outcomes}
+\end{frame}
 
@@
 \begin{frame}
-\frametitle{Figure}
-Uncomment the code on this slide to include your own image from the same directory as the template .TeX file.
-%\begin{figure}
-%\includegraphics[width=0.8\linewidth]{test}
-%\end{figure}
-\end{frame}
-
-%------------------------------------------------
-
-\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
-\frametitle{Citation}
-An example of the \verb|\cite| command to cite within the presentation:\\~
-
-This statement requires citation \cite{p1}.
-\end{frame}
-
-%------------------------------------------------
-
-\begin{frame}
-\frametitle{References}
-\footnotesize{
-\begin{thebibliography}{99} % Beamer does not support BibTeX so references must be inserted manually as below
-\bibitem[Smith, 2012]{p1} John Smith (2012)
-\newblock Title of the publication
-\newblock \emph{Journal Name} 12(3), 45 -- 678.
-\end{thebibliography}
-}
-\end{frame}
-
-%------------------------------------------------
-
-\begin{frame}
 \Huge{\centerline{The End}}

### 2016-02-13 Presentations/viva_presentation.tex
Viva Presentation - Nearly Complete - For next step, back this up and halve the number of slides

@@
 	\framesubtitle{Could this be a source of spurious correlation?}
+
 	\[Corrected\ \dot{V}_{O_2} (ml/kg/min) = \frac{Absolute\ \dot{V}_{O_2}\ (litres/min) * 1000}{Total\ body\ weight\ (kg)}\]
+
+	\vfill
+
+	For example, if the absolute $\dot{V}_{O_2}$AT was 0.84 litres/min, the corrected $\dot{V}_{O_2}$AT (ml/kg/min) would vary with patient weight as follows:
+
+	\begin{table}
+	\begin{tabular}{c c c c c c}
+		     Weight       & kg        & 70 & 80   & 90  & 100 \\
+		$\dot{V}_{O_2}$AT & ml/kg/min & 12 & 10.5 & 9.3 & 8.4
+	\end{tabular}
+	\end{table}
 
@@
 	\frametitle{CPET, Body composition vs. Postop Inflammation }
-	\framesubtitle{Section 5.3.4, 5.3.5 Pages 132-133 }
+	\framesubtitle{Section 5.3.4, 5.3.5 Pages 132-133, Tables 5.8-5.10 Pages 145-147 }
 	\begin{itemize}
 		\item No relationship between $\dot{V}_{O_2}$ and post op. CRP or neutrophil count
+		\vfill
 		\item Serum albumin levels lower in patients with low VO2
@@
 		\end{itemize}
+		\vfill
 		\item No relationship between body composition and postoperative CRP/neutrophil counts
@@
 \begin{frame}
-	\frametitle{Postoperative Systemic Inflammation and Outcomes}
+	\frametitle{Chapter 6}
+	\framesubtitle{Postoperative CRP and complications}
+	\begin{block}{Title}
+		An investigation into the relationship between postoperative systemic inflammation and complications after pancreaticoduodenectomy.
+	\end{block}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Does post op. CRP predict severity of POPF?}
+	\framesubtitle{No, it does not. \\ Moreover, diagnosis of POPF is made on the third day using drain amylase. }
+
+	\begin{columns}
+		\column[c]{0.45\textwidth}
+			\begin{figure}
+				\includegraphics[width=\textwidth]{../Figures/crp_comp_crp_popf_yes_no}
+			\end{figure}
+
+		\column[c]{0.45\textwidth}
+			\begin{figure}
+				\includegraphics[width=\textwidth]{../Figures/crp_comp_crp_popf_isgps}
+			\end{figure}
+	\end{columns}
+
+\end{frame}
+
+\begin{frame}
+	\frametitle{Does post op. CRP predict infective complications?}
+	\framesubtitle{Yes, it does. \\ But, only in the absence of a pancreatic fistula. }
+
+	\begin{columns}
+		\column[c]{0.45\textwidth}
+		\begin{figure}
+			\includegraphics[width=\textwidth]{../Figures/crp_comp_infective_leak0}
+			\\ Patients with NO POPF
+		\end{figure}
+
+		\column[c]{0.45\textwidth}
+		\begin{figure}
+			\includegraphics[width=\textwidth]{../Figures/crp_comp_infective_leak1}
+			\\ Patients with POPF
+		\end{figure}
+	\end{columns}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Receiver operating characteristics analysis}
+	\framesubtitle{CRP thresholds for predicting infective complications in the absence of a POPF. }
+	\begin{columns}
+		\column[c]{0.45\textwidth}
+			\begin{figure}
+				\includegraphics[width=\textwidth]{../Figures/crp_comp_ROC_infection_D3}
+				\\{\scriptsize D3 178 mg/dl; AUC 0.64; NPV 0.79; \\Spec 0.66; Sens 0.64; p=0.011}
+			\end{figure}
+		\column[c]{0.45\textwidth}
+			\begin{figure}
+				\includegraphics[width=\textwidth]{../Figures/crp_comp_ROC_infection_D4}
+				\\{\scriptsize D4 125 mg/dl; AUC 0.71; NPV 0.83; \\Spec 0.64; Sens 0.74; p$<$0.001}
+			\end{figure}
+	\end{columns}
 \end{frame}

### 2016-02-13 Presentations/viva_presentation_short.tex
Viva Presentation - Rename all sections - Move case study to end of presentation - Next - start cutting slides
Viva Presentation - Short
- TexStudio crashed for the first time today!!!!

--- 2016-02-13 Presentations/viva_presentation.tex
@@
 %------------------------------------------------
-\section{Introduction}
-\subsection{Pancreatic cancer}
+%\section{Chapter 1/ Introduction}
+\section[Chapter 1]{Introduction}
+
 \begin{frame}
@@
 
-\subsection{CPET}
 \begin{frame}
@@
 
-\section{Methods}
 \begin{frame}
@@
 
-% Discuss the results in the context of a patient.
-% 63 yr old female with BMI of 35 presents with painless OJ
-% PMH - HTN, T2DM, smoked 10/day for 40 yrs
-% Bloods: Bil 180, CRP 22, Alb 30
-% Investigations reveal an operable tumour of the head of the pancreas
-% How would this patient have been managed 10 years ago?
-\section{Case study}
-\begin{frame}
-	\frametitle{A typical patient}
-	\begin{itemize}
-		\item 63 yr old female with BMI of 35 presents with painless OJ
-		\item PMH - HTN, T2DM, smoked 10/day for 40 yrs
-		\item Bloods: Bil 180, CRP 22, Alb 30
-		\item Imaging: Resectable tumour of head of pancreas
-		\item How would this patient have been managed 10 years ago?
-	\end{itemize}
-\end{frame}
-
-\begin{frame}
-	\frametitle{10 years ago...}
-	\begin{itemize}
-		\item Non-dynamic measures of fitness - ECG, PFTs, ASA, ECHO
-		\item Routine biliary drainage
-		\item No objective methods to predict complications other than pancreatic leak
-		\item Impact of postoperative SIRS on complications - no clinical utility
-		\item Limited ability to identifying patients at risk of complications
-		\item ERAS principles not universally applied
-		\item No prehabilitation
-	\end{itemize}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Now... }
-	\begin{itemize}
-		\item CPET in all or selected cases (evidence from other patient groups)
-		\item Routine biliary drainage no longer indicated
-		\item Enhanced recovery more widespread
-		\medskip
-		\hrule
-		\medskip
-		\item Reluctance to perform CPET in jaundiced patients
-		\item No recognised thresholds for CPET parameters
-		\item No clear thresholds for post-op inflammatory markers in predicting complications
-		\item Poor understanding of impact of obesity on preop function and postop outcomes
-		\item No prehabilitation
-	\end{itemize}
-\end{frame}
-
 %--CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2----CHAPTER 2--
-\section{Chapter 2}
+\section[Chapter 2]{CPET and Postoperative Outcomes}
 \begin{frame}
 	\frametitle{CPET and Postoperative Outcomes}
-	\framesubtitle{Chapter 2 Page 56; (n=100)}
+	\framesubtitle{Does CPET have a role in risk stratification of patients?}
 	\begin{block}{\textbf{Complications}}
@@
 \begin{frame}
-	\frametitle{CPET and Postoperative Outcomes}
-	\framesubtitle{The Glasgow Cohort}
+	\frametitle{CPET in the Glasgow Cohort}
+	\framesubtitle{Why was 'aerobic capacity' so poor in these patients yet outcomes were comparable?}
 	\begin{figure}
@@
 	\frametitle{CPET and Postoperative Outcomes}
-	\framesubtitle{Citations}
+	\framesubtitle{Publication, Citations and more questions!}
+
+	\begin{figure}
+		\includegraphics[width=0.6\linewidth]{cpet_hpb_publication}
+	\end{figure}
+
 	\begin{columns}[c]
@@
 			\begin{figure}
-			\centering
-			\includegraphics[width=\textwidth]{cpet_citations}
+				\includegraphics[width=0.6\textwidth]{cpet_citations}
 			\end{figure}
@@
 		\column{0.6\textwidth}
-	Why is $\dot{V}_{O_2}$AT so low in patients undergoing pancreatic surgery?\\
+			{\footnotesize 	Why is $\dot{V}_{O_2}$AT so low in patients undergoing pancreatic surgery?\\
 				\medskip
@@
 				\medskip
-	Does $\dot{V}_{O_2}$AT truely reflect cardiopulmonary function in this cohort of patients?\\	
+				Does $\dot{V}_{O_2}$AT truly reflect cardiopulmonary function in this cohort of patients?\\	
+			}
+
 	\end{columns}
@@
 %--CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3----CHAPTER 3--
-\section{Chapter 3}
+\section[Chapter 3]{CPET, Jaundice and Preoperative Pathophysiology}
 
 \begin{frame}
-	\frametitle{Chapter 3}
-	\framesubtitle{CPET, Jaundice and Preoperative Pathophysiology - Page 76-96 }
-	\begin{block}{Title}
-	An investigation into the relationship between cardiopulmonary exercise testing, preoperative pathophysiology and obstructive jaundice in patients undergoing pancreaticoduodenectomy.
-	\end{block}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Distribution of Jaundice in the study cohort }
-	\framesubtitle{n=138}
-
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.7\linewidth]{jaundice_distribution}
-		\caption{}
-		\label{fig:jaundice_distribution}
-	\end{figure}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Jaundice and patient characteristics}
-	\framesubtitle{Table 3.1 Page 83, Table 3.2 Page 85 }
-	\begin{figure}
-		\includegraphics[height=0.4\textheight]{jaundice_vs_patient_factors}
-	\end{figure}
-	{\scriptsize			
-	Severity of systemic inflammation proportionate to severity of jaundice. \\
-	\medskip
-
-	Obstructive jaundice associated with electrolyte abnormalities.\\
-	\medskip
-
-	Jaundiced patients were more likely to be anaemic (p$<$0.001) with lower haematocrit (p$<$0.001) and lower mean corpuscular volume (p=0.001).\\
-	\medskip
-
-	No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
-	}
-\end{frame}
-
-\begin{frame}
-	\frametitle{CPET and Jaundice}
-	\framesubtitle{Figure 3.1 Page 87}
-	\begin{columns}[t]
-		\column{0.5\textwidth}
-			\begin{figure}
-				\centering
-				\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_at_bil}
-				\caption{$\dot{V}_{O_2}$AT versus serum bilirubin}
-				\label{fig:cpet_oj_scatter_at_bil}
-			\end{figure}
-
-		\column{0.5\textwidth}
-			\begin{figure}
-				\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_peak_bil}
-				\caption{$\dot{V}_{O_2}$Peak versus serum bilirubin}
-				\label{fig:cpet_oj_scatter_peak_bil}
-			\end{figure}
-	\end{columns}
-\end{frame}
-
-\begin{frame}
-	\frametitle{$\dot{V}_{O_2}$ vs Patient Factors }
-	\framesubtitle{Table 3.6 Page 91 - Multivariate binary logistic regression}
-
+	\frametitle{CPET, Jaundice and Preoperative Pathophysiology}
+	\framesubtitle{What were the preoperative patient factors that were associated with a low $\dot{V}_{O_2}$?}
 	Factors associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min
@@
 
+\begin{frame}
+	\frametitle{What was the effect of jaundice on $\dot{V}_{O_2}$?}
+	\framesubtitle{None on multivariate analysis. Weak correlation on scatter-plot analysis.}
+	\begin{columns}[t]
+		\column{0.5\textwidth}
+		\begin{figure}
+			\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_at_bil}
+			\\ $\dot{V}_{O_2}$AT versus serum bilirubin
+		\end{figure}
+
+		\column{0.5\textwidth}
+		\begin{figure}
+			\includegraphics[width=\textwidth]{../Figures/cpet_oj_scatter_peak_bil}
+			\\ $\dot{V}_{O_2}$Peak versus serum bilirubin
+		\end{figure}
+	\end{columns}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Jaundice and other patient characteristics}
+	\framesubtitle{What was the relationship between obstructive jaundice and other parameters?}
+
+	\begin{columns}
+	\column[c]{0.5\textwidth}
+
+	\begin{figure}
+		\includegraphics[width=\textwidth]{jaundice_distribution}
+	\end{figure}
+
+	\begin{figure}
+		\includegraphics[width=\textwidth]{jaundice_vs_patient_factors}
+	\end{figure}
+
+	\column[c]{0.5\textwidth}		
+	{\scriptsize			
+	Severity of systemic inflammation proportionate to severity of jaundice. \\
+	\medskip
+
+	Obstructive jaundice associated with electrolyte abnormalities.\\
+	\medskip
+
+	Jaundiced patients were more likely to be anaemic, with lower haematocrit and lower mean corpuscular volume.\\
+	\medskip
+
+	No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
+	}
+
+	\end{columns}
+\end{frame}
+
 %--CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4----CHAPTER 4--
-\section{Chapter 4}
+\section[Chapter 4]{CPET and Body Composition}
 
@@
 
-\section{Chapter 5}
+\section[Chapter 5]{Factors affecting post op inflammation}
 
@@
 
-\section{Chapter 6}
+\section[Chapter 6]{Postoperative CRP and complications}
 \begin{frame}
@@
 
+\section[Chapter 7]{Discussion}
+
+% Discuss the results in the context of a patient.
+% 63 yr old female with BMI of 35 presents with painless OJ
+% PMH - HTN, T2DM, smoked 10/day for 40 yrs
+% Bloods: Bil 180, CRP 22, Alb 30
+% Investigations reveal an operable tumour of the head of the pancreas
+% How would this patient have been managed 10 years ago?
+\section{Case study}
+\begin{frame}
+	\frametitle{A typical patient}
+	\begin{itemize}
+		\item 63 yr old female with BMI of 35 presents with painless OJ
+		\item PMH - HTN, T2DM, smoked 10/day for 40 yrs
+		\item Bloods: Bil 180, CRP 22, Alb 30
+		\item Imaging: Resectable tumour of head of pancreas
+		\item How would this patient have been managed 10 years ago?
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{10 years ago...}
+	\begin{itemize}
+		\item Non-dynamic measures of fitness - ECG, PFTs, ASA, ECHO
+		\item Routine biliary drainage
+		\item No objective methods to predict complications other than pancreatic leak
+		\item Impact of postoperative SIRS on complications - no clinical utility
+		\item Limited ability to identifying patients at risk of complications
+		\item ERAS principles not universally applied
+		\item No prehabilitation
+	\end{itemize}
+\end{frame}
+
+\begin{frame}
+	\frametitle{Now... }
+	\begin{itemize}
+		\item CPET in all or selected cases (evidence from other patient groups)
+		\item Routine biliary drainage no longer indicated
+		\item Enhanced recovery more widespread
+		\medskip
+		\hrule
+		\medskip
+		\item Reluctance to perform CPET in jaundiced patients
+		\item No recognised thresholds for CPET parameters
+		\item No clear thresholds for post-op inflammatory markers in predicting complications
+		\item Poor understanding of impact of obesity on preop function and postop outcomes
+		\item No prehabilitation
+	\end{itemize}
+\end{frame}
+
 %------------------------------------------------

### 2016-02-13 Presentations/viva_presentation_short.tex
Viva Presentation
- Down to 35 slides
- Can remove a lot from the intro

@@
 \begin{frame}
-	\frametitle{Parameters measured during CPET }
-	\framesubtitle{ Section 1.6 Page 25}
-	\begin{itemize}
-		\item
-		\item
-		\item
-	\end{itemize}
-\end{frame}
-
-\begin{frame}
 	\frametitle{Role of CPET in preoperative assessment}
@@
 	\end{enumerate}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Patients and Methods}
-	\begin{itemize}
-		\item Patients undergoing pancreaticoduodenectomy at Glasgow Royal Infirmary
-		\item August 2008 - August 2012
-		\item Prospectively maintained MS Access database
-		\item Preoperative data - Bloods, co-morbidity including POSSUM Physiology Score, stent information
-		\item Cardiopulmonary exercise testing
-		\item Prospective recording of complications using ISGPS or Clavie-Dindo classification
-		\item Serial measurements of biochemical parameters during first postoperative week
-	\end{itemize}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{CPET, Jaundice and Preoperative Pathophysiology}
-	\framesubtitle{What were the preoperative patient factors that were associated with a low $\dot{V}_{O_2}$?}
-	Factors associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min
-	\begin{table}
-		\begin{tabular}{l l l l}
-			& HR   & 95\% CI    & p     \\ \hline
-			Female sex & 3.75 & 1.57-8.95  & 0.003 \\
-			BMI$>$25   & 3.65 & 1.61-8.26  & 0.002 \\
-			Cancer     & 4.02 & 1.33-12.16 & 0.014 \\
-			CRP$>$10   & 2.98 & 1.29-6.86  & 0.010
-		\end{tabular}
-	\end{table}
+	\frametitle{Jaundice and other patient characteristics}
+	\framesubtitle{What was the relationship between obstructive jaundice and other parameters?}
 
-	Factors associated with $\dot{V}_{O_2}$Peak $<$ 16ml/kg/min
-	\begin{table}
-		\begin{tabular}{l l l l}
-			& HR   & 95\% CI   & p        \\ \hline
-			Female sex  & 7.57 & 3.09-18.5 & $<$0.001 \\
-			BMI$>$25    & 2.57 & 1.18-5.63 & 0.018    \\
-			Haemoglobin & 2.43 & 1.05-5.63 & 0.038
-		\end{tabular}
-	\end{table}
+	\begin{columns}
+		\column[c]{0.5\textwidth}
+
+		\begin{figure}
+			\includegraphics[width=\textwidth]{jaundice_distribution}
+		\end{figure}
+
+		\begin{figure}
+			\includegraphics[width=\textwidth]{jaundice_vs_patient_factors}
+		\end{figure}
+
+		\column[c]{0.5\textwidth}		
+		{\scriptsize			
+			Severity of systemic inflammation proportionate to severity of jaundice. \\
+			\medskip
+			Obstructive jaundice associated with electrolyte abnormalities.\\
+			\medskip
+			Jaundiced patients were more likely to be anaemic, with lower haematocrit and lower mean corpuscular volume.\\
+			\medskip
+			No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
+		}
+	\end{columns}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Jaundice and other patient characteristics}
-	\framesubtitle{What was the relationship between obstructive jaundice and other parameters?}
-
-	\begin{columns}
-	\column[c]{0.5\textwidth}
-
-	\begin{figure}
-		\includegraphics[width=\textwidth]{jaundice_distribution}
-	\end{figure}
-
-	\begin{figure}
-		\includegraphics[width=\textwidth]{jaundice_vs_patient_factors}
-	\end{figure}
-
-	\column[c]{0.5\textwidth}		
-	{\scriptsize			
-	Severity of systemic inflammation proportionate to severity of jaundice. \\
+	\frametitle{What preop factors were associated with low $\dot{V}_{O_2}$?}
+	\framesubtitle{Multivariate binary logistic regression analysis}
+	Factors associated with $\dot{V}_{O_2}$AT $<$ 10 ml/kg/min
+	\begin{table}
+		\begin{tabular}{l| l l l}
+			& HR   & 95\% CI    & p     \\ \hline
+			Female sex & 3.75 & 1.57-8.95  & 0.003 \\
+			BMI$>$25   & 3.65 & 1.61-8.26  & 0.002 \\
+			Cancer     & 4.02 & 1.33-12.16 & 0.014 \\
+			CRP$>$10   & 2.98 & 1.29-6.86  & 0.010
+		\end{tabular}
+	\end{table}
+	\hrule
 	\medskip
-
-	Obstructive jaundice associated with electrolyte abnormalities.\\
-	\medskip
-
-	Jaundiced patients were more likely to be anaemic, with lower haematocrit and lower mean corpuscular volume.\\
-	\medskip
-
-	No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
-	}
-
-	\end{columns}
+	Factors associated with $\dot{V}_{O_2}$Peak $<$ 16ml/kg/min
+	\begin{table}
+		\begin{tabular}{l| l l l}
+			& HR   & 95\% CI   & p        \\ \hline
+			Female sex  & 7.57 & 3.09-18.5 & $<$0.001 \\
+			BMI$>$25    & 2.57 & 1.18-5.63 & 0.018    \\
+			Haemoglobin & 2.43 & 1.05-5.63 & 0.038
+		\end{tabular}
+	\end{table}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Chapter 4}
-	\framesubtitle{CPET and Body Composition - Page 97-122 }
-	\begin{block}{Title}
-		An investigation into the relationship between cardiopulmonary exercise testing and body composition in patients undergoing pancreaticoduodenectomy.
-	\end{block}
-\end{frame}
-
-\begin{frame}
-	\frametitle{What causes a low $\dot{V}_{O_2}$? }
-	\framesubtitle{Section 4.1.2 Page 98-99 }
+	\frametitle{ CPET and Body Composition}
+	\framesubtitle{What causes a low $\dot{V}_{O_2}$? }
 	\begin{itemize}
@@
 		\item $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak are most commonly used parameters
-		\item Both are normalised by dividing using patient's total body weight
-		\item But, during exercise $\dot{V}_{O_2}$ is a function of skeletal muscle mass
-		\item This can result in 'spurious correlation' unfairly penalising obese subjects
+		\item During exercise $\dot{V}_{O_2}$ is a function of skeletal muscle mass
+		\item But, both are normalised by against patient's total body weight
+		\item 'Spurious correlation' unfairly penalises obese subjects
 		\item Most patients in this cohort did not have overt cardiac/respiratory disease that could explain the low $\dot{V}_{O_2}$.
@@
 \begin{frame}
-	\frametitle{Body composition estimation}
-	\framesubtitle{Figure 4.1 Page 103}
+	\frametitle{Body composition differences between the sexes}
+	\framesubtitle{Significant differences in skeletal muscle area across all BMIs.}
 	\begin{figure}
 		\centering
-		\includegraphics[width=\linewidth]{bc_gimp}
+		\includegraphics[height=0.95\textheight]{../Figures/bc_gender_bmi_pie}
 		\label{fig:bc_gimp}
@@
 \begin{frame}
-	\frametitle{Chapter 5}
-	\begin{block}{Title}
-		An investigation into the relationship between preoperative clinico-pathological characteristics and post-operative systemic inflammatory response in patients undergoing pancreaticoduodenectomy.
-	\end{block}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Background}
-	\framesubtitle{Section 5.1 Page 124}
+	\frametitle{Impact of perioperative systemic inflammation?}
+	\framesubtitle{Inflammation before and after surgery affects short and long term outcomes}
 	\begin{itemize}
@@
 	\frametitle{Preop inflammation vs. Postop CRP}
-	\framesubtitle{Figure 5.3 Page 142}
+	\framesubtitle{Post op CRP trends are influenced by both preop CRP and preop albumin.}
 	\begin{figure}
@@
 	\frametitle{Obstructive jaundice vs. Postop CRP}
-	\framesubtitle{ }
+	\framesubtitle{Obstructive jaundice is associated with an attenuated inflammatory response.}
 	\begin{figure}
@@
 \section[Chapter 6]{Postoperative CRP and complications}
-\begin{frame}
-	\frametitle{Chapter 6}
-	\framesubtitle{Postoperative CRP and complications}
-	\begin{block}{Title}
-		An investigation into the relationship between postoperative systemic inflammation and complications after pancreaticoduodenectomy.
-	\end{block}
-\end{frame}
 
@@
 \begin{frame}
-	\frametitle{A typical patient}
+	\frametitle{A case study}
+	\framesubtitle{How does this work influence decision-making?}
 	\begin{itemize}
@@
 		\item Imaging: Resectable tumour of head of pancreas
-		\item How would this patient have been managed 10 years ago?
+		\medskip
+		\item Should this patient undergo CPET?
+		\item Should this patient undergo biliary drainage?
+		\item Should we wait until resolution of jaundice before CPET?
+		\item How can the CPET results be used to influence her care?
+		\item How can the postoperative inflammatory markers be used to manage her care?
 	\end{itemize}
@@
 \begin{frame}
-	\frametitle{10 years ago...}
-	\begin{itemize}
-		\item Non-dynamic measures of fitness - ECG, PFTs, ASA, ECHO
-		\item Routine biliary drainage
-		\item No objective methods to predict complications other than pancreatic leak
-		\item Impact of postoperative SIRS on complications - no clinical utility
-		\item Limited ability to identifying patients at risk of complications
-		\item ERAS principles not universally applied
-		\item No prehabilitation
-	\end{itemize}
+	\frametitle{A case study}
+	\begin{description}
+	{\scriptsize 	
+		\item [Ch 2] CPET may allow identifying if she is at increased risk of complications, prolonged hospital stay and non-receipt of adjuvant therapy. 
+					\\ CPET should not be used on its own to decline surgery. 
+					\hrule
+		\item [Ch 3] CPET can be performed in the presence of jaundice. 
+					\\ Jaundice does not adversely affect dynamic cardiopulmonary response to and this cannot be a justification for biliary drainage. 
+					\\ Taken together with evidence from trials on preperative biliary drainage, she should proceed direct to surgery.
+						\hrule
+		\item [Ch 4] Interpret low $\dot{V}_{O_2}$ in her case with caution. 
+					\\ Her aerobic capacity may be better than that suggested by her test results. 
+					\\ Beware of \textit{spurious correlation} with total body weight.
+						\hrule
+		\item [Ch 5] Jaundice and hypo-albuminemia attenuate post-op inflammatory response.
+					\\ However, this does not translate into increased complications. 
+					\\ Aerobic capacity does not affect post-op inflammatory response.
+						\hrule
+		\item [Ch 6] If her drain amylase on D3 is normal and D3 CRP$<$178 or D4 CRP$<$125, she is likely to have an uneventful postoperative course and must continue on 'enhanced-recovery' pathway.
+				\\ If she develops a POPF, the early inflammatory response does not predict outcome.
+	}
+	\end{description}
 \end{frame}
@@
 \begin{frame}
-	\frametitle{Now... }
+	\frametitle{Future directions}
 	\begin{itemize}
-		\item CPET in all or selected cases (evidence from other patient groups)
-		\item Routine biliary drainage no longer indicated
-		\item Enhanced recovery more widespread
-		\medskip
-		\hrule
-		\medskip
-		\item Reluctance to perform CPET in jaundiced patients
-		\item No recognised thresholds for CPET parameters
-		\item No clear thresholds for post-op inflammatory markers in predicting complications
-		\item Poor understanding of impact of obesity on preop function and postop outcomes
-		\item No prehabilitation
+		\item Assessing impact of prehabilitation
+		\item Assessing impact of neo-adjuvant treatments
+		\item Does CPET predict long-term survival?
+		\item Improvements in aerobic capacity and body composition through nutritional supplementation and physical exercise
+		\item Correction of anaemia through preoperative iron supplementation
+		\item Early recognition and aggressive management of complications
+		\item Assessing suitability for 'fast-track' to surgery and 'enhanced-recovery'
 	\end{itemize}
+
 \end{frame}
-
-%------------------------------------------------
-
-\begin{frame}[fragile] % Need to use the fragile option when verbatim is used in the slide
-\frametitle{Verbatim}
-\begin{example}[Theorem Slide Code]
-\begin{verbatim}
-\begin{frame}
-\frametitle{Theorem}
-\begin{theorem}[Mass--energy equivalence]
-$E = mc^2$
-\end{theorem}
-\end{frame}\end{verbatim}
-\end{example}
-\end{frame}
-
-%------------------------------------------------
 

### 2016-02-13 Presentations/viva_presentation_short.tex
Viva Presentation - First Draft Complete
- Email this to Donny
- Take a break
- Shower and clear toxins

@@
 \begin{frame}
+	\frametitle{My Supervisors - Thank you!}
+	I wish to thank the following people for their unwavering support through the past few years.
+I thank them especially for their faith that I will see this through and if they had any doubts, for not telling me about them.\\
+	\bigskip
+	Professor Donald C McMillan,\\
+	University Department of Surgery, Glasgow Royal Infirmary\\
+	\bigskip
+	Professor Paul Horgan,\\
+	University Department of Surgery, Glasgow Royal Infirmary\\
+\end{frame}
+
+\begin{frame}
 \frametitle{Overview} % Table of contents slide, comment this block out to remove it
@@
 			\item Patient fitness is as important as tumour factors
+			\item Adjuvant treatment improves survival
 		\end{itemize}
@@
 	\frametitle{Pancreaticoduodenectomy}
-	\begin{columns}[t]
-		\column{0.45\textwidth}
-		\begin{figure}
-			\centering
-			\includegraphics[width=0.8\linewidth]{whipple_schematic_pre}
-			\label{fig:whipple_schematic_pre}
-		\end{figure}
-		{\scriptsize
-			\textbf{Structures removed:}
-			\begin{itemize}
-			\item Head of pancreas
-			\item Distal stomach, duodenum, proximal jejunum
-			\item  Common bile duct, gallbladder
-			\item Lymphadenectomy
-			\item Venous resection in some
-		\end{itemize}}
-
-		\column{0.45\textwidth}
-		\begin{figure}
-			\centering
-			\includegraphics[width=0.8\linewidth]{whipple_schematic_post}
-			\label{fig:whipple_schematic_post}
-		\end{figure}
-		{\scriptsize
-		\textbf{Reconstruction:}
-		\begin{itemize}
-			\item Pancreaticojejunostomy (25\% leak rate)
-			\item Hepaticojejunostomy
-			\item Gastrojejunostomy
-		\end{itemize}
-		}
-	\end{columns}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Complications of Surgery}
-	\framesubtitle{Section 1.2.4 Pages 8-11}
+	\framesubtitle{Postoperative complications}
 	\begin{itemize}
@@
 		\item Clavien-Dindo classification of complications
-	\end{itemize}
-\end{frame}
-
-\begin{frame}
-	\frametitle{Adjuvant therapy}
-	\framesubtitle{Section 1.3 Pages 11,13 }
-	\begin{itemize}
-		\item Adjuvant therapy improves survival
-		\item Currently standard of care and subject of multiple trials
-		\item 25\% - 50\% of patients do not receive adjuvant treatment
-		\item Multiple factors including:
-		\begin{itemize}
-			\item Comorbidity
-			\item Complications of surgery
-			\item Favourable pathological features
-		\end{itemize}
 	\end{itemize}
@@
 		\item Do not provide targets for prehabilitation
+		\item Preoperative management of obstructive jaundice remains controversial
 		\item Pancreas specific risk factors of POPF
 		\begin{itemize}
-			\item Soft pancreas
-			\item Narrow pancreatic duct
-			\item Pathology not obstructing PD
+			\item Soft pancreas, narrow pancreatic duct, non-obstructing pathology
 			\item These factors are not modifiable
@@
 \begin{frame}
-	\frametitle{What is CPET?}
-	\framesubtitle{Section 1.5 Page 18}
+	\frametitle{Cardiopulmonary exercise testing}
+	\framesubtitle{What, how, why?}
 	\begin{itemize}
+
 		\item Objective measure of cardiopulmonary response to exercise
-		\item Dynamic test measuring numerous parameters
+		\item Dynamic test measuring multiple aspects of oxygen delivery
 		\item Cardiac, respiratory, vascular and mitochondrial function
@@
 		\item Widely used in general, oesophago-gastric, vascular and thoracic surgery
-	\end{itemize}
-\end{frame}
-
-\begin{frame}
-	\frametitle{CPET Methodology}
-	\framesubtitle{Section 1.5.2 Page 20 }
+		\item CPET Methodology
 		\begin{itemize}
-		\item Dynamic exercise test
-		\item Cycle ergometer with electronic braking and incremental workload
+			\item Cycle ergometer with electronic braking
 			\item Breath-by-breath gas analysis using a tight-fitting face mask
@@
 			\item Incremental work-load test to volitional fatigue
-		%include picture here
+		\end{itemize}
 	\end{itemize}
@@
 	\frametitle{Physiological basis of CPET} 
+	\begin{columns}
+	\column[c]{0.5\textwidth}
+
 		\begin{figure}
-		\centering
-		\includegraphics[width=0.5\linewidth]{CellRespiration}
-		\label{fig:CellRespiration}
+			\includegraphics[width=\linewidth]{CellRespiration}
 		\end{figure}
-	\begin{block}
-	\centering
-	$H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2$
-	\end{block}
+		{\scriptsize $H^+ + HCO3^- \Longleftrightarrow H_2CO_3 \Longleftrightarrow H_2O + CO_2$}
 
-\end{frame}
+	\column[c]{0.5\textwidth}
+	\begin{figure}[htbp]
+		\includegraphics[width=\linewidth]{../Figures/cpet_vslope}
+	\end{figure}
 
-\begin{frame}
-	\frametitle{Determination of Anaerobic Threshold}
-	\framesubtitle{Section 1.5.3 Page 22}
-	\begin{figure}[htbp]
-		\centering
-		\includegraphics[width=0.7\linewidth]{../Figures/cpet_vslope}
-		%\caption{Determination of $\dot{V}_{O_2}$AT by the V-slope method.}
-		\label{fig:cpet_vslope}
-	\end{figure}
-\end{frame}
+	\end{columns}
 
-\begin{frame}
-	\frametitle{Role of CPET in preoperative assessment}
-	\begin{itemize}
-		\item General Surgery
-		\item Oesophago-gastric surgery
-		\item Bariatric surgery
-		\item Liver transplantation
-		\item Thoracic surgery
-		\medskip
-		\item Little evidence in pancreatic surgery in 2010
-
-	\end{itemize}
 \end{frame}
@@
 	\frametitle{CPET in the Glasgow Cohort}
-	\framesubtitle{Why was 'aerobic capacity' so poor in these patients yet outcomes were comparable?}
+	\framesubtitle{Why was 'aerobic capacity' so poor in these patients, yet outcomes were comparable?}
 	\begin{figure}
@@
 
+\section{The End}
+
 \begin{frame}
-\Huge{\centerline{The End}}
+\frametitle{The End}
+\framesubtitle{Tools and Roles}
+
+\begin{table}
+	\begin{tabular}{r l}
+		        database & MS Access with VBA            \\
+		data collection: & VBA + Dynamic Data Exchange   \\
+		     statistics: & SPSS versions 19, 22          \\
+		   bibliography: & Zotero, Bibus                 \\
+		 thesis writing: & LaTeX using TeXstudio         \\
+		version control: & Git + Dropbox                 \\
+		                 &  \\
+		         salary: & Surgical Critical Care Fellow \\
+		                 & Surgical SpR on-call at GRI   \\
+		                 & \\
+		         sanity: & Staff on Surgical HDU, Donny
+	\end{tabular}
+\end{table}
+
 \end{frame}

### 2016-02-14 Presentations/viva_presentation_short.tex
Viva Presentation - Donny comments
- Next add transitions

@@
 \documentclass[10pt]{beamer}
+%\documentclass[handout, 10pt]{beamer}
+\setbeamercovered{transparent}
 
@@
 
-\begin{frame}
-	\frametitle{My Supervisors - Thank you!}
-	I wish to thank the following people for their unwavering support through the past few years.
+%\begin{frame}
+%	\frametitle{My Supervisors - Thank you!}
+%	I wish to thank the following people for their unwavering support through the past few years.
 I thank them especially for their faith that I will see this through and if they had any doubts, for not telling me about them.\\
-	\bigskip
-	Professor Donald C McMillan,\\
-	University Department of Surgery, Glasgow Royal Infirmary\\
-	\bigskip
-	Professor Paul Horgan,\\
-	University Department of Surgery, Glasgow Royal Infirmary\\
-\end{frame}
+%	\bigskip
+%	Professor Donald C McMillan,\\
+%	University Department of Surgery, Glasgow Royal Infirmary\\
+%	\bigskip
+%	Professor Paul Horgan,\\
+%	University Department of Surgery, Glasgow Royal Infirmary\\
+%\end{frame}
 
@@
 		\begin{itemize}
-			\item Complex surgery with 50\% morbidity
+			\item Complex surgery with 40\% morbidity
 			\item Performed only in specialist centres
@@
 		\item Clavien-Dindo classification of complications
+		\medskip
+		\item Complications may preclude adjuvant treatment
+		\item Median survival is approximately 18 months
+		\item Adjuvant treatment improves this by 2-3 months
 	\end{itemize}
@@
 			\end{figure}
+
 		\column[c]{0.45\textwidth}
@@
 \section[Chapter 7]{Discussion}
+\begin{frame}
+	\frametitle{Future directions}
+	\begin{itemize}
+		\item Assessing impact of prehabilitation
+		\item Improving in aerobic capacity and body composition through nutritional supplementation and physical exercise
+		\item Assessing impact of neo-adjuvant treatments
+		\item Customising goal-directed therapy based on CPET
+		\item Correction of anaemia through preoperative iron supplementation
+		\item Early recognition and aggressive management of complications
+		\item Assessing suitability for 'fast-track' to surgery and 'enhanced-recovery'
+	\end{itemize}
+\end{frame}
 
-% Discuss the results in the context of a patient.
-% 63 yr old female with BMI of 35 presents with painless OJ
-% PMH - HTN, T2DM, smoked 10/day for 40 yrs
-% Bloods: Bil 180, CRP 22, Alb 30
-% Investigations reveal an operable tumour of the head of the pancreas
-% How would this patient have been managed 10 years ago?
 \section{Case study}
@@
 		\item 63 yr old female with BMI of 35 presents with painless OJ
-		\item PMH - HTN, T2DM, smoked 10/day for 40 yrs
+		\item PMH - HTN, T2DM, ex-smoker
 		\item Bloods: Bil 180, CRP 22, Alb 30
@@
 		\medskip
+
 		\item Should this patient undergo CPET?
@@
 \begin{frame}
-	\frametitle{Future directions}
-	\begin{itemize}
-		\item Assessing impact of prehabilitation
-		\item Assessing impact of neo-adjuvant treatments
-		\item Does CPET predict long-term survival?
-		\item Improvements in aerobic capacity and body composition through nutritional supplementation and physical exercise
-		\item Correction of anaemia through preoperative iron supplementation
-		\item Early recognition and aggressive management of complications
-		\item Assessing suitability for 'fast-track' to surgery and 'enhanced-recovery'
-	\end{itemize}
-
+\centerline{Thank You}
 \end{frame}
 
-\section{The End}
-
 \begin{frame}
-\frametitle{The End}
-\framesubtitle{Tools and Roles}
+\frametitle{Tools and Roles}
+%\framesubtitle{Tools and Roles}
 

### 2016-02-14 Presentations/viva_talking_points.tex
Viva presentation - New file - Talking points

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\author{Vishnu V Chandrabalan}

\renewcommand{\familydefault}{\sfdefault}

\begin{document}

\section{Chapter 3}

\subsection{OJ vs $\dot{V}_{O_2}$}
	- OJ was thought to be associated with adverse haemodynamic events

	- Routine PBD was advised based on this

	- Multiple studies have reported adverse events after PBD

	- Management of Bil>250 is unclear

	- Here, OJ not associated with VO2

	- Parker et al, Junejo et al

	- \textbf{NO JUSTIFICATION FOR PBD}

	-Figures on P87, Tables p91, 92

\subsection{VO2 vs Female/BMI}
	- Not clear why this is the case

	- Body composition differences between gender and increasing BMI

	- Do females and obese subjects need different thresholds as suggested by some authors?

\subsection{OJ vs biochemistry}
	Jaundice associated with systemic inflammation, anaemia and electrolyte abnormalities

\subsection{ VO2AT associated with systemic inflammation}
	Sultan et al: NLR associated with low VO2AT in colorectal patients

\subsection{Peak Vo2 was related to hemoglobin - p92}
	- Heart failure patients

	- Healthy volunteers after blood transfusion

	- Blood loss associated with adverse outcomes

	- Is there a role for preoperative optimisation?

\subsection{vo2 vs cancer}
- Does sarcopenia play a role?

- Does the metabolic effects of cancer including inflammation, poor nutrition, etc. impair aerobic capacity?

\clearpage

\section{Chapter 4}

\subsection{BC vs clinicopathological char T4.1 p106}

- Females had more subcutn fat, men had more visceral fat and skeltal muscle

- Lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$peak -- more subcutn fat but not visceral fat

- Lower $\dot{V}_{O_2}$peak -- lower skeletal muscle

- Lower Hb -- lower skeletal muscle

\subsection{BC vs sex/bmi - F4.2 p107}

- As BMI increases, body composition differences between men and women becomes apparent.

- The proportion of skeletal muscle in comparison to fat is low in women (32 vs 38) even at normal BMI.

- This proportion falls even further in obese women to 14\% while obese men have 22\% muscle contributing to their weight

- Obese men and obese women are very different in their body composition

\subsection{BC vs CPET - T4.2 p110}

- All calculations were controlled for the effect of gender

- Pulmonary function tests were controlled for effect of gender and age

- FVC/FEV positively correlated with skeletal muscle but not fat

- Differentiate between absolute and corrected $\dot{V}_{O_2}$.

- Differentiate between the 3 phases of exercise

- Positive correlation between skeletal muscle and load, tidal volume, abs $\dot{V}_{O_2}$, o2pulse

- Negative correlation between corrected $\dot{V}_{O_2}$ and fat - why is this?

\subsection{Scatter plots - f4.4 p111}

- solid line is absolute $\dot{V}_{O_2}$ before normalising for weight

- dotted line is corrected $\dot{V}_{O_2}$ after normalising for weight

- the negative correlation that appears with adipose tissue is 'spurious'

\subsection{Scaling for other factors t4.3 p113, Fig. 4.5 p114}

- the ideal scaling factor should remove any correlation with body habitus

- weight, height squared, BMI, skeletal muscle and eLBM were used

- Correction by eLBM resulted in loss of all correlations and probably best

\clearpage

\end{document}

### 2016-02-14 Presentations/viva_talking_points.tex
Viva Presentations - Talking points complete

@@
 \subsection{OJ vs $\dot{V}_{O_2}$}
-	- OJ was thought to be associated with adverse haemodynamic events
-
-	- Routine PBD was advised based on this
-
-	- Multiple studies have reported adverse events after PBD
-
-	- Management of Bil>250 is unclear
-
-	- Here, OJ not associated with VO2
-
-	- Parker et al, Junejo et al
-
-	- \textbf{NO JUSTIFICATION FOR PBD}
-
+	- OJ was thought to be associated with adverse haemodynamic events\\
+	- Routine PBD was advised based on this\\
+	- Multiple studies have reported adverse events after PBD\\
+	- Management of Bil>250 is unclear\\
+	- Here, OJ not associated with VO2\\
+	- Parker et al, Junejo et al\\
+	- \textbf{NO JUSTIFICATION FOR PBD}\\
 	-Figures on P87, Tables p91, 92
 
-\subsection{VO2 vs Female/BMI}
-	- Not clear why this is the case
-
-	- Body composition differences between gender and increasing BMI
-
+\subsection{$\dot{V}_{O_2}$ vs Female/BMI}
+	- Not clear why this is the case\\
+	- Body composition differences between gender and increasing BMI\\
 	- Do females and obese subjects need different thresholds as suggested by some authors?
@@
 
-\subsection{ VO2AT associated with systemic inflammation}
+\subsection{ $\dot{V}_{O_2}$AT associated with systemic inflammation}
 	Sultan et al: NLR associated with low VO2AT in colorectal patients
 
-\subsection{Peak Vo2 was related to hemoglobin - p92}
-	- Heart failure patients
-
-	- Healthy volunteers after blood transfusion
-
-	- Blood loss associated with adverse outcomes
-
+\subsection{Peak $\dot{V}_{O_2}$ was related to hemoglobin - p92}
+	- Heart failure patients\\
+	- Healthy volunteers after blood transfusion\\
+	- Blood loss associated with adverse outcomes\\
 	- Is there a role for preoperative optimisation?
 
-\subsection{vo2 vs cancer}
-- Does sarcopenia play a role?
-
+\subsection{$\dot{V}_{O_2}$ vs cancer}
+- Does sarcopenia play a role?\\
 - Does the metabolic effects of cancer including inflammation, poor nutrition, etc. impair aerobic capacity?
@@
 
-- Females had more subcutn fat, men had more visceral fat and skeltal muscle
-
-- Lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$peak -- more subcutn fat but not visceral fat
-
-- Lower $\dot{V}_{O_2}$peak -- lower skeletal muscle
-
+- Females had more subcutn fat, men had more visceral fat and skeltal muscle\\
+- Lower $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$peak -- more subcutn fat but not visceral fat\\
+- Lower $\dot{V}_{O_2}$peak -- lower skeletal muscle\\
 - Lower Hb -- lower skeletal muscle
@@
 
-- As BMI increases, body composition differences between men and women becomes apparent.
-
-- The proportion of skeletal muscle in comparison to fat is low in women (32 vs 38) even at normal BMI.
-
-- This proportion falls even further in obese women to 14\% while obese men have 22\% muscle contributing to their weight
-
+- As BMI increases, body composition differences between men and women becomes apparent.\\
+- The proportion of skeletal muscle in comparison to fat is low in women (32 vs 38) even at normal BMI. \\
+- This proportion falls even further in obese women to 14\% while obese men have 22\% muscle contributing to their weight\\
 - Obese men and obese women are very different in their body composition
@@
 
-- All calculations were controlled for the effect of gender
-
-- Pulmonary function tests were controlled for effect of gender and age
-
-- FVC/FEV positively correlated with skeletal muscle but not fat
-
-- Differentiate between absolute and corrected $\dot{V}_{O_2}$.
-
-- Differentiate between the 3 phases of exercise
-
-- Positive correlation between skeletal muscle and load, tidal volume, abs $\dot{V}_{O_2}$, o2pulse
-
+- All calculations were controlled for the effect of gender\\
+- Pulmonary function tests were controlled for effect of gender and age\\
+- FVC/FEV positively correlated with skeletal muscle but not fat\\
+- Differentiate between absolute and corrected $\dot{V}_{O_2}$.\\
+- Differentiate between the 3 phases of exercise\\
+- Positive correlation between skeletal muscle and load, tidal volume, abs $\dot{V}_{O_2}$, o2pulse\\
 - Negative correlation between corrected $\dot{V}_{O_2}$ and fat - why is this?
@@
 
-- solid line is absolute $\dot{V}_{O_2}$ before normalising for weight
-
-- dotted line is corrected $\dot{V}_{O_2}$ after normalising for weight
-
+- solid line is absolute $\dot{V}_{O_2}$ before normalising for weight\\
+- dotted line is corrected $\dot{V}_{O_2}$ after normalising for weight\\
 - the negative correlation that appears with adipose tissue is 'spurious'
@@
 
-- the ideal scaling factor should remove any correlation with body habitus
-
-- weight, height squared, BMI, skeletal muscle and eLBM were used
-
+- the ideal scaling factor should remove any correlation with body habitus\\
+- weight, height squared, BMI, skeletal muscle and eLBM were used\\
 - Correction by eLBM resulted in loss of all correlations and probably best
@@
 
+\section{Chapter 5}
+\subsection{Preop inflammation vs Postop CRP}
+- Preop CRP and albumin affected the inflammatory response in the first week\\
+- This is best seen in Fig 5.3 and Table 5.7 on p142\\
+- Line 0: absence of a preop SIRS; Line 1: only CRP abnormal;\\
+- Line 2: only Albumin abnormal; Line 3:both are abnormal\\
+- Postop CRP response was elevated if preop CRP was high with normal albumin\\
+- Hypoalbuminemia resulted in a dampened inflammatory response
+
+\subsection{OJ vs post op inflammation}
+- Postop CRP significantly lower in jaundiced patients\\
+- Preop CRP greater in OJ. Hence this shows a reversal of this phenomenon\\
+- Postop albumin lower- but this was a continuation of preop hypo-albuminemia
+
+\subsection{CPET, comorbidity, biochemistry, etc. vs postop inflammation}
+- No relationship with aerobic capacity, comorbidity and other factors\\
+- Probably due to dominant effect of preop inflammation and OJ\\
+- Low alb associated with multiple abnormalities as a continuum from preop
+
+\subsection{BC vs postop inflammation}
+- No association with CRP/neutrophil count\\
+- Low albumin $\rightarrow$ low skeletal muscle
+
+\subsection{Discussion}
+- Preop systemic inflammation affects postop SIRS\\
+- When hypoalbuminemia is present, effect of CRP is lost and postop SIRS is dampened\\
+- hypoalbuminemia is a hallmark of CARS\\
+- Preop jaundice affects postop SIRS\\
+- Immune dysregulation/impaired CRP response are seen with liver dysfunction\\
+- Bernal 2014: Liver transplantation - hypoalbuminemia - low $\dot{V}_{O_2}$\\
+- Richards 2012: prep inflammation - low  skeletal muscle\\
+- Hassen 2007: low skeletal muscle - greater postop inflammation after AAA\\
+- Sarcopenic-obesity $\rightarrow$ inflammation
+
+\clearpage
+
+\section{Chapter 6}
+
+\subsection{Numbers t6.4 p165}
+- Total 188; POPF 62; Infections 84
+\subsection{Infective complications, preop factors and other outcomes}
+- $\dot{V}_{O_2}$AT $\rightarrow$ infective complications\\
+- infective complications $\rightarrow$ other complications, critical care, mortality
+
+\subsection{CRP vs POPF p163}
+- CRP does not predict severity of POPF; stress on importance of drain amylase
+
+\subsection{CRP vs infective complications p 165}
+- Useful in the absence of POPF\\
+- POPF triggers other complications and early SIRS no longer a major factor
+
+\subsection{Albumin, WCC}
+- Not particularly useful clinically in spite of some associations later in the week
+
+\subsection{ROC analysis p171}
+- D3 CRP 178 NPV 0.79; D4 CRP 125 NPV 0.83
+
+\subsection{D4 CRP and Length of Stay t6.9 p172}
+- 2/3rds stayed less than 14 days\\
+- Nearly 90\% left critical care in under 1 week\\
+- Only 7.7\% were readmited to crit care against 21\% if CRP over 125\\
+- These were in the absence of enhanced recovery or CRP being included in discharge criteria
+
+\subsection{Discussion}
+- POPF $\rightarrow$ outcomes; ISGPF\\
+- CARS $\rightarrow$ SIRS;D4 CRP $\rightarrow$ surgical trauma
+
 \end{document}

### 2016-02-16 Chapters/CPET_Jaundice.tex
Viva Recommendations:
List of abbreviations
All tables should have units, legends
All values and ranges need defining
Chapter 1
 - More discussion of key RCTs
- Short discussion of etiology
Chapter 2
- Review patient numbers on p63 - pathologies dont add up to total
- Aim and end points must match
- Error in AT numbers in adjuvant therapy table
- P72 Remove unjustifed/unsupported conclusion
- P73 Remove GPS conclusion
Chapter 3 p96 Remove conclusions not based on thesis
Chapter 5-6 - Reduce to more undestandable key issues. All tables need legends. Several irrelevant tables should be removed.

@@
 \begin{figure}[htbp]
+
+	\caption{Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak.}
+	\label{fig:cpet_oj_scatter}
+
 	\centering
@@
 
-	\caption{Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak.}
-	\label{fig:cpet_oj_scatter}
+	\begin{flushleft}
+		The relationship between serum bilirubin and $\dot{V}_{O_2}$AT was weak (Pearson's $\rho$ = - 0.187, $\rho^2$ = 0.035, p = 0.028).
+There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)
+	\end{flushleft}
 
@@
 
-Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.
+%Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.
+%\todo{Remove this conclusion not based on thesis}
 

### 2016-02-16 Tables/cpet_oj_pft.tex
Table cpet_oj_pft
- Add units, legend
- Change PREDICTED to predicted

@@
 		\hline
-		                    &              \multicolumn{4}{c}{Preoperative Serum Bilirubin}              &  \\
-		                    & $\leq$ 17        & 18-35            & 35-250            & $>$ 250          & \textit{p} \\ \hline
-		FVC                 & 4.09 (3.49-4.69) & 3.76 (3.38-4.59) & 3.76 (3.16-4.05)  & 3.35 (2.85-4.38) & 0.092      \\
-		FEV1                & 2.95 (2.39-3.51) & 2.90 (2.12-3.34) & 2.68 (2.37-3.07)  & 2.72 (2.2-3.28)  & 0.556      \\
-		PREDICTED FEV1 (\%) & 105.0 (91-116)   & 98.50 (87-114)   & 103.0 (95-111.5)  & 101.0 (94-116)   & 0.761      \\
+		                        &       \multicolumn{4}{c}{Preoperative Serum Bilirubin ($\mu$mol/L)}        &  \\
+		Pulmonary Function Test & $\leq$ 17        & 18-35            & 35-250            & $>$ 250          & \textit{p} \\ \hline
+		FVC (litres)            & 4.09 (3.49-4.69) & 3.76 (3.38-4.59) & 3.76 (3.16-4.05)  & 3.35 (2.85-4.38) & 0.092      \\
+		FEV1 (litres)           & 2.95 (2.39-3.51) & 2.90 (2.12-3.34) & 2.68 (2.37-3.07)  & 2.72 (2.2-3.28)  & 0.556      \\
+		Predicted FEV1 (\%)     & 105.0 (91-116)   & 98.50 (87-114)   & 103.0 (95-111.5)  & 101.0 (94-116)   & 0.761      \\
 		FEV1/FVC                & 72.0 (65-77)     & 73.0 (65-78)     & 75.50 (70.5-79.5) & 78.0 (70-82)     & 0.115      \\
-		PREDICTED FEV1/FVC  & 94.0 (87-102)    & 96.0 (86-100)    & 99.0 (93-103)     & 102.0 (88-108)   & 0.107      \\ \hline
+		Predicted FEV1/FVC (\%) & 94.0 (87-102)    & 96.0 (86-100)    & 99.0 (93-103)     & 102.0 (88-108)   & 0.107      \\ \hline
 		\multicolumn{6}{l}{Values are median (inter-quartile range); \textit{p} - Kruskal-Wallis test}
 	\end{tabular}
+	\medskip
+	\begin{flushleft}
+		Pulmonary function tests performed immediately prior to cardiopulmonary exercise testing were compared between the non-jaundiced, jaundiced and severely jaundiced patients (Total n = 138).
+This did not show any relationship between obstructive jaundice and preoperative pulmonary function tests.
+\\
+		FVC - Forced Vital Capacity, FEV1 - Forced Expiratory Volume in 1 second.
+	\end{flushleft}
 \end{sidewaystable}

### 2016-02-16 Chapters/BodyComp.tex
Body comp - Post viva changes
- All figures now have extensive legends - may need trimming
- Rearranged tables and figures by adding a few \clearpages
- Recheck layout before final version

@@
 		\includegraphics[width=\textwidth]{Figures/bc_ct_sat}
-		\caption{Subcutaneous Adipose Tissue$^*$}
+		\caption{Subcutaneous Adipose Tissue}
 		\label{fig:bc_ct_sat}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_ct_vat}
-		\caption{Visceral Adipose Tissue$^*$}
+		\caption{Visceral Adipose Tissue}
 		\label{fig:bc_ct_vat}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_ct_sm}
-		\caption{Skeletal Muscle$^*$}
+		\caption{Skeletal Muscle}
 		\label{fig:bc_ct_sm}
@@
 
-	\caption{Selection of components of body composition from CT images using GIMP.}($^*$ The selected area has been removed for representation purposes. 
-The inset confirms the area selected.)
+	\caption[Selection of components of body composition from CT images using GIMP.]
+	{Selection of components of body composition from CT images using the GNU Image Manipulation Program (GIMP).\\ The total cross-sectional area was selected by first selecting the empty space outside the image followed by inverting this selection as depicted in (A).
+The 'Fuzzy Select' tool was then used to select individual body composition components with visual confirmation using the 'layer selection' tool.
+This is shown in B, C and D.
+The selected area has been removed for representation purposes.
+The insets in each image confirm the area selected.}
 	\label{fig:bc_ct_gimp}
-
 \end{figure}
@@
 	\input{Tables/c4_t1}
+\clearpage
 
@@
 	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
-	\caption{Differences in body composition according to sex and body mass index.}
+	\caption[Differences in body composition according to sex and body mass index.]
+	{Differences in body composition according to sex and body mass index.
+	\\Body composition differed significantly between men and women and with increasing body mass index (BMI).
+The proportion of skeletal muscle area decreased with increasing BMI in both sexes but there was a larger decrease in women than men.
+Visceral adipose tissue contributed more to weight in obese men than subcutaneous adipose tissue while the opposite was true in women.}
 	\label{fig:bc_gender_bmi}
@@
 68\% SAT).
+
+\clearpage
 
@@
 \subsection{Body composition vs. exercise load}
+
+Exercise loads achieved at anaerobic threshold and at peak exercise capacity were plotted against skeletal muscle area to create scatter-plots(Fig.
+\ref{fig:bc_load_vs_sm} on page \pageref{fig:bc_load_vs_sm}).
+Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($\rho^2 = 0.284, p < 0.001$, Fig.
+\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($\rho^2 = 0.350, p < 0.001$, Fig.
+\ref{fig:bc_pk_load_vs_sm}).
+However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($\rho^2 = 0.004, p = 0.587$) or peak exercise ($\rho^2 = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.
 
@@
 	\end{subfigure}
-	\caption{Correlation between exercise load and skeletal muscle area.}	
+	\caption[Correlation between exercise load and skeletal muscle area.]
+	{Correlation between exercise load and skeletal muscle area.
+		\\ Scatter plots A and B show a positive correlation between exercise load achieved and skeletal muscle area both at anaerobic threshold ($p < 0.001$) and peak exercise ($p < 0.001$) respectively.}	
 	\label{fig:bc_load_vs_sm}
@@
 
-Exercise loads achieved at anaerobic threshold and at peak exercise capacity were plotted against skeletal muscle area to create scatter-plots(Fig.
-\ref{fig:bc_load_vs_sm} on page \pageref{fig:bc_load_vs_sm}).
-Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($\rho^2 = 0.284, p < 0.001$, Fig.
-\ref{fig:bc_at_load_vs_sm}) and at peak exercise ($\rho^2 = 0.350, p < 0.001$, Fig.
-\ref{fig:bc_pk_load_vs_sm}).
-However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($\rho^2 = 0.004, p = 0.587$) or peak exercise ($\rho^2 = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.
-
 %This would suggest as one would expect that exercise capacity is mostly related to skeletal muscle volume rather than subcutaneous adipose tissue. - Move this to discussion
@@
 \subsection{Body composition vs. oxygen consumption}
-
-\input{Tables/bc_cpet}
 
@@
 	\end{subfigure}
-	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.}	
+	\caption[Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.]
+	{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
+	\\ Scatter-plots were generated comparing absolute $\dot{V}_{O_2}$ (litres/minute, solid lines) and $\dot{V}_{O_2}$ normalised for total body weight (corrected $\dot{V}_{O_2}$, ml/kg/min, dotted lines) against skeletal muscle area (A) and total adipose tissue area (B).
+The solid line in A shows a positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area ($p<0.001$) while in B it shows no correlation with total adipose tissue as expected of exercise physiology.
+However, normalisation of $\dot{V}_{O_2}$ using total body weight results in a 'spurious' negative correlation with total adipose tisue area as depicted by the dotted line in B ($p<0.001$).
+This indicates that patients with high body mass index and excess adipose tissue will have low corrected $\dot{V}_{O_2}$, the value that is commonly used in clinical practice.}
 	\label{fig:bc_scatter_VO2_bodycomp_reversal}
@@
 
+\input{Tables/bc_cpet}
+\clearpage
 \subsection{Scaling oxygen consumption to body composition and other factors}
@@
 	\end{subfigure}
-	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.}
+	\caption[Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.]
+	{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.
+	\\ Fig.
+A shows that there is little difference in the relationship between skeletal muscle area and $\dot{V}_{O_2}$ corrected using either total body weight or estimated lean body mass.
+However, while $\dot{V}_{O_2}$ corrected for total body weight had a 'spurious' negative correlation with total adipose tissue area (solid line in B), this 'spurious' relationship was no longer present when $\dot{V}_{O_2}$ was normalised using estimated lean body mass (dotted line in B).}
 	\label{fig:bc_scatter_VO2_bodycomp_elbm}

### 2016-02-16 Chapters/BodyComp.tex
Body comp - more changes
- Some page layout changes

@@
 \subsection{Body composition vs. BMI/sex}
-The body composition differences between male and female patients with increasing body mass index is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. 
-There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females. 
-Men generally had less subcutaneous fat but more visceral fat and skeletal muscle areas. 
-However, the proportion of skeletal muscle in both males and females decreased significantly with increasing body mass index.
 
-\begin{figure}[h]
+\begin{figure}[htb]
 	\centering
@@
 
+The body composition differences between male and female patients with increasing body mass index is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. 
+There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females. 
+Men generally had less subcutaneous fat but more visceral fat and skeletal muscle areas. 
+However, the proportion of skeletal muscle in both males and females decreased significantly with increasing body mass index.
+
 The proportion of skeletal muscle area decreased from 38\% in male patients with normal BMI to 22\% in obese males. 
@@
 
-\clearpage
+%\clearpage
 

### 2016-02-16 Tables/bc_at_new_indices.tex
Body comp - Paper review complete
- All is well.....I think
- Await final comments from Donny
Body Comp Tables
- Legends, units, etc.
- I think this is enough for now

@@
 
-		Variable              & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p     & $\rho$ &    p    & $\rho$ &    p \\ \hline
-		Subcut. fat (cm2)     & 0.191  &  0.089   & -0.326 &  0.003   & 0.169  &  0.132   & -0.217 &  0.051   & 0.018  &  0.872  & -0.041 & 0.716 \\
-		Visceral fat (cm2)    & 0.184  &  0.103   & -0.365 &  0.001   & 0.154  &  0.168   & -0.279 &  0.012   & -0.065 &  0.563   & -0.108 & 0.339\\
-		Total fat (cm2)       & 0.231  &  0.038   & -0.400 & $<$0.001 & 0.191  &  0.088   & -0.286 &  0.010   & -0.021 &  0.851   & -0.082 & 0.466\\
-		Skeletal Muscle (cm2) & 0.463  & $<$0.001 & -0.078 &  0.487   & 0.329  &  0.003   & 0.105  &  0.353   & -0.429 & $<$0.001 & 0.133 & 0.237 \\
-		Height                & 0.446  & $<$0.001 & 0.019  &  0.870   & 0.001  &  0.993   & 0.475  & $<$0.001 & 0.105  &  0.350   & 0.019 & 0.865\\
-		Weight                & 0.545  & $<$0.001 & -0.302 &  0.006   & 0.358  &  0.001   & -0.043 &  0.703   & 0.001  &  0.995   & 0.033 & 0.770\\
-		Body Mass Index       & 0.357  &  0.001   & -0.363 &  0.001   & 0.425  & $<$0.001 & -0.345 &  0.002   & -0.059 &  0.603   & 0.035 & 0.756\\ \hline
-		\multicolumn{11}{l}{$\rho$ - Pearson's r adjusted for sex.}
+		Variable              	& $\rho$ &\textit{p}& $\rho$ &\textit{p}& $\rho$ &\textit{p}& $\rho$ &\textit{p}& $\rho$ &\textit{p}& $\rho$ &\textit{p} \\ \hline
+		Subcut. fat (cm$^2$)     & 0.191  &  0.089   & -0.326 &  0.003   & 0.169  &  0.132   & -0.217 &  0.051   & 0.018  &  0.872   & -0.041 & 0.716 \\
+		Visceral fat (cm$^2$)    & 0.184  &  0.103   & -0.365 &  0.001   & 0.154  &  0.168   & -0.279 &  0.012   & -0.065 &  0.563   & -0.108 & 0.339\\
+		Total fat (cm$^2$)       & 0.231  &  0.038   & -0.400 & $<$0.001 & 0.191  &  0.088   & -0.286 &  0.010   & -0.021 &  0.851   & -0.082 & 0.466\\
+		Skeletal Muscle (cm$^2$) & 0.463  & $<$0.001 & -0.078 &  0.487   & 0.329  &  0.003   & 0.105  &  0.353   & -0.429 & $<$0.001 & 0.133 & 0.237 \\
+		Height (cm)               & 0.446  & $<$0.001 & 0.019  &  0.870   & 0.001  &  0.993   & 0.475  & $<$0.001 & 0.105  &  0.350   & 0.019 & 0.865\\
+		Weight (kg)               & 0.545  & $<$0.001 & -0.302 &  0.006   & 0.358  &  0.001   & -0.043 &  0.703   & 0.001  &  0.995   & 0.033 & 0.770\\
+		Body Mass Index (kg/m$^2$)      & 0.357  &  0.001   & -0.363 &  0.001   & 0.425  & $<$0.001 & -0.345 &  0.002   & -0.059 &  0.603   & 0.035 & 0.756\\ \hline
+		\multicolumn{11}{l}{$\rho$ - Pearson's partial correlation adjusted for sex.}
 	\end{tabular}
+	\medskip
+	\begin{flushleft}
+		$\dot{V}_{O_2}$AT normalised using different measures of body habitus (including weight, square of height, body mass index, skeletal muscle area and estimated lean body mass) was compared to body composition as well as weight, height and BMI in an exploratory analysis.
+Scaling $\dot{V}_{O_2}$AT using estimated lean body mass was the only method studied that did not result in any 'spurious' correlations with body composition or body habitus.
+This is also depicted in Fig.
+\ref{fig:bc_scatter_VO2_bodycomp_elbm} on page \pageref{fig:bc_scatter_VO2_bodycomp_elbm}.
+	\end{flushleft}
 \end{sidewaystable}

### 2016-02-19 Chapters/CRP_Complications.tex
Post viva changes  - CRP Complications
- Legends, units to tables/figures
- Minor layout adjustments

@@
 
+\clearpage
+
 %CRP trends and pancreatic fistula
@@
 
+\subsection{Postoperative CRP vs. infective complications}
+The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
+\ref{fig:crp_comp_infective_leak0}).
+%Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
+However, there was no such association between CRP and infective complications in patients who developed a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
+
+For instance, the median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
+(p$<$0.001).
+However, in patients who had a POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
+(p = 0.559)
+
+Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_infective_leak} on page \pageref{fig:crp_comp_infective_leak}.
+
+\clearpage
 %========================CRP vs POPF============================================
 %CRP trends in a 2-panel figure and table showing p-values below it
-\clearpage
 \begin{figure}[t]
-	\caption{Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula (POPF).}
-	\label{fig:crp_comp_crp_popf}
 	\centering
@@
 	\end{subfigure}
+	\caption[Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula.]
+	{Serum CRP levels in the first week after pancreaticoduodenectomy in patients with postoperative pancreatic fistula (POPF).
+		\\ Median daily CRP levels in the first postoperative week in 188 patients undergoing pancreaticoduodenectomy were compared between patients who developed a postoperative pancreatic fistula (POPF) and those who did not develop a POPF.
+Figure A shows a significantly raised CRP in the presence of a pancreatic fistula.
+However, severity of POPF did not affect CRP levels in the first postoperative week as shown in Figure B.
+Error bars represent 95\% confidence interval for the median.
+	}
+	\label{fig:crp_comp_crp_popf}
 \end{figure}
-\hfill
+%\hfill
 \input{Tables/crp_comp_vs_POPF_ISGPS_p_values_only}
@@
 
-\subsection{Postoperative CRP vs. infective complications}
-The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
-\ref{fig:crp_comp_infective_leak0}).
-Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
-However, there was no such association in patients who developed a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
-
-For instance, the median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
-(p$<$0.001).
-However, in patients who had a POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
-(p = 0.559)
-
-Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_infective_leak} on page \pageref{fig:crp_comp_infective_leak}.
-
 %========================CRP vs Infections====================================
 %CRP trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
-\clearpage
 \begin{figure}[t]
-	\caption{Relationship between postoperative CRP and clinically significant infective complications in patients without (A) and with (B) POPF.}
 	\label{fig:crp_comp_infective_leak}
@@
 	\end{subfigure}
+	\caption[Relationship between postoperative CRP and clinically significant infective complications in patients without (A) and with (B) POPF.]
+	{Relationship between postoperative CRP and clinically significant infective complications in patients without (A) and with (B) POPF.
+		\\Median CRP was significantly higher in patients who developed infective complications in the absence of a pancreatic fistula (A).
+However, this difference was not seen in the presence of a pancreatic fistula (B).
+Error bars represent 95\% confidence interval for the median.
+	}
 \end{figure}
-\vfill
+
 \input{Tables/crp_comp_vs_infections_popf_y1n0}
@@
 \begin{figure}[t]
-	\caption{Receiver operating characteristics analysis of C-reactive protein as a marker of postoperative infective complications.}
-	\label{fig:crp_comp_ROC_infection}
-
 	\centering
@@
 	\end{subfigure}
+	\caption[Receiver operating characteristics analysis of C-reactive protein as a marker of postoperative infective complications.]
+	{Receiver operating characteristics analysis of C-reactive protein as a marker of postoperative infective complications.
+		\\In the 126 patients who did not develop a postoperative pancreatic fistula, C-reactive protein levels as early as the third and fourth postoperative days had a high negative predictive value of 0.79 (for Day 3  CRP $\leq$ 178 mg/l) and 0.83 (for Day 4 CRP $\leq$125 mg/l ) respectively for the prediction of infective complications.}
+	\label{fig:crp_comp_ROC_infection}
 \end{figure}

### 2016-02-19 Chapters/CRP_Complications.tex
Using \caption*

--- 2015-08-27 Chapters/CRP_Complications.tex
@@
 
+\subsection{Postoperative CRP vs. infective complications}
+The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
+\ref{fig:crp_comp_infective_leak0}).
+%Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
+However, there was no such association between CRP and infective complications in patients who developed a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
+
+For instance, the median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
+(p$<$0.001).
+However, in patients who had a POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
+(p = 0.559)
+
+Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
+\ref{fig:crp_comp_infective_leak} on page \pageref{fig:crp_comp_infective_leak}.
+
+\clearpage
 %========================CRP vs POPF============================================
 %CRP trends in a 2-panel figure and table showing p-values below it
-\clearpage
 \begin{figure}[t]
@@
 	\end{subfigure}
+
+	\caption*{Median daily CRP levels in the first postoperative week in 188 patients undergoing pancreaticoduodenectomy were compared between patients who developed a postoperative pancreatic fistula (POPF) and those who did not develop a POPF.
+Figure A shows a significantly raised CRP in the presence of a pancreatic fistula.
+However, severity of POPF did not affect CRP levels in the first postoperative week as shown in Figure B.
+Error bars represent 95\% confidence interval for the median.}
 \end{figure}
-\hfill
+%\hfill
 \input{Tables/crp_comp_vs_POPF_ISGPS_p_values_only}
@@
 
-\subsection{Postoperative CRP vs. infective complications}
-The median CRP levels on or after postoperative day 3 were significantly higher in patients who developed clinically significant infective complications in the absence of a POPF (Fig.
-\ref{fig:crp_comp_infective_leak0}).
-Moreover, the median difference in CRP also increases from 56 mg/L on the third day to 79 mg/L on the seventh day. 
-However, there was no such association in patients who developed a POPF (Fig. \ref{fig:crp_comp_infective_leak1}).
-
-For instance, the median CRP on Day 4 in patients who did not develop a POPF and did not have an infective complication was 97 mg/L (IQR 71 - 174) but was 173 mg/L if they had an infective complication.
-(p$<$0.001).
-However, in patients who had a POPF, the CRP on day 4 was 214 mg/L (IQR 151 - 250) in the absence of infective complications and 190 mg/L (IQR 137 - 252) in the presence of infective complications.
-(p = 0.559)
-
-Therefore, CRP on or after the third postoperative day appears to be useful in predicting infective complications only in patients who did not develop a POPF and these results are shown in Table \ref{table:crp_comp_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_infective_leak} on page \pageref{fig:crp_comp_infective_leak}.
-
 %========================CRP vs Infections====================================
 %CRP trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
-\clearpage
 \begin{figure}[t]
@@
 	\end{subfigure}
+	\caption*{Median CRP was significantly higher in patients who developed infective complications in the absence of a pancreatic fistula (A).
+This difference was not seen in the presence of POPF(B).
+Error bars - 95\% confidence interval for the median.}
 \end{figure}
-\vfill
+
 \input{Tables/crp_comp_vs_infections_popf_y1n0}
@@
 	\label{fig:crp_comp_ROC_infection}
-
 	\centering
@@
 	\end{subfigure}
+	\caption*{In the 126 patients who did not develop a postoperative pancreatic fistula, C-reactive protein levels as early as the third and fourth postoperative days had a high negative predictive value of 0.79 (for Day 3  CRP $\leq$ 178 mg/l) and 0.83 (for Day 4 CRP $\leq$125 mg/l ) respectively for the prediction of infective complications.}
 \end{figure}

### 2016-02-19 Chapters/CPET_Jaundice.tex
\caption* in CPET_jaundice
- replace all legends with \caption*

@@
 \begin{figure}[htbp]
-
-	\caption{Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak.}
-	\label{fig:cpet_oj_scatter}
-
 	\centering
@@
 	\end{subfigure}
-
-	\begin{flushleft}
-		The relationship between serum bilirubin and $\dot{V}_{O_2}$AT was weak (Pearson's $\rho$ = - 0.187, $\rho^2$ = 0.035, p = 0.028).
-There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)
-	\end{flushleft}
+	\caption{Scatter-plot analysis comparing serum bilirubin versus $\dot{V}_{O_2}$AT and $\dot{V}_{O_2}$Peak.}
+	\label{fig:cpet_oj_scatter}
+	\caption*{The relationship between serum bilirubin and $\dot{V}_{O_2}$AT was weak (Pearson's $\rho$ = - 0.187, $\rho^2$ = 0.035, p = 0.028).
+There was no correlation between serum bilirubin and $\dot{V}_{O_2}$Peak (Pearson's $\rho$ = - 0.132, $\rho^2$ = 0.017, p = 0.123)}
 

### 2016-02-19 Tables/cpet_oj_pft.tex
\caption* in CPET_jaundice
- replace all legends with \caption*

@@
 	\medskip
-	\begin{flushleft}
-		Pulmonary function tests performed immediately prior to cardiopulmonary exercise testing were compared between the non-jaundiced, jaundiced and severely jaundiced patients (Total n = 138).
+	\caption*{Pulmonary function tests performed immediately prior to cardiopulmonary exercise testing were compared between the non-jaundiced, jaundiced and severely jaundiced patients (Total n = 138).
 This did not show any relationship between obstructive jaundice and preoperative pulmonary function tests.
-\\
-		FVC - Forced Vital Capacity, FEV1 - Forced Expiratory Volume in 1 second.
-	\end{flushleft}
+FVC - Forced Vital Capacity, FEV1 - Forced Expiratory Volume in 1 second.}
 \end{sidewaystable}

### 2016-02-19 Chapters/BodyComp.tex
\caption* in Body comp

@@
 
-	\caption[Selection of components of body composition from CT images using GIMP.]
-	{Selection of components of body composition from CT images using the GNU Image Manipulation Program (GIMP).\\ The total cross-sectional area was selected by first selecting the empty space outside the image followed by inverting this selection as depicted in (A).
+	\caption{Selection of components of body composition from CT images using GIMP.}
+	\caption*{The total cross-sectional area was selected by first selecting the empty space outside the image followed by inverting this selection as depicted in (A).
 The 'Fuzzy Select' tool was then used to select individual body composition components with visual confirmation using the 'layer selection' tool.
@@
 	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
-	\caption[Differences in body composition according to sex and body mass index.]
-	{Differences in body composition according to sex and body mass index.
-		\\Body composition differed significantly between men and women and with increasing body mass index (BMI).
+	\caption{Differences in body composition according to sex and BMI.}
+	\caption*{Body composition differed significantly between men and women and with increasing body mass index (BMI).
 The proportion of skeletal muscle area decreased with increasing BMI in both sexes but there was a larger decrease in women than men.
@@
 	\end{subfigure}
-	\caption[Correlation between exercise load and skeletal muscle area.]
-	{Correlation between exercise load and skeletal muscle area.
-		\\ Scatter plots A and B show a positive correlation between exercise load achieved and skeletal muscle area both at anaerobic threshold ($p < 0.001$) and peak exercise ($p < 0.001$) respectively.}	
+	\caption{Correlation between exercise load and skeletal muscle area.}
+	\caption*{Scatter plots A and B show a positive correlation between exercise load achieved and skeletal muscle area both at anaerobic threshold ($p < 0.001$) and peak exercise ($p < 0.001$) respectively.}
 	\label{fig:bc_load_vs_sm}
@@
 		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT}
-		\caption{$\dot{V}_{O_2}$AT vs. 
-Total Adipose Tissue}
+		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
 		\label{fig:bc_scatter_VO2_TAT}
 	\end{subfigure}
-	\caption[Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.]
-	{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
-	\\ Scatter-plots were generated comparing absolute $\dot{V}_{O_2}$ (litres/minute, solid lines) and $\dot{V}_{O_2}$ normalised for total body weight (corrected $\dot{V}_{O_2}$, ml/kg/min, dotted lines) against skeletal muscle area (A) and total adipose tissue area (B).
+	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.}
+	\caption*{%Scatter-plots were generated comparing absolute $\dot{V}_{O_2}$ (litres/minute, solid lines) and $\dot{V}_{O_2}$ normalised for total body weight (corrected $\dot{V}_{O_2}$, ml/kg/min, dotted lines) against skeletal muscle area (A) and total adipose tissue area (B).
 	 The solid line in A shows a positive correlation between absolute $\dot{V}_{O_2}$ and skeletal muscle area ($p<0.001$) while in B it shows no correlation with total adipose tissue as expected of exercise physiology.
@@
 	\end{subfigure}
-	\caption[Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.]
-	{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.
-	\\ Fig.
+	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.}
+	\caption*{Fig.
 A shows that there is little difference in the relationship between skeletal muscle area and $\dot{V}_{O_2}$ corrected using either total body weight or estimated lean body mass.

### 2016-02-19 Tables/bc_at_new_indices.tex
\caption* in Body comp

@@
 	\medskip
-	\begin{flushleft}
-		$\dot{V}_{O_2}$AT normalised using different measures of body habitus (including weight, square of height, body mass index, skeletal muscle area and estimated lean body mass) was compared to body composition as well as weight, height and BMI in an exploratory analysis.
+	\caption*{$\dot{V}_{O_2}$AT normalised using different measures of body habitus (including weight, square of height, body mass index, skeletal muscle area and estimated lean body mass) was compared to body composition as well as weight, height and BMI in an exploratory analysis.
 Scaling $\dot{V}_{O_2}$AT using estimated lean body mass was the only method studied that did not result in any 'spurious' correlations with body composition or body habitus.
 This is also depicted in Fig.
-\ref{fig:bc_scatter_VO2_bodycomp_elbm} on page \pageref{fig:bc_scatter_VO2_bodycomp_elbm}.
-	\end{flushleft}
+\ref{fig:bc_scatter_VO2_bodycomp_elbm} on page \pageref{fig:bc_scatter_VO2_bodycomp_elbm}.}
 \end{sidewaystable}

### 2016-02-19 Chapters/CPET_Outcomes.tex
Viva Recommendations:
List of abbreviations
All tables should have units, legends
All values and ranges need defining
Chapter 1
 - More discussion of key RCTs
- Short discussion of etiology
Chapter 2
- Review patient numbers on p63 - pathologies dont add up to total
- Aim and end points must match
- Error in AT numbers in adjuvant therapy table
- P72 Remove unjustifed/unsupported conclusion
- P73 Remove GPS conclusion
Chapter 3 p96 Remove conclusions not based on thesis
Chapter 5-6 - Reduce to more undestandable key issues. All tables need legends. Several irrelevant tables should be removed.

@@
 \subsection{Aim}
-The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative adverse events as well as fitness for adjuvant therapy in patients undergoing major pancreatic surgery.
-
+The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative length of stay, major postoperative adverse events including operative mortality and fitness to undergo adjuvant therapy when indicated after pancreaticoduodenectomy.
+%Fixed this to match primary outcomes on 19/02/16
 \clearpage
@@
 Of these, 100 patients (pancreaticoduodenectomy - 98, sub-total/total pancreatectomy - 2) had undergone cardiopulmonary exercise testing as part of their preoperative assessment and were included in the study.
-Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=4) or chronic pancreatitis (n=2).
+Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=6) or chronic pancreatitis (n=5).
+\todo{These numbers don't add up - check them.}
 
@@
 Fifty-five patients were included in the analysis. 
-Patients were excluded if chemotherapy was not indicated (n=28), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6).
+Patients were excluded if chemotherapy was not indicated (n=28, ampullary carcinoma=8, neuroendocrine tumours=7, IPMN=4, other neoplasia=5, chronic pancreatitis=4), in the event of operative mortality (n=7), if chemotherapy was offered but declined by the patient (n=4), or where they had not been seen by an oncologist yet (n=6).
 On binary logistic regression analysis, $\dot{V}_{O_2}$AT less than 10 ml/kg/min was the only preoperative factor that was associated with with non-receipt of adjuvant therapy (HR 6.30, 95\% CI 1.25-31.75, p=0.026).
@@
 Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.
+\todo{Remove unjustified/unsupported conclusion}
+\todo{Remove GPS conclusion}
 

### 2016-03-13 MD_corrections_response.tex
Response letter

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\author{Vishnu V Chandrabalan}
\begin{document}

\section{Chapter 1 - More discussion of key RCTs}

The following paragraph has been included as Section 1.x in Chapter 1.

\section{Chapter 1 - Short discussion of aetiology needed}

\section{Chapter 2 - Review patient numbers on page 63}

Patient numbers have been updated to account for the missing information and this section now reads as below accounting for the 100 patients included in this study.

\textit{Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=6) or chronic pancreatitis (n=5).}

\section{Chapter 2 - Aim and end points must match}

The description of the aims of this chapter has been updated to match the end-points described later in methods and results sections and now reads as below.

\textit{The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative length of stay, major postoperative adverse events including operative mortality and fitness to undergo adjuvant therapy when indicated after pancreaticoduodenectomy.}

\section{Page 69 - Error in AT numbers}

The results presented in this table have been reanalysed using the primary database and have been found to be accurate. In the group of patients where adjuvant chemotherapy was indicated, anaerobic threshold was low in a greater proportion. However, low anaerobic threshold was associated with non-receipt of adjuvant therapy in spite of being indicated as shown in this table.

\section{Page 72 - Remove unjustified/unsupported conclusion, Page 73 - Remove GPS conclusion}

I am not entirely sure how to address this. Should I completely remove this sentence?

The following sentence has been removed from the discussion in Chapter 2.

\textit{Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.}

\section{Chapter 3, page 96 - Remove conclusions not based on thesis}

The following sentence has been removed from the discussion in Chapter 3.

\textit{Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.}

\section{Chapter 5,6 - Reduce to more understandable key issues. Several irrelevant tables should be removed.}

\section{All acronyms/abbreviations on list}

A list of abbreviations has been included.

\section{All tables should have units (e.g. mg/L, etc) and legends.}

All tables have been updated to include units. A brief description has been added as a legend to all figures and tables.

\section{All values and ranges need defining}

\end{document}

### 2016-03-14 response_to_reviewer’s_comment.md
Response letter in Markdown
- response.tex deleted
- new markdown file created

# Response to Reviewer’s comments

**An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.**

Dear Sir,

I am grateful for the opportunity to respond to the examiners' comments. I have considered the examiners' comments carefully and these have improved the thesis. Please find my responses to the comments below.

## 1. Chapter 1 - More discussion of key RCTs

The following paragraph has been included as Section 1.x in Chapter 1.

## 2. Chapter 1 - Short discussion of aetiology needed

Point taken. The following subsection titled ’Aetiology of pancreatic cancer’ has been included in Chapter 1 under Section 1.1.
Several factors have been identified as being associated with an increased risk of pancreatic cancer. The most important risk factor associated with increased incidence of pancreatic cancer is tobacco. Obesity has also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms are unclear. Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes. It is of note that some of the important risk factors such as smoking, obesity, poor diet and increasing age may also be associated with other medical comorbidities including cardiorespiratory diseases that may have a significant impact on patient fitness.

## 3. Chapter 2 - Review patient numbers on page 63

Point taken. Patient numbers have been updated to account for the missing information and this section now reads as below accounting for all the 100 patients included in this study.

Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=6) or chronic pancreatitis (n=5).

## 4 Chapter 2 - Aim and end points must match

Point taken. The aims of this chapter have been updated as below to match the end-points described later in methods and results sections.

The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative length of stay, major postoperative adverse events including operative mortality and fitness to undergo adjuvant therapy when indicated after pancreaticoduodenectomy.

## 5 Page 69 - Error in AT numbers

The results presented in this table have been re-analysed using the primary database and have been found to be accurate. In the group of patients where adjuvant chemotherapy was indicated, anaerobic threshold was low in a greater proportion. However, low anaerobic threshold was associated with non-receipt of adjuvant therapy in spite of being indicated as shown in this table.

# 6 Page 72 - Remove unjustified/unsupported conclusion, Page 73 - Remove GPS conclusion

I am not entirely sure how to address this. Should I completely remove this sentence?

Point taken. The following sentence has been removed from the discussion in Chapter 2.

Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.

## 7 Chapter 3, page 96 - Remove conclusions not based on thesis
Point taken. The following sentence has been removed from the discussion in Chapter 3.

Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.

## 8 Chapter 5,6 - Reduce to more understandable key issues. Several irrelevant tables should be removed.
Can we talk about this over the phone, please?

## 9 All acronyms/abbreviations on list
A list of abbreviations has been included.

## 10 All tables should have units (e.g. mg/L, etc) and legends.
Point taken. All tables have been updated to include units. Legends have been added to all figures and tables providing a brief description of the results
depicted.

## 11 All values and ranges need defining.
Are they looking for normal ranges for the various parameters like CRP,albumin, etc.?

### 2016-03-25 Chapters/Introduction.tex
Viva Recommendations:
List of abbreviations
All tables should have units, legends
All values and ranges need defining
Chapter 1
 - More discussion of key RCTs
- Short discussion of etiology
Chapter 2
- Review patient numbers on p63 - pathologies dont add up to total
- Aim and end points must match
- Error in AT numbers in adjuvant therapy table
- P72 Remove unjustifed/unsupported conclusion
- P73 Remove GPS conclusion
Chapter 3 p96 Remove conclusions not based on thesis
Chapter 5-6 - Reduce to more undestandable key issues. All tables need legends. Several irrelevant tables should be removed.

@@
 %----------------------------------------------------------------------------------------
-
+\todo{More discussion of key rcts}
 \section{Pancreatic neoplasia}
@@
 The ability of the cardiac and respiratory systems in particular, to cope with the increased physiological demand placed by general anaesthesia and major pancreatic surgery plays an important role in determining outcome after surgery.
+
+\todo{Short discussion of etiology}
+\subsection{Aetiology of pancreatic cancer}
+Several factors have been identified as being associated with an increased risk of pancreatic cancer.
+The most important risk factor associated with increased incidence of pancreatic cancer is tobacco. 
+Obesity has also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms are unclear.
+Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes.
+It is of note that some of the important risk factors such as smoking, obesity, poor diet and increasing age may also be associated with other medical comorbidities including cardiorespiratory diseases that may have a significant impact on patient fitness.
 
@@
 
+Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery verus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma \parencite{oettle_adjuvant_2007}.
+The median disease free survival in patient who received gemcitabine chemotherapy was 13.4 months while it was only 6.9 months in the surgery-only group.
+However, there was no difference in overall survival.
+
 However, 25\% to 50\% of patients who undergo surgery for pancreatic ductal adenocarcinoma either do not receive or do not complete a full course of adjuvant chemotherapy \parencite{sohn_resected_2000, neoptolemos_randomized_2004, oettle_adjuvant_2007}.
@@
 Lin and co-workers reported that in a study of 1891 patients undergoing pancreaticoduodenectomy, only soft pancreatic gland texture and coronary disease were related to postoperative pancreatic fistula on multivariate analysis \parencite{lin_risk_2004}.
-These pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.
+
+Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
+A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival \parencite{diener_pancreaticoduodenectomy_2008}.
+A prospective randomised trial comparing pancreaticogastrostomy versus pancreaticojejunostomy in patients undergoing pancreaticoduodenectomy reported that there was no difference in the incidence of overall complications or pancreatic fistula.
+However, patients who had a pancreaticogastrostomy had a significantly lower rate of biliary fistula, postoperative collections, delayed gastric emptying as well as lower incidence of multiple surgical complications \parencite{bassi_reconstruction_2005}.
+Surgical drains after pancreaticoduodenectomy have been a subject of considerable debate. 
+A randomised trial comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications.
+Hospital stay and costs were also lower \parencite{bassi_early_2010}.
+
+However, these pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.
 

### 2016-03-25 response_to_reviewer’s_comment.tex
Response to reviewers comments Tex file
- Finally settled for tex version of this.
- Don't effing dick around with md or txt again

--- 2016-03-14 response_to_reviewer’s_comment.md
@@
-# Response to Reviewer’s comments
+\documentclass[12pt,a4paper]{article}
+\usepackage[utf8]{inputenc}
+%\usepackage{amsmath}
+%\usepackage{amsfonts}
+%\usepackage{amssymb}
+%\usepackage{graphicx}
+\author{Vishnu V Chandrabalan}
+\begin{document}
 
-**An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.**
+\textbf{Response to Reviewer’s comments}
+
+\textbf{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.}
 
@@
 
-## 1. Chapter 1 - More discussion of key RCTs
+\textbf{Chapter 1 - More discussion of key RCTs}
 
-The following paragraph has been included as Section 1.x in Chapter 1.
+	Point taken.	
+	The following paragraph has been included under Section 1.4.4 in Chapter 1 and discusses some of the key randomised trials of operative strategies aimed at reducing postoperative complications after pancreaticoduodenectomy.
 
-## 2. Chapter 1 - Short discussion of aetiology needed
+	Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
+	A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival.
+	A prospective randomised trial comparing pancreaticogastrostomy versus pancreaticojejunostomy in patients undergoing pancreaticoduodenectomy reported that there was no difference in the incidence of overall complications or pancreatic fistula.
+	However, patients who had a pancreaticogastrostomy had a significantly lower rate of biliary fistula, postoperative collections, delayed gastric emptying as well as lower incidence of multiple surgical complications.
+	Surgical drains after pancreaticoduodenectomy have been a subject of considerable debate. 
+	A randomised trial comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications.
+	Hospital stay and costs were also lower.
 
-Point taken. The following subsection titled ’Aetiology of pancreatic cancer’ has been included in Chapter 1 under Section 1.1.
+	The following has been added to Section 1.3 in Chapter 1 and adds to the already detailed discussion of some of the key randomised trials in adjuvant therapy after pancreaticoduodenectomy.
+
+	Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery verus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma.
+	The median disease free survival in patient who received gemcitabine chemotherapy was 13.4 months while it was only 6.9 months in the surgery-only group.
+However, there was no difference in overall survival.
+
+\textbf{Chapter 1 - Short discussion of aetiology needed}
+
+	Point taken. The following subsection titled 'Aetiology of pancreatic cancer' has been included in Chapter 1 under Section 1.1.
+
 	Several factors have been identified as being associated with an increased risk of pancreatic cancer. 
@@
 
-## 3. Chapter 2 - Review patient numbers on page 63
+\textbf{Chapter 2 - Review patient numbers on page 63}
 
@@
 
-## 4 Chapter 2 - Aim and end points must match
+\textbf{Chapter 2 - Aim and end points must match }
 
@@
 
-## 5 Page 69 - Error in AT numbers
+\textbf{Page 69 - Error in AT numbers}
 
-The results presented in this table have been re-analysed using the primary database and have been found to be accurate.
-In the group of patients where adjuvant chemotherapy was indicated, anaerobic threshold was low in a greater proportion.
-However, low anaerobic threshold was associated with non-receipt of adjuvant therapy in spite of being indicated as shown in this table.
+	The results presented in this table have been re-analysed using the primary database and have been found to be accurate. Therefore, no changes have been made.
 
-# 6 Page 72 - Remove unjustified/unsupported conclusion, Page 73 - Remove GPS conclusion
-
-I am not entirely sure how to address this. Should I completely remove this sentence?
+\textbf{Page 72 - Remove unjustified/unsupported conclusion, Page 73 - Remove GPS conclusion}
 
@@
 
-## 7 Chapter 3, page 96 - Remove conclusions not based on thesis
+\textbf{Chapter 3, page 96 - Remove conclusions not based on thesis}
+
 	Point taken. The following sentence has been removed from the discussion in Chapter 3.
@@
 
-## 8 Chapter 5,6 - Reduce to more understandable key issues. Several irrelevant tables should be removed.
+\textbf{Chapter 5,6 - Reduce to more understandable key issues. Several irrelevant tables should be removed.}
+
 	Can we talk about this over the phone, please?
 
-## 9 All acronyms/abbreviations on list
+\textbf{All acronyms/abbreviations on list}
+
 	A list of abbreviations has been included.
 
-## 10 All tables should have units (e.g. mg/L, etc) and legends.
-Point taken. All tables have been updated to include units. Legends have been added to all figures and tables providing a brief description of the results
+\textbf{All tables should have units (e.g. mg/L, etc) and legends.}
+
+	Point taken. 
+	All tables have been updated to include units. 
+	Legends have been added to all figures and tables providing a brief description of the results
 	depicted.
 
-## 11 All values and ranges need defining.
+\textbf{All values and ranges need defining.}
+
 	Are they looking for normal ranges for the various parameters like CRP,albumin, etc.?
+\end{document}

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
Comment out following:
- NLR in methods
- CRP vs postop alb
- CRP vs postop neut
- Preop hypoalb vs postop alb, neut

@@
 The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on page \pageref{table:mGPS} in Chapter \ref{ch_intro}. 
-The neutrophil-lymphocyte ratio was calculated by dividing the preoperative neutrophil count by the preoperative lymphocyte count and a threshold of 5.0 was used to dichotomise this variable. 
+%The neutrophil-lymphocyte ratio was calculated by dividing the preoperative neutrophil count by the preoperative lymphocyte count and a threshold of 5.0 was used to dichotomise this variable. 
 Standard thresholds were used to categorise other biochemical parameters. 
@@
 However, this association was not present after the first postoperative day.
-Elevated preoperative CRP was associated with a persistently lower serum albumin level for the entire duration of the first postoperative week (p$<$0.001 to p$<$0.002).
-It was also associated with a higher neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
+%Elevated preoperative CRP was associated with a persistently lower serum albumin level for the entire duration of the first postoperative week (p$<$0.001 to p$<$0.002).
+%It was also associated with a higher neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
 
@@
 This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
-Preoperative hypoalbuminemia persisted postoperatively throughout the first week (p$<$0.001, Fig. \ref{fig:sirs_alb_alb}).
-There was no relationship between preoperative serum albumin levels and postoperative neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
+%Preoperative hypoalbuminemia persisted postoperatively throughout the first week (p$<$0.001, Fig. \ref{fig:sirs_alb_alb}).
+%There was no relationship between preoperative serum albumin levels and postoperative neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
 

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
More major edits
- Remove more references to postop albumin and neutrophil counts

--- 2015-08-26 Chapters/Pre_Post_SIRS.tex
@@
 
-%Post-operative CRP levels have been reported to be associated with the magnitude of surgery as well as to predict infectious complications after neurosurgery \parencite{al-jabi_value_2010}.
+\clearpage
 
-\clearpage
 \section{Patients and methods}
@@
 The modified Glasgow Prognostic Score (mGPS) was calculated as shown in Table \ref{table:mGPS} on page \pageref{table:mGPS} in Chapter \ref{ch_intro}. 
-The neutrophil-lymphocyte ratio was calculated by dividing the preoperative neutrophil count by the preoperative lymphocyte count and a threshold of 5.0 was used to dichotomise this variable. 
 Standard thresholds were used to categorise other biochemical parameters. 
@@
 However, this association was not present after the first postoperative day.
-Elevated preoperative CRP was associated with a persistently lower serum albumin level for the entire duration of the first postoperative week (p$<$0.001 to p$<$0.002).
-It was also associated with a higher neutrophil count (p$<$0.01) from the 3$^{rd}$ to 6$^{th}$ postoperative days with a less but still significant association on the 7$^{th}$ postoperative day.
 
@@
 This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
-Preoperative hypoalbuminemia persisted postoperatively throughout the first week (p$<$0.001, Fig. \ref{fig:sirs_alb_alb}).
-There was no relationship between preoperative serum albumin levels and postoperative neutrophil count (Fig. \ref{fig:sirs_alb_neut}).
 
@@
 Obstructive jaundice was associated with significant differences in the postoperative systemic inflammatory response. 
-Postoperative serum albumin levels were lower in jaundiced patients with severely jaundiced patients having the lowest levels (p$<$0.001, Fig. \ref{fig:sirs_bil_alb}). 
-There were no significant changes in the trends of albumin levels and it is likely that preoperative hypoalbuminemia persisted postoperatively in these patients.
-
-However, the trends in postoperative CRP were significantly different between the jaundiced and non-jaundiced patients. 
+The trends in postoperative CRP were significantly different between the jaundiced and non-jaundiced patients. 
 Obstructive jaundice was associated with a significantly lower peak CRP on POD 2 (p=0.001) and this persisted until POD 6 (p=0.005, Table \ref{table:sirs_crp}, \ref{table:sirs_crp_pvalues}).
 Median postoperative CRP had an inverse relationship with severity of preoperative obstructive jaundice (Fig. \ref{fig:sirs_bil_crp}).
-Obstructive jaundice was associated with a delayed rise in the neutrophil count on POD 6 (p$<$0.05) and POD 7 (p$<$0.05) although this relationship was less significant (Fig. \ref{fig:sirs_bil_neut}).
 
@@
 $\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
-There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP or neutrophil count. 
-Serum albumin levels were lower in patients with $\dot{V}_{O_2}$AT$<$10 mls/kg/min (Fig. \ref{fig:sirs_at}).
+There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP.% or neutrophil count.
 
 There was no significant, persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
-However, postoperative hypoalbuminemia was associated with all the preoperative characteristics studied with the exception of BMI and SIMD (Tables \ref{table:sirs_alb} and \ref{table:sirs_alb_pvalues}).
 

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
Body composition and tables
- Remove body comp vs alb/neut
- Remove some of the tables

@@
 
-There was no significant, persistent relationship between postoperative CRP and neutrophil count and preoperative neutrophil-lymphocyte ratio, SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
+There was no significant, persistent relationship between postoperative CRP and preoperative SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
 
@@
 
-The relationship between body composition and postoperative CRP, albumin and neutrophil count is shown in Tables \ref{table:sirs_bodycomp_crp}, \ref{table:sirs_bodycomp_alb} and \ref{table:sirs_bodycomp_neut} respectively.
-Postoperative CRP and neutrophil count were not related to any of the components of body composition aside from a weak linear association between neutrophil count on POD 2 and visceral fat area (p=0.046).
-However, lower preoperative skeletal muscle area was associated with lower postoperative serum albumin on days 1 to 5 (all p$<$0.05).
+The relationship between body composition and postoperative CRP is shown in Table \ref{table:sirs_bodycomp_crp}.
+Postoperative CRP was not related to any of the components of body composition.
 
@@
 \input{Tables/sirs_crp_pvalues}
-
-\input{Tables/sirs_albumin}
-\input{Tables/sirs_alb_pvalues}
-
-\input{Tables/sirs_neut}
-\input{Tables/sirs_neut_pvalues}
 

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
Remove more figures and tables
- Remove multiple figures
- Combine all postop CRP figures vs preop crp, alb, bil into one figure
- Remove bodycomp vs alb/neut tables

@@
 \begin{figure}[p]
-	\caption{Relationship between preoperative CRP levels and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
+	\caption{Relationship between preoperative CRP levels and postoperative CRP in the first week after pancreaticoduodenectomy.}
 	\label{fig:sirs_crp}
@@
 		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_crp_alb}
-		\caption{Preop. CRP vs. postop. Albumin}
+		\includegraphics[width=\textwidth]{Figures/sirs_alb_crp}
+		\caption{Preop. Albumin vs. postop. CRP}
 		\label{fig:sirs_crp_alb}
@@
 		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_crp_neut}
-		\caption{Preop. CRP vs. postop. Neutrophil count}
+		\includegraphics[width=\textwidth]{Figures/sirs_bil_crp}
+		\caption{Preop. Bilirubin vs. postop. CRP}
 		\label{fig:sirs_crp_neut}
-	\end{subfigure}	
-\end{figure}
-%==============================================================================
-
-%========================Albumin vs Post-op SIRS============================================
-
-\begin{figure}[p]
-	\caption{Relationship between preoperative Albumin levels and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
-	\label{fig:sirs_alb}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_alb_crp}
-		\caption{Preop. Albumin vs. postop. CRP}
-		\label{fig:sirs_alb_crp}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_alb_alb}
-		\caption{Preop. Albumin vs. postop. Albumin}
-		\label{fig:sirs_alb_alb}
-	\end{subfigure}
-
-	\vspace{1cm}
-
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_alb_neut}
-		\caption{Preop. Albumin vs. postop. Neutrophil count}
-		\label{fig:sirs_alb_neut}
 	\end{subfigure}	
@@
 
-%========================Bilirubin vs Post-op SIRS============================================
-
-\begin{figure}[p]
-	\caption{Relationship between obstructive jaundice and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
-	\label{fig:sirs_bilirubin}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_bil_crp}
-		\caption{Preop. Bilirubin vs. postop. CRP}
-		\label{fig:sirs_bil_crp}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_bil_alb}
-		\caption{Preop. Bilirubin vs. postop. Albumin}
-		\label{fig:sirs_bil_alb}
-	\end{subfigure}
-
-	\vspace{1cm}
-
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_bil_neut}
-		\caption{Preop. Bilirubin vs. postop. Neutrophil count}
-		\label{fig:sirs_bil_neut}
-	\end{subfigure}	
-\end{figure}
-%==============================================================================
-
-%========================AT vs Post-op SIRS============================================
-
-\begin{figure}[p]
-	\caption{Relationship between $\dot{V}_{O_2}$AT and postoperative inflammatory markers in the first week after pancreaticoduodenectomy.}
-	\label{fig:sirs_at}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_at_crp}
-		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. CRP}
-		\label{fig:sirs_at_crp}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_at_alb}
-		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Albumin}
-		\label{fig:sirs_at_alb}
-	\end{subfigure}
-
-	\vspace{1cm}
-
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/sirs_at_neut}
-		\caption{Preop. $\dot{V}_{O_2}$AT vs. postop. Neutrophil count}
-		\label{fig:sirs_at_neut}
-	\end{subfigure}	
-\end{figure}
-%==============================================================================
-
 \input{Tables/sirs_bodycomp_crp}
-\input{Tables/sirs_bodycomp_alb}
-\input{Tables/sirs_bodycomp_neut}
 

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
Update results
- Remove references to old tables and figures
- Correct references to reorganised figures

@@
 
-The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels (Tables \ref{table:sirs_crp} and \ref{table:sirs_crp_pvalues}), postoperative serum albumin (Tables \ref{table:sirs_alb} and \ref{table:sirs_alb_pvalues}) and postoperative neutrophil count (Tables \ref{table:sirs_neut} and \ref{table:sirs_neut_pvalues}) are presented.
+The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels is shown in Tables \ref{table:sirs_crp} and \ref{table:sirs_crp_pvalues}).
 
 \subsection{Preoperative vs. postoperative systemic inflammation}
-The relationship between preoperative CRP and postoperative systemic inflammation is also depicted graphically in Fig. \ref{fig:sirs_crp}. 
+The relationship between preoperative CRP and postoperative systemic inflammation is depicted graphically in Fig. \ref{fig:sirs_crp_crp} and Fig. \ref{fig:sirs_crp_alb} on page \pageref{fig:sirs_crp}.
 The median CRP levels on the day of surgery and the first postoperative day (POD) were significantly higher in patients who had an elevated preoperative CRP. 
@@
 Body composition data was available for 90 patients.
-Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, 128.9 cm$^2$, 119.0-150.8) and high (n=30, 241.2 cm$^2$, 221.9-313.1).
+Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, median 128.9 cm$^2$, IQR 119.0-150.8) and high (n=30, median 241.2 cm$^2$, IQR 221.9-313.1).
 Subcutaneous adipose tissue area was divided into tertiles as low (n=30, median 94.1 cm$^2$, IQR 63.2-102.5), moderate (n=30, median 152.5 cm$^2$, IQR 141.3-175.3) and high (n=30, median 253.9 cm$^2$, IQR 203.2-290.3).
@@
 		\caption{Preop. Albumin vs. postop. CRP}
-		\label{fig:sirs_crp_alb}
+		\label{fig:sirs_alb_crp}
 	\end{subfigure}
@@
 		\caption{Preop. Bilirubin vs. postop. CRP}
-		\label{fig:sirs_crp_neut}
+		\label{fig:sirs_bil_crp}
 	\end{subfigure}	

### 2016-03-25 Chapters/Pre_Post_SIRS.tex
Discussion update

@@
 The lack of association between aerobic capacity, POSSUM Physiology Score and postoperative CRP in this study may have been due to the dominant effect of preoperative systemic inflammation and obstructive jaundice on the postoperative inflammatory response.
-However, the association of postoperative hypoalbuminemia with low aerobic capacity, POSSUM Physiology Score and anaemia appears to suggest a link between comorbidity and systemic inflammation and/or nutritional status.
-Bernal and co-workers studied 399 patients undergoing liver transplantation who had cardiopulmonary exercise testing as part of their preoperative work-up and reported that hypoalbuminemia was associated with both low $\dot{V}_{O_2}$AT and low $\dot{V}_{O_2}$Peak \parencite{bernal_aerobic_2014}.
 
 %Body Composition
-The association between hypoalbuminemia and low aerobic capacity may be explained by lower skeletal muscle and greater subcutaneous fat in patients with persistent postoperative hypoalbuminemia.
+
 Richards and co-workers reported that elevated preoperative systemic inflammation was related to low skeletal muscle index \parencite{richards_relationships_2012}.
 Hassen and co-workers reported that fat-free mass and skeletal muscle were inversely related to the severity of the postoperative inflammatory response following endo-vascular surgery for abdominal aortic aneurysm \parencite{hassen_preoperative_2007}.
-
-Elevated systemic inflammation is related to obesity as well as sarcopenia. 
+Elevated systemic inflammation has been reported to be related to obesity as well as sarcopenia. 
 `Sarcopenic-obesity' in patients with pancreatic cancer where there is relative loss of skeletal muscle with preservation of adipose tissue puts these patients at particular risk for immune dysfunction \parencite{berg_adipose_2005, reisinger_sarcopenia_2015}, post-pancreatectomy complications \parencite{joglekar_sarcopenia_2015} and shorter long-term survival \parencite{tan_sarcopenia_2009, peng_impact_2012}.
+However, the present study did not identify any relationship between body composition and postoperative CRP levels.
+This may have been due to several reasons including the effect of preoperative obstructive jaundice, hypoalbuminemia as well as cancer related inflammation and body composition changes.
 

### 2016-03-26 Chapters/Pre_Post_SIRS.tex
Layout changes
- Move tables and figures closer to where they are discussed in text
- Remove a few empty lines

@@
 \subsection{Statistics}
-
 Continuous variables are reported as median (inter-quartile range).
@@
 \clearpage
+
 \section{Results}
@@
 
-The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels is shown in Tables \ref{table:sirs_crp} and \ref{table:sirs_crp_pvalues}).
+The relationship between preoperative clinico-pathological characteristics and postoperative C-reactive protein levels is shown in Tables \ref{table:sirs_crp} and \ref{table:sirs_crp_pvalues}.
+
+\input{Tables/sirs_crp}
+\input{Tables/sirs_crp_pvalues}
 
 \subsection{Preoperative vs. postoperative systemic inflammation}
-The relationship between preoperative CRP and postoperative systemic inflammation is depicted graphically in Fig. \ref{fig:sirs_crp_crp} and Fig. \ref{fig:sirs_crp_alb} on page \pageref{fig:sirs_crp}.
+The relationship between preoperative CRP and postoperative systemic inflammation is depicted graphically in Fig. \ref{fig:sirs_crp_crp} and Fig. \ref{fig:sirs_alb_crp} on page \pageref{fig:sirs_crp}.
 The median CRP levels on the day of surgery and the first postoperative day (POD) were significantly higher in patients who had an elevated preoperative CRP. 
@@
 
-\subsection{CPET, comorbidity vs. postoperative inflammation}
-Cardiopulmonary exercise testing was performed in 130 patients. 
-$\dot{V}_{O_2}$AT could not be estimated in one patient.
-$\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
-There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP.% or neutrophil count.
-
-There was no significant, persistent relationship between postoperative CRP and preoperative SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
-
-\subsection{Body composition vs. postoperative inflammation}
-Body composition data was available for 90 patients.
-Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, median 128.9 cm$^2$, IQR 119.0-150.8) and high (n=30, median 241.2 cm$^2$, IQR 221.9-313.1).
-Subcutaneous adipose tissue area was divided into tertiles as low (n=30, median 94.1 cm$^2$, IQR 63.2-102.5), moderate (n=30, median 152.5 cm$^2$, IQR 141.3-175.3) and high (n=30, median 253.9 cm$^2$, IQR 203.2-290.3).
-Skeletal muscle area was also divided into tertiles as low (n=30, median 96.7 cm$^2$, IQR 91.9-102.5), moderate (n=30, median 121.7 cm$^2$, IQR 116.4-130.9) and high (n=30, median 153.0 cm$^2$, IQR 145.3-167.6).
-
-The relationship between body composition and postoperative CRP is shown in Table \ref{table:sirs_bodycomp_crp}.
-Postoperative CRP was not related to any of the components of body composition.
-
-\input{Tables/sirs_crp}
-\input{Tables/sirs_crp_pvalues}
-
-\clearpage
-%========================CRP vs Post-op SIRS============================================
+%======================== Pre-op factors vs Post-op CRP ============================================
 
@@
 
-%==============================================================================
 \clearpage
@@
 %==============================================================================
+
+\subsection{CPET, comorbidity vs. postoperative inflammation}
+Cardiopulmonary exercise testing was performed in 130 patients. 
+$\dot{V}_{O_2}$AT could not be estimated in one patient.
+$\dot{V}_{O_2}$AT was less than 10 ml/kg/min in 52 (40\%) of patients indicating reduced aerobic capacity in these patients.
+There was no significant relationship between $\dot{V}_{O_2}$AT, $\dot{V}_{O_2}$Peak and postoperative CRP.% or neutrophil count.
+
+There was no significant, persistent relationship between postoperative CRP and preoperative SIMD score, body mass index, preoperative haemoglobin levels or the POSSUM Physiology score.
+
+\subsection{Body composition vs. postoperative inflammation}
+Body composition data was available for 90 patients.
+Visceral adipose tissue area (cm$^2$) was divided into tertiles as low (n=30, median 57.1 cm$^2$, IQR 29.0-62.4), moderate (n=30, median 128.9 cm$^2$, IQR 119.0-150.8) and high (n=30, median 241.2 cm$^2$, IQR 221.9-313.1).
+Subcutaneous adipose tissue area was divided into tertiles as low (n=30, median 94.1 cm$^2$, IQR 63.2-102.5), moderate (n=30, median 152.5 cm$^2$, IQR 141.3-175.3) and high (n=30, median 253.9 cm$^2$, IQR 203.2-290.3).
+Skeletal muscle area was also divided into tertiles as low (n=30, median 96.7 cm$^2$, IQR 91.9-102.5), moderate (n=30, median 121.7 cm$^2$, IQR 116.4-130.9) and high (n=30, median 153.0 cm$^2$, IQR 145.3-167.6).
+
+The relationship between body composition and postoperative CRP is shown in Table \ref{table:sirs_bodycomp_crp}.
+Postoperative CRP was not related to any of the components of body composition.
 

### 2016-03-26 Chapters/Pre_Post_SIRS.tex
END OF MAJOR REVISION
- This ends the major revision of this chapter
- Merge this with the main branch
Captions for figures and tables

@@
 \begin{figure}[p]
-	\caption{Relationship between preoperative CRP levels and postoperative CRP in the first week after pancreaticoduodenectomy.}
+	\caption{Relationship between preoperative factors and postoperative CRP in the first week after pancreaticoduodenectomy.}
 	\label{fig:sirs_crp}
@@
 	\end{subfigure}	
+	\caption*{Elevated preoperative CRP was associated with a greater postoperative CRP on the first postoperative day. 
+		Preoperative hypoalbuminemia was associated with significantly lower postoperative CRP levels during the first postoperative week after pancreaticoduodenectomy. 
+		Preoperative obstructive jaundice was also associated with a significantly dampened postoperative systemic inflammatory response.}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=\textwidth]{Figures/sirs_crp_crp_alb}
 	\caption{Relationship between preoperative CRP and postoperative CRP in the presence or absence of preoperative hypoalbuminemia.}
 	\label{fig:sirs_crp_crp_alb}
+	\includegraphics[width=\textwidth]{Figures/sirs_crp_crp_alb}
+	\caption*{Elevated preoperative CRP in the absence of hypoalbuminemia was associated with an exaggerated postoperative inflammatory response (line 1). 
+		However, preoperative hypoalbuminemia resulted in a dampened postoperative inflammatory response regardless of the preoperative CRP (lines 2,3).}
 \end{figure}
@@
 The results also appear to show that preoperative obstructive jaundice is associated with a dampened systemic inflammatory response as measured by CRP levels. 
-%However, neither aerobic capacity or the POSSUM Physiology Score were related to postoperative systemic inflammation. 
 Taken together, these results appear to suggest that the preoperative inflammatory status of the patient plays an important role in modulating postoperative systemic inflammation which may not simply be due to the effect of surgery and its sequelae.

### 2016-03-26 response_to_reviewer’s_comment.tex
Response letter - Pre_Post_SIRS

@@
 
-	Can we talk about this over the phone, please?
+	Point taken. 
+	Chapter 5 has been restructured and simplified as follows. 
+	Postoperative CRP alone has been used as a marker of postoperative systemic inflammation and postoperative albumin and neutrophil counts have been ignored.
+	All tables and figures referring to postoperative albumin and postoperative neutrophil counts have been removed. 
+	The text has been modified to reflect these changes. 
+	All figures and tables have been updated with a brief description of their contents. 
+	This has resulted in a substantially clearer and more understandable chapter that focuses on the relationship between preoperative patient factors and the postoperative CRP response.
 

### 2016-03-26 Chapters/CRP_Complications.tex
CRP_Complications_MAJOR
- Remove figures and tables referring to postop albumin, WCC and neutrophils
- Remove the results section discussing these as well

@@
 %==============================================================================
-\clearpage
 
-\subsection{Postoperative WCC, albumin vs. infective complications}
-
-Patients who developed infective complications in the absence of POPF had a raised total white cell count on postoperative days 5, 6 and 7 which were statistically significant.
-However, these differences were small with wide, overlapping 95\% confidence intervals (Fig. \ref{fig:crp_comp_WCC_infective_leak0}) and not clinically significant.
-
-For instance, the median white cell count on the fifth postoperative day in patients who did not have an infective complication was 8.9 (IQR 6.6 - 11) while in patients who had an infective complication it was 9.8 (IQR 7.9 - 13).
-While this difference was statistically significant (p = 0.028), both medians were within the normal limits for healthy adults. 
-There was no association between white cell counts and infective complications in patients who had a POPF.
-(Table \ref{table:crp_comp_WCC_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_WCC_infective_leak} on page \pageref{fig:crp_comp_WCC_infective_leak})
-
-Similar statistically significant differences were noted in the postoperative neutrophil counts on postoperative days 5, 6 and 7 in patients without a POPF while there was no difference in patients with a POPF (Table \ref{table:crp_comp_Neutrophil_vs_infections_popf_y1n0}).
-However, these differences were associated with a large overlap between the 95\% confidence intervals (error bars in Fig. \ref{fig:crp_comp_Neutrophil_infective_leak0}).
-
-Patients with infective complications in the absence of POPF had a lower median serum albumin  during the entire first postoperative week with statistically significant differences from days 3 through 7.
-There was no association between serum albumin and infective complications in patients who had a POPF.
-(Table \ref{table:crp_comp_Albumin_vs_infections_popf_y1n0} and Fig.
-\ref{fig:crp_comp_Albumin_infective_leak} on page \pageref{fig:crp_comp_Albumin_infective_leak})
-
-%========================WCC vs Infections====================================
-%WCC trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
-\clearpage
-\begin{figure}[t]
-	\caption{Relationship between postoperative white cell count and clinically significant infective complications in patients without (A) and with (B) POPF.}
-	\label{fig:crp_comp_WCC_infective_leak}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_WCC_infective_leak0}
-		\caption{Patients with \textbf{\underline{no}} POPF}
-		\label{fig:crp_comp_WCC_infective_leak0}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_WCC_infective_leak1}
-		\caption{Patients with POPF}
-		\label{fig:crp_comp_WCC_infective_leak1}
-	\end{subfigure}
-\end{figure}
-\vfill
-\input{Tables/crp_comp_WCC_vs_infections_popf_y1n0}
-%==============================================================================
-
-%==========Neutrophil-Neutrophil-Neutrophil-Neutrophil-Neutrophil-Neutrophil=====
-%Neutrophil trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
-\clearpage
-\begin{figure}[t]
-	\caption{Relationship between postoperative neutrophil count and clinically significant infective complications in patients without (A) and with (B) POPF.}
-	\label{fig:crp_comp_Neutrophil_infective_leak}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_Neutrophil_infective_leak0}
-		\caption{Patients with \textbf{\underline{no}} POPF}
-		\label{fig:crp_comp_Neutrophil_infective_leak0}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_Neutrophil_infective_leak1}
-		\caption{Patients with POPF}
-		\label{fig:crp_comp_Neutrophil_infective_leak1}
-	\end{subfigure}
-\end{figure}
-\vfill
-\input{Tables/crp_comp_Neutrophil_vs_infections_popf_y1n0}
-%==============================================================================
-
-%==========Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-Albumin-=====
-%Albumin trends vs infections in patients with or without popf in a 2-panel figure and table showing median, iqr and p-values below it
-\clearpage
-\begin{figure}[t]
-	\caption{Relationship between postoperative serum albumin and clinically significant infective complications in patients without (A) and with (B) POPF.}
-	\label{fig:crp_comp_Albumin_infective_leak}
-	\centering
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_Albumin_infective_leak0}
-		\caption{Patients with \textbf{\underline{no}} POPF}
-		\label{fig:crp_comp_Albumin_infective_leak0}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}{0.48\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{Figures/crp_comp_Albumin_infective_leak1}
-		\caption{Patients with POPF}
-		\label{fig:crp_comp_Albumin_infective_leak1}
-	\end{subfigure}
-\end{figure}
-\vfill
-\input{Tables/crp_comp_Albumin_vs_infections_popf_y1n0}
-%=============================================================================
 \clearpage

### 2016-03-26 Chapters/CRP_Complications.tex
CRP_Complications
- Remove other mentions of albumin, wcc or neutrophil counts

@@
 
-Patients had routine measurements of inflammatory markers including C-reactive protein (CRP) and differential white cell counts from the day before surgery and every day for at least a week after surgery.
+Patients had routine measurements of inflammatory markers including C-reactive protein (CRP) from the day before surgery and every day for at least a week after surgery. 
 These results were collected from the hospital laboratory database using an automated MS Access application as outlined in Appendix \ref{AppendixAccessDatabase}.
@@
 However, there is no useful clinical application for the association between CRP in the early postoperative period and  pancreatic fistula as the diagnosis of POPF is based on the amylase content in surgical drains on the third postoperative day.
-Moreover, early postoperative CRP did not predict the severity of the pancreatic fistula either.
+Moreover, early postoperative CRP did not predict the severity of the pancreatic fistula.
 
@@
 
-The difference in white cell count and neutrophil count was not clinically significant between patients who did or did not have an infective complication. 
+The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation. 
 This finding is similar to one of the earliest reports on the role of serial CRP measurements in predicting septic complications after surgery \parencite{mustard_c-reactive_1987} as well as in several subsequent reports \parencite{matthiessen_increase_2008, welsch_persisting_2008, dutta_persistent_2011}.
-The temporal association between CRP and infective complications may reflect the sensitivity of CRP as a marker of systemic inflammation. 
 This study was not designed to look at inflammatory markers later in the postoperative period. 

### 2016-03-26 response_to_reviewer’s_comment.tex
Response letter - CRP_Complications
- also remove last point re values and ranges needing defining

@@
 
+	Chapter 6 has been revised and simplified as follows.
+	The chapter has been rewritten to focus on the role of postoperative CRP in predicting complications after pancreaticoduodenectomy.
+	All references to postoperative albumin, white cell count and neutrophil counts have been removed.
+	All redundant/irrelevant tables and figures that relate to postoperative white cell count, neutrophil count and albumin have been removed.
+	This has resulted in a substantially clearer and more concise chapter that emphasises the importance of early postoperative CRP in predicting infective complications after pancreaticoduodenectomy.
+
 \textbf{All acronyms/abbreviations on list}
@@
 
-\textbf{All values and ranges need defining.}
-
-	Are they looking for normal ranges for the various parameters like CRP,albumin, etc.?
 \end{document}

### 2016-03-26 Chapters/CPET_Outcomes.tex
Send this to Donny
- Removed todos
- Finalised response letter
- Updated zot.bib

@@
 Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=6) or chronic pancreatitis (n=5).
-\todo{These numbers don't add up - check them.}
+%\todo{These numbers don't add up - check them.}
 
@@
 
-Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.
-\todo{Remove unjustified/unsupported conclusion}
-\todo{Remove GPS conclusion}
+%Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.
+%\todo{Remove unjustified/unsupported conclusion}
+%\todo{Remove GPS conclusion}
 

### 2016-04-16 Chapters/Introduction.tex
Send this to Donny
- Removed todos
- Finalised response letter
- Updated zot.bib
Expanded dicussion of RCTs in pancreatic cancer surgery

@@
 %----------------------------------------------------------------------------------------
-\todo{More discussion of key rcts}
+%\todo{More discussion of key rcts}
 \section{Pancreatic neoplasia}
@@
 
-\todo{Short discussion of etiology}
+%\todo{Short discussion of etiology}
 \subsection{Aetiology of pancreatic cancer}
@@
 Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
+A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
 A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival \parencite{diener_pancreaticoduodenectomy_2008}.
+
 A prospective randomised trial comparing pancreaticogastrostomy versus pancreaticojejunostomy in patients undergoing pancreaticoduodenectomy reported that there was no difference in the incidence of overall complications or pancreatic fistula.
 However, patients who had a pancreaticogastrostomy had a significantly lower rate of biliary fistula, postoperative collections, delayed gastric emptying as well as lower incidence of multiple surgical complications \parencite{bassi_reconstruction_2005}.
+A meta-analysis of eleven studies (one randomised trial, 2 non-randomised prospective trials and eight observational studies) appeared to show improved outcomes in patients undergoing pancreaticogastrostomy rather than a pancreaticojejunostomy \parencite{mckay_meta-analysis_2006}.
+However, an earlier trial by Yeo and co-workers failed to demonstrate any difference in outcomes between either technique \parencite{yeo_prospective_1995}.
+
+In spite of clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
+While some early randomised trials reported the benefits of octreotide in decreasing the risk of postoperative pancreatic fistula \parencite{montorsi_efficacy_1995,nakatsuka_octreotide_2000}, other more recent trials have not reproduced these results \parencite{lowy_prospective_1997, yeo_does_2000, kollmar_prophylactic_2008}.
+
 Surgical drains after pancreaticoduodenectomy have been a subject of considerable debate. 
-A randomised trial comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications.
-Hospital stay and costs were also lower \parencite{bassi_early_2010}.
+A prospective randomised controlled trial of 179 patients in 2001 reported that intra-abdominal drains placed at the time of surgery did not reduce the incidence of complications \parencite{conlon_prospective_2001}.
+
+Two randomised trials comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications \parencite{kawai_early_2006,bassi_early_2010}.
 

### 2016-04-16 Chapters/Introduction.tex
Expanded aetiology

@@
 The most important risk factor associated with increased incidence of pancreatic cancer is tobacco. 
-Obesity has also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms are unclear.
-Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes.
+Tobacco smoking has been reported to be associated with at least a 2-fold increased risk of pancreatic cancer and this risk increases to 5-fold in patients with over 40 pack-years of smoking \parencite{raimondi_early_2007, iodice_tobacco_2008}.
+This increased risk persists for at least 10 years after cessation of smoking \parencite{iodice_tobacco_2008}.
+
+The association between alcohol consumption and pancreatic cancer is less clear. 
+While alcohol consumption on its own has not been shown to result in an increased risk of pancreatic cancer, the incidence of pancreatic cancer in heavy drinkers is greater than in the general population.
+This may be due to the confounding effect of cigarette smoking in this cohort of patients \parencite{jiao_alcohol_2009, rohrmann_ethanol_2009}.
+
+Obesity and high caloric intake have also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms and the effects of confounding factors are poorly understood \parencite{berrington_de_gonzalez_meta-analysis_2003, larsson_body_2007, li_body_2009}.
+
+Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes \parencite{raimondi_epidemiology_2009, maisonneuve_epidemiology_2010}.
 It is of note that some of the important risk factors such as smoking, obesity, poor diet and increasing age may also be associated with other medical comorbidities including cardiorespiratory diseases that may have a significant impact on patient fitness.

### 2016-04-16 Chapters/BodyComp.tex
Citation edit for ATS/ACCP across multiple chapters

@@
 This assertion is repeated in their later work on 548 patients which also showed a clear association between low $\dot{V}_{O_2}$AT (ml/kg/min) and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}.
-The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.
+The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{_ats/accp_2003}.
 

### 2016-04-16 Chapters/CPET_Jaundice.tex
Citation edit for ATS/ACCP across multiple chapters

@@
 Peak oxygen consumption achieved at this stage was defined as $\dot{V}_{O_2}$Peak. 
-The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{society_ats/accp_2003} methods.
+The $\dot{V}_{O_2}$AT was calculated using the V-slope \parencite{beaver_new_1986,sue_metabolic_1988} and ventilatory equivalents \parencite{_ats/accp_2003} methods.
 $\dot{V}_{O_2}$AT less than 10 ml/kg/min was considered to be low based on previous work by us (Chapter \ref{ch_cpet_outcomes}) as well as Ausania and co-workers \parencite{ausania_effects_2012} which has shown increased incidence of complications in patients with $\dot{V}_{O_2}$AT below this threshold.

### 2016-04-16 Chapters/Introduction.tex
Citation edit for ATS/ACCP across multiple chapters

@@
 Most centres, like ours, use both methods supplemented by information from a variety of other parameters to enable accurate determination of the anaerobic threshold.
-This approach is also recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{society_ats/accp_2003}.
+This approach is also recommended by the American Thoracic Society/American College of Chest Physicians Statement on cardiopulmonary exercise testing \parencite{_ats/accp_2003}.
 

### 2016-04-16 response_letter.tex
New response letter
- As a chapter in thesis
- Allows bibliography for letter

--- 2016-03-26 response_to_reviewer’s_comment.tex
@@
-\documentclass[12pt,a4paper]{article}
-\usepackage[utf8]{inputenc}
-%\usepackage{amsmath}
-%\usepackage{amsfonts}
-%\usepackage{amssymb}
-%\usepackage{graphicx}
-\author{Vishnu V Chandrabalan}
-\begin{document}
+% Chapter 01 - Introduction
 
-\textbf{Response to Reviewer’s comments}
+\chapter*{Response to examiners' comments}
+
+\lhead{\emph{Response to examiners' comments}} % This is for the header on each page - perhaps a shortened title
+
+\bigskip
 
 \textbf{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.}
+
+\bigskip
+
+\textbf{Mr. Vishnu V Chandrabalan}
+
+\textbf{University of Department of Surgery}
+
+\textbf{University of Glasgow}
+
+\clearpage
+%----------------------------------------------------------------------------------------
 
@@
 I am grateful for the opportunity to respond to the examiners' comments. 
-I have considered the examiners' comments carefully and these have improved the thesis. 
+I have considered the examiners' comments carefully and the resulting changes have improved the thesis. 
 Please find my responses to the comments below.
@@
 	Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
-	A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival.
+	A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
+	A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival \parencite{diener_pancreaticoduodenectomy_2008}.
+
 	A prospective randomised trial comparing pancreaticogastrostomy versus pancreaticojejunostomy in patients undergoing pancreaticoduodenectomy reported that there was no difference in the incidence of overall complications or pancreatic fistula.
-	However, patients who had a pancreaticogastrostomy had a significantly lower rate of biliary fistula, postoperative collections, delayed gastric emptying as well as lower incidence of multiple surgical complications.
+	However, patients who had a pancreaticogastrostomy had a significantly lower rate of biliary fistula, postoperative collections, delayed gastric emptying as well as lower incidence of multiple surgical complications \parencite{bassi_reconstruction_2005}.
+	A meta-analysis of eleven studies (one randomised trial, 2 non-randomised prospective trials and eight observational studies) appeared to show improved outcomes in patients undergoing pancreaticogastrostomy rather than a pancreaticojejunostomy \parencite{mckay_meta-analysis_2006}.
+	However, an earlier trial by Yeo and co-workers failed to demonstrate any difference in outcomes between either technique \parencite{yeo_prospective_1995}.
+
+	In spite of clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
+	While some early randomised trials reported the benefits of octreotide in decreasing the risk of postoperative pancreatic fistula \parencite{montorsi_efficacy_1995,nakatsuka_octreotide_2000}, other more recent trials have not reproduced these results \parencite{lowy_prospective_1997, yeo_does_2000, kollmar_prophylactic_2008}.
+
 	Surgical drains after pancreaticoduodenectomy have been a subject of considerable debate. 
-	A randomised trial comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications.
-	Hospital stay and costs were also lower.
+	A prospective randomised controlled trial of 179 patients in 2001 reported that intra-abdominal drains placed at the time of surgery did not reduce the incidence of complications \parencite{conlon_prospective_2001}.
+
+	Two randomised trials comparing early versus delayed drain removal in patients with low risk of pancreatic fistula after pancreaticoduodenectomy reported that early drain removal was associated with lower incidence of pancreatic fistula, abdominal complications and pulmonary complications \parencite{kawai_early_2006,bassi_early_2010}.
+
+	However, these pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.
 
@@
 	The most important risk factor associated with increased incidence of pancreatic cancer is tobacco. 
-	Obesity has also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms are unclear. 
-	Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes. 
+	Tobacco smoking has been reported to be associated with at least a 2-fold increased risk of pancreatic cancer and this risk increases to 5-fold in patients with over 40 pack-years of smoking \parencite{raimondi_early_2007, iodice_tobacco_2008}.
+	This increased risk persists for at least 10 years after cessation of smoking \parencite{iodice_tobacco_2008}.
+
+	The association between alcohol consumption and pancreatic cancer is less clear. 
+	While alcohol consumption on its own has not been shown to result in an increased risk of pancreatic cancer, the incidence of pancreatic cancer in heavy drinkers is greater than in the general population.
+	This may be due to the confounding effect of cigarette smoking in this cohort of patients \parencite{jiao_alcohol_2009, rohrmann_ethanol_2009}.
+
+	Obesity and high caloric intake have also been reported to be associated with increased risk of pancreatic cancer although the underlying mechanisms and the effects of confounding factors are poorly understood \parencite{berrington_de_gonzalez_meta-analysis_2003, larsson_body_2007, li_body_2009}.
+
+	Other possible risk factors include increasing age, diet high in saturated fats, chronic pancreatitis, family history of pancreatic cancer and genetic abnormalities resulting in cancer syndromes \parencite{raimondi_epidemiology_2009, maisonneuve_epidemiology_2010}.
 	It is of note that some of the important risk factors such as smoking, obesity, poor diet and increasing age may also be associated with other medical comorbidities including cardiorespiratory diseases that may have a significant impact on patient fitness.
@@
 
-	Point taken. The following sentence has been removed from the discussion in Chapter 2.
+	Point taken. The following sentence has been \textit{removed} from the discussion in Chapter 2.
 
@@
 
-	Point taken. The following sentence has been removed from the discussion in Chapter 3.
+	Point taken. The following sentence has been \textit{removed} from the discussion in Chapter 3.
 
@@
 	All tables have been updated to include units. 
-	Legends have been added to all figures and tables providing a brief description of the results
-	depicted.
-
-\end{document}
+	Legends have been added to all figures and tables providing a brief description of the results and the statistical tests used.

### 2016-04-16 Chapters/Introduction.tex
Introduction - Expanded RCT of adjuvant/neoadjuvant rx

@@
 While surgery is the primary modality of curative treatment in patients with resectable pancreatic cancer, there is clear evidence that adjuvant chemotherapy improves survival and is now the standard of care for patients with pancreatic ductal adenocarcinoma.
+
+Early randomised trials reported improved survival in patients with inoperable pancreatic cancer \parencite{mallinson_chemotherapy_1980} as well as in patients who had undergone potentially curative surgery \parencite{bakkevold_adjuvant_1993}.
+
 The European Study Group for Pancreatic Cancer (ESPAC) has conducted some of the largest randomised controlled trials of adjuvant chemotherapy and chemoradiotherapy versus observation after potentially curative surgery.
@@
 A number of factors have been identified for the non-receipt of adjuvant therapy including comorbidity, postoperative complications and favourable pathological features \parencite{spitz_preoperative_1997, aloia_delayed_2007, oettle_adjuvant_2007, merchant_adjuvant_2009, russ_impact_2009}.
+
+Increasingly, neo-adjuvant therapies are being used in patients with initially unresectable or 'borderline-resectable' cancers in an attempt to improve resectability. 
+In a meta-analysis of 111 studies, Gillen and co-workers reported that approximately one-third of initially unresectable patients can be expected to have resectable tumours after neo-adjuvant treatment \parencite{gillen_preoperative/neoadjuvant_2010}.
+The authors recommended that such patients should be actively included in neo-adjuvant protocols and re-staged for resection after treatment.
 

### 2016-04-16 response_letter.tex
Update response letter
- Also new version of zot.bib

@@
 
-\chapter*{Response to examiners' comments}
+\chapter*{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.}
 
@@
 
-\bigskip
+\vfill
 
-\textbf{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.}
+\hrule
+\vfill
+{\LARGE \textbf{Response to examiners' comments}}
+\vfill
+\hrule
 
-\bigskip
+\vfill
 
@@
 
+\vfill
+
 \clearpage
 %----------------------------------------------------------------------------------------
-
 Dear Sir,
@@
 
-	In spite of clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
+	In spite of a lack of any clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
 	While some early randomised trials reported the benefits of octreotide in decreasing the risk of postoperative pancreatic fistula \parencite{montorsi_efficacy_1995,nakatsuka_octreotide_2000}, other more recent trials have not reproduced these results \parencite{lowy_prospective_1997, yeo_does_2000, kollmar_prophylactic_2008}.
@@
 
-	Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery verus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma.
+	Early randomised trials reported improved survival in patients with inoperable pancreatic cancer \parencite{mallinson_chemotherapy_1980} as well as in patients who had undergone potentially curative surgery \parencite{bakkevold_adjuvant_1993}.
+
+	Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery verus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma \parencite{oettle_adjuvant_2007}.
 	The median disease free survival in patient who received gemcitabine chemotherapy was 13.4 months while it was only 6.9 months in the surgery-only group. 
 	However, there was no difference in overall survival.
+
+	Increasingly, neo-adjuvant therapies are being used in patients with initially unresectable or 'borderline-resectable' cancers in an attempt to improve resectability. 
+	In a meta-analysis of 111 studies, Gillen and co-workers reported that approximately one-third of initially unresectable patients can be expected to have resectable tumours after neo-adjuvant treatment \parencite{gillen_preoperative/neoadjuvant_2010}.
+	The authors recommended that such patients should be actively included in neo-adjuvant protocols and re-staged for resection after treatment.
 

### 2016-06-05 Chapters/Introduction.tex
Spellcheck - Introduction

@@
 Amylase content in the drain fluid was measured on the third post-operative day to aid in the diagnosis of post-operative pancreatic fistula. 
-Surgical drains and naso-gastric/naso-jeujunal tubes were removed at the surgeon's discretion. 
+Surgical drains and naso-gastric/naso-jejunal tubes were removed at the surgeon's discretion. 
 Patients were discharged when medically fit.
@@
 
-Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery verus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma \parencite{oettle_adjuvant_2007}.
+Oettle and co-workers reported on the results of a European multi-center, randomised controlled phase 3 trial comparing the results of surgery versus surgery and adjuvant chemotherapy with 6 cycles of gemcitabine in patients with pancreatic ductal adenocarcinoma \parencite{oettle_adjuvant_2007}.
 The median disease free survival in patient who received gemcitabine chemotherapy was 13.4 months while it was only 6.9 months in the surgery-only group.
@@
 Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
-A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
+A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any significant difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
 A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival \parencite{diener_pancreaticoduodenectomy_2008}.
@@
 
-In spite of clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
+In spite of a lack of any clear evidence of efficacy, octreotide continues to be used in many centres including ours to reduce the incidence of postoperative pancreatic fistula. 
 While some early randomised trials reported the benefits of octreotide in decreasing the risk of postoperative pancreatic fistula \parencite{montorsi_efficacy_1995,nakatsuka_octreotide_2000}, other more recent trials have not reproduced these results \parencite{lowy_prospective_1997, yeo_does_2000, kollmar_prophylactic_2008}.
@@
 $\dot{V}_{O_2}$ or oxygen uptake is measured breath-by-breath using digital analysis of the inspired and expired gases. 
-This is then averaged, usually over time, to smooth-out any signficant breath-by-breath variation.
+This is then averaged, usually over time, to smooth-out any significant breath-by-breath variation.
 $\dot{V}_{O_2}$ increases with increasing work load and is influenced by several factors that have a role in the transport and utilisation of oxygen.
@@
 They performed choledocho-caval anastomoses in 5 dogs and studied cardiac function before and 2 weeks after this procedure. 
-They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimmetic agents. 
+They reported that 'cholemia' was associated with impaired left ventricular function and blunted response to sympathomimetic agents. 
 Similar findings have been reported by other authors in animal studies \parencite{binah_obstructive_1985,bomzon_systemic_1986}.
@@
 
-More recently, TNF-$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF-$\alpha$ agent restored myocardial contractility \parencite{yang_mechanisms_2010}.
+More recently, TNF$\alpha$ levels have been reported to mediate cardiac dysfunction in animal studies and treatment with an anti-TNF$\alpha$ agent restored myocardial contractility \parencite{yang_mechanisms_2010}.
 Obstructive jaundice was also associated with systemic hypotension and there is increasing evidence that some of this may be mediated by bile acid receptors on the vascular tree \parencite{green_systemic_1995, lefebvre_role_2009}.
@@
 
-A recent randomised controlled trial of 202 patients reported that there was no significant difference in post-operative morbidity (47\% vs, 37\%, p=0.14), mortality or length of stay between the group that underwent preoperative biliary drainage and the group that underwent early surgery \parencite{van_der_gaag_preoperative_2010}.
+A recent randomised controlled trial of 202 patients reported that there was no significant difference in post-operative morbidity (47\% vs 37\%, p=0.14), mortality or length of stay between the group that underwent preoperative biliary drainage and the group that underwent early surgery \parencite{van_der_gaag_preoperative_2010}.
 There was a high incidence of procedure related complications in the drainage group (46\%). 
@@
 
-The paucity of good quality evidence on the role of preoperative biliary drainage before pancreatic resection \parencite{wang_preoperative_2008} and the limitations in our understanding of the complex patho-physiological effects of jaundice on multiple organ systems has resulted in persistent wide variation in the perioperative management of jaundiced patients across different centres.
+The paucity of good quality evidence on the role of preoperative biliary drainage before pancreatic resection \parencite{wang_preoperative_2008} and the limitations in our understanding of the complex pathophysiological effects of jaundice on multiple organ systems has resulted in persistent wide variation in the perioperative management of jaundiced patients across different centres.
 
@@
 It has been a commonly held belief that obese patients had greater incidence of complications. 
-Some of the adverse outcomes such as wound complications and respiratory dysfunction where considered to be the direct result of the increased body mass index while other complications where related to the associated comorbidities such as cardiovascular disease, obstructive sleep apnea or diabetes mellitus.
+Some of the adverse outcomes such as wound complications and respiratory dysfunction where considered to be the direct result of the increased body mass index while other complications where related to the associated comorbidities such as cardiovascular disease, obstructive sleep apnoea or diabetes mellitus.
 
@@
 
-Tsai and co-workers reported that while obese patients had similar tumour characteristics to non-obese patients and comparable perioperative outomes, long-term survival was better in obese subjects \parencite{tsai_impact_2010}.
+Tsai and co-workers reported that while obese patients had similar tumour characteristics to non-obese patients and comparable perioperative outcomes, long-term survival was better in obese subjects \parencite{tsai_impact_2010}.
 This study of 795 patients undergoing potentially curative surgery for pancreatic adenocarcinoma at the Johns Hopkins Medical Institutions in the United States is in contrast to other similar studies.

### 2016-06-05 Appendices/AppendixAccessDatabase.tex
Grammar correction in Appendix 1

@@
 
-The RealLink for Windows (RFW, http://www.nps-reality.com/reallink) application used at Glasgow Royal Infirmary allowed access form VBA via Dynamic Data Exchange.
-This was used to collect and parse laboratory results for preoperative and postoperative blood results efficiently and without error.
+The RealLink for Windows (RFW, http://www.nps-reality.com/reallink) application used at Glasgow Royal Infirmary allowed access from VBA via Dynamic Data Exchange.
+This was used to collect and parse laboratory data for preoperative and postoperative blood tests efficiently and without error.
 

### 2016-06-17 Presentations/viva_presentation_short.tex

@@
 	\framesubtitle{What, how, why?}
+	{\footnotesize
 	\begin{itemize}
-
 		\item Objective measure of cardiopulmonary response to exercise
@@
 		\item Widely used in general, oesophago-gastric, vascular and thoracic surgery
-		\item CPET Methodology
-		\begin{itemize}
-			\item Cycle ergometer with electronic braking
-			\item Breath-by-breath gas analysis using a tight-fitting face mask
-			\item Continuous monitoring of 12-lead ECG
-			\item Initial 3-minute rest period on bike for baseline parameters
-			\item Incremental work-load test to volitional fatigue
 	\end{itemize}
-	\end{itemize}
+	\vfill
+
+	\begin{columns}
+		\column[c]{0.6\textwidth}
+			\textbf{CPET Methodology}
+			\\ - Cycle ergometer with electronic braking
+			\\ - Breath-by-breath gas analysis
+			\\ - 12-lead ECG monitoring
+			\\ - Initial 3-minute rest for baseline
+			\\ - Incremental load to volitional fatigue
+
+		\column[c]{0.4\textwidth}
+
+			\includegraphics[height=0.4\textheight]{../Figures/cpet-zan}
+
+	\end{columns}
+	}
 \end{frame}
@@
 		\item To evaluate the clinical utility of preoperative CPET in predicting postoperative adverse events after pancreaticoduodenectomy.
+\pause
 		\item To examine the patient factors that are related to cardiopulmonary exercise physiology with particular attention to the effect of obstructive jaundice and body composition.
+\pause
 		\item To examine the effect of preoperative systemic inflammation and poor aerobic capacity on the magnitude of the post-operative systemic inflammatory response.
+\pause
 		\item To examine the value of serial daily postoperative markers of systemic inflammatory response in the prediction of post-operative complications.
@@
 	\end{block}
-
+\pause	
 	\begin{block}{\textbf{Length of stay}}
@@
 	\end{block}
-
+\pause	
 	\begin{block}{\textbf{Adjuvant Therapy}}
@@
 		\column{0.6\textwidth}
+\pause
 			{\footnotesize 	Why is $\dot{V}_{O_2}$AT so low in patients undergoing pancreatic surgery?\\
 				\medskip
+
 				What factors are associated with a low $\dot{V}_{O_2}$AT?\\
 				\medskip
+\pause
 				Why was $\dot{V}_{O_2}$AT not associated with cardio-respiratory complications?\\
 				\medskip
+
 				Does $\dot{V}_{O_2}$AT truly reflect cardiopulmonary function in this cohort of patients?\\	
@@
 \section[Chapter 3]{CPET, Jaundice and Preoperative Pathophysiology}
-
-\begin{frame}
-	\frametitle{Jaundice and other patient characteristics}
-	\framesubtitle{What was the relationship between obstructive jaundice and other parameters?}
-
-	\begin{columns}
-		\column[c]{0.5\textwidth}
-
-		\begin{figure}
-			\includegraphics[width=\textwidth]{jaundice_distribution}
-		\end{figure}
-
-		\begin{figure}
-			\includegraphics[width=\textwidth]{jaundice_vs_patient_factors}
-		\end{figure}
-
-		\column[c]{0.5\textwidth}		
-		{\scriptsize			
-			Severity of systemic inflammation proportionate to severity of jaundice. \\
-			\medskip
-			Obstructive jaundice associated with electrolyte abnormalities.\\
-			\medskip
-			Jaundiced patients were more likely to be anaemic, with lower haematocrit and lower mean corpuscular volume.\\
-			\medskip
-			No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
-		}
-	\end{columns}
-\end{frame}
 
@@
 \begin{frame}
+	\frametitle{Jaundice and other patient characteristics}
+	\framesubtitle{What was the relationship between obstructive jaundice and other parameters?}
+
+	\begin{columns}
+		\column[c]{0.5\textwidth}
+
+		\begin{figure}
+			\includegraphics[width=\textwidth]{jaundice_distribution}
+		\end{figure}
+
+		\begin{figure}
+			\includegraphics[width=\textwidth]{jaundice_vs_patient_factors}
+		\end{figure}
+		\pause		
+		\column[c]{0.5\textwidth}		
+		{\scriptsize			
+			Severity of systemic inflammation proportionate to severity of jaundice. \\
+			\medskip
+			Obstructive jaundice associated with electrolyte abnormalities.\\
+			\medskip
+			Jaundiced patients were more likely to be anaemic, with lower haematocrit and lower mean corpuscular volume.\\
+			\medskip
+			No significant difference in the preoperative renal function, prothrombin time and white cell count between jaundiced and non-jaundiced patients.\\
+		}
+	\end{columns}
+\end{frame}
+
+\begin{frame}
 	\frametitle{What preop factors were associated with low $\dot{V}_{O_2}$?}
@@
 	\hrule
+\pause
 	\medskip
@@
 	\vfill
-
+\pause	
 	For example, if the absolute $\dot{V}_{O_2}$AT was 0.84 litres/min, the corrected $\dot{V}_{O_2}$AT (ml/kg/min) would vary with patient weight as follows:
@@
 \begin{frame}
+	\frametitle{$\dot{V}_{O_2}$AT vs Total Adipose Tissue}
+	\framesubtitle{Figure 4.4(B) Page 111, Figure 4.5(B) on Page 114}
+
+	\begin{figure}
+		\includegraphics[height=0.4\textheight]{../Figures/bc_scatter_VO2_TAT}
+	\end{figure}
+
+	\begin{figure}
+		\includegraphics[height=0.45\textheight]{../Figures/bc_scatter_VO2_TAT_elbm}
+	\end{figure}
+
+\end{frame}
+
+\begin{frame}
 	\frametitle{$\dot{V}_{O_2}$AT vs Skeletal Muscle}
@@
 
-\begin{frame}
-	\frametitle{$\dot{V}_{O_2}$AT vs Total Adipose Tissue}
-	\framesubtitle{Figure 4.4(B) Page 111, Figure 4.5(B) on Page 114}
-
-	\begin{figure}
-		\includegraphics[height=0.4\textheight]{../Figures/bc_scatter_VO2_TAT}
-	\end{figure}
-	\begin{figure}
-		\includegraphics[height=0.45\textheight]{../Figures/bc_scatter_VO2_TAT_elbm}
-	\end{figure}
-
-\end{frame}
-
 \section[Chapter 5]{Factors affecting post op inflammation}
@@
 	\begin{itemize}
-		\item Perioperative systemic inflammation influences short and long-term outcomes
+		\item Perioperative systemic inflammation affects short and long-term outcomes
 		\item Elevated preoperative systemic inflammation is associated with postoperative complications in colorectal surgery, after oesophagectomy, liver resections and cardiac surgery. 
@@
 		\item Imaging: Resectable tumour of head of pancreas
-		\medskip
-
+		\medskip \hrule \medskip
+\pause
 		\item Should this patient undergo CPET?
@@
 	{\scriptsize 	
-		\item [Ch 2] CPET may allow identifying if she is at increased risk of complications, prolonged hospital stay and non-receipt of adjuvant therapy. 
+		\item [Chapter 2] CPET may allow identifying if she is at increased risk of complications, prolonged hospital stay and non-receipt of adjuvant therapy. 
 					\\ CPET should not be used on its own to decline surgery. 
 					\hrule
-		\item [Ch 3] CPET can be performed in the presence of jaundice. 
+\pause
+		\item [Chapter 3] CPET can be performed in the presence of jaundice. 
 					\\ Jaundice does not adversely affect dynamic cardiopulmonary response to and this cannot be a justification for biliary drainage. 
@@
 						\hrule
-		\item [Ch 4] Interpret low $\dot{V}_{O_2}$ in her case with caution. 
-					\\ Her aerobic capacity may be better than that suggested by her test results. 
+\pause
+		\item [Chapter 4] Interpret low $\dot{V}_{O_2}$ in her case with caution. 
+					\\ Her aerobic capacity may be better than that suggested by her results. 
 					\\ Beware of \textit{spurious correlation} with total body weight.
 						\hrule
-		\item [Ch 5] Jaundice and hypo-albuminemia attenuate post-op inflammatory response.
+\pause
+		\item [Chapter 5] Jaundice and hypo-albuminemia attenuate post-op inflammatory response.
 					\\ However, this does not translate into increased complications. 
@@
 						\hrule
-		\item [Ch 6] If her drain amylase on D3 is normal and D3 CRP$<$178 or D4 CRP$<$125, she is likely to have an uneventful postoperative course and must continue on 'enhanced-recovery' pathway.
+\pause
+		\item [Chapter 6] If her drain amylase on D3 is normal and D3 CRP$<$178 or D4 CRP$<$125, she is likely to have an uneventful postoperative course and must continue on 'enhanced-recovery' pathway.
 				\\ If she develops a POPF, the early inflammatory response does not predict outcome.

### 2016-06-17 response_letter.tex

@@
 	Point taken.	
-	The following paragraph has been included under Section 1.4.4 in Chapter 1 and discusses some of the key randomised trials of operative strategies aimed at reducing postoperative complications after pancreaticoduodenectomy.
+	The following paragraphs have been included under Section 1.4.4 in Chapter 1 and discusses some of the key randomised trials of operative strategies aimed at reducing postoperative complications after pancreaticoduodenectomy.
 
+	\textquotedblleft
 	Several studies have attempted to evaluate various surgical strategies in improving perioperative outcomes after pancreaticoduodenectomy. 
-	A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
+	A prospective, multi-center, randomised controlled trial of 170 patients undergoing pancreaticoduodenectomy did not demonstrate any significant difference between a Whipple's classical pancreaticoduodenectomy  and pylorus preserving pancreaticoduodenectomy \parencite{tran_pylorus_2004}.
 	A Cochrane review in 2008 reported on 7 randomised controlled trials comparing classical pancreaticoduodenectomy with pylorus-preserving pancreaticoduodenectomy  and found no difference in morbidity, mortality or survival \parencite{diener_pancreaticoduodenectomy_2008}.
@@
 	However, these pancreas-specific risk factors are not modifiable before surgery, have been studied extensively before and are not the subject of this thesis.
+	\textquotedblright
 
+	\textquotedblleft		
 	The following has been added to Section 1.3 in Chapter 1 and adds to the already detailed discussion of some of the key randomised trials in adjuvant therapy after pancreaticoduodenectomy.
@@
 	The authors recommended that such patients should be actively included in neo-adjuvant protocols and re-staged for resection after treatment.
+	\textquotedblright
 
@@
 
+	\textquotedblleft
 	Several factors have been identified as being associated with an increased risk of pancreatic cancer.
@@
 	It is of note that some of the important risk factors such as smoking, obesity, poor diet and increasing age may also be associated with other medical comorbidities including cardiorespiratory diseases that may have a significant impact on patient fitness.
+	\textquotedblright
 
@@
 
+	\textquotedblleft
 	Pathological examination of the resected specimen showed pancreatic ductal adenocarcinoma (n=37), ampullary adenocarcinoma (n=18), cholangiocarcinoma (n=17), duodenal adenocarcinoma (n=6), intra-ductal papillary mucinous neoplasia (n=4), neuroendocrine tumours (n=7), other neoplasia (n=6) or chronic pancreatitis (n=5).
+	\textquotedblright
 
@@
 
+	\textquotedblleft
 	The aim of the present study was to evaluate the role of various measures of patient physiological fitness including cardiopulmonary exercise testing in predicting postoperative length of stay, major postoperative adverse events including operative mortality and fitness to undergo adjuvant therapy when indicated after pancreaticoduodenectomy.
+	\textquotedblright
 
@@
 
-	Point taken. The following sentence has been \textit{removed} from the discussion in Chapter 2.
+	Point taken. The following sentence has been \underline{\textit{removed}} from the discussion in Chapter 2.
 
+	\textquotedblleft
 	Therefore, it would appear that objective measurement of patient physiological fitness using cardiopulmonary exercise testing is superior to conventional measures of patient fitness including the POSSUM Physiology Score or the modified Glasgow Prognostic Score and may have a role in predicting short-term outcome which in turn affects the overall management of these patients including receipt of adjuvant therapy.
+	\textquotedblright
 
@@
 
-	Point taken. The following sentence has been \textit{removed} from the discussion in Chapter 3.
+	Point taken. The following sentence has been \underline{\textit{removed}} from the discussion in Chapter 3.
 
+	\textquotedblleft
 	Taken together, these results appear to suggest that preoperative biliary drainage and resolution of jaundice are unlikely to improve aerobic capacity and cardiopulmonary physiology in these patients.
+	\textquotedblright
 

### 2016-06-17 response_to_reviewer’s_comment.tex

@@
 \usepackage[utf8]{inputenc}
-%\usepackage{amsmath}
-%\usepackage{amsfonts}
-%\usepackage{amssymb}
-%\usepackage{graphicx}
+\usepackage[left=2.00cm, right=2.00cm, top=2.00cm, bottom=2.00cm]{geometry}
+\usepackage{setspace}
+\usepackage{parskip}
 \author{Vishnu V Chandrabalan}
 \begin{document}
+\doublespacing
 
-\textbf{Response to Reviewer’s comments}
+\textbf{Response to examiners' comments}
+
+\bigskip
 
 \textbf{An investigation of the clinical utility of preoperative cardiopulmonary exercise testing in patients undergoing major pancreatic surgery.}
+
+\bigskip
 
@@
 	All tables have been updated to include units. 
-	Legends have been added to all figures and tables providing a brief description of the results
-	depicted.
+	Legends have been added to all figures and tables providing a brief description of the results and the statistical tests used.
 

### 2016-06-17 Chapters/BodyComp.tex

@@
 Detailed breath-by-breath data on a variety of physiological and gas-exchange parameters measured at cardiopulmonary exercise testing were also collected from a prospectively maintained database. 
-Cardiopulmonary exercise testing methodolgy is discussed in Section \ref{sec:cpx_method} (page \pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (page \pageref{sec:cpx_parameters}).
+Cardiopulmonary exercise testing methodology is discussed in Section \ref{sec:cpx_method} (page \pageref{sec:cpx_parameters})and a description of the parameters measured at cardiopulmonary exercise testing is provided in Section \ref{sec:cpx_parameters} (page \pageref{sec:cpx_parameters}).
 A complete dataset for one sample patient is shown in Appendix \ref{AppendixCPETRawData} on page \pageref{AppendixCPETRawData}.
@@
 
-Swart and Carlisle als0 reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. 
+Swart and Carlisle also reported that $\dot{V}_{O_2}$AT $<$ 11 ml/kg/min in patients undergoing open colorectal surgery was associated with adverse outcomes \parencite{swart_case-controlled_2012}. 
 However, the proportion of females in the low $\dot{V}_{O_2}$AT group was significantly greater than that in the normal $\dot{V}_{O_2}$AT group (24\% vs 51\%). 

### 2016-06-17 Chapters/CPET_Outcomes.tex

@@
 While postoperative mortality after pancreatic surgery has steadily improved over the years with major improvements in the quality of surgical and critical care over the past decade \parencite{winter_1423_2006} even in elderly patients \parencite{makary_pancreaticoduodenectomy_2006}, postoperative morbidity remains high \parencite{mann_review_2010}.
-The results of this study show that poor preoperative fitnees is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.
+The results of this study show that poor preoperative fitness is not only associated with a protracted protracted postoperative course with complications but also with non-receipt of adjuvant therapy.
 

### 2016-06-17 Chapters/Pre_Post_SIRS.tex

@@
 Preoperative hypoalbuminemia on the other hand was associated with lower median postoperative CRP levels starting with POD 3. 
-The median difference in postoperative CRP between patients with or without preoperative hypoalbuminemia was 30 mg/l on POD 3 (p$<$0.045) and 53 mg/l on POD 6 (p=0.002) with lower CRP levels in patients with hypoalbuminemia.
+The median difference in postoperative CRP between patients with or without preoperative hypoalbuminemia was 30 g/l on POD 3 (p$<$0.045) and 53 g/l on POD 6 (p=0.002) with lower CRP levels in patients with hypoalbuminemia.
 This relationship is shown in Fig. \ref{fig:sirs_alb_crp}.
